---
title: "Bipolar Disorder & Suicidality Risk Variable Data Wrangling"
author: "Sam A. Sievertsen"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(warning = FALSE, message = FALSE, comment = "")

```

```{r environment, echo = FALSE, include = FALSE, warning = FALSE}

# Load necessary packages + environmental variables
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(skimr)
options(scipen = 999, digits = 8)

# Read in raw data generated in DEAP
raw_data <- read.csv("../../../data/data_raw/dataset.csv")

# Load the shared clustering feature definitions (single source of truth for the complete-case sample)
source("../1_clustering/cluster_configs.R")

# Fail early and informatively if any raw column required by the revision derivations is absent
required_raw_cols <- c(
  "mh_p_cbcl__synd__anxdep_sum", "mh_p_cbcl__synd__anxdep_nm",
  "mh_p_cbcl__synd__attn_sum", "mh_p_cbcl__synd__aggr_sum",
  "mh_y_bisbas__bas__rr_sum__v01", "mh_y_bisbas__bas__dr_sum",
  "mh_y_bisbas__bis_sum__v01", "mh_y_bisbas__bas__fs_sum",
  "ph_p_sds__dims_sum", "ph_p_sds__sbd_sum", "ph_p_sds__da_sum",
  "ph_p_sds__swtd_sum", "ph_p_sds__does_sum", "ph_p_sds__hyphy_sum",
  "mh_p_famhx__halluc_001", "mh_p_famhx__hosp_001",
  "fc_y_fes__confl_005", "fc_p_fes__confl_005"
)
missing_raw_cols <- setdiff(required_raw_cols, names(raw_data))
if (length(missing_raw_cols) > 0) {
  stop("dataset.csv is missing required columns: ", paste(missing_raw_cols, collapse = ", "))
}

```

## Risk Variable Data Wrangling

This section completes the following data wrangling objectives:

  1. Filtering just the baseline assessment timepoint data for use in baseline risk clustering
  
  2. Creating variables from raw ABCD data to be used in clustering
  
  3. Retaining columns of interest for validation & clustering

```{r data wrangling, warning = FALSE}

## Data Wrangling ##

#1. Filter the trimmed data to only contain rows at the baseline assessment timepoint
cat("Filtering just the baseline assessment timepoint data for use in baseline risk clustering", "\n")
raw_data_baseline <- raw_data %>% 
  filter(session_id == "ses-00A")

#2. Preliminarily create relevant risk clustering variables
cat("Creating secondary variables from raw ABCD data to be used in clustering", "\n")
#2.1 Family (first-degree relative) history of depression
raw_data_baseline_clean <- raw_data_baseline %>%
  mutate(
    family_history_depression = case_when(
      
      #2.1.1 Assign 1 if any first-degree relative has a history of depression
      replace_na(mh_p_famhx__dep__fath_001 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__moth_001 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__young_001 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__young_002 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__young_003 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__young_004 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__young_005 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__old_001 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__old_002 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__old_003 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__old_004 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__old_005 == 1, FALSE) ~ 1,
      
      #2.1.2 Assign 0 if "mh_p_famhx__dep_001" (any relative) is 1 but no first-degree relative has a history of depression
      mh_p_famhx__dep_001 == 1 ~ 0,
      
      #2.1.3 Assign 0 if "mh_p_famhx__dep_001" is 0
      mh_p_famhx__dep_001 == 0 ~ 0,
      
      #2.1.4 Otherwise, assign NA
      TRUE ~ NA_real_
    )
  )

#2.2 Family (first-degree relative) history of mania
raw_data_baseline_clean <- raw_data_baseline_clean %>%
  mutate(
    family_history_mania = case_when(
      
      #2.2.1 Assign 1 if any first-degree relative has a history of mania
      replace_na(mh_p_famhx__mania__fath_001 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__moth_001 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__young_001 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__young_002 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__young_003 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__young_004 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__young_005 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__old_001 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__old_002 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__old_003 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__old_004 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__old_005 == 1, FALSE) ~ 1,
      
      #2.2.2 Assign 0 if "mh_p_famhx__mania_001" (any relative) is 1 but no first-degree relative has a history of mania
      mh_p_famhx__mania_001 == 1 ~ 0,
      
      #2.2.3 Assign 0 if "mh_p_famhx__mania_001" is 0
      mh_p_famhx__mania_001 == 0 ~ 0,
      
      #2.2.4 Otherwise, assign NA
      TRUE ~ NA_real_
    )
  )

#2.3 Sum ACE Score Index
raw_data_baseline_clean <- raw_data_baseline_clean %>%
  mutate(
    
    #2.3.1.1 Calculate each ACE component
    ACE_1 = if_else(mh_p_famhx__troub_001 == 1, 1, 0, missing = NA),
    ACE_2 = if_else(fc_y_crpbi__cg1_005 %in% c(1, 2), 1, 0, missing = NA),
    ACE_3 = if_else(mh_p_famhx__suic_001 == 1 | mh_p_famhx__dep_001 == 1, 1, 0, missing = NA),
    ACE_4 = if_else(mh_p_famhx__alc_001 == 1 | mh_p_famhx__drg_001 == 1, 1, 0, missing = NA),
    ACE_5 = if_else(ksads_ptsd_raw_766_p == 1 | fc_y_fes__confl_006 == 1 | fc_y_fes__confl_003 == 1, 1, 0, missing = NA),
    ACE_6 = if_else(ab_p_demo__marital__slf_001 %in% c(2, 3, 4), 1, 0, missing = NA),
    ACE_7 = if_else(ksads_ptsd_raw_762_p == 1 | ksads_ptsd_raw_763_p == 1, 1, 0, missing = NA),
    ACE_8 = if_else(fc_y_pm_001 %in% c(1, 2) | fc_y_pm_003 %in% c(1, 2), 1, 0, missing = NA),
    ACE_9 = if_else(ksads_ptsd_raw_767_p == 1 | ksads_ptsd_raw_768_p == 1, 1, 0, missing = NA),
    
    #2.3.1.2 Sum the ACE components for each participant
    ACE_index_sum_score = rowSums(across(starts_with("ACE_")), na.rm = TRUE),
    
    #2.3.2 Count the number of missing values for the ACE components
    ACE_index_sum_score_n_missing = rowSums(
      across(c(
        mh_p_famhx__troub_001, 
        fc_y_crpbi__cg1_005, 
        mh_p_famhx__suic_001, 
        mh_p_famhx__dep_001, 
        mh_p_famhx__alc_001, 
        mh_p_famhx__drg_001, 
        ksads_ptsd_raw_766_p, 
        fc_y_fes__confl_006, 
        fc_y_fes__confl_003,
        ab_p_demo__marital__slf_001, 
        ksads_ptsd_raw_762_p, 
        ksads_ptsd_raw_763_p, 
        fc_y_pm_001, 
        fc_y_pm_003, 
        ksads_ptsd_raw_767_p, 
        ksads_ptsd_raw_768_p
      ), ~ . %in% c(NA, "", 999, 7, 777))),
    
    #2.3.3 Set sum score to NA if any components are missing
    ACE_index_sum_score = if_else(ACE_index_sum_score_n_missing >= 1, NA_real_, ACE_index_sum_score)
  )

#2.4 History of Bullying (Parent and/or Youth Reported)
raw_data_baseline_clean <- raw_data_baseline_clean %>%
  
  #2.4.1 Determine whether the parent or youth endorsed the youth being bullied
  mutate(bullying = case_when(
    mh_p_kbi__bully_001 == 1 | mh_y_kbi__bully_001 == 1 ~ 1, 
    is.na(mh_p_kbi__bully_001) & is.na(mh_y_kbi__bully_001) ~ NA_real_,
    mh_p_kbi__bully_001 == "" & mh_y_kbi__bully_001 == "" ~ NA_real_,
    TRUE ~ 0)
    )

#2.5 SDSC Total (Sum) Score
#2.5.1 Define the column names to be summed
sds_columns <- c(
  "ph_p_sds__da_001", "ph_p_sds__da_002", "ph_p_sds__da_003",
  "ph_p_sds__dims_001", "ph_p_sds__dims_002", "ph_p_sds__dims_003",
  "ph_p_sds__dims_004", "ph_p_sds__dims_005", "ph_p_sds__dims_006",
  "ph_p_sds__dims_007", "ph_p_sds__does_001", "ph_p_sds__does_002",
  "ph_p_sds__does_003", "ph_p_sds__does_004", "ph_p_sds__does_005",
  "ph_p_sds__hyphy_001", "ph_p_sds__hyphy_002", "ph_p_sds__sbd_001",
  "ph_p_sds__sbd_002", "ph_p_sds__sbd_003", "ph_p_sds__swtd_001",
  "ph_p_sds__swtd_002", "ph_p_sds__swtd_003", "ph_p_sds__swtd_004",
  "ph_p_sds__swtd_005", "ph_p_sds__swtd_006"
)

#2.5.2 Write a function to calculate total SDSC score for each participant row
calculate_sds_total <- function(data) {
  
  #2.5.2.1 For each row, check if ANY column has NA
  has_missing <- apply(data[, sds_columns], 1, function(row) any(is.na(row)))
  
  #2.5.2.2 Calculate the sum for each row
  raw_sums <- rowSums(data[, sds_columns], na.rm = TRUE)
  
  #2.5.2.3 Set the sum to NA for any row with missing values
  data$sds_total <- ifelse(has_missing, NA, raw_sums)
  
  #2.524 Return the data with the sum SDSC score
  return(data)
}

#2.5.3 Apply the SDSC score calculation function to the baseline data
raw_data_baseline_clean <- calculate_sds_total(raw_data_baseline_clean)

#2.6 Family (first-degree relative) history of psychosis ("saw visions or heard voices or thought people were spying on them or plotting against them")
raw_data_baseline_clean <- raw_data_baseline_clean %>%
  mutate(
    family_history_psychosis = case_when(
      
      #2.6.1 Assign 1 if any first-degree relative has a history of psychosis
      replace_na(mh_p_famhx__halluc__fath_001 == 1, FALSE) |
        replace_na(mh_p_famhx__halluc__moth_001 == 1, FALSE) |
        replace_na(mh_p_famhx__halluc__full__young_001 == 1, FALSE) |
        replace_na(mh_p_famhx__halluc__full__young_002 == 1, FALSE) |
        replace_na(mh_p_famhx__halluc__full__young_003 == 1, FALSE) |
        replace_na(mh_p_famhx__halluc__full__young_004 == 1, FALSE) |
        replace_na(mh_p_famhx__halluc__full__young_005 == 1, FALSE) |
        replace_na(mh_p_famhx__halluc__full__old_001 == 1, FALSE) |
        replace_na(mh_p_famhx__halluc__full__old_002 == 1, FALSE) |
        replace_na(mh_p_famhx__halluc__full__old_003 == 1, FALSE) |
        replace_na(mh_p_famhx__halluc__full__old_004 == 1, FALSE) |
        replace_na(mh_p_famhx__halluc__full__old_005 == 1, FALSE) ~ 1,
      
      #2.6.2 Assign 0 if "mh_p_famhx__halluc_001" (any relative) is 1 but no first-degree relative has a history of psychosis
      mh_p_famhx__halluc_001 == 1 ~ 0,
      
      #2.6.3 Assign 0 if "mh_p_famhx__halluc_001" is 0
      mh_p_famhx__halluc_001 == 0 ~ 0,
      
      #2.6.4 Otherwise, assign NA
      TRUE ~ NA_real_
    )
  )

#2.7 Family (first-degree relative) history of psychiatric hospitalization ("hospitalized because of emotional or mental problems, or drug or alcohol problems")
raw_data_baseline_clean <- raw_data_baseline_clean %>%
  mutate(
    family_history_hospitalization = case_when(
      
      #2.7.1 Assign 1 if any first-degree relative has a history of psychiatric hospitalization
      replace_na(mh_p_famhx__hosp__fath_001 == 1, FALSE) |
        replace_na(mh_p_famhx__hosp__moth_001 == 1, FALSE) |
        replace_na(mh_p_famhx__hosp__full__young_001 == 1, FALSE) |
        replace_na(mh_p_famhx__hosp__full__young_002 == 1, FALSE) |
        replace_na(mh_p_famhx__hosp__full__young_003 == 1, FALSE) |
        replace_na(mh_p_famhx__hosp__full__young_004 == 1, FALSE) |
        replace_na(mh_p_famhx__hosp__full__young_005 == 1, FALSE) |
        replace_na(mh_p_famhx__hosp__full__old_001 == 1, FALSE) |
        replace_na(mh_p_famhx__hosp__full__old_002 == 1, FALSE) |
        replace_na(mh_p_famhx__hosp__full__old_003 == 1, FALSE) |
        replace_na(mh_p_famhx__hosp__full__old_004 == 1, FALSE) |
        replace_na(mh_p_famhx__hosp__full__old_005 == 1, FALSE) ~ 1,
      
      #2.7.2 Assign 0 if "mh_p_famhx__hosp_001" (any relative) is 1 but no first-degree relative has a history of hospitalization
      mh_p_famhx__hosp_001 == 1 ~ 0,
      
      #2.7.3 Assign 0 if "mh_p_famhx__hosp_001" is 0
      mh_p_famhx__hosp_001 == 0 ~ 0,
      
      #2.7.4 Otherwise, assign NA
      TRUE ~ NA_real_
    )
  )

#2.8 Abuse indicators separated from the ACE composite
raw_data_baseline_clean <- raw_data_baseline_clean %>%
  mutate(
    
    #2.8.1 Physical abuse: ACE_7, set to NA if either constituent KSADS item is missing (same rule as the published composite)
    abuse_physical = if_else(
      rowSums(across(c(ksads_ptsd_raw_762_p, ksads_ptsd_raw_763_p), ~ . %in% c(NA, "", 999, 7, 777))) >= 1,
      NA_real_, ACE_7),
    
    #2.8.2 Sexual abuse: ACE_9, set to NA if either constituent KSADS item is missing
    abuse_sexual = if_else(
      rowSums(across(c(ksads_ptsd_raw_767_p, ksads_ptsd_raw_768_p), ~ . %in% c(NA, "", 999, 7, 777))) >= 1,
      NA_real_, ACE_9),
    
    #2.8.3 Emotional abuse proxy ("family members often criticize each other"), collapsed across reporters as for bullying
    abuse_emotional = case_when(
      fc_y_fes__confl_005 == 1 | fc_p_fes__confl_005 == 1 ~ 1,
      is.na(fc_y_fes__confl_005) & is.na(fc_p_fes__confl_005) ~ NA_real_,
      fc_y_fes__confl_005 == "" & fc_p_fes__confl_005 == "" ~ NA_real_,
      TRUE ~ 0)
  )

#2.9 ACE composites excluding abuse
raw_data_baseline_clean <- raw_data_baseline_clean %>%
  mutate(
    
    #2.9.1 Count missing raw items among the non-abuse ACE components only (ACE_1 to ACE_6, ACE_8)
    ACE_index_nonabuse_n_missing = rowSums(
      across(c(
        mh_p_famhx__troub_001, 
        fc_y_crpbi__cg1_005, 
        mh_p_famhx__suic_001, 
        mh_p_famhx__dep_001, 
        mh_p_famhx__alc_001, 
        mh_p_famhx__drg_001, 
        ksads_ptsd_raw_766_p, 
        fc_y_fes__confl_006, 
        fc_y_fes__confl_003,
        ab_p_demo__marital__slf_001, 
        fc_y_pm_001, 
        fc_y_pm_003
      ), ~ . %in% c(NA, "", 999, 7, 777))),
    
    #2.9.2 Primary: 7-factor non-abuse composite, NA if any constituent raw item is missing
    ACE_index_nonabuse_sum_score = if_else(
      ACE_index_nonabuse_n_missing >= 1, NA_real_,
      ACE_1 + ACE_2 + ACE_3 + ACE_4 + ACE_5 + ACE_6 + ACE_8),
    
    #2.9.3 Fallback: 8-factor composite adding the emotional abuse proxy; physical and sexual abuse remain separate
    ACE_index_nonabuse_emo_sum_score = ACE_index_nonabuse_sum_score + abuse_emotional
  )

#3. Retain columns of interest for risk clustering
cat("Retaining columns of interest for validation & clustering", "\n")
raw_data_baseline_clean_trim <- raw_data_baseline_clean %>% 
  dplyr::select(c(participant_id, session_id, ab_g_dyn__design_site, ab_g_dyn__visit_age, ab_g_stc__cohort_ethnrace__meim, ab_g_stc__cohort_sex, ab_g_stc__design_id__fam, mh_p_cbcl__dsm__dep_tscore, mh_p_cbcl__dsm__dep_nm, mh_p_cbcl__dsm__anx_tscore, mh_p_cbcl__dsm__anx_nm, mh_p_cbcl__synd__attn_tscore, mh_p_cbcl__synd__attn_nm, mh_p_cbcl__synd__aggr_tscore, mh_p_cbcl__synd__aggr_nm, mh_p_gbi_sum, mh_y_upps__nurg_sum, mh_y_upps__nurg_nm, mh_y_upps__purg_sum, mh_y_upps__purg_nm, le_l_coi__addr1__coi__total__national_zscore, fc_p_nsc__ns_mean, fc_p_nsc__ns_nm, sds_total, family_history_depression, family_history_mania, bullying, nc_y_nihtb__lswmt__uncor_score, nc_y_nihtb__lswmt__agecor_score, nc_y_nihtb__flnkr__uncor_score, nc_y_nihtb__flnkr__agecor_score, nc_y_nihtb__pttcp__uncor_score, nc_y_nihtb__pttcp__agecor_score, ACE_1, ACE_2, ACE_3, ACE_4, ACE_5, ACE_6, ACE_7, ACE_8, ACE_9, ACE_index_sum_score, ACE_index_sum_score_n_missing,
                
                #3.1 Revision additions: CBCL dysregulation profile raw syndrome sums and their missingness counts
                mh_p_cbcl__synd__anxdep_sum, mh_p_cbcl__synd__anxdep_nm, mh_p_cbcl__synd__attn_sum, mh_p_cbcl__synd__aggr_sum,
                
                #3.2 Revision additions: BIS/BAS (validated sums; computed by ABCD only when no items are missing)
                mh_y_bisbas__bas__rr_sum__v01, mh_y_bisbas__bas__dr_sum, mh_y_bisbas__bis_sum__v01, mh_y_bisbas__bas__fs_sum,
                
                #3.3 Revision additions: family history of psychosis and psychiatric hospitalization
                family_history_psychosis, family_history_hospitalization,
                
                #3.4 Revision additions: abuse indicators and non-abuse ACE composites
                abuse_physical, abuse_sexual, abuse_emotional, ACE_index_nonabuse_sum_score, ACE_index_nonabuse_emo_sum_score, ACE_index_nonabuse_n_missing,
                
                #3.5 Revision additions: SDSC subscales (descriptive only; not clustered)
                ph_p_sds__dims_sum, ph_p_sds__sbd_sum, ph_p_sds__da_sum, ph_p_sds__swtd_sum, ph_p_sds__does_sum, ph_p_sds__hyphy_sum))

```

## Risk Variable Cleaning

This section cleans the variables to be used for risk clustering by:

  1. Altering column names and coded values within columns to match desired formatting
  
  2. Removing risk variable values from the data that were generated using any missing data (e.g., incomplete CBCL -> t scores)

```{r variable cleaning, warning = FALSE}

## Data Wrangling ##

#1. Alter column names and coded values within columns to match desired formatting
cat("Altering column names and coded values within columns to match desired formatting", "\n")
#1.1 Alter the biological sex column name + values
colnames(raw_data_baseline_clean_trim)[colnames(raw_data_baseline_clean_trim) == "ab_g_stc__cohort_sex"] <- "sex"
raw_data_baseline_clean_trim$sex[raw_data_baseline_clean_trim$sex == 1] <- "M"
raw_data_baseline_clean_trim$sex[raw_data_baseline_clean_trim$sex == 2] <- "F"
raw_data_baseline_clean_trim$sex[raw_data_baseline_clean_trim$sex == 3] <- "Non-Binary"

#1.2 Alter the race-ethnicity column name & coded values
colnames(raw_data_baseline_clean_trim)[colnames(raw_data_baseline_clean_trim) == "ab_g_stc__cohort_ethnrace__meim"] <- "race_ethnicity"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 0] <- "Other Ethnicity"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 1] <- "White/Caucasian"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 2] <- "Western European"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 3] <- "Eastern European"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 4] <- "Hispanic/Latino"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 5] <- "Black/African American"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 6] <- "Afro-Carribean/Indo-Carribbean/West Indian"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 7] <- "East Asian"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 8] <- "South Asian"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 9] <- "Southeast Asian"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 10] <- "American Indian/Alaska Native"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 11] <- "Middle Eastern/North African"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 12] <- "Native Hawaiian or Pacific Islander"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 13] <- "Multiracial/Multiethnic"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 14] <- "Other Ethnicity"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == 15] <- "None"
raw_data_baseline_clean_trim$race_ethnicity[raw_data_baseline_clean_trim$race_ethnicity == ""] <- "Unsure"
raw_data_baseline_clean_trim$race_ethnicity[is.na(raw_data_baseline_clean_trim$race_ethnicity)] <- "Unsure"

#1.3 Change the name of the family id column
colnames(raw_data_baseline_clean_trim)[colnames(raw_data_baseline_clean_trim) == "ab_g_stc__design_id__fam"] <- "family_id"

#1.4 Change the name of the site id column
colnames(raw_data_baseline_clean_trim)[colnames(raw_data_baseline_clean_trim) == "ab_g_dyn__design_site"] <- "site"

#1.5 Change the name of the age column 
colnames(raw_data_baseline_clean_trim)[colnames(raw_data_baseline_clean_trim) == "ab_g_dyn__visit_age"] <- "age"

#2. Use nm columns to turn any respective instances of values created using missing data to NA
cat("Removing risk variable values from the data that were generated using any missing data (e.g., incomplete CBCL -> t scores)", "\n")
raw_data_baseline_clean_trim <- raw_data_baseline_clean_trim %>%
  mutate(
    
    #2.1 CBCL Depression
    mh_p_cbcl__dsm__dep_tscore = if_else(mh_p_cbcl__dsm__dep_nm >= 1, NA_real_, mh_p_cbcl__dsm__dep_tscore),
    
    #2.2 CBCL Anxiety
    mh_p_cbcl__dsm__anx_tscore = if_else(mh_p_cbcl__dsm__anx_nm >= 1, NA_real_, mh_p_cbcl__dsm__anx_tscore),
    
    #2.3 CBCL Attention
    mh_p_cbcl__synd__attn_tscore = if_else(mh_p_cbcl__synd__attn_nm >= 1, NA_real_, mh_p_cbcl__synd__attn_tscore),
    
    #2.4 CBCL Aggression
    mh_p_cbcl__synd__aggr_tscore = if_else(mh_p_cbcl__synd__aggr_nm >= 1, NA_real_, mh_p_cbcl__synd__aggr_tscore),
    
    #2.5 UPPS Negative Urgency
    mh_y_upps__nurg_sum = if_else(mh_y_upps__nurg_nm >= 1, NA_real_, mh_y_upps__nurg_sum),
    
    #2.6 UPPS Positive Urgency
    mh_y_upps__purg_sum = if_else(mh_y_upps__purg_nm >= 1, NA_real_, mh_y_upps__purg_sum),
    
    #2.7 Neighborhood Safety Score
    fc_p_nsc__ns_mean = if_else(fc_p_nsc__ns_nm >= 1, NA_real_, fc_p_nsc__ns_mean),
    
    #2.8 CBCL Anxious/Depressed raw syndrome sum
    mh_p_cbcl__synd__anxdep_sum = if_else(mh_p_cbcl__synd__anxdep_nm >= 1, NA_real_, as.numeric(mh_p_cbcl__synd__anxdep_sum)),
    
    #2.9 CBCL Attention Problems raw syndrome sum (shares the missingness count with its T-score)
    mh_p_cbcl__synd__attn_sum = if_else(mh_p_cbcl__synd__attn_nm >= 1, NA_real_, as.numeric(mh_p_cbcl__synd__attn_sum)),
    
    #2.10 CBCL Aggressive Behavior raw syndrome sum (shares the missingness count with its T-score)
    mh_p_cbcl__synd__aggr_sum = if_else(mh_p_cbcl__synd__aggr_nm >= 1, NA_real_, as.numeric(mh_p_cbcl__synd__aggr_sum)),
    
    #2.11 CBCL dysregulation profile raw sum, per Giampetruzzi et al. (2026); NA if any constituent sum is NA
    cbcl_dp_raw_sum = mh_p_cbcl__synd__anxdep_sum + mh_p_cbcl__synd__attn_sum + mh_p_cbcl__synd__aggr_sum
  )

```

## Risk Variable Data Validation & Trimming

This section validates and trims the data to be used for risk clustering by:

  1. Creating a dataframe containing columns of interest to the clustering process
  
  2. Creating a copy of the data which only contains subjects who have complete data (i.e., can be clustered)
  
  3. Cross checking (using a distinct function) whether any secondary variables created herein were done so correctly
  
  4. Summarizing the properties of the final dataset via EDA functions from skimr

```{r variable validation + trimming, warning = FALSE}

## Data Cleaning ##

#1. Create a dataframe containing only ordered variables of interest to risk group clustering
cat("Creating a dataframe containing columns of interest to the clustering process", "\n")
#1.1 Identifier and demographic columns, required for every analysis
cols_id_demo <- c("participant_id", "session_id", "site", "age", "race_ethnicity", "sex", "family_id")

#1.2 Columns retained for description and continuity with the submitted analysis, but not required to be complete
#1.2.1 The 4 CBCL T-scores and 9-factor ACE composite from the submitted 17-feature solution, plus SDSC subscales
cols_descriptive <- c(
  "mh_p_cbcl__dsm__dep_tscore", "mh_p_cbcl__dsm__anx_tscore",
  "mh_p_cbcl__synd__attn_tscore", "mh_p_cbcl__synd__aggr_tscore",
  "ACE_index_sum_score",
  "ph_p_sds__dims_sum", "ph_p_sds__sbd_sum", "ph_p_sds__da_sum",
  "ph_p_sds__swtd_sum", "ph_p_sds__does_sum", "ph_p_sds__hyphy_sum"
)

#1.3 Assemble the dataframe: identifiers, every clustering feature across all configurations, then descriptive columns
risk_variable_data <- raw_data_baseline_clean_trim %>% 
  dplyr::select(all_of(c(cols_id_demo, feats_all_configs, cols_descriptive)))

#2. Retain only subjects with complete data in columns of interest
#2.1 Completeness is required on identifiers/demographics and the union of all configuration features, so the primary,
#    backup, and fallback configurations are clustered on identical participants; descriptive columns may be missing
cat("Creating a copy of the data which only contains subjects who have complete data (i.e., can be clustered)", "\n")
cols_required <- c(cols_id_demo, feats_all_configs)
risk_variable_data_clean <- risk_variable_data %>% 
  filter(if_all(all_of(cols_required), ~ !is.na(.)) & !if_any(all_of(cols_required), ~ . %in% ""))

#2.2 Report the analytic sample size and exclusions
cat(sprintf("Analytic sample: %d of %d baseline participants retained (%d excluded, %.1f%%)\n",
            nrow(risk_variable_data_clean), nrow(risk_variable_data),
            nrow(risk_variable_data) - nrow(risk_variable_data_clean),
            100 * (nrow(risk_variable_data) - nrow(risk_variable_data_clean)) / nrow(risk_variable_data)))

#2.3 Report per-feature missingness in the unfiltered data, to identify which features drive exclusion
risk_variable_data %>%
  summarise(across(all_of(feats_all_configs), ~ sum(is.na(.) | . %in% ""))) %>%
  tidyr::pivot_longer(everything(), names_to = "Feature", values_to = "N Missing") %>%
  mutate(`Percent Missing` = round(100 * `N Missing` / nrow(risk_variable_data), 2)) %>%
  arrange(desc(`N Missing`)) %>%
  kable(caption = "Baseline missingness across all revision clustering features") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#3. Create a function ("validate_cluster_vars") to verify all derived clustering variables
cat("Cross checking (using a distinct function) whether any secondary variables created herein were done so correctly", "\n")
validate_cluster_vars <- function(raw_clean, final_data) {
  
  #3.1 Family history of depression
  expected_fhd <- with(raw_clean, case_when(
    replace_na(mh_p_famhx__dep__fath_001 == 1, FALSE) |
      replace_na(mh_p_famhx__dep__moth_001 == 1, FALSE) |
      replace_na(mh_p_famhx__dep__full__young_001 == 1, FALSE) |
      replace_na(mh_p_famhx__dep__full__young_002 == 1, FALSE) |
      replace_na(mh_p_famhx__dep__full__young_003 == 1, FALSE) |
      replace_na(mh_p_famhx__dep__full__young_004 == 1, FALSE) |
      replace_na(mh_p_famhx__dep__full__young_005 == 1, FALSE) |
      replace_na(mh_p_famhx__dep__full__old_001 == 1, FALSE) |
      replace_na(mh_p_famhx__dep__full__old_002 == 1, FALSE) |
      replace_na(mh_p_famhx__dep__full__old_003 == 1, FALSE) |
      replace_na(mh_p_famhx__dep__full__old_004 == 1, FALSE) |
      replace_na(mh_p_famhx__dep__full__old_005 == 1, FALSE) ~ 1,
    mh_p_famhx__dep_001 == 1 ~ 0,
    mh_p_famhx__dep_001 == 0 ~ 0,
    TRUE ~ NA_real_))
  mismatches_fhd <- sum(expected_fhd != final_data$family_history_depression, na.rm = TRUE)
  
  #3.2 Family history of mania
  expected_fhm <- with(raw_clean, case_when(
    replace_na(mh_p_famhx__mania__fath_001 == 1, FALSE) |
      replace_na(mh_p_famhx__mania__moth_001 == 1, FALSE) |
      replace_na(mh_p_famhx__mania__full__young_001 == 1, FALSE) |
      replace_na(mh_p_famhx__mania__full__young_002 == 1, FALSE) |
      replace_na(mh_p_famhx__mania__full__young_003 == 1, FALSE) |
      replace_na(mh_p_famhx__mania__full__young_004 == 1, FALSE) |
      replace_na(mh_p_famhx__mania__full__young_005 == 1, FALSE) |
      replace_na(mh_p_famhx__mania__full__old_001 == 1, FALSE) |
      replace_na(mh_p_famhx__mania__full__old_002 == 1, FALSE) |
      replace_na(mh_p_famhx__mania__full__old_003 == 1, FALSE) |
      replace_na(mh_p_famhx__mania__full__old_004 == 1, FALSE) |
      replace_na(mh_p_famhx__mania__full__old_005 == 1, FALSE) ~ 1,
    mh_p_famhx__mania_001 == 1 ~ 0,
    mh_p_famhx__mania_001 == 0 ~ 0,
    TRUE ~ NA_real_))
  mismatches_fhm <- sum(expected_fhm != final_data$family_history_mania, na.rm = TRUE)
  
  #3.3 Bullying endorsement (parent or youth)
  expected_bully <- with(raw_clean, case_when(
    mh_p_kbi__bully_001 == 1 | mh_y_kbi__bully_001 == 1 ~ 1,
    (is.na(mh_p_kbi__bully_001) & is.na(mh_y_kbi__bully_001)) ~ NA_real_,
    (mh_p_kbi__bully_001 == "" & mh_y_kbi__bully_001 == "") ~ NA_real_,
    TRUE ~ 0))
  mismatches_bully <- sum(expected_bully != final_data$bullying, na.rm = TRUE)
  
  #3.4 ACE index sum
  #3.4.1 Recompute each ACE component
  ace_components <- raw_clean %>%
    transmute(
      ACE_1 = if_else(mh_p_famhx__troub_001 == 1, 1, 0, missing = NA_real_),
      ACE_2 = if_else(fc_y_crpbi__cg1_005 %in% c(1,2), 1, 0, missing = NA_real_),
      ACE_3 = if_else(mh_p_famhx__suic_001 == 1 | mh_p_famhx__dep_001 == 1, 1, 0, missing = NA_real_),
      ACE_4 = if_else(mh_p_famhx__alc_001 == 1 | mh_p_famhx__drg_001 == 1, 1, 0, missing = NA_real_),
      ACE_5 = if_else(ksads_ptsd_raw_766_p == 1 | fc_y_fes__confl_006 == 1 | fc_y_fes__confl_003 == 1, 1, 0, missing = NA_real_),
      ACE_6 = if_else(ab_p_demo__marital__slf_001 %in% c(2,3,4), 1, 0, missing = NA_real_),
      ACE_7 = if_else(ksads_ptsd_raw_762_p == 1 | ksads_ptsd_raw_763_p == 1, 1, 0, missing = NA_real_),
      ACE_8 = if_else(fc_y_pm_001 %in% c(1,2) | fc_y_pm_003 %in% c(1,2), 1, 0, missing = NA_real_),
      ACE_9 = if_else(ksads_ptsd_raw_767_p == 1 | ksads_ptsd_raw_768_p == 1, 1, 0, missing = NA_real_))
  
  #3.4.2 Count missing ACE components
  missing_count <- raw_clean %>%
    transmute(n_miss = rowSums(across(c(
      mh_p_famhx__troub_001, fc_y_crpbi__cg1_005, mh_p_famhx__suic_001,
      mh_p_famhx__dep_001, mh_p_famhx__alc_001, mh_p_famhx__drg_001,
      ksads_ptsd_raw_766_p, fc_y_fes__confl_006, fc_y_fes__confl_003,
      ab_p_demo__marital__slf_001, ksads_ptsd_raw_762_p, ksads_ptsd_raw_763_p,
      fc_y_pm_001, fc_y_pm_003, ksads_ptsd_raw_767_p, ksads_ptsd_raw_768_p
    ), ~ . %in% c(NA, "", 999, 7, 777))))
  
  #3.4.3 Compute expected ACE index
  expected_ace_index <- if_else(
    missing_count$n_miss >= 1, NA_real_,
    rowSums(ace_components, na.rm = TRUE))
  mismatches_ace <- sum(expected_ace_index != final_data$ACE_index_sum_score, na.rm = TRUE)
  
  #3.5 SDSC total score
  #3.5.1 Identify SDSC items and missing rows
  sds_cols <- c(
    "ph_p_sds__da_001","ph_p_sds__da_002","ph_p_sds__da_003",
    "ph_p_sds__dims_001","ph_p_sds__dims_002","ph_p_sds__dims_003",
    "ph_p_sds__dims_004","ph_p_sds__dims_005","ph_p_sds__dims_006",
    "ph_p_sds__dims_007","ph_p_sds__does_001","ph_p_sds__does_002",
    "ph_p_sds__does_003","ph_p_sds__does_004","ph_p_sds__does_005",
    "ph_p_sds__hyphy_001","ph_p_sds__hyphy_002","ph_p_sds__sbd_001",
    "ph_p_sds__sbd_002","ph_p_sds__sbd_003","ph_p_sds__swtd_001",
    "ph_p_sds__swtd_002","ph_p_sds__swtd_003","ph_p_sds__swtd_004",
    "ph_p_sds__swtd_005","ph_p_sds__swtd_006")
  
  #3.5.2 Recompute SDSC and compare
  has_missing_sds <- apply(raw_clean[sds_cols], 1, function(r) any(is.na(r)))
  expected_sds <- ifelse(has_missing_sds, NA_real_, rowSums(raw_clean[sds_cols], na.rm = TRUE))
  mismatches_sds <- sum(expected_sds != final_data$sds_total, na.rm = TRUE)
  
  #3.6 CBCL, UPPS, and NS mean cleaning
  expected_dep_t <- if_else(raw_clean$mh_p_cbcl__dsm__dep_nm >= 1, NA_real_, raw_clean$mh_p_cbcl__dsm__dep_tscore)
  expected_anx_t <- if_else(raw_clean$mh_p_cbcl__dsm__anx_nm >= 1, NA_real_, raw_clean$mh_p_cbcl__dsm__anx_tscore)
  expected_attn_t <- if_else(raw_clean$mh_p_cbcl__synd__attn_nm >= 1, NA_real_, raw_clean$mh_p_cbcl__synd__attn_tscore)
  expected_aggr_t <- if_else(raw_clean$mh_p_cbcl__synd__aggr_nm >= 1, NA_real_, raw_clean$mh_p_cbcl__synd__aggr_tscore)
  expected_nurg <- if_else(raw_clean$mh_y_upps__nurg_nm >= 1, NA_real_, raw_clean$mh_y_upps__nurg_sum)
  expected_purg <- if_else(raw_clean$mh_y_upps__purg_nm >= 1, NA_real_, raw_clean$mh_y_upps__purg_sum)
  expected_ns_mean <- if_else(raw_clean$fc_p_nsc__ns_nm >= 1, NA_real_, raw_clean$fc_p_nsc__ns_mean)
  
  mismatches_cont <- tibble(
    Variable = c(
      "mh_p_cbcl__dsm__dep_tscore","mh_p_cbcl__dsm__anx_tscore",
      "mh_p_cbcl__synd__attn_tscore","mh_p_cbcl__synd__aggr_tscore",
      "mh_y_upps__nurg_sum","mh_y_upps__purg_sum","fc_p_nsc__ns_mean"),
    Mismatches = c(
      sum(expected_dep_t != final_data$mh_p_cbcl__dsm__dep_tscore, na.rm = TRUE),
      sum(expected_anx_t != final_data$mh_p_cbcl__dsm__anx_tscore, na.rm = TRUE),
      sum(expected_attn_t != final_data$mh_p_cbcl__synd__attn_tscore, na.rm = TRUE),
      sum(expected_aggr_t != final_data$mh_p_cbcl__synd__aggr_tscore, na.rm = TRUE),
      sum(expected_nurg != final_data$mh_y_upps__nurg_sum, na.rm = TRUE),
      sum(expected_purg != final_data$mh_y_upps__purg_sum, na.rm = TRUE),
      sum(expected_ns_mean != final_data$fc_p_nsc__ns_mean, na.rm = TRUE))
    )
  
  #3.7 Direct continuous variables (no transformation expected)
  #3.7.1 Specify direct variables
  direct_vars <- c(
    "mh_p_gbi_sum","le_l_coi__addr1__coi__total__national_zscore",
    "nc_y_nihtb__lswmt__uncor_score","nc_y_nihtb__flnkr__uncor_score",
    "nc_y_nihtb__pttcp__uncor_score"
    )
  
  #3.7.2 Compare raw vs final
  mismatches_direct <- tibble(
    Variable   = direct_vars,
    Mismatches = sapply(direct_vars, function(v) {
      sum(raw_clean[[v]] != final_data[[v]], na.rm = TRUE)})
    )
  
  #3.7.3 Revision additions: family history of psychosis and hospitalization, recomputed from raw items
  first_degree_any <- function(data, stem) {
    cols <- c(sprintf("mh_p_famhx__%s__fath_001", stem), sprintf("mh_p_famhx__%s__moth_001", stem),
              sprintf("mh_p_famhx__%s__full__young_%03d", stem, 1:5), sprintf("mh_p_famhx__%s__full__old_%03d", stem, 1:5))
    Reduce(`|`, lapply(cols, function(cc) replace_na(data[[cc]] == 1, FALSE)))
  }
  expected_fhp <- case_when(
    first_degree_any(raw_clean, "halluc") ~ 1,
    raw_clean$mh_p_famhx__halluc_001 %in% c(0, 1) ~ 0,
    TRUE ~ NA_real_)
  expected_fhh <- case_when(
    first_degree_any(raw_clean, "hosp") ~ 1,
    raw_clean$mh_p_famhx__hosp_001 %in% c(0, 1) ~ 0,
    TRUE ~ NA_real_)
  
  #3.7.4 Revision additions: abuse indicators, recomputed from raw items
  miss_codes <- c(NA, "", 999, 7, 777)
  expected_phys <- with(raw_clean, if_else(
    ksads_ptsd_raw_762_p %in% miss_codes | ksads_ptsd_raw_763_p %in% miss_codes, NA_real_,
    as.numeric(ksads_ptsd_raw_762_p == 1 | ksads_ptsd_raw_763_p == 1)))
  expected_sex <- with(raw_clean, if_else(
    ksads_ptsd_raw_767_p %in% miss_codes | ksads_ptsd_raw_768_p %in% miss_codes, NA_real_,
    as.numeric(ksads_ptsd_raw_767_p == 1 | ksads_ptsd_raw_768_p == 1)))
  expected_emo <- with(raw_clean, case_when(
    fc_y_fes__confl_005 == 1 | fc_p_fes__confl_005 == 1 ~ 1,
    is.na(fc_y_fes__confl_005) & is.na(fc_p_fes__confl_005) ~ NA_real_,
    fc_y_fes__confl_005 == "" & fc_p_fes__confl_005 == "" ~ NA_real_,
    TRUE ~ 0))
  
  #3.7.5 Revision additions: non-abuse ACE composites, recomputed from the independently computed ACE components above
  nonabuse_items <- c("mh_p_famhx__troub_001", "fc_y_crpbi__cg1_005", "mh_p_famhx__suic_001", "mh_p_famhx__dep_001",
                      "mh_p_famhx__alc_001", "mh_p_famhx__drg_001", "ksads_ptsd_raw_766_p", "fc_y_fes__confl_006",
                      "fc_y_fes__confl_003", "ab_p_demo__marital__slf_001", "fc_y_pm_001", "fc_y_pm_003")
  nonabuse_missing <- rowSums(sapply(nonabuse_items, function(cc) raw_clean[[cc]] %in% miss_codes)) >= 1
  expected_ace_nonabuse <- if_else(
    nonabuse_missing, NA_real_,
    rowSums(ace_components[, c("ACE_1", "ACE_2", "ACE_3", "ACE_4", "ACE_5", "ACE_6", "ACE_8")]))
  expected_ace_nonabuse_emo <- expected_ace_nonabuse + expected_emo
  
  #3.7.6 Revision additions: CBCL raw syndrome sums under the number-missing rule, and the dysregulation profile raw sum
  expected_anxdep <- if_else(raw_clean$mh_p_cbcl__synd__anxdep_nm >= 1, NA_real_, as.numeric(raw_clean$mh_p_cbcl__synd__anxdep_sum))
  expected_attn <- if_else(raw_clean$mh_p_cbcl__synd__attn_nm >= 1, NA_real_, as.numeric(raw_clean$mh_p_cbcl__synd__attn_sum))
  expected_aggr <- if_else(raw_clean$mh_p_cbcl__synd__aggr_nm >= 1, NA_real_, as.numeric(raw_clean$mh_p_cbcl__synd__aggr_sum))
  expected_dp <- expected_anxdep + expected_attn + expected_aggr
  
  #3.7.7 Count mismatches, and separately count disagreement in missingness, which != with na.rm would otherwise hide
  count_mismatch <- function(expected, observed) {
    sum(expected != observed, na.rm = TRUE) + sum(is.na(expected) != is.na(observed))
  }
  mismatches_revision <- tibble(
    Variable = c("family_history_psychosis", "family_history_hospitalization",
                 "abuse_physical", "abuse_sexual", "abuse_emotional",
                 "ACE_index_nonabuse_sum_score", "ACE_index_nonabuse_emo_sum_score",
                 "mh_p_cbcl__synd__anxdep_sum", "mh_p_cbcl__synd__attn_sum", "mh_p_cbcl__synd__aggr_sum",
                 "cbcl_dp_raw_sum"),
    Mismatches = c(
      count_mismatch(expected_fhp, final_data$family_history_psychosis),
      count_mismatch(expected_fhh, final_data$family_history_hospitalization),
      count_mismatch(expected_phys, final_data$abuse_physical),
      count_mismatch(expected_sex, final_data$abuse_sexual),
      count_mismatch(expected_emo, final_data$abuse_emotional),
      count_mismatch(expected_ace_nonabuse, final_data$ACE_index_nonabuse_sum_score),
      count_mismatch(expected_ace_nonabuse_emo, final_data$ACE_index_nonabuse_emo_sum_score),
      count_mismatch(expected_anxdep, final_data$mh_p_cbcl__synd__anxdep_sum),
      count_mismatch(expected_attn, final_data$mh_p_cbcl__synd__attn_sum),
      count_mismatch(expected_aggr, final_data$mh_p_cbcl__synd__aggr_sum),
      count_mismatch(expected_dp, final_data$cbcl_dp_raw_sum))
  )
  
  #3.7.8 Revision additions: BIS/BAS are carried through unchanged from the raw export
  bas_vars <- c("mh_y_bisbas__bas__rr_sum__v01", "mh_y_bisbas__bas__dr_sum", "mh_y_bisbas__bis_sum__v01", "mh_y_bisbas__bas__fs_sum")
  mismatches_bas <- tibble(
    Variable = bas_vars,
    Mismatches = sapply(bas_vars, function(v) count_mismatch(raw_clean[[v]], final_data[[v]])))
  
  #3.7.9 Structural check: the 9-factor published composite must equal the non-abuse composite plus the two abuse indicators
  structural_ace <- count_mismatch(
    final_data$ACE_index_sum_score,
    final_data$ACE_index_nonabuse_sum_score + final_data$abuse_physical + final_data$abuse_sexual)
  
  #3.8 Combine all mismatch counts into a single summary
  mismatch_summary <- bind_rows(
    tibble(Variable = "family_history_depression", Mismatches = mismatches_fhd),
    tibble(Variable = "family_history_mania", Mismatches = mismatches_fhm),
    tibble(Variable = "bullying", Mismatches = mismatches_bully),
    tibble(Variable = "ACE_index_sum_score", Mismatches = mismatches_ace),
    tibble(Variable = "sds_total", Mismatches = mismatches_sds),
    mismatches_cont,
    mismatches_direct,
    mismatches_revision,
    mismatches_bas,
    tibble(Variable = "ACE 9-factor = nonabuse + physical + sexual", Mismatches = structural_ace)
  )
  
  #3.9 Return the mismatch summary
  return(mismatch_summary)
}

#3.10 Call the function with your pre- and post-wrangled data
validation_results <- validate_cluster_vars(
  raw_clean = raw_data_baseline_clean,
  final_data = risk_variable_data
)

#3.11 View the mismatch summary
kable(validation_results) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))

#3.12 Assert that all mismatches are zero, halting the render otherwise so no unverified data reach clustering
if (all(validation_results$Mismatches == 0)) {
  cat("All derived variables match expected values.\n")
} else {
  stop("Derivation mismatches detected: ",
       paste(validation_results$Variable[validation_results$Mismatches > 0], collapse = ", "))
}

#3.13 Confirm every configuration's features are present in the cleaned data
stopifnot(all(feats_all_configs %in% names(risk_variable_data_clean)))

#4. Summarize final cleaned dataset
cat("Summarizing the properties of the final dataset via EDA functions from skimr", "\n")
skimr::skim(risk_variable_data_clean)

```

## Writing Output to Processed Data Directory

```{r output, warning = FALSE}

## Output ##

#1. Write the cleaned and filtered risk variable data as a csv file to be used in clustering
write.csv(risk_variable_data_clean, "../../../data/data_processed/risk_variable_data.csv", row.names = FALSE)

#2. Write the cleaned and not filtered (i.e., subjects w/o complete data not excluded) as a csv file to be used later in missignness analyses
write.csv(risk_variable_data, "../../../data/data_processed/risk_variable_data_unfiltered.csv", row.names = FALSE)

```
