---
title: "Supplementary Outcome Variables: Umbrella Dx + 4-Group Dx Variable"
author: "Sam A. Sievertsen"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    number_sections: false
    df_print: paged
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(cache = FALSE, message = TRUE, warning = TRUE, results = "markup", verbose = TRUE, comment = "")

```

```{r environment, echo = FALSE, message = FALSE, warning = FALSE}

# Load packages
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(skimr)
library(ggplot2)
library(here)

# Set env options
options(scipen = 999, digits = 8)
set.seed(123)

# Paths
REPO <- here()
proc_dir <- file.path(REPO, "data", "data_processed")
kproto_file <- file.path(proc_dir, "kproto_results", sprintf("kproto_%s.rds", "z_score"))
raw_path <- file.path(REPO, "data", "data_raw", "dataset.csv")
outcome_path <- file.path(proc_dir, "outcome_variable_data.csv")
out_path <- file.path(proc_dir, "supplementary_outcome_variable_data.csv")

# Read in raw data
raw_data <- read.csv(raw_path)

# Assess whether previous outcome data exists
if (!file.exists(outcome_path)) {
  stop(
    "Missing file: data/data_processed/outcome_variable_data.csv\n",
    "Run outcome_variable_data_wrangling.Rmd first so we can merge validated any_bsd, etc.")
}

# Read in validated BD/SI outcomes generated by outcome_variable_data_wrangling.Rmd
outcome_data <- read.csv(outcome_path)

# Merge previously generated outcomes into the raw dataset
dat <- raw_data %>%
  left_join(outcome_data, by = c("participant_id", "session_id"))

# Read in cluster label inputs
kp_list <- readRDS(kproto_file)
pos <- 1
kp <- kp_list[[pos]]

# Create a function to extract participant_id and corresponding cluster labels from kproto$data 
get_cluster_labels <- function(kp, k_value) {
  dat <- as.data.frame(kp$data)
  
  # Prioritize rownames of the kp obj first
  if (!is.null(rownames(dat)) && all(nzchar(rownames(dat)))) {
    ids <- rownames(dat)
  } else if ("participant_id" %in% names(dat)) {
    ids <- as.character(dat$participant_id)
  } else {
    stop("Could not infer participant_id for cluster labels from kproto$data. ",
         "Ensure rownames carry IDs or participant_id is a column.")
  }
  cl <- factor(kp$cluster, levels = seq_len(k_value), labels = paste0("C", seq_len(k_value)))
  tibble(participant_id = as.character(ids), cluster = cl)
}

# Extract cluster labels
labels_tbl <- get_cluster_labels(kp, 2)

# Merge cluster labels into dataset to be wrangled
dat <- dat %>% 
  left_join(labels_tbl, by = "participant_id")

```

## Overview

This script:

1.  Normalizes KSADS dx coding for the *new* raw diagnosis columns (888 -> 0; 555/999/""/NA -> NA)

2.  Creates umbrella (Sub_Domain) dx variables (0/1/NA) for conditions commonly co morbid with bipolar disorder (BD). Specifically:

    -   ADHD
    -   GAD
    -   Social Anxiety
    -   Separation Anxiety
    -   Panic Disorder
    -   OCD,
    -   DMDD
    -   MDD
    -   PDD
    -   PTSD
    -   ODD
    -   Conduct Disorder

3.  Creates a 4-group categorical variable at each session:

    1.  No Dx (in the dx set of interest)
    2.  BSD only
    3.  BSD + other Dx
    4.  Other Dx only (no BSD)

4.  Writes a single processed CSV containing all columns + new dx vars

## Define raw column sets for each umbrella diagnosis

```{r column mapping, echo = FALSE, warning = FALSE}

#2.1 ADHD (Parent)
adhd_vars <- c(
"mh_p_ksads__adhd__past_dx",
"mh_p_ksads__adhd__pres_dx",
"mh_p_ksads__adhd__partrem_dx",
"mh_p_ksads__adhd__unspec_dx",
"mh_p_ksads__adhd__oth_dx")

#2.2 GAD (Parent + Youth)
gad_vars <- c(
"mh_p_ksads__gad__past_dx",
"mh_p_ksads__gad__pres_dx",
"mh_y_ksads__gad__past_dx",
"mh_y_ksads__gad__pres_dx")

#2.3 Social Anxiety (Parent + Youth)
socanx_vars <- c(
"mh_p_ksads__socanx__past_dx",
"mh_p_ksads__socanx__pres_dx",
"mh_y_ksads__socanx__past_dx",
"mh_y_ksads__socanx__pres_dx")

#2.4 Separation Anxiety (Parent)
sepanx_vars <- c(
"mh_p_ksads__sepanx__past_dx",
"mh_p_ksads__sepanx__pres_dx")

#2.5 Panic Disorder (Parent + Youth)
panic_vars <- c(
"mh_p_ksads__panic__past_dx",
"mh_p_ksads__panic__pres_dx",
"mh_p_ksads__panic__partrem_dx",
"mh_y_ksads__panic__past_dx",
"mh_y_ksads__panic__pres_dx",
"mh_y_ksads__panic__partrem_dx")

#2.6 OCD (Parent + Youth)
ocd_vars <- c(
"mh_p_ksads__ocd__past_dx",
"mh_p_ksads__ocd__pres_dx",
"mh_y_ksads__ocd__past_dx",
"mh_y_ksads__ocd__pres_dx")

#2.7 DMDD (Parent + Youth; present only)
dmdd_vars <- c(
"mh_p_ksads__dmdd__dmdd__pres_dx",
"mh_y_ksads__dmdd__dmdd__pres_dx")

#2.8 MDD (Parent + Youth)
mdd_vars <- c(
"mh_p_ksads__dep__mdd__past_dx",
"mh_p_ksads__dep__mdd__pres_dx",
"mh_p_ksads__dep__mdd__partrem_dx",
"mh_y_ksads__dep__mdd__past_dx",
"mh_y_ksads__dep__mdd__pres_dx",
"mh_y_ksads__dep__mdd__partrem_dx")

#2.9 PDD / Dysthymia (Parent + Youth)
pdd_vars <- c(
"mh_p_ksads__dep__pdd__past_dx",
"mh_p_ksads__dep__pdd__pres_dx",
"mh_p_ksads__dep__pdd__partrem_dx",
"mh_y_ksads__dep__pdd__past_dx",
"mh_y_ksads__dep__pdd__pres_dx",
"mh_y_ksads__dep__pdd__partrem_dx")

#2.10 PTSD (Parent + Youth)
ptsd_vars <- c(
"mh_p_ksads__ptsd__past_dx",
"mh_p_ksads__ptsd__pres_dx",
"mh_p_ksads__ptsd__partrem_dx",
"mh_y_ksads__ptsd__past_dx",
"mh_y_ksads__ptsd__pres_dx",
"mh_y_ksads__ptsd__partrem_dx")

#2.11 ODD (Parent)
odd_vars <- c(
"mh_p_ksads__odd__past_dx",
"mh_p_ksads__odd__pres_dx")

#2.12 Conduct Disorder (Parent + Youth; adolescent onset + childhood onset)
cd_vars <- c(
"mh_p_ksads__cond__onset__adoles__past_dx",
"mh_p_ksads__cond__onset__adoles__pres_dx",
"mh_p_ksads__cond__onset__chld__past_dx",
"mh_p_ksads__cond__onset__chld__pres_dx",
"mh_y_ksads__cond__onset__adoles__past_dx",
"mh_y_ksads__cond__onset__adoles__pres_dx",
"mh_y_ksads__cond__onset__chld__past_dx",
"mh_y_ksads__cond__onset__chld__pres_dx")

#2.13 Collect all raw dx columns and sanity check existence
#2.13.1 Create a vector of the generated dx cols
dx_raw_cols <- unique(
  c(
    adhd_vars,
    gad_vars,
    socanx_vars,
    sepanx_vars,
    panic_vars,
    ocd_vars,
    dmdd_vars,
    mdd_vars,
    pdd_vars,
    ptsd_vars,
    odd_vars,
    cd_vars))

#2.13.2 Check whether any of the generated dx cols are missing necessary input cols
missing_cols <- setdiff(dx_raw_cols, names(dat))
if (length(missing_cols) > 0) {
  stop(
    "These expected KSADS dx columns are missing from dataset.csv:\n",
    paste0(" - ", missing_cols, collapse = "\n"))
}

```

## Helper functions + code normalization

```{r helper functions, echo = FALSE, warning = FALSE}

#3.1 Normalize KSADS codes, where 888 (module not administered) -> 0; and 555 / 999 / "" / NA -> NA
normalize_ksads_dx <- function(x) {
  x <- suppressWarnings(as.numeric(x))
  dplyr::case_when(x == 888 ~ 0,
                   is.na(x) ~ NA_real_,
                   x %in% c(555, 999) ~ NA_real_,
                   TRUE ~ x)
}

#3.2 Build umbrella dx (1 if any raw ==1; 0 if all observed ==0; NA if all missing)
mk_umbrella_dx <- function(df, vars) {
  case_when(
    rowSums(dplyr::select(df, dplyr::all_of(vars)) == 1, na.rm = TRUE) > 0 ~ 1,
    rowSums(!is.na(dplyr::select(
      df, dplyr::all_of(vars)
    ))) == 0 ~ NA_real_,
    TRUE ~ 0)
}

```

## Create umbrella dx variables + 4-group categorical variable

```{r create dx, echo = FALSE, warning = FALSE}

#4.1 Normalize raw dx columns in-place (processed file is allowed to deviate from raw)
dat2 <- dat %>%
  mutate(across(all_of(dx_raw_cols), ~ normalize_ksads_dx(.)))

#4.2 Create umbrella dx variables (0/1/NA)
dat2 <- dat2 %>%
  mutate(
    
    #4.2.1 Sub_Domain umbrella dx variables
    dx_adhd = mk_umbrella_dx(., adhd_vars),
    dx_gad = mk_umbrella_dx(., gad_vars),
    dx_socanx = mk_umbrella_dx(., socanx_vars),
    dx_sepanx = mk_umbrella_dx(., sepanx_vars),
    dx_panic = mk_umbrella_dx(., panic_vars),
    dx_ocd = mk_umbrella_dx(., ocd_vars),
    dx_dmdd = mk_umbrella_dx(., dmdd_vars),
    dx_mdd = mk_umbrella_dx(., mdd_vars),
    dx_pdd = mk_umbrella_dx(., pdd_vars),
    dx_ptsd = mk_umbrella_dx(., ptsd_vars),
    dx_odd = mk_umbrella_dx(., odd_vars),
    dx_cd = mk_umbrella_dx(., cd_vars))

#4.3 Any "other Dx of interest" composite (excluding BSD)
#4.3.1 Create a vector of the comorbidity dx
other_dx_umbrella_cols <- c(
  "dx_adhd",
  "dx_gad",
  "dx_socanx",
  "dx_sepanx",
  "dx_panic",
  "dx_ocd",
  "dx_dmdd",
  "dx_mdd",
  "dx_pdd",
  "dx_ptsd",
  "dx_odd",
  "dx_cd")

#4.3.2 Use the comorbidity vector to create the other dx of interest variable
dat2 <- dat2 %>%
  mutate(dx_other_any = case_when(
    rowSums(dplyr::select(., all_of(
      other_dx_umbrella_cols
    )) == 1, na.rm = TRUE) > 0 ~ 1,
    rowSums(!is.na(dplyr::select(
      ., all_of(other_dx_umbrella_cols)
    ))) == 0 ~ NA_real_,
    TRUE ~ 0))

#4.4 4-group categorical variable (uses validated any_bsd from outcome_variable_data.csv). If we can't determine BSD status or other-dx status -> NA
dat2 <- dat2 %>%
  mutate(
    dx_group4_code = case_when(
      any_bsd == 0 &
        dx_other_any == 0 ~ 1,
      
      #4.4.1 No Psych Dx (in dx set of interest)
      any_bsd == 1 & dx_other_any == 0 ~ 2,
      
      #4.4.2 BSD only
      any_bsd == 1 & dx_other_any == 1 ~ 3,
      
      #4.4.3 BSD + another Dx
      any_bsd == 0 & dx_other_any == 1 ~ 4,
      
      #4.4.4 Other Dx (no BSD)
      TRUE ~ NA_real_),
    dx_group4 = factor(
      dx_group4_code,
      levels = c(1, 2, 3, 4),
      labels = c(
        "No Dx (Dx-set of interest)",
        "BSD only",
        "BSD + other Dx",
        "Other Dx only (no BSD)"))
  )

#4.5 Create covariates needed for the supplementary occurrence analyses
#4.5.1 Establish ordered wave levels + a numeric wave index
wave_levels <- c("ses-00A", "ses-01A", "ses-02A", "ses-03A", "ses-04A", "ses-05A", "ses-06A")

#4.5.2 Create cleaned covariates to be used in analyses
dat2 <- dat2 %>%
  mutate(

    #4.5.2.1 Ensure session_id is ordered
    session_id = factor(session_id, levels = wave_levels),

    #4.5.2.2 Numeric wave index (00, 01, 02...) extracted from session_id
    wave_num = suppressWarnings(as.numeric(substr(as.character(session_id), 5, 6))),

    #4.5.2.3 Time-varying age covariate
    age_wave = suppressWarnings(as.numeric(ab_g_dyn__visit_age)),

    #4.5.2.4 Grand-mean center age across all observed age_wave values
    age_wave_gmc = age_wave - mean(age_wave, na.rm = TRUE),

    #4.5.2.5 Static covariates
    #4.5.2.5.1 Family ID
    family_id = as.character(ab_g_stc__design_id__fam),

    #4.5.2.5.2 Sex at birth
    sex = case_when(
      ab_g_stc__cohort_sex %in% c(1, "1", "M", "Male") ~ "Male",
      ab_g_stc__cohort_sex %in% c(2, "2", "F", "Female") ~ "Female",
      ab_g_stc__cohort_sex %in% c(3, "3", "Other", "Intersex") ~ "Intersex",
      TRUE ~ as.character(ab_g_stc__cohort_sex)),

    #4.5.2.5.3 Site (can fluctuate between visits but is rare)
    site_factor = factor(paste0("Site_", as.character(ab_g_dyn__design_site))),

    #4.5.2.5.4 Race/ethnicity (not to be used as covar)
    race_ethnicity = factor(as.character(ab_g_stc__cohort_ethnrace__meim))) %>%
  group_by(participant_id) %>%
  mutate(

    #4.5.2.6 Person-mean age + within-person deviation in case we later want age decomposition
    age_wave_person_mean = mean(age_wave, na.rm = TRUE),
    age_wave_cwc = age_wave - age_wave_person_mean) %>%
  ungroup() %>%
  mutate(

    #4.5.2.7 Between-person age component
    age_wave_between = age_wave_person_mean - mean(age_wave, na.rm = TRUE)
  )

#4.6 Create baseline diagnosis status covariates (to control for pre-existing dx at cluster formation)
#4.6.1 Extract baseline statuses at ses-00A
baseline_status_tbl <- dat2 %>%
  filter(session_id == "ses-00A") %>%
  select(
    participant_id,
    baseline_any_bsd = any_bsd,
    baseline_dx_other_any = dx_other_any,
    baseline_dx_group4_code = dx_group4_code,
    baseline_dx_group4 = dx_group4) %>%
  distinct(participant_id, .keep_all = TRUE) %>%
  mutate(

    #4.6.1.1 Binary baseline dx flag: any dx in the set of interest (BSD and/or other dx umbrellas)
    baseline_any_dx = case_when(
      is.na(baseline_dx_group4_code) ~ NA_real_,
      baseline_dx_group4_code == 1 ~ 0,
      baseline_dx_group4_code %in% c(2, 3, 4) ~ 1,
      TRUE ~ NA_real_)
  )

#4.6.2 Attach baseline statuses back to all waves
dat2 <- dat2 %>%
  left_join(baseline_status_tbl, by = "participant_id")

```

## Quick QC tables

```{r QC check, echo = FALSE, warning = FALSE}

#5.1 Sanity check of whether we have any_bsd
if (!("any_bsd" %in% names(dat2))) {
  stop(
    "any_bsd not present after merge. Check participant_id/session_id keys and outcome_variable_data.csv.")
}

#5.2 Prevalence by wave/session
#5.2.0 Define analytic waves (for this project, these are the waves we will actually analyze)
analytic_waves <- c("ses-00A", "ses-02A", "ses-04A", "ses-06A")

#5.2.1 Create a QC-only dataset restricted to analytic waves (prevents NA-heavy intermediate waves from dominating output)
dat2_qc <- dat2 %>%
  dplyr::filter(session_id %in% analytic_waves)

#5.2.2 Log prevalence with explicit denominators (prevents NaN when a wave has no observed data)
qc_by_wave <- dat2_qc %>%
  group_by(session_id) %>%
  summarise(
    n = n(),
    n_any_bsd_nonmiss = sum(!is.na(any_bsd)),
    any_bsd_pct = if_else(n_any_bsd_nonmiss > 0, mean(any_bsd == 1, na.rm = TRUE), NA_real_),
    n_other_any_nonmiss = sum(!is.na(dx_other_any)),
    other_dx_any_pct = if_else(n_other_any_nonmiss > 0, mean(dx_other_any == 1, na.rm = TRUE), NA_real_),
    n_group4_nonmiss = sum(!is.na(dx_group4_code)),
    group4_na_pct = 1 - (n_group4_nonmiss / n),
    .groups = "drop"
  )

#5.2.3 Print QC prevalence table
knitr::kable(qc_by_wave, caption = "QC: BSD / Other Dx / Group4 Missingness (analytic waves only)", digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#5.3 Group4 distribution by wave/session (exclude NA rows from the distribution table)
#5.3.1 Log the composition of the 4-group categorical var by wave (observed only)
group_dist <- dat2_qc %>%
  dplyr::filter(!is.na(dx_group4)) %>%
  count(session_id, dx_group4) %>%
  group_by(session_id) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

#5.3.2 Print distribution table
knitr::kable(group_dist, caption = "QC: dx_group4 Distribution by Session (observed only; analytic waves)", digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#5.4 Skim just the derived columns
#5.4.1 Create a list of columns to check with skimr
derived_cols <- c(
  "participant_id",
  "session_id",
  other_dx_umbrella_cols,
  "dx_other_any",
  "any_bsd",
  "dx_group4_code",
  "dx_group4")

#5.4.2 Check relevant column distributions with skimr
suppressWarnings(skimr::skim_without_charts(
  dat2 %>%
    filter(session_id %in% analytic_waves) %>%
    dplyr::select(any_of(derived_cols))))

#5.5 Join integrity checks (protect against unintended row duplication from the left_join)
#5.5.1 Check for duplicate keys in raw_data
raw_key_dups <- raw_data %>%
  count(participant_id, session_id) %>%
  filter(n > 1)

#5.5.2 Check for duplicate keys in outcome_data
outcome_key_dups <- outcome_data %>%
  count(participant_id, session_id) %>%
  filter(n > 1)

#5.5.3 Print diagnostics + hard-stop if join would expand rows
cat("\n#5.5 Join integrity\n")
cat("raw_data nrow:", nrow(raw_data), "\n")
cat("outcome_data nrow:", nrow(outcome_data), "\n")
cat("dat (joined) nrow:", nrow(dat), "\n\n")

#5.5.3.1 Print whether any subject IDs were duplicated in the merging process
if (nrow(outcome_key_dups) > 0) {
  cat("WARNING: Duplicate keys in outcome_data (showing first 10):\n")
  print(head(outcome_key_dups, 10))
}

#5.5.3.2 Print whether the number of rows in the merged data matches the number of rows in the raw data
if (nrow(dat) != nrow(raw_data)) {
  stop(
    "Join integrity failure: nrow(dat) != nrow(raw_data). This suggests a join expansion.\n",
    "Resolve duplicate (participant_id, session_id) keys before interpreting dx distributions."
  )
}

#5.6 Verify that dx mapping does not double-count any *raw* KSADS column across umbrellas
#5.6.1 Create a named mapping list of umbrella dx to raw columns
dx_map <- list(
  ADHD = adhd_vars,
  GAD = gad_vars,
  SOCANX = socanx_vars,
  SEPANX = sepanx_vars,
  PANIC = panic_vars,
  OCD = ocd_vars,
  DMDD = dmdd_vars,
  MDD = mdd_vars,
  PDD = pdd_vars,
  PTSD = ptsd_vars,
  ODD = odd_vars,
  CD = cd_vars)

#5.6.2 Identify any raw columns that appear in more than one umbrella list
raw_col_membership <- tibble(
  raw_col = unlist(dx_map),
  umbrella = rep(names(dx_map), lengths(dx_map))) %>%
  count(raw_col, name = "n_umbrellas") %>%
  dplyr::filter(n_umbrellas > 1)

#5.6.2.1 Print whether there were any variables that were used duplicatively
if (nrow(raw_col_membership) == 0) {
  cat("\n#5.6 Raw-column overlap check: PASS (no raw KSADS columns reused across umbrellas)\n")
} else {
  cat("\n#5.6 Raw-column overlap check: FAIL (raw columns reused across umbrellas)\n")
  print(raw_col_membership)
}

#5.7 Composition QC of how many + which umbrellas are driving dx_other_any and dx_group4
#5.7.1 Create per-row count of endorsed umbrella dx (among the 12 comorbidity umbrellas)
dat2 <- dat2 %>%
  mutate(
    dx_n_other = rowSums(dplyr::select(., all_of(other_dx_umbrella_cols)) == 1, na.rm = TRUE),
    dx_n_other_nonmiss = rowSums(!is.na(dplyr::select(., all_of(other_dx_umbrella_cols)))))

#5.7.2 Verify group logic has no internal contradictions. "No Dx" + "BSD only" groups should have dx_n_other == 0 by construction
contradictions <- dat2 %>%
  filter(dx_group4 %in% c("No Dx (Dx-set of interest)", "BSD only")) %>%
  filter(dx_n_other > 0)

cat("\n#5.7 Group logic contradictions (should be 0 rows):", nrow(contradictions), "\n")
if (nrow(contradictions) > 0) {
  print(head(contradictions %>%
               select(participant_id, session_id, dx_group4, dx_n_other, all_of(other_dx_umbrella_cols)), 10))
}

#5.7.3 Create a table with the overall umbrella dx prevalence by session (with denominators; analytic waves only)
dx_prev_overall <- dat2_qc %>%
  dplyr::select(session_id, all_of(other_dx_umbrella_cols)) %>%
  pivot_longer(cols = all_of(other_dx_umbrella_cols),
               names_to = "umbrella_dx",
               values_to = "val") %>%
  group_by(session_id, umbrella_dx) %>%
  summarise(
    n_nonmiss = sum(!is.na(val)),
    n_yes = sum(val == 1, na.rm = TRUE),
    pct_yes = if_else(n_nonmiss > 0, n_yes / n_nonmiss, NA_real_),
    .groups = "drop") %>%
  dplyr::filter(n_nonmiss > 0) %>%
  arrange(session_id, desc(pct_yes))

#5.7.3.1 Print a kable of the overall umbrella dx prevalence by session
knitr::kable(dx_prev_overall, caption = "QC: Umbrella Dx Prevalence by Session (overall; denominators exclude NA)", digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#5.7.4 Create a table of the umbrella dx prevalence within dx_group4 strata (focus on strata where other dx should exist)
dx_prev_by_group <- dat2_qc %>%
  dplyr::filter(dx_group4 %in% c("BSD + other Dx", "Other Dx only (no BSD)")) %>%
  dplyr::select(session_id, dx_group4, all_of(other_dx_umbrella_cols)) %>%
  pivot_longer(cols = all_of(other_dx_umbrella_cols),
               names_to = "umbrella_dx",
               values_to = "val") %>%
  group_by(session_id, dx_group4, umbrella_dx) %>%
  summarise(
    n_nonmiss = sum(!is.na(val)),
    n_yes = sum(val == 1, na.rm = TRUE),
    pct_yes = if_else(n_nonmiss > 0, n_yes / n_nonmiss, NA_real_),
    .groups = "drop") %>%
  dplyr::filter(n_nonmiss > 0) %>%
  arrange(session_id, dx_group4, desc(pct_yes))

#5.7.4.1 Print a kable of the umbrella dx prevalence within dx_group4 strata
knitr::kable(
  dx_prev_by_group,
  caption = "QC: Umbrella Dx Prevalence by Session within Other-Dx Strata (denominators exclude NA)",
  digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#5.7.5 Create a table of distribution of number of comorbidity umbrellas endorsed (dx_n_other) by session + dx_group4
dx_count_dist <- dat2 %>%
  filter(dx_group4 %in% c("BSD + other Dx", "Other Dx only (no BSD)")) %>%
  group_by(session_id, dx_group4, dx_n_other) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(session_id, dx_group4) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

#5.7.5.1 Print the kable of umbrella comorbidity distributions
knitr::kable(dx_count_dist, caption = "QC: Number of Umbrella Dx Endorsed among Other-Dx Strata", digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#5.7.6 Plot umbrella dx prevalence within the two "other dx present" strata (analytic waves; observed only)
#5.7.6.1 Create a dataframe for the umbrella dx prevalence plot
dx_prev_plot_dat <- dx_prev_by_group %>%
  mutate(pct_yes = pct_yes * 100) %>%
  dplyr::filter(!is.na(pct_yes))

#5.7.6.2 Plot the umbrella dx prevalence
ggplot(dx_prev_plot_dat, aes(x = umbrella_dx, y = pct_yes)) +
  geom_col() +
  coord_flip() +
  facet_grid(dx_group4 ~ session_id) +
  labs(
    title = "Umbrella Dx Prevalence within Other-Dx Strata (analytic waves; observed only)",
    x = "Umbrella Dx",
    y = "Percent endorsed (among non-missing)"
  )

#5.8 Characterize missingness for dx_group4 on analytic waves
#5.8.1 Create a vector of which waves we are interested in characterizing
analytic_waves <- c("ses-00A","ses-02A","ses-04A","ses-06A")

#5.8.2 Create a table outlining why variables are missing at each timepoint of interest
dx_group4_na_reason <- dat2 %>%
  filter(session_id %in% analytic_waves) %>%
  mutate(
    any_bsd_na = is.na(any_bsd),
    other_any_na = is.na(dx_other_any),
    group4_na = is.na(dx_group4_code),
    na_reason = case_when(
      !group4_na ~ "Observed",
      any_bsd_na & other_any_na ~ "Missing: any_bsd + dx_other_any",
      any_bsd_na ~ "Missing: any_bsd",
      other_any_na ~ "Missing: dx_other_any",
      TRUE ~ "Missing: other")) %>%
  count(session_id, na_reason) %>%
  group_by(session_id) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

#5.8.3 Print a kable outlining why variables are missing at each timepoint of interest
knitr::kable(dx_group4_na_reason, caption = "QC: Reasons dx_group4 is Missing (analytic waves only)", digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#5.9 Check top comorbidity combinations among BSD cases (by wave) to sanity-check patterns. To note, NA dx values are simply not counted as present in the combo string
bsd_combo_top <- dat2_qc %>%
  dplyr::filter(any_bsd == 1) %>%
  rowwise() %>%
  mutate(
    dx_combo = {
      vals <- c_across(all_of(other_dx_umbrella_cols))
      n_nonmiss <- sum(!is.na(vals))
      idx_yes <- which(vals == 1)
      
      if (n_nonmiss == 0) {
        NA_character_
      } else if (length(idx_yes) == 0) {
        "BSD only (no other dx endorsed among observed)"
      } else {
        paste(other_dx_umbrella_cols[idx_yes], collapse = " + ")
      }
    }
  ) %>%
  ungroup() %>%
  dplyr::filter(!is.na(dx_combo)) %>%
  count(session_id, dx_combo, sort = TRUE) %>%
  group_by(session_id) %>%
  mutate(pct = n / sum(n)) %>%
  slice_max(order_by = n, n = 10, with_ties = FALSE) %>%
  ungroup()

#5.9.1 Print top comorbidity combo kable
knitr::kable(bsd_combo_top, caption = "QC: Top 10 Comorbidity Combinations Among BSD Cases (analytic waves; observed only)", digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#5.10 Comorbidity structure QC for "Other Dx only (no BSD)" (analytic waves; observed only)
#5.10.1 Top combinations among Other-Dx-only group
other_only_combo_top <- dat2_qc %>%
  dplyr::filter(dx_group4 == "Other Dx only (no BSD)") %>%
  rowwise() %>%
  mutate(
    dx_combo = {
      vals <- c_across(all_of(other_dx_umbrella_cols))
      n_nonmiss <- sum(!is.na(vals))
      idx_yes <- which(vals == 1)
      
      if (n_nonmiss == 0) {
        NA_character_
      } else if (length(idx_yes) == 0) {
        "Other-Dx-only, but no umbrella dx endorsed (among observed)"
      } else {
        paste(other_dx_umbrella_cols[idx_yes], collapse = " + ")
      }
    }
  ) %>%
  ungroup() %>%
  dplyr::filter(!is.na(dx_combo)) %>%
  count(session_id, dx_combo, sort = TRUE) %>%
  group_by(session_id) %>%
  mutate(pct = n / sum(n)) %>%
  slice_max(order_by = n, n = 10, with_ties = FALSE) %>%
  ungroup()

#5.10.2 Print a kable of top combinations among Other-Dx-only group
knitr::kable(other_only_combo_top, caption = "QC: Top 10 Comorbidity Combinations Among Other-Dx-Only Group (analytic waves; observed only)", digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#5.10.3 Function to calculate top dyads (pairs) among Other-Dx-only group
calc_top_dyads <- function(df, dx_cols, top_k = 15) {
  pairs <- combn(dx_cols, 2, simplify = FALSE)
  out <- lapply(pairs, function(p) {
    a <- df[[p[1]]]
    b <- df[[p[2]]]
    n_nonmiss <- sum(!is.na(a) & !is.na(b))
    n_both <- sum(a == 1 & b == 1, na.rm = TRUE)
    tibble(
      dx1 = p[1],
      dx2 = p[2],
      n_nonmiss = n_nonmiss,
      n_both = n_both,
      pct_both = if_else(n_nonmiss > 0, n_both / n_nonmiss, NA_real_))}) %>%
    bind_rows() %>%
    dplyr::filter(n_nonmiss > 0) %>%
    arrange(desc(pct_both), desc(n_both)) %>%
    slice_head(n = top_k)
  out
}

#5.10.4 Create a table of the cleaned top dyads (no BSD)
other_only_top_dyads <- dat2_qc %>%
  dplyr::filter(dx_group4 == "Other Dx only (no BSD)") %>%
  group_by(session_id) %>%
  group_modify(~ calc_top_dyads(.x, other_dx_umbrella_cols, top_k = 12)) %>%
  ungroup()

#5.10.5 Print a kable outlining the cleaned top dyads (no BSD) 
knitr::kable(other_only_top_dyads, caption = "QC: Top Umbrella Dx Dyads in Other-Dx-Only Group (analytic waves; observed only)", digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```

## Write supplementary processed dataset

```{r write data, echo = FALSE, warning = FALSE}

#6.1 Only retain essential columns for the supplementary analysis
supplementary_outcome_data <- dat2 %>%
  dplyr::select(

    #6.1.1 Subject ID keys
    participant_id,
    session_id,
    wave_num,

    #6.1.2 Covariates (raw + cleaned)
    cluster,
    ab_g_dyn__design_site,
    ab_g_dyn__visit_age,
    ab_g_stc__cohort_ethnrace__meim,
    ab_g_stc__cohort_sex,
    ab_g_stc__design_id__fam,
    site_factor,
    race_ethnicity,
    sex,
    family_id,
    age_wave,
    age_wave_gmc,
    age_wave_person_mean,
    age_wave_cwc,
    age_wave_between,

    #6.1.3 Baseline-status covariates
    baseline_any_dx,
    baseline_any_bsd,
    baseline_dx_other_any,
    baseline_dx_group4_code,
    baseline_dx_group4,

    #6.1.4 Suicidality outcomes
    si_passive,
    si_active,
    sa,
    nssi,

    #6.1.5 BD outcomes
    bipolar_I,
    bipolar_II,
    bd_nos,
    any_bsd,

    #6.1.6 Umbrella dx + composites + 4-group variable
    dx_adhd,
    dx_gad,
    dx_socanx,
    dx_sepanx,
    dx_panic,
    dx_ocd,
    dx_dmdd,
    dx_mdd,
    dx_pdd,
    dx_ptsd,
    dx_odd,
    dx_cd,
    dx_other_any,
    dx_group4_code,
    dx_group4)

#6.2 Write the full dataset (raw + validated outcomes + derived dx vars)
write.csv(supplementary_outcome_data, out_path, row.names = FALSE)

#6.3 Print that the dataset was written and where
cat("Wrote:", out_path, "\n")

```
