---
title: "Mood Disorder & Suicidality ~ Clustered Risk Group Modeling"
author: "Sam A. Sievertsen"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(warning = FALSE, message = NA, comment = "", dpi = 720)

```

```{r environment, echo = FALSE, include = FALSE, warning = FALSE}

# Load necessary packages + environmental variables
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(easystats)
library(performance)
library(cluster)
library(clustMixType)
library(ggplot2)
library(viridis)
library(lme4)
library(glmmTMB)
library(lmerTest)
library(lmtest)
library(sandwich)
library(car)
library(emmeans)
options(scipen = 999, digits = 8)

# Read in clustered risk variable + outcome data for analysis
bipolar_suicidality_risk_data <- read.csv("../../data/data_processed/bipolar_suicidality_risk_analysis_data.csv")

```

## Data Wrangling

```{r data wrangling, warning = FALSE}

## Data Wrangling for Analyses ## 

#1. Create a dataframe for primary analyses
#1.1 Retain only the columns of interest to the analyses
bipolar_suicidality_risk_primary_analysis_data <- bipolar_suicidality_risk_data %>% 
  dplyr::select(c(src_subject_id, sex, family_id, site_name, cluster, Y4_bipolar_I_dx, Y4_bipolar_II_dx, Y4_any_bipolar_dx, Y4_suicidal_thinking, Y4_suicide_attempt, Y4_nssi, baseline_bipolar_I_dx, baseline_bipolar_II_dx, baseline_any_bipolar_dx, baseline_suicidal_thinking, baseline_suicide_attempt, baseline_nssi))

#1.2 Alter the names of clusters to better reflect their membership during modeling
bipolar_suicidality_risk_primary_analysis_data <- bipolar_suicidality_risk_primary_analysis_data %>%
  mutate(cluster = paste0("cluster_", cluster))

#1.3 Ensure columns of interest are the correct data type
#1.31 Set relevant columns to factor type
bipolar_suicidality_risk_primary_analysis_data <- bipolar_suicidality_risk_primary_analysis_data %>% 
  mutate(across(c("sex", "family_id", "site_name", "cluster", "Y4_bipolar_I_dx", "Y4_bipolar_II_dx", "Y4_any_bipolar_dx", "Y4_suicidal_thinking", "Y4_suicide_attempt", "Y4_nssi", "baseline_bipolar_I_dx", "baseline_bipolar_II_dx", "baseline_any_bipolar_dx", "baseline_suicidal_thinking", "baseline_suicide_attempt", "baseline_nssi"), as.factor))

#1.32 Set reference level of the cluster variable
bipolar_suicidality_risk_primary_analysis_data$cluster <- relevel(bipolar_suicidality_risk_primary_analysis_data$cluster, ref = "cluster_3")

#2. Create a dataframe for secondary analyses
#2.1 Retain only the columns of interest to the analyses
dimensional_psychopathology_risk_secondary_analysis_data <- bipolar_suicidality_risk_data %>%
  dplyr::select(c(src_subject_id, sex, family_id, site_name, cluster, baseline_cbcl_scr_dsm5_depress_t, baseline_cbcl_scr_dsm5_anxdisord_t, baseline_cbcl_scr_syn_attention_t, baseline_cbcl_scr_syn_aggressive_t, cbcl_scr_dsm5_depress_t, cbcl_scr_dsm5_anxdisord_t, cbcl_scr_syn_attention_t, cbcl_scr_syn_aggressive_t))

#2.2 Alter the names of clusters to better reflect their membership during modeling
dimensional_psychopathology_risk_secondary_analysis_data <- dimensional_psychopathology_risk_secondary_analysis_data %>%
  mutate(cluster = paste0("cluster_", cluster))

#2.3 Ensure columns of interest are the correct data type
#2.31 Set relevant columns to factor type
dimensional_psychopathology_risk_secondary_analysis_data <- dimensional_psychopathology_risk_secondary_analysis_data %>% 
  mutate(across(c("sex", "family_id", "site_name", "cluster"), as.factor))

#2.32 Set reference level of the cluster variable
dimensional_psychopathology_risk_secondary_analysis_data$cluster <- relevel(dimensional_psychopathology_risk_secondary_analysis_data$cluster, ref = "cluster_3")

#3. Create a dataframe for demographic data summarization
bipolar_suicidality_risk_demographic_data <- bipolar_suicidality_risk_data %>%
  dplyr::select(c())

```

### Note on Data to be Used in This Analysis

As outlined in `bipolar_suicidality_risk_analysis_data_wrangling`, we have weighed multiple considerations related to using aggregated diagnostic data from follow-up years 1-3 (Y1-3) vs data collected solely at year 4 (Y4) followup. 

For the purpose of analyses to be presented at SOBP 2025, we have elected to use the Y4 data as part of release version 5.1. This is primarily due to the following reasons subset from those hinted at above: 

- The availability of mood disorders at just Y2 followup, and availability of suicidality data at Y1-Y3 being a difficult methodological question to grapple with/justify the use of

- The unexpectedly high rate of bipolar disorders at Y2 [which has been noted before and is advised to be validated/corrected using other diagnostic information](https://www.sciencedirect.com/science/article/pii/S1878929321001201), coupled with the subsequent need to address that issue. Long term this is a good strategy, but likely too time/resource intensive for right now

- Out of all available timepoints released for the ABCD study thus far, Y4 most closely aligns with when (theoretically) we might start to see an uptick in the emergence of bipolar disorders

  - Should release 6.0 occur prior to SOBP, it will be much easier to re-analyze the 4Y (and/or 6Y) data prior to creating the poster than it would be to switch the pipeline from Y1-Y3

## Primary Analysis: Risk for Bipolar Disorder(s) & Suicidality

### Note on Model(s) to be Used in This Analysis

In preliminary testing of the analysis methods, logistic regressions were ran using both a standard `glm()` with fixed effects of cluster, sex, and diagnosis of interest included; as well as a `glmer()` including both the aforementioned fixed effects and random effects for site and family ID. 

After iteratively testing the performance of these methods, when including site as a random effect in the `glmer()` model, I observed model convergence issues and poor performance, suggesting that site-level variation did not meaningfully contribute to explaining the outcome; as well as issues with the random effect of family ID reducing the estimates of the fixed effects in the model. 

Given this, I tested an alternative mixed-effects model using `glmmTMB()`, which provides more flexible parameter estimation. However, this approach yielded implausible odds ratios and an inflated intraclass correlation coefficient (*ICC* ≈ 1.0), indicating that the model was over-partitioning variance into the random effect of family, effectively absorbing meaningful between-subject variability and distorting fixed effect estimates.

To address these issues, I implemented a generalized linear model (`glm()`) with cluster-robust standard errors (`vcovCL()`), which adjusts for within-family correlation while maintaining the interpretability of fixed effects. This approach ensures valid statistical inference by correcting standard errors without absorbing variance into the random effect structure, ultimately providing more stable, reliable, and interpretable effect estimates.

TL;DR: Given the observed convergence issues with `glmer()`, the poorly calibrated effect estimates from `glmmTMB()`, and the (relative) statistical robustness of `glm()` with `vcovCL()`, the latter emerged as the optimal analytic approach for our data. This is summarized in the table below: 

| Model | Pros | Cons |
|--------|------|------|
| **`glm()` (without robust SEs)** | ✅ Simple, ✅ Fast, ✅ Interpretable | ❌ Ignores family structure |
| **`glmer()` (random effect for family)** | ✅ Accounts for clustering, ✅ Random effects | ❌ Overfits to family ID, ❌ Convergence issues, ❌ Shrinks fixed effects |
| **`glmmTMB()` (random effect for family)** | ✅ More stable than `glmer()` | ❌ Still over-explains family ID, ❌ Produces unrealistic ORs |
| **`glm()` + `vcovCL()`** (**Best Option**) | ✅ Accounts for family clusters, ✅ Keeps estimates stable, ✅ No convergence issues | ❌ Doesn't estimate within-family variance explicitly |

## Primary Analysis: Logistic Regression for Binary Diagnostic Outcomes

**Outcomes Examined:**  
  
Six diagnostic outcomes were evaluated: *bipolar I dx*, *bipolar II dx*, *any bipolar dx*, *suicidal thinking*, *suicide attempt*, and *nssi*.

**Model Specification:**  
  
For each outcome, a logistic regression model was fit using predictors that include a cluster variable, the corresponding baseline diagnosis covariate, and sex. The binomial family was specified to model binary diagnostic outcomes.

**Template model:** 

  $$
  \log\left(\frac{P(Y=1)}{1-P(Y=1)}\right) = \beta_0 + \beta_1\,\text{Cluster} + \beta_2\,\text{Baseline} + \beta_3\,\text{Sex} + \varepsilon
  $$
    
  where: 
    
  - \(Y\) is the binary diagnostic outcome
  
  - \(\beta_0\) is the intercept
  
  - \(\beta_1\) represents the effect of cluster membership
  
  - \(\beta_2\) represents the effect of the corresponding baseline covariate (derived by replacing the "Y4\_" prefix with "baseline\_")
  
  - \(\beta_3\) represents the effect of sex
  
  - \(\varepsilon\) is the error term

**Robust Standard Error Analysis:**  
  
Robust (clustered) standard errors were computed using a clustered variance estimator (vcovCL) for effects of *family_id*. The coefficient tests (via coeftest) were then used to extract the robust p-values, which were combined with the exponentiated coefficients to obtain odds ratios and corresponding 95% confidence intervals.

**Result Extraction and FDR Correction:**  

For each outcome, the odds ratios, standard errors, lower and upper 95% confidence bounds, and p-values were extracted. P-values were then adjusted for multiple comparisons using the false discovery rate (FDR) procedure applied within each outcome analysis. 

**Compilation of Final Results:**  

The results were compiled into a single, clean table where results were reformatted and cluster information was extracted (with Cluster 3 relabeled as the reference group). The final output is presented below: 

```{r primary analysis, echo = FALSE, warning = FALSE}

## Logistic Regression for Binary Diagnostic Outcomes ##

#1. Generate relevant summary statistics for each diagnosis of interest by cluster membership
#1.1 Define diagnostic outcome variables
diagnostic_outcome_vars <- c("Y4_bipolar_I_dx", "Y4_bipolar_II_dx", "Y4_any_bipolar_dx",
                  "Y4_suicidal_thinking", "Y4_suicide_attempt", "Y4_nssi")

#1.2 Create an outcome variable (prevalence) summary table by cluster membership
primary_analysis_variable_summary <- bipolar_suicidality_risk_primary_analysis_data %>%
  dplyr::select(cluster, all_of(diagnostic_outcome_vars)) %>%
  
  #1.21 Pivot the diagnosis columns into a long format
  pivot_longer(cols = all_of(diagnostic_outcome_vars), 
               names_to = "Diagnosis", 
               values_to = "Status") %>%
  
  #1.22 Clean the Diagnosis column: remove "Y4_" prefix and replace underscores with spaces
  mutate(Diagnosis = gsub("_", " ", gsub("^Y4_", "", Diagnosis))) %>%
  
  #1.23 Remove missing values in the diagnosis variable
  filter(!is.na(Status)) %>%
  
  #1.24 Group by cluster and diagnosis, and calculate counts and percentages
  group_by(cluster, Diagnosis) %>%
  summarize(N = sum(Status == 1, na.rm = TRUE),
            Total = n(),
            Percent = round((N / Total) * 100, 1),
            .groups = "drop") %>%
  
  #1.25 Create a formatted summary column
  mutate(Summary = paste0(N, " (", Percent, "%)")) %>%
  dplyr::select(cluster, Diagnosis, Summary) %>%
  
  #1.26 Pivot wider so that each cluster is a separate column
  pivot_wider(names_from = cluster, values_from = Summary)

#1.3 Print the primary analysis summary table
primary_analysis_variable_summary %>%
  kable(caption = "Prevalence of Diagnoses by Cluster Membership",
        align = "l", 
        col.names = c("Diagnosis", "Cluster 3 (Ref)", "Cluster 1", "Cluster 2", "Cluster 4", "Cluster 5")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#2. Run the logistic regressions for each diagnostic outcome variable 
#2.1 Define the function to run logistic regression with robust standard errors
run_logistic_models <- function(data, diagnostic_outcomes, cluster_var) {
  
  #2.11 Initialize an empty list to store model results
  logistic_results_list <- list()
  
  #2.12 Iterate over each diagnostic outcome variable
  for (diagnostic_outcome in diagnostic_outcomes) {
    
    #2.121 Define the corresponding baseline covariate name
    baseline_var <- gsub("^Y4_", "baseline_", diagnostic_outcome)
    
    #2.122 Ensure the baseline variable exists in the dataset
    if (!baseline_var %in% colnames(data)) {
      warning(paste("Skipping", diagnostic_outcome, "as baseline variable", baseline_var, "not found in dataset"))
      next
    }
    
    #2.123 Fit the logistic regression model
    logistic_regression_model <- glm(as.formula(paste(diagnostic_outcome, "~ cluster +", baseline_var, "+ sex")), 
                 data = data, 
                 na.action = na.omit, 
                 family = "binomial")
    
    #2.124 Compute robust standard errors using clustered variance estimator
    robust_se <- vcovCL(logistic_regression_model, cluster = data[[cluster_var]])
    
    #2.125 Extract robust coefficients
    robust_coeffs <- coeftest(logistic_regression_model, vcov = robust_se)
    
    #2.126 Compute odds ratios, se, and confidence intervals
    odds_ratios <- exp(coef(logistic_regression_model))
    se <- sqrt(diag(robust_se)) 
    conf_intervals <- exp(confint(logistic_regression_model, vcov = robust_se))
    p_values <- robust_coeffs[, 4]
    
    #2.127 Store results in a dataframe
    logistic_regression_model_results <- data.frame(
      Outcome = diagnostic_outcome,
      Predictor = names(odds_ratios),
      Odds_Ratio = odds_ratios,
      SE = se,
      CI_Lower = conf_intervals[, 1],
      CI_Upper = conf_intervals[, 2],
      P_Value = p_values
    )
    
    #2.128 Append results to the list
    logistic_results_list[[diagnostic_outcome]] <- logistic_regression_model_results
  }
  
  #2.13 Combine all results into a single dataframe
  merged_logistic_regression_results <- bind_rows(logistic_results_list)
  
  return(merged_logistic_regression_results)
}

#2.2 Run the logistic regression function on the primary analysis dataset
primary_analysis_results_raw <- run_logistic_models(data = bipolar_suicidality_risk_primary_analysis_data, 
                                        diagnostic_outcomes = diagnostic_outcome_vars, 
                                        cluster_var = "family_id")

#3. Clean + display the results of the logistic regression models
#3.1 Filter the primary analysis results to only include those pertaining to the clusters of interest
primary_cluster_analysis_results <- primary_analysis_results_raw %>%
  filter(grepl("^cluster", Predictor))

#3.2 FDR correct p values within each outcome analysis
primary_cluster_analysis_results <- primary_cluster_analysis_results %>%
  group_by(Outcome) %>%
  mutate(FDR_P_Value = p.adjust(P_Value, method = "fdr")) %>%
  ungroup()

#3.3 Clean and reformat the logistic regression results
#3.31 Create a table with cleaned + reformatted OR, SE, and p values
primary_analysis_results_table <- primary_cluster_analysis_results %>%
  
  #3.311 Clean Outcome names: remove "Y4_" prefix and change underscores to spaces
  mutate(Outcome = gsub("^Y4_", "", Outcome),
         Outcome = gsub("_", " ", Outcome)) %>%
  
  #3.312 Extract the cluster level from the Predictor column and relabel the reference
  mutate(Cluster = sub(".*_(\\d+)$", "\\1", Predictor),
         Cluster = paste("Cluster", Cluster),
         Cluster = ifelse(Cluster == "Cluster 3", "Cluster 3 (Ref)", Cluster)) %>%
  
  #3.313 Format numeric values: round odds ratios, SE, and confidence limits
  mutate(Odds_Ratio = round(Odds_Ratio, 2),
         SE = round(SE, 2),
         CI_Lower = round(CI_Lower, 2),
         CI_Upper = round(CI_Upper, 2),
         
         #3.3131 Format the original p-values: show values < 0.001 as "<0.001"
         P_Value = ifelse(P_Value < 0.001, "<0.001", round(P_Value, 3)),
         
         #3.3132 Format the FDR corrected p-values similarly
         FDR_P_Value = ifelse(FDR_P_Value < 0.001, "<0.001", round(FDR_P_Value, 3))) %>%
  
  #3.314 Combine CI_Lower and CI_Upper into a single column for the 95% confidence interval
  mutate(`95% CI` = paste0("(", CI_Lower, ", ", CI_Upper, ")")) %>%
  
  #3.315 Select the columns for the final table, with FDR P-Value as the last column
  dplyr::select(Outcome, Cluster, Odds_Ratio, SE, `95% CI`, P_Value, FDR_P_Value)

#3.32 Print the cleaned and reformatted logistic regression results table
primary_analysis_results_table %>%
  kable(caption = "Logistic Regression Results: Outcome ~ Cluster (Relative to Cluster 3)",
        align = "l", 
        col.names = c("Outcome", "Cluster", "Odds Ratio", "SE", "95% CI", "P-Value", "FDR P-Value"),
        row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```

### Plotting Primary Analysis Results

```{r primary analysis plots, echo = FALSE, warning = FALSE, fig.width = 10, fig.height = 6}

## Plot the Results of the Primary Logistic Regression Analyses ##

#1. Prepare the data by cleaning Outcome names, extracting Cluster info, and flagging significance
primary_analysis_plot_data <- primary_cluster_analysis_results %>%
  mutate(
    
    #1.1 Clean outcome names: remove "Y4_" prefix and replace underscores with spaces
    Outcome = gsub("^Y4_", "", Outcome),
    Outcome = gsub("_", " ", Outcome),
    
    #1.2 Extract the cluster number from the Predictor column and format it
    Cluster = sub(".*_(\\d+)$", "\\1", Predictor),
    Cluster = paste("Cluster", Cluster),
    
    #1.3 Flag significant odds ratios (using p < 0.05 as the threshold)
    Significant = if_else(P_Value < 0.05, "Significant p<0.05", "Not Significant p>0.05")
  ) %>%
  
  #1.3 For rows with a missing lower CI (e.g., bipolar II dx, Cluster 4), set Odds_Ratio to NA.
  mutate(Odds_Ratio = ifelse(is.na(CI_Lower), NA, Odds_Ratio))

#2. Plot the primary analysis results (OR)
#2.1 Create the odds ratio forest plot of primary analysis results
primary_analysis_plot <- ggplot(primary_analysis_plot_data, aes(x = Odds_Ratio, y = Cluster, color = Cluster, shape = Significant)) +
  geom_point(size = 3) +
  
  #2.12 Only draw error bars where the lower confidence bound is available
  geom_errorbarh(data = filter(primary_analysis_plot_data, !is.na(CI_Lower)),
                 aes(xmin = CI_Lower, xmax = CI_Upper), height = 0.2) +
  
  #2.13 Facet by Outcome with free x scales for interpretability
  facet_wrap(~ Outcome, scales = "free_x", ncol = 3) +
  
  #2.14 Add a vertical dashed line at 1.0 that stands out in red
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", size = 1) +
  
  #2.15 Set x-axis on log scale with the specified breaks
  scale_x_log10(breaks = c(0.5, 1, 2, 5, 10)) +
  labs(
    x = "Odds Ratio",
    y = "Cluster",
    title = "Odds Ratios for Diagnostic Outcomes by Cluster (Relative to Cluster 3)"
  ) +
  
  #2.16 Use the viridis color palette for a deep and vibrant scheme
  scale_color_viridis_d(option = "viridis") +
  
  #2.17 Map shapes: use circles (16) for non-significant and triangles (17) for significant estimates
  scale_shape_manual(values = c("Not Significant p>0.05" = 16, "Significant p<0.05" = 17)) +
  theme_bw(base_size = 12) +
  theme(
    
    #2.181 Bold axis titles and text
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(face = "bold", size = 10),
    
    #2.182 Only show major vertical gridlines (x-axis) at the specified breaks
    panel.grid.major.x = element_line(color = "grey80", size = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    
    #2.183 Add a clear border around each facet panel
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    
    #2.184 Bold the facet strip text and add a white background with a black border
    strip.background = element_rect(fill = "white", color = "black", size = 1),
    strip.text = element_text(face = "bold", size = 12),
    
    #2.185 Center and bold the plot title
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

#2.2 Print the primary analysis results plot
print(primary_analysis_plot)

```

## Secondary Analysis: Associations Between Risk Group + Dimensional Psychopathology

**Rationale for Secondary Continuous Outcome Modeling Methods**

For the secondary analyses, continuous measures of psychopathology - specifically, DSM‑5 Oriented Depression Problems, DSM‑5 Oriented Anxiety Problems, Attention Problems, and Aggression Problems (as measured by CBCL scores)—were modeled using linear mixed‐effects regression. Each model regressed the outcome on cluster membership, the corresponding baseline CBCL score (included because baseline scores contributed to the derivation of the risk clusters), and sex. Importantly, a random intercept for **family_id** was incorporated to account for the within-family correlation - which is different than the previous

The inclusion of **family_id** as a random effect performed very well: it effectively partitioned between-family variance, especially in comparison to the logistic regression models above. This structure ensured that individual-level variability was not over-absorbed by the random effects, thereby preserving the interpretability and stability of the fixed effect estimates.

Following model fitting with `lmer()`, an ANCOVA (Type II) was conducted using `car::Anova()` to assess the omnibus effect of cluster on each outcome metric. When the omnibus effect was significant (p < 0.05), post-hoc pairwise contrasts were computed using `emmeans` with FDR adjustment.

TL;DR: Linear mixed-effects models with a random intercept for family_id were chosen to robustly account for within-family correlations in continuous CBCL outcomes. The models controlled for baseline scores and sex, and subsequent ANCOVA and post-hoc tests provided nuanced insight into how cluster membership relates to dimensional psychopathology measures.

```{r secondary analyses, echo = FALSE, warning = FALSE}

## Secondary Analyses: Continuous CBCL Scores by Risk Group Cluster ##

#1. Generate relevant summary statistics for each continuous outcome of interest by cluster membership
#1.1 Define continuous outcome variables
secondary_outcome_vars <- c("cbcl_scr_dsm5_depress_t", 
                  "cbcl_scr_dsm5_anxdisord_t", 
                  "cbcl_scr_syn_attention_t", 
                  "cbcl_scr_syn_aggressive_t")

#1.2 Create a summary table by cluster membership for the continuous outcomes
secondary_continuous_summary_table <- dimensional_psychopathology_risk_secondary_analysis_data %>%
  dplyr::select(cluster, all_of(secondary_outcome_vars)) %>%
  
  #1.21 Pivot outcomes into a long format
  pivot_longer(cols = all_of(secondary_outcome_vars), 
               names_to = "Outcome", 
               values_to = "Score") %>%
  
  #1.22 Exclude missing scores
  filter(!is.na(Score)) %>%
  
  #1.23 Group by cluster and outcome to compute summary statistics
  group_by(cluster, Outcome) %>%
  summarize(
    Mean = round(mean(Score, na.rm = TRUE), 2),
    SD = round(sd(Score, na.rm = TRUE), 2),
    Median = round(median(Score, na.rm = TRUE), 2),
    Min = round(min(Score, na.rm = TRUE), 2),
    Max = round(max(Score, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  
  #1.24 Create a range column and clean the outcome names
  mutate(
    Range = paste0(Min, " - ", Max),
    Outcome = gsub("^cbcl_scr_", "", Outcome),
    Outcome = gsub("syn", "", Outcome),
    Outcome = gsub("_", " ", Outcome)
  ) %>%
  
  #1.25 Select desired columns for the final table
  dplyr::select(Outcome, cluster, Mean, SD, Median, Range)

#1.3 Print the secondary outcomes summary table by cluster membership
secondary_continuous_summary_table %>%
  kable(caption = "Summary Statistics for Secondary CBCL Continuous Outcomes by Cluster Membership",
        align = "l", 
        col.names = c("Outcome", "Cluster", "Mean", "SD", "Median", "Range"),
        row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#2. Run the linear mixed effect regressions for each continuous outcome variable 
#2.1 Initialize a list to store results for each outcome
secondary_analysis_results <- list()

#2.2 Create a function to iterate over each secondary outcome variable
for (secondary_outcome in secondary_outcome_vars) {
  
  #2.21 Define the corresponding baseline variable
  baseline_var <- paste0("baseline_", secondary_outcome)
  
  #2.22 Fit the linear mixed-effects model
  linear_mem <- lmer(as.formula(paste(secondary_outcome, "~ cluster +", baseline_var, "+ sex + (1|family_id)")),
                na.action = na.omit,
                data = dimensional_psychopathology_risk_secondary_analysis_data)
  
  #2.23 Run ANCOVA (Type II) to assess omnibus fixed effects
  ancova_result <- car::Anova(linear_mem, type = "II", test.statistic = "F")
  
  #2.24 Extract the omnibus p-value for the cluster effect
  cluster_p <- ancova_result["cluster", "Pr(>F)"]
  
  #2.25 Initialize post hoc results
  posthoc_results <- NA
  
  #2.26 If the omnibus effect is significant (p < 0.05), run post hoc pairwise contrasts
  if (!is.na(cluster_p) && cluster_p < 0.05) {
    emmeans_obj <- emmeans(linear_mem, ~ cluster)
    posthoc_results <- contrast(emmeans_obj, method = "pairwise", adjust = "fdr")
  }
  
  #2.27 Store the modeling results
  secondary_analysis_results[[secondary_outcome]] <- list(
    model = linear_mem,
    ancova = ancova_result,
    cluster_p = cluster_p,
    posthoc = posthoc_results
  )
}

#3. Clean + display the results of the linear mixed effects regression models
#3.1 Create a function to clean the omnibus ANCOVA results
omnibus_results_list <- lapply(names(secondary_analysis_results), function(outcome_name) {
  
  #3.11 Extract the results for this outcome
  omnibus_result <- secondary_analysis_results[[outcome_name]]
  
  #3.12 Extract the row name "cluster" which corresponds to the cluster effect in the ANCOVA
  ancova_df <- omnibus_result$ancova
  omnibus_row <- ancova_df[rownames(ancova_df) == "cluster", ]
  
  data.frame(
    Outcome = outcome_name,
    Comparison = "Omnibus (cluster)",
    Statistic = round(omnibus_row["F"], 2),
    df = paste0(omnibus_row["Df"], ", ", omnibus_row["Df.res"]),
    p_value = as.character(ifelse(omnibus_row["Pr(>F)"] < 0.001, "<0.001", round(omnibus_row["Pr(>F)"], 3))),
    Test_Type = "Omnibus",
    stringsAsFactors = FALSE
  )
})

omnibus_results_table <- bind_rows(omnibus_results_list) %>%
  mutate(
    Outcome = gsub("^cbcl_scr_", "", Outcome),
    Outcome = gsub("syn", "", Outcome),
    Outcome = gsub("_", " ", Outcome)
  )

#3.2 Clean the pairwise post hoc comparison results
posthoc_results_list <- lapply(names(secondary_analysis_results), function(outcome_name) {
  res <- secondary_analysis_results[[outcome_name]]
  posthoc_result <- NULL
  
  #3.21 Check if posthoc results are available (i.e. not just an NA placeholder)
  if (!inherits(res$posthoc, "logical")) {
    posthoc_df <- as.data.frame(res$posthoc)
    if (nrow(posthoc_df) > 0) {
      posthoc_result <- posthoc_df %>%
        mutate(
          Outcome = outcome_name,
          Comparison = contrast,
          Statistic = round(estimate, 2),
          SE = round(SE, 2),
          p_value = as.character(ifelse(p.value < 0.001, "<0.001", round(p.value, 3))),
          Test_Type = "Posthoc"
        ) %>%
        dplyr::select(Outcome, Comparison, Statistic, SE, p_value, Test_Type)
    }
  }
  return(posthoc_result)
})

posthoc_results_table <- bind_rows(posthoc_results_list) %>%
  mutate(
    Outcome = gsub("^cbcl_scr_", "", Outcome),
    Outcome = gsub("syn", "", Outcome),
    Outcome = gsub("_", " ", Outcome)
  )

#3.3 Print the cleaned + reformatted secondary analysis results tables
#3.31 Omnibus ANCOVA Results Table
omnibus_results_table %>%
  kable(caption = "Omnibus ANCOVA Results for Continuous CBCL Outcomes by Cluster",
        align = "l",
        col.names = c("Outcome", "Comparison", "F Statistic", "df", "P-Value", "Test Type"),
        row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#3.32 Pairwise Post Hoc Results Table
posthoc_results_table %>%
  kable(caption = "Pairwise Post Hoc Results for Continuous CBCL Outcomes by Cluster",
        align = "l",
        col.names = c("Outcome", "Comparison", "Estimate", "SE", "FDR P-Value", "Test Type"),
        row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```

### Plotting Secondary Analysis Results

```{r secondary analysis plots, echo = FALSE, warning = FALSE, fig.width = 10, fig.height = 6}

## Plot the Results of the Secondary Linear Mixed Effects Regression Analyses ##

#1. Prepare the data for plotting
secondary_analysis_plot_data <- dimensional_psychopathology_risk_secondary_analysis_data %>%
  pivot_longer(
    cols = all_of(secondary_outcome_vars), 
    names_to = "Outcome", 
    values_to = "Score"
  ) %>%
  mutate(
    Outcome = gsub("^cbcl_scr_", "", Outcome),
    Outcome = gsub("syn", "", Outcome),
    Outcome = gsub("_", " ", Outcome),
    Cluster = paste("Cluster", cluster)
  )

#2 Generate a faceted boxplot of secondary continuous outcomes
#2.1 Create the facet boxplot
secondary_analysis_plot <- ggplot(secondary_analysis_plot_data, aes(x = Cluster, y = Score, fill = Cluster)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  facet_wrap(~ Outcome, scales = "free_y", ncol = 2) +
  scale_fill_viridis_d(option = "viridis") +
  # Remove the redundant "Cluster" prefix from x-axis labels.
  scale_x_discrete(labels = function(x) gsub("Cluster ", "", x)) +
  labs(
    x = "Cluster",
    y = "Score",
    title = "Continuous CBCL Outcomes by Cluster"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(face = "bold", size = 10),
    strip.background = element_rect(fill = "white", color = "black", size = 1),
    strip.text = element_text(face = "bold", size = 12),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "none"
  )

#2.2 Print the secondary analysis plot
print(secondary_analysis_plot)

```
