---
title: "Mood Disorder & Suicidality Outcome Variable EDA"
author: "Sam A. Sievertsen"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(warning = FALSE, message = FALSE, comment = "")

```

# Exploratory Data Analysis (EDA) of variables to be used as outcome metrics related to mood disorders and suicidality in the ABCD dataset

The most up to date list of variables to used as outcomes in the current study, including their role, domain, details of each variable as it relates to the ABCD dataset, etc., can be [found here](https://ohsuitg-my.sharepoint.com/:x:/r/personal/sievertsen_ohsu_edu/Documents/Huber_Lab/FYP/Project_Planning/ABCD_Mood_Disorder_Risk_Model_Variables_010624_SS.xlsx?d=w286583f8914b4ab5a402e36ce8c36291&csf=1&web=1&e=ieCAKt&nav=MTVfezIxMDEwOEMxLTFFMjktNERDMi1CQTE4LTM0QTE2NzA0MEQ2RX0). 

The following report contains information regarding the: 

  1. Integrity of the data
  
  2. Assessment of outliers in the data
  
  3. Skewness + appropriateness of the mean as the measure of central tendency for continuous variables
  
  4. Shape of each variables' distribution

```{r environment, echo = FALSE, include = FALSE, warning = FALSE}

# Load necessary packages + environmental variables
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(moments)
options(scipen = 999, digits = 8)

# Read in raw data generated in DEAP
raw_data <- read.csv("../../data/data_raw/dataset.csv")

```

## Data Wrangling 

This section prepares the data for EDA of outcomes 

```{r data wrangling, echo = FALSE, include = FALSE, warning = FALSE}

## Data Wrangling ##

#1. Clean the outcome data 
#1.1 Retain only columns of interest
raw_data_trimmed <- raw_data %>% 
  dplyr::select(c(participant_id, session_id, mh_p_cbcl__dsm__anx_tscore, mh_p_cbcl__dsm__dep_tscore, mh_p_cbcl__synd__aggr_tscore, mh_p_cbcl__synd__attn_tscore, mh_p_ksads__suic__pass__past_dx, mh_p_ksads__suic__actv__past_dx, mh_p_ksads__suic__actv__mthd__past_dx, mh_p_ksads__suic__actv__int__past_dx, mh_p_ksads__suic__actv__plan__past_dx, mh_p_ksads__suic__pass__pres_dx, mh_p_ksads__suic__actv__pres_dx, mh_p_ksads__suic__actv__mthd__pres_dx, mh_p_ksads__suic__actv__int__pres_dx, mh_p_ksads__suic__actv__plan__pres_dx, mh_y_ksads__suic__pass__past_dx, mh_y_ksads__suic__actv__past_dx, mh_y_ksads__suic__actv__mthd__past_dx, mh_y_ksads__suic__actv__int__past_dx, mh_y_ksads__suic__actv__plan__past_dx, mh_y_ksads__suic__pass__pres_dx, mh_y_ksads__suic__actv__pres_dx, mh_y_ksads__suic__actv__mthd__pres_dx, mh_y_ksads__suic__actv__int__pres_dx, mh_y_ksads__suic__actv__plan__pres_dx, mh_p_ksads__suic__atmpt__intr__past_dx, mh_p_ksads__suic__atmpt__abrt__past_dx, mh_p_ksads__suic__atmpt__past_dx, mh_p_ksads__suic__atmpt__intr__pres_dx, mh_p_ksads__suic__atmpt__abrt__pres_dx, mh_p_ksads__suic__atmpt__pres_dx, mh_y_ksads__suic__atmpt__intr__past_dx, mh_y_ksads__suic__atmpt__abrt__past_dx, mh_y_ksads__suic__atmpt__past_dx, mh_y_ksads__suic__atmpt__intr__pres_dx, mh_y_ksads__suic__atmpt__abrt__pres_dx, mh_y_ksads__suic__atmpt__pres_dx, mh_p_ksads__suic__slfinj__nosi__pres_dx, mh_p_ksads__suic__slfinj__nosi__past_dx, mh_y_ksads__suic__slfinj__nosi__pres_dx, mh_y_ksads__suic__slfinj__nosi__past_dx, mh_p_ksads__bpd__bpd1__curdep__partrem_dx, mh_p_ksads__bpd__bpd1__rcnt__manicmix__pres_dx, mh_p_ksads__bpd__bpd1__curmanic__pres_dx, mh_p_ksads__bpd__bpd1__curdep__pres_dx, mh_y_ksads__bpd__bpd1__curmanic__pres_dx, mh_y_ksads__bpd__bpd1__curdep__pres_dx, mh_y_ksads__bpd__bpd1__rcnt__manic__pres_dx, mh_p_ksads__bpd__bpd1__curhypo__pres_dx, mh_p_ksads__bpd__bpd1__rcnt__dep__pres_dx, mh_p_ksads__bpd__bpd1__rcnt__manic__pres_dx, mh_y_ksads__bpd__bpd1__rcnt__dep__pres_dx, mh_y_ksads__bpd__bpd1__curhypo__pres_dx, mh_y_ksads__bpd__bpd1__rcnt__manicmix__pres_dx, mh_y_ksads__bpd__bpd1__curdep__partrem_dx, mh_p_ksads__bpd__bpd2__curhypo__pres_dx, mh_p_ksads__bpd__bpd2__rcnt__hypo__pres_dx, mh_p_ksads__bpd__bpd2__curdep__pres_dx, mh_p_ksads__bpd__bpd2__curdep__partrem_dx, mh_y_ksads__bpd__bpd2__curdep__pres_dx, mh_y_ksads__bpd__bpd2__curdep__partrem_dx, mh_y_ksads__bpd__bpd2__rcnt__hypo__pres_dx, mh_y_ksads__bpd__bpd2__curhypo__pres_dx, mh_p_ksads__bpd__bpd2__rcnt__hypomix__pres_dx, mh_y_ksads__bpd__bpd2__rcnt__hypomix__pres_dx, mh_p_ksads__bpd__bpd2__oth__mindur_dx, mh_p_ksads__bpd__oth__mindur_dx, mh_y_ksads__bpd__bpd2__oth__mindur_dx, mh_y_ksads__bpd__oth__mindur_dx, mh_p_ksads__bpd__unspec__pres_dx, mh_p_ksads__bpd__unspec__past_dx, mh_y_ksads__bpd__unspec__pres_dx, mh_y_ksads__bpd__unspec__past_dx))

#1.2 Turn all instances of empty, missing, 555, and 999 to NA; and 888 to 0 in KSADS columns
raw_data_trimmed <- raw_data_trimmed %>%
  mutate(across(starts_with("mh_"), ~ na_if(replace(., . %in% c(555, 999, ""), NA), 888) %>% replace_na(0)))

#2. Preliminarily create relevant outcome variables of interest
#2.1 Disorder Variables 
#2.11 Lifetime Bipolar I Disorder
# Create the bipolar_I_dx variable
outcome_data_merged_trimmed <- outcome_data_merged_trimmed %>%
  mutate(
    bipolar_I_dx = case_when(
      rowSums(select(., c(
        "ksads2_2_937_p", "ksads2_2_800_p", "ksads2_2_798_p", 
        "ksads2_2_799_p", "ksads2_2_798_t", "ksads2_2_799_t", 
        "ksads2_2_800_t", "ksads2_2_937_t"
      )) == 1, na.rm = TRUE) > 0 ~ 1, # Any variable == 1
      rowSums(!is.na(select(., c(
        "ksads2_2_937_p", "ksads2_2_800_p", "ksads2_2_798_p", 
        "ksads2_2_799_p", "ksads2_2_798_t", "ksads2_2_799_t", 
        "ksads2_2_800_t", "ksads2_2_937_t"
      )))) == 0 ~ NA_real_, # All variables are NA
      TRUE ~ 0               # All variables == 0
    )
  )

#2.12 Lifetime Bipolar II Disorder
outcome_data_merged_trimmed <- outcome_data_merged_trimmed %>%
  mutate(
    bipolar_II_dx = case_when(
      rowSums(select(., c(
        "ksads2_2_801_p", "ksads2_2_802_p", "ksads2_2_931_p",
        "ksads2_2_936_p", "ksads2_2_931_t", "ksads2_2_936_t",
        "ksads2_2_802_t", "ksads2_2_801_t"
      )) == 1, na.rm = TRUE) > 0 ~ 1, # Any variable == 1
      rowSums(!is.na(select(., c(
        "ksads2_2_801_p", "ksads2_2_802_p", "ksads2_2_931_p",
        "ksads2_2_936_p", "ksads2_2_931_t", "ksads2_2_936_t",
        "ksads2_2_802_t", "ksads2_2_801_t"
      )))) == 0 ~ NA_real_, # All variables are NA
      TRUE ~ 0               # All variables == 0
    )
  )

#2.13 Lifetime Any Bipolar Disorder
outcome_data_merged_trimmed <- outcome_data_merged_trimmed %>%
  mutate(
    any_bipolar_dx = case_when(
      rowSums(select(., c(
        "bipolar_I_dx", "bipolar_II_dx", "ksads2_2_803_p",
        "ksads2_2_933_p", "ksads2_2_803_t", "ksads2_2_933_t"
      )) == 1, na.rm = TRUE) > 0 ~ 1, # Any variable == 1
      rowSums(!is.na(select(., c(
        "bipolar_I_dx", "bipolar_II_dx", "ksads2_2_803_p",
        "ksads2_2_933_p", "ksads2_2_803_t", "ksads2_2_933_t"
      )))) == 0 ~ NA_real_, # All variables are NA
      TRUE ~ 0               # All variables == 0
    )
  )

#2.2 Suicide & self-harm related variables
#2.21 Lifetime Suicidal Thinking
outcome_data_merged_trimmed <- outcome_data_merged_trimmed %>%
  mutate(
    suicidal_ideation = case_when(
      rowSums(select(., c(
        "ksads2_23_917_p", "ksads2_23_918_p", "ksads2_23_919_p", 
        "ksads2_23_920_p", "ksads2_23_921_p", "ksads2_23_906_p", 
        "ksads2_23_907_p", "ksads2_23_908_p", "ksads2_23_909_p", 
        "ksads2_23_910_p", "ksads2_23_917_t", "ksads2_23_918_t", 
        "ksads2_23_919_t", "ksads2_23_920_t", "ksads2_23_921_t", 
        "ksads2_23_906_t", "ksads2_23_907_t", "ksads2_23_908_t", 
        "ksads2_23_909_t", "ksads2_23_910_t"
      )) == 1, na.rm = TRUE) > 0 ~ 1, # Any variable == 1
      rowSums(!is.na(select(., c(
        "ksads2_23_917_p", "ksads2_23_918_p", "ksads2_23_919_p", 
        "ksads2_23_920_p", "ksads2_23_921_p", "ksads2_23_906_p", 
        "ksads2_23_907_p", "ksads2_23_908_p", "ksads2_23_909_p", 
        "ksads2_23_910_p", "ksads2_23_917_t", "ksads2_23_918_t", 
        "ksads2_23_919_t", "ksads2_23_920_t", "ksads2_23_921_t", 
        "ksads2_23_906_t", "ksads2_23_907_t", "ksads2_23_908_t", 
        "ksads2_23_909_t", "ksads2_23_910_t"
      )))) == 0 ~ NA_real_, # All variables are NA
      TRUE ~ 0               # All variables == 0
    )
  )

#2.22 Lifetime Suicide Attempt (i.e., Suicidal Behavior)
outcome_data_merged_trimmed <- outcome_data_merged_trimmed %>%
  mutate(
    suicide_attempt = case_when(
      rowSums(select(., c(
        "ksads2_23_923_p", "ksads2_23_924_p", "ksads2_23_925_p", 
        "ksads2_23_912_p", "ksads2_23_913_p", "ksads2_23_914_p", 
        "ksads2_23_923_t", "ksads2_23_924_t", "ksads2_23_925_t", 
        "ksads2_23_912_t", "ksads2_23_913_t", "ksads2_23_914_t"
      )) == 1, na.rm = TRUE) > 0 ~ 1, # Any variable == 1
      rowSums(!is.na(select(., c(
        "ksads2_23_923_p", "ksads2_23_924_p", "ksads2_23_925_p", 
        "ksads2_23_912_p", "ksads2_23_913_p", "ksads2_23_914_p", 
        "ksads2_23_923_t", "ksads2_23_924_t", "ksads2_23_925_t", 
        "ksads2_23_912_t", "ksads2_23_913_t", "ksads2_23_914_t"
      )))) == 0 ~ NA_real_, # All variables are NA
      TRUE ~ 0               # All variables == 0
    )
  )

#2.23 Lifetime Non-Suicidal Self Injury
outcome_data_merged_trimmed <- outcome_data_merged_trimmed %>%
  mutate(
    nssi = case_when(
      rowSums(select(., c(
        "ksads2_23_905_p", "ksads2_23_916_p", "ksads2_23_905_t",
        "ksads2_23_916_t"
      )) == 1, na.rm = TRUE) > 0 ~ 1, # Any variable == 1
      rowSums(!is.na(select(., c(
        "ksads2_23_905_p", "ksads2_23_916_p", "ksads2_23_905_t",
        "ksads2_23_916_t"
      )))) == 0 ~ NA_real_, # All variables are NA
      TRUE ~ 0               # All variables == 0
    )
  )

```
