---
title: "Mood Disorder & Suicidality Outcome Variable EDA"
author: "Sam A. Sievertsen"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(warning = FALSE, message = FALSE, comment = "")

```

# Exploratory Data Analysis (EDA) of variables to be used as outcome metrics related to mood disorders and suicidality in the ABCD dataset

The most up to date list of variables to used as outcomes in the current study, including their role, domain, details of each variable as it relates to the ABCD dataset, etc., can be [found here](https://ohsuitg-my.sharepoint.com/:x:/r/personal/sievertsen_ohsu_edu/Documents/Huber_Lab/FYP/Project_Planning/ABCD_Mood_Disorder_Risk_Model_Variables_010624_SS.xlsx?d=w286583f8914b4ab5a402e36ce8c36291&csf=1&web=1&e=ieCAKt&nav=MTVfezIxMDEwOEMxLTFFMjktNERDMi1CQTE4LTM0QTE2NzA0MEQ2RX0). 

The following report contains information regarding the: 

  1. Integrity of the data
  
  2. Assessment of outliers in the data
  
  3. Skewness + appropriateness of the mean as the measure of central tendency for continuous variables
  
  4. Shape of each variables' distribution

```{r environment, echo = FALSE, include = FALSE, warning = FALSE}

# Load necessary packages + environmental variables
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(moments)
options(scipen = 999, digits = 8)

# Read in parent report ksads-comp data
parent_ksads_data <- read.csv("../../data/data_raw/mh_p_ksads_ss.csv")

# Read in youth report ksads-comp data
youth_ksads_data <- read.csv("../../data/data_raw/mh_y_ksads_ss.csv")

# Read in cbcl data
cbcl_data <- read.csv("../../data/data_raw/mh_p_cbcl.csv")

```

```{r data wrangling, echo = FALSE, include = FALSE, warning = FALSE}

## Data Wrangling ##

#1. Merge the data
#1.1 Collapse the data frames containing outcome variables of interest
outcome_data_merged <- Reduce(
  function(x, y)
    merge(x, y,
      by = c("src_subject_id", "eventname"),
      all = TRUE),
  list(parent_ksads_data, youth_ksads_data, cbcl_data))

#1.2 Verify the merge did not duplicate any columns and resulted in an expected subject N
#1.21 Check for duplicate columns
duplicate_columns <- grep("\\.x$|\\.y$", colnames(outcome_data_merged), value = TRUE)
if (length(duplicate_columns) > 0) {
  message("Duplicate columns found: ", paste(duplicate_columns, collapse = ", "))
} else {
  message("No duplicate columns ending found.")
}

#1.22 Check the number of unique subjects to ensure no data was lost/erroneously created during merging
unique_count <- length(unique(outcome_data_merged$src_subject_id))
if (unique_count >= 11800 && unique_count <= 12000) {
  message("The number of unique src_subject_id (", unique_count, ") is within the acceptable range.")
} else {
  message("The number of unique src_subject_id (", unique_count, ") is outside the acceptable range!")
}

#2. Clean the merged data 
#2.1 Retain only columns of interest
outcome_data_merged_trimmed <- outcome_data_merged %>% 
  dplyr::select(c(src_subject_id, eventname, cbcl_scr_dsm5_depress_t, cbcl_scr_dsm5_anxdisord_t, cbcl_scr_syn_attention_t, cbcl_scr_syn_aggressive_t, cbcl_scr_dsm5_depress_nm, cbcl_scr_dsm5_anxdisord_nm, cbcl_scr_syn_attention_nm, cbcl_scr_syn_aggressive_nm, ksads2_2_937_p, ksads2_2_800_p, ksads2_2_798_p, ksads2_2_799_p, ksads2_2_798_t, ksads2_2_799_t, ksads2_2_800_t, ksads2_2_937_t, ksads2_2_801_p, ksads2_2_802_p, ksads2_2_931_p,ksads2_2_936_p, ksads2_2_931_t, ksads2_2_936_t,ksads2_2_802_t, ksads2_2_801_t, ksads2_2_803_p, ksads2_2_933_p, ksads2_2_803_t, ksads2_2_933_t, ksads2_23_917_p, ksads2_23_918_p, ksads2_23_919_p, ksads2_23_920_p, ksads2_23_921_p, ksads2_23_906_p, ksads2_23_907_p, ksads2_23_908_p, ksads2_23_909_p, ksads2_23_910_p, ksads2_23_917_t, ksads2_23_918_t, ksads2_23_919_t, ksads2_23_920_t, ksads2_23_921_t, ksads2_23_906_t, ksads2_23_907_t, ksads2_23_908_t, ksads2_23_909_t, ksads2_23_910_t, ksads2_23_923_p, ksads2_23_924_p, ksads2_23_925_p, ksads2_23_912_p, ksads2_23_913_p, ksads2_23_914_p, ksads2_23_923_t, ksads2_23_924_t, ksads2_23_925_t, ksads2_23_912_t, ksads2_23_913_t, ksads2_23_914_t, ksads2_23_905_p, ksads2_23_916_p, ksads2_23_905_t, ksads2_23_916_t))

#2.2 Turn all instances of empty, missing, 555, and 999 to NA; and 888 to 0 in KSADS columns
outcome_data_merged_trimmed <- outcome_data_merged_trimmed %>%
  mutate(across(starts_with("ksads"), ~ na_if(replace(., . %in% c(555, 999, ""), NA), 888) %>% replace_na(0)))

#2.3 Filter the trimmed data to only contain rows at the baseline assessment timepoint
outcome_data_merged_trimmed <- outcome_data_merged_trimmed %>% 
  filter(eventname == "4_year_follow_up_y_arm_1")

#3. Preliminarily create relevant outcome variables of interest
#3.1 Disorder Variables 
#3.11 Bipolar I Disorder
# Create the bipolar_I_dx variable
outcome_data_merged_trimmed <- outcome_data_merged_trimmed %>%
  mutate(
    bipolar_I_dx = case_when(
      rowSums(select(., c(
        "ksads2_2_937_p", "ksads2_2_800_p", "ksads2_2_798_p", 
        "ksads2_2_799_p", "ksads2_2_798_t", "ksads2_2_799_t", 
        "ksads2_2_800_t", "ksads2_2_937_t"
      )) == 1, na.rm = TRUE) > 0 ~ 1, # Any variable == 1
      rowSums(!is.na(select(., c(
        "ksads2_2_937_p", "ksads2_2_800_p", "ksads2_2_798_p", 
        "ksads2_2_799_p", "ksads2_2_798_t", "ksads2_2_799_t", 
        "ksads2_2_800_t", "ksads2_2_937_t"
      )))) == 0 ~ NA_real_, # All variables are NA
      TRUE ~ 0               # All variables == 0
    )
  )

#3.12 Bipolar II Disorder
outcome_data_merged_trimmed <- outcome_data_merged_trimmed %>%
  mutate(
    bipolar_II_dx = case_when(
      rowSums(select(., c(
        "ksads2_2_801_p", "ksads2_2_802_p", "ksads2_2_931_p",
        "ksads2_2_936_p", "ksads2_2_931_t", "ksads2_2_936_t",
        "ksads2_2_802_t", "ksads2_2_801_t"
      )) == 1, na.rm = TRUE) > 0 ~ 1, # Any variable == 1
      rowSums(!is.na(select(., c(
        "ksads2_2_801_p", "ksads2_2_802_p", "ksads2_2_931_p",
        "ksads2_2_936_p", "ksads2_2_931_t", "ksads2_2_936_t",
        "ksads2_2_802_t", "ksads2_2_801_t"
      )))) == 0 ~ NA_real_, # All variables are NA
      TRUE ~ 0               # All variables == 0
    )
  )

#3.13 Any Bipolar Disorder
outcome_data_merged_trimmed <- outcome_data_merged_trimmed %>%
  mutate(
    any_bipolar_dx = case_when(
      rowSums(select(., c(
        "bipolar_I_dx", "bipolar_II_dx", "ksads2_2_803_p",
        "ksads2_2_933_p", "ksads2_2_803_t", "ksads2_2_933_t"
      )) == 1, na.rm = TRUE) > 0 ~ 1, # Any variable == 1
      rowSums(!is.na(select(., c(
        "bipolar_I_dx", "bipolar_II_dx", "ksads2_2_803_p",
        "ksads2_2_933_p", "ksads2_2_803_t", "ksads2_2_933_t"
      )))) == 0 ~ NA_real_, # All variables are NA
      TRUE ~ 0               # All variables == 0
    )
  )

#3.2 Suicide & self-harm related variables
#3.21 Suicidal Thinking
outcome_data_merged_trimmed <- outcome_data_merged_trimmed %>%
  mutate(
    suicidal_thinking = case_when(
      rowSums(select(., c(
        "ksads2_23_917_p", "ksads2_23_918_p", "ksads2_23_919_p", 
        "ksads2_23_920_p", "ksads2_23_921_p", "ksads2_23_906_p", 
        "ksads2_23_907_p", "ksads2_23_908_p", "ksads2_23_909_p", 
        "ksads2_23_910_p", "ksads2_23_917_t", "ksads2_23_918_t", 
        "ksads2_23_919_t", "ksads2_23_920_t", "ksads2_23_921_t", 
        "ksads2_23_906_t", "ksads2_23_907_t", "ksads2_23_908_t", 
        "ksads2_23_909_t", "ksads2_23_910_t"
      )) == 1, na.rm = TRUE) > 0 ~ 1, # Any variable == 1
      rowSums(!is.na(select(., c(
        "ksads2_23_917_p", "ksads2_23_918_p", "ksads2_23_919_p", 
        "ksads2_23_920_p", "ksads2_23_921_p", "ksads2_23_906_p", 
        "ksads2_23_907_p", "ksads2_23_908_p", "ksads2_23_909_p", 
        "ksads2_23_910_p", "ksads2_23_917_t", "ksads2_23_918_t", 
        "ksads2_23_919_t", "ksads2_23_920_t", "ksads2_23_921_t", 
        "ksads2_23_906_t", "ksads2_23_907_t", "ksads2_23_908_t", 
        "ksads2_23_909_t", "ksads2_23_910_t"
      )))) == 0 ~ NA_real_, # All variables are NA
      TRUE ~ 0               # All variables == 0
    )
  )

#3.22 Suicide Attempt (i.e., Suicidal Behavior)
# Create the suicide_attempt variable
outcome_data_merged_trimmed <- outcome_data_merged_trimmed %>%
  mutate(
    suicide_attempt = case_when(
      rowSums(select(., c(
        "ksads2_23_923_p", "ksads2_23_924_p", "ksads2_23_925_p", 
        "ksads2_23_912_p", "ksads2_23_913_p", "ksads2_23_914_p", 
        "ksads2_23_923_t", "ksads2_23_924_t", "ksads2_23_925_t", 
        "ksads2_23_912_t", "ksads2_23_913_t", "ksads2_23_914_t"
      )) == 1, na.rm = TRUE) > 0 ~ 1, # Any variable == 1
      rowSums(!is.na(select(., c(
        "ksads2_23_923_p", "ksads2_23_924_p", "ksads2_23_925_p", 
        "ksads2_23_912_p", "ksads2_23_913_p", "ksads2_23_914_p", 
        "ksads2_23_923_t", "ksads2_23_924_t", "ksads2_23_925_t", 
        "ksads2_23_912_t", "ksads2_23_913_t", "ksads2_23_914_t"
      )))) == 0 ~ NA_real_, # All variables are NA
      TRUE ~ 0               # All variables == 0
    )
  )

#3.23 Non-Suicidal Self Injury
outcome_data_merged_trimmed <- outcome_data_merged_trimmed %>%
  mutate(
    nssi = case_when(
      rowSums(select(., c(
        "ksads2_23_905_p", "ksads2_23_916_p", "ksads2_23_905_t",
        "ksads2_23_916_t"
      )) == 1, na.rm = TRUE) > 0 ~ 1, # Any variable == 1
      rowSums(!is.na(select(., c(
        "ksads2_23_905_p", "ksads2_23_916_p", "ksads2_23_905_t",
        "ksads2_23_916_t"
      )))) == 0 ~ NA_real_, # All variables are NA
      TRUE ~ 0               # All variables == 0
    )
  )

```

## 1. Integrity of the Data

## 1. Define the N, percent, and range of outcome variables of interest at the 4Y assessment timepoint

```{r EDA 1, echo = FALSE, warning = FALSE}

## Exploratory Data Analysis ##

#1. Define the N, percent, and range of outcome variables of interest at the 4Y assessment timepoint
#1.1 Define the columns to analyze
#1.11 Categorical columns to analyze
categorical_columns <- c("bipolar_I_dx", "bipolar_II_dx", "any_bipolar_dx", "suicidal_thinking", "suicide_attempt", "nssi")

#1.12 Continuous columns to analyze
continuous_columns <- c("cbcl_scr_dsm5_depress_t", "cbcl_scr_dsm5_anxdisord_t", "cbcl_scr_syn_attention_t", "cbcl_scr_syn_aggressive_t")

#1.13 Continuous missing variable columns to analyze
missing_columns <- c("cbcl_scr_dsm5_depress_nm", "cbcl_scr_dsm5_anxdisord_nm", "cbcl_scr_syn_attention_nm", "cbcl_scr_syn_aggressive_nm")

#1.2 Create a function to compute EDA metrics for both categorical and continuous columns
get_missingness_metrics <- function(data, categorical_columns, continuous_columns, missing_columns) {
  
  #1.21 Helper function for categorical columns
  get_categorical_metrics <- function(column_name) {
    column_data <- data[[column_name]]
    n_non_na <- sum(!is.na(column_data) & column_data != "")
    percent_non_na <- (n_non_na / nrow(data)) * 100
    unique_values <- unique(column_data[!is.na(column_data) & column_data != ""])
    list(
      column = column_name,
      n_non_na = n_non_na,
      percent_non_na = percent_non_na,
      range_or_unique = unique_values
    )
  }
  
  #1.22 Helper function for continuous columns
  get_continuous_metrics <- function(column_name, missing_column_name) {
    column_data <- data[[column_name]]
    missing_data <- data[[missing_column_name]]
    n_non_na <- sum(!is.na(column_data) & column_data != "")
    percent_non_na <- (n_non_na / nrow(data)) * 100
    value_range <- range(column_data, na.rm = TRUE)
    n_missing_scores <- sum(missing_data > 1, na.rm = TRUE)
    list(
      column = column_name,
      n_non_na = n_non_na,
      percent_non_na = percent_non_na,
      range_or_unique = value_range,
      n_missing_scores = n_missing_scores
    )
  }
  
  #1.23 Process categorical columns
  categorical_metrics <- lapply(categorical_columns, get_categorical_metrics)
  
  #1.24 Process continuous columns with missing value checks
  continuous_metrics <- mapply(
    get_continuous_metrics,
    continuous_columns,
    missing_columns,
    SIMPLIFY = FALSE
  )
  
  #1.25 Combine results into a single list
  metrics <- c(categorical_metrics, continuous_metrics)
  
  #1.26 Convert to a dataframe for printing in output
  outcome_integrity_metrics <- do.call(rbind, lapply(metrics, function(x) {
    data.frame(
      column = x$column,
      n_non_na = x$n_non_na,
      percent_non_na = x$percent_non_na,
      range_or_unique = I(list(x$range_or_unique)),
      n_missing_scores = ifelse(is.null(x$n_missing_scores), NA, x$n_missing_scores)
    )
  }))
  
  return(outcome_integrity_metrics)
}

#1.3 Apply the function to compute EDA metrics for both categorical and continuous columns to the outcome variable data
outcome_variable_integrity <- get_missingness_metrics(outcome_data_merged_trimmed, categorical_columns, continuous_columns, missing_columns)

# Print the results
kable(outcome_variable_integrity, col.names = c("Variable", "N Non-Missing", "Percent Non-Missing", "Range/Unique Values", "Missing Scores > 1"), digits = 2, row.names = FALSE)

```

## 2. Visually and Quantitatively Assess for any Outliers in the Data

```{r EDA 2, echo = FALSE, warning = FALSE}

## Exploratory Data Analysis ##

#2. Visually and Quantitatively Assess for any Outliers in the Data
#2.1 Create boxplots with outlier indicators for continuous variables in the dataset
#2.11 Filter only continuous variables
continuous_outcome_variables <- outcome_data_merged_trimmed %>%
  dplyr::select(cbcl_scr_dsm5_depress_t, cbcl_scr_dsm5_anxdisord_t, cbcl_scr_syn_attention_t, cbcl_scr_syn_aggressive_t) %>%
  gather(key = "Variable", value = "Value")

#2.12 Create boxplot with facets for continuous variables
continuous_outcome_variables_boxplot <- ggplot(continuous_outcome_variables, aes(x = Variable, y = Value)) + 
  geom_boxplot(fill = "lightblue", color = "darkblue") + 
  facet_wrap(~ Variable, scales = "free") + 
  theme_minimal() + 
  theme(axis.text.x = element_blank(),
        strip.text = element_text(size = 6, face = "bold")) +
  labs(title = "Boxplots of Continuous Outcome Variables",
       y = "Values",
       x = "")

#2.13 Print the plot
print(continuous_outcome_variables_boxplot)

```

Based on the boxplots of the continuous (secondary) outcome variables, it does not appear as though any values are out of range or otherwise unexpected/impactful outliers.

## 3. Appropriateness of the Mean as a Measure of Central Tendency

While clustering methods used in this study do not require parametric assumptions, the secondary continuous outcome measures - `cbcl_scr_dsm5_depress_t`, `cbcl_scr_dsm5_anxdisord_t`, `cbcl_scr_syn_attention_t`, and `cbcl_scr_syn_aggressive_t` - will serve as dependent variables in statistical models that do. Therefore, examining their skewness is essential to assess their symmetry and appropriateness for said analyses.

Below, skewness values for the four secondary outcome measures are presented.

```{r EDA 3, echo = FALSE, warning = FALSE}

## Exploratory Data Analysis ##

#3.1 Calculate skewness for secondary continuous outcome measures
outcome_measures_skewness_table <- outcome_data_merged_trimmed %>%
  dplyr::select(
    cbcl_scr_dsm5_depress_t,
    cbcl_scr_dsm5_anxdisord_t,
    cbcl_scr_syn_attention_t,
    cbcl_scr_syn_aggressive_t
  ) %>%
  summarise(across(everything(), ~ moments::skewness(.x, na.rm = TRUE))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Skewness")

#3.2 Display the skewness table
kable(outcome_measures_skewness_table, caption = "Skewness Values for Secondary Continuous Outcome Measures", digits = 2)

```

Evidenced by the table above, the CBCL aggression problems scale score (`cbcl_scr_syn_aggressive_t`) may exhibit problematic skew (i.e., > 3). When assessing results of the linear mixed effects models, this will need to be taken into account, and applicability of transformation of variable (e.g., through box-cox or log transform) should be evaluated.

## 4. Shape of Each Variables' Distribution

The shape of the distribution for each continuous variable and frequency distribution of each categorical variable is outlined below, both in terms of summary statistics derived from and a graphical histogram with density overlay of relevant data:

```{r EDA 4, echo = FALSE, warning = FALSE, fig.height=10, fig.width=10}

## Exploratory Data Analysis ##

#4. Create summary tables and visualizations for the distributions of categorical and continuous variables included as outcomes
#4.1 Create a summary table for continuous and categorical variables
summarize_data <- function(data, continuous_cols, categorical_cols) {
  summary_list <- list()
  
  #4.11 Continuous variables summary
  for (col in continuous_cols) {
    column_data <- data[[col]]
    valid_values <- column_data[!is.na(column_data)]
    summary_list[[col]] <- data.frame(
      Variable = col,
      Type = "Continuous",
      Mean = mean(valid_values, na.rm = TRUE),
      Median = median(valid_values, na.rm = TRUE),
      SD = sd(valid_values, na.rm = TRUE),
      Min = min(valid_values, na.rm = TRUE),
      Max = max(valid_values, na.rm = TRUE),
      Levels = NA,
      Frequencies = NA,
      stringsAsFactors = FALSE
    )
  }
  
  #4.12 Categorical variables summary
  for (col in categorical_cols) {
    column_data <- data[[col]]
    freq_table <- table(column_data, useNA = "no")
    summary_list[[col]] <- data.frame(
      Variable = col,
      Type = "Categorical",
      Mean = NA,
      Median = NA,
      SD = NA,
      Min = NA,
      Max = NA,
      Levels = paste(names(freq_table), collapse = ", "),
      Frequencies = paste(freq_table, collapse = ", "),
      stringsAsFactors = FALSE
    )
  }
  
  do.call(rbind, summary_list)
}

#4.2 Generate the summary table
summary_table <- summarize_data(outcome_data_merged_trimmed, continuous_columns, categorical_columns)

#4.3 Print the summary table in the RMD
kable(summary_table, caption = "Summary Statistics for Risk Model Variables", row.names = FALSE)

#4.4 Create plots visualizing distributions of risk variables
#4.41 Generate a combined dataset with an indicator for variable type
plot_data <- list()

#4.42 Add continuous variables
for (col in continuous_columns) {
  plot_data[[col]] <- outcome_data_merged_trimmed %>%
    select(Value = !!sym(col)) %>%
    mutate(Type = "Continuous", Variable = col)
}

#4.43 Add categorical variables
for (col in categorical_columns) {
  plot_data[[col]] <- outcome_data_merged_trimmed %>%
    select(Value = !!sym(col)) %>%
    mutate(Type = "Categorical", Variable = col)
}

#4.44 Combine all into a single data frame
plot_data_combined <- bind_rows(plot_data)

#4.441 Generate facet plot for continuous variables
continuous_plot <- ggplot(
  plot_data_combined %>% filter(Type == "Continuous"),
  aes(x = Value)) +
  geom_histogram(aes(y = ..density..), binwidth = 1, fill = "blue", alpha = 0.6, color = "black") +
  geom_density(color = "red", size = 1, adjust = 1.5) +
  facet_wrap(~Variable, ncol = 3, scales = "free") +
  theme_minimal() +
  labs(title = "Distributions of Continuous Variables",
       x = "Value",
       y = "Density") +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(size = 8),
    panel.spacing = unit(1, "lines"))

#4.442 Print the continuous variable plot
print(continuous_plot)

#4.451 Generate facet plot for categorical variables
categorical_plot <- ggplot(
  plot_data_combined %>% filter(Type == "Categorical"),
  aes(x = as.factor(Value))) +  
  geom_bar(fill = "skyblue", color = "black", alpha = 0.8) +
  geom_text(
    stat = "count", 
    aes(label = ..count..),  
    vjust = -0.5,            
    size = 3                
  ) +
  facet_wrap(~Variable, ncol = 3, scales = "free") +
  theme_minimal() +
  labs(title = "Distributions of Categorical Variables",
       x = "Categories",
       y = "Count") +
  scale_x_discrete(limits = c("0", "1"), labels = c("No", "Yes")) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(size = 8),
    panel.spacing = unit(1, "lines"))

#4.452 Print the categorical variable plot
print(categorical_plot)

```

Distributions of continuous variables and rates of categorical variables are shown above. While continuous variables exhibit noticeable skew, and the CBCL aggression problem scale results will need to be interpreted with caution/re-evaluated during modeling, no other problematic distributions are not noted herein. 
