---
title: "Bipolar Disorder & Suicidality Risk Variable EDA"
author: "Sam A. Sievertsen"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(warning = FALSE, message = FALSE, comment = "")

```

# Exploratory Data Analysis (EDA) of variables to be included in the clustering of groups at potential risk for the development of mood disorders and suicidality in the ABCD dataset

The most up to date list of variables to be incorporated in the risk clustering, including their role, domain, details of each variable as it relates to the ABCD dataset, empirical justification for inclusion in the clustering, etc., can be [found here](https://ohsuitg-my.sharepoint.com/:x:/r/personal/sievertsen_ohsu_edu/Documents/Huber_Lab/FYP/Project_Planning/ABCD_Mood_Disorder_Risk_Model_Variables_010624_SS.xlsx?d=w286583f8914b4ab5a402e36ce8c36291&csf=1&web=1&e=wmfy1l&nav=MTVfezk2MDA5QjlDLUQyNkQtNDFBRS04NTA5LUFDQUQwOEZGMUJFN30). 

Additionally, details regarding the incorporation of the sum adverse childhood experience (ACE) index and specific variables within it, as first recommended by [Hoffman et al., (2019)](https://www.sciencedirect.com/science/article/pii/S2352289518300821?via%3Dihub) and iterated upon by [Raney et al., (2022)](https://pmc.ncbi.nlm.nih.gov/articles/PMC9288265/) can be [located here](https://ohsuitg-my.sharepoint.com/:x:/r/personal/sievertsen_ohsu_edu/Documents/Huber_Lab/FYP/Project_Planning/ABCD_Mood_Disorder_Risk_Model_Variables_010624_SS.xlsx?d=w286583f8914b4ab5a402e36ce8c36291&csf=1&web=1&e=mr4ikT&nav=MTVfe0Y1OUEwNTY2LTY2MTUtNDNCOC1BQUQ4LTgwNkI0QkVDODZEQn0). 

The following report contains information regarding the: 

  1. Integrity of the baseline assessment timepoint data
  
  2. Assessment of outliers in the data + thoughts on how to address them if relevant
  
  3. Skewness + kurtosis + mean as the measure of central tendency for continuous variables
  
  4. Shape of each variables' distribution
  
  5. A summary of the overall availability of variables of interest to the current study
  
```{r environment, echo = FALSE, include = FALSE, warning = FALSE}

# Load necessary packages + environmental variables
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(ggplot2)
library(psych)
library(moments)
library(skimr)
options(scipen = 999, digits = 8)

# Read in raw data generated in DEAP
raw_data <- read.csv("../../data/data_raw/dataset.csv")

```

## Data Wrangling

Prior to performing EDA, we will examine the number of unique subjects to ensure no erroneous duplication has occurred. Since we are clustering the risk for bipolar disorder (BD) variables at baseline, we will then create a copy of the data at the baseline assessment timepoint. Finally, for clustering variables that require multiple raw input variables to create them (e.g., first degree relative depression history), we will create a singular, cleaned variable for each. 

```{r data wrangling, echo = FALSE, include = FALSE, warning = FALSE}

## Data Wrangling ##

#1. Check the number of unique subjects to ensure no data was lost/erroneously created during merging
unique_count <- length(unique(raw_data$participant_id))
if (unique_count >= 11800 && unique_count <= 12000) {
  message("The number of unique src_subject_id (", unique_count, ") is within the acceptable range.")
} else {
  message("The number of unique src_subject_id (", unique_count, ") is outside the acceptable range!")
}

#2. Clean the raw data 
#2.1 Filter the trimmed data to only contain rows at the baseline assessment timepoint
raw_data_baseline <- raw_data %>% 
  filter(session_id == "ses-00A")

#3. Preliminarily create relevant risk clustering variables
#3.1 Family (first-degree relative) history of depression
raw_data_baseline_clean <- raw_data_baseline %>%
  mutate(
    family_history_depression = case_when(
      
      #3.1.1 Assign 1 if any first-degree relative has a history of depression
      replace_na(mh_p_famhx__dep__fath_001 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__moth_001 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__young_001 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__young_002 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__young_003 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__young_004 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__young_005 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__old_001 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__old_002 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__old_003 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__old_004 == 1, FALSE) |
        replace_na(mh_p_famhx__dep__full__old_005 == 1, FALSE) ~ 1,
      
      #3.1.2 Assign 0 if "mh_p_famhx__dep_001" (any relative) is 1 but no first-degree relative has a history of depression
      mh_p_famhx__dep_001 == 1 ~ 0,
      
      #3.1.3 Assign 0 if "mh_p_famhx__dep_001" is 0
      mh_p_famhx__dep_001 == 0 ~ 0,
      
      #3.1.4 Otherwise, assign NA
      TRUE ~ NA_real_
    )
  )

#3.2 Family (first-degree relative) history of mania
raw_data_baseline_clean <- raw_data_baseline_clean %>%
  mutate(
    family_history_mania = case_when(
      
      #3.2.1 Assign 1 if any first-degree relative has a history of mania
      replace_na(mh_p_famhx__mania__fath_001 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__moth_001 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__young_001 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__young_002 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__young_003 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__young_004 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__young_005 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__old_001 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__old_002 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__old_003 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__old_004 == 1, FALSE) |
        replace_na(mh_p_famhx__mania__full__old_005 == 1, FALSE) ~ 1,
      
      #3.2.2 Assign 0 if "mh_p_famhx__mania_001" (any relative) is 1 but no first-degree relative has a history of mania
      mh_p_famhx__mania_001 == 1 ~ 0,
      
      #3.2.3 Assign 0 if "mh_p_famhx__mania_001" is 0
      mh_p_famhx__mania_001 == 0 ~ 0,
      
      #3.2.4 Otherwise, assign NA
      TRUE ~ NA_real_
    )
  )

#3.3 Sum ACE Score Index
raw_data_baseline_clean <- raw_data_baseline_clean %>%
  mutate(
    
    #3.3.1.1 Calculate each ACE component
    ACE_1 = if_else(mh_p_famhx__troub_001 == 1, 1, 0, missing = NA),
    ACE_2 = if_else(fc_y_crpbi__cg1_005 %in% c(1, 2), 1, 0, missing = NA),
    ACE_3 = if_else(mh_p_famhx__suic_001 == 1 | mh_p_famhx__dep_001 == 1, 1, 0, missing = NA),
    ACE_4 = if_else(mh_p_famhx__alc_001 == 1 | mh_p_famhx__drg_001 == 1, 1, 0, missing = NA),
    ACE_5 = if_else(ksads_ptsd_raw_766_p == 1 | fc_y_fes__confl_006 == 1 | fc_y_fes__confl_003 == 1, 1, 0, missing = NA),
    ACE_6 = if_else(ab_p_demo__marital__slf_001 %in% c(2, 3, 4), 1, 0, missing = NA),
    ACE_7 = if_else(ksads_ptsd_raw_762_p == 1 | ksads_ptsd_raw_763_p == 1, 1, 0, missing = NA),
    ACE_8 = if_else(fc_y_pm_001 %in% c(1, 2) | fc_y_pm_003 %in% c(1, 2), 1, 0, missing = NA),
    ACE_9 = if_else(ksads_ptsd_raw_767_p == 1 | ksads_ptsd_raw_768_p == 1, 1, 0, missing = NA),
    
    #3.3.1.2 Sum the ACE components for each participant
    ACE_index_sum_score = rowSums(across(starts_with("ACE_")), na.rm = TRUE),
    
    #3.3.2 Count the number of missing values for the ACE components
    ACE_index_sum_score_n_missing = rowSums(
      across(c(
        mh_p_famhx__troub_001, 
        fc_y_crpbi__cg1_005, 
        mh_p_famhx__suic_001, 
        mh_p_famhx__dep_001, 
        mh_p_famhx__alc_001, 
        mh_p_famhx__drg_001, 
        ksads_ptsd_raw_766_p, 
        fc_y_fes__confl_006, 
        fc_y_fes__confl_003,
        ab_p_demo__marital__slf_001, 
        ksads_ptsd_raw_762_p, 
        ksads_ptsd_raw_763_p, 
        fc_y_pm_001, 
        fc_y_pm_003, 
        ksads_ptsd_raw_767_p, 
        ksads_ptsd_raw_768_p
      ), ~ . %in% c(NA, "", 999, 7, 777))),
    
    #3.3.3 Set sum score to NA if any components are missing
    ACE_index_sum_score = if_else(ACE_index_sum_score_n_missing >= 1, NA_real_, ACE_index_sum_score)
  )

#3.4 History of Bullying (Parent and/or Youth Reported)
raw_data_baseline_clean <- raw_data_baseline_clean %>%
  
  #3.4.1 Determine whether the parent or youth endorsed the youth being bullied
  mutate(bullying = case_when(
    mh_p_kbi__bully_001 == 1 | mh_y_kbi__bully_001 == 1 ~ 1, 
    is.na(mh_p_kbi__bully_001) & is.na(mh_y_kbi__bully_001) ~ NA_real_,
    mh_p_kbi__bully_001 == "" & mh_y_kbi__bully_001 == "" ~ NA_real_,
    TRUE ~ 0)
    )

#3.5 SDSC Total (Sum) Score
#3.5.1 Define the column names to be summed
sds_columns <- c(
  "ph_p_sds__da_001", "ph_p_sds__da_002", "ph_p_sds__da_003",
  "ph_p_sds__dims_001", "ph_p_sds__dims_002", "ph_p_sds__dims_003",
  "ph_p_sds__dims_004", "ph_p_sds__dims_005", "ph_p_sds__dims_006",
  "ph_p_sds__dims_007", "ph_p_sds__does_001", "ph_p_sds__does_002",
  "ph_p_sds__does_003", "ph_p_sds__does_004", "ph_p_sds__does_005",
  "ph_p_sds__hyphy_001", "ph_p_sds__hyphy_002", "ph_p_sds__sbd_001",
  "ph_p_sds__sbd_002", "ph_p_sds__sbd_003", "ph_p_sds__swtd_001",
  "ph_p_sds__swtd_002", "ph_p_sds__swtd_003", "ph_p_sds__swtd_004",
  "ph_p_sds__swtd_005", "ph_p_sds__swtd_006"
)

#3.5.2 Write a function to calculate total SDSC score for each participant row
calculate_sds_total <- function(data) {
  
  #3.5.2.1 For each row, check if ANY column has NA
  has_missing <- apply(data[, sds_columns], 1, function(row) any(is.na(row)))
  
  #3.5.2.2 Calculate the sum for each row
  raw_sums <- rowSums(data[, sds_columns], na.rm = TRUE)
  
  #3.5.2.3 Set the sum to NA for any row with missing values
  data$sds_total <- ifelse(has_missing, NA, raw_sums)
  
  #3.524 Return the data with the sum SDSC score
  return(data)
}

#3.5.3 Apply the SDSC score calculation function to the baseline data
raw_data_baseline_clean <- calculate_sds_total(raw_data_baseline_clean)

#4. Retain columns of interest for risk clustering
#4.1 Create a dataframe containing columns of interest & any respective nm columns
raw_data_baseline_clean_trim <- raw_data_baseline_clean %>% 
  dplyr::select(c(participant_id, session_id, mh_p_cbcl__dsm__dep_tscore, mh_p_cbcl__dsm__dep_nm, mh_p_cbcl__dsm__anx_tscore, mh_p_cbcl__dsm__anx_nm, mh_p_cbcl__synd__attn_tscore, mh_p_cbcl__synd__attn_nm, mh_p_cbcl__synd__aggr_tscore, mh_p_cbcl__synd__aggr_nm, mh_p_gbi_sum, mh_y_upps__nurg_sum, mh_y_upps__nurg_nm, mh_y_upps__purg_sum, mh_y_upps__purg_nm, le_l_coi__addr1__coi__total__national_zscore, fc_p_nsc__ns_mean, fc_p_nsc__ns_nm, sds_total, family_history_depression, family_history_mania, bullying, nc_y_nihtb__lswmt__uncor_score, nc_y_nihtb__lswmt__agecor_score, nc_y_nihtb__flnkr__uncor_score, nc_y_nihtb__flnkr__agecor_score, nc_y_nihtb__pttcp__uncor_score, nc_y_nihtb__pttcp__agecor_score, ACE_1, ACE_2, ACE_3, ACE_4, ACE_5, ACE_6, ACE_7, ACE_8, ACE_9, ACE_index_sum_score, ACE_index_sum_score_n_missing))

#4.2 Use nm columns to turn any respective instances of values created using missing data to NA
raw_data_baseline_clean_trim <- raw_data_baseline_clean_trim %>%
  mutate(
    
    #4.2.1 CBCL Depression
    mh_p_cbcl__dsm__dep_tscore = if_else(mh_p_cbcl__dsm__dep_nm >= 1, NA_real_, mh_p_cbcl__dsm__dep_tscore),
    
    #4.2.2 CBCL Anxiety
    mh_p_cbcl__dsm__anx_tscore = if_else(mh_p_cbcl__dsm__anx_nm >= 1, NA_real_, mh_p_cbcl__dsm__anx_tscore),
    
    #4.2.3 CBCL Attention
    mh_p_cbcl__synd__attn_tscore = if_else(mh_p_cbcl__synd__attn_nm >= 1, NA_real_, mh_p_cbcl__synd__attn_tscore),
    
    #4.2.4 CBCL Aggression
    mh_p_cbcl__synd__aggr_tscore = if_else(mh_p_cbcl__synd__aggr_nm >= 1, NA_real_, mh_p_cbcl__synd__aggr_tscore),
    
    #4.2.5 UPPS Negative Urgency
    mh_y_upps__nurg_sum = if_else(mh_y_upps__nurg_nm >= 1, NA_real_, mh_y_upps__nurg_sum),
    
    #4.2.6 UPPS Positive Urgency
    mh_y_upps__purg_sum = if_else(mh_y_upps__purg_nm >= 1, NA_real_, mh_y_upps__purg_sum),
    
    #4.2.7 Neighborhood Safety Score
    fc_p_nsc__ns_mean = if_else(fc_p_nsc__ns_nm >= 1, NA_real_, fc_p_nsc__ns_mean)
  )

#4.3 Further trim the cleaned data to only include columns of interest to risk clustering
raw_data_baseline_clean_trim <- raw_data_baseline_clean_trim %>% 
  dplyr::select(c(participant_id, session_id, mh_p_cbcl__dsm__dep_tscore, mh_p_cbcl__dsm__anx_tscore, mh_p_cbcl__synd__attn_tscore, mh_p_cbcl__synd__aggr_tscore, mh_p_gbi_sum, mh_y_upps__nurg_sum, mh_y_upps__purg_sum, le_l_coi__addr1__coi__total__national_zscore, fc_p_nsc__ns_mean, sds_total, family_history_depression, family_history_mania, bullying, nc_y_nihtb__lswmt__uncor_score, nc_y_nihtb__lswmt__agecor_score, nc_y_nihtb__flnkr__uncor_score, nc_y_nihtb__flnkr__agecor_score, nc_y_nihtb__pttcp__uncor_score, nc_y_nihtb__pttcp__agecor_score, ACE_1, ACE_2, ACE_3, ACE_4, ACE_5, ACE_6, ACE_7, ACE_8, ACE_9, ACE_index_sum_score))

```

## 1. Define the N, percent, and range of risk variables of interest at the baseline assessment timepoint

```{r EDA 1, echo = FALSE, warning = FALSE}

## Exploratory Data Analysis ##

#1. Define the N, percent, and range of risk variables of interest at the baseline assessment timepoint
#1.1 Define the columns to analyze
columns_to_check <- c(
  "mh_p_cbcl__dsm__dep_tscore", 
  "mh_p_cbcl__dsm__anx_tscore", 
  "mh_p_cbcl__synd__attn_tscore", 
  "mh_p_cbcl__synd__aggr_tscore", 
  "mh_p_gbi_sum", 
  "mh_y_upps__nurg_sum", 
  "mh_y_upps__purg_sum", 
  "le_l_coi__addr1__coi__total__national_zscore", 
  "fc_p_nsc__ns_mean", 
  "sds_total", 
  "family_history_depression", 
  "family_history_mania", 
  "bullying",
  "nc_y_nihtb__lswmt__uncor_score", 
  "nc_y_nihtb__lswmt__agecor_score", 
  "nc_y_nihtb__flnkr__uncor_score", 
  "nc_y_nihtb__flnkr__agecor_score", 
  "nc_y_nihtb__pttcp__uncor_score", 
  "nc_y_nihtb__pttcp__agecor_score", 
  "ACE_1", 
  "ACE_2", 
  "ACE_3", 
  "ACE_4", 
  "ACE_5", 
  "ACE_6",
  "ACE_7",
  "ACE_8", 
  "ACE_9",
  "ACE_index_sum_score")

#1.2 Create a function to compute the required metrics for each column
get_column_metrics <- function(column_name, data, total_subjects = nrow(data)) {
  
  #1.2.1 Extract each column
  column_data <- data[[column_name]]
  
  #1.2.2 Define universal missing value codes instead of variable-specific ones
  missing_values <- c(999, NA, NaN, "")
  
  #1.2.3 Filter out missing values
  valid_values <- column_data[!is.na(column_data) &
                               column_data != "" &
                               !column_data %in% c(999, NaN)]
  
  #1.2.4 Compute metrics
  n_non_na <- length(valid_values)
  percent_non_na <- (n_non_na / total_subjects) * 100
  
  if (is.numeric(valid_values)) {
    range_or_unique <- range(valid_values, na.rm = TRUE)
  } else {
    range_or_unique <- unique(valid_values)
  }
  
  #1.2.5 Return the metrics as a list
  list(
    column = column_name,
    n_non_na = n_non_na,
    percent_non_na = percent_non_na,
    range_or_unique = range_or_unique
  )
}

#1.3 Apply the function to all columns and store risk data missingness in a list
raw_data_baseline_clean_trim_missingness_list <- lapply(columns_to_check, get_column_metrics, data = raw_data_baseline_clean_trim)

#1.4 Convert the risk data missingness into a dataframe
raw_data_baseline_clean_trim_missingness_df <- do.call(rbind, lapply(raw_data_baseline_clean_trim_missingness_list, function(x) {
  data.frame(
    column = x$column,
    n_non_na = x$n_non_na,
    percent_non_na = x$percent_non_na,
    range_or_unique = I(list(x$range_or_unique))  # Use I() to preserve list structure for unique/range
  )
}))

#1.5 Print the final dataframe of raw_data_baseline_clean_trim_missingness_
kable(raw_data_baseline_clean_trim_missingness_df, col.names = c("Risk Variable", "N True Values", "Percent True Values", "Range or Unique Values in Data"), digits = 2) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))

```

## 2. Visually and Quantitatively Assess for any Outliers in the Data

```{r EDA 2, echo = FALSE, warning = FALSE, fig.height=10, fig.width=10}

## Exploratory Data Analysis ##

#2. Visually and Quantitatively Assess for any Outliers in the Data
#2.1 Create boxplots with outlier indicators for continuous variables in the dataset
#2.11 Filter only continuous variables (excluding ID and gender)
risk_data_continuous_variables <- raw_data_baseline_clean_trim %>%
  dplyr::select(mh_p_cbcl__dsm__dep_tscore, mh_p_cbcl__dsm__anx_tscore, mh_p_cbcl__synd__attn_tscore, mh_p_cbcl__synd__aggr_tscore, mh_p_gbi_sum, mh_y_upps__nurg_sum, mh_y_upps__purg_sum, le_l_coi__addr1__coi__total__national_zscore, fc_p_nsc__ns_mean, sds_total, nc_y_nihtb__lswmt__uncor_score, nc_y_nihtb__lswmt__agecor_score, nc_y_nihtb__flnkr__uncor_score, nc_y_nihtb__flnkr__agecor_score, nc_y_nihtb__pttcp__uncor_score, nc_y_nihtb__pttcp__agecor_score, ACE_index_sum_score) %>%
  gather(key = "Variable", value = "Value")

#2.12 Create boxplot with facets for continuous variables
risk_data_continuous_variables_boxplot <- ggplot(risk_data_continuous_variables, aes(x = Variable, y = Value)) + 
  geom_boxplot(fill = "lightblue", color = "darkblue") + 
  facet_wrap(~ Variable, scales = "free") + 
  theme_minimal() + 
  theme(axis.text.x = element_blank(),
        strip.text = element_text(size = 6, face = "bold")) +
  labs(title = "Boxplots of Continuous Variables",
       y = "Values",
       x = "")

#2.13 Print the plot
print(risk_data_continuous_variables_boxplot)

```

From the boxplots for continuous variables, it is evident that some of the NIHTB scores, particularly those that have been age-corrected, are likely out of the expected range (within ~3 SD of the mean, with mean centered ~ 100). While no guidelines within the [overview of the ABCD baseline neurocognition battery](https://www.sciencedirect.com/science/article/pii/S1878929317302384?via%3Dihub#tbl0005) or mention of the extraneously high/low scores within [previous release notes](https://wiki.abcdstudy.org/release-notes/non-imaging/neurocognition.html#nih-toolbox-cognition) could be located, it is likely that winsorization or other outlier removal methods would need to be applied to handle values beyond the expected range. 

Following review of the [NIH Toolbox (Cognition) Instrument Description](https://wiki.abcdstudy.org/release-notes/non-imaging/neurocognition.html#nih-toolbox-cognition) (see **"Notes and Special Considerations"**), it is noted by the neurocognitive working group that "For longitudinal analyses, we recommend using either uncorrected Scaled Scores or raw scores." Following review of the distributions below, a tentative decision regarding which strategy to use will be chosen.

No other anomalous or outlier values are noteworthy or unexpected.

## 3. Appropriateness of the Mean as a Measure of Central Tendency

Though proposed clustering methods for the current study do not have parametric assumptions, assessment of kurtosis & skewness provides information about the asymmetry of variable distributions, which may help identify variables with extreme outliers or long tails. Such variables can disproportionately influence cluster formation when using euclidean distance (as in k-prototypes for continuous variables), and thus it is important to examine them. The skewness & kurtosis values for continuous variables herein are as such: 

```{r EDA 3, echo = FALSE, warning = FALSE}

## Exploratory Data Analysis ##

#3.1 Calculate skewness & kurtosis for continuous variables
risk_data_continuous_variables_skewness_kurtosis_table <- raw_data_baseline_clean_trim %>%
  dplyr::select(
    mh_p_cbcl__dsm__dep_tscore,
    mh_p_cbcl__dsm__anx_tscore,
    mh_p_cbcl__synd__attn_tscore,
    mh_p_cbcl__synd__aggr_tscore,
    mh_p_gbi_sum,
    mh_y_upps__nurg_sum,
    mh_y_upps__purg_sum,
    le_l_coi__addr1__coi__total__national_zscore,
    fc_p_nsc__ns_mean,
    sds_total,
    nc_y_nihtb__lswmt__uncor_score,
    nc_y_nihtb__lswmt__agecor_score,
    nc_y_nihtb__flnkr__uncor_score,
    nc_y_nihtb__flnkr__agecor_score,
    nc_y_nihtb__pttcp__uncor_score,
    nc_y_nihtb__pttcp__agecor_score,
    ACE_index_sum_score) %>%
  summarise(
    
  #3.1.1 Calculate skewness & kurtosis for each variable
    across(everything(), list(
      skewness = ~ skewness(.x, na.rm = TRUE),
      kurtosis = ~ kurtosis(.x, na.rm = TRUE)))) %>%
  
  #3.1.2 Reshape from wide to long format
  pivot_longer(
    cols = everything(),
    names_to = c("Variable", "Statistic"),
    names_pattern = "(.*)_(skewness|kurtosis)",
    values_to = "Value") %>%
  
  #3.1.3 Reshape back to have skewness and kurtosis as separate columns
  pivot_wider(names_from = Statistic, values_from = Value)

#3.2 Display the skewness & kurtosis table
kable(risk_data_continuous_variables_skewness_kurtosis_table, caption = "Skewness & Kurtosis Values for Continuous Variables", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))

```

Results from the assessment of skewness and kurtosis indicate several variables exhibiting substantial deviations from normality. Specifically, variables such as mh_p_gbi_sum (GBI Mania Sum Score; skewness = 3.49, kurtosis = 18.58), mh_p_cbcl__synd__aggr_tscore (CBCL Aggression T-score; skewness = 2.76, kurtosis = 12.04), mh_p_cbcl__synd__attn_tscore (CBCL Attention Problems T-score; skewness = 2.47, kurtosis = 10.62), and mh_p_cbcl__dsm__anx_tscore (CBCL DSM-5 Anxiety T-score; skewness = 2.51, kurtosis = 10.11) display notable positive skewness and leptokurtic distributions. This indicates that these measures are characterized by an abundance of lower-end values and a heavy tail with more extreme values, consistent with the expected distributions of symptom-related scales in a community sample where most participants report minimal psychopathology (and which is consistent with expected [low rates of mania in this age range](https://pmc.ncbi.nlm.nih.gov/articles/PMC8233426/)).

Additionally, variables assessing cognitive and socioeconomic factors, such as le_l_coi__addr1__coi__total__national_zscore (Child Opportunity Index; skewness = -0.78, kurtosis = 3.08), fc_p_nsc__ns_mean (Neighborhood Safety Mean Score; skewness = -0.88, kurtosis = 3.34), and nc_y_nihtb__flnkr__uncor_score (NIH Toolbox Flanker Uncorrected Score; skewness = -1.00, kurtosis = 4.49), exhibit moderate negative skewness, indicating higher scores are more frequent, which aligns with expected performance and demographic trends in such a sample.

None of the skewness or kurtosis values alone necessarily warrant immediate transformations. Given the expected nature of these distributions in the context of symptom and cognitive scores within this developmental population, downstream analyses—particularly the clustering procedures and subsequent longitudinal survival and logistic regression analyses—should incorporate awareness of these distributional characteristics during parameter tuning, but are still likely appropriate to use. Monitoring the impact of these skewed and leptokurtic distributions on analytical outcomes will be essential, and transformation or robust analytic methods may be considered if these distributions compromise statistical assumptions or model performance.

## 4. Shape of Each Variables' Distribution

The shape of the distribution for each continuous variable and frequency distribution of each categorical variable is outlined below, both in terms of summary statistics derived from and a graphical histogram with density overlay of relevant data:

```{r EDA 4, echo = FALSE, warning = FALSE, fig.height=10, fig.width=10}

## Exploratory Data Analysis ##

#4. Create summary tables and visualizations for the distributions of categorical and continuous variables included in the risk model
#4.1 Create a summary table of distribution information
#4.11 Specify continuous columns explicitly
continuous_columns <- c(
  "mh_p_cbcl__dsm__dep_tscore", "mh_p_cbcl__dsm__anx_tscore", 
  "mh_p_cbcl__synd__attn_tscore", "mh_p_cbcl__synd__aggr_tscore", 
  "mh_p_gbi_sum", "mh_y_upps__nurg_sum", 
  "mh_y_upps__purg_sum", "le_l_coi__addr1__coi__total__national_zscore", 
  "fc_p_nsc__ns_mean", "sds_total", 
  "nc_y_nihtb__lswmt__uncor_score", "nc_y_nihtb__lswmt__agecor_score", 
  "nc_y_nihtb__flnkr__uncor_score", "nc_y_nihtb__flnkr__agecor_score", 
  "nc_y_nihtb__pttcp__uncor_score", "nc_y_nihtb__pttcp__agecor_score", 
  "ACE_index_sum_score")

#4.12 Define all columns to check
columns_to_check <- colnames(raw_data_baseline_clean_trim)

#4.131 Derive categorical columns as those not in the continuous list
categorical_columns <- setdiff(columns_to_check, continuous_columns)
categorical_columns <- categorical_columns[!categorical_columns %in% c("participant_id", "session_id")]

#4.132 Define new titles for individual ACE variables
categorical_title_map <- c(
  "ACE_1" = "Criminal Household Member",
  "ACE_2" = "Emotional neglect",
  "ACE_3" = "Household Depression or SI",
  "ACE_4" = "Household substance abuse",
  "ACE_5" = "Mother treated violently",
  "ACE_6" = "Parental separation or divorce",
  "ACE_7" = "Physical abuse",
  "ACE_8" = "Physical neglect",
  "ACE_9" = "Sexual Abuse"
  )

#4.14 Create a summary table for continuous and categorical variables
# Adjust summary function for categorical columns
summarize_data <- function(data, continuous_cols, categorical_cols) {
  summary_list <- list()
  
  # Continuous variables summary
  for (col in continuous_cols) {
    column_data <- data[[col]]
    valid_values <- column_data[!is.na(column_data)]
    summary_list[[col]] <- data.frame(
      Variable = col,
      Type = "Continuous",
      Mean = mean(valid_values, na.rm = TRUE),
      Median = median(valid_values, na.rm = TRUE),
      SD = sd(valid_values, na.rm = TRUE),
      Min = min(valid_values, na.rm = TRUE),
      Max = max(valid_values, na.rm = TRUE),
      Levels = NA,
      Frequencies = NA,
      stringsAsFactors = FALSE
    )
  }
  
  # Categorical variables summary
  for (col in categorical_cols) {
    column_data <- data[[col]]
    freq_table <- table(column_data, useNA = "no")
    summary_list[[col]] <- data.frame(
      Variable = col,
      Type = "Categorical",
      Mean = NA,
      Median = NA,
      SD = NA,
      Min = NA,
      Max = NA,
      Levels = paste(names(freq_table), collapse = ", "),
      Frequencies = paste(freq_table, collapse = ", "),
      stringsAsFactors = FALSE
    )
  }
  
  do.call(rbind, summary_list)
}

#4.151 Generate the summary table
summary_table <- summarize_data(raw_data_baseline_clean_trim, continuous_columns, categorical_columns)

#4.152 Rename the ACE variables in the summary table
summary_table <- summary_table %>%
  mutate(Variable = ifelse(Variable %in% names(categorical_title_map), categorical_title_map[Variable], Variable))

#4.153 Print the summary table in the RMD
kable(summary_table, caption = "Summary Statistics for Risk Model Variables", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))

#4.2 Create plots visualizing distributions of risk variables
#4.21 Generate a combined dataset with an indicator for variable type
plot_data <- list()

#4.22 Add continuous variables
for (col in continuous_columns) {
  plot_data[[col]] <- raw_data_baseline_clean_trim %>%
    select(Value = !!sym(col)) %>%
    mutate(Type = "Continuous", Variable = col)
}

#4.23 Add categorical variables
for (col in categorical_columns) {
  plot_data[[col]] <- raw_data_baseline_clean_trim %>%
    select(Value = !!sym(col)) %>%
    mutate(Type = "Categorical", Variable = col)
}

#4.24 Combine all into a single data frame
plot_data_combined <- bind_rows(plot_data)

#4.25 Generate and print the continuous variable facet plot
#4.251 Notate the COI variable for the purpose of focusing the distribution in the COI subplot
coi_var  <- "le_l_coi__addr1__coi__total__national_zscore"

#4.252 Generate facet plot for continuous variables
continuous_plot <- ggplot() +
  
  #4.2521 Plots for all continuous variables except COI, full range
  geom_histogram(
    data = plot_data_combined %>% filter(Type == "Continuous" & Variable != coi_var),
    aes(x = Value, y = ..density..),
    binwidth = 1,
    fill = "blue",
    alpha = 0.6,
    color = "black") +
  geom_density(
    data = plot_data_combined %>% filter(Type == "Continuous" & Variable != coi_var),
    aes(x = Value),
    color = "red",
    size = 1,
    adjust = 1.5) +
  
  #4.2522 COI variable plot, restricted to range of COI variable (which for some reason free x scale doesn't capture accurately)
  geom_histogram(
    data = plot_data_combined %>%
      filter(
        Type == "Continuous",
        Variable == coi_var,
        between(Value, -0.02, 0.07)),
    aes(x = Value, y = ..density..),
    binwidth = 0.005,
    fill = "blue",
    alpha = 0.6,
    color = "black") +
  geom_density(
    data = plot_data_combined %>%
      filter(
        Type == "Continuous",
        Variable == coi_var,
        between(Value, -0.02, 0.07)),
    aes(x = Value),
    color = "red",
    size = 1,
    adjust = 1.5) +
  
  #4.2523 Facet wrap all subplots and add title/thematic components
  facet_wrap(~ Variable, ncol = 3, scales = "free") +
  theme_minimal() +
  labs(title = "Distributions of Continuous Variables", x = "Value", y = "Density") +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(size = 8),
    panel.spacing = unit(1, "lines"))

#4.252 Print the continuous variable plot
print(continuous_plot)

#4.261 Generate facet plot for categorical variables
#4.2611 Generate the categorical variable plot
categorical_plot <- ggplot(
  plot_data_combined %>% filter(Type == "Categorical"),
  aes(x = as.factor(Value))) +
  geom_bar(fill = "skyblue", color = "black", alpha = 0.8) +
  geom_text(aes(label = ..count..), stat = "count", vjust = -0.5, size = 3) + # Add count labels above each bar
  facet_wrap(~Variable, ncol = 3, scales = "free",
             labeller = labeller(Variable = categorical_title_map)) +
  theme_minimal() +
  labs(
    title = "Distributions of Categorical Variables",
    x = "Categories",
    y = "Count") +
  scale_x_discrete(limits = c("0", "1"), labels = c("No", "Yes")) +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(size = 8),
    panel.spacing = unit(1, "lines"))

#4.2612 Expand the y‐axis on the categorical plot so labels don’t get cut off
categorical_plot <- categorical_plot +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  theme(
    plot.margin = margin(5, 5, 15, 5))

#4.2613 Print the categorical variable plot
print(categorical_plot)

```

Distributions of continuous variables and rates of categorical variables (including the individual ACEs comprising the sum ACE index score) are shown above. While some previously mentioned variables exhibit noticeable skew, no problematic distributions are notable here. 

Per previous mention in section 2 (assessing outliers in the data), following review of the distributions above, I make the tentative choice of utilizing the uncorrected standard scores (as opposed to winsorizing or otherwise subjectively selecting a procedure to clean up the age corrected standard scores). This choice is for two key reasons: 

  1. This selection aligns with the [NIH Toolbox (Cognition) Instrument Description](https://wiki.abcdstudy.org/release-notes/non-imaging/neurocognition.html#nih-toolbox-cognition) (see **"Notes and Special Considerations"**), at least for longitudinal usage of the NIH Toolbox cognitive measures, and; 
  
  2. The means, standard deviations, and full ranges of the uncorrected standard scores align more closely with the expected descriptive statistics of a measure of IQ (i.e., μ = 100; σ = 15), and contain less significant outliers than the age corrected standard scores of interest
  
This decision is tentative, however, and subject to future review/change in the event new information comes to light. 

## 5. Overall Availability of Risk Model Variables for Clustering

```{r EDA 5, echo = FALSE, warning = FALSE}

## Exploratory Data Analysis ##

#5. Determine the broad availability of variables to be included in the risk model
#5.1 Determine the number of unique subjects in the dataset
unique_subject_count <- as.numeric(length(unique(raw_data_baseline_clean_trim$participant_id)))

#5.2 Create a vector of counts for complete cases across specified columns
complete_cases_count <- raw_data_baseline_clean_trim %>%
  dplyr::select(matches("mh_p_cbcl__dsm__dep_tscore|mh_p_cbcl__dsm__anx_tscore|mh_p_cbcl__synd__attn_tscore|mh_p_cbcl__synd__aggr_tscore|mh_p_gbi_sum|mh_y_upps__nurg_sum|mh_y_upps__purg_sum|le_l_coi__addr1__coi__total__national_zscore|fc_p_nsc__ns_mean|sds_total|nc_y_nihtb__lswmt__uncor_score|nc_y_nihtb__lswmt__agecor_score|nc_y_nihtb__flnkr__uncor_score|nc_y_nihtb__flnkr__agecor_score|nc_y_nihtb__pttcp__uncor_score|nc_y_nihtb__pttcp__agecor_score|ACE_index_sum_score")) %>%
  filter_all(all_vars(!is.na(.) & . != "" & . != 999 & . != 777)) %>%
  nrow()

```

When looking across all relevant risk variables in the dataset, `r paste(complete_cases_count)` `r paste0("(", round((complete_cases_count/unique_subject_count)*100, 2), "%)")` subjects have complete (i.e., fully reported, non-missing) data. 