---
title: "Mood Disorder & Suicidality Risk Variable EDA"
author: "Sam A. Sievertsen"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(warning = FALSE, message = FALSE, comment = "")

```

# Exploratory Data Analysis (EDA) of variables to be included in the clustering of groups at potential risk for the development of mood disorders and suicidality in the ABCD dataset

The most up to date list of variables to be incorporated in the risk clustering, including their role, domain, details of each variable as it relates to the ABCD dataset, empirical justification for inclusion in the clustering, etc., can be [found here](https://ohsuitg-my.sharepoint.com/:x:/r/personal/sievertsen_ohsu_edu/Documents/Huber_Lab/FYP/Project_Planning/ABCD_Mood_Disorder_Risk_Model_Variables_010624_SS.xlsx?d=w286583f8914b4ab5a402e36ce8c36291&csf=1&web=1&e=wmfy1l&nav=MTVfezk2MDA5QjlDLUQyNkQtNDFBRS04NTA5LUFDQUQwOEZGMUJFN30). 

Additionally, details regarding the incorporation of the sum adverse childhood experience (ACE) index and specific variables within it, as first recommended by [Hoffman et al., (2019)](https://www.sciencedirect.com/science/article/pii/S2352289518300821?via%3Dihub) and iterated upon by [Raney et al., (2022)](https://pmc.ncbi.nlm.nih.gov/articles/PMC9288265/) can be [located here](https://ohsuitg-my.sharepoint.com/:x:/r/personal/sievertsen_ohsu_edu/Documents/Huber_Lab/FYP/Project_Planning/ABCD_Mood_Disorder_Risk_Model_Variables_010624_SS.xlsx?d=w286583f8914b4ab5a402e36ce8c36291&csf=1&web=1&e=mr4ikT&nav=MTVfe0Y1OUEwNTY2LTY2MTUtNDNCOC1BQUQ4LTgwNkI0QkVDODZEQn0). 

The following report contains information regarding the: 

  1. Integrity of the data
  
  2. Assessment of outliers in the data + preliminary thoughts on how to address them
  
  3. Skewness + appropriateness of the mean as the measure of central tendency for continuous variables
  
  4. Shape of each variables' distribution
  
  5. A summary of the overall availability of variables of interest to the current study

```{r environment, echo = FALSE, include = FALSE, warning = FALSE}

# Load necessary packages + environmental variables
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(moments)
library(skimr)
options(scipen = 999, digits = 8)

# Read in cbcl data
cbcl_data <- read.csv("../../data/data_raw/mh_p_cbcl.csv")

# Read in general behavior inventory data
gbi_data <- read.csv("../../data/data_raw/mh_p_gbi.csv")

# Read in upps-p data
uppsp_data <- read.csv("../../data/data_raw/mh_y_upps.csv")

# Read in sleep disturbance scale data
sleep_disturbance_data <- read.csv("../../data/data_raw/ph_p_sds.csv")

# Read in family history data
family_hx_data <- read.csv("../../data/data_raw/mh_p_fhx.csv")

# Read in childhood opportunity index data
coi_data <- read.csv("../../data/data_raw/led_l_coi.csv")

# Read in parent KSADS-COMP background data
parent_ksads_bg_data <- read.csv("../../data/data_raw/mh_p_ksads_bg.csv")

# Read in youth KSADS-COMP background data
youth_ksads_bg_data <- read.csv("../../data/data_raw/mh_y_ksads_bg.csv")

# Read in neighborhood safety data
neighborhood_safety_data <- read.csv("../../data/data_raw/ce_p_nsc.csv")

# Read in NIH-toolbox data
nihtb_data <- read.csv("../../data/data_raw/nc_y_nihtb.csv")

# Read in ACE Index Data
# KSADS-COMP PTSD Questions
ksads_ptsd_data <- read.csv("../../data/data_raw/mh_p_ksads_ptsd.csv")

# Parent reported demographics survey - family conflict subscale
parent_demo_conflict_data <- read.csv("../../data/data_raw/ce_p_fes.csv")

# Youth reported parental monitoring survey
youth_parental_monitoring_data <- read.csv("../../data/data_raw/ce_y_fes.csv")

# Family history data read in above

# Parent reported demographic data
parent_demographic_data <- read.csv("../../data/data_raw/abcd_p_demo.csv")

# Youth caregiver acceptance
youth_crpbi_data <- read.csv("../../data/data_raw/ce_y_crpbi.csv")

# Parental monitoring survey
parental_monitoring_data <- read.csv("../../data/data_raw/ce_y_pm.csv")

```



```{r data wrangling, echo = FALSE, include = FALSE, warning = FALSE}

## Data Wrangling ##

#1. Merge the data
#1.1 Collapse the data frames containing risk variables of interest
risk_data_merged <- Reduce(
  function(x, y)
    merge(x, y,
      by = c("src_subject_id", "eventname"),
      all = TRUE),
  list(cbcl_data, gbi_data, uppsp_data, sleep_disturbance_data, family_hx_data, coi_data, parent_ksads_bg_data, youth_ksads_bg_data, neighborhood_safety_data, nihtb_data, ksads_ptsd_data, parent_demo_conflict_data, youth_parental_monitoring_data, parent_demographic_data, youth_crpbi_data, parental_monitoring_data))

#1.2 Verify the merge did not duplicate any columns and resulted in an expected 
#1.21 Check for duplicate columns
duplicate_columns <- grep("\\.x$|\\.y$", colnames(risk_data_merged), value = TRUE)
if (length(duplicate_columns) > 0) {
  message("Duplicate columns found: ", paste(duplicate_columns, collapse = ", "))
} else {
  message("No duplicate columns ending found.")
}

#1.22 Check the number of unique subjects to ensure no data was lost/erroneously created during merging
unique_count <- length(unique(risk_data_merged$src_subject_id))
if (unique_count >= 11800 && unique_count <= 12000) {
  message("The number of unique src_subject_id (", unique_count, ") is within the acceptable range.")
} else {
  message("The number of unique src_subject_id (", unique_count, ") is outside the acceptable range!")
}

#2. Clean the merged data 
#2.1 Retain only columns of interest
risk_data_merged_trimmed <- risk_data_merged %>% 
  dplyr::select(c(src_subject_id, eventname, cbcl_scr_dsm5_depress_t, cbcl_scr_dsm5_anxdisord_t,	cbcl_scr_syn_attention_t, cbcl_scr_syn_aggressive_t, pgbi_p_ss_score, upps_y_ss_negative_urgency, upps_y_ss_positive_urgency, sds_p_ss_total, fam_history_6_yes_no, fam_history_q6a_depression, fam_history_q6d_depression, q6k_full_sib_young1_depression, q6k_full_sib_young2_depression, q6k_full_sib_young3_depression, q6k_full_sib_young4_depression, q6k_full_sib_young5_depression, q6l_full_sib_old1_depression, q6l_full_sib_old2_depression, q6l_full_sib_old3_depression, q6l_full_sib_old4_depression, q6l_full_sib_old5_depression, famhx_ss_momdad_ma_p, reshist_addr1_coi_z_coi_nat, kbi_p_c_bully, ksads_bully_raw_26, nsc_p_ss_mean_3_items, nihtbx_list_agecorrected, nihtbx_flanker_agecorrected, nihtbx_pattern_agecorrected, ksads_ptsd_raw_762_p, ksads_ptsd_raw_763_p,  ksads_ptsd_raw_767_p, ksads_ptsd_raw_768_p, ksads_ptsd_raw_766_p, fam_enviro6_p, fam_enviro3_p, fes_youth_q6, fes_youth_q3, famhx_4_p, fam_history_13_yes_no, fam_history_5_yes_no, fam_history_6_yes_no, demo_prnt_marital_v2, fam_history_9_yes_no, crpbi_parent4_y, parent_monitor_q1_y, parent_monitor_q3_y))

#2.2 Filter the trimmed data to only contain rows at the baseline assessment timepoint
risk_data_merged_trimmed <- risk_data_merged_trimmed %>% 
  filter(eventname == "baseline_year_1_arm_1")

#3. Preliminarily create relevant risk clustering variables
#3.1 Family (first-degree relative) history of depression
risk_data_merged_trimmed <- risk_data_merged_trimmed %>%
  mutate(
    family_history_depression = case_when(
      
      #3.11 Assign 1 if any first-degree relative has depression
      replace_na(fam_history_q6a_depression == 1, FALSE) |
        replace_na(fam_history_q6d_depression == 1, FALSE) |
        replace_na(q6k_full_sib_young1_depression == 1, FALSE) |
        replace_na(q6k_full_sib_young2_depression == 1, FALSE) |
        replace_na(q6k_full_sib_young3_depression == 1, FALSE) |
        replace_na(q6k_full_sib_young4_depression == 1, FALSE) |
        replace_na(q6k_full_sib_young5_depression == 1, FALSE) |
        replace_na(q6l_full_sib_old1_depression == 1, FALSE) |
        replace_na(q6l_full_sib_old2_depression == 1, FALSE) |
        replace_na(q6l_full_sib_old3_depression == 1, FALSE) |
        replace_na(q6l_full_sib_old4_depression == 1, FALSE) |
        replace_na(q6l_full_sib_old5_depression == 1, FALSE) ~ 1,
      
      #3.12 Assign 0 if "fam_history_6_yes_no" is 1 but no first-degree relative has depression
      fam_history_6_yes_no == 1 ~ 0,
      
      #3.13 Assign 0 if "fam_history_6_yes_no" is 0
      fam_history_6_yes_no == 0 ~ 0,
      
      #3.14 Otherwise, assign NA
      TRUE ~ NA_real_
    )
  )

#3.2 Sum ACE Score Index
risk_data_merged_trimmed <- risk_data_merged_trimmed %>%
  mutate(
    
    #3.211 Calculate each ACE component
    ACE_1 = if_else(fam_history_9_yes_no == 1, 1, 0, missing = 0),
    ACE_2 = if_else(crpbi_parent4_y %in% c(1, 2), 1, 0, missing = 0),
    ACE_3 = if_else(fam_history_13_yes_no == 1 | fam_history_6_yes_no == 1, 1, 0, missing = 0),
    ACE_4 = if_else(famhx_4_p == 1 | fam_history_5_yes_no == 1, 1, 0, missing = 0),
    ACE_5 = if_else(ksads_ptsd_raw_766_p == 1 | fes_youth_q6 == 1 | fes_youth_q3 == 1, 1, 0, missing = 0),
    ACE_6 = if_else(demo_prnt_marital_v2 %in% c(2, 3, 4), 1, 0, missing = 0),
    ACE_7 = if_else(ksads_ptsd_raw_762_p == 1 | ksads_ptsd_raw_763_p == 1, 1, 0, missing = 0),
    ACE_8 = if_else(parent_monitor_q1_y %in% c(1, 2) | parent_monitor_q3_y %in% c(1, 2), 1, 0, missing = 0),
    ACE_9 = if_else(ksads_ptsd_raw_767_p == 1 | ksads_ptsd_raw_768_p == 1, 1, 0, missing = 0),
    
    #3.212 Sum the ACE components for each participant
    ACE_index_sum_score = rowSums(across(starts_with("ACE_")), na.rm = TRUE),
    
    #3.22 Count the number of missing values for the ACE components
    ACE_index_sum_score_n_missing = rowSums(
      across(c(
        fam_history_9_yes_no, crpbi_parent4_y, fam_history_13_yes_no, fam_history_6_yes_no, 
        famhx_4_p, fam_history_5_yes_no, ksads_ptsd_raw_766_p, fes_youth_q6, fes_youth_q3,
        demo_prnt_marital_v2, ksads_ptsd_raw_762_p, ksads_ptsd_raw_763_p, 
        parent_monitor_q1_y, parent_monitor_q3_y, ksads_ptsd_raw_767_p, ksads_ptsd_raw_768_p
      ), ~ . %in% c(NA, "", 999, 7, 777))
    )
  )

```

## 1. Define the N, percent, and range of risk variables of interest at the baseline assessment timepoint

```{r EDA 1, echo = FALSE, warning = FALSE}

## Exploratory Data Analysis ##

#1. Define the N, percent, and range of risk variables of interest at the baseline assessment timepoint
#1.1 Define the columns to analyze
columns_to_check <- c(
  "cbcl_scr_dsm5_depress_t",
  "cbcl_scr_dsm5_anxdisord_t",
  "cbcl_scr_syn_attention_t",
  "cbcl_scr_syn_aggressive_t",
  "pgbi_p_ss_score",
  "upps_y_ss_negative_urgency",
  "upps_y_ss_positive_urgency",
  "sds_p_ss_total",
  "family_history_depression",
  "fam_history_q6a_depression",
  "fam_history_q6d_depression",
  "q6k_full_sib_young1_depression",
  "q6k_full_sib_young2_depression",
  "q6k_full_sib_young3_depression",
  "q6k_full_sib_young4_depression",
  "q6k_full_sib_young5_depression",
  "q6l_full_sib_old1_depression",
  "q6l_full_sib_old2_depression",
  "q6l_full_sib_old3_depression",
  "q6l_full_sib_old4_depression",
  "q6l_full_sib_old5_depression",
  "famhx_ss_momdad_ma_p",
  "reshist_addr1_coi_z_coi_nat",
  "kbi_p_c_bully",
  "ksads_bully_raw_26",
  "nsc_p_ss_mean_3_items",
  "nihtbx_list_agecorrected",
  "nihtbx_flanker_agecorrected",
  "nihtbx_pattern_agecorrected",
  "ksads_ptsd_raw_762_p",
  "ksads_ptsd_raw_763_p",
  "ksads_ptsd_raw_767_p",
  "ksads_ptsd_raw_768_p",
  "ksads_ptsd_raw_766_p",
  "fam_enviro6_p",
  "fam_enviro3_p",
  "fes_youth_q6",
  "fes_youth_q3",
  "famhx_4_p",
  "fam_history_13_yes_no",
  "fam_history_5_yes_no",
  "fam_history_6_yes_no",
  "demo_prnt_marital_v2",
  "fam_history_9_yes_no",
  "crpbi_parent4_y",
  "parent_monitor_q1_y",
  "parent_monitor_q3_y",
  "ACE_index_sum_score")

# 1.2 Create a function to compute the required metrics for each column
get_column_metrics <- function(column_name, data, total_subjects = nrow(data)) {
  
  # 1.21 Extract each column
  column_data <- data[[column_name]]
  
  # 1.22 Define variable-specific missing value codes
  missing_values <- c(999, 7)  # Default missing codes
  if (column_name == "demo_prnt_marital_v2") {
    missing_values <- c(777)
  } else if (column_name %in% c("fam_history_q6a_depression", "fam_history_q6d_depression",
                                "q6k_full_sib_young1_depression", "q6k_full_sib_young2_depression",
                                "q6k_full_sib_young3_depression", "q6k_full_sib_young4_depression",
                                "q6k_full_sib_young5_depression", "q6l_full_sib_old1_depression",
                                "q6l_full_sib_old2_depression", "q6l_full_sib_old3_depression",
                                "q6l_full_sib_old4_depression", "q6l_full_sib_old5_depression")) {
    missing_values <- c(7, 999)
  } else if (column_name %in% c("famhx_4_p", "fam_history_13_yes_no", "fam_history_5_yes_no",
                                "fam_history_6_yes_no", "fam_history_9_yes_no")) {
    missing_values <- c(7, 999)
  }
  
  # 1.23 Filter non-NA, non-empty, and non-invalid values
  valid_values <- column_data[!is.na(column_data) &
                                column_data != "" &
                                !column_data %in% missing_values]
  
  # 1.24 Compute metrics
  n_non_na <- length(valid_values)
  percent_non_na <- (n_non_na / total_subjects) * 100
  
  if (is.numeric(valid_values)) {
    range_or_unique <- range(valid_values, na.rm = TRUE)
  } else {
    range_or_unique <- unique(valid_values)
  }
  
  # 1.25 Return the metrics as a list
  list(
    column = column_name,
    n_non_na = n_non_na,
    percent_non_na = percent_non_na,
    range_or_unique = range_or_unique
  )
}

# 1.3 Apply the function to all columns and store risk data missingness in a list
risk_data_missingness_list <- lapply(columns_to_check, get_column_metrics, data = risk_data_merged_trimmed)

# 1.4 Convert the risk data missingness into a dataframe for easier viewing
risk_data_missingness_df <- do.call(rbind, lapply(risk_data_missingness_list, function(x) {
  data.frame(
    column = x$column,
    n_non_na = x$n_non_na,
    percent_non_na = x$percent_non_na,
    range_or_unique = I(list(x$range_or_unique))  # Use I() to preserve list structure for unique/range
  )
}))

#1.5 Print the final dataframe of risk_data_missingness_
kable(risk_data_missingness_df, col.names = c("Risk Variable", "N True Values", "Percent True Values", "Range or Unique Values in Data"), digits = 2)

```

## 2. Visually and Quantitatively Assess for any Outliers in the Data

```{r EDA 2, echo = FALSE, warning = FALSE}

## Exploratory Data Analysis ##

#2. Visually and Quantitatively Assess for any Outliers in the Data
#2.1 Create boxplots with outlier indicators for continuous variables in the dataset
#2.11 Filter only continuous variables (excluding ID and gender)
risk_data_continuous_variables <- risk_data_merged_trimmed %>%
  dplyr::select(cbcl_scr_dsm5_depress_t, cbcl_scr_dsm5_anxdisord_t, cbcl_scr_syn_attention_t, cbcl_scr_syn_aggressive_t, pgbi_p_ss_score, upps_y_ss_negative_urgency, upps_y_ss_positive_urgency, sds_p_ss_total, reshist_addr1_coi_z_coi_nat, nsc_p_ss_mean_3_items, nihtbx_list_agecorrected, nihtbx_flanker_agecorrected, nihtbx_pattern_agecorrected, ACE_index_sum_score) %>%
  gather(key = "Variable", value = "Value")

#2.12 Create boxplot with facets for continuous variables
risk_data_continuous_variables_boxplot <- ggplot(risk_data_continuous_variables, aes(x = Variable, y = Value)) + 
  geom_boxplot(fill = "lightblue", color = "darkblue") + 
  facet_wrap(~ Variable, scales = "free") + 
  theme_minimal() + 
  theme(axis.text.x = element_blank(),
        strip.text = element_text(size = 6, face = "bold")) +
  labs(title = "Boxplots of Continuous Variables",
       y = "Values",
       x = "")

#2.13 Print the plot
print(risk_data_continuous_variables_boxplot)

```

From the boxplots for continuous variables, it is evident that some of the NIHTB scores are likely out of the expected range. While no guidelines within the [overview of the ABCD baseline neurocognition battery](https://www.sciencedirect.com/science/article/pii/S1878929317302384?via%3Dihub#tbl0005) or mention of the extraneously high/low scores within [current release notes](https://wiki.abcdstudy.org/release-notes/non-imaging/neurocognition.html#nih-toolbox-cognition) could be located, it is likely that winsorization or other outlier removal methods will be applied to handle values beyond the expected range. This topic to be re-visited ASAP during initial stages of clustering. 

No other anomalous or outlier values are noteworthy. 

## 3. Appropriateness of the Mean as a Measure of Central Tendency

Though proposed clustering methods herein do not have parametric assumptions, assessment of skewness provides information about the asymmetry of variable distributions, which may help identify variables with extreme outliers or long tails. Such variables can disproportionately influence cluster formation when using euclidean distances (as in k-prototypes for continuous variables), and thus it is important to examine them. The skewness values for continuous variables herein are as such: 

```{r EDA 3, echo = FALSE, warning = FALSE}

## Exploratory Data Analysis ##

#3.1 Calculate skewness for continuous variables
risk_data_continuous_variables_skewness_table <- risk_data_merged_trimmed %>%
  dplyr::select(
    cbcl_scr_dsm5_depress_t,
    cbcl_scr_dsm5_anxdisord_t,
    cbcl_scr_syn_attention_t,
    cbcl_scr_syn_aggressive_t,
    pgbi_p_ss_score,
    upps_y_ss_negative_urgency,
    upps_y_ss_positive_urgency,
    sds_p_ss_total,
    reshist_addr1_coi_z_coi_nat,
    nsc_p_ss_mean_3_items,
    nihtbx_list_agecorrected,
    nihtbx_flanker_agecorrected,
    nihtbx_pattern_agecorrected,
    ACE_index_sum_score) %>%
  summarise(across(everything(), ~ skewness(.x, na.rm = TRUE))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Skewness")

#3.2 Display the skewness table
kable(risk_data_continuous_variables_skewness_table, caption = "Skewness Values for Continuous Variables", digits = 2)

```

Results from the assessment of skewness values suggest that the only variable which may be problematically skewed is *pgbi_p_ss_score* (GBI Mania Sum Score, skewness = `r round(risk_data_continuous_variables_skewness_table %>% filter(Variable == "pgbi_p_ss_score") %>% pull(Skewness), digits = 2)`); though, it is expected that the variable would be positively skewed given [low rates of mania in this age range](https://pmc.ncbi.nlm.nih.gov/articles/PMC8233426/). 

As such, it is likely that no variable transformations are necessary, and downstream trends of this variable during clustering and subsequent frequentist analyses will just need to be monitored with this skew in mind. 

## 4. Shape of Each Variables' Distribution

The shape of the distribution for each continuous variable and frequency distribution of each categorical variable is outlined below, both in terms of summary statistics derived from and a graphical histogram with density overlay of relevant data:

```{r EDA 4, echo = FALSE, warning = FALSE, fig.height=10, fig.width=10}

## Exploratory Data Analysis ##

#4. Create summary tables and visualizations for the distributions of categorical and continuous variables included in the risk model
#4.1 Create a summary table of distribution information
#4.11 Specify continuous columns explicitly
continuous_columns <- c(
  "cbcl_scr_dsm5_depress_t", "cbcl_scr_dsm5_anxdisord_t", 
  "cbcl_scr_syn_attention_t", "cbcl_scr_syn_aggressive_t", 
  "pgbi_p_ss_score", "upps_y_ss_negative_urgency", 
  "upps_y_ss_positive_urgency", "sds_p_ss_total", 
  "reshist_addr1_coi_z_coi_nat", "nsc_p_ss_mean_3_items", 
  "nihtbx_list_agecorrected", "nihtbx_flanker_agecorrected", 
  "nihtbx_pattern_agecorrected", "ACE_index_sum_score")

#4.12 Define all columns to check
columns_to_check <- colnames(risk_data_merged_trimmed)

#4.131 Derive categorical columns as those not in the continuous list
categorical_columns <- setdiff(columns_to_check, continuous_columns)
categorical_columns <- categorical_columns[!categorical_columns %in% c("src_subject_id", "eventname", "fam_history_q6a_depression", "fam_history_q6d_depression", "q6k_full_sib_young1_depression", "q6k_full_sib_young2_depression", "q6k_full_sib_young3_depression", "q6k_full_sib_young4_depression", "q6k_full_sib_young5_depression", "q6l_full_sib_old1_depression", "q6l_full_sib_old2_depression", "q6l_full_sib_old3_depression", "q6l_full_sib_old4_depression", "q6l_full_sib_old5_depression", "ksads_ptsd_raw_762_p", "ksads_ptsd_raw_763_p",  "ksads_ptsd_raw_767_p", "ksads_ptsd_raw_768_p", "ksads_ptsd_raw_766_p", "fam_enviro6_p", "fam_enviro3_p", "fes_youth_q6", "fes_youth_q3", "famhx_4_p", "fam_history_13_yes_no", "fam_history_5_yes_no", "fam_history_6_yes_no", "demo_prnt_marital_v2", "fam_history_9_yes_no", "crpbi_parent4_y", "parent_monitor_q1_y", "parent_monitor_q3_y", "ACE_index_sum_score_n_missing")]

#4.132 Define new titles for individual ACE variables
categorical_title_map <- c(
  "ACE_1" = "Criminal Household Member",
  "ACE_2" = "Emotional neglect",
  "ACE_3" = "Household Depression or SI",
  "ACE_4" = "Household substance abuse",
  "ACE_5" = "Mother treated violently",
  "ACE_6" = "Parental separation or divorce",
  "ACE_7" = "Physical abuse",
  "ACE_8" = "Physical neglect",
  "ACE_9" = "Sexual Abuse")

#4.14 Create a summary table for continuous and categorical variables
# Adjust summary function for categorical columns
summarize_data <- function(data, continuous_cols, categorical_cols) {
  summary_list <- list()
  
  # Continuous variables summary
  for (col in continuous_cols) {
    column_data <- data[[col]]
    valid_values <- column_data[!is.na(column_data)]
    summary_list[[col]] <- data.frame(
      Variable = col,
      Type = "Continuous",
      Mean = mean(valid_values, na.rm = TRUE),
      Median = median(valid_values, na.rm = TRUE),
      SD = sd(valid_values, na.rm = TRUE),
      Min = min(valid_values, na.rm = TRUE),
      Max = max(valid_values, na.rm = TRUE),
      Levels = NA,
      Frequencies = NA,
      stringsAsFactors = FALSE
    )
  }
  
  # Categorical variables summary
  for (col in categorical_cols) {
    column_data <- data[[col]]
    freq_table <- table(column_data, useNA = "no")
    summary_list[[col]] <- data.frame(
      Variable = col,
      Type = "Categorical",
      Mean = NA,
      Median = NA,
      SD = NA,
      Min = NA,
      Max = NA,
      Levels = paste(names(freq_table), collapse = ", "),
      Frequencies = paste(freq_table, collapse = ", "),
      stringsAsFactors = FALSE
    )
  }
  
  do.call(rbind, summary_list)
}

#4.151 Generate the summary table
summary_table <- summarize_data(risk_data_merged_trimmed, continuous_columns, categorical_columns)

#4.152 Rename the ACE variables in the summary table
summary_table <- summary_table %>%
  mutate(Variable = ifelse(Variable %in% names(categorical_title_map), categorical_title_map[Variable], Variable))

#4.153 Print the summary table in the RMD
kable(summary_table, caption = "Summary Statistics for Risk Model Variables", row.names = FALSE)

#4.2 Create plots visualizing distributions of risk variables
#4.21 Generate a combined dataset with an indicator for variable type
plot_data <- list()

#4.22 Add continuous variables
for (col in continuous_columns) {
  plot_data[[col]] <- risk_data_merged_trimmed %>%
    select(Value = !!sym(col)) %>%
    mutate(Type = "Continuous", Variable = col)
}

#4.23 Add categorical variables
for (col in categorical_columns) {
  plot_data[[col]] <- risk_data_merged_trimmed %>%
    select(Value = !!sym(col)) %>%
    mutate(Type = "Categorical", Variable = col)
}

#4.24 Combine all into a single data frame
plot_data_combined <- bind_rows(plot_data)

#4.251 Generate facet plot for continuous variables
continuous_plot <- ggplot(
  plot_data_combined %>% filter(Type == "Continuous"),
  aes(x = Value)) +
  geom_histogram(aes(y = ..density..), binwidth = 1, fill = "blue", alpha = 0.6, color = "black") +
  geom_density(color = "red", size = 1, adjust = 1.5) +
  facet_wrap(~Variable, ncol = 3, scales = "free") +
  theme_minimal() +
  labs(title = "Distributions of Continuous Variables",
       x = "Value",
       y = "Density") +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(size = 8),
    panel.spacing = unit(1, "lines"))

#4.252 Print the continuous variable plot
print(continuous_plot)

#4.261 Generate facet plot for categorical variables
#4.2611 Alter reverse coded values for the "kbi_p_c_bully" variable for plotting purposes
plot_data_combined <- plot_data_combined %>%
  mutate(Value = ifelse(Variable == "kbi_p_c_bully", recode(Value, `1` = 1, `2` = 0), Value))

#4.2612 Generate the categorical variable plot
categorical_plot <- ggplot(
  plot_data_combined %>% filter(Type == "Categorical"),
  aes(x = as.factor(Value))) +  # Treat Value as a factor
  geom_bar(fill = "skyblue", color = "black", alpha = 0.8) +
  facet_wrap(~Variable, ncol = 3, scales = "free", 
             labeller = labeller(Variable = categorical_title_map)) +
  theme_minimal() +
  labs(title = "Distributions of Categorical Variables",
       x = "Categories",
       y = "Count") +
  scale_x_discrete(limits = c("0", "1"), labels = c("No", "Yes")) + # Set specific x-axis breaks and labels
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(size = 8),
    panel.spacing = unit(1, "lines"))

#4.2613 Print the categorical variable plot
print(categorical_plot)

```

Distributions of continuous variables and rates of categorical variables (including the individual ACEs comprising the sum ACE index score) are shown above. While some previously mentioned variables exhibit noticeable skew, no problematic distributions or outlier values are notable here. 

## 5. Overall Availability of Risk Model Variables for Clustering

```{r EDA 5, echo = FALSE, warning = FALSE}

## Exploratory Data Analysis ##

#5. Determine the broad availability of variables to be included in the risk model
#5.1 Determine the number of unique subjects in the dataset
unique_subject_count <- as.numeric(length(unique(risk_data_merged_trimmed$src_subject_id)))

#5.2 Create a vector of counts for complete cases across specified columns
complete_cases_count <- risk_data_merged_trimmed %>%
  dplyr::select(matches("cbcl_scr|pgbi_p_ss_score|upps_y_ss|sds_p_ss_total|family_history_depression|famhx_ss_momdad_ma_p|reshist_addr1_coi_z_coi_nat|nsc_p_ss_mean_3_items|kbi_p_c_bully|ksads_bully_raw_26|nihtbx_|ACE_index_sum_score")) %>%
  filter_all(all_vars(!is.na(.) & . != "" & . != 999 & . != 777)) %>%
  nrow()

```

When looking across all relevant risk variables in the dataset, `r paste0(complete_cases_count)` (`r paste0(round((complete_cases_count / unique_subject_count) * 100, 2), "%")`) subjects have complete (i.e., fully reported, non-missing) data. 