---
title: "k = 2 Solution Diagnostics Across Revision Configurations"
author: "Sam A. Sievertsen"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
params:
  scaling_method: "z_score"
  legacy_anchor_file: "kproto_z_score.rds"
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, message = TRUE, warning = TRUE, results = "markup", comment = "")

```

```{r environment, echo = FALSE, include = FALSE, warning = FALSE}

# Load necessary packages + environment
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(clustMixType)
library(mclust)
library(effectsize)
library(glue)
library(here)

# Set numerical options
options(scipen = 999, digits = 8)

# List project repository location and load the shared configuration definitions and silhouette helpers
REPO <- here()
source(file.path(REPO, "scripts/main_analysis/1_clustering/cluster_configs.R"))
source(file.path(REPO, "scripts/main_analysis/1_clustering/silhouette_helpers.R"))

# Paths
kproto_dir <- file.path(REPO, "data/data_processed", "kproto_results")
diag_dir <- file.path(REPO, "data/data_processed", "validation_results", "k2_diagnostics")
dir.create(diag_dir, recursive = TRUE, showWarnings = FALSE)

# Read in the unscaled risk variable data, used for orientation and feature profiles
risk_variable_data <- read.csv(file.path(REPO, "data/data_processed", "risk_variable_data.csv"))

```

## Purpose

All three revision configurations selected *k* = 2, but they differ materially in how the sample is partitioned, and in each the smaller cluster has a negative mean silhouette width. This report asks four questions of the cached *k* = 2 fits, none of which require refitting:

  1. **Does the emotional abuse proxy drive the primary partition?** Primary and fallback differ only in whether emotional abuse is a standalone binary or folded into the ACE composite

  2. **How much do the configurations agree** with one another and with the published 17-feature solution?

  3. **Are the negative silhouette widths misassignments, or a consequence of unequal cluster dispersion?** The mean Huang distance from participant *i* to cluster *C* decomposes exactly into (distance to C's centroid) + (C's internal numeric spread) + (a categorical frequency term), whereas k-prototypes assigns by distance to the prototype alone. A participant can therefore be correctly assigned to a diffuse cluster yet have a(i) > b(i)

  4. **Was this already true of the published solution?**

```{r load, warning = FALSE}

## Load the k = 2 fits ##

#1. Load each revision configuration with a cache, validating its fingerprint
fits <- list()
for (cfg in names(cluster_configs)) {
  f <- file.path(kproto_dir, sprintf("kproto_%s_%s.rds", cfg, params$scaling_method))
  if (!file.exists(f)) { cat("No cache for", cfg, "; skipping\n"); next }
  kl <- readRDS(f)
  fp <- attr(kl, "fingerprint")
  if (is.null(fp) || !identical(fp$features, cluster_configs[[cfg]])) stop("Fingerprint mismatch for ", cfg)
  fits[[cfg]] <- kl$k2
}

#2. Load the published 17-feature anchor solution, if present, for comparison
legacy_path <- file.path(kproto_dir, params$legacy_anchor_file)
if (file.exists(legacy_path)) {
  anchor_k2 <- readRDS(legacy_path)$k2
  n_shared <- sum(rownames(anchor_k2$data) %in% risk_variable_data$participant_id)
  if (n_shared == 0) {
    cat("Published anchor rows are not keyed by participant_id; anchor comparisons skipped\n")
  } else {
    fits[["published_anchor"]] <- anchor_k2
    cat("Loaded published anchor from", basename(legacy_path), "(", n_shared, "participants shared with the revision sample )\n")
  }
} else {
  cat("Published anchor not found at", legacy_path, "; anchor comparisons skipped\n")
}

#3. Orient every solution so the cluster with the higher mean CBCL dysregulation profile raw sum is labelled higher-risk
orient <- function(kp) {
  ids <- rownames(kp$data)
  dp <- risk_variable_data$cbcl_dp_raw_sum[match(ids, risk_variable_data$participant_id)]
  means <- tapply(dp, kp$cluster, mean, na.rm = TRUE)
  higher <- as.integer(names(which.max(means)))
  tibble(participant_id = ids, raw_cluster = kp$cluster,
         risk_group = factor(ifelse(kp$cluster == higher, "Higher-risk", "Lower-risk"), levels = c("Higher-risk", "Lower-risk")))
}
labels <- purrr::map(fits, orient)

```

## 1. Solution Overview

```{r overview, warning = FALSE}

## Size, silhouette, and per-group silhouette for every solution ##

#1. Compute individual widths (exact closed form, verified identical to kproto2silhouette to ~1e-15 on these data)
sils <- purrr::map(fits, silhouette_exact_huang)

#2. Summarize by oriented risk group
overview <- purrr::imap_dfr(fits, function(kp, cfg) {
  s <- sils[[cfg]][, "sil_width"]; g <- labels[[cfg]]$risk_group
  tibble(
    Config = cfg,
    `N` = length(g),
    `N feat` = ncol(kp$data),
    lambda = mean(kp$lambda),
    `Higher-risk n` = sum(g == "Higher-risk"),
    `Higher-risk %` = 100 * mean(g == "Higher-risk"),
    `Mean silhouette` = mean(s),
    `Higher-risk mean width` = mean(s[g == "Higher-risk"]),
    `Higher-risk % negative` = 100 * mean(s[g == "Higher-risk"] < 0),
    `Lower-risk mean width` = mean(s[g == "Lower-risk"]),
    `Lower-risk % negative` = 100 * mean(s[g == "Lower-risk"] < 0)
  )
})

#3. Print
overview %>%
  kable(caption = "k = 2 solutions, oriented so the higher-risk group has the higher mean CBCL dysregulation profile raw sum", digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 12)

```

## 2. Agreement Across Solutions

Adjusted Rand index on shared participants, and the share of each row solution's higher-risk group that is also higher-risk in the column solution.

```{r agreement, warning = FALSE}

## Pairwise ARI and higher-risk overlap ##

#1. Pairwise ARI on the participants shared by each pair
cfgs <- names(labels)
ari <- matrix(NA_real_, length(cfgs), length(cfgs), dimnames = list(cfgs, cfgs))
overlap <- ari
for (a in cfgs) for (b in cfgs) {
  m <- inner_join(labels[[a]], labels[[b]], by = "participant_id", suffix = c(".a", ".b"))
  ari[a, b] <- adjustedRandIndex(m$risk_group.a, m$risk_group.b)
  overlap[a, b] <- 100 * mean(m$risk_group.b[m$risk_group.a == "Higher-risk"] == "Higher-risk")
}

#2. Print ARI
round(ari, 3) %>%
  kable(caption = "Adjusted Rand index between k = 2 solutions (shared participants)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#3. Print higher-risk overlap
round(overlap, 1) %>%
  kable(caption = "% of the ROW solution's higher-risk group that is also higher-risk in the COLUMN solution") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```

## 3. What Separates the Groups

Every feature in the union of all configurations, plus the four CBCL T-scores from the submitted solution, profiled under each solution's own labels. Binary features are converted to a Cohen's *d* equivalent (`effectsize::oddsratio_to_d()`) so all features rank on one scale. Positive values are elevated in the higher-risk group.

```{r profiles, warning = FALSE}

## Feature separation under each solution ##

#1. Features to profile
profile_feats <- unique(c(feats_all_configs, "mh_p_cbcl__dsm__dep_tscore", "mh_p_cbcl__dsm__anx_tscore",
                          "mh_p_cbcl__synd__attn_tscore", "mh_p_cbcl__synd__aggr_tscore"))

#2. Create a helper returning a d-equivalent separation for one feature under one solution's labels
separation <- function(var, lab) {
  v <- risk_variable_data[[var]][match(lab$participant_id, risk_variable_data$participant_id)]
  ok <- !is.na(v); v <- v[ok]; g <- lab$risk_group[ok]
  if (var %in% feats_categorical) {
    tab <- table(g, factor(v, levels = c(1, 0)))
    if (any(tab == 0)) return(NA_real_)
    as.numeric(effectsize::oddsratio_to_d(effectsize::oddsratio(tab)$Odds_ratio))
  } else {
    as.numeric(effectsize::cohens_d(v ~ g)$Cohens_d)
  }
}

#3. Build the profile matrix and rank features within each solution
profile <- purrr::map_dfc(labels, function(lab) purrr::map_dbl(profile_feats, separation, lab = lab))
profile <- bind_cols(tibble(feature = profile_feats), profile)
ranks <- profile %>% mutate(across(-feature, ~ rank(-abs(.x), na.last = "keep"), .names = "rank_{.col}"))

#4. Print the profile, sorted by the primary solution if present
sort_col <- if (primary_config %in% names(profile)) primary_config else names(profile)[2]
profile %>%
  arrange(desc(abs(.data[[sort_col]]))) %>%
  kable(caption = glue("Separation (d-equivalent, higher-risk minus lower-risk) of each feature under each k = 2 solution; sorted by {sort_col}"), digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 11)

#5. Emotional abuse specifically: prevalence by group and its rank within each solution
emo <- purrr::imap_dfr(labels, function(lab, cfg) {
  v <- risk_variable_data$abuse_emotional[match(lab$participant_id, risk_variable_data$participant_id)]
  tibble(Config = cfg,
         `Higher-risk % endorsed` = 100 * mean(v[lab$risk_group == "Higher-risk"] == 1, na.rm = TRUE),
         `Lower-risk % endorsed` = 100 * mean(v[lab$risk_group == "Lower-risk"] == 1, na.rm = TRUE),
         `Rank among features` = ranks[[paste0("rank_", cfg)]][ranks$feature == "abuse_emotional"],
         `Is a clustering feature` = if (cfg %in% names(cluster_configs)) "abuse_emotional" %in% cluster_configs[[cfg]] else FALSE)
})
emo %>%
  kable(caption = "Emotional abuse proxy (FES 'family members often criticize each other', either reporter) by risk group", digits = 1) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#6. Write the profile to CSV
write.csv(profile, file.path(diag_dir, "k2_feature_profiles.csv"), row.names = FALSE)

```

## 4. Why Are Silhouette Widths Negative in the Smaller Group?

For each participant, a(i) and b(i) are the mean Huang distances to their own and the nearest other cluster. Each decomposes exactly as *centroid distance + that cluster's numeric spread + a categorical frequency term*. Two tests:

- **Assignment check:** among participants with negative widths, the share who are nonetheless closer to their own k-prototypes prototype than to the other (i.e., correctly assigned by the algorithm's own criterion)
- **Spread-removed check:** the share of widths that remain negative after subtracting each cluster's numeric spread constant from a(i) and b(i). If negatives largely vanish, they reflect the higher-risk group's greater dispersion rather than misassignment

```{r decomposition, warning = FALSE}

## Decompose silhouette widths into centroid distance and cluster spread ##

#1. Create a helper computing the decomposition for one solution
decompose <- function(kp, lab, s_ref) {
  stopifnot(length(table(kp$cluster)) == 2)
  x <- kp$data; cl <- kp$cluster; lambda <- kp$lambda
  numvars <- vapply(x, is.numeric, logical(1)); catvars <- vapply(x, is.factor, logical(1))
  w_num <- if (length(lambda) == 1) rep(1, sum(numvars)) else lambda[numvars]
  w_cat <- if (length(lambda) == 1) rep(lambda, sum(catvars)) else lambda[catvars]
  X <- as.matrix(x[, numvars, drop = FALSE]); n <- nrow(x); K <- length(table(cl))
  full <- matrix(0, n, K); spread_num <- numeric(K); spread_cat <- numeric(K)
  for (i in seq_len(K)) {
    m <- which(cl == i); Xc <- X[m, , drop = FALSE]; mu <- colMeans(Xc)
    spread_num[i] <- sum(w_num * colMeans(sweep(Xc, 2, mu)^2))
    cat_term <- numeric(n)
    for (jj in seq_along(which(catvars))) {
      v <- x[[which(catvars)[jj]]]; f <- tabulate(as.integer(v[m]), nlevels(v)) / length(m)
      cat_term <- cat_term + w_cat[jj] * (1 - f[as.integer(v)])
    }
    spread_cat[i] <- mean(cat_term[m])
    full[, i] <- colSums(w_num * (t(X) - mu)^2) + spread_num[i] + cat_term
  }
  own <- cbind(seq_len(n), cl); oth <- cbind(seq_len(n), 3 - cl)
  a <- full[own]; b <- full[oth]
  s <- (b - a) / pmax(a, b)
  stopifnot(max(abs(s - s_ref)) < 1e-10)  # the decomposition must reproduce the verified widths exactly
  s_nospread <- ((b - spread_num[3 - cl]) - (a - spread_num[cl])) / pmax(a - spread_num[cl], b - spread_num[3 - cl])
  correct <- kp$dists[own] < kp$dists[oth]
  g <- lab$risk_group
  purrr::map_dfr(c("Higher-risk", "Lower-risk"), function(grp) {
    idx <- g == grp; ci <- unique(cl[idx])
    tibble(
      `Risk group` = grp, n = sum(idx),
      `Numeric spread` = spread_num[ci], `Categorical spread` = spread_cat[ci],
      `% negative` = 100 * mean(s[idx] < 0),
      `% of negatives correctly assigned` = if (any(idx & s < 0)) 100 * mean(correct[idx & s < 0]) else NA_real_,
      `% negative, spread removed` = 100 * mean(s_nospread[idx] < 0))
  })
}

#2. Run and print for every solution, checking each decomposition against the verified widths
decomp <- purrr::imap_dfr(fits, ~ mutate(decompose(.x, labels[[.y]], sils[[.y]][, "sil_width"]), Config = .y, .before = 1))
decomp %>%
  kable(caption = "Silhouette decomposition by risk group (spread = mean within-cluster distance to centroid)", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 12)

write.csv(decomp, file.path(diag_dir, "k2_silhouette_decomposition.csv"), row.names = FALSE)

```

## 5. Summary

```{r summary, echo = FALSE, results = "asis"}

## Auto-generated summary of the diagnostic questions ##
#1. Adopted primary, read from the roles defined in cluster_configs.R
if (primary_config %in% names(labels)) {
  po <- overview %>% filter(Config == primary_config)
  top3 <- profile %>% transmute(feature, v = .data[[primary_config]]) %>% arrange(desc(abs(v))) %>% slice(1:3)
  cat(glue(
    "- **Adopted primary ({primary_config}):** higher-risk n = {po$`Higher-risk n`} ({round(po$`Higher-risk %`, 1)}%), ",
    "mean silhouette = {round(po$`Mean silhouette`, 3)}; strongest separators: ",
    "{paste(sprintf('%s (%.2f)', top3$feature, top3$v), collapse = ', ')}\n\n"))
}

#2. The emotional abuse contingency: the superseded standalone specification versus the adopted primary
if (all(c("primary", primary_config) %in% names(labels))) {
  e <- emo %>% filter(Config == "primary")
  cat(glue(
    "- **Superseded specification (emotional abuse as a standalone feature):** endorsed by {round(e$`Higher-risk % endorsed`, 1)}% of the ",
    "higher-risk vs {round(e$`Lower-risk % endorsed`, 1)}% of the lower-risk group, ranking #{e$`Rank among features`} of {nrow(profile)} features; ",
    "ARI vs the adopted primary = {round(ari['primary', primary_config], 3)}\n\n"))
}
pct_correct <- decomp$`% of negatives correctly assigned`
if (all(is.na(pct_correct))) {
  cat("- **Negative widths:** no solution has negative silhouette widths\n\n")
} else {
  cat(glue(
    "- **Negative widths:** across solutions, {round(min(pct_correct, na.rm = TRUE), 1)}% to ",
    "{round(max(pct_correct, na.rm = TRUE), 1)}% of negative-width participants are correctly ",
    "assigned by the k-prototypes criterion\n\n"))
}
if ("published_anchor" %in% names(labels)) {
  pa <- overview %>% filter(Config == "published_anchor")
  cat(glue("- **Published anchor:** higher-risk mean width = {round(pa$`Higher-risk mean width`, 3)}, ",
           "{round(pa$`Higher-risk % negative`, 1)}% negative\n"))
}

```
