---
title: "Bipolar Disorder & Suicidality Risk Group Cluster Stability Analysis: `r params$config` (`r params$mode`)"
author: "Sam A. Sievertsen"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
params:
  config: "fallback_emo_ace"
  mode: "chunk"
  chunk_id: 1
  n_chunks: 10
  scaling_method: "z_score"
  k_value: 2
  n_boot: 1000
  stab_nstart: 25
  seed: 123
  overwrite: false
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, message = TRUE, warning = TRUE, results = "markup", verbose = TRUE, comment = "")

```

```{r environment, echo = FALSE, include = FALSE, warning = FALSE}

## Load packages, data, & set env ##

# Load necesasry packages
library(dplyr)
library(purrr)
library(tidyr)
library(knitr)
library(kableExtra)
library(tibble)
library(ggplot2)
library(readr)
library(glue)
library(scales)
library(clustMixType)
library(here)

# Validate params before doing any work
stopifnot(params$mode %in% c("chunk", "report"), params$n_boot %% params$n_chunks == 0,
          params$chunk_id >= 1, params$chunk_id <= params$n_chunks)

# Set paths to be used in reading in data & objects
REPO <- here()
datapath <- file.path(REPO, "data", "data_processed")

# Load the shared configuration definitions and verify the requested configuration exists
source(file.path(REPO, "scripts/main_analysis/1_clustering/cluster_configs.R"))
if (!params$config %in% names(cluster_configs)) stop("Unknown config: ", params$config)

# Establish the path to the kproto object, namespaced by configuration so no other feature set's fit is reused
kproto_file <- file.path(datapath, "kproto_results", sprintf("kproto_%s_%s.rds", params$config, params$scaling_method))

# Establish the path to the stability results and write a new directory if it doesn't exist
stability_dir <- file.path(datapath, "stability_results")
chunk_dir <- file.path(stability_dir, "chunks_in_progress")
dir.create(chunk_dir, recursive = TRUE, showWarnings = FALSE)

# Set up logging (from SLURM)
log_file <- Sys.getenv("DETAILED_LOG", unset = NA)

# Set seed for reproducibility
set.seed(params$seed)

# Load k-prototype list, confirming its fingerprint belongs to this configuration
kp_list <- readRDS(kproto_file)
fp <- attr(kp_list, "fingerprint")
if (is.null(fp) || !identical(fp$config, params$config) || !identical(fp$features, cluster_configs[[params$config]])) {
  stop("kproto cache fingerprint does not match config ", params$config)
}

# Establish the current kproto object for the requested k by name (list elements are named k2 to k8)
kp_name <- paste0("k", params$k_value)
stopifnot(kp_name %in% names(kp_list))
kp <- kp_list[[kp_name]]

# Strip row names from the data embedded in the kproto object since stability_kproto() trips over them
#   (it matches bootstrap assignments by character row position, so participant-ID row names return NA)
if (!is.null(rownames(kp$data))) {
  rownames(kp$data) <- NULL
}

# Compose output file stems, namespaced by configuration
#   (the number of random starts per bootstrap refit is part of the stem, so results from different settings never mix)
stem <- glue("stability_{params$config}_{params$scaling_method}_k{params$k_value}_nb{params$n_boot}_ns{params$stab_nstart}")
rds_out <- file.path(stability_dir, glue("{stem}.rds"))
csv_out <- file.path(stability_dir, glue("{stem}.csv"))
chunk_file <- function(i) file.path(chunk_dir, glue("{stem}_chunk{i}of{params$n_chunks}.rds"))

# Establish guards to skip compute if cached & overwrite=FALSE
if (file.exists(rds_out) && isFALSE(params$overwrite)) {
  if (!is.na(log_file)) cat(
    sprintf("%s|STAB_SKIP|%s\n", format(Sys.time(), "%Y-%m-%d %H:%M:%OS3"),
            basename(rds_out)), file = log_file, append = TRUE)
  boot_res <- readRDS(rds_out)
} else {
  boot_res <- NULL
}

```

## Compute Stability of Specified Cluster Solution

This code evaluates the stability of the chosen clustering solution by: 

  1. Performing (B = `r params$n_boot`) bootstrap resampling with replacement analysis using Rand and Jaccard indices to measure cluster consistency across resampled datasets, split into `r params$n_chunks` chunks run as independent SLURM array tasks
  
  2. Pooling the chunks, which reproduces a single B-bootstrap run exactly: `stability_kproto()` reports the mean of the per-bootstrap Rand and Jaccard values, so stacking the per-bootstrap rows from every chunk and averaging yields the same estimator
  
  3. Extracting and organizing the bootstrap results into structured data frames containing individual bootstrap iterations and overall stability metrics
  
  4. Computing summary statistics including means, standard deviations, and 95% confidence intervals for both stability indices
  
  5. Outputting results displaying the stability assessment for the specified configuration, k-value, and scaling method
  
To note, higher Rand and Jaccard values (i.e., closer to 1.0) indicate more stable, reproducible clustering solutions

```{r bootstrap chunk, echo = FALSE, warning = FALSE, eval = (params$mode == "chunk")}

## Bootstrap stability for one chunk (Rand & Jaccard) ##

#1. Skip if this chunk already exists and overwrite is not requested
if (file.exists(chunk_file(params$chunk_id)) && isFALSE(params$overwrite)) {
  if (!is.na(log_file)) cat(
    sprintf("%s|STAB_CHUNK_SKIP|config=%s|chunk=%d\n", format(Sys.time(), "%Y-%m-%d %H:%M:%OS3"),
            params$config, params$chunk_id), file = log_file, append = TRUE)
} else {

  #1.1 Place this chunk on its own L'Ecuyer-CMRG stream so chunks never overlap and each is reproducible
  RNGkind("L'Ecuyer-CMRG")
  set.seed(params$seed)
  stream <- .Random.seed
  for (i in seq_len(params$chunk_id - 1)) stream <- parallel::nextRNGStream(stream)
  assign(".Random.seed", stream, envir = globalenv())

  #1.2 Announce start to log
  b_chunk <- params$n_boot / params$n_chunks
  if (!is.na(log_file)) cat(
    sprintf("%s|STAB_CHUNK_START|config=%s|k=%d|chunk=%d|B=%d\n",
            format(Sys.time(), "%Y-%m-%d %H:%M:%OS3"),
            params$config, params$k_value, params$chunk_id, b_chunk),
    file = log_file, append = TRUE
  )

  #1.3 Build call safely (function arg name for boots is 'B' in current clustMixType)
  stab_formals <- names(formals(clustMixType::stability_kproto))
  stab_args <- list(
    object = kp,
    method = c("rand","jaccard"),
    nstart = params$stab_nstart,
    verbose = TRUE
  )
  
  #1.4 Pass this chunk's share of the bootstraps under whichever argument name the installed version uses
  if ("B" %in% stab_formals) {
    stab_args$B <- b_chunk
  } else if ("nboot" %in% stab_formals) {
    stab_args$nboot <- b_chunk
  } else {
    stop("stability_kproto() has no recognized bootstrap-count argument; cannot control B per chunk")
  }

  #1.5 Run bootstrap stability - to note, at this stage, kproto object has no rownames
  chunk_res <- tryCatch(
    do.call(clustMixType::stability_kproto, stab_args),
    error = function(e) {
      stop(glue("stability_kproto() failed: {conditionMessage(e)}"))
    }
  )

  #1.6 Fail loudly if any bootstrap returned an undefined index (e.g., if row names were not stripped)
  if (anyNA(chunk_res[[2]][, c("R", "J")])) stop("NA stability values in chunk ", params$chunk_id)

  #1.7 Persist the raw chunk result
  saveRDS(chunk_res, chunk_file(params$chunk_id))

  #1.8 Announce end of the chunk and store in log
  if (!is.na(log_file)) cat(
    sprintf("%s|STAB_CHUNK_DONE|config=%s|chunk=%d|mean_J=%.4f|saved=%s\n", format(Sys.time(), "%Y-%m-%d %H:%M:%OS3"),
            params$config, params$chunk_id, mean(chunk_res[[2]][, "J"]), basename(chunk_file(params$chunk_id))),
    file = log_file, append = TRUE
  )
}

```

```{r cluster stability, echo = FALSE, warning = FALSE, eval = (params$mode == "report")}

## Bootstrap stability for chosen k (Rand & Jaccard), pooled across chunks ##

#1. Pool the chunks into one stability object, unless a pooled result is already cached
if (is.null(boot_res)) {

  #1.1 Require every chunk to be present
  missing_chunks <- which(!file.exists(vapply(seq_len(params$n_chunks), chunk_file, character(1))))
  if (length(missing_chunks) > 0) stop("Missing stability chunks: ", paste(missing_chunks, collapse = ", "))

  #1.2 Stack the per-bootstrap rows from every chunk and renumber them 1 to B
  chunks <- lapply(seq_len(params$n_chunks), function(i) readRDS(chunk_file(i)))
  boot_mat <- do.call(rbind, lapply(chunks, `[[`, 2))
  boot_mat[, "B"] <- seq_len(nrow(boot_mat))
  stopifnot(nrow(boot_mat) == params$n_boot)

  #1.3 Rebuild the structure stability_kproto() returns: overall means, then the per-bootstrap matrix
  boot_res <- list(colMeans(boot_mat[, -1, drop = FALSE]), boot_mat)

  #1.4 Persist the pooled result
  saveRDS(boot_res, rds_out)

  #1.5 Announce end of pooling and store in log
  if (!is.na(log_file)) cat(
    sprintf("%s|STAB_POOLED|config=%s|k=%d|B=%d|saved=%s\n", format(Sys.time(), "%Y-%m-%d %H:%M:%OS3"),
            params$config, params$k_value, params$n_boot, rds_out),
    file = log_file, append = TRUE
  )
}

#2. Unpack bootstrapped stability results into tidy objects
overall_vec <- boot_res[[1]]
boot_tbl <- as.data.frame(boot_res[[2]])
colnames(boot_tbl) <- c("B","Rand","Jaccard")

#3. Create a summary stats table outlining the results of the stability analysis
sum_tbl <- tibble(
  config = params$config,
  k = params$k_value,
  nstart = params$stab_nstart,
  boots = nrow(boot_tbl),
  overall_Rand = unname(overall_vec["R"]),
  overall_Jacc = unname(overall_vec["J"]),
  mean_Rand = mean(boot_tbl$Rand, na.rm = TRUE),
  mean_Jacc = mean(boot_tbl$Jaccard, na.rm = TRUE),
  pct_Jacc_ge_0.90 = 100 * mean(boot_tbl$Jaccard >= 0.90, na.rm = TRUE),
  sd_Rand = sd(boot_tbl$Rand, na.rm = TRUE),
  sd_Jacc = sd(boot_tbl$Jaccard, na.rm = TRUE),
  q05_Rand = quantile(boot_tbl$Rand, 0.05, na.rm = TRUE),
  q95_Rand = quantile(boot_tbl$Rand, 0.95, na.rm = TRUE),
  q05_Jacc = quantile(boot_tbl$Jaccard, 0.05, na.rm = TRUE),
  q95_Jacc = quantile(boot_tbl$Jaccard, 0.95, na.rm = TRUE)
)

#4. Write a CSV summary of the stability analysis results
write_csv(sum_tbl, csv_out)

#5. Print the summary table
knitr::kable(
  sum_tbl,
  caption = glue("Bootstrap stability summary ({params$config}, scaling={params$scaling_method}, k={params$k_value}, B={params$n_boot}, nstart={params$stab_nstart})"),
  digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))

```

## Plot Results of Stability Analysis

The following plots visualize the distribution and consistency of clustering stability metrics derived from bootstrap resampling (with replacement). The density plots with rug marks show the distribution of Rand and Jaccard indices across bootstrap samples, with reference lines indicating overall and mean values-higher, more concentrated distributions near 1.0 suggest stable clustering

The violin-boxplot compares the two stability metrics side-by-side, revealing their central tendencies and variability

The empirical cumulative distribution functions (ECDF) illustrate the probability that stability values fall below given thresholds, with steeper curves at higher values indicating better stability

Finally ,the scatter plot with density contours examines the correlation between Rand and Jaccard indices across bootstrap iterations, helping assess whether both metrics agree on clustering quality

Together, these visualizations provide some decent empirical visualizations for evaluating whether the chosen clustering solution produces consistent, reproducible partitions across resampled data 

```{r cluster stability plots, echo = FALSE, warning = FALSE, eval = (params$mode == "report")}

## Create plots for Cluster Stability Metric Results ##

#1. Generate rug + density plots for the Rand & Jaccard bootstrap results
#1.1 Create an aesthetics helper for the density and rug plots
vline_labs <- function(x, lbl) {
  list(
    geom_vline(xintercept = x, linetype = "dashed", linewidth = 0.7),
    annotate("label", x = x, y = Inf, vjust = 1.2,
             label = lbl, size = 3.5)
  )
}

#1.2 Density + rug for Rand
p_rand <- ggplot(boot_tbl, aes(x = Rand)) +
  geom_density(linewidth = 1) +
  geom_rug(alpha = 0.2) +
  vline_labs(sum_tbl$overall_Rand, "overall") +
  vline_labs(sum_tbl$mean_Rand,    "mean") +
  labs(
    title = glue("Bootstrap Rand index (k={params$k_value})"),
    x = "Rand", y = "Density") +
  coord_cartesian(xlim = c(0,1)) +
  theme_minimal(base_size = 14)

#1.3 Density + rug for Jaccard
p_jacc <- ggplot(boot_tbl, aes(x = Jaccard)) +
  geom_density(linewidth = 1) +
  geom_rug(alpha = 0.2) +
  vline_labs(sum_tbl$overall_Jacc, "overall") +
  vline_labs(sum_tbl$mean_Jacc,    "mean") +
  labs(
    title = glue("Bootstrap Jaccard index (k={params$k_value})"),
    x = "Jaccard", y = "Density") +
  coord_cartesian(xlim = c(0,1)) +
  theme_minimal(base_size = 14)

#2. Create a violin boxplot for both metrics
#2.1 Create a long pivoted version of the data for plotting
long_tbl <- boot_tbl %>%
  pivot_longer(c(Rand, Jaccard), names_to = "Metric", values_to = "Value")

#2.2 Create a violin boxplot representing both metrics
p_violin <- ggplot(long_tbl, aes(x = Metric, y = Value)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, outlier.size = 0.8) +
  coord_cartesian(ylim = c(0,1)) +
  labs(
    title = glue("Distribution of stability metrics (k={params$k_value})"),
    x = NULL, y = "Value") +
  theme_minimal(base_size = 14)

#3. Create Empirical Cumulative Distribution Functions (ECDF's) side-by-side for both stability indices
p_ecdf <- ggplot(long_tbl, aes(x = Value, color = Metric)) +
  stat_ecdf(linewidth = 1) +
  coord_cartesian(xlim = c(0,1)) +
  labs(
    title = glue("ECDF of stability metrics (k={params$k_value})"),
    x = "Value", y = "ECDF") +
  theme_minimal(base_size = 14)

#4. Create Rand vs Jaccard + density contours with scatter
p_scatter <- ggplot(boot_tbl, aes(x = Rand, y = Jaccard)) +
  geom_point(alpha = 0.4, size = 1.2) +
  stat_density_2d(color = "grey50", alpha = 0.5) +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  labs(
    title = glue("Rand vs Jaccard across bootstraps (k={params$k_value})"),
    x = "Rand", y = "Jaccard") +
  theme_minimal(base_size = 14)

#5. Print all plots for the stability analysis results
#5.1 Print the density and rug plot for the Rand index
print(p_rand)

#5.2 #Print the density and rug plot for the Jaccard index
print(p_jacc)

#5.3 Print the violin boxplot displaying the distribution of both indices
print(p_violin)

#5.4 Print the side-by-side ECDF for both indices
print(p_ecdf)

#5.5 Print the density contours with scatter for both indices
print(p_scatter)

#6. Log successful completion of the stability analysis
if (!is.na(log_file)) cat(
  sprintf("%s|STAB_REPORT_DONE|k=%d|B=%d|rds=%s|csv=%s\n", format(Sys.time(), "%Y-%m-%d %H:%M:%OS3"),
          params$k_value, params$n_boot, basename(rds_out), basename(csv_out)),
  file = log_file, append = TRUE
)

```
