---
title: "5. Parsimonious Feature Selection (MVFS; k-prototypes)"
author: "Sam A. Sievertsen"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    df_print: paged
params:
  mode: "auto"                 # "auto" | "step" | "aggregate"
                               # "auto": run step-level if SLURM_ARRAY_TASK_ID set, else aggregate
  search: "forward"            # "forward" | "backward" | "domain"
  step_index: !r NULL          # integer step to run (meaning depends on search)
  k_opt: 2                     # fixed k to match anchor solution
  nstart: 8                    # multi-start control (kproto retains best objective)
  lambda_mode: "estimate"      # "estimate" | "anchor_fixed"
  lambda_sensitivity: true     # if TRUE, compare aggregate summaries across lambda modes (if both traces exist)
  boot_B: 1000                 # bootstrap reps per candidate set
  boot_seed: 123               # base seed for bootstrap RNG
  ari_target: 0.80
  jaccard_target: 0.80
  silhouette_method: "validation_kproto"
  silhouette_type: "huang"               
  risk_signature_feats:        # used to define "higher-risk" anchor cluster + directionality checks
    - mh_p_cbcl__dsm__dep_tscore
    - mh_p_cbcl__synd__aggr_tscore
    - mh_p_cbcl__synd__attn_tscore
    - mh_p_cbcl__dsm__anx_tscore
    - mh_p_gbi_sum
    - sds_total
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(cache = FALSE, message = TRUE, warning = TRUE, results = "markup", verbose = TRUE, comment = "")

```

## Overview & Plan

This script implements step 5 of the clustering pipeline: identifying a Minimally Viable Feature Set (MVFS) that preserves the anchor (i.e., 17-feature) k-prototypes partition while remaining stable

Core outputs:

* MVFS feature priority table (MCR primary + centroid/SHAP two-vote local overrides)

* Step-level trace files: ARI vs anchor + clusterwise bootstrap Jaccard vs |S|

* Selected MVFS (smallest S meeting ARI + min-Jaccard thresholds)

* Sensitivity traces: backward elimination + domain-coverage MVFS + lambda handling sensitivity

### 1. Setup & Environment

**Goal:** Load packages, create output paths, configure parallel workers from SLURM, initialize logging, and generate a reproducible seed list shared across array tasks. Also define IO helper(s)

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}

## 1. Set-Up and Environment ## 
cat("Set-up beginning\n")

#1.1 Load packages
suppressPackageStartupMessages({
library(here)
library(knitr)
library(tibble)
library(data.table)
library(magrittr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(readr)
library(purrr)
library(stringr)
library(glue)
library(clustMixType)
library(FeatureImpCluster)
library(mclust)
library(clue)
library(cluster)
library(jsonlite)
library(ggplot2)
})

#1.2 Define canonical project I/O paths
paths <- list(
  
#1.2.1 Anchor clustering object
kproto_rds = here("data", "data_processed", "kproto_results", "kproto_z_score.rds"),

#1.2.2 Feature importance outputs computed in feature importance analysis script
fi_sum_dir = here("results", "main_analysis", "1_clustering", "4_feature_importance", "summary"),

#1.2.3 Outputs for this script
out_dir = here("results", "main_analysis", "1_clustering", "5_parsimonious_feature_selection"),
step_dir = here("results", "main_analysis", "1_clustering", "5_parsimonious_feature_selection", "step_level"),
sum_dir = here("results", "main_analysis", "1_clustering", "5_parsimonious_feature_selection", "summary"),
fig_dir = here("results", "main_analysis", "1_clustering", "5_parsimonious_feature_selection", "figures")
)

dir.create(paths$out_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$step_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$sum_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$fig_dir, showWarnings = FALSE, recursive = TRUE)

#1.3 Safe CSV writer
write_csv2 <- function(x, path) {
  dir.create(dirname(path), TRUE, TRUE)
  readr::write_csv(x, path)
}

cat("Set-up complete\n")

```

### 2. Load Anchor Clustering Solution

```{r load data, echo = FALSE, warning = FALSE}

#2.1 Load anchor k-prototypes object (k and lambda already chosen upstream)
kp_path <- paths$kproto_rds
if (!file.exists(kp_path))
  stop(glue("Anchor kproto RDS not found: {kp_path}"))
kp_any <- readRDS(kp_path)

#2.2 Helper: locate the desired kproto object
pick_kproto <- function(obj, k_target) {
  
  #2.2.1 Case 1: single kproto object
  if (inherits(obj, "kproto"))
    return(obj)
  
  #2.2.2 Case 2: list of kproto objects
  stopifnot(is.list(obj))
  nm <- names(obj)
  
  #2.2.2.1 Prefer exact name match ("k2", "k3", ...)
  if (!is.null(nm)) {
    nm_hit <- paste0("k", k_target)
    if (nm_hit %in% nm &&
        inherits(obj[[nm_hit]], "kproto"))
      return(obj[[nm_hit]])
  }
  
  #2.2.2.2 Fallback: first element whose size length matches k_target
  for (el in obj) {
    if (inherits(el, "kproto")) {
      sz <- tryCatch(
        el$size,
        error = function(...)
          NULL
      )
      if (!is.null(sz) && length(sz) == k_target)
        return(el)
    }
  }
  stop(glue("Could not locate a k-prototypes object for k={k_target} in {kp_path}" ))
}

#2.3 Select anchor object + canonical feature frame used at fit time
kp0 <- pick_kproto(kp_any, params$k_opt)
X_fit <- kp0$data
stopifnot(!is.null(X_fit), nrow(X_fit) > 0)

risk_dt <- as.data.table(X_fit)
features_all <- names(risk_dt)

#2.4 Anchor assignments on the canonical frame
anchor_assign <- if (!is.null(kp0$cluster)) kp0$cluster else {
pr <- predict(kp0, newdata = as.data.frame(risk_dt))
if (is.list(pr)) pr$cluster else pr
}
stopifnot(length(anchor_assign) == nrow(risk_dt))

#2.5 Anchor lambda (for lambda sensitivity)
lambda_anchor <- tryCatch(
  kp0$lambda,
  error = function(...)
    NA_real_)
if (!is.finite(lambda_anchor)) {
  message("kp0$lambda not found; estimating anchor lambda via lambdaest(kp0$data).")
  lambda_anchor <- clustMixType::lambdaest(as.data.frame(risk_dt))
}

#2.6 Define the anchor "higher-risk" cluster once using signature indicators (pre-specified)
sig_feats <- params$risk_signature_feats
sig_feats <- sig_feats[sig_feats %in% features_all]
if (!length(sig_feats)) stop("No risk_signature_feats found in anchor feature set.")

sig_tbl_anchor <- risk_dt[, ..sig_feats] |>
  as_tibble() |>
  mutate(.cluster = factor(anchor_assign)) |>
  pivot_longer(-.cluster, names_to = "feature", values_to = "value") |>
  group_by(.cluster, feature) |>
  summarise(mu = mean(value, na.rm = TRUE), .groups = "drop") |>
  group_by(.cluster) |>
  summarise(sig_mean = mean(mu, na.rm = TRUE), .groups = "drop")

anchor_high_risk_cluster <- sig_tbl_anchor |>
  arrange(desc(sig_mean)) |>
  slice(1) |>
  pull(.cluster) |>
  as.character()

message(glue("Anchor higher-risk cluster (by signature mean): {anchor_high_risk_cluster}"))

```

### 3. Build MVFS Candidate Ordering

```{r mvfs order, echo = FALSE, warning = FALSE}

#3.1 Load MCR summary
mcr_path_a <- file.path(paths$fi_sum_dir, "feature_importance_summary.csv")
mcr_path_b <- file.path(paths$fi_sum_dir, "feature_importance_summary_feature_spec_mcr.csv")

if (file.exists(mcr_path_a)) {
  mcr_raw <- readr::read_csv(mcr_path_a, show_col_types = FALSE)
} else if (file.exists(mcr_path_b)) {
  mcr_raw <- readr::read_csv(mcr_path_b, show_col_types = FALSE)
} else {
  stop(glue("Missing MCR summary file(s) in {paths$fi_sum_dir}"))
}

mcr_tab <- mcr_raw |>
  rename_with( ~ "feature", .cols = any_of(c(
    "feature", "var", "variable", "feature_name"))) |>
  rename_with( ~ "mcr_mean", .cols = any_of(c("mcr_mean", "mean_mcr", "mcr"))) |>
  mutate(mcr_mean = as.numeric(mcr_mean)) |>
  dplyr::select(feature, mcr_mean, everything())

stopifnot("feature" %in% names(mcr_tab), "mcr_mean" %in% names(mcr_tab))

#3.2 Load centroid separation summary
cent_path <- file.path(paths$fi_sum_dir, "centroid_separation_summary.csv")
if (!file.exists(cent_path)) stop(glue("Missing centroid_separation_summary.csv in {paths$fi_sum_dir}"))
cent_raw <- readr::read_csv(cent_path, show_col_types = FALSE)

cent_tab <- cent_raw |>
  rename_with(~"feature", .cols = any_of(c("feature", "var", "variable", "feature_name"))) |>
  mutate(
    delta_z = as.numeric(delta_z),
    max_cohens_d = as.numeric(max_cohens_d)) |>
  transmute(
    feature,
    delta_z = delta_z,
    max_cohens_d = max_cohens_d,
    sep_abs = abs(delta_z))

#3.3 Load SHAP global importance
shap_path <- file.path(paths$fi_sum_dir, "shap_global_importance.csv")
if (!file.exists(shap_path)) stop(glue("Missing shap_global_importance.csv in {paths$fi_sum_dir}"))
shap_raw <- readr::read_csv(shap_path, show_col_types = FALSE)

shap_tab <- shap_raw |>
  rename_with( ~ "feature", .cols = any_of(c(
    "feature", "var", "variable", "feature_name"))) |>
  rename_with( ~ "mean_abs_shap", .cols = any_of(c(
    "mean_abs_shap", "mean_shap", "shap_mean_abs"))) |>
  mutate(mean_abs_shap = as.numeric(mean_abs_shap)) |>
  select(feature, mean_abs_shap)

#3.4 Merge on FULL anchor feature set (prevents accidental feature drop)
base_feat <- tibble::tibble(feature = features_all)
rank_tab <- base_feat |>
  left_join(mcr_tab |> dplyr::select(feature, mcr_mean), by = "feature") |>
  left_join(cent_tab, by = "feature") |>
  left_join(shap_tab, by = "feature") |>
  mutate(
    
    #3.4.1 Missing => rank-last (keeps feature but deprioritizes)
    mcr_mean = ifelse(is.na(mcr_mean), -Inf, mcr_mean),
    sep_abs = ifelse(is.na(sep_abs), -Inf, sep_abs),
    mean_abs_shap = ifelse(is.na(mean_abs_shap), -Inf, mean_abs_shap),
    rank_MCR = rank(-mcr_mean, ties.method = "min"),
    rank_centroid = rank(-sep_abs, ties.method = "min"),
    rank_SHAP = rank(-mean_abs_shap, ties.method = "min")) |>
  arrange(rank_MCR)

#3.4.2 Hard stop if somehow we still don’t have all anchor features
stopifnot(length(rank_tab$feature) == length(features_all))

#3.5 Two-vote local-swap pass (consensus promotion)
consensus_promote <- function(tab) {

ord <- tab$feature[order(tab$rank_MCR)]
rC <- setNames(tab$rank_centroid, tab$feature)
rS <- setNames(tab$rank_SHAP, tab$feature)

swapped <- TRUE
while (swapped) {
  
  swapped <- FALSE
  
  for (i in seq_len(length(ord) - 1L)) {
    A <- ord[i]
    B <- ord[i + 1L]
    if (is.finite(rC[A]) &&
        is.finite(rC[B]) && is.finite(rS[A]) && is.finite(rS[B])) {
      
      # Swap A and B iff centroid and SHAP both prefer B
      if (rC[B] < rC[A] && rS[B] < rS[A]) {
        ord[i] <- B
        ord[i + 1L] <- A
        swapped <- TRUE
      }
    }
  }
}

ord
}

final_order <- consensus_promote(rank_tab)

rank_tab <- rank_tab |>
  mutate(rank_final_priority = match(feature, final_order)) |>
  arrange(rank_final_priority)

#3.6 Write priority list table
prio_path <- file.path(paths$sum_dir, "mvfs_feature_priority_table.csv")
if (params$mode %in% c("aggregate", "auto")) {
  if (!file.exists(prio_path)) write_csv2(rank_tab, prio_path)
}

#3.7 Convenience vector for downstream
priority_features <- rank_tab$feature

rank_tab |>
  arrange(rank_final_priority) |>
  knitr::kable(digits = 3, caption = "MVFS Priority List (MCR primary + 2-vote local override)") |>
  kableExtra::kable_styling(full_width = FALSE)

```

### 4. Helper Functions

```{r helper functions, echo = FALSE, warning = FALSE}

#4.1 Align newdata to model (basic type/level harmonization)
align_newdata_to_model <- function(model, newdata) {
  X <- as.data.frame(model$data)
  Y <- as.data.frame(newdata)
  common <- intersect(names(X), names(Y))
  for (nm in common) {
    if (is.factor(X[[nm]])) {
      Y[[nm]] <- factor(as.character(Y[[nm]]), levels = levels(X[[nm]]))
    } else if (is.numeric(X[[nm]]) && !is.numeric(Y[[nm]])) {
      Y[[nm]] <- suppressWarnings(as.numeric(Y[[nm]]))
    }
  }
  Y
}

#4.2 Predict cluster labels as integer vector
pred_kproto <- function(kp, newdata) {
  nd <- align_newdata_to_model(kp, newdata)
  pr <- predict(kp, newdata = nd)
  if (is.list(pr))
    pr$cluster
  else
    pr
}

#4.3 Extract objective value (best-effort; used only for debugging / tie checks)
kproto_objective <- function(kp) {
  for (nm in c("tot.withinss",
               "tot.withindist",
               "tot.within.dist",
               "tot.within")) {
    if (!is.null(kp[[nm]]) &&
        is.finite(kp[[nm]]))
      return(as.numeric(kp[[nm]]))
  }
  NA_real_
}

#4.4 Fit k-prototypes with multi-start and retain best (kproto handles nstart internally)
fit_kproto_best <- function(X_df,
                            k,
                            lambda,
                            nstart = 8L,
                            seed = NULL) {
  if (!is.null(seed))
    set.seed(seed)
  kp <- clustMixType::kproto(
    x = X_df,
    k = k,
    lambda = lambda,
    nstart = as.integer(nstart),
    verbose = FALSE)
  list(model = kp, obj = kproto_objective(kp))
}

#4.5 Hungarian mapping between two partitions: map labels in 'from' -> 'to' (max overlap)
hungarian_map <- function(from, to) {
  from <- as.integer(from)
  to <- as.integer(to)
  K1 <- length(unique(from))
  K2 <- length(unique(to))
  K <- max(K1, K2)
  
  tab <- table(from, to)
  tab_full <- matrix(0, nrow = K, ncol = K)
  tab_full[seq_len(nrow(tab)), seq_len(ncol(tab))] <- tab
  
  #4.5.1 Maximize overlap => minimize cost = max(tab) - tab
  maxv <- max(tab_full)
  cost <- maxv - tab_full
  assign <- clue::solve_LSAP(cost)
  as.integer(assign)
}

#4.6 Apply mapping vector to labels (labels assumed 1..K)
apply_map <- function(lbl, mapping) {
  lbl <- as.integer(lbl)
  out <- lbl
  ok <- lbl >= 1 & lbl <= length(mapping)
  out[ok] <- mapping[lbl[ok]]
  out
}

#4.7 Clusterwise Jaccard between two labelings (after mapping)
clusterwise_jaccard <- function(lbl_a, lbl_b, K = NULL) {
  lbl_a <- as.integer(lbl_a)
  lbl_b <- as.integer(lbl_b)
  if (is.null(K))
    K <- max(length(unique(lbl_a)), length(unique(lbl_b)))
  out <- rep(NA_real_, K)
  for (k in seq_len(K)) {
    A <- which(lbl_a == k)
    B <- which(lbl_b == k)
    den <- length(union(A, B))
    out[k] <- if (den == 0)
      NA_real_
    else
      length(intersect(A, B)) / den
  }
  out
}

#4.8 Directionality check vs anchor on signature indicators (post mapping)
signature_direction_check <- function(dt,
                                      lbl_mapped,
                                      anchor_lbl,
                                      sig_feats,
                                      anchor_hi_label_chr) {
  cand <- dt[, ..sig_feats] |>
    as_tibble() |>
    mutate(.cluster = factor(lbl_mapped)) |>
    pivot_longer(-.cluster, names_to = "feature", values_to = "value") |>
    group_by(.cluster, feature) |>
    summarise(mu = mean(value, na.rm = TRUE), .groups = "drop")
  
  anch <- dt[, ..sig_feats] |>
    as_tibble() |>
    mutate(.cluster = factor(anchor_lbl)) |>
    pivot_longer(-.cluster, names_to = "feature", values_to = "value") |>
    group_by(.cluster, feature) |>
    summarise(mu = mean(value, na.rm = TRUE), .groups = "drop")
  
  cl_levels <- sort(unique(as.character(anchor_lbl)))
  if (length(cl_levels) < 2)
    return(list(prop_match = NA_real_, detail = NULL))
  
  hi <- anchor_hi_label_chr
  lo <- setdiff(cl_levels, hi)[1]
  
  delta_anchor <- anch |>
    filter(.cluster %in% c(hi, lo)) |>
    pivot_wider(names_from = .cluster, values_from = mu) |>
    mutate(delta = .data[[hi]] - .data[[lo]]) |>
    dplyr::select(feature, delta)
  
  delta_cand <- cand |>
    filter(.cluster %in% c(hi, lo)) |>
    pivot_wider(names_from = .cluster, values_from = mu) |>
    mutate(delta = .data[[hi]] - .data[[lo]]) |>
    dplyr::select(feature, delta)
  
  chk <- delta_anchor |>
    rename(delta_anchor = delta) |>
    left_join(delta_cand |> rename(delta_cand = delta), by = "feature") |>
    mutate(
      sign_match = sign(delta_anchor) == sign(delta_cand),
      sign_match = ifelse(is.na(delta_anchor) | is.na(delta_cand), NA, sign_match))
  
  list(prop_match = mean(chk$sign_match, na.rm = TRUE),
       detail = chk)
}

#4.9 Extract a validation_kproto index robustly
extract_validation_index <- function(x) {
  if (is.numeric(x) && length(x) == 1L) return(as.numeric(x))
  if (is.list(x) && "index" %in% names(x)) return(as.numeric(x$index))
  if (is.list(x) && "index_opt" %in% names(x)) return(as.numeric(x$index_opt))
  NA_real_
}

#4.10 Silhouette via clustMixType::validation_kproto() on a fitted kproto object
silhouette_validation_kproto <- function(kp, type = "huang") {
  out <- tryCatch(
    clustMixType::validation_kproto(method = "silhouette", object = kp, type = type),
    error = function(e) NA)
  extract_validation_index(out)
}

#4.11 Bootstrap stability (clusterwise Jaccard) for a candidate set
bootstrap_stability <- function(dt_full,
                                cand_lbl,
                                k,
                                lambda,
                                nstart,
                                B = 200L,
                                seed = 123L) {
  n <- nrow(dt_full)
  K <- length(unique(cand_lbl))
  set.seed(seed)
  
  J <- matrix(NA_real_, nrow = as.integer(B), ncol = as.integer(K))
  
  for (b in seq_len(as.integer(B))) {
    
    #4.11.1 Bootstrap sample
    idx <- sample(seq_len(n), n, replace = TRUE)
    dt_boot <- dt_full[idx, , drop = FALSE]
    
    #4.11.2 Fit on bootstrap sample
    kp_b <- fit_kproto_best(
      as.data.frame(dt_boot),
      k = k,
      lambda = lambda,
      nstart = nstart,
      seed = seed + b)$model
    
    #4.11.3 Predict labels on full data (stable universe)
    boot_lbl <- pred_kproto(kp_b, as.data.frame(dt_full))
    
    #4.11.4 Map bootstrap labels -> candidate labels by overlap
    mapping <- hungarian_map(from = boot_lbl, to = cand_lbl)
    boot_lbl_mapped <- apply_map(boot_lbl, mapping)
    
    #4.11.5 Clusterwise Jaccard
    J[b, ] <- clusterwise_jaccard(cand_lbl, boot_lbl_mapped, K = K)
    
  }
  
  list(
    jaccard_by_cluster = colMeans(J, na.rm = TRUE),
    jaccard_mean = mean(J, na.rm = TRUE),
    jaccard_min = min(colMeans(J, na.rm = TRUE), na.rm = TRUE))
}

#4.12 Secondary full-set concordance (descriptive): among features in S, quantify how often the candidate hi-vs-lo direction matches the anchor
fullset_concordance_check <- function(dt,
                                      cand_lbl_mapped,
                                      anchor_lbl,
                                      feats_S,
                                      anchor_hi_label_chr) {
  feats_S <- feats_S[feats_S %in% names(dt)]
  if (!length(feats_S))
    return(list(
      prop_match = NA_real_,
      n_eval = 0L,
      detail = NULL))
  
  cl_levels <- sort(unique(as.character(anchor_lbl)))
  if (length(cl_levels) < 2)
    return(list(
      prop_match = NA_real_,
      n_eval = 0L,
      detail = NULL))
  
  hi <- anchor_hi_label_chr
  lo <- setdiff(cl_levels, hi)[1]
  
  detail <- lapply(feats_S, function(f) {
    x <- dt[[f]]
    a_hi <- x[as.character(anchor_lbl) == hi]
    a_lo <- x[as.character(anchor_lbl) == lo]
    c_hi <- x[as.character(cand_lbl_mapped) == hi]
    c_lo <- x[as.character(cand_lbl_mapped) == lo]
    
    #4.12.1 Numeric: sign of mean difference
    if (is.numeric(x)) {
      da <- mean(a_hi, na.rm = TRUE) - mean(a_lo, na.rm = TRUE)
      dc <- mean(c_hi, na.rm = TRUE) - mean(c_lo, na.rm = TRUE)
      ok <- ifelse(is.na(da) | is.na(dc), NA, sign(da) == sign(dc))
      return(tibble::tibble(
        feature = f,
        type = "numeric",
        match = ok))
    }
    
    #4.12.2 Categorical: require informative anchor separation (different modes)
    if (is.factor(x) || is.character(x)) {
      mode_fun <- function(v) {
        v <- v[!is.na(v)]
        if (!length(v))
          return(NA_character_)
        names(sort(table(v), decreasing = TRUE))[1]
      }
      a_m_hi <- mode_fun(a_hi)
      a_m_lo <- mode_fun(a_lo)
      c_m_hi <- mode_fun(c_hi)
      c_m_lo <- mode_fun(c_lo)
      
      #4.12.3 If anchor mode doesn't differ across clusters, direction is uninformative
      if (is.na(a_m_hi) ||
          is.na(a_m_lo) || identical(a_m_hi, a_m_lo)) {
        return(tibble::tibble(
          feature = f,
          type = "categorical",
          match = NA))
      }
      
      ok <- identical(c_m_hi, a_m_hi) && identical(c_m_lo, a_m_lo)
      return(tibble::tibble(
        feature = f,
        type = "categorical",
        match = ok))
    }
    
    tibble::tibble(feature = f,
                   type = "other",
                   match = NA)}) |>
    dplyr::bind_rows()
  
  prop <- mean(detail$match, na.rm = TRUE)
  n_eval <- sum(!is.na(detail$match))
  
  list(prop_match = prop,
       n_eval = as.integer(n_eval),
       detail = detail)
}

#4.13 Count variable types (based on current column classes)
count_types <- function(df) {
  df <- as.data.frame(df)
  is_num <- vapply(df, is.numeric, logical(1))
  is_fac <- vapply(df, is.factor,  logical(1))
  list(
    n_num = sum(is_num),
    n_fac = sum(is_fac),
    is_num = is_num,
    is_fac = is_fac
  )
}

#4.14 Ensure kproto input has at least one numeric AND one factor column.
#     (Adds constant "dummy" columns when a candidate set is single-type.)
#     This does NOT change distances because the dummy column contributes 0 to all pairwise distances.
prep_for_kproto <- function(df) {
  df <- as.data.frame(df)
  ct <- count_types(df)

  add_num <- (ct$n_num == 0L)
  add_fac <- (ct$n_fac == 0L)

  if (add_num) df$.mvfs_dummy_num <- 0
  if (add_fac) df$.mvfs_dummy_fac <- factor("A")

  list(
    x = df,
    add_num = add_num,
    add_fac = add_fac
  )
}

#4.15 Safe lambda rule:
# - anchor_fixed: use anchor lambda (for sensitivity)
# - otherwise:
#   * if |S|<2 OR single-type: lambda not identifiable/relevant => use 1
#   * else: use lambdaest on the ORIGINAL X (not dummy-augmented)
lambda_safe <- function(X_orig, lambda_mode, lambda_anchor) {
  if (identical(lambda_mode, "anchor_fixed")) {
    return(list(lambda = lambda_anchor, reason = "anchor_fixed"))
  }

  ct <- count_types(X_orig)

  if (ncol(X_orig) < 2L) {
    return(list(lambda = 1, reason = "p<2: lambdaest undefined; lambda irrelevant"))
  }

  if (ct$n_num == 0L || ct$n_fac == 0L) {
    return(list(lambda = 1, reason = "single-type set: lambda irrelevant"))
  }

  lam <- clustMixType::lambdaest(X_orig, verbose = FALSE)
  list(lambda = lam, reason = "lambdaest (mixed-type set)")
}

```

### 5. Candidate Set Builder 

```{r candidate feature set, echo = FALSE, warning = FALSE}

#5.1 Domain map (Sensitivity 3): pre-specified domains
domain_map <- tibble::tribble(
~feature, ~domain,

#5.1.1 Clinical/prodromal
"mh_p_cbcl__synd__aggr_tscore", "clinical_prodromal",
"mh_p_cbcl__dsm__anx_tscore", "clinical_prodromal",
"mh_p_cbcl__synd__attn_tscore", "clinical_prodromal",
"mh_p_cbcl__dsm__dep_tscore", "clinical_prodromal",
"mh_p_gbi_sum", "clinical_prodromal",
"mh_y_upps__nurg_sum", "clinical_prodromal",
"mh_y_upps__purg_sum", "clinical_prodromal",
"sds_total", "clinical_prodromal",

#5.1.2 Environmental
"ACE_index_sum_score", "environmental",
"bullying", "environmental",
"le_l_coi__addr1__coi__total__national_zscore", "environmental",
"fc_p_nsc__ns_mean", "environmental",

#5.1.3 Familial
"family_history_depression", "familial",
"family_history_mania", "familial",

#5.1.4 Neurocognitive
"nc_y_nihtb__flnkr__uncor_score", "neurocognitive",
"nc_y_nihtb__lswmt__uncor_score", "neurocognitive",
"nc_y_nihtb__pttcp__uncor_score", "neurocognitive")

#5.2 Helper: build the feature set S for a given step
build_feature_set <- function(search,
                              step_idx,
                              priority_features,
                              domain_map) {
  p <- length(priority_features)
  stopifnot(step_idx >= 1L)
  
  if (identical(search, "forward")) {
    
    #5.2.1 Forward: step_idx = |S|
    s <- min(step_idx, p)
    return(priority_features[seq_len(s)])
  }
  
  if (identical(search, "backward")) {

    #5.2.2 Backward: step_idx = 1..p corresponds to keeping p..1 features
    s <- p - step_idx + 1L
    s <- max(1L, min(s, p))
    return(priority_features[seq_len(s)])
  
  }
  
  if (identical(search, "domain")) {
    dm <- domain_map |>
      filter(feature %in% priority_features)
    
    #5.2.3 Identify the top-ranked feature per domain
    top_per_domain <- dm |>
      mutate(rank = match(feature, priority_features)) |>
      group_by(domain) |>
      slice_min(rank, n = 1, with_ties = FALSE) |>
      ungroup() |>
      arrange(rank) |>
      pull(feature)
    
    base <- unique(top_per_domain)
    
    #5.2.4 Continue adding in priority order until size = step_idx
    if (length(base) >= step_idx)
      return(base[seq_len(step_idx)])
    
    rest <- setdiff(priority_features, base)
    need <- step_idx - length(base)
    add  <- head(rest, need)
    return(c(base, add))
    
  }
  
  stop(glue("Unknown search mode: {search}"))
}

```

### 6. Run a Single Test Step

```{r test step, echo = FALSE, warning = FALSE}

#6.1 Run one candidate set and write a step-level CSV artifact
run_step <- function(step_idx,
                     search = params$search,
                     lambda_mode = params$lambda_mode) {
  
  #6.1.1 Build candidate feature set
  S <- build_feature_set(search, step_idx, priority_features, domain_map)
  S <- S[S %in% features_all]
  stopifnot(length(S) >= 1)
  
  dt_S <- risk_dt[, ..S]
  X_S <- as.data.frame(dt_S)
  
  #6.1.2 Lambda handling (safe)
  lam_info <- lambda_safe(X_S, lambda_mode = lambda_mode, lambda_anchor = lambda_anchor)
  lambda_S <- lam_info$lambda
  lambda_reason <- lam_info$reason
  
  #6.1.2.1 Ensure kproto can run even for single-type sets (implementation detail)
  prep <- prep_for_kproto(X_S)
  X_S_fit <- prep$x
  added_dummy <- (prep$add_num || prep$add_fac)

  
  #6.1.3 Fit k-prototypes (multi-start; retains best objective)
  fit <- fit_kproto_best(X_S_fit,
                       k = params$k_opt,
                       lambda = lambda_S,
                       nstart = params$nstart,
                       seed = 1000 + step_idx)
  kpS <- fit$model
  obj <- fit$obj
  
  #6.1.4 Candidate labels on full data
  cand_lbl <- pred_kproto(kpS, X_S_fit)
  k_observed <- length(unique(cand_lbl))
  min_cluster_n <- min(tabulate(cand_lbl))

  #6.1.4.1 Optionally treat non-2-cluster solutions as automatic failure
  if (k_observed < params$k_opt) {
    ari <- NA_real_
    sil <- NA_real_
    stab <- list(jaccard_mean = NA_real_, jaccard_min = NA_real_)
  }
  
  #6.1.4.2 If k-prototypes collapses to <k clusters, treat as failure and skip expensive metrics
  if (k_observed < params$k_opt || min_cluster_n < 2) {
    out <- tibble::tibble(
      search = search,
      lambda_mode = lambda_mode,
      step_idx = step_idx,
      n_features = length(S),
      features = paste(S, collapse = ";"),
      lambda = lambda_S,
      objective = obj,
      ari = NA_real_,
      jaccard_mean = NA_real_,
      jaccard_min = NA_real_,
      silhouette_mean = NA_real_,
      sig_direction_match = NA_real_,
      full_direction_match = NA_real_,
      full_direction_n_eval = NA_integer_,
      lambda_reason = lambda_reason,
      added_dummy = added_dummy,
      k_observed = k_observed,
      min_cluster_n = min_cluster_n,
      fail_reason = "collapsed_solution")
    stub <- glue("mvfs_step_{search}_{lambda_mode}_n{length(S)}.csv")
    write_csv2(out, file.path(paths$step_dir, stub))
    return(invisible(out))
  }

  #6.1.5 Similarity-to-anchor (ARI; label-invariant)
  ari <- mclust::adjustedRandIndex(cand_lbl, anchor_assign)
  
  #6.1.6 Mapping candidate clusters -> anchor clusters (for interpretation ONLY)
  map_c2a <- hungarian_map(from = cand_lbl, to = anchor_assign)
  cand_lbl_mapped <- apply_map(cand_lbl, map_c2a)
  
  #6.1.7 Risk label inheritance + directionality check
  sig_check <- signature_direction_check(
    dt = risk_dt,
    lbl_mapped = cand_lbl_mapped,
    anchor_lbl = anchor_assign,
    sig_feats = sig_feats,
    anchor_hi_label_chr = as.character(anchor_high_risk_cluster))
  
  #6.1.7.1 Secondary full-set concordance (descriptive) on features in S
  full_chk <- fullset_concordance_check(
    dt = risk_dt,
    cand_lbl_mapped = cand_lbl_mapped,
    anchor_lbl = anchor_assign,
    feats_S = S,
    anchor_hi_label_chr = as.character(anchor_high_risk_cluster))
  
  #6.1.8 Bootstrap stability (clusterwise Jaccard against candidate partition)
  stab <- bootstrap_stability(
    dt_full = as.data.table(X_S_fit),
    cand_lbl = cand_lbl,
    k = params$k_opt,
    lambda = lambda_S,
    nstart = params$nstart,
    B = params$boot_B,
    seed = params$boot_seed + 10L * step_idx)
  
  #6.1.9 Silhouette once per candidate set; validation_kproto on fitted object
  sil <- silhouette_validation_kproto(
    kpS,
    type = params$silhouette_type)
  
  #6.1.10 Output row
  out <- tibble::tibble(
    search = search,
    lambda_mode = lambda_mode,
    step_idx = step_idx,
    n_features = length(S),
    features = paste(S, collapse = ";"),
    lambda = lambda_S,
    objective = obj,
    ari = ari,
    jaccard_mean = stab$jaccard_mean,
    jaccard_min = stab$jaccard_min,
    silhouette_mean = sil,
    sig_direction_match = sig_check$prop_match,
    full_direction_match = full_chk$prop_match,
    full_direction_n_eval = full_chk$n_eval,
    lambda_reason = lambda_reason,
    added_dummy = added_dummy,
    k_observed = k_observed,
    min_cluster_n = min_cluster_n)
  
  #6.1.11 Write step-level csv (one file per step x lambda_mode)
  stub <- glue("mvfs_step_{search}_{lambda_mode}_n{length(S)}.csv")
  write_csv2(out, file.path(paths$step_dir, stub))
  
  invisible(out)
}

```

### 7. Dispatch Step vs Aggregate Job

```{r dispatch, echo = FALSE, warning = FALSE}

#7.1 Determine whether we are running a single step (SLURM array) or aggregating
if (params$mode %in% c("auto", "step")) {
  arr <- Sys.getenv("SLURM_ARRAY_TASK_ID")
  idx <- if (!is.null(params$step_index))
    params$step_index
  else if (nzchar(arr))
    as.integer(arr)
  else
    NA_integer_
  
  if (is.finite(idx)) {
    
    #7.1.1 Step mode: run one step and exit
    run_step(
      step_idx = idx,
      search = params$search,
      lambda_mode = params$lambda_mode)
    knitr::knit_exit()
    
  }
}

#7.2 Auto local fallback: if not on SLURM and no step files exist, run full trace now
if (params$mode == "auto") {
  step_files_now <- list.files(paths$step_dir, pattern = "^mvfs_step_.*\\.csv$", full.names = TRUE)
  arr_now <- Sys.getenv("SLURM_ARRAY_TASK_ID")

  if (!nzchar(arr_now) && !length(step_files_now)) {
    message("No SLURM array detected and no step files found; running full trace locally (1..P).")
    P <- length(priority_features)
    for (i in seq_len(P)) {
      run_step(step_idx = i, search = params$search, lambda_mode = params$lambda_mode)
    }
  }
}

#7.3 Aggregate mode: continue below (reads step-level files)
if (params$mode %in% c("auto", "aggregate")) {
  
  #7.3.1 Load all step-level csvs
  step_files <- list.files(paths$step_dir, pattern = "^mvfs_step_.*\\.csv$", full.names = TRUE)
  if (!length(step_files))
    stop(glue("No step-level MVFS files found in: {paths$step_dir}"))
  
  mvfs_df <- lapply(step_files, readr::read_csv, show_col_types = FALSE) |>
    bind_rows() |>
    arrange(search, lambda_mode, n_features)
  
  write_csv2(mvfs_df, file.path(paths$sum_dir, "mvfs_trace_all.csv"))
  
  #7.3.2 Identify MVFS for each trace (first n_features meeting ARI + Jaccard targets)
  mvfs_pick <- mvfs_df |>
    group_by(search, lambda_mode) |>
    arrange(n_features) |>
    mutate(pass = (ari >= params$ari_target) &
             (jaccard_min >= params$jaccard_target)) |>
    summarise(
      mvfs_n = ifelse(
        any(pass, na.rm = TRUE),
        min(n_features[pass], na.rm = TRUE),
        NA_real_),
      mvfs_row = ifelse(any(pass, na.rm = TRUE), which(pass)[1], NA_integer_),
      .groups = "drop")
  
  write_csv2(mvfs_pick, file.path(paths$sum_dir, "mvfs_selected_summary.csv"))
  
    # 7.3.2.1 Refit + save the selected MVFS solution(s)
    mvfs_pick_ok <- mvfs_pick |> dplyr::filter(!is.na(mvfs_n))
  
    if (nrow(mvfs_pick_ok) > 0) {
  
      mvfs_solutions <- purrr::pmap(
        mvfs_pick_ok |> dplyr::select(search, lambda_mode, mvfs_n),
        function(search, lambda_mode, mvfs_n) {
  
          # Build selected feature set
          S <- build_feature_set(search, mvfs_n, priority_features, domain_map)
          S <- S[S %in% features_all]
  
          X_S <- as.data.frame(risk_dt[, ..S])
  
          # Lambda handling + kproto input prep
          lam_info <- lambda_safe(X_S, lambda_mode = lambda_mode, lambda_anchor = lambda_anchor)
          prep <- prep_for_kproto(X_S)
  
          # Refit once (no bootstrap here)
          kp <- fit_kproto_best(
            X_df   = prep$x,
            k      = params$k_opt,
            lambda = lam_info$lambda,
            nstart = params$nstart,
            seed   = 9000 + mvfs_n
          )$model
  
          # Labels + mapping to anchor labels (for interpretability)
          lbl <- pred_kproto(kp, prep$x)
          map_c2a <- hungarian_map(from = lbl, to = anchor_assign)
          lbl_mapped <- apply_map(lbl, map_c2a)
  
          # Trim stored data to reduce file size
          kp$data <- NULL
  
          list(
            search = search,
            lambda_mode = lambda_mode,
            mvfs_n = mvfs_n,
            features = S,
            lambda = lam_info$lambda,
            lambda_reason = lam_info$reason,
            model = kp,
            labels_raw = lbl,
            labels_mapped_to_anchor = lbl_mapped,
            anchor_assign = anchor_assign,
            map_c2a = map_c2a,
            anchor_high_risk_cluster = anchor_high_risk_cluster
          )
        }
      )
  
      saveRDS(
        mvfs_solutions,
        file.path(paths$sum_dir, "mvfs_selected_solutions.rds")
      )
  
      # Also write a flat CSV of assignments (first solution if only one trace)
      sol1 <- mvfs_solutions[[1]]
  
      assign_tbl <- tibble::tibble(
        row_id = seq_len(nrow(risk_dt)),
        anchor_cluster = as.integer(sol1$anchor_assign),
        mvfs_cluster_mapped = as.integer(sol1$labels_mapped_to_anchor)
      )
  
      if (!is.null(rownames(risk_dt))) {
        assign_tbl <- dplyr::mutate(assign_tbl, rowname = rownames(risk_dt))
      }
  
      write_csv2(
        assign_tbl,
        file.path(paths$sum_dir, "mvfs_selected_assignments.csv")
      )
    }
  
  #7.3.3 Plot traces (ARI and stability over |S|)
  mvfs_pick_ok <- mvfs_pick |> dplyr::filter(!is.na(mvfs_n))

  mvfs_point <- mvfs_df |>
    dplyr::inner_join(mvfs_pick_ok, by = c("search", "lambda_mode")) |>
    dplyr::filter(n_features == mvfs_n) |>
    dplyr::mutate(label = paste0("MVFS = ", mvfs_n))

  p_ari <- ggplot(mvfs_df, aes(x = n_features, y = ari)) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1.6) +
    facet_grid(search ~ lambda_mode, scales = "free_x") +
    geom_hline(yintercept = params$ari_target, linetype = 2) +
    geom_vline(data = mvfs_pick_ok, aes(xintercept = mvfs_n), linetype = 2) +
    geom_point(data = mvfs_point, size = 2.6) +
    geom_text(
      data = mvfs_point,
      aes(label = label),
      vjust = -0.8,
      size = 3
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
      title = "Agreement with anchor partition across candidate set size",
      subtitle = glue("Search = {paste(unique(mvfs_df$search), collapse=', ')};
        lambda = {paste(unique(mvfs_df$lambda_mode), collapse=', ')};
        Selection rule: ARI ≥ {params$ari_target}, min Jaccard ≥ {params$jaccard_target}"),
      x = "Candidate set size, |S| (number of features)",
      y = "Adjusted Rand index vs anchor") +
    theme_classic(base_size = 12) +
    theme(
      strip.background = element_rect(fill = "white", colour = "black"),
      axis.line = element_line(),
      plot.title.position = "plot")

  ggsave(
    file.path(paths$fig_dir, "mvfs_trace_ari.png"),
    p_ari,
    width = 9,
    height = 5,
    dpi = 600,
    bg = "white"
  )

  
}

```

### 8. Reproducibility Log

```{r reproducibility, echo = FALSE, warning = FALSE}

log <- list(
  timestamp = as.character(Sys.time()),
  mode = params$mode,
  search = params$search,
  k = params$k_opt,
  n = nrow(risk_dt),
  p_anchor = length(features_all),
  lambda_anchor = lambda_anchor,
  lambda_mode = params$lambda_mode,
  nstart = params$nstart,
  boot_B = params$boot_B,
  boot_seed = params$boot_seed,
  ari_target = params$ari_target,
  jaccard_target = params$jaccard_target,
  container = "abcd-mds-risk-r_0.1.7.sif")

writeLines(jsonlite::toJSON(log, pretty = TRUE), con = file.path(paths$sum_dir, "run_log.json"))

```

