---
title: "7. Missingness Diagnostics + Imputation (missForest; mice) + k-prototypes Refit (k=2:8)"
author: "Sam A. Sievertsen"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    df_print: paged
params:
  # I/O
  REPO: "/home/exacloud/gscratch/NagelLab/staff/sam/projects/ABCD_MDS_Risk"

  # Unfiltered baseline risk frame (includes rows excluded previously due to missingness)
  risk_unfiltered_path: "/home/exacloud/gscratch/NagelLab/staff/sam/projects/ABCD_MDS_Risk/data/data_processed/risk_variable_data_unfiltered.csv"

  # Baseline anchor k-prototypes object (z-scored pipeline)
  kproto_anchor_rds: "/home/exacloud/gscratch/NagelLab/staff/sam/projects/ABCD_MDS_Risk/data/data_processed/kproto_results/kproto_z_score.rds"

  # Outputs
  out_dir: "/home/exacloud/gscratch/NagelLab/staff/sam/projects/ABCD_MDS_Risk/results/main_analysis/1_clustering/7_imputation_kproto/"

  # Missingness inclusion rule for imputation
  # Require at least this proportion of the 17 risk variables observed for a row to be imputable.
  # (Set to 0 to impute everyone; set higher to be more conservative.)
  min_obs_prop: 0.50

  # Imputation settings
  do_missForest: true
  missForest_seed: 123
  missForest_maxiter: 10
  missForest_ntree: 200

  do_mice: true
  mice_seed: 123
  mice_m: 20
  mice_maxit: 20

  # Clustering settings post-imputation
  k_grid: [2,3,4,5,6,7,8]
  nstart: 8

  # Lambda handling
  lambda_mode: "estimate"   # "estimate" | "fixed"
  lambda_fixed: 2.5

  # Bootstrap stability for each k
  boot_B: 250
  boot_seed: 123

  # Misc
  keep_model_data: true
---

```{r global, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, message = TRUE, warning = TRUE, results = "markup", verbose = TRUE, comment = "")
```

## 7.1 Setup

```{r setup, include = FALSE, echo = FALSE, warning = FALSE}

## 7.1 Setup ##

cat("7.1 Setup beginning\n")

#7.1.1 Resolve REPO + out_dir safely since YAML can't reference params$REPO via !r
REPO <- Sys.getenv("REPO", unset = params$REPO)

if (!nzchar(params$out_dir)) {
  params$out_dir <- file.path(REPO, "results", "main_analysis", "1_clustering", "7_imputation_kproto")
}

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(readr)
  library(glue)
  library(purrr)
  library(tibble)
  library(data.table)
  library(ggplot2)
  library(naniar)      # mcar_test + missingness visuals
  library(clustMixType)
  library(cluster)
  library(clue)
  library(mclust)
  library(missForest)  # primary imputation
  library(mice)        # sensitivity imputation
})

#7.1.2 Output dirs
paths <- list(
  out_dir = params$out_dir,
  fig_dir = file.path(params$out_dir, "figures"),
  tab_dir = file.path(params$out_dir, "tables"),
  rds_dir = file.path(params$out_dir, "rds"))

dir.create(paths$out_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$fig_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$tab_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$rds_dir, showWarnings = FALSE, recursive = TRUE)

#7.1.3 Safe writers
write_csv2 <- function(x, path) {
  dir.create(dirname(path), TRUE, TRUE)
  readr::write_csv(x, path)
}
save_rds2 <- function(x, path) {
  dir.create(dirname(path), TRUE, TRUE)
  saveRDS(x, path)
}

cat("7.1 Setup complete\n")

```

## 7.2 Load Anchor + Define Canonical 17-Feature Set

```{r load anchor, echo = FALSE, warning = FALSE}

## 7.2 Load anchor clustering object and recover canonical feature set ##

#7.2.1 Load kp obj into env
kp_path <- params$kproto_anchor_rds
if (!file.exists(kp_path)) stop(glue("Anchor kproto RDS not found: {kp_path}"))

kp_any <- readRDS(kp_path)


#7.2.2 Select & parse target kproto solution
pick_kproto <- function(obj, k_target) {
  if (inherits(obj, "kproto")) return(obj)
  stopifnot(is.list(obj))
  nm <- names(obj)
  if (!is.null(nm)) {
    hit <- paste0("k", k_target)
    if (hit %in% nm && inherits(obj[[hit]], "kproto")) return(obj[[hit]])
  }
  for (el in obj) {
    if (inherits(el, "kproto")) {
      sz <- tryCatch(el$size, error = function(...) NULL)
      if (!is.null(sz) && length(sz) == k_target) return(el)
    }
  }
  stop(glue("Could not locate kproto object for k={k_target} in {kp_path}"))
}

kp2 <- pick_kproto(kp_any, 2)

#7.2.3 Pull data from target kproto solution
X_fit <- kp2$data
if (is.null(X_fit) || nrow(X_fit) == 0) stop("kp2$data is missing/empty; cannot recover canonical feature frame.")

risk_anchor_df <- as.data.frame(X_fit)
risk_anchor_dt <- as.data.table(risk_anchor_df)
risk_features_17 <- names(risk_anchor_dt)

cat(glue("Recovered p={length(risk_features_17)} canonical clustering features from anchor.\n"))

#7.2.4 Baseline anchor labels (k=2)
anchor_lbl <- if (!is.null(kp2$cluster)) kp2$cluster else {
  pr <- predict(kp2, newdata = as.data.frame(risk_anchor_dt))
  if (is.list(pr)) pr$cluster else pr
}
stopifnot(length(anchor_lbl) == nrow(risk_anchor_dt))

#7.2.5 Ensure participant_id is stored in rownames of kp2$data
anchor_id <- rownames(risk_anchor_df)
if (is.null(anchor_id) || anyNA(anchor_id) || length(anchor_id) != nrow(risk_anchor_df)) {
  stop("Could not recover participant_id from rownames(kp2$data). Please confirm rownames are participant IDs.")
}

#7.2.6 Store anchor clustering IDs
anchor_id <- as.character(anchor_id)

```

## 7.3 Load Unfiltered Baseline Data + Align to Canonical 17

```{r load data, echo = FALSE, warning = FALSE}

## 7.3 Load unfiltered baseline risk frame (full baseline N) ##

#7.3.1 Load in unfiltered baseline risk df to be used in imputation
risk_path <- params$risk_unfiltered_path
if (!file.exists(risk_path)) stop(glue("Unfiltered risk CSV not found: {risk_path}"))

risk_df_raw <- readr::read_csv(risk_path, show_col_types = FALSE)
risk_dt_raw <- as.data.table(risk_df_raw)

if (!("participant_id" %in% names(risk_dt_raw))) {
  stop("Unfiltered risk frame does not include participant_id; needed for alignment and downstream merges.")
}

#7.3.2 Restrict to canonical 17 features + participant_id
missing_feats <- setdiff(risk_features_17, names(risk_dt_raw))
if (length(missing_feats)) {
  stop(glue("Unfiltered frame is missing canonical features: {paste(missing_feats, collapse=', ')}"))
}

risk_dt <- risk_dt_raw[, c("participant_id", risk_features_17), with = FALSE]
risk_dt[, .row_id := .I]

cat(glue("Loaded unfiltered baseline frame: N={nrow(risk_dt)}; p={length(risk_features_17)}\n"))

```

## 7.4 Missingness Diagnostics (MCAR-oriented summary + pattern checks)

```{r missingness, echo = FALSE, warning = FALSE}

## 7.4 Missingness diagnostics ##

X0 <- as.data.frame(risk_dt[, risk_features_17, with = FALSE])

#7.4.1 Per-variable missingness
miss_var <- tibble::tibble(
  feature = names(X0),
  n_miss = vapply(X0, function(x) sum(is.na(x)), numeric(1)),
  pct_miss = n_miss / nrow(X0)) %>% arrange(desc(pct_miss))

write_csv2(miss_var, file.path(paths$tab_dir, "missingness_by_variable.csv"))
knitr::kable(head(miss_var, 25), digits = 3, caption = "Top missingness (variables)") %>%
  kableExtra::kable_styling(full_width = FALSE)

#7.4.2 Per-row observed proportion across the 17 features
obs_prop <- rowMeans(!is.na(X0))
risk_dt[, obs_prop := obs_prop]

miss_row <- tibble::tibble(
  obs_prop = obs_prop) %>%
  summarise(
    n = n(),
    min = min(obs_prop),
    p10 = quantile(obs_prop, 0.10),
    p25 = quantile(obs_prop, 0.25),
    median = median(obs_prop),
    p75 = quantile(obs_prop, 0.75),
    p90 = quantile(obs_prop, 0.90),
    max = max(obs_prop))

write_csv2(miss_row, file.path(paths$tab_dir, "missingness_row_obsprop_summary.csv"))
knitr::kable(miss_row, digits = 3, caption = "Row-level observed proportion summary") %>%
  kableExtra::kable_styling(full_width = FALSE)

#7.4.3 Visuals: missingness overview
p_miss_var <- ggplot(miss_var, aes(x = reorder(feature, pct_miss), y = pct_miss)) +
  geom_col() +
  coord_flip() +
  theme_classic(base_size = 12) +
  labs(title = "Missingness by risk feature", x = NULL, y = "Proportion missing")

ggsave(file.path(paths$fig_dir, "missingness_by_variable.png"),
       p_miss_var, width = 7.2, height = 6.2, dpi = 600, bg = "white")

#7.4.4 Missingness pattern heatmap (subset for readability if huge N)
set.seed(1)
idx_plot <- if (nrow(X0) > 12000) sample(seq_len(nrow(X0)), 12000) else seq_len(nrow(X0))
p_vis <- naniar::vis_miss(X0[idx_plot, , drop = FALSE]) +
  theme_classic(base_size = 12) +
  labs(title = "Missingness heatmap (sampled rows)")

ggsave(file.path(paths$fig_dir, "missingness_heatmap_vis_miss.png"),
       p_vis, width = 9.0, height = 5.2, dpi = 600, bg = "white")

#7.4.5 Little's MCAR test (most interpretable on numeric-only)
is_num <- vapply(X0, is.numeric, logical(1))
X_num <- X0[, is_num, drop = FALSE]

mcar_out <- NULL
if (ncol(X_num) >= 2) {
  mcar_out <- tryCatch(naniar::mcar_test(X_num), error = function(e) NULL)
}

if (!is.null(mcar_out)) {

  write_csv2(as.data.frame(mcar_out), file.path(paths$tab_dir, "little_mcar_test_numeric_only.csv"))
  knitr::kable(mcar_out, digits = 3, caption = "Little's MCAR test (numeric features only)") %>%
    kableExtra::kable_styling(full_width = FALSE)
} else {
  cat("Little MCAR test not run (need >=2 numeric columns and successful computation).\n")
}

#7.4.6 Practical MAR-style check: do missingness indicators relate to other observed variables?
#     For each feature f, regress I(missing_f) on all other features (simple GLM on model.matrix).
#     This is an exploratory screen and not a definitive MAR test
make_mm <- function(df) {
  mm <- model.matrix(~ . - 1, data = df)
  mm[, apply(mm, 2, function(x) stats::sd(x) > 0), drop = FALSE]
}

X_mm <- make_mm(X0)

glm_screen <- lapply(names(X0), function(f) {

  y <- as.integer(is.na(X0[[f]]))
  if (all(y == 0) || all(y == 1)) return(NULL)

  #7.4.6.1 Predictors exclude the target feature's own columns as best-effort: we drop raw f from df and rebuild mm
  df_pred <- X0[, setdiff(names(X0), f), drop = FALSE]
  mm_pred <- make_mm(df_pred)

  #7.4.6.2 Keep it light; use a ridge-like workaround by limiting predictors if enormous
  if (ncol(mm_pred) > 200) {
    set.seed(1)
    keep <- sample(seq_len(ncol(mm_pred)), 200)
    mm_pred <- mm_pred[, keep, drop = FALSE]
  }

  fit <- tryCatch(glm(y ~ mm_pred, family = binomial()), error = function(e) NULL)
  if (is.null(fit)) return(NULL)

  #7.4.6.3 Omnibus LRT vs intercept-only
  fit0 <- glm(y ~ 1, family = binomial())
  lr <- tryCatch(anova(fit0, fit, test = "Chisq"), error = function(e) NULL)
  if (is.null(lr) || nrow(lr) < 2) return(NULL)

  tibble::tibble(
    feature = f,
    n_miss = sum(y == 1),
    pct_miss = mean(y == 1),
    df = lr$Df[2],
    chisq = lr$Deviance[2],
    p = lr$`Pr(>Chi)`[2])
}) %>% bind_rows()

if (nrow(glm_screen) > 0) {
  glm_screen <- glm_screen %>%
    mutate(p_fdr = p.adjust(p, method = "fdr")) %>%
    arrange(p_fdr)

  write_csv2(glm_screen, file.path(paths$tab_dir, "missingness_indicator_glm_screen.csv"))
  knitr::kable(head(glm_screen, 25), digits = 3, caption = "Missingness-indicator GLM screen (exploratory)") %>%
    kableExtra::kable_styling(full_width = FALSE)
}

```

## 7.5 Helpers: Types, Z-scoring, Lambda, Stability, Fit Loop

```{r helper functions, echo = FALSE, warning = FALSE}

## 7.5 Helper functions ##

#7.5.1 Force known categorical risk variables to factor
force_known_categoricals <- function(df) {
  df <- as.data.frame(df)

  known_cat <- intersect(
    c("family_history_depression", "family_history_mania", "bullying"),
    names(df))

  for (nm in known_cat) {
    df[[nm]] <- factor(df[[nm]])
  }

  df
}

#7.5.2 Coerce types: numeric stays numeric; character/logical -> factor; small-int -> factor
coerce_types <- function(df) {
  df <- as.data.frame(df)
  for (nm in names(df)) {
    x <- df[[nm]]
    if (is.numeric(x)) next
    if (is.logical(x)) { df[[nm]] <- factor(x, levels = c(FALSE, TRUE)); next }
    if (is.character(x)) { df[[nm]] <- factor(x); next }
    if (is.integer(x)) {
      u <- unique(x[!is.na(x)])
      if (length(u) <= 10) df[[nm]] <- factor(x) else df[[nm]] <- as.numeric(x)
      next
    }
    if (!is.factor(x)) df[[nm]] <- factor(x)
  }
  df
}

#7.5.3 Z-score numeric columns
zscore_numeric <- function(df) {
  df <- as.data.frame(df)
  is_num <- vapply(df, is.numeric, logical(1))
  for (nm in names(df)[is_num]) {
    x <- df[[nm]]
    mu <- mean(x, na.rm = TRUE)
    sdv <- stats::sd(x, na.rm = TRUE)
    if (is.finite(sdv) && sdv > 0) df[[nm]] <- (x - mu) / sdv else df[[nm]] <- 0
  }
  df
}

#7.5.4 Ensure >=1 numeric and >=1 factor column for kproto
count_types <- function(df) {
  df <- as.data.frame(df)
  is_num <- vapply(df, is.numeric, logical(1))
  is_fac <- vapply(df, is.factor,  logical(1))
  list(n_num = sum(is_num), n_fac = sum(is_fac), is_num = is_num, is_fac = is_fac)
}

prep_for_kproto <- function(df) {
  df <- as.data.frame(df)
  ct <- count_types(df)
  add_num <- (ct$n_num == 0L)
  add_fac <- (ct$n_fac == 0L)
  if (add_num) df$.dummy_num <- 0
  if (add_fac) df$.dummy_fac <- factor("A")
  list(x = df, added_dummy = (add_num || add_fac), ct = ct)
}

#7.5.5 Lambda rule
lambda_safe <- function(X_orig, lambda_mode, lambda_fixed = 2.5) {
  if (identical(lambda_mode, "fixed")) return(list(lambda = as.numeric(lambda_fixed), reason = "fixed"))
  ct <- count_types(X_orig)
  if (ncol(X_orig) < 2L) return(list(lambda = 1, reason = "p<2: lambdaest undefined; lambda irrelevant"))
  if (ct$n_num == 0L || ct$n_fac == 0L) return(list(lambda = 1, reason = "single-type set: lambda irrelevant"))
  lam <- clustMixType::lambdaest(X_orig, verbose = FALSE)
  list(lambda = lam, reason = "lambdaest (mixed-type set)")
}

#7.5.6 Objective + silhouette
kproto_objective <- function(kp) {
  for (nm in c("tot.withinss","tot.withindist","tot.within.dist","tot.within")) {
    if (!is.null(kp[[nm]]) && is.finite(kp[[nm]])) return(as.numeric(kp[[nm]]))
  }
  NA_real_
}

extract_validation_index <- function(x) {
  if (is.numeric(x) && length(x) == 1L) return(as.numeric(x))
  if (is.list(x) && "index" %in% names(x)) return(as.numeric(x$index))
  if (is.list(x) && "index_opt" %in% names(x)) return(as.numeric(x$index_opt))
  NA_real_
}

silhouette_validation_kproto <- function(kp, type = "huang") {
  out <- tryCatch(
    clustMixType::validation_kproto(method = "silhouette", object = kp, type = type),
    error = function(e) NA)
  extract_validation_index(out)
}

#7.5.7 Clean kproto object for stability_kproto()
clean_kproto_for_stability <- function(kp) {
  kp2 <- kp

  #7.5.7.1 Ensure data is a plain data.frame, no rownames
  kp2$data <- as.data.frame(kp2$data)
  rownames(kp2$data) <- NULL

  #7.5.7.2 Ensure cluster is a plain integer vector, no names
  kp2$cluster <- as.integer(kp2$cluster)
  names(kp2$cluster) <- NULL

  #7.5.7.3 Ensure lambda is scalar numeric
  if (!is.null(kp2$lambda)) kp2$lambda <- as.numeric(kp2$lambda)[1]

  kp2
}

```

## 7.6 Build Imputation Frame

```{r build imputation data, echo = FALSE, warning = FALSE}

## 7.6 Define which rows are imputable ##

#7.6.1 Apply inclusion rule for imputation clustering; rows with too few observed values are excluded from imputation + downstream
min_obs_prop <- params$min_obs_prop
risk_dt[, imputable := (obs_prop >= min_obs_prop)]

tab_imputable <- risk_dt[, .(
  N_total = .N,
  N_imputable = sum(imputable),
  pct_imputable = mean(imputable),
  min_obs_prop = min_obs_prop)]

write_csv2(as.data.frame(tab_imputable), file.path(paths$tab_dir, "imputation_inclusion_summary.csv"))
knitr::kable(tab_imputable, digits = 3, caption = "Imputation inclusion summary") %>%
  kableExtra::kable_styling(full_width = FALSE)

risk_imp_dt <- risk_dt[imputable == TRUE]
X_imp0 <- as.data.frame(risk_imp_dt[, risk_features_17, with = FALSE])

```

## 7.7 Imputation 

### 7.7.1 Primary missForest Imputation

```{r missForest, echo = FALSE, warning = FALSE}

## 7.7.1 missForest Primary Imputation ##

#7.7.1.1 Store empty object to store results
imp_mf <- NULL

#7.7.1.2 Dynamically run missForest imputation
if (isTRUE(params$do_missForest)) {

  cat("Running missForest...\n")

  #7.7.1.2.1 Prep data
  X_mf <- X_imp0 
  X_mf <- force_known_categoricals(X_mf) 
  X_mf <- coerce_types(X_mf)

  #7.7.1.2.2 Set seed for reproducibility
  set.seed(params$missForest_seed)
  
  #7.7.1.2.3 Run missForest model on data
  imp_mf <- missForest::missForest(
    xmis = X_mf,
    maxiter = params$missForest_maxiter,
    ntree = params$missForest_ntree,
    verbose = TRUE)

  #7.7.1.2.4 Store completed data
  X_mf_comp <- imp_mf$ximp

  #7.7.1.2.5 Post-imputation: z-score numeric to match kproto z-score convention
  X_mf_comp <- zscore_numeric(X_mf_comp)

  #7.7.1.2.6 Save output
  save_rds2(imp_mf, file.path(paths$rds_dir, "imputation_missForest_object.rds"))
  write_csv2(
    tibble::tibble(
      method = "missForest",
      OOBerror = if (!is.null(imp_mf$OOBerror)) imp_mf$OOBerror else NA_real_),
    file.path(paths$tab_dir, "imputation_missForest_oob.csv"))
}

```

### 7.7.2 MICE Sensitivity

```{r MICE, echo = FALSE, warning = FALSE}

## 7.7.2 MICE Sensitivity Imputation ##

#7.7.2.1 Create objects to store MICE results
imp_mice <- NULL
X_mice_comp <- NULL

#7.7.2.2 Dynamically run MICE
if (isTRUE(params$do_mice)) {

  cat("Running mice...\n")

  #7.7.2.2.1 Prep data for MICE
  X_mi <- X_imp0
  X_mi <- force_known_categoricals(X_mi)
  X_mi <- coerce_types(X_mi)

  #7.7.2.2.2 Define mice methods by type
  meth <- mice::make.method(X_mi)

  is_num <- vapply(X_mi, is.numeric, logical(1))
  is_fac <- vapply(X_mi, is.factor, logical(1))

  #7.7.2.2.3 For numeric vars, run predictive mean matching
  meth[is_num] <- "pmm"

  #7.7.2.2.4 For factor vars binary logistic if 2 levels; else polyreg
  for (nm in names(X_mi)[is_fac]) {
    k <- nlevels(X_mi[[nm]])
    if (k == 2) meth[nm] <- "logreg" else meth[nm] <- "polyreg"
  }

  #7.7.2.2.5 Run MICE
  set.seed(params$mice_seed)
  imp_mice <- mice::mice(
    data = X_mi,
    m = params$mice_m,
    maxit = params$mice_maxit,
    method = meth,
    printFlag = TRUE)

  #7.7.2.2.6 Use first completed dataset for clustering sensitivity
  X_mice_comp <- mice::complete(imp_mice, action = 1)
  X_mice_comp <- zscore_numeric(X_mice_comp)

  save_rds2(imp_mice, file.path(paths$rds_dir, "imputation_mice_mids.rds"))

  ##7.7.2.2.7 Basic convergence plot artifact
  png(file.path(paths$fig_dir, "mice_traceplot.png"), width = 1400, height = 900, res = 180)
  plot(imp_mice)
  dev.off()
}

```

## 7.8 Refit k-prototypes (k=2:8) on Imputed Data + Compare to Baseline Anchor

```{r refit kproto, echo = FALSE, warning = FALSE}

## 7.8 k-prototypes grid on imputed datasets ##

run_k_grid <- function(X_completed, label, participant_id_vec) {

  #7.8.1 Lambda
  lam_info <- lambda_safe(
    X_completed,
    lambda_mode = params$lambda_mode,
    lambda_fixed = params$lambda_fixed)

  #7.8.2 Prep input (dummy cols if needed)
  prep <- prep_for_kproto(X_completed)
  X_fit <- prep$x

  #7.8.3 Fit across k
  res_list <- list()
  model_list <- list()
  assign_list <- list()

  #7.8.3 Overlap with anchor IDs for ARI comparison
  pid <- as.character(participant_id_vec)
  in_anchor <- pid %in% anchor_id
  subj_overlap <- pid[in_anchor]
  anchor_lbl_overlap <- anchor_lbl[match(subj_overlap, anchor_id)]

  for (k in params$k_grid) {

    set.seed(7000 + k)

    kp <- clustMixType::kproto(
      x = X_fit,
      k = as.integer(k),
      lambda = lam_info$lambda,
      nstart = as.integer(params$nstart),
      verbose = FALSE)

    lbl <- kp$cluster
    k_obs <- length(unique(lbl))
    min_n <- min(tabulate(lbl))

    sil <- if (k_obs >= 2 && min_n >= 2) silhouette_validation_kproto(kp, type = "huang") else NA_real_
    obj <- kproto_objective(kp)

    #7.8.3.1 Stability via stability_kproto: Rand + Jaccard
    if (k_obs >= 2 && min_n >= 2) {
    
      kp_stab <- clean_kproto_for_stability(kp)
    
      set.seed(params$boot_seed + 100L * k)
    
      stab_formals <- names(formals(clustMixType::stability_kproto))
      stab_args <- list(
        object = kp_stab,
        method = c("rand", "jaccard"),
        nstart = as.integer(params$nstart),
        verbose = FALSE)
    
      if ("B" %in% stab_formals) {
        stab_args$B <- as.integer(params$boot_B)
      } else if ("nboot" %in% stab_formals) {
        stab_args$nboot <- as.integer(params$boot_B)
      }
    
      stab_raw <- tryCatch(
        do.call(clustMixType::stability_kproto, stab_args),
        error = function(e) {
          stop(glue("stability_kproto() failed at {label}, k={k}: {conditionMessage(e)}"))
        }
      )
    
      stab_overall <- stab_raw[[1]]
      stab_tbl <- as.data.frame(stab_raw[[2]])
      colnames(stab_tbl) <- c("B", "Rand", "Jaccard")
    
    } else {
      stab_overall <- c(R = NA_real_, J = NA_real_)
      stab_tbl <- data.frame(B = integer(0), Rand = numeric(0), Jaccard = numeric(0))
    }

    #7.8.3.2 Agreement with baseline anchor (ARI) on the overlapping subject set
    ari_vs_anchor <- NA_real_
    if (sum(in_anchor) > 10) {
      idx_overlap <- match(subj_overlap, pid)
      lbl_overlap <- lbl[idx_overlap]
      ari_vs_anchor <- mclust::adjustedRandIndex(lbl_overlap, anchor_lbl_overlap)
    }

    res_list[[as.character(k)]] <- tibble::tibble(
      imputation = label,
      k = as.integer(k),
      k_observed = as.integer(k_obs),
      min_cluster_n = as.integer(min_n),
      lambda = as.numeric(lam_info$lambda),
      lambda_reason = lam_info$reason,
      added_dummy = prep$added_dummy,
      objective = as.numeric(obj),
      silhouette_huang = as.numeric(sil),
      rand_overall = as.numeric(stab_overall["R"]),
      jacc_overall = as.numeric(stab_overall["J"]),
      rand_mean = mean(stab_tbl$Rand, na.rm = TRUE),
      jacc_mean = mean(stab_tbl$Jaccard, na.rm = TRUE),
      rand_sd = sd(stab_tbl$Rand, na.rm = TRUE),
      jacc_sd = sd(stab_tbl$Jaccard, na.rm = TRUE),,
      ari_vs_anchor_k2 = as.numeric(ari_vs_anchor),
      n_overlap_anchor = as.integer(sum(in_anchor)))

    #7.8.3.3 Assignments
    assign_list[[as.character(k)]] <- tibble::tibble(
      participant_id = participant_id_vec,
      imputation = label,
      k = as.integer(k),
      cluster = as.integer(lbl),
      in_anchor = in_anchor)

    if (!isTRUE(params$keep_model_data)) kp$data <- NULL
    model_list[[as.character(k)]] <- kp
  }

  res_df <- bind_rows(res_list) %>% arrange(k)
  assign_df <- bind_rows(assign_list)

  #7.8.4 Write artifacts
  write_csv2(res_df, file.path(paths$tab_dir, glue("kgrid_metrics_{label}.csv")))
  write_csv2(assign_df, file.path(paths$tab_dir, glue("kgrid_assignments_{label}.csv")))
  save_rds2(model_list, file.path(paths$rds_dir, glue("kgrid_models_{label}.rds")))

  #7.8.5 Plots
  p1 <- ggplot(res_df, aes(x = k, y = silhouette_huang)) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1.8) +
    theme_classic(base_size = 12) +
    labs(
      title = glue("Silhouette across k ({label})"),
      subtitle = glue("lambda_mode={params$lambda_mode}; lambda={round(res_df$lambda[1], 3)}; nstart={params$nstart}"),
      x = "k", y = "Silhouette (Huang)")
  ggsave(file.path(paths$fig_dir, glue("silhouette_by_k_{label}.png")),
         p1, width = 7, height = 4.2, dpi = 600, bg = "white")

  p2 <- ggplot(res_df, aes(x = k, y = jaccard_mean)) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1.8) +
    theme_classic(base_size = 12) +
    labs(
      title = glue("Bootstrap stability across k ({label})"),
      subtitle = glue("Metric = min clusterwise Jaccard; B={params$boot_B}"),
      x = "k", y = "min Jaccard")
  ggsave(file.path(paths$fig_dir, glue("jaccard_mean_by_k_{label}.png")),
         p2, width = 7, height = 4.2, dpi = 600, bg = "white")

  p3 <- ggplot(res_df, aes(x = k, y = ari_vs_anchor_k2)) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1.8) +
    theme_classic(base_size = 12) +
    labs(
      title = glue("Agreement vs baseline anchor labels across k ({label})"),
      subtitle = "ARI computed on subjects overlapping the original anchor clustering frame",
      x = "k", y = "ARI vs anchor labels")
  ggsave(file.path(paths$fig_dir, glue("ari_vs_anchor_by_k_{label}.png")),
         p3, width = 7, height = 4.2, dpi = 600, bg = "white")

  list(metrics = res_df, assignments = assign_df)
}

#7.8.6 Run missForest & MICE k grids
runs <- list()

if (isTRUE(params$do_missForest) && !is.null(imp_mf)) {
  runs$missForest <- run_k_grid(
    X_completed = X_mf_comp,
    label = "missForest",
    participant_id_vec = risk_imp_dt$participant_id)
}

if (isTRUE(params$do_mice) && !is.null(imp_mice)) {
  runs$mice <- run_k_grid(
    X_completed = X_mice_comp,
    label = "mice",
    participant_id_vec = risk_imp_dt$participant_id)
}

```

## 7.9 Cross-Method Summary Table

```{r summary, echo = FALSE, warning = FALSE}

## 7.9 Summary Table ##

if (length(runs) > 0) {

  sum_df <- bind_rows(lapply(runs, function(x) x$metrics)) %>%
    arrange(imputation, k)

  write_csv2(sum_df, file.path(paths$tab_dir, "kgrid_metrics_all_imputation_methods.csv"))

  #7.9.1 Recommendation rule: maximize silhouette, tie-break by higher jaccard_mean, then smaller k
  best_df <- sum_df %>%
    filter(is.finite(silhouette_huang)) %>%
    group_by(imputation) %>%
    arrange(desc(silhouette_huang), desc(jacc_mean), k) %>%
    slice(1) %>%
    ungroup()

  write_csv2(best_df, file.path(paths$tab_dir, "k_recommendation_per_imputation_method.csv"))

  knitr::kable(best_df, digits = 3, caption = "Recommended k per imputation method (max silhouette; tie-break by mean bootstrap Jaccard; then smaller k)") %>%
    kableExtra::kable_styling(full_width = FALSE)
}

```

## 7.10 Reproducibility Log

```{r reproducibility log, echo = FALSE, warning = FALSE}

## 7.10 Store Reproducibility Log ##

log <- list(
  timestamp = as.character(Sys.time()),
  REPO = params$REPO,
  risk_unfiltered_path = params$risk_unfiltered_path,
  kproto_anchor_rds = params$kproto_anchor_rds,
  risk_features_17 = risk_features_17,
  min_obs_prop = params$min_obs_prop,
  missForest = list(
    run = params$do_missForest,
    seed = params$missForest_seed,
    maxiter = params$missForest_maxiter,
    ntree = params$missForest_ntree),
  mice = list(
    run = params$do_mice,
    seed = params$mice_seed,
    m = params$mice_m,
    maxit = params$mice_maxit),
  clustering = list(
    k_grid = params$k_grid,
    nstart = params$nstart,
    lambda_mode = params$lambda_mode,
    lambda_fixed = params$lambda_fixed,
    boot_B = params$boot_B,
    boot_seed = params$boot_seed))

writeLines(jsonlite::toJSON(log, pretty = TRUE), con = file.path(paths$out_dir, "run_log.json"))

```
