---
title: "8. Baseline Mixed-Data Clustering Benchmark: k-prototypes vs KAMILA vs clustMD"
author: "Sam A. Sievertsen"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    df_print: paged
params:
  # I/O
  REPO: "/home/exacloud/gscratch/NagelLab/staff/sam/projects/ABCD_MDS_Risk"
  risk_data_path: "/home/exacloud/gscratch/NagelLab/staff/sam/projects/ABCD_MDS_Risk/data/data_processed/risk_variable_data.csv"
  kproto_anchor_rds: "/home/exacloud/gscratch/NagelLab/staff/sam/projects/ABCD_MDS_Risk/data/data_processed/kproto_results/kproto_z_score.rds"
  out_dir: "/home/exacloud/gscratch/NagelLab/staff/sam/projects/ABCD_MDS_Risk/results/main_analysis/1_clustering/8_algorithm_benchmark"

  # Search grid
  k_grid: [2,3,4,5,6,7,8]

  # Light-weight optimization settings
  nstart: 8
  seed: 123

  # k-prototypes lambda handling
  lambda_mode: "estimate"   # "estimate" | "fixed"
  lambda_fixed: 2.5

  # KAMILA settings
  kamila_num_init: 8
  kamila_num_pred_strt: 25
  kamila_max_iter: 100

  # KAMILA algorithm-native k selection
  run_kamila_ps: true
  kamila_ps_threshold: 0.80
  kamila_ps_cv_runs: 10

  # clustMD settings
  run_clustMD: true
  clustmd_model: "EII"
  clustmd_max_iter: 200
  clustmd_nnorms: 2000

  # UMAP settings
  umap_neighbors: 15
  umap_min_dist: 0.1
  umap_seed: 123

  # Misc
  keep_model_data: true
---

```{r global, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, message = TRUE, warning = TRUE, results = "markup", verbose = TRUE, comment = "")
```

## 8.1 Setup

```{r setup, include = FALSE, echo = FALSE, warning = FALSE}

## 8.1 Setup ##

cat("8.1 Setup beginning\n")

REPO <- Sys.getenv("REPO", unset = params$REPO)
OUT_DIR <- if (nzchar(params$out_dir)) params$out_dir else {
  file.path(REPO, "results", "main_analysis", "1_clustering", "8_algorithm_benchmark")
}

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(readr)
  library(glue)
  library(purrr)
  library(tibble)
  library(data.table)
  library(ggplot2)
  library(cluster)
  library(mclust)
  library(clue)
  library(uwot)
  library(kableExtra)
  library(jsonlite)
  library(clustMixType)
  library(kamila)
  library(clustMD)
})

paths <- list(
  out_dir = OUT_DIR,
  fig_dir = file.path(OUT_DIR, "figures"),
  tab_dir = file.path(OUT_DIR, "tables"),
  rds_dir = file.path(OUT_DIR, "rds"))

dir.create(paths$out_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(paths$fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(paths$tab_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(paths$rds_dir, recursive = TRUE, showWarnings = FALSE)

write_csv2 <- function(x, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  readr::write_csv(x, path)
}

save_rds2 <- function(x, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  saveRDS(x, path)
}

set.seed(params$seed)

cat("8.1 Setup complete\n")

```

## 8.2 Load Data + Recover Canonical 17-Feature Baseline Frame

```{r load data, echo = FALSE, warning = FALSE}

## 8.2 Load canonical clustering frame ##

risk_path <- params$risk_data_path
if (!file.exists(risk_path)) stop(glue("Risk data CSV not found: {risk_path}"))

risk_df_raw <- readr::read_csv(risk_path, show_col_types = FALSE)
risk_dt_raw <- as.data.table(risk_df_raw)

if (!("participant_id" %in% names(risk_dt_raw))) {
  stop("participant_id not found in risk data.")
}

#8.2.1 Load anchor k-prototypes object and recover canonical feature names
kp_any <- readRDS(params$kproto_anchor_rds)

pick_kproto <- function(obj, k_target) {
  if (inherits(obj, "kproto")) return(obj)
  stopifnot(is.list(obj))
  nm <- names(obj)
  if (!is.null(nm)) {
    hit <- paste0("k", k_target)
    if (hit %in% nm && inherits(obj[[hit]], "kproto")) return(obj[[hit]])
  }
  for (el in obj) {
    if (inherits(el, "kproto")) {
      sz <- tryCatch(el$size, error = function(...) NULL)
      if (!is.null(sz) && length(sz) == k_target) return(el)
    }
  }
  stop(glue("Could not locate kproto object for k={k_target}."))
}

kp2_anchor <- pick_kproto(kp_any, 2)
anchor_df <- as.data.frame(kp2_anchor$data)
anchor_features <- names(anchor_df)

#8.2.2 Restrict baseline frame to participant_id + canonical clustering features
missing_feats <- setdiff(anchor_features, names(risk_dt_raw))
if (length(missing_feats) > 0) {
  stop(glue("Risk data missing canonical features: {paste(missing_feats, collapse=', ')}"))
}

risk_dt <- risk_dt_raw[, c("participant_id", anchor_features), with = FALSE]
risk_dt[, .row_id := .I]

cat(glue("Loaded N={nrow(risk_dt)} rows and p={length(anchor_features)} canonical clustering features.\n"))

#8.2.3 Recover anchor labels for ARI comparison
anchor_lbl <- if (!is.null(kp2_anchor$cluster)) kp2_anchor$cluster else {
  pr <- predict(kp2_anchor, newdata = as.data.frame(kp2_anchor$data))
  if (is.list(pr)) pr$cluster else pr
}

anchor_id <- rownames(anchor_df)
if (is.null(anchor_id) || length(anchor_id) != nrow(anchor_df)) {
  stop("Could not recover participant IDs from rownames(anchor_df).")
}

anchor_id <- as.character(anchor_id)

```

## 8.3 Helper Functions for Type Handling + Common Validation

```{r helper functions, echo = FALSE, warning = FALSE}

## 8.3 Helper functions ##

#8.3.1 Force known categorical variables to factor
force_known_categoricals <- function(df) {
  df <- as.data.frame(df)

  known_cat <- intersect(
    c("family_history_depression", "family_history_mania", "bullying"),
    names(df))

  for (nm in known_cat) {
    df[[nm]] <- factor(df[[nm]])
  }

  df
}

#8.3.2 Coerce remaining variable classes
coerce_types <- function(df) {
  df <- as.data.frame(df)

  for (nm in names(df)) {
    x <- df[[nm]]

    if (is.numeric(x)) next

    if (is.logical(x)) {
      df[[nm]] <- factor(x, levels = c(FALSE, TRUE))
      next
    }

    if (is.character(x)) {
      df[[nm]] <- factor(x)
      next
    }

    if (is.integer(x)) {
      u <- unique(x[!is.na(x)])
      if (length(u) <= 10) {
        df[[nm]] <- factor(x)
      } else {
        df[[nm]] <- as.numeric(x)
      }
      next
    }

    if (!is.factor(x)) df[[nm]] <- factor(x)
  }

  df
}

#8.3.3 Z-score numeric columns
zscore_numeric <- function(df) {
  df <- as.data.frame(df)
  is_num <- vapply(df, is.numeric, logical(1))

  for (nm in names(df)[is_num]) {
    x <- df[[nm]]
    mu <- mean(x, na.rm = TRUE)
    sdv <- stats::sd(x, na.rm = TRUE)

    if (is.finite(sdv) && sdv > 0) {
      df[[nm]] <- (x - mu) / sdv
    } else {
      df[[nm]] <- 0
    }
  }

  df
}

#8.3.4 Count numeric / factor columns
count_types <- function(df) {
  df <- as.data.frame(df)
  is_num <- vapply(df, is.numeric, logical(1))
  is_fac <- vapply(df, is.factor, logical(1))
  list(
    n_num = sum(is_num),
    n_fac = sum(is_fac),
    is_num = is_num,
    is_fac = is_fac
  )
}

#8.3.5 Safe lambda rule for k-prototypes
lambda_safe <- function(X_orig, lambda_mode, lambda_fixed = 2.5) {
  if (identical(lambda_mode, "fixed")) {
    return(list(lambda = as.numeric(lambda_fixed), reason = "fixed"))
  }

  ct <- count_types(X_orig)
  if (ncol(X_orig) < 2L) {
    return(list(lambda = 1, reason = "p<2: lambdaest undefined; lambda irrelevant"))
  }
  if (ct$n_num == 0L || ct$n_fac == 0L) {
    return(list(lambda = 1, reason = "single-type set: lambda irrelevant"))
  }

  lam <- clustMixType::lambdaest(X_orig, verbose = FALSE)
  list(lambda = lam, reason = "lambdaest (mixed-type set)")
}

#8.3.6 Common Gower distance and silhouette
gower_silhouette_mean <- function(df, lbl) {
  df <- as.data.frame(df)
  lbl <- as.integer(lbl)

  if (length(unique(lbl)) < 2L) return(NA_real_)
  if (min(tabulate(lbl)) < 2L) return(NA_real_)

  d <- cluster::daisy(df, metric = "gower")
  sil <- cluster::silhouette(lbl, dmatrix = as.matrix(d))
  mean(sil[, "sil_width"], na.rm = TRUE)
}

#8.3.7 Hungarian alignment helper
hungarian_map <- function(from, to) {
  from <- as.integer(from)
  to <- as.integer(to)
  K <- max(length(unique(from)), length(unique(to)))
  tab <- table(from, to)
  tab_full <- matrix(0, nrow = K, ncol = K)
  tab_full[seq_len(nrow(tab)), seq_len(ncol(tab))] <- tab
  maxv <- max(tab_full)
  cost <- maxv - tab_full
  assign <- clue::solve_LSAP(cost)
  as.integer(assign)
}

#8.3.8 KAMILA data prep for continuous + categorical split
prep_kamila <- function(df) {
  df <- as.data.frame(df)
  is_num <- vapply(df, is.numeric, logical(1))
  is_fac <- vapply(df, is.factor, logical(1))

  x_con <- as.data.frame(df[, is_num, drop = FALSE])
  x_cat <- as.data.frame(df[, is_fac, drop = FALSE])

  for (nm in names(x_cat)) x_cat[[nm]] <- factor(x_cat[[nm]])

  list(x_con = x_con, x_cat = x_cat)
}

#8.3.9 Build a version-safe kamila() argument list
build_kamila_args <- function(x_con,
                              x_cat,
                              k,
                              calc_mode = "none",
                              num_init = 8,
                              num_pred_strt = 25,
                              max_iter = 100,
                              ps_threshold = 0.80,
                              ps_cv_runs = 10,
                              verbose = FALSE) {

  fm <- names(formals(kamila::kamila))
  args <- list()

  #8.3.9.1 Core data arguments
  if ("conVar" %in% fm) {
    args$conVar <- x_con
  } else if ("xCon" %in% fm) {
    args$xCon <- x_con
  } else {
    stop("Could not find continuous-data argument name for kamila::kamila().")
  }

  if ("catFactor" %in% fm) {
    args$catFactor <- x_cat
  } else if ("xCat" %in% fm) {
    args$xCat <- x_cat
  } else {
    stop("Could not find categorical-data argument name for kamila::kamila().")
  }

  #8.3.9.2 Number of clusters
  if ("numClust" %in% fm) {
    args$numClust <- as.integer(k)
  } else if ("K" %in% fm) {
    args$K <- as.integer(k)
  }

  #8.3.9.3 Initialization count
  if ("numInit" %in% fm) {
    args$numInit <- as.integer(num_init)
  } else if ("nstart" %in% fm) {
    args$nstart <- as.integer(num_init)
  }

  #8.3.9.4 Prediction-strength starts
  if ("numPredStrt" %in% fm) {
    args$numPredStrt <- as.integer(num_pred_strt)
  }

  #8.3.9.5 Iteration cap
  if ("maxIter" %in% fm) {
    args$maxIter <- as.integer(max_iter)
  } else if ("max.iter" %in% fm) {
    args$max.iter <- as.integer(max_iter)
  }

  #8.3.9.6 Algorithm-native k selection
  if ("calcNumClust" %in% fm) {
    args$calcNumClust <- calc_mode
  }

  if (identical(calc_mode, "ps")) {
    if ("numPredStrCvRun" %in% fm) {
      args$numPredStrCvRun <- as.integer(ps_cv_runs)
    }
    if ("predStrThresh" %in% fm) {
      args$predStrThresh <- as.numeric(ps_threshold)
    }
  }

  if ("verbose" %in% fm) {
    args$verbose <- verbose
  }

  args
}

#8.3.10 Extract labels from kamila fit robustly across versions
extract_kamila_labels <- function(fit) {
  if (!is.null(fit$finalMemb)) return(as.integer(fit$finalMemb))
  if (!is.null(fit$cluster)) return(as.integer(fit$cluster))
  if (!is.null(fit$clustering)) return(as.integer(fit$clustering))
  stop("Could not extract cluster labels from KAMILA fit.")
}

#8.3.11 clustMD data prep for continuous first, then binary, then nominal
prep_clustMD <- function(df) {
  df <- as.data.frame(df)

  is_num <- vapply(df, is.numeric, logical(1))
  is_fac <- vapply(df, is.factor, logical(1))

  fac_names <- names(df)[is_fac]
  num_names <- names(df)[is_num]

  bin_names <- fac_names[vapply(df[fac_names], nlevels, integer(1)) == 2L]
  nom_names <- fac_names[vapply(df[fac_names], nlevels, integer(1)) > 2L]

  X_num <- df[, num_names, drop = FALSE]
  X_bin <- df[, bin_names, drop = FALSE]
  X_nom <- df[, nom_names, drop = FALSE]

  if (ncol(X_bin) > 0) X_bin[] <- lapply(X_bin, function(x) as.integer(factor(x)))
  if (ncol(X_nom) > 0) X_nom[] <- lapply(X_nom, function(x) as.integer(factor(x)))

  X_all <- cbind(X_num, X_bin, X_nom)

  list(
    X = as.matrix(X_all),
    CnsIndx = ncol(X_num),
    OrdIndx = ncol(X_num) + ncol(X_bin))
}

#8.3.12 Extract labels from clustMD fit
extract_clustMD_labels <- function(fit) {
  out <- NULL
  if (!is.null(fit$cl)) out <- fit$cl
  if (is.null(out) && !is.null(fit$map)) out <- fit$map
  if (is.null(out) && !is.null(fit$class)) out <- fit$class
  if (is.null(out)) stop("Could not extract cluster labels from clustMD fit.")
  as.integer(out)
}

#8.3.13 Objective-like extractors
#8.3.13.1 Kproto
kproto_objective <- function(kp) {
  for (nm in c("tot.withinss", "tot.withindist", "tot.within.dist", "tot.within")) {
    if (!is.null(kp[[nm]]) && is.finite(kp[[nm]])) return(as.numeric(kp[[nm]]))
  }
  NA_real_
}

#8.3.13.2 KAMILA
kamila_objective <- function(fit) {
  if (!is.null(fit$objective)) return(as.numeric(fit$objective))
  if (!is.null(fit$crit)) return(as.numeric(fit$crit))
  NA_real_
}

#8.3.13.3 clustMD
clustMD_bic <- function(fit) {
  if (!is.null(fit$BIC)) return(as.numeric(fit$BIC))
  NA_real_
}

#8.3.14 UMAP design matrix for mixed-data visualization
make_umap_matrix <- function(df) {
  mm <- model.matrix(~ . - 1, data = as.data.frame(df))
  scale(mm)
}

```

## 8.4 Prepare Benchmark Input Frame

```{r prepare frame, echo = FALSE, warning = FALSE}

## 8.4 Prepare complete-case benchmark frame ##

X0 <- risk_dt[, anchor_features, with = FALSE] %>% as.data.frame()

#8.4.1 Mirror original complete-case baseline clustering frame
cc_idx <- stats::complete.cases(X0)
bench_dt <- risk_dt[cc_idx == TRUE]
X_cc <- X0[cc_idx, , drop = FALSE]

#8.4.2 Coerce types to match mixed-data algorithms
X_cc <- force_known_categoricals(X_cc)
X_cc <- coerce_types(X_cc)
X_cc <- zscore_numeric(X_cc)

cat(glue("Benchmark frame complete-case N={nrow(X_cc)}.\n"))

#8.4.3 Anchor overlap for ARI comparison
pid_cc <- as.character(bench_dt$participant_id)
in_anchor <- pid_cc %in% anchor_id
subj_overlap <- pid_cc[in_anchor]
anchor_lbl_overlap <- anchor_lbl[match(subj_overlap, anchor_id)]

```

## 8.5 Fit k-prototypes Across k = 2:8

```{r fit kproto, echo = FALSE, warning = FALSE}

## 8.5 k-prototypes benchmark ##

run_kproto_grid <- function(X_df) {

  lam_info <- lambda_safe(
    X_df,
    lambda_mode = params$lambda_mode,
    lambda_fixed = params$lambda_fixed)

  res_list <- list()
  model_list <- list()
  assign_list <- list()

  for (k in params$k_grid) {

    set.seed(1000 + k)

    fit <- clustMixType::kproto(
      x = X_df,
      k = as.integer(k),
      lambda = lam_info$lambda,
      nstart = as.integer(params$nstart),
      verbose = FALSE)

    lbl <- as.integer(fit$cluster)

    idx_overlap <- match(subj_overlap, pid_cc)
    ari_anchor <- mclust::adjustedRandIndex(lbl[idx_overlap], anchor_lbl_overlap)

    res_list[[as.character(k)]] <- tibble::tibble(
      algorithm = "kproto",
      k = as.integer(k),
      objective = kproto_objective(fit),
      bic = NA_real_,
      gower_silhouette = gower_silhouette_mean(X_df, lbl),
      ari_vs_anchor_k2 = ari_anchor,
      min_cluster_n = min(tabulate(lbl)),
      max_cluster_n = max(tabulate(lbl)),
      lambda = as.numeric(lam_info$lambda),
      lambda_reason = lam_info$reason)

    assign_list[[as.character(k)]] <- tibble::tibble(
      participant_id = pid_cc,
      algorithm = "kproto",
      k = as.integer(k),
      cluster = lbl)

    if (!isTRUE(params$keep_model_data)) fit$data <- NULL
    model_list[[as.character(k)]] <- fit
  }

  list(
    metrics = bind_rows(res_list),
    assignments = bind_rows(assign_list),
    models = model_list)
}

kproto_res <- run_kproto_grid(X_cc)

write_csv2(kproto_res$metrics, file.path(paths$tab_dir, "kproto_grid_metrics.csv"))
write_csv2(kproto_res$assignments, file.path(paths$tab_dir, "kproto_grid_assignments.csv"))
save_rds2(kproto_res$models, file.path(paths$rds_dir, "kproto_grid_models.rds"))

```

## 8.6 Fit KAMILA Across k = 2:8

```{r fit kamila, echo = FALSE, warning = FALSE}

## 8.6 KAMILA benchmark ##

run_kamila_grid <- function(X_df) {

  kam_in <- prep_kamila(X_df)

  if (ncol(kam_in$x_con) < 1 || ncol(kam_in$x_cat) < 1) {
    stop("KAMILA requires both continuous and categorical variables.")
  }

  res_list <- list()
  model_list <- list()
  assign_list <- list()

  for (k in params$k_grid) {

    set.seed(2000 + k)

    kam_args <- build_kamila_args(
      x_con = kam_in$x_con,
      x_cat = kam_in$x_cat,
      k = k,
      calc_mode = "none",
      num_init = params$kamila_num_init,
      num_pred_strt = params$kamila_num_pred_strt,
      max_iter = params$kamila_max_iter,
      ps_threshold = params$kamila_ps_threshold,
      ps_cv_runs = params$kamila_ps_cv_runs,
      verbose = FALSE)
    
    fit <- do.call(kamila::kamila, kam_args)
    
    lbl <- extract_kamila_labels(fit)

    idx_overlap <- match(subj_overlap, pid_cc)
    ari_anchor <- mclust::adjustedRandIndex(lbl[idx_overlap], anchor_lbl_overlap)

    res_list[[as.character(k)]] <- tibble::tibble(
      algorithm = "kamila",
      k = as.integer(k),
      objective = kamila_objective(fit),
      bic = NA_real_,
      gower_silhouette = gower_silhouette_mean(X_df, lbl),
      ari_vs_anchor_k2 = ari_anchor,
      min_cluster_n = min(tabulate(lbl)),
      max_cluster_n = max(tabulate(lbl)),
      lambda = NA_real_,
      lambda_reason = NA_character_)

    assign_list[[as.character(k)]] <- tibble::tibble(
      participant_id = pid_cc,
      algorithm = "kamila",
      k = as.integer(k),
      cluster = lbl)

    model_list[[as.character(k)]] <- fit
  }

  list(
    metrics = bind_rows(res_list),
    assignments = bind_rows(assign_list),
    models = model_list)
}

kamila_res <- run_kamila_grid(X_cc)

write_csv2(kamila_res$metrics, file.path(paths$tab_dir, "kamila_grid_metrics.csv"))
write_csv2(kamila_res$assignments, file.path(paths$tab_dir, "kamila_grid_assignments.csv"))
save_rds2(kamila_res$models, file.path(paths$rds_dir, "kamila_grid_models.rds"))

```

## 8.7 KAMILA Algorithm-Native k Selection via Prediction Strength

```{r kamila prediction strength, echo = FALSE, warning = FALSE}

## 8.7 KAMILA calcNumClust = 'ps' ##

kamila_ps_tbl <- NULL
kamila_ps_fit <- NULL

if (isTRUE(params$run_kamila_ps)) {

  kam_in <- prep_kamila(X_cc)

  set.seed(params$seed + 5000)

  kam_ps_args <- build_kamila_args(
  x_con = kam_in$x_con,
  x_cat = kam_in$x_cat,
  k = max(params$k_grid),
  calc_mode = "ps",
  num_init = params$kamila_num_init,
  num_pred_strt = params$kamila_num_pred_strt,
  max_iter = params$kamila_max_iter,
  ps_threshold = params$kamila_ps_threshold,
  ps_cv_runs = params$kamila_ps_cv_runs,
  verbose = FALSE)

  kamila_ps_fit <- do.call(kamila::kamila, kam_ps_args)

  save_rds2(kamila_ps_fit, file.path(paths$rds_dir, "kamila_prediction_strength_object.rds"))

  if (!is.null(kamila_ps_fit$nClust)) {
    kamila_ps_tbl <- as.data.frame(kamila_ps_fit$nClust)
    write_csv2(kamila_ps_tbl, file.path(paths$tab_dir, "kamila_prediction_strength_results.csv"))
  }
}

if (!is.null(kamila_ps_tbl)) {
  knitr::kable(
    kamila_ps_tbl,
    digits = 3,
    caption = "KAMILA algorithm-native cluster-number results (prediction strength)") %>%
    kableExtra::kable_styling(full_width = FALSE)
}

```

## 8.8 Fit clustMD Across k = 2:8

```{r fit clustmd, echo = FALSE, warning = FALSE}

## 8.8 clustMD benchmark ##

clustmd_res <- NULL

if (isTRUE(params$run_clustMD)) {

  run_clustMD_grid <- function(X_df) {

    clmd_in <- prep_clustMD(X_df)

    res_list <- list()
    model_list <- list()
    assign_list <- list()

    for (k in params$k_grid) {

      set.seed(3000 + k)

      fit <- clustMD::clustMD(
        X = clmd_in$X,
        G = as.integer(k),
        CnsIndx = clmd_in$CnsIndx,
        OrdIndx = clmd_in$OrdIndx,
        Nnorms = as.integer(params$clustmd_nnorms),
        MaxIter = as.integer(params$clustmd_max_iter),
        model = params$clustmd_model,
        store.params = FALSE,
        scale = FALSE,
        startCL = "hc_mclust",
        autoStop = FALSE)

      lbl <- extract_clustMD_labels(fit)

      idx_overlap <- match(subj_overlap, pid_cc)
      ari_anchor <- mclust::adjustedRandIndex(lbl[idx_overlap], anchor_lbl_overlap)

      res_list[[as.character(k)]] <- tibble::tibble(
        algorithm = "clustMD",
        k = as.integer(k),
        objective = NA_real_,
        bic = clustMD_bic(fit),
        gower_silhouette = gower_silhouette_mean(X_df, lbl),
        ari_vs_anchor_k2 = ari_anchor,
        min_cluster_n = min(tabulate(lbl)),
        max_cluster_n = max(tabulate(lbl)),
        lambda = NA_real_,
        lambda_reason = NA_character_)

      assign_list[[as.character(k)]] <- tibble::tibble(
        participant_id = pid_cc,
        algorithm = "clustMD",
        k = as.integer(k),
        cluster = lbl)

      model_list[[as.character(k)]] <- fit
    }

    list(
      metrics = bind_rows(res_list),
      assignments = bind_rows(assign_list),
      models = model_list)
  }

  clustmd_res <- run_clustMD_grid(X_cc)

  write_csv2(clustmd_res$metrics, file.path(paths$tab_dir, "clustMD_grid_metrics.csv"))
  write_csv2(clustmd_res$assignments, file.path(paths$tab_dir, "clustMD_grid_assignments.csv"))
  save_rds2(clustmd_res$models, file.path(paths$rds_dir, "clustMD_grid_models.rds"))
}

```

## 8.9 Summarize Benchmark Results

```{r summarize, echo = FALSE, warning = FALSE}

## 8.9 Summary of algorithm benchmark ##

metric_list <- list(
  kproto_res$metrics,
  kamila_res$metrics)

if (!is.null(clustmd_res)) {
  metric_list <- c(metric_list, list(clustmd_res$metrics))
}

bench_metrics <- bind_rows(metric_list) %>%
  arrange(algorithm, k)

write_csv2(bench_metrics, file.path(paths$tab_dir, "algorithm_benchmark_metrics_all.csv"))

#8.9.1 Primary recommendation rule:
# maximize common Gower silhouette; tie-break by smaller k
best_df <- bench_metrics %>%
  filter(is.finite(gower_silhouette)) %>%
  group_by(algorithm) %>%
  arrange(desc(gower_silhouette), k) %>%
  slice(1) %>%
  ungroup()

write_csv2(best_df, file.path(paths$tab_dir, "algorithm_benchmark_recommendation.csv"))

knitr::kable(
  best_df,
  digits = 3,
  caption = "Recommended k per algorithm (primary benchmark = max Gower silhouette; smaller k breaks ties)") %>%
  kableExtra::kable_styling(full_width = FALSE)

```

## 8.10 Plots

```{r plots, echo = FALSE, warning = FALSE}

## 8.10 Visualization of benchmark results ##

p_sil <- ggplot(bench_metrics, aes(x = k, y = gower_silhouette, color = algorithm)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.9) +
  theme_classic(base_size = 12) +
  labs(
    title = "Common internal validation across algorithms",
    subtitle = "Primary benchmark metric = average silhouette on Gower distance",
    x = "k",
    y = "Gower silhouette")

ggsave(file.path(paths$fig_dir, "benchmark_gower_silhouette_by_k.png"), p_sil, width = 7.6, height = 4.6, dpi = 600, bg = "white")

p_ari <- ggplot(bench_metrics, aes(x = k, y = ari_vs_anchor_k2, color = algorithm)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.9) +
  theme_classic(base_size = 12) +
  labs(
    title = "Agreement with original k-prototypes k=2 anchor",
    subtitle = "Descriptive comparability metric only; not the optimization target",
    x = "k",
    y = "ARI vs anchor")

ggsave(file.path(paths$fig_dir, "benchmark_ari_vs_anchor_by_k.png"), p_ari, width = 7.6, height = 4.6, dpi = 600, bg = "white")

p_size <- ggplot(bench_metrics, aes(x = k, y = min_cluster_n, color = algorithm)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.9) +
  theme_classic(base_size = 12) +
  labs(
    title = "Minimum cluster size across algorithms",
    x = "k",
    y = "Minimum cluster size")

ggsave(file.path(paths$fig_dir, "benchmark_min_cluster_size_by_k.png"), p_size, width = 7.6, height = 4.6, dpi = 600, bg = "white")

if ("clustMD" %in% bench_metrics$algorithm) {
  p_bic <- bench_metrics %>%
    filter(algorithm == "clustMD", is.finite(bic)) %>%
    ggplot(aes(x = k, y = bic)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.9) +
    theme_classic(base_size = 12) +
    labs(
      title = "clustMD model-based fit",
      subtitle = "BIC shown descriptively for the model-based comparator",
      x = "k",
      y = "BIC")

  ggsave(file.path(paths$fig_dir, "benchmark_clustMD_bic_by_k.png"), p_bic, width = 7.6, height = 4.6, dpi = 600, bg = "white")
}

```

## 8.11 UMAP Visualization of Each Algorithm's Preferred Solution

```{r umap, echo = FALSE, warning = FALSE}

## 8.11 UMAP of preferred solutions ##

best_by_alg <- best_df %>%
  select(algorithm, k)

mm <- make_umap_matrix(X_cc)

set.seed(params$umap_seed)
emb <- uwot::umap(
  mm,
  n_neighbors = params$umap_neighbors,
  min_dist = params$umap_min_dist,
  metric = "euclidean",
  verbose = TRUE)

get_labels_for_best <- function(alg, k_best) {
  if (alg == "kproto") {
    return(as.integer(kproto_res$models[[as.character(k_best)]]$cluster))
  }
  if (alg == "kamila") {
  return(extract_kamila_labels(kamila_res$models[[as.character(k_best)]]))
  }
  if (alg == "clustMD") {
    return(extract_clustMD_labels(clustmd_res$models[[as.character(k_best)]]))
  }
  stop("Unknown algorithm.")
}

umap_plots <- lapply(seq_len(nrow(best_by_alg)), function(i) {
  alg <- best_by_alg$algorithm[i]
  k_best <- best_by_alg$k[i]
  lbl <- get_labels_for_best(alg, k_best)

  umap_df <- tibble::tibble(
    UMAP1 = emb[, 1],
    UMAP2 = emb[, 2],
    cluster = factor(lbl))

  write_csv2(
    umap_df %>% mutate(algorithm = alg, k = k_best),
    file.path(paths$tab_dir, glue("umap_{alg}_k{k_best}.csv")))

  ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cluster)) +
    geom_point(alpha = 0.6, size = 0.8) +
    theme_classic(base_size = 12) +
    labs(
      title = glue("UMAP: {alg} preferred solution"),
      subtitle = glue("k = {k_best}"),
      x = "UMAP1",
      y = "UMAP2",
      color = "Cluster")
})

for (i in seq_along(umap_plots)) print(umap_plots[[i]])

```

## 8.12 Reproducibility Log

```{r reproducibility, echo = FALSE, warning = FALSE}

## 8.12 Reproducibility Log ##

log <- list(
  timestamp = as.character(Sys.time()),
  REPO = REPO,
  risk_data_path = risk_path,
  kproto_anchor_rds = params$kproto_anchor_rds,
  out_dir = paths$out_dir,
  k_grid = params$k_grid,
  nstart = params$nstart,
  lambda_mode = params$lambda_mode,
  lambda_fixed = params$lambda_fixed,
  kamila = list(
    num_init = params$kamila_num_init,
    num_pred_strt = params$kamila_num_pred_strt,
    max_iter = params$kamila_max_iter,
    run_ps = params$run_kamila_ps,
    ps_threshold = params$kamila_ps_threshold,
    ps_cv_runs = params$kamila_ps_cv_runs),
  clustMD = list(
    run = params$run_clustMD,
    model = params$clustmd_model,
    max_iter = params$clustmd_max_iter,
    nnorms = params$clustmd_nnorms),
  umap = list(
    n_neighbors = params$umap_neighbors,
    min_dist = params$umap_min_dist,
    seed = params$umap_seed))

writeLines(
  jsonlite::toJSON(log, pretty = TRUE),
  con = file.path(paths$out_dir, "run_log.json"))

```
