---
title: "Preliminary Expanded Feature Clustering Sensitivity"
author: "Sam A. Sievertsen"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
params:
  scaling_method: "z_score"
  k_max: 8
  n_boot: 25
  stab_nstart: 8
  seed: 123
  overwrite: false
  run_validation_kproto: false
  sil_method: "full"
  sil_subsample_n: 2000
  sil_subsample_reps: 5
  cbcl_dp_cutoff: 65
  base_config: "bas2_fhx"
  descriptive_config: "full_maximal"
  profile_configs: !r c("bas2_fhx", "cbcl_dp_rawsum", "full_maximal", "full_maximal_dpsum", "anchor_nocbcl", "full_maximal_nocbcl")
  risk_data_file: "risk_variable_data.csv"
  raw_data_file: "dataset.csv"
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, message = TRUE, warning = TRUE, results = "markup", verbose = TRUE, comment = "")

```

```{r environment, echo = FALSE, include = FALSE, warning = FALSE}

## Load packages, data, & set env ##

# Load necessary packages
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(furrr)
library(parallelly)
library(readr)
library(clustMixType)
library(cluster)
library(mclust)
library(effectsize)
library(moments)
library(ggplot2)
library(glue)
library(here)

# Configure parallel backend across the k values being fit, respecting the SLURM/cgroup allocation
n_workers <- min(as.integer(parallelly::availableCores()), length(2:params$k_max))
if (n_workers > 1) {
  plan(multisession, workers = n_workers)
} else {
  plan(sequential)
}
options(future.globals.maxSize = 5e9)
cat(glue::glue("Parallel workers: {n_workers}\n"))

# Set numerical options
options(scipen = 999, digits = 8)

# Set seed for reproducible random starts
set.seed(params$seed)

# Set scaling method from piped params argument
scaling_method <- params$scaling_method

# Set paths to be used in reading in data & objects
REPO <- here()
datapath <- file.path(REPO, "data", "data_processed")

# Establish the path to the preliminary results dir and write it if it does not exist
prelim_dir <- file.path(datapath, "expanded_feature_preliminary")
dir.create(prelim_dir, recursive = TRUE, showWarnings = FALSE)

# Set up logging (from SLURM), falling back to a local file when unset
log_file <- Sys.getenv("DETAILED_LOG", unset = file.path(REPO, "logs", "expanded_feature_preliminary.log"))
dir.create(dirname(log_file), recursive = TRUE, showWarnings = FALSE)

# Create a logging helper matching the pipe delimited convention used elsewhere
log_event <- function(tag, ...) {
  cat(
    sprintf("%s|%s|%s\n", format(Sys.time(), "%Y-%m-%d %H:%M:%OS3"), tag, paste(..., collapse = "|")),
    file = log_file, append = TRUE
  )
}

# Read in risk variable data for clustering
risk_variable_data <- read.csv(file.path(datapath, params$risk_data_file))

```

## Purpose

Preliminary sensitivity run resolving the open feature-set questions empirically before committing to the full revision pipeline. Eleven configurations, most varying one element from a common base so differences are attributable:

| Config | Question it answers |
|---|---|
| `anchor` | Published 17 features, refit on common rows. Validity check. |
| `bas2_fhx` | **BASE.** BAS RR (v01) + Drive, plus FHx psychosis and hospitalization |
| `bas4_fhx` | Does adding BIS (v01) and Fun Seeking change anything? |
| `cbcl3_dim` | Does swapping DSM-5 depression + anxiety for the Anxious/Depressed syndrome scale change anything? |
| `cbcl3_bin` | Does the classic binary dysregulation profile (all 3 scales T >= cutoff) change anything? |
| `cbcl_dp_rawsum` | **Does replacing the 4 CBCL scales with the DP raw sum (Giampetruzzi et al., 2026) change anything?** |
| `cbcl_dp_pc1` | **Does replacing the 4 CBCL scales with the DP first principal component change anything?** |
| `ace_split` | Does partitioning abuse out of the ACE composite change anything? |
| `full_maximal` | Everything at once: all current features plus BIS, Fun Seeking, FHx additions, and the ACE abuse partition, holding CBCL as-is |
| `full_maximal_dpsum` | **Everything at once, but with the 4 CBCL scales replaced by the DP raw sum** |
| `anchor_nocbcl` | What remains of the published solution with the CBCL scales removed entirely? |
| `full_maximal_nocbcl` | What remains of the maximal solution with the CBCL scales removed entirely? |

The two new DP configurations follow Giampetruzzi et al. (2026, *JCPP*), who operationalized the dysregulation profile two ways: as the sum of the three raw subscale scores, and as the first principal component of the three subscales (centered and scaled). Both replace the four CBCL scales with a single continuous variable, which also isolates whether the `cbcl3_bin` collapse was due to dichotomization or to reducing four features to one. Note that Giampetruzzi used the youth self-report YSR; the CBCL here is parent report.

Read the **CBCL Dysregulation Profile Variables**, **Decision Aid**, and **Feature Separation Profiles** sections first. Everything else is supporting detail.

Existing cached solutions are reused. The DP variables are derived from the CBCL subscales already in the feature union, so the analytic row set should be unchanged; a fingerprint check halts execution if it is not.

Bootstrap replicates default to `r params$n_boot`. Rerun the chosen configuration at `n_boot = 1000` before reporting anything externally.

## Data Prep

```{r data prep, warning = FALSE, echo = FALSE}

## Data Prep ##

#1. Define the feature sets using the column names written by the wrangling script
#1.1 The published 17 clustering features
feats_anchor <- c(
  "mh_p_cbcl__dsm__dep_tscore", "mh_p_cbcl__dsm__anx_tscore",
  "mh_p_cbcl__synd__attn_tscore", "mh_p_cbcl__synd__aggr_tscore",
  "mh_p_gbi_sum", "mh_y_upps__nurg_sum", "mh_y_upps__purg_sum",
  "le_l_coi__addr1__coi__total__national_zscore", "fc_p_nsc__ns_mean", "sds_total",
  "family_history_depression", "family_history_mania", "bullying",
  "nc_y_nihtb__lswmt__uncor_score", "nc_y_nihtb__flnkr__uncor_score",
  "nc_y_nihtb__pttcp__uncor_score", "ACE_index_sum_score"
)

#1.2 The four CBCL scales, isolated so they can be removed as a block
feats_cbcl <- c(
  "mh_p_cbcl__dsm__dep_tscore", "mh_p_cbcl__dsm__anx_tscore",
  "mh_p_cbcl__synd__attn_tscore", "mh_p_cbcl__synd__aggr_tscore"
)

#1.3 The three CBCL syndrome raw sums that compose the dysregulation profile
feats_cbcl_dp_raw <- c("mh_p_cbcl__synd__anxdep_sum", "mh_p_cbcl__synd__attn_sum", "mh_p_cbcl__synd__aggr_sum")

#1.4 The BAS scales, per Pagliaccio et al. (2016) revised scoring
feats_bas2 <- c("mh_y_bisbas__bas__rr_sum__v01", "mh_y_bisbas__bas__dr_sum")
feats_bas4 <- c(feats_bas2, "mh_y_bisbas__bis_sum__v01", "mh_y_bisbas__bas__fs_sum")

#1.5 The two derived family history variables
feats_fhx <- c("family_history_psychosis", "family_history_hospitalization")

#1.6 The SDSC subscales, reported descriptively only and never clustered
sdsc_subscales <- c(
  "ph_p_sds__dims_sum", "ph_p_sds__sbd_sum", "ph_p_sds__da_sum",
  "ph_p_sds__swtd_sum", "ph_p_sds__does_sum", "ph_p_sds__hyphy_sum"
)

#1.7 Raw columns needed to derive the abuse indicators and the emotional abuse proxy
#1.7.1 KSADS PTSD module items indexing physical and sexual abuse (release 5.1 names, no 6.x equivalent)
ksads_abuse_items <- c(
  physical = "ksads_ptsd_raw_762_p", physical2 = "ksads_ptsd_raw_763_p",
  sexual   = "ksads_ptsd_raw_767_p", sexual2   = "ksads_ptsd_raw_768_p"
)

#1.7.2 Family Environment Scale conflict item used as an emotional abuse proxy, both reporters
fes_emotional_items <- c("fc_y_fes__confl_005", "fc_p_fes__confl_005")

#2. Derive the two new family history variables from the raw DEAP export
#2.1 Map the desired variable names onto their ABCD family history module stems
famhx_stems <- c(
  family_history_psychosis = "halluc",
  family_history_hospitalization = "hosp"
)

#2.2 Build the first-degree relative source column names for a given stem
famhx_source_cols <- function(stem) {
  c(
    sprintf("mh_p_famhx__%s__fath_001", stem),
    sprintf("mh_p_famhx__%s__moth_001", stem),
    sprintf("mh_p_famhx__%s__full__young_%03d", stem, 1:5),
    sprintf("mh_p_famhx__%s__full__old_%03d", stem, 1:5)
  )
}

#2.3 Build the any-relative gate column name for a given stem
famhx_gate_col <- function(stem) sprintf("mh_p_famhx__%s_001", stem)

#2.4 Assemble every raw family history column needed
famhx_raw_cols <- unlist(
  lapply(famhx_stems, function(s) c(famhx_source_cols(s), famhx_gate_col(s))),
  use.names = FALSE
)

#2.5 List the BAS, SDSC, and CBCL columns to be pulled from the raw export
new_columns <- c(feats_bas4, sdsc_subscales, "mh_p_cbcl__synd__anxdep_tscore", feats_cbcl_dp_raw)

#2.6 Read only the needed columns from the raw export and filter to baseline
raw_data_baseline <- readr::read_csv(
  file.path(REPO, "data", "data_raw", params$raw_data_file),
  col_select = all_of(c("participant_id", "session_id", new_columns, famhx_raw_cols,
                        unname(ksads_abuse_items), fes_emotional_items)),
  show_col_types = FALSE
) %>%
  filter(session_id == "ses-00A")

#2.7 Confirm one baseline row per participant before deriving or joining
stopifnot(!any(duplicated(raw_data_baseline$participant_id)))

#2.8 Derive a first-degree relative history indicator, mirroring the depression derivation
derive_family_history <- function(data, stem) {

  #2.8.1 Flag any first-degree relative endorsement, treating NA as not endorsed
  any_first_degree <- Reduce(
    `|`,
    lapply(famhx_source_cols(stem), function(cc) tidyr::replace_na(data[[cc]] == 1, FALSE))
  )

  #2.8.2 Assign 1 for any first-degree endorsement, 0 where the gate item was answered, NA otherwise
  dplyr::case_when(
    any_first_degree ~ 1,
    data[[famhx_gate_col(stem)]] == 1 ~ 0,
    data[[famhx_gate_col(stem)]] == 0 ~ 0,
    TRUE ~ NA_real_
  )
}

#2.9 Apply the derivation for each stem and report endorsement
for (nm in names(famhx_stems)) {
  raw_data_baseline[[nm]] <- derive_family_history(raw_data_baseline, famhx_stems[[nm]])
  cat(glue(
    "{nm}: {sum(raw_data_baseline[[nm]] == 1, na.rm = TRUE)} endorsed, ",
    "{sum(is.na(raw_data_baseline[[nm]]))} missing of {nrow(raw_data_baseline)}\n"
  ))
}

#3. Derive the abuse indicators used by the ace_split and maximal configurations
#3.1 Physical abuse factor: either KSADS physical abuse item endorsed
raw_data_baseline$abuse_physical <- dplyr::case_when(
  raw_data_baseline[[ksads_abuse_items[["physical"]]]] == 1 |
    raw_data_baseline[[ksads_abuse_items[["physical2"]]]] == 1 ~ 1,
  !is.na(raw_data_baseline[[ksads_abuse_items[["physical"]]]]) |
    !is.na(raw_data_baseline[[ksads_abuse_items[["physical2"]]]]) ~ 0,
  TRUE ~ NA_real_
)

#3.2 Sexual abuse factor: either KSADS sexual abuse item endorsed
raw_data_baseline$abuse_sexual <- dplyr::case_when(
  raw_data_baseline[[ksads_abuse_items[["sexual"]]]] == 1 |
    raw_data_baseline[[ksads_abuse_items[["sexual2"]]]] == 1 ~ 1,
  !is.na(raw_data_baseline[[ksads_abuse_items[["sexual"]]]]) |
    !is.na(raw_data_baseline[[ksads_abuse_items[["sexual2"]]]]) ~ 0,
  TRUE ~ NA_real_
)

#3.3 Emotional abuse proxy: either reporter endorses the family criticism item, mirroring the bullying collapse
raw_data_baseline$abuse_emotional <- dplyr::case_when(
  raw_data_baseline[[fes_emotional_items[1]]] == 1 |
    raw_data_baseline[[fes_emotional_items[2]]] == 1 ~ 1,
  !is.na(raw_data_baseline[[fes_emotional_items[1]]]) |
    !is.na(raw_data_baseline[[fes_emotional_items[2]]]) ~ 0,
  TRUE ~ NA_real_
)

#3.4 Report endorsement for each abuse indicator
feats_abuse <- c("abuse_physical", "abuse_sexual", "abuse_emotional")
for (nm in feats_abuse) {
  cat(glue(
    "{nm}: {sum(raw_data_baseline[[nm]] == 1, na.rm = TRUE)} endorsed ",
    "({round(100 * mean(raw_data_baseline[[nm]] == 1, na.rm = TRUE), 2)}%), ",
    "{sum(is.na(raw_data_baseline[[nm]]))} missing\n"
  ))
}

#4. Join the derived and new columns onto the clustering data
risk_variable_data <- risk_variable_data %>%
  left_join(
    raw_data_baseline %>%
      dplyr::select(all_of(c("participant_id", new_columns, names(famhx_stems), feats_abuse))),
    by = "participant_id"
  )

#5. Derive the reduced ACE sum, the CBCL dysregulation profile binary, and the DP raw sum
#5.1 Reduced ACE sum removes the physical and sexual abuse factors from the composite
#5.1.1 The emotional abuse proxy is not currently in the composite, so nothing is removed for it
risk_variable_data <- risk_variable_data %>%
  mutate(
    ACE_index_reduced = ACE_index_sum_score - abuse_physical - abuse_sexual
  )

#5.2 Verify the reduced sum is coherent before it is used as a feature
ace_check <- risk_variable_data %>%
  filter(!is.na(ACE_index_reduced))
cat(glue(
  "ACE reduced sum: range {min(ace_check$ACE_index_reduced)} to {max(ace_check$ACE_index_reduced)}; ",
  "{sum(ace_check$ACE_index_reduced < 0)} negative values (should be 0)\n"
))
if (any(ace_check$ACE_index_reduced < 0)) {
  stop("Negative values in ACE_index_reduced. The abuse factor derivation does not match the composite; stop and reconcile against the wrangling script before proceeding.")
}

#5.3 CBCL dysregulation profile binary: all three syndrome scales at or above the cutoff
risk_variable_data <- risk_variable_data %>%
  mutate(
    cbcl_dysregulation_binary = dplyr::case_when(
      mh_p_cbcl__synd__anxdep_tscore >= params$cbcl_dp_cutoff &
        mh_p_cbcl__synd__attn_tscore >= params$cbcl_dp_cutoff &
        mh_p_cbcl__synd__aggr_tscore >= params$cbcl_dp_cutoff ~ 1,
      !is.na(mh_p_cbcl__synd__anxdep_tscore) &
        !is.na(mh_p_cbcl__synd__attn_tscore) &
        !is.na(mh_p_cbcl__synd__aggr_tscore) ~ 0,
      TRUE ~ NA_real_
    )
  )

#5.4 Report dysregulation profile endorsement
cat(glue(
  "CBCL dysregulation profile (all 3 scales T >= {params$cbcl_dp_cutoff}): ",
  "{sum(risk_variable_data$cbcl_dysregulation_binary == 1, na.rm = TRUE)} endorsed ",
  "({round(100 * mean(risk_variable_data$cbcl_dysregulation_binary == 1, na.rm = TRUE), 2)}%)\n"
))

#5.5 CBCL dysregulation profile raw sum, per Giampetruzzi et al. (2026)
risk_variable_data <- risk_variable_data %>%
  mutate(cbcl_dp_raw_sum = mh_p_cbcl__synd__anxdep_sum + mh_p_cbcl__synd__attn_sum + mh_p_cbcl__synd__aggr_sum)

#5.6 Confirm the raw sums carry no missingness beyond the T-scores they generate
raw_vs_t_missing <- sum(!is.na(risk_variable_data$mh_p_cbcl__synd__aggr_tscore) & is.na(risk_variable_data$cbcl_dp_raw_sum))
cat(glue("Rows with CBCL T-scores but missing a DP raw sum: {raw_vs_t_missing} (should be 0)\n"))
if (raw_vs_t_missing > 0) {
  warning("DP raw sum has missingness beyond the T-scores; the analytic row set may change and invalidate cached fits.")
}

#6. Sanity check that the ABCD SDSC subscales reconstruct the internally computed total
sds_check <- risk_variable_data %>%
  filter(if_all(all_of(c("sds_total", sdsc_subscales)), ~ !is.na(.))) %>%
  mutate(subscale_sum = rowSums(across(all_of(sdsc_subscales))))
cat(glue(
  "SDSC total vs subscale sum: {sum(sds_check$sds_total == sds_check$subscale_sum)} of ",
  "{nrow(sds_check)} exact matches (r = {round(cor(sds_check$sds_total, sds_check$subscale_sum), 4)})\n"
))

#7. Define the eleven configurations
#7.1 The leading candidate base
feats_base <- c(feats_anchor, feats_bas2, feats_fhx)

#7.2 The maximal set, defined separately so its no-CBCL variant can be derived from it
feats_maximal <- c(
  setdiff(feats_anchor, "ACE_index_sum_score"),
  feats_bas4, feats_fhx,
  "ACE_index_reduced", feats_abuse
)

#7.3 Assemble the configuration list
config_feats <- list(

  #7.3.1 Published 17 features, refit on common rows as a validity check
  anchor = feats_anchor,

  #7.3.2 Leading candidate base
  bas2_fhx = feats_base,

  #7.3.3 Full BIS/BAS, adding BIS (v01) and Fun Seeking
  bas4_fhx = c(feats_anchor, feats_bas4, feats_fhx),

  #7.3.4 Three dimensional CBCL scales, swapping DSM-5 dep and anx for the Anxious/Depressed syndrome scale
  cbcl3_dim = c(
    setdiff(feats_base, c("mh_p_cbcl__dsm__dep_tscore", "mh_p_cbcl__dsm__anx_tscore")),
    "mh_p_cbcl__synd__anxdep_tscore"
  ),

  #7.3.5 Classic binary dysregulation profile replacing all four CBCL scales
  cbcl3_bin = c(
    setdiff(feats_base, feats_cbcl),
    "cbcl_dysregulation_binary"
  ),

  #7.3.6 DP raw sum replacing all four CBCL scales
  cbcl_dp_rawsum = c(
    setdiff(feats_base, feats_cbcl),
    "cbcl_dp_raw_sum"
  ),

  #7.3.7 DP first principal component replacing all four CBCL scales; derived below after row restriction
  cbcl_dp_pc1 = c(
    setdiff(feats_base, feats_cbcl),
    "cbcl_dp_pc1"
  ),

  #7.3.8 ACE composite partitioned into abuse binaries plus a reduced sum
  ace_split = c(
    setdiff(feats_base, "ACE_index_sum_score"),
    "ACE_index_reduced", feats_abuse
  ),

  #7.3.9 Maximal set, holding the CBCL scales as they are
  full_maximal = feats_maximal,
  
  #7.3.10 Maximal set with the four CBCL scales replaced by the DP raw sum
  full_maximal_dpsum = c(
    setdiff(feats_maximal, feats_cbcl),
    "cbcl_dp_raw_sum"
  ),

  #7.3.11 Published feature set with the four CBCL scales removed entirely
  anchor_nocbcl = setdiff(feats_anchor, feats_cbcl),

  #7.3.12 Maximal feature set with the four CBCL scales removed entirely
  full_maximal_nocbcl = setdiff(feats_maximal, feats_cbcl)
)

#7.4 Label each configuration with what it varies, for readable output
config_meta <- tibble::tribble(
  ~config,                ~varies_from_base,
  "anchor",               "Published 17 features (validity check)",
  "bas2_fhx",             "BASE: + BAS RR & Drive, + FHx psychosis & hospitalization",
  "bas4_fhx",             "BASE + BIS & BAS Fun Seeking",
  "cbcl3_dim",            "BASE, DSM-5 dep+anx swapped for Anxious/Depressed syndrome",
  "cbcl3_bin",            glue("BASE, 4 CBCL scales replaced by binary DP (all T >= {params$cbcl_dp_cutoff})"),
  "cbcl_dp_rawsum",       "BASE, 4 CBCL scales replaced by DP raw sum (anxdep + attn + aggr)",
  "cbcl_dp_pc1",          "BASE, 4 CBCL scales replaced by DP PC1 (scaled PCA of 3 raw sums)",
  "ace_split",            "BASE, ACE partitioned into 3 abuse binaries + reduced sum",
  "full_maximal",         "All accepted additions combined, CBCL held as-is",
  "full_maximal_dpsum",   "MAXIMAL, 4 CBCL scales replaced by DP raw sum",
  "anchor_nocbcl",        "ANCHOR with all 4 CBCL scales REMOVED",
  "full_maximal_nocbcl",  "MAXIMAL with all 4 CBCL scales REMOVED"
)

#7.5 Record which features each configuration introduces relative to the base, for the decision aid
config_introduces <- list(
  anchor              = character(0),
  bas2_fhx            = c(feats_bas2, feats_fhx),
  bas4_fhx            = setdiff(feats_bas4, feats_bas2),
  cbcl3_dim           = "mh_p_cbcl__synd__anxdep_tscore",
  cbcl3_bin           = "cbcl_dysregulation_binary",
  cbcl_dp_rawsum      = "cbcl_dp_raw_sum",
  cbcl_dp_pc1         = "cbcl_dp_pc1",
  ace_split           = c("ACE_index_reduced", feats_abuse),
  full_maximal        = c(setdiff(feats_bas4, feats_bas2), "ACE_index_reduced", feats_abuse),
  full_maximal_dpsum  = c(setdiff(feats_bas4, feats_bas2), "ACE_index_reduced", feats_abuse, "cbcl_dp_raw_sum"),
  anchor_nocbcl       = character(0),
  full_maximal_nocbcl = character(0)
)

#8. Verify every column required by any configuration is present, excepting PC1 which is derived below
missing_cols <- setdiff(
  setdiff(unique(c(unlist(config_feats), sdsc_subscales)), "cbcl_dp_pc1"),
  names(risk_variable_data)
)
if (length(missing_cols) > 0) {
  stop(glue("Missing expected columns: {paste(missing_cols, collapse = ', ')}"))
}

#9. Ensure dichotomization of categorical data for clustering
#9.1 Assemble the full categorical feature list across all configurations
feats_categorical <- c("family_history_depression", "family_history_mania", "bullying",
                       feats_fhx, feats_abuse, "cbcl_dysregulation_binary")

#9.2 Coerce to factor
risk_variable_data <- risk_variable_data %>%
  mutate(across(all_of(feats_categorical), as.factor))

#10. Derive the DP first principal component on the rows that will form the analytic sample
#10.1 Identify the union of all features excluding PC1, which does not yet exist
feats_union <- unique(unlist(config_feats))
feats_union_pre_pca <- setdiff(feats_union, "cbcl_dp_pc1")

#10.2 Identify rows complete on every other clustering feature; PC1 will be complete on exactly these rows
pca_rows <- complete.cases(risk_variable_data[, feats_union_pre_pca])

#10.3 Fit the PCA on centered and scaled raw subscale sums, per Giampetruzzi et al. (2026)
cbcl_dp_pca <- prcomp(risk_variable_data[pca_rows, feats_cbcl_dp_raw], center = TRUE, scale. = TRUE)

#10.4 Orient PC1 so higher scores indicate greater dysregulation, since prcomp sign is arbitrary
pc1_scores <- cbcl_dp_pca$x[, 1]
if (cor(pc1_scores, risk_variable_data$cbcl_dp_raw_sum[pca_rows]) < 0) {
  cbcl_dp_pca$rotation[, 1] <- -cbcl_dp_pca$rotation[, 1]
  cbcl_dp_pca$x[, 1] <- -cbcl_dp_pca$x[, 1]
  pc1_scores <- cbcl_dp_pca$x[, 1]
}

#10.5 Write PC1 back onto the data frame, NA outside the analytic rows
risk_variable_data$cbcl_dp_pc1 <- NA_real_
risk_variable_data$cbcl_dp_pc1[pca_rows] <- pc1_scores

#11. Restrict to rows complete across the union of all configuration features
#11.1 Record the pre-restriction n
n_before <- nrow(risk_variable_data)

#11.2 Drop rows incomplete on clustering features only, leaving SDSC subscale missingness alone
risk_variable_data <- risk_variable_data %>%
  filter(if_all(all_of(feats_union), ~ !is.na(.)))

#11.3 Report and log the retained n
n_after <- nrow(risk_variable_data)
cat(glue("Rows retained: {n_after} of {n_before} ({n_before - n_after} dropped on new feature missingness)\n"))
log_event("PREP", scaling_method, glue("n_before={n_before}"), glue("n_after={n_after}"))

#12. Guard cached solutions against a silently changed analytic row set
#12.1 The cached kproto fits are only valid if the retained participants are identical
fingerprint_file <- file.path(prelim_dir, sprintf("row_fingerprint_%s.rds", scaling_method))

#12.2 Compare against the stored fingerprint, or write it on first run
if (file.exists(fingerprint_file) && isFALSE(params$overwrite)) {
  stored_ids <- readRDS(fingerprint_file)
  if (!identical(sort(as.character(stored_ids)), sort(as.character(risk_variable_data$participant_id)))) {
    stop(glue(
      "Analytic row set has changed since the cached solutions were written ",
      "(cached n={length(stored_ids)}, current n={nrow(risk_variable_data)}). ",
      "Cached kproto fits are no longer comparable. Re-run with overwrite = TRUE, ",
      "or delete the contents of {prelim_dir}."
    ))
  }
  cat(glue("Row fingerprint matches cached solutions (n = {length(stored_ids)}); cached fits will be reused\n"))
} else {
  saveRDS(risk_variable_data$participant_id, fingerprint_file)
  cat(glue("Row fingerprint written for n = {nrow(risk_variable_data)}\n"))
}

#13. Build the clustering frame with participant IDs as row names
#13.1 Subset to the union of clustering features
risk_variable_clustering_data <- risk_variable_data[, feats_union, drop = FALSE]

#13.2 Set subject IDs as row names
row.names(risk_variable_clustering_data) <- risk_variable_data$participant_id

#14. Identify continuous and categorical variables
continuous_vars <- names(risk_variable_clustering_data)[sapply(risk_variable_clustering_data, is.numeric)]
categorical_vars <- names(risk_variable_clustering_data)[sapply(risk_variable_clustering_data, is.factor)]

#15. Scaling function to be matched with established param
apply_scaling <- function(df, method) {
  out <- df
  if (method != "none") {
    cont_cols <- intersect(continuous_vars, names(df))
    cont <- df[cont_cols]
    scaled <- switch(method,
      z_score = scale(cont),
      min_max = purrr::map_df(cont, ~ (.x - min(.x, na.rm=TRUE)) / (max(.x, na.rm=TRUE) - min(.x, na.rm=TRUE))),
      percentile = purrr::map_df(cont, ~ rank(.x, na.last="keep") / sum(!is.na(.x)) * 100),
      max_absolute = purrr::map_df(cont, ~ .x / max(abs(.x), na.rm=TRUE)),
      robust = purrr::map_df(cont, ~ (.x - median(.x, na.rm=TRUE)) / (IQR(.x, na.rm=TRUE) %||% 1))
    )
    out[cont_cols] <- scaled
  }
  out
}

#16. Scale each configuration according to the param set in the script call
configs <- purrr::map(config_feats, ~ apply_scaling(risk_variable_clustering_data[.x], scaling_method))

#17. Report the composition of each configuration
purrr::imap_dfr(configs, ~ tibble(
  config = .y,
  n_features = ncol(.x),
  n_numeric = sum(sapply(.x, is.numeric)),
  n_categorical = sum(sapply(.x, is.factor))
)) %>%
  left_join(config_meta, by = "config") %>%
  knitr::kable(caption = "Feature composition by configuration") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#18. Confirm the expected feature counts before proceeding
stopifnot(
  ncol(configs$anchor) == 17,
  ncol(configs$anchor_nocbcl) == 13,
  ncol(configs$cbcl_dp_rawsum) == 18,
  ncol(configs$cbcl_dp_pc1) == 18,
  ncol(configs$full_maximal_nocbcl) == ncol(configs$full_maximal) - 4,
  ncol(configs$full_maximal_dpsum) == ncol(configs$full_maximal) - 3
)

```

## CBCL Dysregulation Profile Variables

Descriptive characterization of the two continuous DP operationalizations from Giampetruzzi et al. (2026), computed on the analytic sample. The raw sum implicitly weights subscales by their item count (Anxious/Depressed 13 items, Attention Problems 10, Aggressive Behavior 18), whereas the scaled PCA weights them by their loadings.

```{r dp variables, warning = FALSE, message = FALSE, echo = FALSE}

## Distributional and structural characterization of the DP raw sum and PC1 ##

#1. PCA summary: variance explained and loadings
#1.1 Variance explained by each component
pca_var <- tibble(
  component = paste0("PC", seq_along(cbcl_dp_pca$sdev)),
  sd = cbcl_dp_pca$sdev,
  variance_explained = cbcl_dp_pca$sdev^2 / sum(cbcl_dp_pca$sdev^2),
  cumulative = cumsum(cbcl_dp_pca$sdev^2 / sum(cbcl_dp_pca$sdev^2))
)

#1.2 Print variance explained
pca_var %>%
  knitr::kable(caption = "PCA of the three CBCL DP raw subscale sums (centered, scaled): variance explained", digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#1.3 Print loadings, with Giampetruzzi's YSR-DP T1 loadings for reference (0.57, 0.57, 0.59)
as.data.frame(cbcl_dp_pca$rotation) %>%
  tibble::rownames_to_column("subscale") %>%
  knitr::kable(caption = "PCA loadings (Giampetruzzi et al. 2026 YSR-DP PC1 loadings at T1: anxdep 0.57, attn 0.57, aggr 0.59)", digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#1.4 Log the PC1 variance explained
log_event("DP_PCA", glue("PC1_var={round(pca_var$variance_explained[1], 3)}"))

#2. Implicit subscale weighting in the raw sum vs the PCA
#2.1 Raw sum weights are proportional to each subscale's SD; PCA weights are loading / subscale SD on the raw scale
subscale_sd <- sapply(risk_variable_data[feats_cbcl_dp_raw], sd)
tibble(
  subscale = feats_cbcl_dp_raw,
  raw_sd = subscale_sd,
  raw_sum_relative_weight = subscale_sd / min(subscale_sd),
  pc1_loading = cbcl_dp_pca$rotation[, 1],
  pc1_relative_weight_on_raw_scale = (cbcl_dp_pca$rotation[, 1] / subscale_sd) / min(cbcl_dp_pca$rotation[, 1] / subscale_sd)
) %>%
  knitr::kable(caption = "Implicit relative weighting of each subscale under the raw sum vs PC1", digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#3. Distributional descriptives for the DP variables and their constituent subscales
#3.1 Create a helper returning a full descriptive row
describe_var <- function(data, var) {
  x <- data[[var]]
  x <- x[!is.na(x)]
  tibble(
    variable = var,
    n = length(x),
    mean = mean(x), sd = sd(x), median = median(x), mad = mad(x),
    min = min(x), p25 = quantile(x, 0.25), p75 = quantile(x, 0.75), max = max(x),
    skew = moments::skewness(x), kurtosis = moments::kurtosis(x) - 3,
    pct_at_min = 100 * mean(x == min(x))
  )
}

#3.2 Describe the DP variables, their raw constituents, and the currently used T-scores for reference
dp_describe_vars <- c("cbcl_dp_raw_sum", "cbcl_dp_pc1", feats_cbcl_dp_raw, feats_cbcl, "mh_p_cbcl__synd__anxdep_tscore")
dp_descriptives <- purrr::map_dfr(dp_describe_vars, ~ describe_var(risk_variable_data, .x))

#3.3 Print
dp_descriptives %>%
  knitr::kable(caption = "Distributional descriptives: DP variables, raw subscale sums, and current T-scores (pct_at_min = floor rate)", digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#3.4 Write to CSV
readr::write_csv(dp_descriptives, file.path(prelim_dir, "cbcl_dp_descriptives.csv"))

#4. Correlation among DP variables, constituents, and currently used CBCL scales
cor_vars <- c("cbcl_dp_raw_sum", "cbcl_dp_pc1", feats_cbcl_dp_raw, feats_cbcl, "mh_p_cbcl__synd__anxdep_tscore")
cor_mat <- cor(risk_variable_data[cor_vars], use = "complete.obs")

#4.1 Print the correlation matrix
round(cor_mat, 3) %>%
  knitr::kable(caption = "Pearson correlations among DP variables, raw subscales, and current CBCL T-scores") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 11)

#4.2 State the key correlation explicitly
cat(glue("\nCorrelation between DP raw sum and DP PC1: {round(cor_mat['cbcl_dp_raw_sum', 'cbcl_dp_pc1'], 4)}\n"))

#5. Distribution plots
#5.1 Histograms of the two DP variables and the three raw subscales
dp_plot_data <- risk_variable_data %>%
  dplyr::select(all_of(c("cbcl_dp_raw_sum", "cbcl_dp_pc1", feats_cbcl_dp_raw))) %>%
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  mutate(variable = factor(variable, levels = c("cbcl_dp_raw_sum", "cbcl_dp_pc1", feats_cbcl_dp_raw)))

dp_hist <- ggplot(dp_plot_data, aes(x = value)) +
  geom_histogram(bins = 40, fill = "grey55", color = "white") +
  facet_wrap(~ variable, scales = "free", nrow = 2) +
  labs(x = NULL, y = "Count", title = "Distributions of DP variables and raw subscale sums") +
  theme_minimal(base_size = 11)

print(dp_hist)

#5.2 Scatter of raw sum vs PC1
dp_scatter <- ggplot(risk_variable_data, aes(x = cbcl_dp_raw_sum, y = cbcl_dp_pc1)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick", linewidth = 0.8) +
  labs(x = "DP raw sum", y = "DP PC1",
       title = glue("DP raw sum vs PC1 (r = {round(cor_mat['cbcl_dp_raw_sum', 'cbcl_dp_pc1'], 3)})")) +
  theme_minimal(base_size = 11)

print(dp_scatter)

#5.3 Overlay the raw sum distribution against the DP binary cutoff group for context
risk_variable_data %>%
  mutate(dp_binary = factor(cbcl_dysregulation_binary, labels = c("Below cutoff", "At/above cutoff"))) %>%
  ggplot(aes(x = cbcl_dp_raw_sum, fill = dp_binary)) +
  geom_histogram(bins = 40, position = "identity", alpha = 0.6, color = "white") +
  scale_fill_manual(values = c("grey60", "firebrick")) +
  labs(x = "DP raw sum", y = "Count", fill = glue("Binary DP (all 3 T >= {params$cbcl_dp_cutoff})"),
       title = "Where the binary DP cutoff falls on the continuous raw sum") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

```

**Reading this section:**

- If PC1 explains roughly two thirds of variance or more with near-equal loadings, the three subscales behave as a single dimension, consistent with Giampetruzzi (67 to 77%, loadings 0.55 to 0.63) and Deutz's bifactor general factor
- If the raw sum and PC1 correlate above roughly 0.97, they are functionally one variable for clustering purposes and the choice between them is presentational
- `pct_at_min` on the raw sum is the floor rate. A large floor in a population sample is expected and is the analogue of T-score truncation at 50
- The implicit weighting table shows how much more the raw sum leans on Aggressive Behavior than PC1 does

## Fit All Configurations

All configurations use `nstart = 8` and the package default data-driven lambda. Configurations already fitted in a prior run load from cache; only the two DP configurations require new fits.

```{r clustering, warning = FALSE, message = FALSE, echo = FALSE}

## Fit k-prototypes across k for all configurations ##

#1. Create a vector containing the k's to test
k_vals <- 2:params$k_max

#2. Fit or load each configuration
kp_lists <- purrr::imap(configs, function(dat, cfg) {

  #2.1 List where to stash / load the kproto list for this configuration
  proto_file <- file.path(prelim_dir, sprintf("kproto_%s_%s.rds", cfg, scaling_method))

  #2.2 Load cached kproto list if it already exists and overwrite is not requested
  if (file.exists(proto_file) && isFALSE(params$overwrite)) {
    log_event("PROTO_SKIP", cfg, scaling_method, glue("loaded {length(k_vals)} ks from cache"))
    cat(glue("{cfg}: loaded from cache\n"))
    return(readRDS(proto_file))
  }

  #2.3 Otherwise start the kproto timer and compute
  proto_start <- Sys.time()
  log_event("PROTO_START", cfg, scaling_method)
  cat(glue("{cfg}: fitting {length(k_vals)} solutions\n"))

  #2.3.1 Execute kproto across k, leaving lambda at the package default
  kp_list <- future_map(
    k_vals,
    ~ kproto(x = dat,
             k = .x,
             nstart = 8,
             verbose = FALSE),
    .options = furrr_options(seed = TRUE)
  )

  #2.3.2 Provide a k suffix to each of the k values
  names(kp_list) <- paste0("k", k_vals)

  #2.3.3 Save the kproto list
  saveRDS(kp_list, proto_file)

  #2.3.4 Log end status for clustering
  proto_end <- Sys.time()
  log_event("PROTO_END", cfg, scaling_method,
            glue("duration={round(as.numeric(difftime(proto_end, proto_start, units = 'mins')), 1)}min"))

  kp_list
})

#3. Extract the data-driven lambda actually used by each configuration for reporting
lambda_used <- purrr::map_dbl(kp_lists, ~ mean(.x[["k2"]]$lambda))
cat("Data-driven lambda by configuration:\n"); print(round(lambda_used, 3))
log_event("LAMBDA", paste(names(lambda_used), round(lambda_used, 3), sep = "=", collapse = "|"))

#4. Align cluster labels across configurations so the higher-risk profile is always level 1
#4.1 Cluster numbering from kproto is arbitrary, so orient every solution by sleep disturbance
align_labels <- function(labels, orient_var) {
  means <- tapply(orient_var, labels, mean, na.rm = TRUE)
  higher <- names(which.max(means))
  factor(ifelse(as.character(labels) == higher, "Higher-risk", "Lower-risk"),
         levels = c("Higher-risk", "Lower-risk"))
}

#4.2 Build an aligned k = 2 label vector for every configuration
aligned_k2 <- purrr::map(kp_lists, ~ align_labels(.x$k2$cluster, risk_variable_data$sds_total))

#4.3 Store the anchor and base partitions for agreement testing
anchor_labels <- aligned_k2$anchor
base_labels <- aligned_k2[[params$base_config]]

#4.4 Confirm the orientation worked by reporting mean sleep disturbance per aligned cluster
purrr::imap_dfr(aligned_k2, ~ tibble(
  config = .y,
  mean_sds_higher = mean(risk_variable_data$sds_total[.x == "Higher-risk"]),
  mean_sds_lower = mean(risk_variable_data$sds_total[.x == "Lower-risk"]),
  n_higher = sum(.x == "Higher-risk")
)) %>%
  knitr::kable(caption = "Label alignment check: higher-risk cluster oriented by SDSC total", digits = 2) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```

## Internal Validation Across k

Gower silhouette is computed on the full analytic sample. The distance matrix depends only on the feature set, not on *k*, so it is computed once per configuration and reused across all candidate solutions.

```{r validation, warning = FALSE, message = FALSE, echo = FALSE}

## Silhouette and WSS across k for all configurations ##

#1. Create a helper computing mean Gower silhouette across all k for one configuration
#1.1 Gower distance is computed once and reused, since it depends only on the feature set
silhouette_across_k <- function(dat, kp_list, cfg) {

  #1.1.1 Compose the output path, keyed by method so full and subsampled caches do not collide
  sil_file <- file.path(prelim_dir, sprintf("silhouette_%s_%s_%s.rds", params$sil_method, cfg, scaling_method))

  #1.1.2 Return the cached vector if it exists and overwrite is not requested
  if (file.exists(sil_file) && isFALSE(params$overwrite)) {
    log_event("SIL_SKIP", cfg, params$sil_method)
    return(readRDS(sil_file))
  }

  #1.1.3 Start the timer
  sil_start <- Sys.time()
  log_event("SIL_START", cfg, params$sil_method, glue("n={nrow(dat)}"))
  cat(glue("{cfg}: computing {params$sil_method}-sample Gower silhouette\n"))

  if (identical(params$sil_method, "full")) {

    #1.1.4 Compute the full-sample Gower distance matrix once
    dd <- cluster::daisy(dat, metric = "gower")

    #1.1.5 Evaluate every candidate k against the same distance matrix
    sils <- vapply(kp_list, function(kp) {
      lab <- as.integer(factor(kp$cluster))
      if (length(unique(lab)) < 2) return(NA_real_)
      mean(cluster::silhouette(lab, dd)[, "sil_width"])
    }, numeric(1))

    #1.1.6 Release the distance matrix before moving to the next configuration
    rm(dd); invisible(gc())

  } else {

    #1.1.7 Fall back to the subsampled estimator when requested
    sils <- vapply(kp_list, function(kp) {
      set.seed(params$seed)
      vals <- vapply(seq_len(params$sil_subsample_reps), function(r) {
        idx <- sample.int(nrow(dat), min(params$sil_subsample_n, nrow(dat)))
        lab <- as.integer(factor(kp$cluster[idx]))
        if (length(unique(lab)) < 2) return(NA_real_)
        dsub <- cluster::daisy(dat[idx, , drop = FALSE], metric = "gower")
        mean(cluster::silhouette(lab, dsub)[, "sil_width"])
      }, numeric(1))
      mean(vals, na.rm = TRUE)
    }, numeric(1))
  }

  #1.1.8 Persist and log
  saveRDS(sils, sil_file)
  log_event("SIL_END", cfg, params$sil_method,
            glue("duration={round(as.numeric(difftime(Sys.time(), sil_start, units = 'mins')), 1)}min"))

  sils
}

#2. Compute silhouette for every configuration, sequentially to bound peak memory
silhouette_by_config <- purrr::imap(configs, ~ silhouette_across_k(.x, kp_lists[[.y]], .y))

#3. Bind silhouette and WSS trajectories into a single frame
metrics_by_k <- purrr::imap_dfr(kp_lists, function(kp_list, cfg) {
  tibble(
    config = cfg,
    k = k_vals,
    gower_silhouette = as.numeric(silhouette_by_config[[cfg]]),
    wss = purrr::map_dbl(kp_list, "tot.withinss")
  )
})

#4. Optionally run the full Huang silhouette via validation_kproto
#4.1 This is O(n^2) per k and can take days at this sample size, so it is off by default
if (isTRUE(params$run_validation_kproto)) {
  validation_results <- purrr::imap(configs, function(dat, cfg) {
    validation_file <- file.path(prelim_dir, sprintf("val_%s_%s_silhouette.rds", cfg, scaling_method))
    if (file.exists(validation_file) && isFALSE(params$overwrite)) {
      log_event("VAL_SKIP", cfg, scaling_method, "silhouette")
      return(readRDS(validation_file))
    }
    vr_sil <- validation_kproto(method = "silhouette", data = dat, k = k_vals,
                                kp_obj = "all", nstart = 8, verbose = FALSE)
    saveRDS(vr_sil, validation_file)
    vr_sil
  })
  metrics_by_k <- metrics_by_k %>%
    mutate(huang_silhouette = purrr::map2_dbl(config, k, ~ as.numeric(validation_results[[.x]]$indices)[which(k_vals == .y)]))
}

#5. Print the silhouette trajectory table
metrics_by_k %>%
  dplyr::select(config, k, gower_silhouette) %>%
  tidyr::pivot_wider(names_from = config, values_from = gower_silhouette) %>%
  knitr::kable(
    caption = glue("Gower silhouette by k ({params$sil_method} sample, n = {nrow(risk_variable_data)}, {scaling_method} scaling)"),
    digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#6. Plot the silhouette trajectories, faceted so each configuration is legible
#6.1 Generate the faceted silhouette plot
silhouette_plot <- metrics_by_k %>%
  mutate(config = factor(config, levels = names(configs))) %>%
  ggplot(aes(x = k, y = gower_silhouette, group = config)) +
  geom_line(linewidth = 0.8, color = "grey40") +
  geom_point(size = 1.8, color = "grey20") +
  geom_vline(xintercept = 2, linetype = "dashed", color = "firebrick") +
  facet_wrap(~ config, nrow = 3) +
  scale_x_continuous(breaks = k_vals) +
  labs(x = "Number of clusters (k)", y = "Mean Gower silhouette (full sample)",
       title = "Silhouette trajectories by candidate feature set") +
  theme_minimal(base_size = 12)

#6.2 Print the silhouette plot
print(silhouette_plot)

```

## Agreement and Stability at k = 2

```{r agreement, warning = FALSE, message = FALSE, echo = FALSE}

## Agreement with the anchor and base partitions, plus bootstrap stability at k = 2 ##

#1. Create a helper running the bootstrap stability analysis for a given kproto object
run_stability <- function(kp, cfg) {

  #1.1 Compose the output path
  stab_file <- file.path(prelim_dir, sprintf("stability_%s_%s_k2_nb%d.rds", cfg, scaling_method, params$n_boot))

  #1.2 Return the cached object if it exists and overwrite is not requested
  if (file.exists(stab_file) && isFALSE(params$overwrite)) {
    log_event("STAB_SKIP", cfg, basename(stab_file))
    return(readRDS(stab_file))
  }

  #1.3 Strip row names from the embedded data since stability_kproto() trips over them
  if (!is.null(rownames(kp$data))) {
    rownames(kp$data) <- NULL
  }

  #1.4 Build the call safely, since the bootstrap arg name varies across clustMixType versions
  stab_formals <- names(formals(clustMixType::stability_kproto))
  stab_args <- list(
    object = kp,
    method = c("rand", "jaccard"),
    nstart = params$stab_nstart,
    verbose = FALSE
  )
  if ("B" %in% stab_formals) {
    stab_args$B <- params$n_boot
  } else if ("nboot" %in% stab_formals) {
    stab_args$nboot <- params$n_boot
  } else {
    log_event("STAB_WARN", cfg, "unknown_boot_arg", "default_used")
  }

  #1.5 Run the bootstrap stability analysis
  log_event("STAB_START", cfg, glue("B={params$n_boot}"))
  boot_res <- tryCatch(
    do.call(clustMixType::stability_kproto, stab_args),
    error = function(e) stop(glue("stability_kproto() failed for {cfg}: {conditionMessage(e)}"))
  )

  #1.6 Persist the raw result and log completion
  saveRDS(boot_res, stab_file)
  log_event("STAB_DONE", cfg, glue("B={params$n_boot}"), glue("saved={basename(stab_file)}"))

  boot_res
}

#2. Run stability for all configurations at k = 2
stability_results <- purrr::imap(kp_lists, ~ run_stability(.x$k2, .y))

#3. Unpack the stability output into tidy per bootstrap tables
boot_tbls <- purrr::imap_dfr(stability_results, function(boot_res, cfg) {
  bt <- as.data.frame(boot_res[[2]])
  colnames(bt) <- c("B", "Rand", "Jaccard")
  bt$config <- cfg
  bt
})

#4. Assemble the decision table
agreement_tbl <- purrr::imap_dfr(kp_lists, function(kp_list, cfg) {

  #4.1 Pull the k = 2 fit, aligned labels, and bootstrap results
  kp2 <- kp_list$k2
  lab <- aligned_k2[[cfg]]
  overall_vec <- stability_results[[cfg]][[1]]
  bt <- boot_tbls %>% filter(config == cfg)

  #4.2 Compute reassignment against the base, using aligned labels
  reassigned_base <- sum(lab != base_labels)

  #4.3 Return a one row tibble with the decision metrics
  tibble(
    config = cfg,
    n_features = ncol(configs[[cfg]]),
    lambda = lambda_used[[cfg]],
    gower_silhouette = metrics_by_k$gower_silhouette[metrics_by_k$config == cfg & metrics_by_k$k == 2],
    ari_vs_anchor = mclust::adjustedRandIndex(as.integer(lab), as.integer(anchor_labels)),
    ari_vs_base = mclust::adjustedRandIndex(as.integer(lab), as.integer(base_labels)),
    pct_reassigned_vs_base = 100 * reassigned_base / length(lab),
    overall_Jacc = unname(overall_vec["J"]),
    mean_Jacc = mean(bt$Jaccard, na.rm = TRUE),
    sd_Jacc = sd(bt$Jaccard, na.rm = TRUE),
    n_higher_risk = sum(lab == "Higher-risk"),
    n_lower_risk = sum(lab == "Lower-risk")
  )
})

#5. Attach the per-configuration optimal k and readable labels
agreement_tbl <- agreement_tbl %>%
  left_join(
    metrics_by_k %>%
      group_by(config) %>%
      slice_max(gower_silhouette, n = 1, with_ties = FALSE) %>%
      dplyr::select(config, k_opt = k),
    by = "config"
  ) %>%
  left_join(config_meta, by = "config") %>%
  mutate(k2_optimal = k_opt == 2)

#6. Write and print the decision tables
#6.1 Write the full decision table to CSV
readr::write_csv(agreement_tbl, file.path(prelim_dir, glue("decision_summary_{scaling_method}_nb{params$n_boot}.csv")))

#6.2 Print the full detail table
agreement_tbl %>%
  dplyr::select(-varies_from_base) %>%
  knitr::kable(caption = "Full detail: all agreement and stability metrics", digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```

## Decision Aid

```{r decision aid, warning = FALSE, message = FALSE, echo = FALSE}

## Compact decision aid across configurations ##

#1. Create a helper returning the effect size of one feature under a given label vector
#1.1 Numeric features return Cohen's d, binary features return an odds ratio
feature_separation <- function(var, labels, data) {

  #1.1.1 Build a temporary frame with the aligned labels
  tmp <- data.frame(cluster = labels, value = data[[var]])
  tmp <- tmp[!is.na(tmp$value), ]

  #1.1.2 Branch on variable type
  if (is.factor(tmp$value) || length(unique(tmp$value)) <= 2) {

    #1.1.2.1 Order the outcome so endorsement is the first column, giving odds of endorsement in higher vs lower risk
    vv <- factor(as.character(tmp$value), levels = c("1", "0"))
    tab <- table(tmp$cluster, vv)
    if (any(dim(tab) < 2) || any(tab == 0)) return(tibble(variable = var, metric = "OR", estimate = NA_real_, d_equivalent = NA_real_))
    orr <- as.numeric(effectsize::oddsratio(tab)$Odds_ratio)
    tibble(variable = var, metric = "OR", estimate = orr,
           d_equivalent = as.numeric(effectsize::oddsratio_to_d(orr)))

  } else {

    #1.1.2.2 Numeric features get Cohen's d, oriented higher-risk minus lower-risk
    dd <- as.numeric(effectsize::cohens_d(value ~ cluster, data = tmp)$Cohens_d)
    tibble(variable = var, metric = "d", estimate = dd, d_equivalent = dd)
  }
}

#2. Compute separation for the features each configuration introduces
separation_tbl <- purrr::imap_dfr(config_introduces, function(vars, cfg) {
  if (length(vars) == 0) return(tibble(config = cfg, variable = NA_character_, metric = NA_character_, estimate = NA_real_, d_equivalent = NA_real_))
  purrr::map_dfr(vars, ~ feature_separation(.x, aligned_k2[[cfg]], risk_variable_data)) %>%
    mutate(config = cfg)
})

#3. Print the per-feature separation detail
separation_tbl %>%
  filter(!is.na(variable)) %>%
  dplyr::select(config, variable, metric, estimate) %>%
  knitr::kable(
    caption = "Separation of newly introduced features, under each configuration's own cluster labels",
    digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#4. Collapse separation into a compact per-configuration summary string
separation_summary <- separation_tbl %>%
  filter(!is.na(variable)) %>%
  group_by(config) %>%
  summarise(
    feature_separation = paste0(sprintf("%s=%.2f", metric, estimate), collapse = "; "),
    .groups = "drop"
  )

#5. Build the headline decision aid
decision_aid <- agreement_tbl %>%
  left_join(separation_summary, by = "config") %>%
  mutate(
    delta_sil_vs_base = gower_silhouette - gower_silhouette[config == params$base_config],
    feature_separation = tidyr::replace_na(feature_separation, "none added")
  ) %>%
  dplyr::select(
    Config = config,
    `What varies` = varies_from_base,
    `N feat` = n_features,
    `k opt` = k_opt,
    `k=2 best?` = k2_optimal,
    `ARI vs base` = ari_vs_base,
    `% moved vs base` = pct_reassigned_vs_base,
    `Silhouette` = gower_silhouette,
    `D sil vs base` = delta_sil_vs_base,
    `Mean Jaccard` = mean_Jacc,
    `Higher-risk n` = n_higher_risk,
    `New feature separation` = feature_separation
  )

#6. Write and print the decision aid
#6.1 Write the decision aid to CSV
readr::write_csv(decision_aid, file.path(prelim_dir, glue("decision_aid_{scaling_method}_nb{params$n_boot}.csv")))

#6.2 Print the decision aid
decision_aid %>%
  knitr::kable(
    caption = glue("DECISION AID: what each candidate change does, relative to the base configuration ({params$base_config})"),
    digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 12)

#7. Print a one line per configuration narrative summary
cat("\n---- Summary ----\n")
for (i in seq_len(nrow(agreement_tbl))) {
  r <- agreement_tbl[i, ]
  cat(glue(
    "{r$config}: {r$n_features} features | k_opt={r$k_opt} | ARI vs base={round(r$ari_vs_base, 3)} | ",
    "{round(r$pct_reassigned_vs_base, 2)}% moved | silhouette={round(r$gower_silhouette, 3)} | ",
    "Jaccard={round(r$mean_Jacc, 3)} | clusters {r$n_higher_risk}/{r$n_lower_risk}\n\n"
  ))
}

```

## Feature Separation Profiles

For each selected configuration, every feature in that configuration is ranked by how strongly it separates the two clusters, under that configuration's own labels. Odds ratios are converted to a Cohen's d equivalent via `effectsize::oddsratio_to_d()` so numeric and binary features can be ranked on a common scale.

```{r separation profiles, warning = FALSE, message = FALSE, results = 'asis', echo = FALSE}

## Rank every feature within selected configurations by cluster separation ##

#1. Restrict to configurations requested via params, that actually exist
profile_configs <- intersect(params$profile_configs, names(configs))

#2. Compute the full separation profile for each requested configuration
profile_tbl <- purrr::map_dfr(profile_configs, function(cfg) {
  purrr::map_dfr(config_feats[[cfg]], ~ feature_separation(.x, aligned_k2[[cfg]], risk_variable_data)) %>%
    mutate(config = cfg) %>%
    arrange(desc(abs(d_equivalent))) %>%
    mutate(rank = row_number())
})

#3. Print one ranked table per configuration
for (cfg in profile_configs) {
  print(
    profile_tbl %>%
      filter(config == cfg) %>%
      dplyr::select(rank, variable, metric, estimate, d_equivalent) %>%
      knitr::kable(
        caption = glue("Feature separation profile: {cfg} ({config_meta$varies_from_base[config_meta$config == cfg]})"),
        digits = 3) %>%
      kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
  )
}

#4. Plot the profiles side by side for visual comparison
#4.1 Generate the profile plot
profile_plot <- profile_tbl %>%
  ggplot(aes(x = abs(d_equivalent), y = reorder(variable, abs(d_equivalent)))) +
  geom_col(fill = "grey55") +
  facet_wrap(~ config, scales = "free_y") +
  labs(x = "Absolute separation (Cohen's d equivalent)", y = NULL,
       title = "Feature separation by configuration") +
  theme_minimal(base_size = 10)

#4.2 Print the profile plot
print(profile_plot)

#5. Write the profile table to CSV
readr::write_csv(profile_tbl, file.path(prelim_dir, glue("separation_profiles_{scaling_method}.csv")))

```

## Descriptives for the Reviewer Response

Computed under the `r params$descriptive_config` configuration, with labels aligned so the higher-risk profile is always listed first.

```{r descriptives, warning = FALSE, message = FALSE, echo = FALSE}

## Cluster descriptives for the reviewer response ##

#1. Attach the selected configuration's aligned cluster labels to the unscaled data
descriptive_data <- risk_variable_data %>%
  mutate(cluster = aligned_k2[[params$descriptive_config]])

#2. Create a helper returning mean, SD, and Cohen's d by cluster for a numeric variable
summarise_numeric_by_cluster <- function(data, var) {

  #2.1 Compute Cohen's d with CI via the effectsize package
  d <- effectsize::cohens_d(stats::reformulate("cluster", response = var), data = data)

  #2.2 Compute per cluster mean and SD, then attach the effect size
  data %>%
    group_by(cluster) %>%
    summarise(summary = sprintf("%.2f ± %.2f", mean(.data[[var]], na.rm = TRUE), sd(.data[[var]], na.rm = TRUE)),
              .groups = "drop") %>%
    tidyr::pivot_wider(names_from = cluster, values_from = summary) %>%
    mutate(variable = var, effect = "Cohen's d", estimate = d$Cohens_d, ci_low = d$CI_low, ci_high = d$CI_high) %>%
    dplyr::select(variable, everything())
}

#3. Create a helper returning endorsement and odds ratio by cluster for a binary variable
summarise_binary_by_cluster <- function(data, var) {

  #3.1 Order the outcome so the odds ratio reflects endorsement in higher vs lower risk
  vv <- factor(as.character(data[[var]]), levels = c("1", "0"))
  tab <- table(data$cluster, vv)

  #3.2 Guard against empty cells, which make the odds ratio undefined
  if (any(tab == 0)) {
    est <- NA_real_; lo <- NA_real_; hi <- NA_real_
  } else {
    orr <- effectsize::oddsratio(tab)
    est <- orr$Odds_ratio; lo <- orr$CI_low; hi <- orr$CI_high
  }

  #3.3 Compute per cluster endorsement counts and percentages
  data %>%
    group_by(cluster) %>%
    summarise(summary = sprintf("%d (%.1f%%)",
                                sum(.data[[var]] == 1, na.rm = TRUE),
                                100 * mean(.data[[var]] == 1, na.rm = TRUE)),
              .groups = "drop") %>%
    tidyr::pivot_wider(names_from = cluster, values_from = summary) %>%
    mutate(variable = var, effect = "Odds ratio", estimate = est, ci_low = lo, ci_high = hi) %>%
    dplyr::select(variable, everything())
}

#4. Summarise all numeric candidate features by cluster
purrr::map_dfr(
  c(feats_bas4, "cbcl_dp_raw_sum", "cbcl_dp_pc1", "mh_p_cbcl__synd__anxdep_tscore", feats_cbcl,
    "ACE_index_sum_score", "ACE_index_reduced", "sds_total", sdsc_subscales),
  ~ summarise_numeric_by_cluster(descriptive_data, .x)
) %>%
  knitr::kable(
    caption = "Numeric candidate features by cluster (SDSC subscales descriptive only, not clustered)",
    digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#5. Summarise the binary candidate features by cluster
purrr::map_dfr(
  c(feats_fhx, feats_abuse, "cbcl_dysregulation_binary"),
  ~ summarise_binary_by_cluster(descriptive_data, .x)
) %>%
  knitr::kable(
    caption = "Binary candidate features by cluster (NA odds ratio indicates an empty cell)",
    digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```

**How to read this run:**

- Compare `cbcl_dp_rawsum` and `cbcl_dp_pc1` against `cbcl3_bin`. All three replace four CBCL features with one. If the two continuous versions hold *k* = 2 with stability near the base while the binary does not, the `cbcl3_bin` collapse was caused by dichotomization rather than by feature reduction. If the continuous versions also degrade, feature count is doing some of the work.
- Compare `cbcl_dp_rawsum` and `cbcl_dp_pc1` against each other. If their raw sum to PC1 correlation is above roughly 0.97, expect near-identical partitions and treat them as one option.
- The **Feature Separation Profiles** for the two DP configurations show whether a single DP variable still dominates the geometry the way the four separate scales do.
- Bootstrap replicates remain at `r params$n_boot`; rerun the chosen configuration at `n_boot = 1000` before reporting externally.
