---
title: "Mood Disorder & Suicidality Risk Group Cluster Characterizations"
author: "Sam A. Sievertsen"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(warning = FALSE, message = NA, comment = "")

```

```{r environment, echo = FALSE, include = FALSE, warning = FALSE}

# Load necessary packages + environmental variables
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(DescTools)
library(skimr)
library(easystats)
library(cluster)
library(clustMixType)
library(factoextra)
library(NbClust)
library(clValid) 
library(ClusterR) 
library(clusterSim)
library(clustertend)
library(hopkins)
library(tidymodels)
library(tidyclust)
library(tidyquant)
library(ggplot2)
library(plotly)
options(scipen = 999, digits = 8)

# Read in risk variable data
risk_variable_data <- read.csv("../../data/preliminary_analysis/data_processed/risk_variable_data.csv")

# Read in clustering results
kproto_results <- readRDS(file = "../../results/preliminary_analysis/kproto_clustering_results.rds")

# Read in individual silhouette scores
silhouette_data <- read.csv("../../results/preliminary_analysis/kproto_individual_silhouette_scores.csv")

# Read in k grouped silhouette means
silhouette_scores <- read.csv("../../results/preliminary_analysis/kproto_silhouette_means.csv")

# Read in elbow plot data
elbow_plot_data <- read.csv("../../results/preliminary_analysis/kproto_elbow_plot_data.csv")

```

```{r data wrangling, echo = FALSE, warning = FALSE}

## Data Wrangling ## 

#1. Z-score continuous data for clustering
#1.1 Create a list of continuous variables to z-score
variables_to_scale <- c(
  "cbcl_scr_dsm5_depress_t",
  "cbcl_scr_dsm5_anxdisord_t",
  "cbcl_scr_syn_attention_t",
  "cbcl_scr_syn_aggressive_t",
  "pgbi_p_ss_score",
  "upps_y_ss_negative_urgency",
  "upps_y_ss_positive_urgency",
  "sds_p_ss_total",
  "nsc_p_ss_mean_3_items",
  "nihtbx_list_uncorrected",
  "nihtbx_flanker_uncorrected",
  "nihtbx_pattern_uncorrected",
  "ACE_index_sum_score")

#1.2 Z-score all continuous variables to be used in clustering
risk_variable_data <- risk_variable_data %>%
  mutate(across(
    all_of(variables_to_scale),
    ~ as.numeric(scale(.)),
    .names = "{.col}_scaled"))

#2. Ensure dichotimization of categorical data for clustering
risk_variable_data <- risk_variable_data %>% 
  mutate(across(c("family_history_depression", "famhx_ss_momdad_ma_p", "bullying"), as.factor))

```

## Review of Clustering Results and Selection of Optimal *k*

Results from the `risk_group_clustering` script suggest it is likely that optimal *k* = 5. Re-iterating the most important information here, this determination was made because the elbow plot suggests a natural elbow wherein reduction in total within sum of squares values levels off after *k* = 5 (*note*: a similar tapering is observed after *k* = 4, but the distinction between the two *k* values will be made in the discussion of the silhouette index results): 

```{r elbow plot, echo = FALSE, warning = FALSE, fig.width = 10, fig.height = 6}

#1. Generate and print the Elbow Plot
#1.1 Generate the elbow plot
elbow_plot <- ggplot(elbow_plot_data, aes(x = k, y = Tot_WithinSS)) +
  geom_line() +
  geom_point() +
  labs(title = "Elbow Plot", x = "Number of Clusters (k)", y = "Total Within Sum of Squares") +
  theme_minimal()

#1.2 Print the Elbow Plot
print(elbow_plot)

```

In addition to the elbow plot, wherein *k = 5* is observed to exhibit the qualities of an elbow, the *k* grouped mean silhouette scores indicate that *k = 5* is the third most optimal clustering solution (μ = 0.113; following *k = 4* closely, wherein μ = 0.123): 

```{r silhouette score means, echo = FALSE, warning = FALSE}

#1. Print the silhouette scores for reference in output
knitr::kable(silhouette_scores, label = "Daisy + cluster::silhouette derived silhouette scores", col.names = c("k", "Mean Silhouette Score"))

```

At this point, despite the slightly greater reduction in total within sum of squares values, it is not evident whether *k = 5* is more optimal than *k = 4*. This becomes more clear, though, when we examine the density plot of silhouette scores at the individual subject within-cluster level: 

```{r silhouette density plot, echo = FALSE, warning = FALSE, fig.width = 10, fig.height = 6}

#1. Generate a silhouette width plot
#1.1 Calculate mean silhouette widths for each cluster and k
mean_silhouette <- silhouette_data %>%
  group_by(k, cluster) %>%
  summarize(mean_width = mean(silhouette_width), .groups = "drop")

#1.2 Create the silhouette density/mean plot
silhouette_plot <- ggplot(silhouette_data, aes(x = silhouette_width, fill = factor(cluster))) +
  geom_density(alpha = 0.6, color = NA) + 
  geom_vline(data = mean_silhouette, aes(xintercept = mean_width, color = factor(cluster)), 
             linetype = "dashed", size = 0.8) + 
  facet_wrap(~k, ncol = 3, scales = "free", labeller = labeller(k = function(k) paste("k =", k))) + 
  labs(
    title = "Silhouette Plot with Density and Mean Widths",
    x = "Silhouette Width",
    y = "Density",
    fill = "Cluster",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom", 
    strip.text = element_text(size = 12)
  )

#1.3 Display the silhouette density/mean plot
print(silhouette_plot)

```

### Assessment of Current Results Using Huang Distance

At *k = 5*, the tight grouping of clusters about silhouette = 0.20 for clusters 1 and 3 and okay (albeit lower and more spread) performance of cluster 2 and 4 (about silhouette = 0) is superior to *k = 4*. Cluster 5 does not indicate good performance according to the silhouette index, but this limitation will be discussed below.

At *k = 4*, cluster 3 performs well (i.e., is grouped about silhouette = 0.25), but cluster 2 performs worse than cluster 2 in the *k = 5* solution, with clusters 1 and 4 actually mean centered below 0 (indicating that most subjects in that cluster would have been more accurately grouped into another cluster). 

It should be noted that the silhouette index as it is currently used is likely not a great indicator of cluster separation/performance, given it tends to favor clustering solutions wherein separation of the categorical variables are prioritized over separation of the continuous variables (as is best done by the gower distance method). Given this bias, our determination of the optimal k value in this iteration of clustering is partially subjective in nature; and while this is acceptable for first round analyses to be submitted to SOBP, these results are tentative, and need to be reviewed more carefully considering: 

- The inclusion of validation indices that provide indicators of within cluster performance (e.g., C index), cluster separation for both categorical and continuous data (e.g., point-biserial coefficient), and something like the weighted silhouette index which takes distance + variance of both types of data into account; 

- Secondary considerations such as the stability of the clusters (e.g., assessed using stability_kproto()) and gap statistics; and

- Optimization of lambda parameters

prior to any publication/more widespread communication of statistical results derived using these clusters. 

## Characterization of *k* = 5 Clusters

### Integration of Clusters Back Into Risk Variable Data

In order to best characterize the risk groups (i.e., according to their original risk variables on their native scale), we will first need to merge the clustered group values back into the original risk data: 

```{r cluster value merging, warning = FALSE}

## Merge Cluster Group Values Back Into Original Data ##

#1. Extract the cluster assignment values from the kproto results
#1.1 Create a dataframe containing each subjects cluster value
subject_cluster_values <- as.data.frame(kproto_results[["k_5"]][["model"]][["cluster"]])

#1.2 Create a column containing each subjects ID
subject_cluster_values$src_subject_id <- row.names(subject_cluster_values)

#1.3 Clean the column names + retain columns of interest
subject_cluster_values <- subject_cluster_values %>% 
  mutate(cluster = `kproto_results[["k_5"]][["model"]][["cluster"]]`) %>% 
  dplyr::select(c(src_subject_id, cluster))

#2. Merge cluster values back into original data
clustered_risk_variable_data <- left_join(risk_variable_data, subject_cluster_values)

#3. Output the merged risk variable + cluster data
write.csv(clustered_risk_variable_data, "../../preliminary_analysis/data/data_processed/clustered_risk_variable_data.csv", row.names = FALSE)

```

### Profiling the Risk Group Clusters According to Risk Variables

Now that cluster assignments have been merged into the original risk data, we can profile the different clusters according to each categorical (binary) risk variable: 

```{r cluster binary variable profiles, warning = FALSE, fig.width = 10, fig.height = 6}

## Characterize the Clusters According to Binary Risk Variables ##

#1. Create a bar chart for each binary variable
#1.1 Define the binary variables
binary_vars <- c("family_history_depression", "famhx_ss_momdad_ma_p", "bullying")

#1.2 Define variable descriptions
binary_variable_labels <- c(
  family_history_depression = "1st Degree Relative History of Dep",
  famhx_ss_momdad_ma_p = "Parent History of Mania",
  bullying = "Experience of Bullying"
)

#1.3 Wrangle data for plotting
binary_plot_data <- clustered_risk_variable_data %>%
  dplyr::select(cluster, family_history_depression, famhx_ss_momdad_ma_p, bullying) %>%
  pivot_longer(-cluster, names_to = "variable", values_to = "value") %>%
  group_by(cluster, variable, value) %>%
  summarise(count = n(), .groups = "drop")

#1.4 Create the binary variable plot
binary_plot <- binary_plot_data %>%
  ggplot(aes(x = as.factor(value), y = count, fill = as.factor(cluster))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  geom_text(aes(label = count), 
            position = position_dodge(width = 0.8), 
            vjust = -0.5, 
            size = 2.5) + 
  facet_wrap(
    ~variable, 
    scales = "free_y", 
    labeller = as_labeller(binary_variable_labels)
  ) +
  labs(x = "Value", y = "Count", fill = "Cluster",
       title = "Cluster Profiles on Binary Risk Variables") +
  scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) + 
  theme_minimal() +
  theme(strip.text = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "top") +
  scale_fill_brewer(palette = "Set3")

#1.4 Print the binary values plot
print(binary_plot)

#2. Create a summary statistics table for the categorical variables
#2.1 Wrangle data for summary statistics
binary_summary_data <- binary_plot_data %>%
  filter(value == 1) %>%
  dplyr::select(cluster, variable, count, total, percent) %>%
  rename(n_endorsed = count, 
         n_total = total, 
         percent_endorsed = percent) %>%
  mutate(percent_endorsed = round(percent_endorsed, 1),
         display = paste0(n_endorsed, " (", percent_endorsed, "%)"))

#2.2 Create the summary statistics table
binary_summary_data %>%
  dplyr::select(cluster, variable, n_endorsed, n_total, percent_endorsed, display) %>%
  pivot_wider(
    id_cols = variable,
    names_from = cluster,
    values_from = display,
    names_prefix = "Cluster "
  ) %>%
  kbl(caption = "Percentage of Subjects with Endorsed Binary Risk Variables by Cluster") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))

```

As well as profile the clusters according to each continuous risk variable: 

```{r cluster continuous variable profiles, warning = FALSE, fig.width = 10, fig.height = 6}

## Characterize the Clusters According to Continuous Risk Variables ##

#1. Create a violin boxplot for each continuous variable
#1.1 Define the continuous variables and descriptions
continuous_vars <- c("cbcl_scr_dsm5_depress_t", "cbcl_scr_dsm5_anxdisord_t",
                     "cbcl_scr_syn_attention_t", "cbcl_scr_syn_aggressive_t",
                     "pgbi_p_ss_score", "upps_y_ss_negative_urgency",
                     "upps_y_ss_positive_urgency", "sds_p_ss_total",
                     "reshist_addr1_coi_z_coi_nat", "nsc_p_ss_mean_3_items",
                     "nihtbx_list_uncorrected", "nihtbx_flanker_uncorrected",
                     "nihtbx_pattern_uncorrected", "ACE_index_sum_score")

#1.2 Define labels for each continuous variable
continuous_variable_labels <- c(
  cbcl_scr_dsm5_depress_t = "DSM-5 Depression (CBCL)",
  cbcl_scr_dsm5_anxdisord_t = "DSM-5 Anxiety (CBCL)",
  cbcl_scr_syn_attention_t = "Attention Problems (CBCL)",
  cbcl_scr_syn_aggressive_t = "Aggression Problems (CBCL)",
  pgbi_p_ss_score = "Mania Sum Score (GBI)",
  upps_y_ss_negative_urgency = "Negative Urgency (UPPS-P)",
  upps_y_ss_positive_urgency = "Positive Urgency (UPPS-P)",
  sds_p_ss_total = "Sleep Disturbance (SDS)",
  reshist_addr1_coi_z_coi_nat = "COI Z-Score",
  nsc_p_ss_mean_3_items = "Neighborhood Safety (Mean)",
  nihtbx_list_uncorrected = "List Sorting (NIHTB)",
  nihtbx_flanker_uncorrected = "Flanker Task (NIHTB)",
  nihtbx_pattern_uncorrected = "Pattern Comparison (NIHTB)",
  ACE_index_sum_score = "ACE Index Sum"
)

#1.3 Wrangle data for plotting
continuous_plot_data <- clustered_risk_variable_data %>%
  dplyr::select(cluster, cbcl_scr_dsm5_depress_t, 
                cbcl_scr_dsm5_anxdisord_t, cbcl_scr_syn_attention_t, 
                cbcl_scr_syn_aggressive_t, pgbi_p_ss_score, 
                upps_y_ss_negative_urgency, upps_y_ss_positive_urgency, 
                sds_p_ss_total, reshist_addr1_coi_z_coi_nat, 
                nsc_p_ss_mean_3_items, nihtbx_list_uncorrected, 
                nihtbx_flanker_uncorrected, nihtbx_pattern_uncorrected, 
                ACE_index_sum_score) %>%
  pivot_longer(-cluster, names_to = "variable", values_to = "value")

#1.4 Create the violin boxplot
continuous_plot <- continuous_plot_data %>%
  ggplot(aes(x = as.factor(cluster), y = value, fill = as.factor(cluster))) +
  geom_violin(trim = TRUE, alpha = 0.7) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9), outlier.size = 0.5) +
  facet_wrap(
    ~variable,
    scales = "free_y",
    labeller = as_labeller(continuous_variable_labels)
  ) +
  labs(x = "Cluster", y = "Value", fill = "Cluster",
       title = "Cluster Profiles on Continuous Risk Variables") +
  theme_minimal() +
  theme(strip.text = element_text(size = 10),
        axis.text = element_text(size = 8),
        legend.position = "top") +
  scale_fill_brewer(palette = "Set3")

#1.5 Print the continuous variable plot
print(continuous_plot)

#2. Create a summary statistics table for the continuous variables
#2.1 Wrangle data for summary statistics
continuous_summary_data <- continuous_plot_data %>%
  group_by(cluster, variable) %>%
  summarize(
    mean = round(mean(value, na.rm = TRUE), 2),
    sd = round(sd(value, na.rm = TRUE), 2),
    median = round(median(value, na.rm = TRUE), 2),
    range = paste0(
      round(min(value, na.rm = TRUE), 2), " - ",
      round(max(value, na.rm = TRUE), 2)
    ),
    .groups = "drop"
  ) %>%
  mutate(variable = as_labeller(continuous_variable_labels)(variable))

#2.2 Create the summary statistics table
continuous_summary_data %>%
  kbl(col.names = c("Cluster", "Variable", "Mean", "SD", "Median", "Range"),
      caption = "Summary Statistics for Continuous Risk Variables by Cluster") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))

```

### **Refined Cluster Characterization and Labeling Based on Risk Variables**  

Each cluster is characterized based on categorical risk factors (e.g., family history of depression, bullying) and continuous risk factors (e.g., DSM-5 anxiety and depression scores, cognitive functioning, and impulsivity). Below is an in-depth profiling of each cluster, refining your pre-existing hypotheses with more detailed insights.

#### **Cluster 1: Moderate Risk with High Family Depression History and Cognitive Strengths**

- **Size:** 1,984 individuals  

- **Categorical Characteristics:**  

  - **Very high family history of depression**  
  
  - **Moderate bullying exposure**  
  
  - **Moderate parental history of mania**  
  
- **Continuous Characteristics:**  

  - **Moderate DSM-5 Anxiety and Depression**  
  
  - **Elevated attention problems and aggression**  
  
  - **High cognitive function** 
  
  - **Moderate childhood adversity**  
  
  - **Moderate sleep disturbances**  
  
  - **Moderate impulsivity** 
  
  - **Decent neighborhood safety**  

#### **Cluster 2: Moderate Risk with High Impulsivity, Low Cognitive Scores, and Socioeconomic Disadvantage**

- **Size:** 2,119 individuals  

- **Categorical Characteristics:**  

  - **Moderate bullying exposure**  
  
  - **Low parental history of mania**  
  
  - **Moderate family history of depression**  
  
- **Continuous Characteristics:** 

  - **Moderate-low DSM-5 Anxiety and Depression**  
  
  - **Low cognitive scores**  
  
  - **Low childhood adversity**  
  
  - **High impulsivity** 
  
  - **Low-moderate sleep disturbances**  
  
  - **Low neighborhood safety**  

#### **Cluster 3: Healthy Low-Risk Group**

- **Size:** 3,498 individuals (largest cluster) 

- **Categorical Characteristics:**

  - **Lowest rates of bullying**  
  
  - **Lowest rates of family history of depression**  
  
  - **Lowest rates of parental mania**  
  
- **Continuous Characteristics:**  

  - **Lowest DSM-5 Anxiety and Depression **  
  
  - **Lowest aggression and attention problems**  
  
  - **Highest cognitive function**  
  
  - **Best neighborhood safety**  
  
  - **Lowest childhood adversity**  
  
  - **Lowest impulsivity**  
  
  - **Lowest sleep disturbance**  

#### **Cluster 4: Severe Emotional & Behavioral Dysregulation with High Adversity**

- **Size:** 427 individuals (smallest cluster)  

- **Categorical Characteristics:** 

  - **Highest rates of bullying** 
  
  - **High family history of depression**  
  
  - **High parental history of mania**  
  
- **Continuous Characteristics:**  

  - **Very high DSM-5 Anxiety and Depression **  
  
  - **Severe attention problems and aggression**  
  
  - **Lower cognitive function** 
  
  - **Highest childhood adversity**
  
  - **Highest impulsivity** 
  
  - **Highest sleep disturbances**
  
  - **Low neighborhood safety**  

#### **Cluster 5: Moderate-High Risk with Attention/Aggression Problems and Cognitive Strengths**

- **Size:** 1,125 individuals  

- **Categorical Characteristics:**  

  - **High bullying exposure**  
  
  - **Moderate family history of depression**  
  
  - **Moderate parental history of mania**  
  
- **Continuous Characteristics:**  

  - **Elevated DSM-5 Anxiety and Depression**  
  
  - **High attention problems and aggression**  
  
  - **Moderate cognitive scores** 
  
  - **Moderate childhood adversity**
  
  - **Moderate impulsivity** 
  
  - **Moderate-high sleep disturbances** 
  
  - **Moderate neighborhood safety**  
  