---
title: "1. BD Discrete-Time Survival (First Onset)"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
params:
  repo: !r Sys.getenv("REPO", unset = here::here())
  data_dir: "data/data_processed/analysis_datasets/"
  out_dir: "results/main_analysis/1_bd_survival"
  bd_pp_rds: "bd_person_period_k2_z_score.rds"
  bd_pp_csv: "bd_person_period_k2_z_score.csv"
  k_value: 2
  run_influence: false
  seed: 123
  link_primary: "logit"
  ages_pred: !r c()
  wave_ref: "ses-04A"
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, message = TRUE, warning = TRUE, results = "markup", verbose = TRUE, comment = "")

```

## Data Wrangling

First, we will read in necessary packages and use parameters to read in our pre-built analysis dataset; then filter it for complete cases (according to the necessary variables to be used in analyses) and print a brief table outlining the unique number of subjects and total observations we possess in our "complete" analysis ready dataset

```{r environment, echo = FALSE, warning = FALSE}

## Load packages, data, & set env ##

# Quiet & safe package loads
suppressPackageStartupMessages({
  library(dplyr); 
  library(tidyr); 
  library(readr);
  library(rlang);
  library(tibble);
  library(lubridate); 
  library(glue); 
  library(forcats)
  library(broom); 
  library(broom.mixed); 
  library(lme4); 
  library(blme);
  library(performance)
  library(emmeans); 
  library(ggeffects); 
  library(pROC); 
  library(yardstick)
  library(clubSandwich);
  library(sandwich);
  library(lmtest);
  library(patchwork); 
  library(here); 
  library(stringr); 
  library(ggplot2);
  library(kableExtra);
  library(DHARMa);
  library(see);
  library(glmmTMB);
  library(logistf);
  library(purrr);
  library(scales);
  library(brglm2)
})

# Set seed for reproducibility
set.seed(params$seed)

# Set numerical handling
options(scipen = 999, digits = 8)

# Ensure proper encoding of output
options(encoding = "UTF-8")
knitr::opts_knit$set(encoding = "UTF-8")
knitr::knit_hooks$set(output = function(x, options) enc2utf8(x))
knitr::knit_hooks$set(inline = function(x) enc2utf8(paste(x, collapse = "")))

# Ensure proper storage of emmeans objects
emmeans::emm_options(infer = c(TRUE, TRUE))

# Establish paths to repo and data (I/O)
REPO <- params$repo
DATA_DIR <- file.path(REPO, params$data_dir)
OUT_DIR <- file.path(REPO, params$out_dir)
dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)

# Create a helper function to read in BD person-period data - prefer RDS; fallback to csv
read_pp <- function(rds_path, csv_path) {
  if (file.exists(rds_path)) readRDS(rds_path) else readr::read_csv(csv_path, show_col_types = FALSE)
}

# Load person-period BD data
bd_pp <- read_pp(file.path(DATA_DIR, params$bd_pp_rds),
  file.path(DATA_DIR, params$bd_pp_csv)) %>%
  mutate(
    participant_id = as.character(participant_id),
    family_id = factor(as.character(family_id)),
    site_factor = factor(site_factor),
    cluster = factor(cluster),
    start_wave = factor(start_wave),
    end_wave = factor(end_wave),
    outcome = factor(outcome),
    sex = factor(sex)
  )

# Keep just the BD outcomes we model in this analysis
bd_outcomes <- c("bipolar_I","bipolar_II","bd_nos","any_bsd")
bd_pp <- bd_pp %>% filter(outcome %in% bd_outcomes)

# Refine pre-built data using a model-ready filter that mirrors QC "post" rules
bd_pp_model <- bd_pp %>%
  filter(event %in% c(0L,1L)) %>%
  filter(!is.na(cluster)) %>%
  filter(!is.na(age_mid), !is.na(sex), !is.na(site_factor), !is.na(family_id))

# Conduct a sanity check of the dataset size (unique ID's and number of observations)
sizes_tbl <- tibble(
  Dataset = "BD person-period (model-ready)",
  Rows = nrow(bd_pp_model),
  Unique_IDs  = dplyr::n_distinct(bd_pp_model$participant_id)
)

# Print the dataset size sanity check table
knitr::kable(sizes_tbl, caption = "Dataset size check")  %>%
  kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))

# Ensure cluster, wave, sex, and site reference levels (consistent displays)
cluster_levels <- levels(bd_pp$cluster)
bd_pp_model <- bd_pp_model %>%
  mutate(
    cluster = forcats::fct_relevel(cluster, "C1"),
    end_wave = forcats::fct_relevel(end_wave, "ses-02A", "ses-04A", "ses-06A"),
    sex = {
      lv <- levels(sex)
      ref <- if ("Female" %in% lv) "Female" else lv[1]
      forcats::fct_relevel(sex, ref)
    },
    site_factor = {
  lv <- levels(site_factor)
  ref <- if ("site01" %in% lv) "site01" else lv[1]
  forcats::fct_relevel(site_factor, ref)
  })

# Establish interval length in years (02A~2, 04A~4, 06A~6); handles missed waves
wave_year <- function(w) {
  as.numeric(stringr::str_extract(as.character(w), "\\d+"))
}
bd_pp_model <- bd_pp_model %>%
  mutate(dt_years = pmax(wave_year(end_wave) - wave_year(start_wave), 1),
    log_dt = log(dt_years))

# Guard interval-lengths (offset) before fitting cloglog
bad_dt <- sum(!is.finite(bd_pp_model$dt_years) | !is.finite(bd_pp_model$log_dt))
if (bad_dt > 0) {
  warning("Dropping ", bad_dt, " rows with undefined interval length.")
  bd_pp_model <- bd_pp_model %>% dplyr::filter(is.finite(dt_years), is.finite(log_dt))
}

# Get descriptive event counts (post-filter; by outcome x wave x cluster)
onset_counts <- bd_pp_model %>%
  group_by(outcome, end_wave, cluster) %>%
  summarise(events = sum(event == 1L, na.rm = TRUE),
    intervals = n(), .groups = "drop") %>%
  arrange(outcome, end_wave, cluster)

# Save post-filter counts to disk and show in HTML
readr::write_csv(onset_counts, file.path(OUT_DIR, "bd_dth_onset_counts.csv"))
knitr::kable(onset_counts, caption = "Event counts by outcome x wave x cluster") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))

## Do a post-filter missingness summary of the model covariates ##

# Create a list of relevant covariates
covars <- c("cluster","age_mid_between","age_mid_cwc","baseline_status",
            "sex","site_factor","end_wave","participant_id","family_id")

# Create a missingness table for the post-filter covariate data
miss_tbl <- bd_pp_model %>%
  mutate(across(all_of(covars), ~ as.numeric(is.na(.)))) %>%
  group_by(outcome) %>%
  summarise(n = n(),
    across(all_of(covars), ~ sum(.), .names = "n_miss_{col}"),
    across(starts_with("n_miss_"), ~ round(./n, 4), .names = "p_{.col}"),
    .groups = "drop")

# Write the missingness table for the post-filter covariate data as a table in the HTML
knitr::kable(miss_tbl, caption = "Missingness (post-filter) by outcome") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))

## Baseline_status QC specific to each outcome ##

# Create a qc table summarizing the baseline status presence for each diagnostic outcome
baseline_qc <- bd_pp_model %>%
  group_by(outcome, baseline_status) %>%
  summarise(n = n(), events = sum(event == 1L), .groups = "drop") %>%
  arrange(outcome, baseline_status)

# Write the qc table summarizing the baseline status presence for each diagnostic outcome in the HTML
knitr::kable(baseline_qc, caption = "Baseline_status QC by outcome") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))

## Baseline prevalence at baseline (participants) by outcome x risk cluster ##

# IMPORTANT: do this BEFORE filtering out baseline-positive so the counts are complete - collapse to one row per participant x outcome x cluster, then count baseline-positive IDs
baseline_prev_ids <- bd_pp_model %>%
  dplyr::group_by(outcome, cluster, participant_id) %>%
  dplyr::summarise(
    baseline_pos = as.integer(any(baseline_status == 1, na.rm = TRUE)),
    .groups = "drop"
  )

# Crate a table outlining the distribution of baseline BD dx endorsements
baseline_prev_tbl <- baseline_prev_ids %>%
  dplyr::group_by(outcome, cluster) %>%
  dplyr::summarise(
    n_participants = dplyr::n(),
    n_baseline_positive = sum(baseline_pos == 1L, na.rm = TRUE),
    pct_baseline_positive = n_baseline_positive / n_participants,
    .groups = "drop") %>%
  dplyr::arrange(outcome, cluster)

# Display the distribution of baseline BD dx endorsements in the HTML
knitr::kable(
  baseline_prev_tbl %>%
    dplyr::mutate(pct_baseline_positive = scales::percent(pct_baseline_positive, accuracy = 0.01)),
  caption = "Baseline prevalence at baseline (participants) by outcome - risk cluster") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))

## Primary risk set for FIRST ONSET (critical) ##

# Build an ID-level baseline flag using the any_bsd outcome (covers BD-I/II/NOS)
baseline_any_bsd_ids <- bd_pp_model %>% 
  dplyr::filter(outcome == "any_bsd") %>%
  dplyr::group_by(participant_id) %>%
  dplyr::summarise(
    all_na = all(is.na(baseline_status)),
    baseline_any_bsd = dplyr::if_else(
      all_na, NA_integer_, as.integer(any(baseline_status == 1L, na.rm = TRUE))),
    .groups = "drop")

# Join to all outcomes, do not assume 0 for missing data
bd_pp_model <- bd_pp_model %>%
  dplyr::left_join(baseline_any_bsd_ids %>% dplyr::select(participant_id, baseline_any_bsd),
    by = "participant_id")

# Count before/after and filter to baseline-negative-for-any-BSD
n_before <- nrow(bd_pp_model)
ids_before <- dplyr::n_distinct(bd_pp_model$participant_id)

# Filter the dataset to contain participants that had not yet had BD
bd_pp_model <- bd_pp_model %>% dplyr::filter(baseline_any_bsd == 0L)

# Print a message containing the new number of observations without the participants who already endorsed BD at baseline
message(
  "Risk-set conditioning: excluded participants baseline-positive (and baseline-unknown) for ANY BSD; ",
  "rows dropped = ", n_before - nrow(bd_pp_model),
  " (IDs from ", ids_before, " to ", dplyr::n_distinct(bd_pp_model$participant_id), "). ",
  "Models omit a baseline term because only baseline-negative youth remain."
)

# Censor after first onset (per outcome x participant); keep all intervals up to and including the first event==1 row; drop any later rows
if ("interval_index" %in% names(bd_pp_model)) {
  bd_pp_model <- bd_pp_model %>%
    dplyr::arrange(outcome, participant_id, interval_index)
} else {
  bd_pp_model <- bd_pp_model %>%
    dplyr::arrange(outcome, participant_id, as.integer(end_wave))
}

bd_pp_model <- bd_pp_model %>%
  dplyr::group_by(outcome, participant_id) %>%
  dplyr::mutate(after_onset = dplyr::lag(cummax(event == 1L), default = FALSE)) %>%
  dplyr::filter(!after_onset) %>%
  dplyr::ungroup() %>%
  dplyr::select(-after_onset)

# Check to make sure that censoring was performed correctly
if ("interval_index" %in% names(bd_pp_model)) {
  bd_pp_model <- bd_pp_model %>% dplyr::arrange(outcome, participant_id, interval_index)
} else {
  bd_pp_model <- bd_pp_model %>% dplyr::arrange(outcome, participant_id, as.integer(end_wave))
}

aud <- bd_pp_model %>%
  group_by(outcome, participant_id) %>%
  mutate(had_onset = cummax(event == 1L)) %>%
  summarise(post_onset_rows = max(0L, sum(had_onset) - 1L), .groups = "drop")
message("Any post-onset rows remain? ", any(aud$post_onset_rows > 0))

# Establish empirical wave-median ages for plotting/prediction
wave_age_medians <- bd_pp_model %>%
  dplyr::group_by(end_wave) %>%
  dplyr::summarise(age_med = median(age_mid, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(match(end_wave, levels(bd_pp_model$end_wave)))
ages_empirical <- wave_age_medians$age_med %>% as.numeric() %>% round(1)

# Save the wave-median ages for later reporting 
readr::write_csv(wave_age_medians, file.path(OUT_DIR, "wave_age_medians.csv"))

# Choose ages to use (robust fallback to guarantee non-empty)
ages_pred_vec <- if (!is.null(params$ages_pred) && length(params$ages_pred) > 0) {
  params$ages_pred
} else if (length(ages_empirical) > 0 && all(is.finite(ages_empirical))) {
  ages_empirical
} else {
  message("ages_empirical empty; falling back to previously determined medians.")
  c(11.004729021633345, 13.116629238715475, 15.2109589041096)
}

```

## Analysis Overview

**Our Goal**

Estimate discrete-time hazards with a logit link for first-onset bipolar outcomes-BD-I, BD-II, BD-NOS, and Any BSD-from 02A onward, using the model-ready person-period datasets built previously

**Why discrete-time hazard (DTH)?**

-   DTH models the probability of an onset within an interval, conditional on no prior onset. With assessments at distinct waves in ABCD, it is good practice here for us to:

    -   Encode each interval as a row (person-period format)
    -   Flexibly estimate the baseline hazard via wave fixed effects (i.e., dummies for 02A, 04A, 06A)
    -   Include static and time-varying predictors alongside the hazard
    -   This matches standard practice for event occurrence with panel waves (e.g., Singer & Willett)

-   To note, intervals with missed waves contribute longer person-periods; our cloglog + offset(log interval) sensitivity shows results are robust to irregular interval length

**Why pooled logistic DTH (GLM) instead of a GLMM as primary?**

-   ABCD follow-ups occur at distinct waves (02A, 04A, 06A), so event times are interval-censored by design; discrete-time hazard models on person-period data are therefore the natural framework (Singer & Willett)

-   In our person-period data, clustering is extremely sparse (1–3 rows/person) with single-event censoring, and the participant random intercept is weakly identified. In mixed-effects fits, Laplace solutions with nAGQ ≥ 1 were unstable / optimizer-dependent and frequently produced degenerate Hessians when wave indicators and within-person age were both included

-   We therefore use pooled logistic regression for primary inference and account for within-person dependence using participant-clustered robust SEs (CR2 primary; HC1 secondary). This yields stable, design-consistent inference without requiring estimation of a poorly-supported random effect structure

-   Sensitivity analyses include: (i) complementary log-log with offset(log interval length), and (ii) mixed-effects versions only when estimable/stable (reported as diagnostics/sensitivity rather than primary inference)

**How we handle age (critical for interpretation)**

-   Our build steps created:

    -   `age_mid`: midpoint age for each interval
    -   `age_mid_cwc`: within-person centered age = `age_mid` person_mean(age_mid)\` (captures aging within individuals)
    -   `age_mid_between`: between-person age = `person_mean(age_mid)` grand_mean(age_mid)\` (captures older vs younger people on average)

-   Best practice in longitudinal models is to include both pieces (within & between) to avoid conflating them. This is the within-between (Mundlak/hybrid) decomposition

    -   Interpretation:

        -   `age_mid_cwc` -\> change in hazard as the same person ages across waves (holding their average age constant)
        -   `age_mid_between` -\> difference in hazard for people who are on average older vs younger in this sample

    -   We originally interacted cluster x age_mid_between to ask: do cluster differences widen or narrow at older average ages? We evaluated cluster x age_mid_between; it was unstable/non-significant and was dropped for parsimony and better diagnostics
    
> In practice here, age_mid_cwc is highly collinear with the wave baseline-hazard indicators given only three intervals (VIF ~15), so our primary model includes age_mid_between plus wave indicators; we include orthogonalized within-person age (age_mid_cwc_orth) only as a sensitivity analysis too though (see below section)

**Model we fit per outcome**

-   Primary marginal DTH model (pooled logistic):
    -   `event ~ cluster + sex + end_wave + age_mid_between`
    -   Inference: participant-clustered robust SEs (CR2 primary via `clubSandwich`; HC1 secondary via `sandwich`)
-   Sensitivity models:
    -   Linear time: replace `end_wave` with `wave_num` (1 df)
    -   Within-person age robustness: include `age_mid_cwc_orth` (within-person midpoint age residualized on wave indicators to avoid collinearity/identifiability issues)
    -   cloglog + offset(log_dt): event \~ ... with binomial(link="cloglog") + offset(log_dt)
    -   Mixed-effects diagnostics only: glmer(... + (1\|participant_id)) at nAGQ=0 to compute ICC; nAGQ=1 and allFit() used to document optimizer dependence / instability (not used for primary inference)

## 1. Create Helper Functions to Use for Conducting Analyses

-   `build_formula_primary()`: Discrete-time hazard GLM with wave fixed effects (baseline hazard), within/between age decomposition, and cluster robust effects for participant ID
-   `fit_glmer_safe()`: Robust GLMM fitter with `bobyqa` + fallback; supports logit (primary) and cloglog (sensitivity)
-   `tidy_wald()`: Fixed-effect ORs with Wald CIs
-   `tidy_cr2()`: Cluster-robust (CR2) SEs and CIs, clustered by `family_id` (recommended small/medium cluster correction)
-   `pred_grid_by_age()`: Marginal predicted first-onset hazards by cluster across chosen ages (empirical medians), fixing within-age = 0 and wave = `params$wave_ref`
-   `calibration_df()`: Decile-bin calibration (predicted vs observed)
-   `auc_interval()`: Interval-level ROC AUC (QC indicator; marginal)
-   `diag_report()`: Singularity and overdispersion checks
-   `cluster_contrasts_by_age()`: `emmeans` pairwise cluster ORs at the chosen ages

Multiple others for QC and plotting

```{r helper function, echo = FALSE, warning = FALSE}

## 1. Create helper functions for building, fitting, and summarizing models ##

#1.1 Build DTH formula, where wave FE (end_wave) = flexible baseline hazard; and formula includes within- and between-person age components
#1.1 Build DTH formulas (primary GLM + sensitivities)
build_formula_primary <- function() {
  stats::as.formula("event ~ cluster + sex + end_wave + age_mid_between")
}

build_formula_lin_time <- function() {
  stats::as.formula("event ~ cluster + sex + wave_num + age_mid_between")
}

build_formula_within_orth <- function() {
  stats::as.formula("event ~ cluster + sex + end_wave + age_mid_between + age_mid_cwc_orth")
}

#1.1.1 Decide if we should simplify (drop cluster:age_mid_between) based on diagnostics
needs_simplify <- function(fit, data) {
  di <- try(diag_report(fit), silent = TRUE)
  if (inherits(di, "try-error")) return(FALSE)
  nevents <- sum(data$event == 1L, na.rm = TRUE)
  p_fixed <- try(length(lme4::fixef(fit)), silent = TRUE); if (inherits(p_fixed, "try-error")) p_fixed <- NA_real_
  flag_singular <- isTRUE(di$is_singular)
  flag_vif <- !is.na(di$max_VIF) && di$max_VIF > 10
  flag_events <- is.finite(nevents) && is.finite(p_fixed) && nevents < (p_fixed + 3)
  flag_singular || flag_vif || flag_events
}

#1.2 Create a safe glmer with robust optimizer fallback
fit_glmer_safe <- function(formula, data, link = c("logit","cloglog")) {
  link <- match.arg(link)
  family_fn <- if (link == "logit") binomial(link = "logit") else binomial(link = "cloglog")
  ctrl <- glmerControl(optimizer = "bobyqa",
    optCtrl = list(maxfun = 2e5),
    check.conv.singular = "ignore")
  fit <- tryCatch(
    glmer(formula, data = data, family = family_fn, control = ctrl, nAGQ = 0),
    error = function(e) {
      message("Retrying with nAGQ=0 (", link, ") due to: ", conditionMessage(e))
      glmer(formula, data = data, family = family_fn, control = ctrl, nAGQ = 0)
    }
  )
  fit
}

#1.2.1 Safe GLM fitter for primary inference
fit_glm_safe <- function(formula, data, link = c("logit","cloglog")) {
  link <- match.arg(link)
  fam <- binomial(link = link)
  glm(formula, data = data, family = fam)
}

#1.3 Create a function to extract Wald OR + CI in tidy format
tidy_wald <- function(fit) {
  broom.mixed::tidy(fit, effects = "fixed", conf.int = TRUE, conf.method = "Wald") %>%
    mutate(term = as.character(term),
      OR = exp(estimate),
      CI_lo= exp(conf.low),
      CI_hi= exp(conf.high),
      SE = std.error) %>%
    dplyr::select(term, estimate, SE, OR, CI_lo, CI_hi, p.value)
}

#1.4 Create a function to also extract CR2 robust in tidy format
tidy_cr2 <- function(fit, cluster_var) {
  test_method <- "Satterthwaite"

  ct <- try(
    clubSandwich::coef_test(fit, vcov = "CR2", cluster = cluster_var, test = test_method),
    silent = TRUE)

  if (inherits(ct, "try-error")) {
    out <- broom.mixed::tidy(fit, effects = "fixed", conf.int = TRUE, conf.method = "Wald") %>%
      dplyr::mutate(
        term = as.character(term),
        OR = exp(estimate),
        CI_lo = exp(conf.low),
        CI_hi = exp(conf.high),
        SE = std.error,
        df = NA_real_,
        p.value = NA_real_,
        note = "CR2_failed") %>%
      dplyr::select(term, estimate, SE, df, OR, CI_lo, CI_hi, p.value, note)
    return(out)
  }

  ctt <- tibble::as_tibble(ct)
  pcol <- intersect(c("p_Satt","p_z","p_Nd","p_t"), names(ctt))[1]
  dfcol <- intersect(c("df_Satt","df"), names(ctt))[1]
  if (is.na(pcol)) ctt$p_any <- NA_real_ else ctt$p_any <- ctt[[pcol]]
  if (is.na(dfcol)) ctt$df <- NA_real_ else ctt$df <- ctt[[dfcol]]

  ctt %>%
    dplyr::mutate(
      term  = as.character(term),
      OR = exp(beta),
      CI_lo = exp(CI_L),
      CI_hi = exp(CI_U)) %>%
    dplyr::select(term, estimate = beta, SE, df, OR, CI_lo, CI_hi, p.value = p_any)
}

#1.4.1 Tidy CR2 inference for GLM objects (participant-clustered)
tidy_glm_cr2 <- function(fit, cluster_var) {
  mf <- model.frame(fit)
  idx <- as.integer(rownames(mf))
  cluster_aligned <- cluster_var[idx]
  V  <- clubSandwich::vcovCR(fit, cluster = cluster_aligned, type = "CR2")
  ct <- clubSandwich::coef_test(fit, vcov = V, test = "Satterthwaite")

  #1.4.1.1 Make it a vanilla data.frame so we can reliably access rownames/cols across versions
  ctd <- as.data.frame(ct)

  #1.4.1.2 Ensure we always have a `term` column
  if (!("term" %in% names(ctd))) {
    if ("Coef" %in% names(ctd)) {
      ctd$term <- as.character(ctd$Coef)
    } else if (!is.null(rownames(ctd)) && all(nzchar(rownames(ctd)))) {
      ctd$term <- rownames(ctd)
    } else {
      ctd$term <- paste0("b", seq_len(nrow(ctd)))
    }
  }

  ct2 <- tibble::as_tibble(ctd)

  #1.4.1.3 Pick p-value / df columns robustly
  pcol  <- intersect(c("p_Satt","p_z","p_Nd","p_t","p_val","p"), names(ct2))[1]
  dfcol <- intersect(c("df_Satt","df","DF"), names(ct2))[1]

  #1.4.1.4 Pick beta/SE/CI columns robustly
  betacol <- intersect(c("beta","Beta","Estimate"), names(ct2))[1]
  secol <- intersect(c("SE","Std. Error","Std.Error","SE_beta"), names(ct2))[1]
  cilocol <- intersect(c("CI_L","CI_L_Satt","lower","Lower","2.5 %"), names(ct2))[1]
  cihicol <- intersect(c("CI_U","CI_U_Satt","upper","Upper","97.5 %"), names(ct2))[1]

  ct2 %>%
    dplyr::transmute(
      term = term,
      estimate = if (!is.na(betacol)) .data[[betacol]] else NA_real_,
      SE = if (!is.na(secol)) .data[[secol]] else NA_real_,
      df = if (!is.na(dfcol)) .data[[dfcol]] else NA_real_,
      OR = exp(estimate),
      CI_lo = if (!is.na(cilocol)) exp(.data[[cilocol]]) else NA_real_,
      CI_hi = if (!is.na(cihicol)) exp(.data[[cihicol]]) else NA_real_,
      p.value = if (!is.na(pcol)) .data[[pcol]] else NA_real_)
}

#1.4.2 Tidy CR2 / HC1 inference for GLM objects (participant-clustered)
tidy_glm_hc1 <- function(fit, cluster_var) {

  #1.4.2.1 Align cluster_var to the rows actually used by the model (robust to non-numeric rownames)
  mf <- model.frame(fit)
  rn <- rownames(mf)
  idx <- suppressWarnings(as.integer(rn))
  if (anyNA(idx)) idx <- seq_len(nrow(mf))
  cluster_aligned <- cluster_var[idx]

  #1.4.2.2 Cluster-robust HC1 vcov (sandwich)
  V <- sandwich::vcovCL(fit, cluster = cluster_aligned, type = "HC1")

  #1.4.2.3 Coefficients
  b <- stats::coef(fit)
  terms <- names(b)

  #1.4.2.4 Ensure vcov conforms to coef ordering
  if (is.null(dim(V)) || nrow(V) != length(b) || ncol(V) != length(b)) {
    stop("vcovCL() returned a vcov with dims ", paste(dim(V), collapse = "x"),
         " but coef(fit) has length ", length(b), ". Check cluster alignment and model matrix.")
  }
  if (!is.null(colnames(V)) && all(terms %in% colnames(V))) {
    V <- V[terms, terms, drop = FALSE]
  }

  se <- sqrt(diag(V))

  #1.4.2.5 If any SEs are non-finite, fail loudly
  if (any(!is.finite(se))) {
    bad <- terms[!is.finite(se)]
    stop("Non-finite robust SE(s) for: ", paste(bad, collapse = ", "),
         ". This usually indicates a vcov problem (e.g., separation / rank deficiency).")
  }

  #1.4.2.6 Wald z tests
  z <- b / se
  p <- 2 * stats::pnorm(-abs(z))

  tibble::tibble(
    term = terms,
    estimate = unname(b),
    SE = unname(se),
    df = NA_real_,
    OR = exp(estimate),
    CI_lo = exp(estimate - 1.96 * SE),
    CI_hi = exp(estimate + 1.96 * SE),
    p.value = unname(p))
}

#1.5 Create a function to calculate marginal predicted hazard by cluster across ages, with an option to align ages to waves
pred_grid_by_age <- function(fit, data, ages = ages_pred_vec, wave_ref = params$wave_ref, align_waves = FALSE) {
  grand_mean <- mean(data$age_mid, na.rm = TRUE)
  age_between_vals <- ages - grand_mean
  
  #1.5.1 Map empirical ages to waves if requested (assumes ages are wave medians in end_wave order)
  waves <- levels(data$end_wave)
  if (align_waves) {
    
    #1.5.2 use min(lengths) to be robust if ages length differs from number of waves
    k <- min(length(ages), length(waves))
    end_waves_vec <- waves[seq_len(k)]
    ages <- ages[seq_len(k)]
    age_between_vals <- ages - grand_mean
    newdat <- do.call(rbind, lapply(seq_len(k), function(i) {
      expand.grid(cluster = levels(data$cluster),
        age_mid_between = age_between_vals[i],
        age_mid_cwc = 0,
        sex = levels(data$sex)[1],
        end_wave = factor(end_waves_vec[i], levels = levels(data$end_wave)))
    }))
    newdat$age_years <- rep(ages, each = length(levels(data$cluster)))
  } else {
    newdat <- expand.grid(cluster = levels(data$cluster),
      age_mid_between = age_between_vals,
      age_mid_cwc = 0,
      sex = levels(data$sex)[1],
      end_wave = factor(wave_ref, levels = levels(data$end_wave)))
    newdat$age_years <- rep(ages, each = length(levels(data$cluster)))
  }
  
  #1.5.3 Predict and store marginal hazards
  newdat$pred <- predict(fit, newdata = newdat, type = "response")
  newdat
}

#1.6 Create a function to conduct calibration (i.e., deciles of predicted hazard vs observed)
calibration_df <- function(fit, data) {
  pr <- predict(fit, type = "response")
  df <- data %>% mutate(pred = pr, dec = dplyr::ntile(pred, 10))
  df %>%
    group_by(dec) %>%
    summarise(
      pred_mean = mean(pred, na.rm = TRUE),
      obs_mean = mean(event == 1L, na.rm = TRUE),
      n = n(),
      .groups = "drop")
}

#1.7 Create a function to calculate the AUC (interval-level; QC only, ignores clustering)
auc_interval <- function(fit, data) {
  pr <- predict(fit, type = "response")
  pROC::roc(response = data$event, predictor = pr, quiet = TRUE)$auc %>% as.numeric()
}

#1.8 Diagnostics: singularity, dispersion, groups, convergence, max VIF
diag_report <- function(fit) {
  is_sing <- tryCatch(lme4::isSingular(fit, tol = 1e-4), error = function(e) NA)
  overK <- tryCatch(performance::check_overdispersion(fit)$dispersion_ratio, error = function(e) NA_real_)
  conv <- tryCatch(performance::check_convergence(fit)$converged, error = function(e) NA)
  vif_max <- tryCatch(max(performance::check_collinearity(fit)$VIF, na.rm = TRUE), error = function(e) NA_real_)
  fl <- tryCatch(lme4::getME(fit, "flist"), error = function(e) NULL)
  n_id <- if (!is.null(fl) && "participant_id" %in% names(fl)) nlevels(fl$participant_id) else NA_integer_
  n_fam <- if (!is.null(fl) && "family_id" %in% names(fl)) nlevels(fl$family_id) else NA_integer_
  n_site <- if (!is.null(fl) && "site_factor" %in% names(fl)) nlevels(fl$site_factor) else NA_integer_
  tibble(is_singular = is_sing,
         converged = conv,
         overdispersion_K = overK,
         max_VIF = vif_max,
         n_groups_participant = n_id,
         n_groups_family = n_fam,
         n_groups_site = n_site)
}

#1.9.1 Build a function to conduct cluster contrasts at representative ages (emmeans on linear predictor). Builds EMMs at: within=0, wave=ref, and between set to represent ages
cluster_contrasts_by_age <- function(fit, data, ages = ages_pred_vec, wave_ref = params$wave_ref) {
  grand_mean <- mean(data$age_mid, na.rm = TRUE)
  age_between_vals <- ages - grand_mean

  #1.9.1.1 Start with both age-within variants and both wave codings; we'll keep only what the model uses.
  ref_list <- list(
    age_mid_between = age_between_vals,
    age_mid_cwc = 0,
    baseline_status = 0,
    end_wave = factor(wave_ref, levels = levels(data$end_wave)),
    wave_num = match(wave_ref, levels(data$end_wave)) - 1L,
    sex = levels(data$sex)[1],
    site_factor = levels(data$site_factor)[1])

  #1.9.1.2 If the fitted model included an offset, supply it (e.g., cloglog sensitivity in other calls)
  mf_names <- try(names(model.frame(fit)), silent = TRUE)
    if (!inherits(mf_names, "try-error") && "log_dt" %in% mf_names) {
      
      #1.9.1.2.1 Use empirical median log_dt for the reference wave used in contrasts
      ref_list$log_dt <- get_log_dt_for_wave(data, factor(wave_ref, levels = levels(data$end_wave)))
    }

  #1.9.1.3 Keep only variables the model actually uses; also drop the unused within-age variant.
  if (!inherits(mf_names, "try-error")) {
    if (!"end_wave" %in% mf_names) ref_list$end_wave <- NULL
    if (!"wave_num" %in% mf_names) ref_list$wave_num <- NULL
    ref_list <- ref_list[names(ref_list) %in% mf_names]
  }

  #1.9.1.4 EMMs on the linear predictor; contrasts -> ORs via exp()
  em <- emmeans::emmeans(fit, ~ cluster | age_mid_between, at = ref_list)
  ct <- emmeans::contrast(em, method = "revpairwise", adjust = "none")
  sm <- summary(ct, infer = c(TRUE, TRUE)) |> tibble::as_tibble()
  ci <- emm_ci_cols(sm)
  sm |>
    dplyr::mutate(
      age_years = round(age_mid_between + grand_mean, 2),
      OR = exp(estimate),
      CI_lo = exp(.data[[ci$lo]]),
      CI_hi = exp(.data[[ci$hi]])) |>
    dplyr::select(age_years, contrast, OR, CI_lo, CI_hi, p.value)
}

#1.9.2 Wave-specific cluster contrasts (OR) at wave-median ages, robust to wave_num
cluster_contrasts_by_wave_aligned <- function(fit, data) {
  waves <- levels(data$end_wave)
  
  #1.9.2.1 Empirical medians per wave
  ages_by_wave <- data %>%
    dplyr::group_by(end_wave) %>%
    dplyr::summarise(age_med = median(age_mid, na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(match(end_wave, waves))
  grand_mean <- mean(data$age_mid, na.rm = TRUE)
  mf_names <- try(names(model.frame(fit)), silent = TRUE)
  out_list <- vector("list", length(waves))
  for (i in seq_along(waves)) {
    w <- waves[i]
    a <- ages_by_wave$age_med[i]
    at <- list(
      cluster = levels(data$cluster),
      age_mid_between = a - grand_mean,
      age_mid_cwc = 0,
      baseline_status = 0,
      end_wave = factor(w, levels = waves),
      sex = levels(data$sex)[1],
      site_factor = levels(data$site_factor)[1])
    if (!inherits(mf_names, "try-error")) {
      
      #1.9.2.1.1 supply wave_num if needed; drop unused within-age variant, and unused wave coding
      if ("wave_num" %in% mf_names) at$wave_num <- i - 1L
      if (!"end_wave" %in% mf_names) at$end_wave <- NULL
      if (!"wave_num" %in% mf_names) at$wave_num <- NULL
      if ("log_dt" %in% mf_names) at$log_dt <- get_log_dt_for_wave(data, factor(w, levels = waves))
      at <- at[names(at) %in% mf_names]
    }
    em <- emmeans::emmeans(fit, ~ cluster, at = at)
    ct <- emmeans::contrast(em, "revpairwise", adjust = "none")
    sm <- summary(ct, infer = c(TRUE, TRUE)) |> tibble::as_tibble()
    ci <- emm_ci_cols(sm)
    out_list[[i]] <- sm |>
      dplyr::transmute(
        end_wave = as.character(w),
        contrast,
        OR = exp(estimate),
        CI_lo = exp(.data[[ci$lo]]),
        CI_hi = exp(.data[[ci$hi]]),
        p.value)
  }
  dplyr::bind_rows(out_list)
}


#1.10 Safe evaluation wrapper used to prevent per-outcome failures from aborting the loop
safe_or_null <- function(expr) {
  out <- try(expr, silent = TRUE)
  if (inherits(out, "try-error")) NULL else out
}

#1.11 UI helper for wrapped plot titles to avoid overflow
wrap_title <- function(x, width = 70) {
  stringr::str_wrap(x, width = width)
}

#1.12 Term filter for tables: keep clusters and core covariates; drop wave FEs and intercept
keep_terms <- function(df) {
  if (!"term" %in% names(df)) return(df)
  df %>%
    dplyr::filter(
      !grepl("^(Intercept)$", term),
      grepl("^cluster", term) |
      grepl("^age_mid_between$", term) |
      grepl("^age_mid_cwc$", term) |
      grepl("^baseline_status$", term) |
      grepl("^sex", term) |
      grepl("^cluster.*:age_mid_between$", term)
    )
}

#1.13 Refit if boundary (drop ~0 variance RE) - but NEVER drop participant_id
refit_if_boundary <- function(fit, form, data, link) {
  vc <- tryCatch(lme4::VarCorr(fit), error = function(e) NULL)
  if (is.null(vc)) return(fit)
  vv <- sapply(vc, function(x) attr(x, "stddev")^2)
  tiny <- names(vv)[vv < 1e-3]
  # do not drop the subject intercept even if tiny
  tiny <- tiny[!grepl("^participant_id", tiny)]
  if (length(tiny) == 0) return(fit)
  message("Dropping near-zero RE: ", paste(tiny, collapse = ", "))
  form2 <- update.formula(form, . ~ .)
  if (any(grepl("^family_id", tiny)))   form2 <- update.formula(form2, . ~ . - (1 | family_id))
  if (any(grepl("^site_factor", tiny))) form2 <- update.formula(form2, . ~ . - (1 | site_factor))
  fit_glmer_safe(form2, data, link = link)
}

#1.13.1 Detect whether a random-effect term for a given group is present in a formula
has_re_term <- function(form, grp) {
  bars <- try(lme4::findbars(form), silent = TRUE)
  if (inherits(bars, "try-error") || length(bars) == 0) return(FALSE)
  any(vapply(bars, function(b) {
    grepl(paste0("\\|\\s*", grp, "\\)"), paste(deparse(b), collapse = " "), perl = TRUE)
  }, logical(1)))
}

#1.13.2 Iteratively simplify random-effects if convergence/singularity persists
simplify_random_effects <- function(fit, form, data, link) {
  di <- try(diag_report(fit), silent = TRUE)
  if (inherits(di, "try-error")) return(fit)
  if (isTRUE(di$converged) && !isTRUE(di$is_singular)) return(fit)

  #1.13.2.1 Drop order: family first (we'll use CR2 by family), then site. Never drop participant.
  candidates <- c("family_id", "site_factor")
  best_fit  <- fit
  best_form <- form
  for (grp in candidates) {
    if (has_re_term(best_form, grp)) {
      form_try <- update.formula(best_form, as.formula(paste0(". ~ . - (1 | ", grp, ")")))
      fit_try  <- safe_or_null(fit_glmer_safe(form_try, data, link = link))
      if (!is.null(fit_try)) {
        di_try <- diag_report(fit_try)
        if (isTRUE(di_try$converged) && !isTRUE(di_try$is_singular)) {
          message("Simplified random-effects by dropping (1|", grp, ").")
          return(fit_try)
        } else {
          best_fit <- fit_try
          best_form <- form_try}
      }
    }
  }
  best_fit
}

#1.13.3 Force a participant-only RE structure
force_participant_only <- function(form, data, link) {
  
  #1.13.3.1 Strip out all REs except (1|participant_id)
  f2 <- update.formula(form, . ~ .)
  if (has_re_term(f2, "family_id"))   f2 <- update.formula(f2, . ~ . - (1 | family_id))
  if (has_re_term(f2, "site_factor")) f2 <- update.formula(f2, . ~ . - (1 | site_factor))
  
  #1.13.3.2 Ensure participant term is present
  if (!has_re_term(f2, "participant_id"))
    f2 <- update.formula(f2, . ~ . + (1 | participant_id))
  fit_glmer_safe(f2, data, link = link)
}

#1.14 Compute Brier score (lower is better; calibration-weighted accuracy)
brier_score <- function(obs, pr) {
  obs <- as.numeric(obs)
  mean((obs - pr)^2, na.rm = TRUE)
}

#1.15 Compute log loss (binary cross-entropy)
log_loss <- function(obs, pr, eps = 1e-12) {
  obs <- as.numeric(obs)
  prc <- pmin(pmax(pr, eps), 1 - eps)
  -mean(obs * log(prc) + (1 - obs) * log(1 - prc), na.rm = TRUE)
}

#1.16 Summarize performance per fit (R2, Brier, LogLoss); robust to GLMMs
perf_metrics <- function(fit, data) {
  pr <- try(predict(fit, type = "response"), silent = TRUE)
  R2m <- R2c <- NA_real_
  if (inherits(try(performance::r2_nakagawa, silent = TRUE), "function")) {
    r2 <- try(performance::r2_nakagawa(fit), silent = TRUE)
    if (!inherits(r2, "try-error")) {
      R2m <- as.numeric(r2$R2_marginal)
      R2c <- as.numeric(r2$R2_conditional)
    }
  }
  bs <- if (!inherits(pr, "try-error")) brier_score(data$event, pr) else NA_real_
  ll <- if (!inherits(pr, "try-error")) log_loss(data$event, pr) else NA_real_
  tibble(R2_marginal = R2m, R2_conditional = R2c, Brier = bs, LogLoss = ll)
}

#1.17 Save performance::check_model() panel to file
save_check_model_plots <- function(fit, tag, width = 9, height = 6, dpi = 220) {
  cm <- try(performance::check_model(fit), silent = TRUE)
  if (inherits(cm, "try-error")) return(NULL)
  p <- try(if (requireNamespace("see", quietly = TRUE)) see::plot(cm) else cm, silent = TRUE)
  if (inherits(p, "try-error")) return(NULL)
  fn <- file.path(OUT_DIR, glue::glue("check_model_{tag}.png"))
  if (requireNamespace("ragg", quietly = TRUE)) {
    ggplot2::ggsave(filename = fn, plot = p, device = ragg::agg_png, width = width, height = height, units = "in", dpi = dpi)
  } else {
    ggplot2::ggsave(filename = fn, plot = p, width = width, height = height, units = "in", dpi = dpi)
  }
  tibble(file = fn)
}

#1.18 Save DHARMa residual plots and tests; guarded if DHARMa missing
save_dharma_plots_and_tests <- function(fit, tag) {
  if (!requireNamespace("DHARMa", quietly = TRUE)) return(NULL)
  set.seed(params$seed)
  sr <- try(DHARMa::simulateResiduals(fit, n = 1000, refit = FALSE), silent = TRUE)
  if (inherits(sr, "try-error")) return(NULL)
  fn <- file.path(OUT_DIR, glue::glue("dharma_{tag}.png"))
  try({
    if (requireNamespace("ragg", quietly = TRUE)) {
      ragg::agg_png(filename = fn, width = 1800, height = 1200, units = "px", res = 144)
    } else {
      png(filename = fn, width = 1800, height = 1200, res = 144)
    }
    plot(sr)
    dev.off()
  }, silent = TRUE)
  td <- try(DHARMa::testDispersion(sr), silent = TRUE)
  tz <- try(DHARMa::testZeroInflation(sr), silent = TRUE)
  tibble(file = fn,
    p_dispersion = if (!inherits(td, "try-error")) td$p.value else NA_real_,
    p_zeroinfl = if (!inherits(tz, "try-error")) tz$p.value else NA_real_)
}

#1.19 Compute cumulative risk across waves for a reference profile
cumrisk_by_wave_ci <- function(fit, data, B = 500) {

  #1.19.1 Build reference newdata with REQUIRED covariates for the primary model
  nd <- expand.grid(
    end_wave = factor(levels(data$end_wave), levels = levels(data$end_wave)),
    cluster = factor(levels(data$cluster), levels = levels(data$cluster)),
    age_mid_between = 0,
    age_mid_cwc = 0,
    sex = factor(levels(data$sex)[1], levels = levels(data$sex))) %>%
    dplyr::arrange(cluster, end_wave)

  #1.19.2 Keep only variables the model actually uses; add any missing ones safely
  mf_names <- try(names(model.frame(fit)), silent = TRUE)

  if (!inherits(mf_names, "try-error")) {
    
    #1.19.2.1 If model uses wave_num, supply it
    if ("wave_num" %in% mf_names && !"wave_num" %in% names(nd)) {
      nd$wave_num <- match(nd$end_wave, levels(data$end_wave)) - 1L
    }

    #1.19.2.2 If model uses an offset/log_dt, supply it
    if ("log_dt" %in% mf_names && !"log_dt" %in% names(nd)) {
      nd$log_dt <- vapply(
        nd$end_wave,
        function(w) get_log_dt_for_wave(data, factor(w, levels = levels(data$end_wave))),
        numeric(1))
    }

    #1.19.2.3 Drop any extra columns not used by the model (prevents weird eval paths)
    nd <- nd[, intersect(names(nd), mf_names), drop = FALSE]
  }

  #1.19.3 Point predictions on response scale
  haz_hat <- as.numeric(predict(fit, newdata = nd, type = "response"))

  to_cum <- function(h) {
    df <- nd
    df$haz <- h
    df %>%
      dplyr::group_by(cluster) %>%
      dplyr::mutate(surv = cumprod(1 - haz), cumrisk = 1 - surv) %>%
      dplyr::ungroup() %>%
      dplyr::select(end_wave, cluster, cumrisk)
  }

  #1.19.4 Parametric bootstrap only for merMod objects; GLM should NOT use bootMer
  if (inherits(fit, "merMod")) {
    sim <- try(
      lme4::bootMer(
        fit,
        FUN = function(fm) as.numeric(predict(fm, newdata = nd, type = "response", re.form = NA)),
        nsim = B, use.u = FALSE, type = "parametric", seed = params$seed
      ),
      silent = TRUE
    )

    if (!inherits(sim, "try-error")) {
      sims <- tibble::as_tibble(sim$t)
      cis <- lapply(seq_len(nrow(nd)), function(j) {
        qs <- quantile(sims[[j]], probs = c(0.025, 0.975), na.rm = TRUE)
        tibble::tibble(row = j, lohaz = qs[[1]], hihaz = qs[[2]])
      }) %>% dplyr::bind_rows()

      nd2 <- nd %>%
        dplyr::mutate(row = dplyr::row_number(), haz = haz_hat) %>%
        dplyr::left_join(cis, by = "row")

      lo <- to_cum(nd2$lohaz) %>% dplyr::rename(cumrisk_lo = cumrisk)
      hi <- to_cum(nd2$hihaz) %>% dplyr::rename(cumrisk_hi = cumrisk)
      pt <- to_cum(nd2$haz)   %>% dplyr::rename(cumrisk    = cumrisk)

      return(
        pt %>%
          dplyr::left_join(lo, by = c("end_wave", "cluster")) %>%
          dplyr::left_join(hi, by = c("end_wave", "cluster"))
      )
    } else {
      warning("bootMer failed; returning point estimates only.")
    }
  }

  #1.19.5 GLM fallback: point estimates, NA CIs
  to_cum(haz_hat) %>% dplyr::mutate(cumrisk_lo = NA_real_, cumrisk_hi = NA_real_)
}

#1.20 Fixed-wave predictions with 95% CIs via emmeans (response scale) with safe fallback
pred_fixedwave_with_ci <- function(fit,
                                   data,
                                   ages = ages_pred_vec,
                                   wave_ref = params$wave_ref) {
  if (missing(ages) ||
      is.null(ages) || length(ages) == 0 || all(!is.finite(ages)))
    ages <- c(11.004729021633345, 13.116629238715475, 15.2109589041096)
  grand_mean <- mean(data$age_mid, na.rm = TRUE)
  at_list <- list(
    cluster = levels(data$cluster),
    age_mid_between = ages - grand_mean,
    age_mid_cwc = 0,
    baseline_status = 0,
    end_wave = factor(wave_ref, levels = levels(data$end_wave)),
    sex = levels(data$sex)[1],
    site_factor = levels(data$site_factor)[1])
  
  #1.20.1 Keep only variables that exist in the fitted model
  mf_names <- try(names(model.frame(fit)), silent = TRUE)
  if (!inherits(mf_names, "try-error")) {
    if (!inherits(mf_names, "try-error")) {
      #1.20.1.1 Map wave_ref -> numeric index if model uses wave_num
      if ("wave_num" %in% mf_names) {
        wlev <- levels(data$end_wave)
        at_list$wave_num <- match(wave_ref, wlev) - 1L
      }
      
      #1.20.1.2 Supply offset value if model uses log_dt for cloglog sensitivity
      if ("log_dt" %in% mf_names) {
        at_list$log_dt <- get_log_dt_for_wave(data, factor(wave_ref, levels = levels(data$end_wave)))
      }
      
      #1.20.1.3 Keep only columns the model actually uses
      at_list <- at_list[names(at_list) %in% mf_names]
    }
    
    #1.20.2 Try emmeans only; if that fails, fallback to predict()
    em <- try(emmeans::emmeans(fit,
                               ~ cluster |
                                 age_mid_between,
                               at = at_list,
                               type = "response"),
              silent = TRUE)
    if (!inherits(em, "try-error")) {
      sm <- tibble::as_tibble(summary(em, infer = c(TRUE, TRUE)))
      resp <- emm_resp_col(sm)
      ci <- emm_ci_cols(sm)
      lo_col <- if (is.null(ci$lo))
        NA_character_
      else
        ci$lo
      hi_col <- if (is.null(ci$hi))
        NA_character_
      else
        ci$hi
      return(
        dplyr::transmute(
          sm,
          cluster,
          age_years = round(age_mid_between + grand_mean, 2),
          pred = if (is.null(resp))
            NA_real_
          else
            .data[[resp]],
          lower = if (is.na(lo_col))
            NA_real_
          else
            .data[[lo_col]],
          upper = if (is.na(hi_col))
            NA_real_
          else
            .data[[hi_col]],
          end_wave = wave_ref))
    }
    
    #1.20.3 Fallback: point predictions (no CI)
    nd <- expand.grid(
      cluster = levels(data$cluster),
      age_mid_between = ages - grand_mean,
      age_mid_cwc = 0,
      end_wave = factor(wave_ref, levels = levels(data$end_wave)),
      sex = levels(data$sex)[1])
    if ("wave_num" %in% names(model.frame(fit))) {
      wlev <- levels(data$end_wave)
      nd$wave_num <- match(wave_ref, wlev) - 1L
    }
    if ("log_dt" %in% names(model.frame(fit))) {
      nd$log_dt <- get_log_dt_for_wave(data, factor(wave_ref, levels = levels(data$end_wave)))
    }
    nd$age_years <- round(nd$age_mid_between + grand_mean, 2)
    nd$pred <- as.numeric(predict(
      fit,
      newdata = nd,
      type = "response",
      re.form = NA))
    nd$lower <- NA_real_
    nd$upper <- NA_real_
    dplyr::select(nd, cluster, age_years, pred, lower, upper, end_wave)
  }
}

#1.21 Wave-aligned predictions (pair age to wave) with safe fallback
pred_aligned_with_ci <- function(fit, data, ages = ages_pred_vec) {
  if (missing(ages) || is.null(ages) || length(ages) == 0 || all(!is.finite(ages)))
    ages <- c(11.004729021633345, 13.116629238715475, 15.2109589041096)
  waves <- levels(data$end_wave)
  grand_mean <- mean(data$age_mid, na.rm = TRUE)
  k <- min(length(ages), length(waves))
  outs <- vector("list", k)
  for (i in seq_len(k)) {
    at_list <- list(
      cluster = levels(data$cluster),
      age_mid_between = ages[i] - grand_mean,
      age_mid_cwc = 0,
      baseline_status = 0,
      end_wave = factor(waves[i], levels = levels(data$end_wave)),
      sex = levels(data$sex)[1],
      site_factor = levels(data$site_factor)[1])
    mf_names <- try(names(model.frame(fit)), silent = TRUE)
    if (!inherits(mf_names, "try-error")) {
      if ("wave_num" %in% mf_names) at_list$wave_num <- i - 1L
      if ("log_dt" %in% mf_names)   at_list$log_dt  <- get_log_dt_for_wave(data, factor(waves[i], levels = waves))
      at_list <- at_list[names(at_list) %in% mf_names]
    }    
    em <- try(emmeans::emmeans(fit, ~ cluster, at = at_list, type = "response"), silent = TRUE)
    if (!inherits(em, "try-error")) {
      sm <- tibble::as_tibble(summary(em, infer = c(TRUE, TRUE)))
      resp <- emm_resp_col(sm); ci <- emm_ci_cols(sm)
      lo_col <- if (is.null(ci$lo)) NA_character_ else ci$lo
      hi_col <- if (is.null(ci$hi)) NA_character_ else ci$hi
      outs[[i]] <- dplyr::transmute(
        sm,
        cluster,
        end_wave = waves[i],
        age_years = round(ages[i], 2),
        pred = if (is.null(resp)) NA_real_ else .data[[resp]],
        lower = if (is.na(lo_col)) NA_real_ else .data[[lo_col]],
        upper = if (is.na(hi_col)) NA_real_ else .data[[hi_col]])
    } else {

      #1.21.1 Fallback: point predictions (no CI)
      nd <- expand.grid(
        cluster = levels(data$cluster),
        age_mid_between = ages[i] - grand_mean,
        age_mid_cwc = 0,
        end_wave = factor(waves[i], levels = levels(data$end_wave)),
        sex = levels(data$sex)[1])
      if ("wave_num" %in% names(model.frame(fit))) nd$wave_num <- i - 1L
      if ("log_dt" %in% names(model.frame(fit))) nd$log_dt <- get_log_dt_for_wave(data, factor(waves[i], levels = waves))
      outs[[i]] <- tibble::tibble(
        cluster = nd$cluster,
        end_wave = waves[i],
        age_years = round(ages[i], 2),
        pred = as.numeric(predict(fit, newdata = nd, type = "response", re.form = NA)),
        lower = NA_real_, upper = NA_real_)}
  }
  dplyr::bind_rows(outs)
}

#1.22 Per-interval hazard table by cluster x wave (95% CI via emmeans) with safe fallback
hazard_by_wave_ci <- function(fit, data) {
  at_list <- list(
    cluster = levels(data$cluster),
    age_mid_between = 0, age_mid_cwc = 0,
    baseline_status = 0,
    end_wave = levels(data$end_wave),
    wave_num = seq_along(levels(data$end_wave)) - 1L,  # handy if needed
    sex = levels(data$sex)[1],
    site_factor = levels(data$site_factor)[1])
  mf_names <- try(names(model.frame(fit)), silent = TRUE)

  #1.22.1 Choose which "wave" variable to condition on
  cond_var <- if (!inherits(mf_names, "try-error") && "end_wave" %in% mf_names) "end_wave"
    else if (!inherits(mf_names, "try-error") && "wave_num" %in% mf_names) "wave_num"
    else NULL
  if (!is.null(cond_var)) {
    at_use <- at_list[names(at_list) %in% mf_names]
    spec <- stats::as.formula(paste0("~ cluster | ", cond_var))
    em <- try(emmeans::emmeans(fit, spec, at = at_use, type = "response"), silent = TRUE)
    if (!inherits(em, "try-error")) {
      sm <- tibble::as_tibble(summary(em, infer = c(TRUE, TRUE)))
      resp <- emm_resp_col(sm); ci <- emm_ci_cols(sm)
      out <- dplyr::transmute(
        sm,
        end_wave = if (cond_var == "end_wave")
          as.character(.data$end_wave)
        else
          levels(data$end_wave)[.data$wave_num + 1L],
        cluster,
        hazard = if (is.null(resp)) NA_real_ else .data[[resp]],
        lo = if (is.null(ci$lo)) NA_real_ else .data[[ci$lo]],
        hi = if (is.null(ci$hi)) NA_real_ else .data[[ci$hi]])
      return(out)
    }
  }

  #1.22.2 Fallback: point preds; include wave_num if present so predict() is happy
  nd <- expand.grid(
    end_wave = levels(data$end_wave),
    cluster = levels(data$cluster),
    age_mid_between = 0, age_mid_cwc = 0,
    sex = levels(data$sex)[1]) %>% 
  dplyr::arrange(end_wave, cluster)
  if (!inherits(mf_names, "try-error") && "wave_num" %in% mf_names)
    nd$wave_num <- match(nd$end_wave, levels(data$end_wave)) - 1L
  if (!inherits(mf_names, "try-error") && "log_dt" %in% mf_names)
    nd$log_dt <- vapply(nd$end_wave, function(w) get_log_dt_for_wave(data, factor(w, levels = levels(data$end_wave))), numeric(1))
  tibble::tibble(
    end_wave = nd$end_wave,
    cluster = nd$cluster,
    hazard = as.numeric(predict(fit, newdata = nd, type = "response", re.form = NA)),
    lo = NA_real_, hi = NA_real_
  )
}

#1.23 Cumulative risk by wave with bootstrap CIs (parametric via bootMer)
cumrisk_by_wave_ci <- function(fit, data, B = 500) {
  
  #1.23.1 Build reference newdata
  nd <- expand.grid(
    end_wave = levels(data$end_wave),
    cluster = levels(data$cluster),
    age_mid_between = 0, age_mid_cwc = 0) %>% dplyr::arrange(cluster, end_wave)

  #1.23.2 If the fitted model used an offset(log_dt) (cloglog sensitivity), supply it
  mf_names <- try(names(model.frame(fit)), silent = TRUE)
  if (!inherits(mf_names, "try-error") && "log_dt" %in% mf_names) {
    nd$log_dt <- vapply(nd$end_wave, function(w) get_log_dt_for_wave(data, factor(w, levels = levels(data$end_wave))), numeric(1))
  }

  #1.23.3 Point predictions on response scale
  haz_hat <- as.numeric(predict(fit, newdata = nd, type = "response", re.form = NA))

  #1.23.4 Helper: hazards -> cumulative risk by wave
  to_cum <- function(h) {
    df <- nd
    df$haz <- h
    df %>%
      dplyr::group_by(cluster) %>%
      dplyr::mutate(surv = cumprod(1 - haz), cumrisk = 1 - surv) %>%
      dplyr::ungroup() %>%
      dplyr::select(end_wave, cluster, cumrisk)
  }

  #1.23.5 Try parametric bootstrap for CIs; fall back cleanly on any error
  can_boot <- isTRUE("bootMer" %in% getNamespaceExports("lme4"))
  if (can_boot) {
    sim <- try(
      lme4::bootMer(
        fit,
        FUN = function(fm) as.numeric(predict(fm, newdata = nd, type = "response", re.form = NA)),
        nsim = B, use.u = FALSE, type = "parametric", seed = params$seed),
      silent = TRUE)
    if (!inherits(sim, "try-error")) {
      sims <- as_tibble(sim$t)
      cis <- lapply(seq_len(nrow(nd)), function(j) {
        qs <- quantile(sims[[j]], probs = c(0.025, 0.975), na.rm = TRUE)
        tibble(row = j, lohaz = qs[[1]], hihaz = qs[[2]])
      }) %>% bind_rows()
      nd2 <- nd %>% dplyr::mutate(row = dplyr::row_number(), haz = haz_hat) %>% dplyr::left_join(cis, by = "row")
      lo <- to_cum(nd2$lohaz) %>% dplyr::rename(cumrisk_lo = cumrisk)
      hi <- to_cum(nd2$hihaz) %>% dplyr::rename(cumrisk_hi = cumrisk)
      pt <- to_cum(nd2$haz)   %>% dplyr::rename(cumrisk    = cumrisk)
      return(pt %>% dplyr::left_join(lo, by = c("end_wave","cluster")) %>% dplyr::left_join(hi, by = c("end_wave","cluster")))
    } else {
      warning("bootMer failed; returning point estimates only.")
    }
  } else {
    warning("bootMer not available; returning point estimates only.")
  }

  #1.23.6 Fallback: point estimates, NA CIs
  to_cum(haz_hat) %>% dplyr::mutate(cumrisk_lo = NA_real_, cumrisk_hi = NA_real_)
}

#1.24 Calibration summary (intercept, slope, ECE, MCE)
calibration_summary <- function(fit, data) {
  pr <- as.numeric(predict(fit, type = "response"))
  dat <- data.frame(event = as.numeric(data$event), pr = pr)
  dat$lp <- qlogis(pmin(pmax(dat$pr, 1e-9), 1-1e-9))

  #1.24.1 Intercept-only with offset => calibration-in-the-large (target 0)
  cil_fit <- glm(event ~ 1 + offset(lp), family = binomial(), data = dat)
  cil <- coef(cil_fit)[1]

  #1.24.2 Slope model (target slope ~1, intercept ~0)
  sl_fit <- glm(event ~ lp, family = binomial(), data = dat)
  slope <- coef(sl_fit)[2]; intercept <- coef(sl_fit)[1]

  #1.24.3 Deciles ECE/MCE
  dd <- dplyr::tibble(event = dat$event, pr = dat$pr) |>
    dplyr::mutate(dec = dplyr::ntile(pr, 10)) |>
    dplyr::group_by(dec) |>
    dplyr::summarise(pred = mean(pr), obs = mean(event), n = dplyr::n(), .groups="drop") |>
    dplyr::mutate(abs_err = abs(obs - pred))
  ece <- with(dd, sum(n / sum(n) * abs_err))
  mce <- max(dd$abs_err)
  tibble(cali_intercept = unname(cil), slope = unname(slope), intercept = unname(intercept), ECE = ece, MCE = mce)
}

#1.25 Extract clean emmeans results
#1.25.1 Pick the CI column names that exist in an emmeans summary
emm_ci_cols <- function(sm) {
  lo <- intersect(c("asymp.LCL","lower.CL","LCL"), names(sm))[1]
  hi <- intersect(c("asymp.UCL","upper.CL","UCL"), names(sm))[1]
  list(lo = lo, hi = hi)
}

#1.25.2 Pick the response-scale column from an emmeans summary
emm_resp_col <- function(sm) {
  # common names across families; last two are safe fallbacks
  out <- intersect(c("response","prob","rate","emmean"), names(sm))
  if (length(out)) out[[1]] else NULL
}

#1.26 Deep fixed-effect design diagnostics
design_diag <- function(form, data, out_name = "unknown") {

  #1.26.1 Build fixed-effect design matrix (no random effects)
  rhs <- lme4::nobars(form)
  X <- model.matrix(rhs, data = data)
  # drop intercept for kappa/VIF calcs
  Xi <- X[, colnames(X) != "(Intercept)", drop = FALSE]
  p <- ncol(Xi); n <- nrow(Xi)
  rk <- tryCatch(qr(Xi)$rank, error = function(e) NA_integer_)

  #1.26.2 Condition numbers (raw & scaled)
  kappa_raw <- tryCatch(kappa(Xi), error = function(e) NA_real_)
  kappa_scaled <- tryCatch(kappa(scale(Xi, center = TRUE, scale = TRUE)), error = function(e) NA_real_)
  
  #1.26.3 Per-term VIF (via performance::check_collinearity)
  vif_tbl <- tryCatch({
    cc <- performance::check_collinearity(stats::glm(form, data = data, family = binomial()))
    tibble::as_tibble(cc)
  }, error = function(e) {
    tryCatch({
      cc <- performance::check_collinearity(lme4::glmer(form, data = data, family = binomial(), nAGQ = 0))
      tibble::as_tibble(cc)
    }, error = function(e2) tibble())
  })
  
  #1.26.4 Top pairwise correlations among design columns (helps pinpoint culprits)
  cor_pairs <- tibble()
  if (p >= 2) {
    C <- tryCatch(cor(scale(Xi), use = "pairwise.complete.obs"), error = function(e) NULL)
    if (!is.null(C)) {
      C[lower.tri(C, diag = TRUE)] <- NA
      cor_pairs <- as.data.frame(as.table(C)) |>
        dplyr::rename(var1 = Var1, var2 = Var2, r = Freq) |>
        dplyr::filter(is.finite(r)) |>
        dplyr::arrange(dplyr::desc(abs(r))) |>
        dplyr::slice(1:min(20, dplyr::n())) |>
        tibble::as_tibble()
    }
  }
  
  #1.26.5Linear-combo detection if caret is present
  lc <- tryCatch({
    if (requireNamespace("caret", quietly = TRUE)) {
      z <- caret::findLinearCombos(Xi)
      tibble::tibble(linear_combos = if (length(z$combos)) sapply(z$combos, paste, collapse = " + ") else character())
    } else tibble()
  }, error = function(e) tibble())
  list(
    meta = tibble::tibble(outcome = out_name, n = n, p = p, rank = rk, kappa_raw = kappa_raw, kappa_scaled = kappa_scaled),
    vif = dplyr::mutate(vif_tbl, outcome = out_name, .before = 1),
    cors = dplyr::mutate(cor_pairs, outcome = out_name, .before = 1),
    lincomb = dplyr::mutate(lc, outcome = out_name, .before = 1)
  )
}

#1.27 Sparse/empty cell scan (common root cause for singularity/separation)
sparse_cells <- function(data, vars = c("end_wave","cluster","sex"), y = "event", min_n = 5) {
  df <- data |>
    dplyr::group_by(dplyr::across(dplyr::all_of(vars))) |>
    dplyr::summarise(
      n = dplyr::n(),
      n_events = sum(.data[[y]] == 1L, na.rm = TRUE),
      n_nonev  = sum(.data[[y]] == 0L, na.rm = TRUE),
      .groups = "drop") |>
    dplyr::mutate(flag_zero_event = n_events == 0,
      flag_zero_nonev = n_nonev == 0,
      flag_small = n < min_n)
  df
}

#1.28 Fixed-part separation screen for GLM only
sep_screen <- function(form, data, family) {
  if (!requireNamespace("detectseparation", quietly = TRUE)) {
    return(tibble::tibble(note = "detectseparation not installed; skipping"))
  }
  out <- tryCatch(
    detectseparation::detect_separation(form, data = data, family = family),
    error = function(e) e)
  if (inherits(out, "error")) {
    return(tibble::tibble(note = paste("detect_separation failed:", conditionMessage(out))))
  }
  tibble::tibble(
    separation = isTRUE(out$separation),
    note = if (isTRUE(out$separation)) "Possible separation in fixed part" else "No separation detected")
}

#1.29 Random-effects boundary report
re_boundary <- function(fit, out_name = "unknown") {
  vc <- tryCatch(lme4::VarCorr(fit), error = function(e) NULL)
  if (is.null(vc)) return(tibble())
  tibble::tibble(outcome = out_name,
    grp = names(vc),
    var = sapply(vc, function(x) attr(x, "stddev")^2),
    sd = sapply(vc, function(x) attr(x, "stddev"))) |>
    dplyr::mutate(flag_tiny = var < 1e-4)
}

#1.30 Collinearity + fit quality checks (used to pick a BD-II candidate)
max_vif <- function(fit) {
  co <- try(performance::check_collinearity(fit, component = "conditional"), silent = TRUE)
  if (inherits(co, "try-error")) return(Inf)
  suppressWarnings(max(co$VIF, na.rm = TRUE))
}

#1.30.1 Check convergence (robust across performance versions)
ok_convergence <- function(fit) {
  cv <- try(performance::check_convergence(fit), silent = TRUE)
  if (inherits(cv, "try-error")) return(TRUE)   # don't block candidate selection on a check error

  #1.30.1.1 Try several possible return shapes from performance::check_convergence()
  val <- tryCatch({
    if (!is.null(cv$converged)) {
      cv$converged
    } else if (is.data.frame(cv) && "converged" %in% names(cv)) {
      cv[["converged"]]
    } else if (!is.null(attr(cv, "converged"))) {
      attr(cv, "converged")
    } else if (is.list(cv) && length(cv) == 1 && is.logical(cv[[1]])) {
      cv[[1]]
    } else if (is.logical(cv) && length(cv) == 1) {
      cv
    } else {

      #1.30.1.2 Fallback on lme4 optinfo: consider "OK" if there are no convergence messages
      oi <- try(fit@optinfo, silent = TRUE)
      if (!inherits(oi, "try-error")) {
        msgs <- c(oi$conv$lme4$messages, oi$conv$lme4$message, oi$conv$messages)
        length(unlist(msgs)) == 0
      } else TRUE
    }
  }, error = function(e) NA)
  isTRUE(val)
}
 
#1.30.2 Residualize a numeric x on a factor f -> removes linear dependence on f levels.
resid_on_factor <- function(x, f) {

  #1.30.2.1 NA-robust residuals from x ~ f
  out <- rep(NA_real_, length(x))
  ok <- is.finite(x) & !is.na(f)
  if (any(ok)) out[ok] <- resid(lm(x ~ f, na.action = na.exclude))
  out
}

#1.30.3 Add modeling-ready derived terms
prep_dth_terms <- function(dat) {

  #1.30.3.1 Ensure wave_num exists
  dat <- dat %>%
    dplyr::mutate(
      wave_num = as.integer(end_wave) - 1L)

  #1.30.3.2 Ensure within-person orthogonalized age exists
  dat <- dat %>%
    dplyr::mutate(
      age_mid_cwc_orth = resid_on_factor(age_mid_cwc, end_wave)
    )

  #1.30.3.3 Ensure dt_years + log_dt exist and are finite, needed for cloglog + offset
  if (!("dt_years" %in% names(dat))) {
    
    #1.30.3.3.1 Attempt to reconstruct using the same wave_year logic we used earlier
    wave_year_local <- function(w) as.numeric(stringr::str_extract(as.character(w), "\\d+"))
    if (all(c("start_wave", "end_wave") %in% names(dat))) {
      dat <- dat %>%
        dplyr::mutate(
          dt_years = pmax(wave_year_local(end_wave) - wave_year_local(start_wave), 1),
          log_dt = log(dt_years))
    }
  }

  #1.30.3.4 Final guards; even if dt_years/log_dt were already present
  if (all(c("dt_years", "log_dt") %in% names(dat))) {
    dat <- dat %>%
      dplyr::mutate(
        dt_years = dplyr::if_else(is.finite(dt_years) & dt_years > 0, dt_years, NA_real_),
        log_dt = dplyr::if_else(is.finite(log_dt), log_dt, NA_real_)) %>%
      dplyr::filter(is.finite(dt_years), is.finite(log_dt))
  }

  dat
}

#1.31 Map each end_wave to an empirical log_dt (median) for a given outcome dataset.
log_dt_by_wave <- function(dat) {
  if (!all(c("end_wave", "log_dt") %in% names(dat))) return(NULL)
  dat %>%
    dplyr::group_by(end_wave) %>%
    dplyr::summarise(log_dt_med = median(log_dt, na.rm = TRUE), .groups = "drop")
}

get_log_dt_for_wave <- function(dat, wave) {
  m <- log_dt_by_wave(dat)
  if (is.null(m) || nrow(m) == 0) return(0)  # last-resort fallback
  v <- m$log_dt_med[m$end_wave == wave]
  if (length(v) == 0 || !is.finite(v[1])) 0 else v[1]
}

#1.32 Create a safe CSV writer that skips empties
write_if_any <- function(df, path) {
  if (is.data.frame(df) && nrow(df) > 0 && ncol(df) > 0) {
    readr::write_csv(df, path)
  } else {
    message("Skipping write for ", basename(path), " (empty).")
  }
}

#1.33 Model decision log documentation scaffold
decision_log_init <- function() {
  tibble::tibble(
    model_family = character(),       # e.g., GLM, GLMM, GAMM, Bayesian
    engine = character(),             # glm / lme4 / glmmTMB / mgcv / brms
    specification = character(),      # formula summary
    outcome = character(),            # bipolar_I / ...
    status = character(),             # used_primary / sensitivity / attempted_failed / attempted_unstable
    key_diagnostics = character(),    # short: ICC, VIF, separation, etc
    decision_rationale = character(), # why kept/dropped
    artifact = character()            # file path(s) if saved
  )
}

add_decision <- function(log_tbl,
                         model_family, engine, specification,
                         outcome = NA_character_,
                         status, key_diagnostics = NA_character_,
                         decision_rationale = NA_character_,
                         artifact = NA_character_) {
  dplyr::bind_rows(
    log_tbl,
    tibble::tibble(
      model_family = model_family,
      engine = engine,
      specification = specification,
      outcome = outcome,
      status = status,
      key_diagnostics = key_diagnostics,
      decision_rationale = decision_rationale,
      artifact = artifact
    )
  )
}

```

## 2. Fit Discrete-Time Hazard Models

-   Iterate over BD outcomes and fit pooled logistic DTH via `glm()` (primary logit link)

-   Compute participant-clustered robust inference (CR2 primary; HC1 secondary)

-   Fit sensitivity models: `wave_num`, `age_mid_cwc_orth`, and cloglog + offset(log_dt)

-   Fit mixed-effects versions only as diagnostics/sensitivity: compute ICC from `glmer(..., nAGQ=0)` and document instability for `nAGQ=1` plus optimizer dependence via `allFit()`

-   Persist all results (i.e., tables + model objects + diagnostic evidence) to `OUT_DIR`

```{r fit dth models, warning = FALSE}

## 2. Fit models (primary GLM + sensitivities), summarize, and save ##

#2.1 Outcomes to iterate on for this analysis
OUTCOMES <- c("bipolar_I","bipolar_II","bd_nos","any_bsd")

#2.2 Containers
fits_primary <- list()
fits_sens_cloglog <- list()
fits_sens_linT <- list()
fits_sens_within <- list()

tbl_cr2_primary <- list()
tbl_hc1_primary <- list()

tbl_cr2_cloglog <- list()
tbl_hc1_cloglog <- list()

tbl_cr2_linT <- list()
tbl_cr2_within <- list()

tbl_cr2_primary_fam <- list()

#2.3 Evidence objects supporting why GLMM is not primary for this analysis
vif_evidence <- list()
icc_evidence <- list()
allfit_evidence <- list()
or_compare <- list()

#2.3.1 Model decision log; seed with previously-attempted families so they appear in the HTML
model_log <- decision_log_init()

model_log <- add_decision(
  model_log,
  model_family = "DTH pooled logistic (primary)",
  engine = "glm + clubSandwich",
  specification = "event ~ cluster + sex + end_wave + age_mid_between; CR2 clustered by participant (HC1 secondary)",
  status = "used_primary",
  key_diagnostics = "Stable fits across outcomes; design-consistent inference with clustered SEs",
  decision_rationale = "Primary model: avoids weakly-identified random intercept with 1–3 rows/person and sparse events")

model_log <- add_decision(
  model_log,
  model_family = "DTH cloglog + offset(log_dt)",
  engine = "glm + clubSandwich",
  specification = "event ~ cluster + sex + end_wave + age_mid_between + offset(log_dt); CR2 clustered by participant",
  status = "sensitivity",
  key_diagnostics = "Robustness to unequal interval lengths (missed waves -> longer dt)",
  decision_rationale = "Sensitivity: approximates continuous-time proportional hazards under discrete intervals")

#2.3.2 Seed previously tried approaches even if not run in this script
model_log <- add_decision(
  model_log,
  model_family = "GLMM (random intercept)",
  engine = "lme4::glmer",
  specification = "event ~ cluster + sex + end_wave + age_mid_between + (1|participant_id) [+ optional (1|family_id)/(1|site_factor)]",
  status = "diagnostic_only",
  key_diagnostics = "Often optimizer-dependent / degenerate Hessians when wave FE + within-age included; RE weakly identified",
  decision_rationale = "Documented as diagnostics; not used for primary inference")

model_log <- add_decision(
  model_log,
  model_family = "GAMM / spline time (previous)",
  engine = "mgcv (e.g., gam/bam)",
  specification = "event ~ cluster + sex + s(age_mid) + (random effects as applicable) + wave terms",
  status = "attempted_previous",
  key_diagnostics = "Add your prior notes/results here (e.g., concurvity, unstable smooths with only 3 waves, separation in cells)",
  decision_rationale = "Not carried forward: limited unique timepoints (3 waves) makes smooth terms poorly supported")

model_log <- add_decision(
  model_log,
  model_family = "glmmTMB (previous)",
  engine = "glmmTMB",
  specification = "Binomial mixed model variants (including alternative RE structures / dispersion handling)",
  status = "attempted_previous",
  key_diagnostics = "Add your prior notes/results here (e.g., convergence warnings, boundary fits, separation)",
  decision_rationale = "Not carried forward if convergence/identifiability issues persisted or estimates were optimizer-dependent")

model_log <- add_decision(
  model_log,
  model_family = "Bayesian (brms) (previous)",
  engine = "brms",
  specification = "Bernoulli/logit or cloglog with priors; random intercepts; optional varying effects",
  status = "attempted_previous",
  key_diagnostics = "Add your prior notes/results here (e.g., divergent transitions, weak ID for RE, sensitivity to priors)",
  decision_rationale = "Not carried forward if computationally heavy without improving stability/interpretability given sparse intervals")

#2.4 Loop per outcome
for (out in OUTCOMES) {

  message("Primary GLM (logit): ", out)
  dat_o <- bd_pp_model %>% dplyr::filter(outcome == out) %>% droplevels() %>% prep_dth_terms()
  
  #2.4.0 Guard: ensure sex has at least 1 non-NA level and remains a factor
  if (!("sex" %in% names(dat_o))) stop("dat_o missing sex for outcome ", out)
  if (!is.factor(dat_o$sex)) dat_o$sex <- factor(dat_o$sex)
  if (nlevels(dat_o$sex) < 1) stop("sex has 0 levels for outcome ", out)

  #2.4.1 Primary pooled logistic DTH
  f_primary <- build_formula_primary()
  m_glm <- fit_glm_safe(f_primary, dat_o, link = "logit")
  fits_primary[[out]] <- m_glm
  
  #2.4.2 GLM diagnostics: separation, influence, binned residuals, check_model panels
  #2.4.2.1 Separation screen for the actual primary model too
  sep_primary <- sep_screen(f_primary, dat_o, family = binomial(link="logit"))
  write_if_any(sep_primary, file.path(OUT_DIR, glue("sep_primary_{out}.csv")))
  
  #2.4.2.2 Append outcome-specific diagnostic snapshot to the model decision log
  nevents <- sum(dat_o$event == 1L, na.rm = TRUE)
  nrows <- nrow(dat_o)
  sep_flag <- if ("separation" %in% names(sep_primary)) as.character(sep_primary$separation[1]) else NA_character_

  model_log <- add_decision(
    model_log,
    model_family = "DTH pooled logistic (primary)",
    engine = "glm + clubSandwich",
    specification = "event ~ cluster + sex + end_wave + age_mid_between; CR2 clustered by participant",
    outcome = out,
    status = "used_primary",
    key_diagnostics = glue::glue("n_rows={nrows}; n_events={nevents}; separation={sep_flag}"),
    decision_rationale = "Primary inference run")
  
  #2.4.2.3 Sparse cell scan; helps interpret separation / instability
  sc <- sparse_cells(dat_o, vars = c("end_wave","cluster","sex"), y = "event", min_n = 5)
  write_if_any(sc, file.path(OUT_DIR, glue("sparse_cells_{out}.csv")))
  
  #2.4.2.4 Influence measures
  infl_tbl <- tibble(
    outcome = out,
    .row = seq_len(nrow(dat_o)),
    hat = hatvalues(m_glm),
    cooksD = cooks.distance(m_glm))
  write_if_any(infl_tbl, file.path(OUT_DIR, glue("influence_basic_{out}.csv")))
  
  #2.4.2.5 Top influential rows
  top_infl <- infl_tbl %>% arrange(desc(cooksD)) %>% slice(1:25)
  write_if_any(top_infl, file.path(OUT_DIR, glue("influence_top25_{out}.csv")))
  
  #2.4.2.6 Binned residual plot (arm::binnedplot if available; otherwise skip)
  if (requireNamespace("arm", quietly = TRUE)) {
    fn <- file.path(OUT_DIR, glue("binned_resid_{out}.png"))
    if (requireNamespace("ragg", quietly = TRUE)) ragg::agg_png(fn, width=1800, height=1200, res=144) else png(fn, width=1800, height=1200, res=144)
    arm::binnedplot(x = fitted(m_glm), y = residuals(m_glm, type = "response"),
                    xlab = "Fitted", ylab = "Response residual")
    dev.off()
  }
  
  #2.4.2.7 check_model panels to file
  save_check_model_plots(m_glm, tag = glue("glm_{out}"))

  #2.4.3 Participant-clustered inference (CR2 primary; HC1 secondary)
  tbl_cr2_primary[[out]] <- safe_or_null(
  tidy_glm_cr2(m_glm, cluster_var = dat_o$participant_id) %>%
    dplyr::mutate(outcome = out, model = "logit_primary", se_type = "CR2", .before = 1))

  tbl_hc1_primary[[out]] <- safe_or_null(
  tidy_glm_hc1(m_glm, cluster_var = dat_o$participant_id) %>%
    dplyr::mutate(outcome = out, model = "logit_primary", se_type = "HC1", .before = 1))
  
  tbl_cr2_primary_fam[[out]] <- tidy_glm_cr2(m_glm, cluster_var = dat_o$family_id) %>%
  mutate(outcome = out, model = "logit_primary", se_type = "CR2_family", .before = 1)

  #2.4.4 Sensitivity A: linear time (wave_num)
  m_linT <- fit_glm_safe(build_formula_lin_time(), dat_o, link = "logit")
  fits_sens_linT[[out]] <- m_linT
  tbl_cr2_linT[[out]] <- tidy_glm_cr2(m_linT, cluster_var = dat_o$participant_id) %>%
    dplyr::mutate(outcome = out, model = "logit_lin_time", se_type = "CR2", .before = 1)

  #2.4.5 Sensitivity B: orthogonalized within-person age
  m_within <- fit_glm_safe(build_formula_within_orth(), dat_o, link = "logit")
  fits_sens_within[[out]] <- m_within
  tbl_cr2_within[[out]] <- tidy_glm_cr2(m_within, cluster_var = dat_o$participant_id) %>%
    dplyr::mutate(outcome = out, model = "logit_within_orth", se_type = "CR2", .before = 1)

  #2.4.6 Sensitivity C: cloglog + offset(log_dt)
  f_cloglog <- update.formula(f_primary, . ~ . + offset(log_dt))
  sep_cloglog <- sep_screen(f_cloglog, dat_o, family = binomial(link="cloglog"))
  m_clog <- fit_glm_safe(f_cloglog, dat_o, link = "cloglog")
  fits_sens_cloglog[[out]] <- m_clog

  tbl_cr2_cloglog[[out]] <- tidy_glm_cr2(m_clog, cluster_var = dat_o$participant_id) %>%
    dplyr::mutate(outcome = out, model = "cloglog_offset", se_type = "CR2", .before = 1)

  tbl_hc1_cloglog[[out]] <- tidy_glm_hc1(m_clog, cluster_var = dat_o$participant_id) %>%
    dplyr::mutate(outcome = out, model = "cloglog_offset", se_type = "HC1", .before = 1)

  #2.4.7 Evidence supporting dropping raw age_mid_cwc from primary when using wave indicators, AKA VIF / collinearity
  vif_tbl <- safe_or_null({
    cc <- performance::check_collinearity(glm(event ~ cluster + sex + end_wave + age_mid_between + age_mid_cwc,
                                              data = dat_o, family = binomial()))
    tibble::as_tibble(cc) %>% dplyr::mutate(outcome = out, .before = 1)
  })
  vif_evidence[[out]] <- vif_tbl

  #2.4.8 Evidence: GLMM ICC + allFit optimizer dependence at nAGQ>=1 (diagnostic only)
  #2.4.8.1 ICC from nAGQ=0 participant-only RE
  glmm0 <- safe_or_null(
    lme4::glmer(event ~ cluster + sex + end_wave + age_mid_between + (1|participant_id),
                data = dat_o, family = binomial(), nAGQ = 0,
                control = glmerControl(optimizer="bobyqa", optCtrl=list(maxfun=2e5)))
  )
  if (!is.null(glmm0)) {
    vc <- as.data.frame(VarCorr(glmm0))$vcov[1]
    icc <- vc / (vc + (pi^2 / 3))
    icc_evidence[[out]] <- tibble::tibble(outcome = out, vcov = vc, icc = icc)
  }

  #2.4.8.2 nAGQ=1 + allFit: store convergence/degeneracy across optimizers if it runs
  glmm1 <- safe_or_null(tryCatch(update(glmm0, nAGQ = 1), error = function(e) NULL))
  if (!is.null(glmm1)) {
    af <- safe_or_null(lme4::allFit(glmm1))
    if (!is.null(af)) {
      ss <- summary(af)
      allfit_evidence[[out]] <- tibble::tibble(
        outcome = out,
        n_ok = ss$OK,
        n_fail = ss$failed,
        n_warn = ss$warnings
      )
      saveRDS(af, file.path(OUT_DIR, glue::glue("allFit_nAGQ1_{out}.rds")))
    }
  }

  #2.4.9 Evidence of point-estimate stability in GLM vs glmer nAGQ=0
  if (!is.null(glmm0)) {
    or_compare[[out]] <- tibble::tibble(
      outcome = out,
      OR_glm = exp(coef(m_glm)[["clusterC2"]]),
      OR_glmer_nAGQ0 = exp(lme4::fixef(glmm0)[["clusterC2"]])
    )
    saveRDS(glmm0, file.path(OUT_DIR, glue::glue("glmer_nAGQ0_diag_{out}.rds")))
  }
}

#2.5 Bind and save outputs
cr2_all <- bind_rows(compact(c(tbl_cr2_primary, tbl_cr2_primary_fam,
                               tbl_cr2_linT, tbl_cr2_within, tbl_cr2_cloglog)))
hc1_all <- dplyr::bind_rows(purrr::compact(c(tbl_hc1_primary, tbl_hc1_cloglog)))

vif_all <- dplyr::bind_rows(purrr::compact(vif_evidence))
icc_all <- dplyr::bind_rows(purrr::compact(icc_evidence))
allfit_all <- dplyr::bind_rows(purrr::compact(allfit_evidence))
or_compare_all <- dplyr::bind_rows(purrr::compact(or_compare))

readr::write_csv(cr2_all, file.path(OUT_DIR, "bd_dth_inference_CR2.csv"))
readr::write_csv(hc1_all, file.path(OUT_DIR, "bd_dth_inference_HC1_participant.csv"))
write_if_any(vif_all, file.path(OUT_DIR, "bd_dth_vif_evidence.csv"))
write_if_any(icc_all, file.path(OUT_DIR, "bd_dth_icc_evidence.csv"))
write_if_any(allfit_all, file.path(OUT_DIR, "bd_dth_allFit_nAGQ1_summary.csv"))
write_if_any(or_compare_all, file.path(OUT_DIR, "bd_dth_or_compare_glm_vs_glmer0.csv"))

#2.5.1 Persist and display the model decision log
readr::write_csv(model_log, file.path(OUT_DIR, "bd_dth_model_decision_log.csv"))

knitr::kable(
  model_log,
  caption = "Model decision log (primary + sensitivity + previously attempted strategies)") %>% kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))

#2.6 Create pred objects for downstream utilization
#2.6.1 Wald tables for reference from model-based SE; not robust
wald_all <- bind_rows(lapply(fits_primary, tidy_wald), .id = "outcome") %>%
  mutate(model = "logit_primary", .before = 1)

#2.6.2 Predictions for plots using primary link
pred_all_fixedwave <- bind_rows(lapply(OUTCOMES, function(out) {
  fit <- fits_primary[[out]]
  dat <- bd_pp_model %>% filter(outcome == out) %>% droplevels() %>% prep_dth_terms()
  df  <- pred_fixedwave_with_ci(fit, dat, ages = ages_pred_vec, wave_ref = params$wave_ref)
  mutate(df, outcome = out, model = params$link_primary, pred_mode = "fixed_wave", .before = 1)
}))

pred_all_aligned <- bind_rows(lapply(OUTCOMES, function(out) {
  fit <- fits_primary[[out]]
  dat <- bd_pp_model %>% filter(outcome == out) %>% droplevels() %>% prep_dth_terms()
  df  <- pred_aligned_with_ci(fit, dat, ages = ages_pred_vec)
  mutate(df, outcome = out, model = params$link_primary, pred_mode = "wave_aligned", .before = 1)
}))

#2.6.3 Calibration deciles using primary link
cal_all <- bind_rows(lapply(OUTCOMES, function(out) {
  fit <- fits_primary[[out]]
  dat <- bd_pp_model %>% filter(outcome == out) %>% droplevels() %>% prep_dth_terms()
  df <- calibration_df(fit, dat)
  mutate(df, outcome = out, model = params$link_primary, .before = 1)
}))

#2.6.4 Hazard-by-wave tables
hazard_tables_all <- bind_rows(lapply(OUTCOMES, function(out) {
  fit <- fits_primary[[out]]
  dat <- bd_pp_model %>% filter(outcome == out) %>% droplevels() %>% prep_dth_terms()
  df  <- hazard_by_wave_ci(fit, dat)
  mutate(df, outcome = out, model = params$link_primary, .before = 1)
}))

#2.6.5 Cumulative risk tables
cumrisk_tables_all <- dplyr::bind_rows(lapply(OUTCOMES, function(out) {
  fit <- fits_primary[[out]]
  dat <- bd_pp_model %>% filter(outcome == out) %>% droplevels() %>% prep_dth_terms()
  df <- tryCatch(
    cumrisk_by_wave_ci(fit, dat, B = 500),
    error = function(e) {
      warning("cumrisk_by_wave_ci failed for ", out, ": ", conditionMessage(e))
      tibble::tibble(
        end_wave = factor(levels(dat$end_wave), levels = levels(dat$end_wave)),
        cluster = factor(rep(levels(dat$cluster), each = length(levels(dat$end_wave))),
                         levels = levels(dat$cluster)),
        cumrisk = NA_real_, cumrisk_lo = NA_real_, cumrisk_hi = NA_real_)
    })

  dplyr::mutate(df, outcome = out, model = params$link_primary, .before = 1)
}))

#2.6.6 Cluster contrasts
emm_all <- bind_rows(lapply(OUTCOMES, function(out) {
  fit <- fits_primary[[out]]
  dat <- bd_pp_model %>% filter(outcome == out) %>% droplevels() %>% prep_dth_terms()
  df  <- cluster_contrasts_by_age(fit, dat, ages = ages_pred_vec, wave_ref = params$wave_ref)
  mutate(df, outcome = out, model = params$link_primary, .before = 1)
}))

emm_wave_primary_all <- bind_rows(lapply(OUTCOMES, function(out) {
  fit <- fits_primary[[out]]
  dat <- bd_pp_model %>% filter(outcome == out) %>% droplevels() %>% prep_dth_terms()
  df  <- cluster_contrasts_by_wave_aligned(fit, dat)
  mutate(df, outcome = out, model = params$link_primary, .before = 1)
}))

#2.6.7 Write all pred objects to disk for reference
write_csv(wald_all, file.path(OUT_DIR, "bd_dth_wald_reference.csv"))
write_csv(pred_all_fixedwave, file.path(OUT_DIR, "bd_dth_preds_fixedwave.csv"))
write_csv(pred_all_aligned, file.path(OUT_DIR, "bd_dth_preds_wave_aligned.csv"))
write_csv(cal_all, file.path(OUT_DIR, "bd_dth_calibration_deciles.csv"))
write_csv(hazard_tables_all, file.path(OUT_DIR, "bd_dth_hazard_by_wave.csv"))
write_csv(cumrisk_tables_all, file.path(OUT_DIR, "bd_dth_cumrisk_by_wave.csv"))
write_csv(emm_all, file.path(OUT_DIR, "bd_dth_emmeans_contrasts_by_age.csv"))
write_csv(emm_wave_primary_all, file.path(OUT_DIR, "bd_dth_emmeans_contrasts_by_wave.csv"))

#2.7 Save models
purrr::iwalk(fits_primary, ~ saveRDS(.x, file.path(OUT_DIR, glue::glue("glm_primary_{.y}.rds"))))
purrr::iwalk(fits_sens_linT, ~ saveRDS(.x, file.path(OUT_DIR, glue::glue("glm_linT_{.y}.rds"))))
purrr::iwalk(fits_sens_within, ~ saveRDS(.x, file.path(OUT_DIR, glue::glue("glm_withinOrth_{.y}.rds"))))
purrr::iwalk(fits_sens_cloglog, ~ saveRDS(.x, file.path(OUT_DIR, glue::glue("glm_cloglogOffset_{.y}.rds"))))

#2.8 Show key evidence tables in HTML
if (nrow(icc_all) > 0) {
  knitr::kable(icc_all, digits = 4, caption = "ICC evidence from participant-only glmer (nAGQ=0; diagnostic)") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))
}
if (nrow(or_compare_all) > 0) {
  knitr::kable(or_compare_all, digits = 4, caption = "Point-estimate stability: OR(cluster) GLM vs glmer nAGQ=0") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))
}
if (nrow(allfit_all) > 0) {
  knitr::kable(allfit_all, digits = 0, caption = "Optimizer dependence: allFit summary for nAGQ=1 fits (diagnostic)") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))
}

```

## 3. Determine Marginal Effects: Hazard by Cluster Across Age

Generates visualizations of predicted first-onset hazard vs age for each outcome and cluster, pairing each empirical age with its corresponding wave (02A/04A/06A) and within-person age deviation at 0. These visualize risk separation and trajectories across the median ages in the data at each timepoint

### How to read the hazard plots (interpretation guide)

-   **Fixed-wave age curves** (first panel): We hold the wave constant at `r params$wave_ref` and vary age.\
    This isolates how the hazard changes with age within a given wave. Lines show model-predicted first-onset hazard per interval; higher lines = higher risk. Compare clusters at the same age to see separation

-   **Wave-aligned curves** (second panel): Each point/age is paired with its actual wave (e.g., median ages at 02A, 04A, 06A).\
    This shows how the baseline hazard changes across waves while also reflecting the typical age at each wave.\
    A downward slope here does *not* contradict rising age risk - it reflects how the *wave baseline* (calendar-time hazard) moves across assessments

-   **QC overlay (model vs empirical)**: Hollow points are empirical per-wave hazards; solid lines are the model predictions on the same x-axis (wave). Close agreement indicates good calibration by wave

-   **Cumulative first-onset risk** (third panel): Converts per-interval hazards into the probability of having had a first onset by each wave

    -   This is the "when do onsets happen" view most would expect - rising curves that separate by cluster indicate earlier and/or more frequent onsets in the higher-risk cluster

```{r marginal effects, echo = FALSE, warning = FALSE, fig.height=11, fig.width=8.5}

## 3. Marginal hazard by cluster across age ##

#3.0.1 ROBUST SCHEMAS so filters/plots don't error if preds are empty
empty_pred_df <- function() {
  tibble(
    outcome = character(),
    model = character(),
    pred_mode = character(),
    end_wave = factor(character(), levels = levels(bd_pp_model$end_wave)),
    cluster = factor(character(), levels = levels(bd_pp_model$cluster)),
    age_years = double(),
    pred = double(),
    lower = double(),
    upper = double()
  )
}

#3.0.2 Ensure the df's exist for plotting, create if not
if (!exists("pred_all_aligned") || !"model" %in% names(pred_all_aligned)) pred_all_aligned <- empty_pred_df()
if (!exists("pred_all_fixedwave") || !"model" %in% names(pred_all_fixedwave)) pred_all_fixedwave <- empty_pred_df()

#3.0.3 Split predicted frames
pred_aligned <- pred_all_aligned %>% filter(model == params$link_primary)
pred_fixedwave <- pred_all_fixedwave %>% filter(model == params$link_primary)

#3.1 Plot helpers
#3.1.1 Fixed age predictions
plot_pred_fixed <- function(df, title) {
  has_ci <- all(c("lower","upper") %in% names(df)) && any(is.finite(df$lower) & is.finite(df$upper))
  ggplot(df, aes(x = age_years, y = pred, color = cluster, group = cluster)) +
    { if (has_ci) geom_ribbon(aes(ymin = lower, ymax = upper, fill = cluster), alpha = 0.15, color = NA) } +
    geom_line(size = 1.0) +
    geom_point(size = 1.8) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
    labs(title = wrap_title(title, 70),
      x = "Age (years)", y = "Predicted first-onset hazard (per interval)",
      color = "Cluster", fill = "Cluster") +
    theme_minimal(base_size = 9) +
    theme(legend.position = "bottom", legend.box = "horizontal", plot.title = element_text(lineheight = 0.95)) +
    guides(color = guide_legend(nrow = 1), fill = guide_legend(nrow = 1))
}

#3.1.2 Wave-aligned predictions (discrete x -> use error bars)
plot_pred_aligned <- function(df, title) {
  has_ci <- all(c("lower","upper") %in% names(df)) && any(is.finite(df$lower) & is.finite(df$upper))
  ggplot(df, aes(x = end_wave, y = pred, color = cluster, group = cluster)) +
    { if (has_ci) geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15, alpha = 0.7) } +
    geom_line(size = 1.0) +
    geom_point(size = 1.8) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
    labs(title = wrap_title(title, 70),
      x = "Wave (interval end)", y = "Predicted first-onset hazard (per interval)",
      color = "Cluster") +
    theme_minimal(base_size = 9) +
    theme(legend.position = "bottom", legend.box = "horizontal", plot.title = element_text(lineheight = 0.95)) +
    guides(color = guide_legend(nrow = 1))
}

#3.2 Fixed-wave age curves (pure age effect at wave_ref)
plots_fixed <- lapply(OUTCOMES, function(out) {
  df <- pred_fixedwave %>% filter(outcome == out)
  ttl <- glue("Predicted hazard by cluster across age (FIXED wave={params$wave_ref}; link={params$link_primary}) - {out}")  
  plot_pred_fixed(df, ttl)
})

#3.2.1 Wrap and print the fixed wave age curves
wrap_plots(plots_fixed, ncol = 2)

#3.3 Wave-aligned curves (age paired to wave)
plots_aligned <- lapply(OUTCOMES, function(out) {
  df <- pred_aligned %>% filter(outcome == out)
  ttl <- glue("Predicted hazard by cluster across wave (age paired to wave; link={params$link_primary}) - {out}")
  plot_pred_aligned(df, ttl)
})

#3.3.1 Wrap and print the wave aligned hazard plots
wrap_plots(plots_aligned, ncol = 2)

#3.4 QC overlay: model vs empirical per-wave hazards (aligned scale on the x-axis)
empirical_haz <- function(df) {
  df %>% group_by(outcome, end_wave, cluster) %>% summarise(emp = mean(event == 1L), .groups = "drop")
}
qc_plots <- lapply(OUTCOMES, function(out) {
  dfm <- pred_aligned %>% filter(outcome == out)
  dfe <- bd_pp_model %>% filter(outcome == out) %>% empirical_haz()
  ggplot() +
    geom_line(data = dfm, aes(x = end_wave, y = pred, color = cluster, group = cluster)) +
    geom_point(data = dfe, aes(x = end_wave, y = emp, color = cluster), shape = 1, size = 2) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
    labs(title = wrap_title(glue("Model vs empirical per-wave hazards - {out}"), 70),
    x = "Wave (interval end)", y = "Hazard (per interval)") +
    theme_minimal(base_size = 9) + theme(legend.position = "bottom")
})

#3.4.1 Wrap and print the QC plots
wrap_plots(qc_plots, ncol = 2)

#3.5 Cumulative incidence by wave (reference profile)
cum_plots <- lapply(OUTCOMES, function(out) {
  fit <- fits_primary[[out]]
  dat <- bd_pp_model %>% filter(outcome == out)

  #3.5.1 Use the CI version
  cr <- tryCatch(
  cumrisk_by_wave_ci(fit, dat, B = 500),
  error = function(e) {
    warning("cumrisk_by_wave_ci failed for plot ", out, ": ", conditionMessage(e))
    tibble::tibble(
      end_wave = factor(levels(dat$end_wave), levels = levels(dat$end_wave)),
      cluster = factor(levels(dat$cluster), levels = levels(dat$cluster))) %>%
      tidyr::crossing(end_wave) %>%
      dplyr::mutate(cumrisk = NA_real_, cumrisk_lo = NA_real_, cumrisk_hi = NA_real_)
  }) 
  has_ci <- "cumrisk_lo" %in% names(cr) && "cumrisk_hi" %in% names(cr) &&
            any(is.finite(cr$cumrisk_lo) & is.finite(cr$cumrisk_hi))
  ggplot(cr, aes(x = end_wave, y = cumrisk, color = cluster, group = cluster)) +
    { if (has_ci) geom_ribbon(aes(ymin = cumrisk_lo, ymax = cumrisk_hi, fill = cluster),
      alpha = 0.15, color = NA) } +
    geom_line() + geom_point(size = 1.8) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, 1)) +
    labs(title = wrap_title(glue("Cumulative first-onset risk by wave (ref profile) - {out}"), 70),
      x = "Wave (interval end)", y = "Cumulative risk") +
    theme_minimal(base_size = 9) +
    theme(legend.position = "bottom", plot.title = element_text(lineheight = 0.95)) +
    guides(color = guide_legend(nrow = 1), fill = guide_legend(nrow = 1))
})

#3.5.2 Wrap and print the cumulative plots
wrap_plots(cum_plots, ncol = 2)

```

## 4. Baseline Hazard (Wave Effects) - Interpretability QC

Uses the fitted model to compute marginal baseline hazards by wave (sessions 02A, 04A, 06A) while holding other covariates at reference. Bar plots provide a quick sanity check that interval risks align with expected cohort timing

```{r baseline hazard wave effects, warning = FALSE, fig.height=11, fig.width=8.5}

## 4. Baseline hazard (wave FE -> marginal hazard by wave) ##

#4.1 Create a helper function to plot marginal hazard effects by assessment timepoint (wave)
plot_wave_hazard <- function(fit, data, title) {
  nd <- expand.grid(end_wave = levels(data$end_wave), cluster = levels(data$cluster)[1], age_mid_between = 0, age_mid_cwc = 0, sex = levels(data$sex)[1])
  nd$pred <- predict(fit, newdata = nd, type = "response", re.form = NA)
  ggplot(nd, aes(x = end_wave, y = pred)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.001)) +
    labs(title = wrap_title(title, 70), x = "Wave (interval end)", y = "Baseline hazard (marginal)") +
    theme_minimal(base_size = 9) +
    theme(plot.title = element_text(lineheight = 0.95))
}

#4.2 Use the helper function to generate marginal hazard effects by wave plots
wave_plots <- lapply(OUTCOMES, function(out) {
  fit <- fits_primary[[out]]
  dat <- bd_pp_model %>% filter(outcome == out)
  plot_wave_hazard(fit, dat, glue("Baseline (wave) hazard - {out} (link={params$link_primary})"))
})

#4.3 Facet wrap the plots
wrap_plots(wave_plots, ncol = 2)

```

## 5. Perform Calibration (Using Decile Bins) for the Primary Link

Creates decile-bin calibration plots (mean predicted vs observed event rate). This checks probability calibration of the discrete-time models and helps flag miscalculation across the risk spectrum

```{r calibration, warning = FALSE, echo = FALSE, fig.height=11, fig.width=8.5}

## 5. Calibration (decile bins) - primary link ##

#5.1 Calibration helper: connect points and wrap title
plot_cal <- function(df, title) {
  df <- dplyr::arrange(df, pred_mean)
  xmax <- max(df$pred_mean, na.rm = TRUE)
  ymax <- max(df$obs_mean,  na.rm = TRUE)
  lim  <- max(0.05, xmax, ymax)
  ggplot(df, aes(x = pred_mean, y = obs_mean)) +
    geom_point(size = 2) +
    geom_line() +
    geom_abline(slope = 1, intercept = 0, linetype = 2) +
    scale_x_continuous(limits = c(0, lim), labels = scales::percent_format(accuracy = 0.001)) +
    scale_y_continuous(limits = c(0, lim), labels = scales::percent_format(accuracy = 0.001)) +
    coord_equal() +
    labs(title = wrap_title(title, 70),
      x = "Mean predicted hazard",
      y = "Observed event rate") +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(lineheight = 0.95))
}

#5.2 Guard cal_all so filter() never errors when nothing to show
if (!exists("cal_all") || !"model" %in% names(cal_all)) {
  cal_all <- tibble(outcome = character(), model = character(),
    pred_mean = double(), obs_mean = double(), n = integer())
}

#5.3 Execute the helper function for decile calibration on each primary link 
plots_cal <- lapply(OUTCOMES, function(out) {
  df <- cal_all %>% filter(outcome == out, model == params$link_primary)
  ttl <- glue("Calibration by decile - {out} (link={params$link_primary})")
  plot_cal(df, ttl)
})

#5.4 Facet wrap the plots
wrap_plots(plots_cal, ncol = 2)

```

## 6. Generate Inference Tables (Wald vs CR2) Primary vs Sensitivity

We report participant-clustered robust inference as primary (CR2; HC1 secondary). Model-based (Wald) SEs from the pooled logistic model are shown only as a reference, since they assume independent rows and can be anti-conservative when within-person clustering is present

To note, odds ratios are per-interval hazard ratios on the logit scale; cluster main effects represent the average cluster gap across the modeled age range unless a cluster*x*age interaction is present (in which case interpret the interaction at the reported ages)

```{r inference tables, warning = FALSE, echo = FALSE, fig.height=11, fig.width=8.5}

## 6. Inference tables (ORs with 95% CI) ##

#6.1 Create a function to create readable, formatted labels for inference tables
pretty_term <- function(x) {
  x %>%
    stringr::str_replace_all(":", " x ") %>%
    stringr::str_replace("^cluster", "Cluster: ") %>%
    stringr::str_replace("^age_mid_between$", "Age (between-person)") %>%
    stringr::str_replace("^age_mid_cwc$", "Age (within-person)") %>%
    stringr::str_replace("^baseline_status$", "Baseline positive") %>%
    stringr::str_replace("^sex", "Sex: ") %>%
    stringr::str_replace("^end_wave", "Wave: ")
}

#6.2 Generate the wald results tables (filtered) using the pretty_term helper function
tbl_wald <- wald_all %>%
  keep_terms() %>%
  mutate(term = pretty_term(term)) %>%
  dplyr::select(model, outcome, term, OR, CI_lo, CI_hi, p.value)

#6.3 Generate the cr2 results tables (filtered) using the pretty_term helper function
tbl_cr2 <- cr2_all %>%
  keep_terms() %>%
  mutate(term = pretty_term(term)) %>%
  dplyr::select(se_type, model, outcome, term, OR, CI_lo, CI_hi, p.value)

#6.4 Print the formatted wald and cr2 results tables (filtered)
knitr::kable(tbl_wald %>% arrange(outcome, model), digits = 3, caption = "Fixed effects (Wald) Odds ratios with 95% CI (filtered)") %>%
  kableExtra::kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover","condensed","responsive"))
knitr::kable(tbl_cr2 %>% arrange(outcome, model), digits = 3, caption = "Fixed effects (cluster-robust; see se_type for participant vs family; CR2 primary) Odds ratios with 95% CI (filtered)") %>%
  kableExtra::kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover","condensed","responsive"))

```

## 7. Cluster Contrasts at Representative Ages (Primary Link)

Reports `emmeans` pairwise cluster contrasts (odds-ratio scale) at `wave_ref` with within-person age = 0 and baseline/sex/site at reference. This isolates how cluster differences manifest at clinically interpretable age points

```{r cluster contrasts, echo = FALSE, warning = FALSE}

## 7. Cluster contrasts (emmeans) at ages = {params$ages_pred} years (primary link) ##

#7.1 Determine which ages were used for contrasts
used_ages <- if (!is.null(params$ages_pred) && length(params$ages_pred) > 0) params$ages_pred else ages_pred_vec

#7.2 Safely display contrasts table (handle empty or missing columns)
if (exists("emm_all") && is.data.frame(emm_all) && nrow(emm_all) > 0) {
  emm_disp <- emm_all
  if (!"model" %in% names(emm_disp)) emm_disp <- dplyr::mutate(emm_disp, model = params$link_primary)
  if (!"outcome" %in% names(emm_disp)) emm_disp <- dplyr::mutate(emm_disp, outcome = NA_character_)
  knitr::kable(
    emm_disp %>% dplyr::filter(model == params$link_primary) %>% dplyr::arrange(outcome, age_years, contrast),
    digits = 3,
    caption = glue::glue("Cluster contrasts (OR) at ages {paste(used_ages, collapse=', ')} - primary link = {params$link_primary}")) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))
} else {
  message("No emmeans contrasts to display (likely singular fits); skipping table.")
}

#7.3 Safely display wave-specific cluster contrasts (primary link; aligned to wave-median ages)
if (exists("emm_wave_primary_all") && is.data.frame(emm_wave_primary_all) && nrow(emm_wave_primary_all) > 0) {
  knitr::kable(
    emm_wave_primary_all %>% dplyr::arrange(outcome, end_wave, contrast),
    digits = 3,
    caption = glue::glue("Cluster contrasts (OR) by wave at each wave's median age - primary link = {params$link_primary}")
  ) %>% kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))
}

```

## Age Notes & Sensitivity

**Why decompose age?**

-   In longitudinal hazards, naive "age" conflates two distinct things:
    -   Aging within the same youth (time passing)
    -   Older vs younger youths on average (between families/ids)
-   Including age_mid_cwc + age_mid_between separates these. Interpretation:
    -   age_mid_cwc: per-person change in hazard as they age (holding their own mean age fixed)
    -   age_mid_between: difference in hazard for a child whose typical age is higher vs lower than the cohort mean

An optional parsimonious alternative would be to use a single centered age. If we end up needing a simpler reporting model, we could fit: `event ~ cluster * age_mid_gmc + baseline_status + sex + site_factor + end_wave + (1|id) + (1|family)`; and keep the decomposed model as a sensitivity analysis (this is currently not run here though, to keep runtime minimal)

## Final Story Template

In the interest of summarizing the many and complex analyses that were ran in service of answering the "when" of BD onset by risk cluster, the following is an attempt to summarize the most pertinent and meaningful analysis results.

```{r final story template, warning = FALSE, echo = FALSE}

has_cols <- function(df, cols) is.data.frame(df) && nrow(df) > 0 && all(cols %in% names(df))

# Absolute hazards by wave (primary link; reference age profile)
haz_story <- if (exists("hazard_tables_all") &&
                 has_cols(hazard_tables_all, c("outcome","end_wave","cluster","hazard","lo","hi"))) {
  hazard_tables_all %>%
    dplyr::transmute(outcome, end_wave, cluster,
      metric = "hazard_per_interval",
      estimate = hazard, lo = lo, hi = hi)
} else tibble()

# Cumulative risk by wave (primary; reference profile)
cum_story <- if (exists("cumrisk_tables_all") &&
                 has_cols(cumrisk_tables_all, c("outcome","end_wave","cluster","cumrisk","cumrisk_lo","cumrisk_hi"))) {
  cumrisk_tables_all %>%
    dplyr::transmute(outcome, end_wave, cluster,
      metric = "cumulative_risk",
      estimate = cumrisk, lo = cumrisk_lo, hi = cumrisk_hi)
} else tibble()

# Cluster contrasts at ages (primary OR)
contr_or_story <- if (exists("emm_all") &&
                      has_cols(emm_all, c("model","outcome","age_years","contrast","OR","CI_lo","CI_hi","p.value"))) {
  emm_all %>%
    dplyr::filter(model == params$link_primary) %>%
    dplyr::transmute(outcome, age_years, contrast,
      effect_scale = "OR",
      estimate = OR, lo = CI_lo, hi = CI_hi, p.value)
} else tibble()

# Cluster contrasts at ages (cloglog HR)
contr_hr_story <- if (exists("emm_cloglog_all") &&
                      has_cols(emm_cloglog_all, c("outcome","age_years","contrast","HR","CI_lo","CI_hi","p.value"))) {
  emm_cloglog_all %>%
    dplyr::transmute(outcome, age_years, contrast,
      effect_scale = "HR",
      estimate = HR, lo = CI_lo, hi = CI_hi, p.value)
} else tibble()

# Per-year first-onset rates from cloglog (reference profile)
rates_story <- if (exists("rates_by_wave") &&
                   has_cols(rates_by_wave, c("outcome","end_wave","cluster","rate_per_year"))) {
  rates_by_wave %>%
    dplyr::transmute(outcome, end_wave, cluster,
      metric = "rate_per_year",
      estimate = rate_per_year, lo = NA_real_, hi = NA_real_)
} else tibble()

# Combine, guarantee sort keys, and print
story <- dplyr::bind_rows(haz_story, cum_story, contr_or_story, contr_hr_story, rates_story)

# If story exists, print in full
if (nrow(story) > 0) {

  # Ensure the columns used for sorting/printing exist & are consistent types
  if (!"age_years" %in% names(story)) story$age_years <- NA_real_
  if (!"end_wave"  %in% names(story)) story$end_wave  <- NA_character_
  story <- story %>%
    dplyr::mutate(
      end_wave = as.character(end_wave),
      
      # Prefer 'metric' if present; otherwise use 'effect_scale' (OR/HR)
      metric_display = dplyr::coalesce(metric, effect_scale))
  knitr::kable(
    story %>%
      dplyr::arrange(outcome, metric_display, age_years, end_wave) %>%
      dplyr::select(
        outcome,
        metric = metric_display,
        dplyr::any_of(c("end_wave","age_years","cluster","contrast","effect_scale", "estimate","lo","hi","p.value"))),
    digits = 3,
    caption = "Final story: key absolute risks & contrasts") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"))
} else {
  message("No 'final story' rows to display.")
}

```
