---
title: "6. Downstream Predictive-Utility Sensitivity Analysis: Cluster Labels vs MVFS vs Full Baseline Risk Variable Set"
author: "Sam A. Sievertsen"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    df_print: paged
params:
  
  # I/O
  repo: !r Sys.getenv("REPO", unset = here::here())
  data_dir: "data/data_processed/analysis_datasets"
  risk_data_rel: "data/data_processed/risk_variable_data.csv"
  out_dir: "results/main_analysis/2_statistical_analysis/6_risk_vars_vs_cluster_utility"

  # Analysis datasets
  bd_pp_rds: "bd_person_period_k2_z_score.rds"
  bd_pp_csv: "bd_person_period_k2_z_score.csv"
  bd_panel_rds: "bd_panel_k2_z_score.rds"
  bd_panel_csv: "bd_panel_k2_z_score.csv"

  # Outcomes
  outcomes: !r c("bipolar_I", "bipolar_II", "bd_nos", "any_bsd")

  # Model settings
  dth_link: "logit"
  gee_link: "logit"
  do_gee_interaction: true
  gee_corstr: "exchangeable"
  wave_ref_dth: "ses-04A"
  wave_ref_prev: "ses-02A"
  seed: 123

  # Diagnostics / memory control
  run_qic: false
  save_full_gee_objects: true
  progress_log_file: "run_progress.log"

  # Risk variables to compare against cluster labels
  risk_vars: !r c(
    "mh_p_cbcl__dsm__dep_tscore",
    "mh_p_cbcl__dsm__anx_tscore",
    "mh_p_cbcl__synd__attn_tscore",
    "mh_p_cbcl__synd__aggr_tscore",
    "mh_p_gbi_sum",
    "mh_y_upps__nurg_sum",
    "mh_y_upps__purg_sum",
    "le_l_coi__addr1__coi__total__national_zscore",
    "fc_p_nsc__ns_mean",
    "sds_total",
    "family_history_depression",
    "family_history_mania",
    "bullying",
    "nc_y_nihtb__lswmt__uncor_score",
    "nc_y_nihtb__flnkr__uncor_score",
    "nc_y_nihtb__pttcp__uncor_score",
    "ACE_index_sum_score")
    
  # MVFS variables to compare against cluster labels and full risk set
  mvfs_vars: !r c(
    "mh_p_cbcl__dsm__dep_tscore",
    "mh_p_cbcl__synd__aggr_tscore",
    "mh_p_cbcl__synd__attn_tscore",
    "mh_p_cbcl__dsm__anx_tscore",
    "mh_p_gbi_sum",
    "sds_total",
    "family_history_depression")
---

```{r global, include = FALSE, echo = FALSE, warning = FALSE}
knitr::opts_chunk$set(cache = FALSE, message = TRUE, warning = TRUE, results = "markup", comment = "")
```

## 6.1 Set Up

```{r setup, echo = FALSE, warning = FALSE}

## 6.1 Set Up ##

#6.1.1 Load packages
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(readr)
  library(tibble)
  library(stringr)
  library(glue)
  library(forcats)
  library(yardstick)
  library(purrr)
  library(data.table)
  library(tidytext)
  library(ggplot2)
  library(patchwork)
  library(geepack)
  library(emmeans)
  library(clubSandwich)
  library(pROC)
  library(sandwich)
  library(lmtest)
  library(broom)
  library(kableExtra)
  library(janitor)
  library(performance)
  library(here)
  library(jsonlite)
})

#6.1.2 Reproducibility and numeric display
set.seed(params$seed)
options(scipen = 999, digits = 8)
options(encoding = "UTF-8")
knitr::opts_knit$set(encoding = "UTF-8")

#6.1.3 Resolve paths
REPO <- params$repo
DATA_DIR <- file.path(REPO, params$data_dir)
RISK_PATH <- file.path(REPO, params$risk_data_rel)
OUT_DIR <- file.path(REPO, params$out_dir)

dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUT_DIR, "tables"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUT_DIR, "figures"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUT_DIR, "rds"), recursive = TRUE, showWarnings = FALSE)

#6.1.4 Safe writers
write_csv2 <- function(x, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  readr::write_csv(x, path)
}

save_rds2 <- function(x, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  saveRDS(x, path)
}

#6.1.5 Readers with RDS preference
read_obj <- function(rds_path, csv_path) {
  if (file.exists(rds_path)) {
    readRDS(rds_path)
  } else {
    readr::read_csv(csv_path, show_col_types = FALSE)
  }
}

#6.1.6 Progress logging helpers
PROGRESS_LOG <- file.path(OUT_DIR, params$progress_log_file)
if (file.exists(PROGRESS_LOG)) file.remove(PROGRESS_LOG)

log_msg <- function(...) {
  txt <- paste0("[", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "] ", paste0(..., collapse = ""))
  cat(txt, "\n")
  cat(txt, "\n", file = PROGRESS_LOG, append = TRUE)
  flush.console()
}

log_gc <- function(tag = "") {
  gc_txt <- paste(capture.output(gc()), collapse = " | ")
  log_msg(tag, " | gc: ", gc_txt)
}

#6.1.7 Pack GEE objects before saving to reduce memory pressure
pack_gee_for_rds <- function(fit) {
  if (isTRUE(params$save_full_gee_objects)) return(fit)

  list(
    class = class(fit),
    call = fit$call,
    formula = stats::formula(fit),
    coefficients = stats::coef(fit),
    corstr = tryCatch(fit$geese$corstr, error = function(e) NULL),
    scale = tryCatch(fit$geese$gamma, error = function(e) NULL))
}

save_gee_rds2 <- function(fit, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  saveRDS(pack_gee_for_rds(fit), path)
}

#6.1.8 Initial checkpoint
log_msg("6.1 Setup complete")
log_gc("6.1 After setup")

```

## 6.2 Load Baseline Risk Data and Prepare Predictors

```{r risk data, echo = FALSE, warning = FALSE}

## 6.2 Load Risk Data ##

#6.2.1 Read baseline risk-variable file
if (!file.exists(RISK_PATH)) stop(glue("Risk data file not found: {RISK_PATH}"))
risk_df_raw <- readr::read_csv(RISK_PATH, show_col_types = FALSE)
log_msg("6.2 Loaded risk_variable_data.csv: N=", nrow(risk_df_raw), ", p=", ncol(risk_df_raw))

#6.2.2 Confirm participant_id exists
if (!("participant_id" %in% names(risk_df_raw))) {
  stop("participant_id not found in risk_variable_data.csv")
}

#6.2.3 Confirm all requested full-risk variables exist
missing_risk_vars <- setdiff(params$risk_vars, names(risk_df_raw))
if (length(missing_risk_vars) > 0) {
  stop(glue("Missing full-risk variables in risk_variable_data.csv: {paste(missing_risk_vars, collapse = ', ')}"))
}

#6.2.4 Confirm all requested MVFS variables exist
missing_mvfs_vars <- setdiff(params$mvfs_vars, names(risk_df_raw))
if (length(missing_mvfs_vars) > 0) {
  stop(glue("Missing MVFS variables in risk_variable_data.csv: {paste(missing_mvfs_vars, collapse = ', ')}"))
}

#6.2.5 Restrict to ID + requested predictors
risk_df <- risk_df_raw %>%
  dplyr::select(participant_id, dplyr::all_of(params$risk_vars)) %>%
  dplyr::mutate(participant_id = as.character(participant_id))

#6.2.6 Force known categorical variables to factor
force_known_categoricals <- function(df) {
  df <- as.data.frame(df)

  known_cat <- intersect(
    c("family_history_depression", "family_history_mania", "bullying"),
    names(df))

  for (nm in known_cat) {
    df[[nm]] <- factor(df[[nm]])
  }

  df
}

#6.2.7 Coerce remaining variable classes conservatively
coerce_types <- function(df) {
  df <- as.data.frame(df)

  for (nm in names(df)) {
    x <- df[[nm]]

    #6.2.7.1 Keep numeric as numeric
    if (is.numeric(x)) next

    #6.2.7.2 Logical to factor
    if (is.logical(x)) {
      df[[nm]] <- factor(x, levels = c(FALSE, TRUE))
      next
    }

    #6.2.7.3 Character to factor
    if (is.character(x)) {
      df[[nm]] <- factor(x)
      next
    }

    #6.2.7.4 Small-cardinality integers to factor
    if (is.integer(x)) {
      u <- unique(x[!is.na(x)])
      if (length(u) <= 10) {
        df[[nm]] <- factor(x)
      } else {
        df[[nm]] <- as.numeric(x)
      }
      next
    }

    #6.2.7.5 Fallback
    if (!is.factor(x)) df[[nm]] <- factor(x)
  }

  df
}

#6.2.8 Z-score numeric columns so the raw-variable model operates on comparable scales
zscore_numeric <- function(df) {
  df <- as.data.frame(df)
  is_num <- vapply(df, is.numeric, logical(1))

  for (nm in names(df)[is_num]) {
    x <- df[[nm]]
    mu <- mean(x, na.rm = TRUE)
    sdv <- stats::sd(x, na.rm = TRUE)

    if (is.finite(sdv) && sdv > 0) {
      df[[nm]] <- (x - mu) / sdv
    } else {
      df[[nm]] <- 0
    }
  }

  df
}

#6.2.9 Build processed predictor frame once
risk_x <- risk_df %>%
  dplyr::select(-participant_id) %>%
  force_known_categoricals() %>%
  coerce_types() %>%
  zscore_numeric()

risk_proc <- dplyr::bind_cols(
  tibble::tibble(participant_id = risk_df$participant_id),
  as_tibble(risk_x))

#6.2.10 Build formula-ready term string for the full 17-variable models
risk_term_string <- paste(params$risk_vars, collapse = " + ")

#6.2.11 Build formula-ready term string for the MVFS models
mvfs_term_string <- paste(params$mvfs_vars, collapse = " + ")

#6.2.12 Quick audit table
risk_type_tbl <- tibble::tibble(
  variable = names(risk_x),
  class = vapply(risk_x, function(x) class(x)[1], character(1)))

write_csv2(risk_type_tbl, file.path(OUT_DIR, "tables", "risk_variable_classes_used.csv"))

knitr::kable(
  risk_type_tbl,
  caption = "Baseline full-risk variables used in the cluster-vs-MVFS-vs-full utility sensitivity analysis") %>%
  kableExtra::kable_styling(full_width = FALSE)

#6.2.13 MVFS audit table
mvfs_type_tbl <- risk_type_tbl %>%
  dplyr::filter(variable %in% params$mvfs_vars)

write_csv2(mvfs_type_tbl, file.path(OUT_DIR, "tables", "mvfs_variable_classes_used.csv"))

knitr::kable(
  mvfs_type_tbl,
  caption = "MVFS variables used in the cluster-vs-MVFS-vs-full utility sensitivity analysis") %>%
  kableExtra::kable_styling(full_width = FALSE)

#6.2.14 Free large intermediate objects
rm(risk_df_raw, risk_df, risk_x)
invisible(gc())

log_msg("6.2 Processed baseline risk predictors: N=", nrow(risk_proc), ", p=", ncol(risk_proc))
log_gc("6.2 After processing risk predictors")

```

## 6.3 Load DTH Person-Period Data and Build Analysis Frame

```{r dth data, echo = FALSE, warning = FALSE}

## 6.3 Load DTH Data ##

#6.3.1 Read person-period dataset
bd_pp <- read_obj(
  file.path(DATA_DIR, params$bd_pp_rds),
  file.path(DATA_DIR, params$bd_pp_csv)) %>%
  dplyr::mutate(
    participant_id = as.character(participant_id),
    family_id = factor(as.character(family_id)),
    site_factor = factor(site_factor),
    cluster = factor(cluster),
    start_wave = factor(start_wave),
    end_wave = factor(end_wave),
    outcome = factor(outcome),
    sex = factor(sex))

#6.3.2 Keep modeled BD outcomes only
bd_pp <- bd_pp %>%
  dplyr::filter(outcome %in% params$outcomes)

#6.3.3 Mirror primary DTH model-ready filtering
bd_pp_model <- bd_pp %>%
  dplyr::filter(event %in% c(0L, 1L)) %>%
  dplyr::filter(!is.na(cluster)) %>%
  dplyr::filter(!is.na(age_mid), !is.na(sex), !is.na(site_factor), !is.na(family_id))

#6.3.4 Set reference levels to match primary DTH script
bd_pp_model <- bd_pp_model %>%
  dplyr::mutate(
    cluster = forcats::fct_relevel(cluster, "C1"),
    end_wave = forcats::fct_relevel(end_wave, "ses-02A", "ses-04A", "ses-06A"),
    sex = {
      lv <- levels(sex)
      ref <- if ("Female" %in% lv) "Female" else lv[1]
      forcats::fct_relevel(sex, ref)
    },
    site_factor = {
      lv <- levels(site_factor)
      ref <- if ("site01" %in% lv) "site01" else lv[1]
      forcats::fct_relevel(site_factor, ref)
    })

#6.3.5 Compute interval length variables used in the main DTH script
wave_year <- function(w) {
  as.numeric(stringr::str_extract(as.character(w), "\\d+"))
}

bd_pp_model <- bd_pp_model %>%
  dplyr::mutate(
    dt_years = pmax(wave_year(end_wave) - wave_year(start_wave), 1),
    log_dt = log(dt_years)) %>%
  dplyr::filter(is.finite(dt_years), is.finite(log_dt))

#6.3.6 Reproduce baseline-negative risk set for any BSD
baseline_any_bsd_ids <- bd_pp_model %>%
  dplyr::filter(outcome == "any_bsd") %>%
  dplyr::group_by(participant_id) %>%
  dplyr::summarise(
    all_na = all(is.na(baseline_status)),
    baseline_any_bsd = dplyr::if_else(
      all_na,
      NA_integer_,
      as.integer(any(baseline_status == 1L, na.rm = TRUE))),
    .groups = "drop")

bd_pp_model <- bd_pp_model %>%
  dplyr::left_join(
    baseline_any_bsd_ids %>% dplyr::select(participant_id, baseline_any_bsd),
    by = "participant_id") %>%
  dplyr::filter(baseline_any_bsd == 0L)

#6.3.7 Censor after first onset within outcome x participant
if ("interval_index" %in% names(bd_pp_model)) {
  bd_pp_model <- bd_pp_model %>%
    dplyr::arrange(outcome, participant_id, interval_index)
} else {
  bd_pp_model <- bd_pp_model %>%
    dplyr::arrange(outcome, participant_id, as.integer(end_wave))
}

bd_pp_model <- bd_pp_model %>%
  dplyr::group_by(outcome, participant_id) %>%
  dplyr::mutate(after_onset = dplyr::lag(cummax(event == 1L), default = FALSE)) %>%
  dplyr::filter(!after_onset) %>%
  dplyr::ungroup() %>%
  dplyr::select(-after_onset)

#6.3.8 Join raw baseline risk variables by participant_id
bd_pp_model <- bd_pp_model %>%
  dplyr::left_join(risk_proc, by = "participant_id")

#6.3.9 Require complete data on the full 17-variable predictor block so that cluster, MVFS, and full-risk models are compared on the exact same rows
n_before_dth_cc <- nrow(bd_pp_model)

bd_pp_model <- bd_pp_model %>%
  tidyr::drop_na(dplyr::all_of(params$risk_vars))

log_msg(
  "6.3 DTH complete-case filter on raw predictors: rows ",
  n_before_dth_cc, " -> ", nrow(bd_pp_model),
  "; unique IDs = ", dplyr::n_distinct(bd_pp_model$participant_id))

log_gc("6.3 After DTH frame assembly")

#6.3.10 Sanity check table
dth_size_tbl <- tibble::tibble(
  Dataset = "DTH sensitivity analysis frame",
  Rows = nrow(bd_pp_model),
  Unique_IDs = dplyr::n_distinct(bd_pp_model$participant_id))

knitr::kable(dth_size_tbl, caption = "DTH sensitivity-analysis dataset size") %>%
  kableExtra::kable_styling(full_width = FALSE)

#6.3.11 Free source person-period object
rm(bd_pp)
invisible(gc())

```

## 6.4 Load Prevalence Panel Data and Build Analysis Frame

```{r prevalence data, echo = FALSE, warning = FALSE}

## 6.4 Load Prevalence Data ##

#6.4.1 Read panel dataset
bd_panel <- read_obj(
  file.path(DATA_DIR, params$bd_panel_rds),
  file.path(DATA_DIR, params$bd_panel_csv)) %>%
  dplyr::mutate(
    participant_id = as.factor(participant_id),
    family_id = as.factor(family_id),
    site_factor = as.factor(site_factor),
    cluster = factor(cluster, levels = c("C1", "C2")),
    wave = factor(wave, levels = c("ses-02A", "ses-04A", "ses-06A")),
    sex = factor(sex, levels = c("Female", "Male")))

#6.4.2 Confirm primary prevalence variables exist
req_prev <- c("outcome", "status", "wave", "cluster", "sex", "participant_id", "age_wave_gmc")
missing_prev <- setdiff(req_prev, names(bd_panel))
if (length(missing_prev) > 0) {
  stop(glue("Prevalence panel data missing required variables: {paste(missing_prev, collapse = ', ')}"))
}

#6.4.3 Keep outcomes of interest and join raw predictors
bd_panel_model <- bd_panel %>%
  dplyr::filter(outcome %in% params$outcomes) %>%
  dplyr::left_join(risk_proc, by = "participant_id")

#6.4.4 Require complete data on the full 17-variable predictor block so that cluster, MVFS, and full-risk models are compared on the exact same rows
n_before_prev_cc <- nrow(bd_panel_model)

bd_panel_model <- bd_panel_model %>%
  tidyr::drop_na(dplyr::all_of(params$risk_vars))

log_msg(
  "6.4 Prevalence complete-case filter on raw predictors: rows ",
  n_before_prev_cc, " -> ", nrow(bd_panel_model),
  "; unique IDs = ", dplyr::n_distinct(bd_panel_model$participant_id))

log_gc("6.4 After prevalence frame assembly")

#6.4.5 Ensure baseline prevalence covariate exists
if (!("baseline_status_bd" %in% names(bd_panel_model))) {
  stop("baseline_status_bd not found in the panel dataset; this sensitivity script currently mirrors the primary GEE prevalence adjustment set.")
}

#6.4.6 Reapply the same factor structure used in the main prevalence script
bd_panel_model <- bd_panel_model %>%
  dplyr::mutate(
    participant_id = as.factor(participant_id),
    family_id = as.factor(family_id),
    site_factor = as.factor(site_factor),
    cluster = factor(cluster, levels = c("C1", "C2")),
    wave = factor(wave, levels = c("ses-02A", "ses-04A", "ses-06A")),
    sex = factor(sex, levels = c("Female", "Male")),
    baseline_status_bd = as.factor(baseline_status_bd)) %>%
  dplyr::arrange(outcome, participant_id, wave)

log_msg("6.4 Reapplied main-script factor structure and sorted prevalence rows")
log_gc("6.4 After prevalence factor reset")

#6.4.6.1 Audit binary coding used for prevalence outcome metrics
status_audit_tbl <- tibble::tibble(
  variable = c("status", "baseline_status_bd"),
  class = c(class(bd_panel_model$status)[1], class(bd_panel_model$baseline_status_bd)[1]),
  example_values = c(
    paste(utils::head(unique(as.character(bd_panel_model$status)), 10), collapse = ", "),
    paste(utils::head(unique(as.character(bd_panel_model$baseline_status_bd)), 10), collapse = ", ")))

write_csv2(status_audit_tbl, file.path(OUT_DIR, "tables", "prevalence_status_audit.csv"))

knitr::kable(
  status_audit_tbl,
  caption = "Audit of binary outcome/covariate coding used in prevalence performance metrics") %>%
  kableExtra::kable_styling(full_width = FALSE)

#6.4.7 Sanity check table
prev_size_tbl <- tibble::tibble(
  Dataset = "Prevalence sensitivity analysis frame",
  Rows = nrow(bd_panel_model),
  Unique_IDs = dplyr::n_distinct(bd_panel_model$participant_id))

knitr::kable(prev_size_tbl, caption = "Prevalence sensitivity-analysis dataset size") %>%
  kableExtra::kable_styling(full_width = FALSE)

#6.4.8 Free source panel object
rm(bd_panel)
invisible(gc())

```

## 6.5 Helpers for Model Fitting and Comparison

```{r helpers, echo = FALSE, warning = FALSE}

## 6.5 Helper Functions ##

#6.5.1 Robust SE (clustered by participant) helpers
#6.5.1.1 Cluster-robust CR2 inference for glm objects
tidy_glm_cr2 <- function(fit, cluster_var) {

  #6.5.1.1.1 Safe helper to get the first matching column name or NA
  pick_first <- function(x) {
    if (length(x) > 0) x[1] else NA_character_
  }

  mf <- model.frame(fit)
  rn <- rownames(mf)
  idx <- suppressWarnings(as.integer(rn))
  if (anyNA(idx)) idx <- seq_len(nrow(mf))
  cluster_aligned <- cluster_var[idx]

  V <- clubSandwich::vcovCR(fit, cluster = cluster_aligned, type = "CR2")
  ct <- clubSandwich::coef_test(fit, vcov = V, test = "Satterthwaite")
  ctd <- as.data.frame(ct)

  #6.5.1.1.2 Ensure a term column exists
  if (!("term" %in% names(ctd))) {
    if ("Coef" %in% names(ctd)) {
      ctd$term <- as.character(ctd$Coef)
    } else if (!is.null(rownames(ctd)) && all(nzchar(rownames(ctd)))) {
      ctd$term <- rownames(ctd)
    } else {
      ctd$term <- paste0("b", seq_len(nrow(ctd)))
    }
  }

  ct2 <- tibble::as_tibble(ctd)

  pcol <- pick_first(intersect(c("p_Satt", "p_z", "p_Nd", "p_t", "p"), names(ct2)))
  dfcol <- pick_first(intersect(c("df_Satt", "df"), names(ct2)))
  betacol <- pick_first(intersect(c("beta", "Beta", "Estimate"), names(ct2)))
  secol <- pick_first(intersect(c("SE", "Std. Error", "Std.Error"), names(ct2)))
  cilocol <- pick_first(intersect(c("CI_L", "CI_L_Satt", "lower", "Lower"), names(ct2)))
  cihicol <- pick_first(intersect(c("CI_U", "CI_U_Satt", "upper", "Upper"), names(ct2)))

  #6.5.1.1.3 Extract core statistics
  est <- if (!is.na(betacol)) as.numeric(ct2[[betacol]]) else rep(NA_real_, nrow(ct2))
  se <- if (!is.na(secol)) as.numeric(ct2[[secol]]) else rep(NA_real_, nrow(ct2))
  dfv <- if (!is.na(dfcol)) as.numeric(ct2[[dfcol]]) else rep(NA_real_, nrow(ct2))

  #6.5.1.1.4 Use reported CR2 CI columns if available; otherwise derive from estimate/SE
  crit <- ifelse(
    is.finite(dfv) & dfv > 0,
    stats::qt(0.975, df = pmax(dfv, 1)),
    stats::qnorm(0.975))

  ci_lo_log <- if (!is.na(cilocol)) {
    as.numeric(ct2[[cilocol]])
  } else {
    est - crit * se
  }

  ci_hi_log <- if (!is.na(cihicol)) {
    as.numeric(ct2[[cihicol]])
  } else {
    est + crit * se
  }

  #6.5.1.1.5 P-values: use reported CR2 p-values if available; otherwise Wald fallback
  pv <- if (!is.na(pcol)) {
    as.numeric(ct2[[pcol]])
  } else {
    z <- est / se
    2 * stats::pnorm(-abs(z))
  }

  tibble::tibble(
    term = ct2$term,
    estimate = est,
    SE = se,
    df = dfv,
    OR = exp(est),
    CI_lo = exp(ci_lo_log),
    CI_hi = exp(ci_hi_log),
    p.value = pv)
}

#6.5.1.2 Cluster-robust HC1 inference for glm objects
tidy_glm_hc1 <- function(fit, cluster_var) {
  mf <- model.frame(fit)
  rn <- rownames(mf)
  idx <- suppressWarnings(as.integer(rn))
  if (anyNA(idx)) idx <- seq_len(nrow(mf))
  cluster_aligned <- cluster_var[idx]

  V <- sandwich::vcovCL(fit, cluster = cluster_aligned, type = "HC1")
  b <- stats::coef(fit)

  #6.5.1.2.1 Align vcov to coefficient order
  V <- V[names(b), names(b), drop = FALSE]

  #6.5.1.2.2 Extract robust SEs and Wald z-tests
  se <- sqrt(diag(V))
  z <- b / se
  p <- 2 * stats::pnorm(-abs(z))

  tibble::tibble(
    term = names(b),
    estimate = unname(b),
    SE = unname(se),
    df = NA_real_,
    OR = exp(estimate),
    CI_lo = exp(estimate - 1.96 * SE),
    CI_hi = exp(estimate + 1.96 * SE),
    p.value = unname(p))
}

#6.5.2 Robust coefficient extraction for GEE
tidy_gee_robust <- function(fit) {
  sm <- summary(fit)$coefficients
  sm <- as.data.frame(sm)
  sm$term <- rownames(sm)

  est_col <- intersect(c("Estimate"), names(sm))[1]
  se_col <- intersect(c("Robust S.E.", "Std.err", "Std.error", "san.se"), names(sm))[1]
  w_col <- intersect(c("Wald", "Wald Statistic"), names(sm))[1]
  p_col <- intersect(c("Pr(>|W|)", "Pr(>|z|)", "p"), names(sm))[1]

  sm %>%
    tibble::as_tibble() %>%
    dplyr::transmute(
      term = term,
      estimate = .data[[est_col]],
      SE = .data[[se_col]],
      OR = exp(estimate),
      CI_lo = exp(estimate - 1.96 * SE),
      CI_hi = exp(estimate + 1.96 * SE),
      Wald = if (!is.na(w_col)) .data[[w_col]] else NA_real_,
      p.value = if (!is.na(p_col)) .data[[p_col]] else NA_real_)
}

#6.5.3 Robust CI column extractor for emmeans
extract_emm_ci_cols <- function(tb) {
  nm <- names(tb)
  if (all(c("lower.CL", "upper.CL") %in% nm)) return(list(low = tb$lower.CL, high = tb$upper.CL))
  if (all(c("asymp.LCL", "asymp.UCL") %in% nm)) return(list(low = tb$asymp.LCL, high = tb$asymp.UCL))
  if (all(c("LCL", "UCL") %in% nm)) return(list(low = tb$LCL, high = tb$UCL))
  list(low = rep(NA_real_, nrow(tb)), high = rep(NA_real_, nrow(tb)))
}

#6.5.4 Cluster C2 vs C1 odds ratio by wave from emmeans
or_by_wave_emm <- function(fit, outcome, model_label) {
  if (inherits(fit, "try-error")) return(NULL)

  emm <- try(emmeans::emmeans(fit, ~ cluster | wave, type = "link"), silent = TRUE)
  if (inherits(emm, "try-error")) return(NULL)

  sm <- try(
    emmeans::contrast(emm, method = "revpairwise") %>% summary(infer = TRUE),
    silent = TRUE)
  if (inherits(sm, "try-error")) return(NULL)

  tb <- tibble::as_tibble(sm)
  ci <- extract_emm_ci_cols(tb)

  p_col <- if ("p.value" %in% names(tb)) "p.value" else NA_character_

  tibble::tibble(
    outcome = outcome,
    model = model_label,
    wave = as.character(tb$wave),
    OR = exp(tb$estimate),
    CI_lo = exp(ci$low),
    CI_hi = exp(ci$high),
    p.value = if (!is.na(p_col)) as.numeric(tb[[p_col]]) else NA_real_)
}

#6.5.5 Likelihood-style pseudo-R2 for glm. Fit objects created through wrappers may not support update(fit, . ~ 1) cleanly because the original data symbol is not recoverable. Refit the null model directly from model.frame(fit) instead
pseudo_r2_mcfadden <- function(fit) {
  mf <- model.frame(fit)

  #6.5.5.1 First column of model.frame(fit) is the response
  resp <- names(mf)[1]

  #6.5.5.2 Refit intercept-only null model on the exact same analytic rows
  null_form <- stats::as.formula(paste(resp, "~ 1"))
  fit_null <- glm(
    formula = null_form,
    data = mf,
    family = stats::family(fit))

  ll_full <- as.numeric(logLik(fit))
  ll_null <- as.numeric(logLik(fit_null))

  1 - (ll_full / ll_null)
}

#6.5.5.3 Strict binary coercion helper
coerce_binary01 <- function(x, var_name = "outcome") {

  #6.5.5.3.1 Logical -> 0/1
  if (is.logical(x)) {
    return(as.integer(x))
  }

  #6.5.5.3.2 Numeric/integer already coded 0/1
  if (is.numeric(x) || is.integer(x)) {
    x_num <- as.numeric(x)

    if (all(stats::na.omit(x_num) %in% c(0, 1))) {
      return(x_num)
    }

    if (all(stats::na.omit(x_num) %in% c(1, 2))) {
      return(x_num - 1)
    }
  }

  #6.5.5.3.3 Factor/character handling
  x_chr <- as.character(x)

  if (all(stats::na.omit(x_chr) %in% c("0", "1"))) {
    return(as.numeric(x_chr))
  }

  if (all(stats::na.omit(x_chr) %in% c("FALSE", "TRUE"))) {
    return(as.integer(x_chr == "TRUE"))
  }

  #6.5.5.3.4 Last-resort two-level factor mapping
  x_fac <- factor(x)
  if (nlevels(x_fac) == 2) {
    return(as.integer(x_fac == levels(x_fac)[2]))
  }

  stop(glue::glue(
    "Could not coerce {var_name} to binary 0/1. ",
    "Class = {class(x)[1]}; levels/values = {paste(utils::head(unique(as.character(x)), 10), collapse = ', ')}"))
}

#6.5.5.4 AUC helper with explicit positive-class handling
auc_binary <- function(obs, pr) {
  y <- coerce_binary01(obs, var_name = "obs")
  p <- as.numeric(pr)

  ok <- is.finite(y) & is.finite(p)

  if (sum(ok) == 0 || length(unique(y[ok])) < 2) {
    return(NA_real_)
  }

  truth_fac <- factor(y[ok], levels = c(0, 1))

  as.numeric(
    yardstick::roc_auc_vec(
      truth = truth_fac,
      estimate = p[ok],
      event_level = "second"))
}

#6.5.5.5 Log-loss helper
log_loss_binary <- function(obs, pr) {
  y <- coerce_binary01(obs, var_name = "obs")
  p <- pmin(pmax(as.numeric(pr), 1e-8), 1 - 1e-8)

  ok <- is.finite(y) & is.finite(p)

  if (sum(ok) == 0) {
    return(NA_real_)
  }

  -mean(y[ok] * log(p[ok]) + (1 - y[ok]) * log(1 - p[ok]))
}

#6.5.5.6 Readable outcome labels
label_outcome <- function(x) {
  dplyr::recode(
    as.character(x),
    "bipolar_I" = "BD-I",
    "bipolar_II" = "BD-II",
    "bd_nos" = "BD-NOS",
    "any_bsd" = "Any BSD",
    .default = as.character(x))
}

#6.5.5.7 Readable predictor labels for tables/figures
risk_label_lookup <- c(
  "mh_p_cbcl__dsm__dep_tscore" = "CBCL Depression",
  "mh_p_cbcl__dsm__anx_tscore" = "CBCL Anxiety",
  "mh_p_cbcl__synd__attn_tscore" = "CBCL Attention Problems",
  "mh_p_cbcl__synd__aggr_tscore" = "CBCL Aggression Problems",
  "mh_p_gbi_sum" = "GBI Mania",
  "mh_y_upps__nurg_sum" = "UPPS Negative Urgency",
  "mh_y_upps__purg_sum" = "UPPS Positive Urgency",
  "le_l_coi__addr1__coi__total__national_zscore" = "Child Opportunity Index",
  "fc_p_nsc__ns_mean" = "Neighborhood Safety",
  "sds_total" = "Sleep Disturbance",
  "family_history_depression" = "Family History of Depression",
  "family_history_mania" = "Family History of Mania",
  "bullying" = "Bullying",
  "nc_y_nihtb__lswmt__uncor_score" = "List Sorting Working Memory",
  "nc_y_nihtb__flnkr__uncor_score" = "Flanker",
  "nc_y_nihtb__pttcp__uncor_score" = "Pattern Comparison",
  "ACE_index_sum_score" = "Adverse Childhood Experiences")

label_risk_term <- function(term) {
  out <- as.character(term)

  for (nm in names(risk_label_lookup)) {
    out <- ifelse(
      startsWith(out, nm),
      risk_label_lookup[[nm]],
      out)
  }

  out
}

#6.5.5.8 Identify nuisance adjustment terms for prevalence coefficient displays
is_nuisance_prev_term <- function(term) {
  term <- as.character(term)

  stringr::str_detect(
    term,
    "^(\\(Intercept\\)|wave($|\\.)|Wave($|\\.)|waveses-02A|waveses-04A|waveses-06A|sex|baseline_status_bd|age_wave_gmc$)"
  ) | is.na(term)
}

#6.5.6 Brier score helper
brier_score <- function(obs, pr) {
  y <- coerce_binary01(obs, var_name = "obs")
  p <- as.numeric(pr)

  ok <- is.finite(y) & is.finite(p)

  if (sum(ok) == 0) {
    return(NA_real_)
  }

  mean((y[ok] - p[ok])^2)
}

#6.5.7 DTH formulas
build_dth_formula_cluster <- function() {
  stats::as.formula("event ~ cluster + sex + end_wave + age_mid_between")
}

build_dth_formula_raw <- function() {
  stats::as.formula(glue("event ~ sex + end_wave + age_mid_between + {risk_term_string}"))
}

build_dth_formula_mvfs <- function() {
  stats::as.formula(glue("event ~ sex + end_wave + age_mid_between + {mvfs_term_string}"))
}

#6.5.8 Prevalence formulas
build_prev_formula_cluster <- function() {
  stats::as.formula("status ~ cluster + wave + sex + baseline_status_bd + age_wave_gmc")
}

build_prev_formula_cluster_int <- function() {
  stats::as.formula("status ~ cluster * wave + sex + baseline_status_bd + age_wave_gmc")
}

build_prev_formula_mvfs <- function() {
  stats::as.formula(glue("status ~ wave + sex + baseline_status_bd + age_wave_gmc + {mvfs_term_string}"))
}

build_prev_formula_raw <- function() {
  stats::as.formula(glue("status ~ wave + sex + baseline_status_bd + age_wave_gmc + {risk_term_string}"))
}

#6.5.9 DTH model fitter
fit_dth_glm <- function(formula, data) {
  glm(
    formula,
    data = data,
    family = binomial(link = params$dth_link),
    model = TRUE,
    x = FALSE,
    y = FALSE)
}

#6.5.10 Prevalence model fitter
#6.5.10.1 Mirror the main prevalence script exactly; do not pass waves= or custom geese.control here
fit_prev_gee <- function(formula, data) {
  geepack::geeglm(
    formula,
    id = participant_id,
    data = data,
    family = binomial(link = params$gee_link),
    corstr = params$gee_corstr,
    std.err = "san.se")
}

#6.5.11 Identify all coefficient names belonging to a variable block
find_block_coef_names <- function(coef_names, vars) {
  out <- unique(unlist(lapply(vars, function(v) {
    coef_names[startsWith(coef_names, v)]
  })))

  setdiff(out, "(Intercept)")
}

#6.5.12 Omnibus Wald chi-square for the raw predictor block in glm using HC1
glm_block_test_hc1 <- function(fit, vars, cluster_var) {
  mf <- model.frame(fit)
  rn <- rownames(mf)
  idx <- suppressWarnings(as.integer(rn))
  if (anyNA(idx)) idx <- seq_len(nrow(mf))
  cluster_aligned <- cluster_var[idx]

  keep <- find_block_coef_names(names(stats::coef(fit)), vars)

  if (length(keep) == 0) {
    return(tibble::tibble(
      chisq = NA_real_,
      df = NA_real_,
      p.value = NA_real_,
      n_terms = 0L))
  }

  V <- sandwich::vcovCL(fit, cluster = cluster_aligned, type = "HC1")
  V <- V[keep, keep, drop = FALSE]
  b <- stats::coef(fit)[keep]

  stat <- tryCatch(
    as.numeric(t(b) %*% solve(V) %*% b),
    error = function(e) NA_real_)

  df <- length(keep)
  p <- if (is.finite(stat)) stats::pchisq(stat, df = df, lower.tail = FALSE) else NA_real_

  tibble::tibble(
    chisq = stat,
    df = df,
    p.value = p,
    n_terms = df)
}

#6.5.13 Omnibus Wald test for the raw predictor block in GEE
gee_block_test <- function(fit, vars) {
  sm <- summary(fit)$coefficients
  rn <- rownames(sm)
  keep <- find_block_coef_names(rn, vars)

  if (length(keep) == 0) {
    return(tibble::tibble(
      chisq = NA_real_,
      df = NA_real_,
      p.value = NA_real_,
      n_terms = 0L))
  }

  b <- stats::coef(fit)[keep]
  V <- vcov(fit)[keep, keep, drop = FALSE]

  stat <- tryCatch(
    as.numeric(t(b) %*% solve(V) %*% b),
    error = function(e) NA_real_)

  df <- length(keep)
  p <- if (is.finite(stat)) stats::pchisq(stat, df = df, lower.tail = FALSE) else NA_real_

  tibble::tibble(
    chisq = stat,
    df = df,
    p.value = p,
    n_terms = df)
}

#6.5.14 Predicted-probability helper for cluster models
pred_cluster_by_wave <- function(fit, data, wave_var, age_var, baseline_var = NULL) {
  ref <- expand.grid(
    cluster = levels(data$cluster),
    wave_tmp = levels(data[[wave_var]]),
    sex = levels(data$sex)[1],
    stringsAsFactors = FALSE)

  names(ref)[names(ref) == "wave_tmp"] <- wave_var
  ref[[wave_var]] <- factor(ref[[wave_var]], levels = levels(data[[wave_var]]))
  ref[[age_var]] <- 0

  if (!is.null(baseline_var)) {
    ref[[baseline_var]] <- levels(data[[baseline_var]])[1]
    ref[[baseline_var]] <- factor(ref[[baseline_var]], levels = levels(data[[baseline_var]]))
  }

  ref$pred <- as.numeric(predict(fit, newdata = ref, type = "response"))
  ref
}

```

## 6.6 Primary DTH Sensitivity Analysis: Cluster Labels vs Raw Predictors

```{r dth models, echo = FALSE, warning = FALSE}

## 6.6 DTH Sensitivity Analysis ##

dth_results <- list()
dth_coef_tbl <- list()
dth_pred_tbl <- list()

for (out in params$outcomes) {

  log_msg("6.6 Starting DTH outcome: ", out)

  #6.6.1 Subset outcome-specific data
  dat_o <- bd_pp_model %>%
    dplyr::filter(outcome == out) %>%
    droplevels()

  if (nrow(dat_o) == 0) {
    log_msg("6.6 ", out, ": no rows available after filtering; skipping")
    next
  }

  #6.6.2 Create minimal model-specific datasets
  dat_o_cluster <- dat_o %>%
  dplyr::select(participant_id, event, cluster, sex, end_wave, age_mid_between)

  dat_o_mvfs <- dat_o %>%
  dplyr::select(participant_id, event, sex, end_wave, age_mid_between, dplyr::all_of(params$mvfs_vars))

  dat_o_raw <- dat_o %>%
  dplyr::select(participant_id, event, sex, end_wave, age_mid_between, dplyr::all_of(params$risk_vars))

  log_msg(
    "6.6 ", out,
    ": rows=", nrow(dat_o_cluster),
    ", ids=", dplyr::n_distinct(dat_o_cluster$participant_id),
    ", events=", sum(dat_o_cluster$event == 1L, na.rm = TRUE))
  
  log_gc(paste0("6.6 ", out, " before fits"))

  #6.6.3 Fit cluster-label model
  log_msg("6.6 ", out, ": fitting cluster GLM")
  fit_cluster <- tryCatch(
    fit_dth_glm(build_dth_formula_cluster(), dat_o_cluster),
    error = function(e) {
      log_msg("6.6 ", out, ": cluster GLM FAILED - ", conditionMessage(e))
      stop(e)
    })
  
  log_msg("6.6 ", out, ": cluster GLM fit complete")
  
  #6.6.4 Fit MVFS-variable model
  log_msg("6.6 ", out, ": fitting MVFS GLM")
  fit_mvfs <- tryCatch(
    fit_dth_glm(build_dth_formula_mvfs(), dat_o_mvfs),
    error = function(e) {
      log_msg("6.6 ", out, ": MVFS GLM FAILED - ", conditionMessage(e))
      stop(e)
    })

  log_msg("6.6 ", out, ": MVFS GLM fit complete")

  #6.6.5 Fit raw-variable model
  log_msg("6.6 ", out, ": fitting raw GLM")
  fit_raw <- tryCatch(
    fit_dth_glm(build_dth_formula_raw(), dat_o_raw),
    error = function(e) {
      log_msg("6.6 ", out, ": raw GLM FAILED - ", conditionMessage(e))
      stop(e)
    })
  
  log_msg("6.6 ", out, ": raw GLM fit complete")
  log_gc(paste0("6.6 ", out, " after GLM fits"))

 #6.6.6 Robust inference table for cluster-label model. Use CR2 to match the primary DTH inference standard, and fall back to HC1 only if CR2 fails unexpectedly
  log_msg("6.6 ", out, ": extracting cluster-model robust coefficients (CR2 preferred)")
  
  coef_cluster <- tryCatch(
    tidy_glm_cr2(fit_cluster, cluster_var = dat_o_cluster$participant_id) %>%
      dplyr::mutate(
        outcome = out,
        model = "cluster_model",
        se_type = "CR2",
        .before = 1),
    error = function(e) {
      log_msg("6.6 ", out, ": CR2 coefficient extraction FAILED - ",
              conditionMessage(e), "; falling back to HC1")
  
      tidy_glm_hc1(fit_cluster, cluster_var = dat_o_cluster$participant_id) %>%
        dplyr::mutate(
          outcome = out,
          model = "cluster_model",
          se_type = "HC1",
          .before = 1)
    })
  
  
  dth_coef_tbl[[paste0(out, "_cluster")]] <- coef_cluster
  log_msg("6.6 ", out, ": cluster-model robust coefficients extracted")
  
    #6.6.6.1 Robust coefficient table for MVFS model (HC1)
  coef_mvfs <- tidy_glm_hc1(fit_mvfs, cluster_var = dat_o_mvfs$participant_id) %>%
    dplyr::mutate(
      outcome = out,
      model = "mvfs_model",
      se_type = "HC1",
      .before = 1
    )

  dth_coef_tbl[[paste0(out, "_mvfs")]] <- coef_mvfs

  #6.6.6.2 Robust coefficient table for full 17-variable model (HC1)
  coef_raw <- tidy_glm_hc1(fit_raw, cluster_var = dat_o_raw$participant_id) %>%
    dplyr::mutate(
      outcome = out,
      model = "raw_model",
      se_type = "HC1",
      .before = 1)

  dth_coef_tbl[[paste0(out, "_raw")]] <- coef_raw

  #6.6.7 Block test for raw predictor set
  log_msg("6.6 ", out, ": running raw-predictor HC1 block test")
  raw_block <- tryCatch(
    glm_block_test_hc1(
      fit = fit_raw,
      vars = params$risk_vars,
      cluster_var = dat_o_raw$participant_id),
    error = function(e) {
      log_msg("6.6 ", out, ": raw HC1 block test FAILED - ", conditionMessage(e))
      tibble::tibble(chisq = NA_real_, df = NA_real_, p.value = NA_real_, n_terms = NA_integer_)
    })
  
  #6.6.8 Run MVFS-predictor HC1 block test
  log_msg("6.6 ", out, ": running MVFS HC1 block test")
  mvfs_block <- tryCatch(
    glm_block_test_hc1(
      fit = fit_mvfs,
      vars = params$mvfs_vars,
      cluster_var = dat_o_mvfs$participant_id),
    error = function(e) {
      log_msg("6.6 ", out, ": MVFS HC1 block test FAILED - ", conditionMessage(e))
      tibble::tibble(chisq = NA_real_, df = NA_real_, p.value = NA_real_, n_terms = NA_integer_)
    })

  log_msg("6.6 ", out, ": MVFS HC1 block test complete")
  
  log_msg("6.6 ", out, ": raw-predictor HC1 block test complete")

  #6.6.9 Model-level summary comparison
  dth_results[[out]] <- tibble::tibble(
    outcome = out,
    n_rows = nrow(dat_o_cluster),
    n_ids = dplyr::n_distinct(dat_o_cluster$participant_id),
    n_events = sum(dat_o_cluster$event == 1L, na.rm = TRUE),
  
    cluster_aic = AIC(fit_cluster),
    cluster_bic = BIC(fit_cluster),
    cluster_loglik = as.numeric(logLik(fit_cluster)),
    cluster_pseudo_r2 = pseudo_r2_mcfadden(fit_cluster),
    cluster_brier = brier_score(dat_o_cluster$event, predict(fit_cluster, type = "response")),
  
    mvfs_aic = AIC(fit_mvfs),
    mvfs_bic = BIC(fit_mvfs),
    mvfs_loglik = as.numeric(logLik(fit_mvfs)),
    mvfs_pseudo_r2 = pseudo_r2_mcfadden(fit_mvfs),
    mvfs_brier = brier_score(dat_o_mvfs$event, predict(fit_mvfs, type = "response")),
  
    raw_aic = AIC(fit_raw),
    raw_bic = BIC(fit_raw),
    raw_loglik = as.numeric(logLik(fit_raw)),
    raw_pseudo_r2 = pseudo_r2_mcfadden(fit_raw),
    raw_brier = brier_score(dat_o_raw$event, predict(fit_raw, type = "response")),
  
    delta_aic_mvfs_minus_cluster = AIC(fit_mvfs) - AIC(fit_cluster),
    delta_bic_mvfs_minus_cluster = BIC(fit_mvfs) - BIC(fit_cluster),
    delta_pseudo_r2_mvfs_minus_cluster = pseudo_r2_mcfadden(fit_mvfs) - pseudo_r2_mcfadden(fit_cluster),
    delta_brier_mvfs_minus_cluster =
      brier_score(dat_o_mvfs$event, predict(fit_mvfs, type = "response")) -
      brier_score(dat_o_cluster$event, predict(fit_cluster, type = "response")),
  
    delta_aic_raw_minus_cluster = AIC(fit_raw) - AIC(fit_cluster),
    delta_bic_raw_minus_cluster = BIC(fit_raw) - BIC(fit_cluster),
    delta_pseudo_r2_raw_minus_cluster = pseudo_r2_mcfadden(fit_raw) - pseudo_r2_mcfadden(fit_cluster),
    delta_brier_raw_minus_cluster =
      brier_score(dat_o_raw$event, predict(fit_raw, type = "response")) -
      brier_score(dat_o_cluster$event, predict(fit_cluster, type = "response")),
  
    delta_aic_raw_minus_mvfs = AIC(fit_raw) - AIC(fit_mvfs),
    delta_bic_raw_minus_mvfs = BIC(fit_raw) - BIC(fit_mvfs),
    delta_pseudo_r2_raw_minus_mvfs = pseudo_r2_mcfadden(fit_raw) - pseudo_r2_mcfadden(fit_mvfs),
    delta_brier_raw_minus_mvfs =
      brier_score(dat_o_raw$event, predict(fit_raw, type = "response")) -
      brier_score(dat_o_mvfs$event, predict(fit_mvfs, type = "response")),
  
    mvfs_block_chisq = mvfs_block$chisq,
    mvfs_block_df = mvfs_block$df,
    mvfs_block_p = mvfs_block$p.value,
    mvfs_block_n_terms = mvfs_block$n_terms,
  
    raw_block_chisq = raw_block$chisq,
    raw_block_df = raw_block$df,
    raw_block_p = raw_block$p.value,
    raw_block_n_terms = raw_block$n_terms)

  #6.6.10 Predicted hazard by cluster across waves for interpretation
  dth_pred_tbl[[out]] <- pred_cluster_by_wave(
    fit = fit_cluster,
    data = dat_o_cluster,
    wave_var = "end_wave",
    age_var = "age_mid_between") %>%
    dplyr::mutate(
      outcome = out,
      model = "cluster_model",
      .before = 1)

  #6.6.11 Save model objects
  log_msg("6.6 ", out, ": saving GLM RDS files")
  save_rds2(fit_cluster, file.path(OUT_DIR, "rds", glue("dth_cluster_model_{out}.rds")))
  save_rds2(fit_mvfs, file.path(OUT_DIR, "rds", glue("dth_mvfs_model_{out}.rds")))
  save_rds2(fit_raw, file.path(OUT_DIR, "rds", glue("dth_raw_model_{out}.rds")))
  log_msg("6.6 ", out, ": GLM RDS files saved")

  #6.6.12 Cleanup
  rm(dat_o, dat_o_cluster, dat_o_mvfs, dat_o_raw,
   fit_cluster, fit_mvfs, fit_raw,
   coef_cluster, coef_mvfs, coef_raw, mvfs_block, raw_block)
  invisible(gc())
  log_gc(paste0("6.6 ", out, " end outcome"))
}

dth_summary_tbl <- bind_rows(dth_results)
dth_coef_tbl <- bind_rows(dth_coef_tbl)
dth_pred_tbl <- bind_rows(dth_pred_tbl)

write_csv2(dth_summary_tbl, file.path(OUT_DIR, "tables", "dth_cluster_vs_mvfs_vs_raw_model_summary.csv"))

write_csv2(dth_coef_tbl, file.path(OUT_DIR, "tables", "dth_all_model_robust_coefficients.csv"))

write_csv2(
  dth_coef_tbl %>% dplyr::filter(model == "mvfs_model"),
  file.path(OUT_DIR, "tables", "dth_mvfs_model_coefficients.csv"))

write_csv2(
  dth_coef_tbl %>% dplyr::filter(model == "raw_model"),
  file.path(OUT_DIR, "tables", "dth_raw_model_coefficients.csv"))

write_csv2(dth_pred_tbl, file.path(OUT_DIR, "tables", "dth_cluster_model_predicted_hazards_by_wave.csv"))

log_msg("6.6 DTH section complete")
log_gc("6.6 After DTH section")

knitr::kable(
  dth_summary_tbl,
  digits = 3,
  caption = "DTH sensitivity analysis: cluster-label model vs MVFS model vs full baseline predictor model") %>%
  kableExtra::kable_styling(full_width = FALSE)

```

## 6.7 Primary Prevalence Sensitivity Analysis: Cluster Labels vs Raw Predictors

```{r prevalence models, echo = FALSE, warning = FALSE}

## 6.7 Prevalence Sensitivity Analysis ##

prev_results <- list()
prev_coef_tbl <- list()
prev_pred_tbl <- list()
prev_or_by_wave_tbl <- list()

for (out in params$outcomes) {

  log_msg("6.7 Starting prevalence outcome: ", out)

  #6.7.1 Subset outcome-specific data
  dat_o <- bd_panel_model %>%
    dplyr::filter(outcome == out) %>%
    droplevels()

  if (nrow(dat_o) == 0) {
    log_msg("6.7 ", out, ": no rows available after filtering; skipping")
    next
  }

  #6.7.2 Create minimal model-specific datasets
  dat_o_cluster <- dat_o %>%
  dplyr::select(participant_id, status, wave, cluster, sex, baseline_status_bd, age_wave_gmc)

  dat_o_mvfs <- dat_o %>%
    dplyr::select(participant_id, status, wave, sex, baseline_status_bd, age_wave_gmc, dplyr::all_of(params$mvfs_vars))

  dat_o_raw <- dat_o %>%
    dplyr::select(participant_id, status, wave, sex, baseline_status_bd, age_wave_gmc, dplyr::all_of(params$risk_vars))
  
  #6.7.2.1 Binary-coded prevalence outcome count for logging/summaries
  n_positive_cur <- sum(coerce_binary01(dat_o_cluster$status, var_name = "status") == 1, na.rm = TRUE)

  log_msg(
    "6.7 ", out,
    ": rows=", nrow(dat_o_cluster),
    ", ids=", dplyr::n_distinct(dat_o_cluster$participant_id),
    ", positives=", n_positive_cur)
  
  log_gc(paste0("6.7 ", out, " before fits"))

  #6.7.3 Fit cluster-label GEE
  log_msg("6.7 ", out, ": fitting cluster GEE")
  fit_cluster <- tryCatch(
    fit_prev_gee(build_prev_formula_cluster(), dat_o_cluster),
    error = function(e) {
      log_msg("6.7 ", out, ": cluster GEE FAILED - ", conditionMessage(e))
      stop(e)
    })
  
  log_msg("6.7 ", out, ": cluster GEE fit complete")
  
  #6.7.3.1 One-time AUC diagnostic for the first prevalence outcome
  if (identical(out, params$outcomes[1])) {

    pr_chk <- as.numeric(predict(fit_cluster, type = "response"))
    y_chk <- coerce_binary01(dat_o_cluster$status, var_name = "status")

    auc_tbl_chk <- tibble::tibble(
      metric = c(
        "min_pred",
        "max_pred",
        "mean_pred_cases",
        "mean_pred_noncases",
        "auc_yardstick",
        "auc_pROC_direction_lt",
        "auc_pROC_direction_gt"),
      value = c(
        min(pr_chk, na.rm = TRUE),
        max(pr_chk, na.rm = TRUE),
        mean(pr_chk[y_chk == 1], na.rm = TRUE),
        mean(pr_chk[y_chk == 0], na.rm = TRUE),
        tryCatch(
          yardstick::roc_auc_vec(
            truth = factor(y_chk, levels = c(0, 1)),
            estimate = pr_chk,
            event_level = "second"),
          error = function(e) NA_real_),
        tryCatch(
          as.numeric(
            pROC::auc(
              pROC::roc(
                response = factor(y_chk, levels = c(0, 1)),
                predictor = pr_chk,
                levels = c("0", "1"),
                direction = "<",
                quiet = TRUE))),
          error = function(e) NA_real_),
        tryCatch(
          as.numeric(
            pROC::auc(
              pROC::roc(
                response = factor(y_chk, levels = c(0, 1)),
                predictor = pr_chk,
                levels = c("0", "1"),
                direction = ">",
                quiet = TRUE))),
          error = function(e) NA_real_)))

    write_csv2(
      auc_tbl_chk,
      file.path(OUT_DIR, "tables", "prevalence_auc_diagnostic_first_outcome.csv"))

    knitr::kable(
      auc_tbl_chk,
      digits = 4,
      caption = "AUC diagnostic for the first prevalence cluster model") %>%
      kableExtra::kable_styling(full_width = FALSE)
  }
  
  #6.7.4 Fit cluster x wave interaction GEE for timepoint-specific cluster ORs
  if (isTRUE(params$do_gee_interaction)) {
    log_msg("6.7 ", out, ": fitting cluster x wave interaction GEE")
  
    fit_cluster_int <- tryCatch(
      fit_prev_gee(build_prev_formula_cluster_int(), dat_o_cluster),
      error = function(e) {
        log_msg("6.7 ", out, ": cluster x wave interaction GEE FAILED - ", conditionMessage(e))
        structure("cluster_interaction_failed", class = "try-error")
      }
    )
  
    if (!inherits(fit_cluster_int, "try-error")) {
      log_msg("6.7 ", out, ": cluster x wave interaction GEE fit complete")
  
      prev_or_by_wave_tbl[[out]] <- or_by_wave_emm(
        fit = fit_cluster_int,
        outcome = out,
        model_label = "cluster_model_int"
      )
    }
  } else {
    fit_cluster_int <- structure("cluster_interaction_off", class = "try-error")
  }
  
  #6.7.5 Fit MVFS-variable GEE
  log_msg("6.7 ", out, ": fitting MVFS GEE")
  fit_mvfs <- tryCatch(
    fit_prev_gee(build_prev_formula_mvfs(), dat_o_mvfs),
    error = function(e) {
      log_msg("6.7 ", out, ": MVFS GEE FAILED - ", conditionMessage(e))
      stop(e)
    })

  log_msg("6.7 ", out, ": MVFS GEE fit complete")

  #6.7.6 Fit raw-variable GEE
  log_msg("6.7 ", out, ": fitting raw GEE")
  fit_raw <- tryCatch(
    fit_prev_gee(build_prev_formula_raw(), dat_o_raw),
    error = function(e) {
      log_msg("6.7 ", out, ": raw GEE FAILED - ", conditionMessage(e))
      stop(e)
    })
  
  log_msg("6.7 ", out, ": raw GEE fit complete")
  log_gc(paste0("6.7 ", out, " after GEE fits"))

  #6.7.7 Coefficient extraction
  #6.7.7.1 Robust coefficient table for cluster-label model
  log_msg("6.7 ", out, ": extracting cluster-model GEE coefficients")
  coef_cluster <- tidy_gee_robust(fit_cluster) %>%
    dplyr::mutate(
      outcome = out,
      model = "cluster_model",
      .before = 1)

  prev_coef_tbl[[paste0(out, "_cluster")]] <- coef_cluster
  
  #6.7.7.2 Robust coefficient table for MVFS model
  coef_mvfs <- tidy_gee_robust(fit_mvfs) %>%
    dplyr::mutate(
      outcome = out,
      model = "mvfs_model",
      .before = 1)

  prev_coef_tbl[[paste0(out, "_mvfs")]] <- coef_mvfs

  #6.7.7.3 Robust coefficient table for full 17-variable model
  coef_raw <- tidy_gee_robust(fit_raw) %>%
    dplyr::mutate(
      outcome = out,
      model = "raw_model",
      .before = 1)

  prev_coef_tbl[[paste0(out, "_raw")]] <- coef_raw

  #6.7.8 Block test for raw predictor set
  log_msg("6.7 ", out, ": running raw-predictor GEE block test")
  raw_block <- tryCatch(
    gee_block_test(fit_raw, vars = params$risk_vars),
    error = function(e) {
      log_msg("6.7 ", out, ": raw GEE block test FAILED - ", conditionMessage(e))
      tibble::tibble(chisq = NA_real_, df = NA_real_, p.value = NA_real_, n_terms = NA_integer_)
    })
  
  log_msg("6.7 ", out, ": raw-predictor GEE block test complete")
  
  #6.7.9 Run MVFS-predictor GEE block test
  log_msg("6.7 ", out, ": running MVFS GEE block test")
  mvfs_block <- tryCatch(
    gee_block_test(fit_mvfs, vars = params$mvfs_vars),
    error = function(e) {
      log_msg("6.7 ", out, ": MVFS GEE block test FAILED - ", conditionMessage(e))
      tibble::tibble(chisq = NA_real_, df = NA_real_, p.value = NA_real_, n_terms = NA_integer_)
    })

  log_msg("6.7 ", out, ": MVFS GEE block test complete")

  #6.7.10 QIC only if explicitly requested
  if (isTRUE(params$run_qic)) {
    log_msg("6.7 ", out, ": computing QIC")
    qic_cluster <- tryCatch(geepack::QIC(fit_cluster), error = function(e) NULL)
    qic_mvfs <- tryCatch(geepack::QIC(fit_mvfs), error = function(e) NULL)
    qic_raw <- tryCatch(geepack::QIC(fit_raw), error = function(e) NULL)

    qic_cluster_val <- if (!is.null(qic_cluster)) unname(qic_cluster[1]) else NA_real_
    qic_mvfs_val <- if (!is.null(qic_mvfs)) unname(qic_mvfs[1]) else NA_real_
    qic_raw_val <- if (!is.null(qic_raw)) unname(qic_raw[1]) else NA_real_
    log_msg("6.7 ", out, ": QIC complete")
  } else {
    qic_cluster_val <- NA_real_
    qic_mvfs_val <- NA_real_
    qic_raw_val <- NA_real_
    log_msg("6.7 ", out, ": skipping QIC because params$run_qic = FALSE")
  }
  
  #6.7.10.1 Predicted probabilities for performance summaries
  pr_cluster <- as.numeric(predict(fit_cluster, type = "response"))
  pr_mvfs <- as.numeric(predict(fit_mvfs, type = "response"))
  pr_raw <- as.numeric(predict(fit_raw, type = "response"))

  #6.7.11 Model-level comparison
  prev_results[[out]] <- tibble::tibble(
    outcome = out,
    n_rows = nrow(dat_o_cluster),
    n_ids = dplyr::n_distinct(dat_o_cluster$participant_id),
    n_positive = n_positive_cur,

    cluster_qic = qic_cluster_val,
    cluster_auc = auc_binary(dat_o_cluster$status, pr_cluster),
    cluster_logloss = log_loss_binary(dat_o_cluster$status, pr_cluster),
    cluster_brier = brier_score(dat_o_cluster$status, pr_cluster),

    mvfs_qic = qic_mvfs_val,
    mvfs_auc = auc_binary(dat_o_mvfs$status, pr_mvfs),
    mvfs_logloss = log_loss_binary(dat_o_mvfs$status, pr_mvfs),
    mvfs_brier = brier_score(dat_o_mvfs$status, pr_mvfs),

    raw_qic = qic_raw_val,
    raw_auc = auc_binary(dat_o_raw$status, pr_raw),
    raw_logloss = log_loss_binary(dat_o_raw$status, pr_raw),
    raw_brier = brier_score(dat_o_raw$status, pr_raw),

    delta_qic_mvfs_minus_cluster = qic_mvfs_val - qic_cluster_val,
    delta_auc_mvfs_minus_cluster = auc_binary(dat_o_mvfs$status, pr_mvfs) -
      auc_binary(dat_o_cluster$status, pr_cluster),
    delta_logloss_mvfs_minus_cluster = log_loss_binary(dat_o_mvfs$status, pr_mvfs) -
      log_loss_binary(dat_o_cluster$status, pr_cluster),
    delta_brier_mvfs_minus_cluster = brier_score(dat_o_mvfs$status, pr_mvfs) -
      brier_score(dat_o_cluster$status, pr_cluster),

    delta_qic_raw_minus_cluster = qic_raw_val - qic_cluster_val,
    delta_auc_raw_minus_cluster = auc_binary(dat_o_raw$status, pr_raw) -
      auc_binary(dat_o_cluster$status, pr_cluster),
    delta_logloss_raw_minus_cluster = log_loss_binary(dat_o_raw$status, pr_raw) -
      log_loss_binary(dat_o_cluster$status, pr_cluster),
    delta_brier_raw_minus_cluster = brier_score(dat_o_raw$status, pr_raw) -
      brier_score(dat_o_cluster$status, pr_cluster),

    delta_qic_raw_minus_mvfs = qic_raw_val - qic_mvfs_val,
    delta_auc_raw_minus_mvfs = auc_binary(dat_o_raw$status, pr_raw) -
      auc_binary(dat_o_mvfs$status, pr_mvfs),
    delta_logloss_raw_minus_mvfs = log_loss_binary(dat_o_raw$status, pr_raw) -
      log_loss_binary(dat_o_mvfs$status, pr_mvfs),
    delta_brier_raw_minus_mvfs = brier_score(dat_o_raw$status, pr_raw) -
      brier_score(dat_o_mvfs$status, pr_mvfs),

    mvfs_block_chisq = mvfs_block$chisq,
    mvfs_block_df = mvfs_block$df,
    mvfs_block_p = mvfs_block$p.value,
    mvfs_block_n_terms = mvfs_block$n_terms,

    raw_block_chisq = raw_block$chisq,
    raw_block_df = raw_block$df,
    raw_block_p = raw_block$p.value,
    raw_block_n_terms = raw_block$n_terms)

  #6.7.12 Predicted prevalence by cluster across waves
  prev_pred_tbl[[out]] <- pred_cluster_by_wave(
    fit = fit_cluster,
    data = dat_o_cluster,
    wave_var = "wave",
    age_var = "age_wave_gmc",
    baseline_var = "baseline_status_bd") %>%
    dplyr::mutate(
      outcome = out,
      model = "cluster_model",
      .before = 1)

  #6.7.13 Save model objects
  log_msg("6.7 ", out, ": saving GEE RDS files")
  save_gee_rds2(fit_cluster, file.path(OUT_DIR, "rds", glue("prev_cluster_model_{out}.rds")))
  
  if (!inherits(fit_cluster_int, "try-error")) {
    save_gee_rds2(fit_cluster_int, file.path(OUT_DIR, "rds", glue("prev_cluster_int_model_{out}.rds")))
  }
  
  save_gee_rds2(fit_mvfs, file.path(OUT_DIR, "rds", glue("prev_mvfs_model_{out}.rds")))
  save_gee_rds2(fit_raw, file.path(OUT_DIR, "rds", glue("prev_raw_model_{out}.rds")))
  log_msg("6.7 ", out, ": GEE RDS files saved")

  #6.7.14 Cleanup
  rm(dat_o, dat_o_cluster, dat_o_mvfs, dat_o_raw,
   fit_cluster, fit_cluster_int, fit_mvfs, fit_raw,
   coef_cluster, coef_mvfs, coef_raw, mvfs_block, raw_block,
   pr_cluster, pr_mvfs, pr_raw, n_positive_cur)
  invisible(gc())
  log_gc(paste0("6.7 ", out, " end outcome"))
}

prev_summary_tbl <- bind_rows(prev_results)
prev_coef_tbl <- bind_rows(prev_coef_tbl)
prev_pred_tbl <- bind_rows(prev_pred_tbl)
prev_or_by_wave_tbl <- bind_rows(prev_or_by_wave_tbl)

write_csv2(prev_summary_tbl, file.path(OUT_DIR, "tables", "prevalence_cluster_vs_mvfs_vs_raw_model_summary.csv"))

write_csv2(prev_coef_tbl, file.path(OUT_DIR, "tables", "prevalence_all_model_coefficients.csv"))

write_csv2(
  prev_coef_tbl %>% dplyr::filter(model == "mvfs_model"),
  file.path(OUT_DIR, "tables", "prevalence_mvfs_model_coefficients.csv"))

write_csv2(
  prev_coef_tbl %>% dplyr::filter(model == "raw_model"),
  file.path(OUT_DIR, "tables", "prevalence_raw_model_coefficients.csv"))

write_csv2(prev_pred_tbl, file.path(OUT_DIR, "tables", "prevalence_cluster_model_predicted_probabilities_by_wave.csv"))

write_csv2(prev_or_by_wave_tbl, file.path(OUT_DIR, "tables", "prevalence_cluster_interaction_OR_by_wave.csv"))

log_msg("6.7 Prevalence section complete")
log_gc("6.7 After prevalence section")

prev_summary_disp <- prev_summary_tbl %>%
  dplyr::mutate(
    dplyr::across(dplyr::ends_with("_auc"), ~ round(.x, 4)),
    dplyr::across(dplyr::ends_with("_logloss"), ~ round(.x, 6)),
    dplyr::across(dplyr::ends_with("_brier"), ~ round(.x, 6)),
    dplyr::across(dplyr::starts_with("delta_auc_"), ~ round(.x, 4)),
    dplyr::across(dplyr::starts_with("delta_logloss_"), ~ round(.x, 6)),
    dplyr::across(dplyr::starts_with("delta_brier_"), ~ round(.x, 6)))

knitr::kable(
  prev_summary_disp,
  caption = "Prevalence sensitivity analysis: cluster-label model vs MVFS model vs full baseline predictor model") %>%
  kableExtra::kable_styling(full_width = FALSE)

if (nrow(prev_or_by_wave_tbl) > 0) {
  knitr::kable(
    prev_or_by_wave_tbl,
    digits = 3,
    caption = "Prevalence cluster C2 vs C1 odds ratios by wave from the cluster x wave interaction GEE") %>%
    kableExtra::kable_styling(full_width = FALSE)
}

```

## 6.8 Extract Cluster-Main-Effect Results

```{r main effect tables, echo = FALSE, warning = FALSE}

## 6.8 Cluster Main-Effect Tables ##

#6.8.1 DTH cluster coefficient
dth_cluster_effects <- dth_coef_tbl %>%
  dplyr::filter(model == "cluster_model", term == "clusterC2") %>%
  dplyr::select(outcome, model, se_type, term, estimate, SE, df, OR, CI_lo, CI_hi, p.value)

#6.8.2 Prevalence cluster coefficient
prev_cluster_effects <- prev_coef_tbl %>%
  dplyr::filter(model == "cluster_model", term == "clusterC2") %>%
  dplyr::select(outcome, model, term, estimate, SE, OR, CI_lo, CI_hi, p.value)

#6.8.3 Write manuscript-facing extracts
write_csv2(dth_cluster_effects, file.path(OUT_DIR, "tables", "dth_cluster_main_effects_for_manuscript.csv"))
write_csv2(prev_cluster_effects, file.path(OUT_DIR, "tables", "prevalence_cluster_main_effects_for_manuscript.csv"))

#6.8.4 Print tables
knitr::kable(
  dth_cluster_effects,
  digits = 3,
  caption = "DTH: higher-risk cluster main effect (C2 vs C1) from the cluster-label utility model (CR2 preferred; HC1 fallback only if needed)") %>%
  kableExtra::kable_styling(full_width = FALSE)

knitr::kable(
  prev_cluster_effects,
  digits = 3,
  caption = "Prevalence: higher-risk cluster main effect (C2 vs C1) from the primary cluster-label model") %>%
  kableExtra::kable_styling(full_width = FALSE)

#6.8.5 Build a compact preview table for cluster-by-wave ORs, MVFS baseline predictors, and the most informative subset of full-model baseline predictors
top_full_prev_terms <- prev_coef_tbl %>%
  dplyr::filter(model == "raw_model", !is_nuisance_prev_term(term)) %>%
  dplyr::mutate(abs_z = abs(estimate / SE)) %>%
  dplyr::group_by(term) %>%
  dplyr::summarise(max_abs_z = max(abs_z, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(max_abs_z)) %>%
  dplyr::slice_head(n = 8) %>%
  dplyr::pull(term)

prev_key_coef_tbl <- dplyr::bind_rows(
  prev_or_by_wave_tbl %>%
  dplyr::mutate(
    model_set = "Cluster model: C2 vs C1 by wave",
    term = dplyr::recode(
      wave,
      "ses-02A" = "Wave 2",
      "ses-04A" = "Wave 4",
      "ses-06A" = "Wave 6")) %>%
  dplyr::select(model_set, outcome, term, OR, CI_lo, CI_hi, p.value),

  prev_coef_tbl %>%
    dplyr::filter(model == "mvfs_model", !is_nuisance_prev_term(term)) %>%
    dplyr::mutate(model_set = "MVFS prevalence model") %>%
    dplyr::select(model_set, outcome, term, OR, CI_lo, CI_hi, p.value),

  prev_coef_tbl %>%
    dplyr::filter(model == "raw_model", term %in% top_full_prev_terms) %>%
    dplyr::mutate(model_set = "Full 17-variable prevalence model (top baseline terms)") %>%
    dplyr::select(model_set, outcome, term, OR, CI_lo, CI_hi, p.value)) %>%
  dplyr::mutate(
    outcome = label_outcome(outcome),
    term = dplyr::if_else(
      model_set == "Cluster model: C2 vs C1 by wave",
      term,
      label_risk_term(term)))

write_csv2(
  prev_key_coef_tbl,
  file.path(OUT_DIR, "tables", "prevalence_key_mvfs_full_and_cluster_coefficients.csv"))

knitr::kable(
  prev_key_coef_tbl,
  digits = 3,
  caption = "Key prevalence coefficients from the cluster-by-wave interaction model, the MVFS model, and the highest-information subset of the full 17-variable model") %>%
  kableExtra::kable_styling(full_width = FALSE)

#6.8.6 Compact prevalence effect-size vs parsimony summary
prev_effectsize_summary <- prev_coef_tbl %>%
  dplyr::filter(model %in% c("mvfs_model", "raw_model"), !is_nuisance_prev_term(term)) %>%
  dplyr::mutate(
    abs_log_or = abs(log(OR)),
    sig = !is.na(p.value) & p.value < 0.05) %>%
  dplyr::group_by(model, outcome) %>%
  dplyr::summarise(
    n_terms = dplyr::n(),
    n_sig = sum(sig, na.rm = TRUE),
    median_abs_log_or = median(abs_log_or, na.rm = TRUE),
    max_or = max(pmax(OR, 1 / OR), na.rm = TRUE),
    .groups = "drop") %>%
  dplyr::mutate(outcome = label_outcome(outcome))

write_csv2(prev_effectsize_summary, file.path(OUT_DIR, "tables", "prevalence_effect_size_vs_parsimony_summary.csv"))

knitr::kable(
  prev_effectsize_summary,
  digits = 3,
  caption = "Prevalence baseline-predictor effect-size summary for the MVFS and full 17-variable models") %>%
  kableExtra::kable_styling(full_width = FALSE)

```

## 6.9 Visualization of Cluster-Model Predicted Risks Across Waves

```{r figures, echo = FALSE, warning = FALSE, fig.width = 8, fig.height = 6}

## 6.9 Predicted Probabilities by Wave for Cluster Models ##

#6.9.1 DTH predicted hazard plot
p_dth <- dth_pred_tbl %>%
  dplyr::mutate(
    end_wave = factor(end_wave, levels = c("ses-02A", "ses-04A", "ses-06A"))) %>%
  ggplot(aes(x = end_wave, y = pred, color = cluster, group = cluster)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  facet_wrap(~ outcome, scales = "free_y") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_classic(base_size = 11) +
  labs(
    title = "DTH cluster-model predicted first-onset hazard by wave",
    x = "Wave",
    y = "Predicted hazard",
    color = "Cluster")

ggsave(filename = file.path(OUT_DIR, "figures", "dth_cluster_model_predicted_hazards_by_wave.png"), plot = p_dth, width = 8, height = 6, dpi = 600, bg = "white")

print(p_dth)

#6.9.2 Prevalence predicted probability plot
p_prev <- prev_pred_tbl %>%
  dplyr::mutate(
    wave = factor(wave, levels = c("ses-02A", "ses-04A", "ses-06A"))) %>%
  ggplot(aes(x = wave, y = pred, color = cluster, group = cluster)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  facet_wrap(~ outcome, scales = "free_y") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_classic(base_size = 11) +
  labs(
    title = "Prevalence cluster-model predicted probability by wave",
    x = "Wave",
    y = "Predicted probability",
    color = "Cluster")

ggsave(filename = file.path(OUT_DIR, "figures", "prevalence_cluster_model_predicted_probabilities_by_wave.png"), plot = p_prev, width = 8, height = 6, dpi = 600, bg = "white")

print(p_prev)

#6.9.3 Linear OR prevalence coefficient figure for cluster, MVFS, and top full-model terms
#6.9.3.1 Helper to select top baseline terms within outcome
select_top_prev_terms <- function(coef_tbl, model_name, n_top = 6) {
  coef_tbl %>%
    dplyr::filter(model == model_name, !is_nuisance_prev_term(term)) %>%
    dplyr::mutate(abs_z = abs(estimate / SE)) %>%
    dplyr::group_by(outcome) %>%
    dplyr::slice_max(order_by = abs_z, n = n_top, with_ties = FALSE) %>%
    dplyr::ungroup() %>%
    dplyr::select(outcome, term)
}

#6.9.3.2 Terms to plot
#6.9.3.2.1 For the MVFS plot, show all non-nuisance MVFS terms for each outcome
mvfs_prev_terms_to_plot <- prev_coef_tbl %>%
  dplyr::filter(model == "mvfs_model", !is_nuisance_prev_term(term)) %>%
  dplyr::distinct(outcome, term)

#6.9.3.2.2 For the full model plot, show only the top baseline terms within each outcome
raw_prev_terms_to_plot <- select_top_prev_terms(
  coef_tbl = prev_coef_tbl,
  model_name = "raw_model",
  n_top = 7)

#6.9.3.3 Cluster main-effect plotting table
cluster_plot_tbl <- prev_coef_tbl %>%
  dplyr::filter(model == "cluster_model", term == "clusterC2") %>%
  dplyr::mutate(
    outcome_lab = factor(
      label_outcome(outcome),
      levels = c("Any BSD", "BD-I", "BD-II", "BD-NOS")),
    term_lab = "Higher-risk cluster (C2 vs C1)",
    sig = dplyr::if_else(!is.na(p.value) & p.value < 0.05, "p < .05", "NS"),
    OR_plot = pmin(OR, 10),
    CI_lo_plot = pmin(CI_lo, 10),
    CI_hi_plot = pmin(CI_hi, 10),
    truncated_hi = OR > 10 | CI_hi > 10)

#6.9.3.4 MVFS plotting table
mvfs_plot_tbl <- prev_coef_tbl %>%
  dplyr::inner_join(mvfs_prev_terms_to_plot, by = c("outcome", "term")) %>%
  dplyr::mutate(
    outcome_lab = factor(
      label_outcome(outcome),
      levels = c("Any BSD", "BD-I", "BD-II", "BD-NOS")),
    term_lab = label_risk_term(term),
    term_plot = tidytext::reorder_within(term_lab, OR, outcome_lab),
    sig = dplyr::if_else(!is.na(p.value) & p.value < 0.05, "p < .05", "NS"),
    OR_plot = pmin(OR, 10),
    CI_lo_plot = pmin(CI_lo, 10),
    CI_hi_plot = pmin(CI_hi, 10),
    truncated_hi = OR > 10 | CI_hi > 10)

#6.9.3.5 Full-model plotting table
raw_plot_tbl <- prev_coef_tbl %>%
  dplyr::inner_join(raw_prev_terms_to_plot, by = c("outcome", "term")) %>%
  dplyr::mutate(
    outcome_lab = factor(
      label_outcome(outcome),
      levels = c("Any BSD", "BD-I", "BD-II", "BD-NOS")),
    term_lab = label_risk_term(term),
    term_plot = tidytext::reorder_within(term_lab, OR, outcome_lab),
    sig = dplyr::if_else(!is.na(p.value) & p.value < 0.05, "p < .05", "NS"),
    OR_plot = pmin(OR, 10),
    CI_lo_plot = pmin(CI_lo, 10),
    CI_hi_plot = pmin(CI_hi, 10),
    truncated_hi = OR > 10 | CI_hi > 10)

#6.9.3.6 Shared plotting pieces
prev_or_theme <- theme_classic(base_size = 11) +
  theme(
    legend.position = "right",
    strip.background = element_rect(fill = "grey95", color = "black"),
    strip.text = element_text(face = "bold", size = 11),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 11),
    panel.spacing = unit(1.1, "lines"))

prev_or_x <- scale_x_continuous(
  breaks = seq(0, 10, by = 2),
  limits = c(0, 10),
  expand = ggplot2::expansion(mult = c(0.01, 0.03)))

prev_or_cols <- scale_color_manual(values = c("NS" = "grey55", "p < .05" = "black"))

#6.9.3.7 Cluster plot
p_prev_cluster_or <- cluster_plot_tbl %>%
  ggplot(aes(x = OR_plot, y = term_lab, color = sig)) +
  geom_vline(xintercept = 1, linetype = 2, linewidth = 0.3) +
  geom_segment(aes(x = CI_lo_plot, xend = CI_hi_plot, yend = term_lab), linewidth = 0.45) +
  geom_point(size = 1.8) +
  geom_point(
    data = ~ dplyr::filter(.x, truncated_hi),
    aes(x = 10, y = term_lab),
    shape = 17,
    size = 2.1,
    inherit.aes = FALSE) +
  facet_wrap(~ outcome_lab, ncol = 2, scales = "free_y") +
  prev_or_x +
  prev_or_cols +
  prev_or_theme +
  labs(
    title = "Cluster-label prevalence model odds ratios",
    subtitle = "Additive-model higher-risk cluster effect (C2 vs C1); triangles mark values exceeding the display cap of 10",
    x = "Odds ratio",
    y = NULL,
    color = NULL)

#6.9.3.8 MVFS plot
p_prev_mvfs_or <- mvfs_plot_tbl %>%
  ggplot(aes(x = OR_plot, y = term_plot, color = sig)) +
  geom_vline(xintercept = 1, linetype = 2, linewidth = 0.3) +
  geom_segment(aes(x = CI_lo_plot, xend = CI_hi_plot, yend = term_plot), linewidth = 0.45) +
  geom_point(size = 1.8) +
  geom_point(
    data = ~ dplyr::filter(.x, truncated_hi),
    aes(x = 10, y = term_plot),
    shape = 17,
    size = 2.1,
    inherit.aes = FALSE) +
  tidytext::scale_y_reordered() +
  facet_wrap(~ outcome_lab, ncol = 2, scales = "free_y") +
  prev_or_x +
  prev_or_cols +
  prev_or_theme +
  labs(
    title = "MVFS prevalence model odds ratios",
    subtitle = "All 7 MVFS baseline predictors shown for each outcome",
    x = "Odds ratio",
    y = NULL,
    color = NULL)

#6.9.3.9 Full-model plot
p_prev_raw_or <- raw_plot_tbl %>%
  ggplot(aes(x = OR_plot, y = term_plot, color = sig)) +
  geom_vline(xintercept = 1, linetype = 2, linewidth = 0.3) +
  geom_segment(aes(x = CI_lo_plot, xend = CI_hi_plot, yend = term_plot), linewidth = 0.45) +
  geom_point(size = 1.8) +
  geom_point(
    data = ~ dplyr::filter(.x, truncated_hi),
    aes(x = 10, y = term_plot),
    shape = 17,
    size = 2.1,
    inherit.aes = FALSE) +
  tidytext::scale_y_reordered() +
  facet_wrap(~ outcome_lab, ncol = 2, scales = "free_y") +
  prev_or_x +
  prev_or_cols +
  prev_or_theme +
  labs(
    title = "Full 17-variable prevalence model odds ratios",
    subtitle = "Top 7 baseline predictors per outcome by absolute Wald z-statistic",
    x = "Odds ratio",
    y = NULL,
    color = NULL)

#6.9.3.10 Combine into one figure
p_prev_coef <- (p_prev_cluster_or / p_prev_mvfs_or / p_prev_raw_or) +
  patchwork::plot_layout(heights = c(0.9, 1.8, 1.8), guides = "collect") &
  theme(legend.position = "right")

ggsave(filename = file.path(OUT_DIR, "figures", "prevalence_cluster_mvfs_fullmodel_linear_or_plot.png"), plot = p_prev_coef, width = 18, height = 16, dpi = 720, bg = "white")

print(p_prev_coef)

```

## 6.10 Plain-Language Takeaway Table

```{r takeaway table, echo = FALSE, warning = FALSE}

## 6.10 Compact Takeaway Table ##

#6.10.1 Combine DTH and prevalence model-comparison summaries
takeaway_tbl <- dplyr::bind_rows(
  dth_summary_tbl %>%
    dplyr::transmute(
      domain = "DTH",
      outcome = label_outcome(outcome),
      primary_metric = "McFadden pseudo-R2",
      cluster_primary = cluster_pseudo_r2,
      mvfs_primary = mvfs_pseudo_r2,
      raw_primary = raw_pseudo_r2,
      secondary_metric = "Brier score",
      cluster_secondary = cluster_brier,
      mvfs_secondary = mvfs_brier,
      raw_secondary = raw_brier,
      delta_mvfs_minus_cluster = delta_pseudo_r2_mvfs_minus_cluster,
      delta_raw_minus_cluster = delta_pseudo_r2_raw_minus_cluster,
      delta_raw_minus_mvfs = delta_pseudo_r2_raw_minus_mvfs,
      mvfs_block_p = mvfs_block_p,
      raw_block_p = raw_block_p),
  prev_summary_tbl %>%
    dplyr::transmute(
      domain = "Prevalence",
      outcome = label_outcome(outcome),
      primary_metric = "AUC",
      cluster_primary = cluster_auc,
      mvfs_primary = mvfs_auc,
      raw_primary = raw_auc,
      secondary_metric = "Log loss",
      cluster_secondary = cluster_logloss,
      mvfs_secondary = mvfs_logloss,
      raw_secondary = raw_logloss,
      delta_mvfs_minus_cluster = delta_auc_mvfs_minus_cluster,
      delta_raw_minus_cluster = delta_auc_raw_minus_cluster,
      delta_raw_minus_mvfs = delta_auc_raw_minus_mvfs,
      mvfs_block_p = mvfs_block_p,
      raw_block_p = raw_block_p))

write_csv2(takeaway_tbl, file.path(OUT_DIR, "tables", "cluster_vs_mvfs_vs_raw_takeaway_table.csv"))

takeaway_disp <- takeaway_tbl %>%
  dplyr::mutate(
    dplyr::across(
      c(cluster_primary, mvfs_primary, raw_primary,
        delta_mvfs_minus_cluster, delta_raw_minus_cluster, delta_raw_minus_mvfs),
      ~ round(.x, 4)),
    dplyr::across(c(cluster_secondary, mvfs_secondary, raw_secondary), ~ round(.x, 6)),
    dplyr::across(c(mvfs_block_p, raw_block_p), ~ round(.x, 4)))

knitr::kable(
  takeaway_disp,
  caption = "Compact comparison of cluster-label, MVFS, and full baseline predictor models") %>%
  kableExtra::kable_styling(full_width = FALSE)

#6.10.2 Dynamic interpretation summary
dth_best_tbl <- dth_summary_tbl %>%
  dplyr::mutate(
    best_model = dplyr::case_when(
      raw_pseudo_r2 >= mvfs_pseudo_r2 & raw_pseudo_r2 >= cluster_pseudo_r2 ~ "full 17-variable model",
      mvfs_pseudo_r2 >= raw_pseudo_r2 & mvfs_pseudo_r2 >= cluster_pseudo_r2 ~ "MVFS model",
      TRUE ~ "cluster model"),
    mvfs_share_of_full_gain = dplyr::if_else(
      is.finite(raw_pseudo_r2 - cluster_pseudo_r2) &
        abs(raw_pseudo_r2 - cluster_pseudo_r2) > 1e-10,
      (mvfs_pseudo_r2 - cluster_pseudo_r2) / (raw_pseudo_r2 - cluster_pseudo_r2),
      NA_real_))

prev_best_tbl <- prev_summary_tbl %>%
  dplyr::mutate(
    best_auc_model = dplyr::case_when(
      raw_auc >= mvfs_auc & raw_auc >= cluster_auc ~ "full 17-variable model",
      mvfs_auc >= raw_auc & mvfs_auc >= cluster_auc ~ "MVFS model",
      TRUE ~ "cluster model"))

auc_range_tbl <- prev_summary_tbl %>%
  dplyr::summarise(
    cluster_auc_min = min(cluster_auc, na.rm = TRUE),
    cluster_auc_max = max(cluster_auc, na.rm = TRUE),
    mvfs_auc_min = min(mvfs_auc, na.rm = TRUE),
    mvfs_auc_max = max(mvfs_auc, na.rm = TRUE),
    raw_auc_min = min(raw_auc, na.rm = TRUE),
    raw_auc_max = max(raw_auc, na.rm = TRUE))

cat("### Dynamic interpretation summary\n\n")

cat(
  "This sensitivity analysis asks whether the predictive utility in the baseline bipolar-risk architecture is carried mainly by the simple two-cluster label, by the reduced 7-variable MVFS, or only by the full 17-variable set.\n\n")

cat(glue::glue(
  "For first-onset risk (DTH), the full 17-variable model had the highest McFadden pseudo-R2 for {sum(dth_best_tbl$best_model == 'full 17-variable model', na.rm = TRUE)} of 4 outcomes, and the MVFS recovered {scales::percent(mean(dth_best_tbl$mvfs_share_of_full_gain, na.rm = TRUE), accuracy = 0.1)} of the incremental pseudo-R2 gain achieved by the full model over the cluster model. This indicates that the cluster label retains a meaningful low-dimensional summary of risk, but that some onset-relevant information is lost when the full baseline feature set is collapsed to either the MVFS or the two-cluster label.\n\n"))

cat(glue::glue(
  "For prevalence, the MVFS had the highest AUC for {sum(prev_best_tbl$best_auc_model == 'MVFS model', na.rm = TRUE)} of 4 outcomes, and both the MVFS and full 17-variable predictor blocks were jointly significant for all 4 outcomes. However, the absolute AUC values remained low across all models (cluster: {round(auc_range_tbl$cluster_auc_min, 3)}-{round(auc_range_tbl$cluster_auc_max, 3)}; MVFS: {round(auc_range_tbl$mvfs_auc_min, 3)}-{round(auc_range_tbl$mvfs_auc_max, 3)}; full 17-variable: {round(auc_range_tbl$raw_auc_min, 3)}-{round(auc_range_tbl$raw_auc_max, 3)}). Accordingly, the prevalence results are best interpreted as evidence of statistically meaningful association and relative parsimony, not as evidence of strong individual-level discrimination.\n\n"))

cat(
  "That distinction matters here: the higher-risk cluster still shows large timepoint-specific prevalence odds ratios, especially at earlier waves, but those strong group-level effect sizes do not translate into high pooled row-level classification performance. In practice, the current results support risk enrichment and clinically interpretable stratification more than screening-grade prediction.\n\n")

cat(
  "Taken together, the current preliminary takeaway is that the cluster label remains clinically interpretable and strongly associated with outcome, the MVFS is the most efficient reduced non-cluster representation of the concurrent prevalence signal, and the full 17-variable set retains the strongest utility for first-onset modeling.\n\n")

```

## 6.11 Reproducibility Log

```{r reproducibility, echo = FALSE, warning = FALSE}

## 6.11 Reproducibility Log ##

log <- list(
  timestamp = as.character(Sys.time()),
  repo = REPO,
  data_dir = DATA_DIR,
  risk_data_path = RISK_PATH,
  out_dir = OUT_DIR,
  outcomes = params$outcomes,
  dth_link = params$dth_link,
  gee_link = params$gee_link,
  do_gee_interaction = params$do_gee_interaction,
  gee_corstr = params$gee_corstr,
  run_qic = params$run_qic,
  save_full_gee_objects = params$save_full_gee_objects,
  progress_log_file = params$progress_log_file,
  risk_vars = params$risk_vars,
  mvfs_vars = params$mvfs_vars)

writeLines(jsonlite::toJSON(log, pretty = TRUE, auto_unbox = TRUE), con = file.path(OUT_DIR, "run_log.json"))

```
