---
title: "6. Predictive Utility of Cluster Labels vs Individual Risk Features (Reviewer 1, Comment 5): `r params$config`"
author: "Sam A. Sievertsen"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    df_print: paged
params:
  # I/O (datasets and outputs are namespaced by clustering configuration)
  repo: !r Sys.getenv("REPO", unset = here::here())
  data_dir: "data/data_processed/analysis_datasets"
  risk_data_rel: "data/data_processed/risk_variable_data.csv"
  out_dir: "results/main_analysis/2_statistical_analysis/6_risk_vars_vs_cluster_utility"
  config: "fallback_emo_ace"
  scaling_method: "z_score"
  k_value: 2
  # Outcomes
  outcomes: !r c("any_bsd", "bipolar_I", "bipolar_II", "bd_nos")
  # Cross-validation and bootstrap
  seed: 123
  n_folds: 10
  boot_B: 500                 # participant bootstrap of out-of-fold predictions, for AUC/Brier differences
  # Selected MVFS to evaluate, from 5_parsimonious_feature_selection (features are read from its output, not listed here)
  mvfs_search: "forward"
  mvfs_lambda_mode: "anchor_fixed"
  # Supplementary within-cluster ridge
  ridge_boot_B: 200
  ridge_outcome: "any_bsd"
  # Logging
  progress_log_file: "run_progress.log"
---

```{r global, include = FALSE, echo = FALSE, warning = FALSE}
knitr::opts_chunk$set(cache = FALSE, message = TRUE, warning = FALSE, results = "markup", comment = "")
```

## Overview

Reviewer 1 asked whether the risk clusters add predictive information beyond the individual features most closely tied to bipolar disorder, specifically parent-reported mania symptoms (GBI) and family history of mania. Nested models are compared for first onset (discrete-time hazard) and occurrence (by wave), each with the primary models' covariates:

| Model | Adds to covariates |
|---|---|
| M0 | nothing |
| M1 | GBI |
| M2 | family history of mania |
| M3 | GBI + family history of mania |
| M4 | GBI + family history of mania + cluster |
| M5 | cluster only |
| M6 | MVFS features |
| M7 | all clustering features |

The key comparison is **M4 vs M3**: the cluster's incremental value beyond the two bipolar-specific markers. M5, M6, and M7 are the cluster-label, MVFS, and full-feature representations compared in the submitted version of this analysis. Discrimination (AUC) and calibration-sensitive accuracy (Brier score) are estimated **out of fold** from `r params$n_folds`-fold cross-validation with folds assigned by family (siblings never split between training and test), using identical folds for every model; differences between models carry 95% intervals from a participant bootstrap of the out-of-fold predictions. In-sample AUC, AIC, and McFadden R2 are reported as secondary metrics. Added terms are tested with Wald tests on the participant-clustered CR2 covariance. Occurrence models are pooled logistic regressions with cluster-robust inference, which share the marginal mean model of a GEE with an independence working correlation and keep cross-validation tractable.

Cluster main effects and model-predicted risks by wave are not re-estimated here; they are produced once, by `1_bd_discrete_time_survival.Rmd` and `2_bd_mixed_effects_logit.Rmd`.

## 6.1 Set Up

```{r setup, echo = FALSE, warning = FALSE}

## 6.1 Set Up ##

#6.1.1 Load packages
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(readr)
  library(tibble)
  library(stringr)
  library(glue)
  library(purrr)
  library(clubSandwich)
  library(kableExtra)
  library(here)
  library(jsonlite)
})

#6.1.2 Reproducibility and numeric display
set.seed(params$seed)
options(scipen = 999, digits = 8)
options(encoding = "UTF-8")
knitr::opts_knit$set(encoding = "UTF-8")

#6.1.3 Resolve paths, namespaced by configuration
REPO <- params$repo
tag <- sprintf("%s_k%d_%s", params$config, as.integer(params$k_value), params$scaling_method)
DATA_DIR <- file.path(REPO, params$data_dir)
RISK_PATH <- file.path(REPO, params$risk_data_rel)
OUT_DIR <- file.path(REPO, params$out_dir, params$config)
ad_dir <- DATA_DIR

dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUT_DIR, "tables"), recursive = TRUE, showWarnings = FALSE)

#6.1.4 Safe writers
write_csv2 <- function(x, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  readr::write_csv(x, path)
}

#6.1.5 Shared configuration definitions (primary features, categorical features)
source(file.path(REPO, "scripts", "main_analysis", "1_clustering", "cluster_configs.R"))

#6.1.6 Confirm the label coding assumed below: C1 = lower-risk, C2 = higher-risk
orient <- readr::read_csv(file.path(DATA_DIR, sprintf("cluster_orientation_%s.csv", tag)), show_col_types = FALSE)
if (!isTRUE(orient$mean_cbcl_dp_raw_sum[orient$oriented_cluster == "C2"] >
            orient$mean_cbcl_dp_raw_sum[orient$oriented_cluster == "C1"])) {
  stop("Cluster coding check failed: C2 must be the higher-risk group (higher mean CBCL dysregulation profile sum)")
}

#6.1.7 Progress logging helpers
PROGRESS_LOG <- file.path(OUT_DIR, params$progress_log_file)
if (file.exists(PROGRESS_LOG)) file.remove(PROGRESS_LOG)

log_msg <- function(...) {
  txt <- paste0("[", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "] ", paste0(..., collapse = ""))
  cat(txt, "\n")
  cat(txt, "\n", file = PROGRESS_LOG, append = TRUE)
  flush.console()
}

log_gc <- function(tag = "") {
  gc_txt <- paste(capture.output(gc()), collapse = " | ")
  log_msg(tag, " | gc: ", gc_txt)
}

#6.1.8 Initial checkpoint
log_msg("6.1 Setup complete")
log_gc("6.1 After setup")

```

## 6.2 Load Baseline Risk Data and Prepare Predictors

```{r risk data, echo = FALSE, warning = FALSE}

## 6.2 Load Risk Data ##


#6.2.1 Baseline features: numeric features standardized (per 1 SD), binary features as 0/1
feats <- cluster_configs[[primary_config]]
risk <- readr::read_csv(RISK_PATH, show_col_types = FALSE)
stopifnot(all(c("participant_id", "age", feats) %in% names(risk)))
feat_df <- risk %>% dplyr::select(participant_id, age_baseline = age, dplyr::all_of(feats))
for (f in feats) {
  feat_df[[f]] <- if (f %in% feats_categorical) as.numeric(as.character(feat_df[[f]])) else as.numeric(scale(feat_df[[f]]))
}

#6.2.2 Selected MVFS for this configuration (the MVFS model is skipped if it is not available yet)
mvfs_rds <- file.path(REPO, "results", "main_analysis", "1_clustering", "5_parsimonious_feature_selection", params$config, "summary", "mvfs_selected_solutions.rds")
mvfs_feats <- NULL
if (file.exists(mvfs_rds)) {
  pick <- Filter(function(s) identical(s$search, params$mvfs_search) && identical(s$lambda_mode, params$mvfs_lambda_mode), readRDS(mvfs_rds))
  if (length(pick) == 1) mvfs_feats <- pick[[1]]$features
}
cat(if (is.null(mvfs_feats)) "MVFS selection not found: M6 is skipped\n" else glue("MVFS ({length(mvfs_feats)} features): {paste(mvfs_feats, collapse = ', ')}\n"))

log_msg("6.2 Risk data loaded: ", nrow(feat_df), " participants, ", length(feats), " features")

```

## 6.3 Load DTH Person-Period Data and Build Analysis Frame

```{r dth data, echo = FALSE, warning = FALSE}

## 6.3 Load DTH Data ##

#6.3.1 First-onset person-period data, prepared as in 1_bd_discrete_time_survival.Rmd
pp <- readRDS(file.path(ad_dir, sprintf("bd_person_period_%s.rds", tag))) %>%
  dplyr::mutate(cluster = factor(cluster, levels = c("C1", "C2")), end_wave = factor(end_wave), sex = factor(sex)) %>%
  dplyr::filter(event %in% c(0L, 1L), !is.na(cluster), !is.na(age_mid), !is.na(sex), !is.na(site_factor), !is.na(family_id)) %>%
  dplyr::inner_join(feat_df, by = "participant_id")
pp$age_base_c <- pp$age_baseline - mean(dplyr::distinct(pp, participant_id, age_baseline)$age_baseline)
base_ids <- pp %>% dplyr::filter(outcome == "any_bsd") %>% dplyr::group_by(participant_id) %>%
  dplyr::summarise(b = if (all(is.na(baseline_status))) NA_integer_ else as.integer(any(baseline_status == 1L, na.rm = TRUE)), .groups = "drop")
pp <- pp %>% dplyr::left_join(base_ids, by = "participant_id") %>% dplyr::filter(b == 0L) %>% dplyr::rename(y = event)

log_msg("6.3 DTH frame: ", nrow(pp), " rows, ", dplyr::n_distinct(pp$participant_id), " participants")

```

## 6.4 Load Occurrence Panel Data and Build Analysis Frame

```{r occurrence data, echo = FALSE, warning = FALSE}

## 6.4 Load Occurrence Data ##

#6.4.1 Occurrence panel data, prepared as in 2_bd_mixed_effects_logit.Rmd
pan <- readRDS(file.path(ad_dir, sprintf("bd_panel_%s.rds", tag))) %>%
  dplyr::mutate(wave = factor(wave, levels = c("ses-02A", "ses-04A", "ses-06A")), cluster = factor(cluster, levels = c("C1", "C2")),
                sex = factor(sex), baseline_status_bd = factor(baseline_status_bd)) %>%
  tidyr::drop_na(status, cluster, wave, sex, participant_id, age_wave_gmc, baseline_status_bd) %>%
  dplyr::inner_join(feat_df, by = "participant_id") %>%
  dplyr::rename(y = status)

#6.4.2 Family-grouped folds, shared by every model and both frameworks
set.seed(params$seed)
fams <- unique(c(as.character(pp$family_id), as.character(pan$family_id)))
fold_of_family <- setNames(sample(rep(seq_len(params$n_folds), length.out = length(fams))), fams)
pp$fold <- fold_of_family[as.character(pp$family_id)]
pan$fold <- fold_of_family[as.character(pan$family_id)]
log_msg("6.4 Occurrence frame: ", nrow(pan), " rows; ", params$n_folds, " family-grouped folds assigned")

```

## 6.5 Helpers for Model Fitting and Comparison

```{r helpers, echo = FALSE, warning = FALSE}


## 6.5 Helpers ##

#6.5.1 Rank-based AUC (Mann-Whitney) and Brier score
auc_fast <- function(y, p) {
  n1 <- sum(y == 1); n0 <- length(y) - n1
  if (n1 == 0 || n0 == 0) return(NA_real_)
  (sum(rank(p)[y == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}
brier <- function(y, p) mean((y - p)^2)

#6.5.2 Out-of-fold predictions for one model (fit on k-1 folds, predict the held-out fold)
oof_predict <- function(fml, d) {
  p <- rep(NA_real_, nrow(d))
  for (k in sort(unique(d$fold))) {
    tr <- d$fold != k
    f <- suppressWarnings(stats::glm(fml, family = binomial(), data = d[tr, , drop = FALSE]))
    p[!tr] <- suppressWarnings(stats::predict(f, newdata = d[!tr, , drop = FALSE], type = "response"))
  }
  p
}

#6.5.3 Wald chi-square for a block of terms on the participant-clustered CR2 covariance
cr2_wald <- function(fit, d, pattern) {
  idx <- grep(pattern, names(stats::coef(fit)))
  if (length(idx) == 0 || anyNA(stats::coef(fit)[idx])) return(c(chi_sq = NA, df = NA, p = NA))
  V <- as.matrix(clubSandwich::vcovCR(fit, cluster = d$participant_id, type = "CR2"))[idx, idx, drop = FALSE]
  b <- stats::coef(fit)[idx]
  chi <- as.numeric(t(b) %*% solve(V) %*% b)
  c(chi_sq = chi, df = length(idx), p = stats::pchisq(chi, length(idx), lower.tail = FALSE))
}

#6.5.4 Participant bootstrap of a difference in out-of-fold AUC and Brier between two models
boot_diff <- function(y, p_a, p_b, pid, B) {
  rows <- split(seq_along(y), pid)
  set.seed(params$seed)
  m <- replicate(B, {
    i <- unlist(rows[sample.int(length(rows), replace = TRUE)], use.names = FALSE)
    c(auc_fast(y[i], p_a[i]) - auc_fast(y[i], p_b[i]), brier(y[i], p_a[i]) - brier(y[i], p_b[i]))
  })
  c(d_auc_lo = stats::quantile(m[1, ], .025, na.rm = TRUE), d_auc_hi = stats::quantile(m[1, ], .975, na.rm = TRUE),
    d_brier_lo = stats::quantile(m[2, ], .025, na.rm = TRUE), d_brier_hi = stats::quantile(m[2, ], .975, na.rm = TRUE))
}

#6.5.5 Model specifications: covariates plus each model's added terms
model_terms <- list(M0 = character(0), M1 = "mh_p_gbi_sum", M2 = "family_history_mania",
                    M3 = c("mh_p_gbi_sum", "family_history_mania"), M4 = c("mh_p_gbi_sum", "family_history_mania", "cluster"),
                    M5 = "cluster", M6 = mvfs_feats, M7 = feats)
if (is.null(mvfs_feats)) model_terms$M6 <- NULL
comparisons <- list(c("M1", "M0"), c("M2", "M0"), c("M3", "M0"), c("M4", "M3"), c("M5", "M0"), c("M6", "M0"), c("M7", "M0"), c("M7", "M5"))
comparisons <- Filter(function(cmp) all(cmp %in% names(model_terms)), comparisons)
added_pattern <- function(a, b) {
  extra <- setdiff(model_terms[[a]], model_terms[[b]])
  paste0("^(", paste(ifelse(extra == "cluster", "clusterC2", extra), collapse = "|"), ")$")
}

#6.5.6 Ridge logistic by penalized IRLS (supplementary 6.8); pen = 0 leaves a column unpenalized (intercept and covariates)
ridge_logit <- function(X, y, pen, lambda, maxit = 100, tol = 1e-8) {
  b <- rep(0, ncol(X)); P <- diag(lambda * pen, ncol(X))
  for (it in seq_len(maxit)) {
    eta <- drop(X %*% b); mu <- stats::plogis(eta); w <- pmax(mu * (1 - mu), 1e-10)
    b_new <- drop(solve(crossprod(X, X * w) + P, crossprod(X, w * (eta + (y - mu) / w))))
    if (max(abs(b_new - b)) < tol) return(b_new)
    b <- b_new
  }
  b
}

```

## 6.6 Nested Model Comparison: Onset and Occurrence

```{r model comparison, echo = FALSE, warning = FALSE, results = "asis"}


## 6.6 Fit every model, out of fold and in sample, for both frameworks and every outcome ##
frameworks <- list(
  `First onset (DTH)` = list(data = pp, covars = "sex + end_wave + age_base_c"),
  `Occurrence` = list(data = pan, covars = "sex + wave + baseline_status_bd + age_wave_gmc"))

metric_rows <- list(); cmp_rows <- list()
for (fw in names(frameworks)) for (oc in params$outcomes) {
  d <- frameworks[[fw]]$data %>% dplyr::filter(outcome == oc) %>% droplevels() %>% as.data.frame()
  if (nrow(d) == 0 || sum(d$y) < 20) { cat(glue("\n\n**{fw}, {oc}:** skipped (fewer than 20 events)\n\n")); next }
  
  log_msg("6.6 ", fw, ", ", oc, ": fitting ", length(model_terms), " models")
  
  #6.6.1 Each model: in-sample fit (AIC, McFadden R2) and out-of-fold predictions
  fits <- list(); oof <- list()
  for (m in names(model_terms)) {
    rhs <- paste(c(frameworks[[fw]]$covars, model_terms[[m]]), collapse = " + ")
    fml <- stats::as.formula(paste("y ~", rhs))
    fits[[m]] <- suppressWarnings(stats::glm(fml, family = binomial(), data = d))
    oof[[m]] <- oof_predict(fml, d)
    metric_rows[[paste(fw, oc, m)]] <- tibble::tibble(
      framework = fw, outcome = oc, model = m, n_rows = nrow(d), events = sum(d$y),
      auc_oof = auc_fast(d$y, oof[[m]]), brier_oof = brier(d$y, oof[[m]]),
      auc_in = auc_fast(d$y, stats::fitted(fits[[m]])), aic = stats::AIC(fits[[m]]),
      mcfadden_r2 = 1 - fits[[m]]$deviance / fits[[m]]$null.deviance)
  }
  
  #6.6.2 Planned comparisons: out-of-fold AUC/Brier differences (bootstrap CIs) and CR2 Wald test for the added terms
  for (cmp in comparisons) {
    a <- cmp[1]; b <- cmp[2]
    ci <- boot_diff(d$y, oof[[a]], oof[[b]], d$participant_id, params$boot_B)
    w <- cr2_wald(fits[[a]], d, added_pattern(a, b))
    cmp_rows[[paste(fw, oc, a, b)]] <- tibble::tibble(
      framework = fw, outcome = oc, comparison = paste(a, "vs", b),
      d_auc = auc_fast(d$y, oof[[a]]) - auc_fast(d$y, oof[[b]]), d_auc_lo = ci[[1]], d_auc_hi = ci[[2]],
      d_brier = brier(d$y, oof[[a]]) - brier(d$y, oof[[b]]), d_brier_lo = ci[[3]], d_brier_hi = ci[[4]],
      wald_chi_sq = w[["chi_sq"]], wald_df = w[["df"]], wald_p = w[["p"]])
  }
}
metric_tbl <- dplyr::bind_rows(metric_rows); cmp_tbl <- dplyr::bind_rows(cmp_rows)
write_csv2(metric_tbl, file.path(OUT_DIR, "tables", "utility_model_metrics.csv"))
write_csv2(cmp_tbl, file.path(OUT_DIR, "tables", "utility_model_comparisons.csv"))

print(metric_tbl %>% kable(digits = 4, caption = "Model performance: out-of-fold AUC and Brier (family-grouped CV), in-sample AUC, AIC, McFadden R2") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 11))
print(cmp_tbl %>% kable(digits = 4, caption = "Planned comparisons: difference in out-of-fold AUC and Brier (first minus second; participant-bootstrap 95% CI) and CR2 Wald test of the added terms") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 11))
log_msg("6.6 Model comparison complete")

```

## 6.7 Feature-Level Associations (First Onset, Any BSD)

```{r feature ors, echo = FALSE, warning = FALSE}


## 6.7 Standardized ORs for every clustering feature, mutually adjusted (M7, first onset, any BSD) ##
d7 <- pp %>% dplyr::filter(outcome == "any_bsd") %>% droplevels() %>% as.data.frame()
f7 <- stats::glm(stats::as.formula(paste("y ~ sex + end_wave + age_base_c +", paste(feats, collapse = " + "))), family = binomial(), data = d7)
V7 <- as.matrix(clubSandwich::vcovCR(f7, cluster = d7$participant_id, type = "CR2"))
feat_or <- tibble::tibble(feature = feats, b = stats::coef(f7)[feats], se = sqrt(diag(V7))[feats]) %>%
  dplyr::mutate(type = ifelse(feature %in% feats_categorical, "binary (present vs absent)", "numeric (per 1 SD)"),
                OR = exp(b), CI_lo = exp(b - 1.959964 * se), CI_hi = exp(b + 1.959964 * se), p = 2 * stats::pnorm(-abs(b / se))) %>%
  dplyr::arrange(p) %>% dplyr::select(feature, type, OR, CI_lo, CI_hi, p)
write_csv2(feat_or, file.path(OUT_DIR, "tables", "feature_standardized_ORs_onset_any_bsd.csv"))
feat_or %>% kable(digits = 3, caption = "Mutually adjusted feature ORs for first onset of any BSD (CR2 by participant)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 11)

```

## 6.8 Supplementary: Within-Cluster Ridge Regression (Exploratory)

Within each cluster, a ridge-penalized pooled logistic model of first onset (`r params$ridge_outcome`) is fit to all clustering features (standardized within cluster), with covariates left unpenalized. The penalty is chosen by family-grouped cross-validated deviance, and 95% intervals come from `r params$ridge_boot_B` participant-bootstrap refits at that penalty. **Interpretive caution:** clusters were formed from these same features, so each feature's range is restricted within a cluster and associations can attenuate or reverse relative to the full sample; these coefficients describe associations within each group and are not evidence of distinct causal pathways.

```{r ridge, echo = FALSE, warning = FALSE}


## 6.8 Within-cluster ridge (exploratory) ##

#6.8.1 ridge_logit() is defined with the other helpers (6.5.6)

ridge_rows <- list()
for (cl in c("C1", "C2")) {
  d <- pp %>% dplyr::filter(outcome == params$ridge_outcome, cluster == cl) %>% droplevels() %>% as.data.frame()
  
  #6.8.2 Design: unpenalized intercept and covariates, then features standardized within this cluster (constant features dropped)
  Xc <- stats::model.matrix(~ sex + end_wave + age_base_c, d)
  keep <- feats[vapply(feats, function(f) stats::sd(d[[f]]) > 0, logical(1))]
  Xf <- scale(as.matrix(d[, keep, drop = FALSE]))
  X <- cbind(Xc, Xf); pen <- c(rep(0, ncol(Xc)), rep(1, ncol(Xf)))
  
  #6.8.3 Penalty chosen by family-grouped cross-validated deviance
  grid <- exp(seq(log(0.1), log(1000), length.out = 30))
  cv_dev <- vapply(grid, function(lam) {
    sum(vapply(sort(unique(d$fold)), function(k) {
      tr <- d$fold != k; b <- ridge_logit(X[tr, ], d$y[tr], pen, lam)
      mu <- pmin(pmax(stats::plogis(drop(X[!tr, ] %*% b)), 1e-12), 1 - 1e-12)
      -2 * sum(d$y[!tr] * log(mu) + (1 - d$y[!tr]) * log(1 - mu))
    }, numeric(1)))
  }, numeric(1))
  lam <- grid[which.min(cv_dev)]
  
  #6.8.4 Fit at the chosen penalty, then participant-bootstrap refits for intervals
  b_hat <- ridge_logit(X, d$y, pen, lam)
  rows <- split(seq_len(nrow(d)), d$participant_id)
  set.seed(params$seed)
  boots <- replicate(params$ridge_boot_B, {
    i <- unlist(rows[sample.int(length(rows), replace = TRUE)], use.names = FALSE)
    ridge_logit(X[i, ], d$y[i], pen, lam)[colnames(Xf)]
  })
  ridge_rows[[cl]] <- tibble::tibble(cluster = ifelse(cl == "C2", "Higher-Risk", "Lower-Risk"), feature = colnames(Xf),
                                     events = sum(d$y), lambda = lam, OR_per_SD = exp(b_hat[colnames(Xf)]),
                                     CI_lo = exp(apply(boots, 1, stats::quantile, .025)), CI_hi = exp(apply(boots, 1, stats::quantile, .975)))
}
ridge_tbl <- dplyr::bind_rows(ridge_rows)
write_csv2(ridge_tbl, file.path(OUT_DIR, "tables", "within_cluster_ridge_onset.csv"))
ridge_tbl %>%
  tidyr::pivot_wider(id_cols = feature, names_from = cluster, values_from = c(OR_per_SD, CI_lo, CI_hi)) %>%
  kable(digits = 3, caption = "Within-cluster ridge: ORs per 1 SD (within cluster) for first onset, with participant-bootstrap 95% CIs") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 11)
log_msg("6.8 Within-cluster ridge complete")

```

## 6.9 Compact Takeaway Table

```{r takeaway, echo = FALSE, warning = FALSE}

## 6.9 Compact Takeaway Table ##

#6.9.1 The three comparisons the manuscript reports, per framework and outcome
takeaway_tbl <- cmp_tbl %>%
  dplyr::filter(comparison %in% c("M4 vs M3", "M5 vs M0", "M7 vs M5")) %>%
  dplyr::mutate(question = dplyr::recode(comparison,
                  `M4 vs M3` = "Cluster beyond GBI + family history of mania",
                  `M5 vs M0` = "Cluster label vs covariates alone",
                  `M7 vs M5` = "All features vs cluster label"),
                `Delta AUC [95% CI]` = sprintf("%.3f [%.3f, %.3f]", d_auc, d_auc_lo, d_auc_hi),
                `Wald p` = ifelse(wald_p < .001, "< .001", sprintf("%.3f", wald_p))) %>%
  dplyr::select(framework, outcome, question, `Delta AUC [95% CI]`, `Wald p`)
write_csv2(takeaway_tbl, file.path(OUT_DIR, "tables", "utility_takeaway_table.csv"))
takeaway_tbl %>%
  kable(caption = "Takeaway: out-of-fold AUC differences and Wald tests for the reported comparisons") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 11)

```

## 6.10 Reproducibility Log

```{r reproducibility, echo = FALSE, warning = FALSE}

## 6.10 Reproducibility Log ##

log <- list(
  timestamp = as.character(Sys.time()),
  repo = REPO,
  config = params$config,
  data_dir = DATA_DIR,
  risk_data_path = RISK_PATH,
  out_dir = OUT_DIR,
  outcomes = params$outcomes,
  n_folds = params$n_folds,
  boot_B = params$boot_B,
  features = feats,
  mvfs_features = mvfs_feats,
  mvfs_selection = paste(params$mvfs_search, params$mvfs_lambda_mode, sep = "/"),
  ridge_boot_B = params$ridge_boot_B)

writeLines(jsonlite::toJSON(log, pretty = TRUE, auto_unbox = TRUE), con = file.path(OUT_DIR, "run_log.json"))

```
