---
title: "Results Compiler - ABCD MDS Risk"
author: "Sam A. Sievertsen"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    code_folding: hide
params:
  seed: 123
  link_primary: "logit"
  wave_ref: "ses-04A"
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(fig.width = 8, fig.height = 5, dpi = 300, fig.align = "center", echo = TRUE, cache = FALSE, message = TRUE, warning = TRUE, results = "markup", verbose = TRUE, comment = "")

```

```{r setup, include=FALSE}

#1. Script Setup
#1.1 Load in dependent packages from renv
#1.1.1 Create a function to load libraries safely
safe_library <- function(pkg) {
  suppressWarnings({
    suppressPackageStartupMessages({
      tryCatch({
        library(pkg, character.only = TRUE)
      }, error = function(e) {
        message(sprintf("[WARN] Package '%s' not available. Some features may be skipped.", pkg))
      })
    })
  })
}

#1.1.2 Load required packages
pkgs <- c(
  "ggplot2", "dplyr", "tidyr", "stringr", "tibble", "clustMixType", "here",
  "glue", "janitor", "readr", "broom", "broom.mixed", "gt", "gtsummary",
  "kableExtra", "patchwork", "cowplot", "ggtext", "scales", "mgcv", "geepack",
  "lme4", "glmmTMB", "clubSandwich", "DHARMa", "performance", "consort",
  "ggrepel", "GGally", "sf", "viridisLite", "webshot2", "purrr",
  "DiagrammeR", "DiagrammeRsvg", "rsvg", "emmeans", "effectsize"
)

#1.1.3 Load all packages safely
invisible(lapply(pkgs, safe_library))

#1.2 Set global options and themes
#1.2.1 Set seed for reproducibility
set.seed(123)

#1.2.2 Set table style to JAMA-like if gtsummary is available
if (requireNamespace("gtsummary", quietly = TRUE)) {
  gtsummary::theme_gtsummary_journal(journal = "jama")
}

#1.2.3 Set ggplot2 text default size
theme_set(theme_minimal(base_size = 12))

#1.3 Establish paths to relevant directories & files as they exist on ARC
#1.3.1 Project root (assumes Rmd is at repo root)
root_dir <- here::here()

#1.3.2 Data roots
data_root <- here::here("data")
data_processed <- here::here("data", "data_processed")
data_raw <- here::here("data", "data_raw")

#1.3.3 Common processed subtrees
analysis_datasets <- here::here("data", "data_processed", "analysis_datasets")
kproto_results <- here::here("data", "data_processed", "kproto_results")
stability_results <- here::here("data", "data_processed", "stability_results")
validation_results<- here::here("data", "data_processed", "validation_results")

#1.3.4 Results (model outputs) roots
results_root <- here::here("results")
results_main <- here::here("results", "main_analysis")
res_bd_surv <- here::here("results", "main_analysis", "1_bd_survival")
res_bd_mixed <- here::here("results", "main_analysis", "2_bd_mixed_logit")

#1.3.5 Assets (for Figure 0 schematic, etc.)
assets_dir <- here::here("assets")

#1.3.6 Export targets (per §5 filenames)
fig_dir <- here::here("results", "figures")
tab_dir <- here::here("results", "tables")

#1.4 Ensure export directories exist
for (d in c(fig_dir, tab_dir)) if (!dir.exists(d)) dir.create(d, recursive = TRUE)

#1.5 Set default color palette for clusters
cluster_cols <- c("#7F3C8D", "#11A579")

#1.6 Helper functions to be used throughout
#1.6.1 Save figure (PDF + PNG) with transparent background
save_fig <- function(plot, base_filename, width = 9, height = 8, dpi = 720) {
  pdf_path <- file.path(fig_dir, paste0(base_filename, ".pdf"))
  png_path <- file.path(fig_dir, paste0(base_filename, ".png"))
  ggplot2::ggsave(pdf_path, plot = plot, width = width, height = height, bg = "transparent")
  ggplot2::ggsave(png_path, plot = plot, width = width, height = height, dpi = dpi, bg = "transparent")
  invisible(list(pdf = pdf_path, png = png_path))
}

#1.6.2 Format CI labels
ci_label <- function(lo, hi, digits = 2) sprintf("(%.2f, %.2f)", round(lo, digits), round(hi, digits))

#1.6.3 Format OR with CI
fmt_or <- function(est, lo, hi, digits = 2) sprintf("%.2f %s", round(est, digits), ci_label(lo, hi, digits))

#1.6.4 Pretty p-value
strip_p <- function(p) ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

#1.6.5 Patchwork legend bottom helper
legend_bottom <- function(p) p + patchwork::plot_layout(guides = "collect") & theme(legend.position = "bottom")

#1.6.6 Panel tag helper
panel_tag <- function(p, tag = "A") p + ggplot2::labs(tag = tag) + theme(plot.tag = element_text(face = "bold", size = 12))

#1.6.7 GT styled table save helper (HTML/PNG/RTF) with fallbacks
save_gt <- function(gt_tbl, base_filename) {
  html_path <- file.path(tab_dir, paste0(base_filename, ".html"))
  png_path <- file.path(tab_dir, paste0(base_filename, ".png"))
  rtf_path <- file.path(tab_dir, paste0(base_filename, ".rtf"))

  gt::gtsave(gt_tbl, html_path)
  if (requireNamespace("webshot2", quietly = TRUE)) {

    #1.6.7.1 On headless ARC nodes, webshot2 should still work; PNG save is best-effort
    try(gt::gtsave(gt_tbl, png_path), silent = TRUE)
  }
  try(gt::gtsave(gt_tbl, rtf_path), silent = TRUE)
  invisible(list(html = html_path, png = png_path, rtf = rtf_path))
}

#1.6.8 Safe read: CSV or RDS; if missing, return NULL
safe_read <- function(path) {
  if (file.exists(path)) {
    ext <- tolower(tools::file_ext(path))
    if (ext == "rds") return(readRDS(path))
    if (ext %in% c("csv", "tsv")) return(readr::read_csv(path, show_col_types = FALSE))
  }
  return(NULL)
}

#1.6.9 Read or TK placeholder by expected columns
read_or_tk <- function(path, tk_cols, tk_n = 10) {
  df <- safe_read(path)
  if (!is.null(df)) return(df)
  out <- tibble::tibble(.rows = tk_n)
  for (nm in tk_cols) out[[nm]] <- NA
  out <- janitor::remove_empty(out, which = c("rows", "cols"))
  attr(out, "TK") <- TRUE
  out
}

#1.6.10 Small utility function to detect TK data or not 
is_tk <- function(df) isTRUE(isTRUE(attr(df, "TK")))

#1.6.11 CONSORT helper to safely label TK captions
tk_caption <- function(base) paste0(base, ifelse(grepl("TK", base, fixed=TRUE) ||
  grepl("TK", paste(capture.output(str(base)), collapse=" "), fixed=TRUE),
  " (TK)", ""))

#1.6.12 Convenience: quick path joiners for later sections. These keep the Inputs chunk simple when pointing to ARC outputs
p_data <- function(...) file.path(data_root, ...)
p_proc <- function(...) file.path(data_processed, ...)
p_analysis <- function(...) file.path(analysis_datasets, ...)
p_kproto <- function(...) file.path(kproto_results, ...)
p_stab <- function(...) file.path(stability_results, ...)
p_valid <- function(...) file.path(validation_results, ...)
p_res <- function(...) file.path(results_root, ...)
p_main <- function(...) file.path(results_main, ...)
p_bd_surv <- function(...) file.path(res_bd_surv, ...)
p_bd_mix <- function(...) file.path(res_bd_mixed, ...)

#1.6.13 Normalize BD outcome names to {BDI, BDII, BSD, BDNOS} (if present)
norm_bd_outcome <- function(x) {
  dplyr::case_when(
    stringr::str_detect(x, stringr::regex("bipolar[_-]?ii", ignore_case = TRUE)) ~ "BDII",
    stringr::str_detect(x, stringr::regex("bipolar[_-]?i(?!i)", ignore_case = TRUE)) ~ "BDI",
    stringr::str_detect(x, stringr::regex("any[_-]?bsd|\\bbd?s?d\\b", ignore_case = TRUE)) ~ "BSD",
    stringr::str_detect(x, stringr::regex("bd[_-]?nos", ignore_case = TRUE)) ~ "BDNOS",
    TRUE ~ x)
}

#1.6.14 Safer directory stacker function: only stack CSVs (or DFs); skip list/vec RDS
stack_dir <- function(dir, pattern, mutate_f = NULL) {
  if (!dir.exists(dir)) return(NULL)
  fs <- list.files(dir, pattern = pattern, full.names = TRUE)
  if (!length(fs)) return(NULL)
  out <- purrr::map(fs, function(f) {
    df <- safe_read(f)

    #1.6.14.1 Skip non-data-frame objects
    if (is.null(df) || !is.data.frame(df)) return(NULL)
    df$file <- basename(f)
    df})
  out <- purrr::compact(out)
  if (!length(out)) return(NULL)
  res <- dplyr::bind_rows(out)
  if (!is.null(mutate_f)) res <- mutate_f(res)
  res
}

#1.6.15 Helper to read val_z_score_silhouette_k*.rds into a tidy DF if possible
read_val_sil_perk <- function(dir) {
  if (!dir.exists(dir)) return(NULL)
  fs <- list.files(dir, pattern = "^val_z_score_silhouette_k[0-9]+\\.rds$", full.names = TRUE)
  if (!length(fs)) return(NULL)
  purrr::map_dfr(fs, function(f) {
    obj <- safe_read(f)
    k <- suppressWarnings(as.integer(stringr::str_match(basename(f), "k(\\d+)")[,2]))
    if (is.data.frame(obj) && "silhouette" %in% names(obj)) {
      tibble::tibble(k = k, scaling = "z_score", silhouette = mean(obj$silhouette, na.rm = TRUE))
    } else if (is.numeric(obj)) {
      tibble::tibble(k = k, scaling = "z_score", silhouette = mean(obj, na.rm = TRUE))
    } else if (is.list(obj)) {
      tibble::tibble(k = k, scaling = "z_score", silhouette = mean(unlist(obj), na.rm = TRUE))
    } else {
      NULL
    }
  })
}

#1.6.16 Coerce ID name
ensure_participant_id <- function(df) {
  if (is.null(df)) return(df)
  nm <- names(df)
  if ("participant_id" %in% nm) return(df)
  if ("participant_id" %in% nm) {
    names(df)[match("participant_id", nm)] <- "participant_id"
  }
  df
}

#1.6.17 Null-coalescing helper
`%||%` <- function(x, y) if (is.null(x) || (is.data.frame(x) && nrow(x) == 0)) y else x

#1.6.18 Helper for file discovery
first_existing <- function(...) {
  xs <- c(...)
  for (p in xs) if (!is.null(p) && nzchar(p) && file.exists(p)) return(p)
  return("")
}

```

## Overview and Inputs

This document compiles all manuscript-ready figures and tables for the ABCD MDS Risk project. It is designed to **knit immediately** using TK placeholder data if any source files are missing. Wherever TK placeholders appear, captions are explicitly marked *(TK)* so you can swap in final CSV/RDS files without editing code.

### Inputs configuration

- Purpose: Load all downstream inputs for figures and tables with robust TK fallbacks so the document always knits
- Inputs: CSV/RDS files under `data/data_processed/**/*` and `results/main_analysis/**/*` produced by earlier scripts
- Outputs: In-memory frames/objects (e.g., `risk_vars`, `outcome_vars`, `bd_pp`, `val_sil_*`); TK frames if files are missing
- Notes: `read_or_tk()` seeds placeholders with expected columns; stats frames are standardized for consistent downstream use

```{r input-config}

#2. Inputs Configuration
#2.1 Read in core datasets
#2.1.1 Read in risk variable data
risk_vars <- read_or_tk(
  first_existing(
    p_proc("risk_variable_data.csv"),
    p_proc("risk_variable_data.rds")),
  tk_cols = c("participant_id","variable","value","domain","stage","cluster")) %>% 
  ensure_participant_id()

#2.1.2 Read in outcome variable data
outcome_vars <- read_or_tk(
  first_existing(
    p_proc("outcome_variable_data.csv"),
    p_proc("outcome_variable_data.rds")),
  tk_cols = c("participant_id","age","sex","site","wave","bd_any")) %>% 
  ensure_participant_id()

#2.1.3 Read in cluster labels
cluster_labels <- read_or_tk(
  first_existing(
    p_analysis("cluster_labels_k2_z_score.csv"),
    p_analysis("cluster_labels_k2_z_score.rds")),
  tk_cols = c("participant_id","cluster")) %>% 
  ensure_participant_id()

#2.1.4 Read in BD panel data
bd_panel <- read_or_tk(first_existing(
  p_analysis("bd_panel_k2_z_score.csv"),
  p_analysis("bd_panel_k2_z_score.rds")), 
  tk_cols = c("participant_id","wave","age","cluster","any_bsd","bipolar_I","bipolar_II","bd_nos")) %>% 
  ensure_participant_id()

#2.1.5 Read in BD person-period data
bd_pp <- read_or_tk(first_existing(
  p_analysis("bd_person_period_k2_z_score.csv"),
  p_analysis("bd_person_period_k2_z_score.rds")), 
  tk_cols = c("participant_id","wave","age","interval_id","cluster","event_any_bsd","event_bipolar_I","event_bipolar_II","event_bd_nos")) %>% ensure_participant_id()

#2.2 Map C1/C2 cluster labels to manuscript labels (C1 = Lower; C2 = Higher)
#2.2.1 Create a coercion function
coerce_cluster <- function(x) {
  x_chr <- as.character(x)
  dplyr::recode(
    x_chr,
    "C1"="Lower-Risk","1"="Lower-Risk","low"="Lower-Risk","lower"="Lower-Risk",
    "C2"="Higher-Risk","2"="Higher-Risk","high"="Higher-Risk","higher"="Higher-Risk",
    .default = NA_character_)
}

#2.2.2 Coerce cluster labels where available
if (!is.null(cluster_labels) && nrow(cluster_labels)) {
  cluster_labels <- cluster_labels %>%
    dplyr::mutate(cluster = coerce_cluster(cluster))
}

#2.3 Read in validation results
#2.3.1 All available clustMixType derived index results
val_sil_all <- safe_read(first_existing(
  p_valid("val_z_score_silhouette.rds")
))

#2.3.2 Robust reader for per-k clustMixType validation index files (skips non-DF RDS)
val_sil_perk <- read_val_sil_perk(p_valid("partial_results_in_progress"))

#2.3.3 Optional other scalings (may be non-DF; used only if DF)
val_sil_z <- safe_read(first_existing(p_valid("val_z_score_silhouette.rds")))
val_sil_none <- safe_read(first_existing(p_valid("val_none_silhouette.rds")))
val_sil_minmax <- safe_read(first_existing(p_valid("val_min_max_silhouette.rds")))
val_sil_pct <- safe_read(first_existing(p_valid("val_percentile_silhouette.rds")))
val_sil_abs <- safe_read(first_existing(p_valid("val_max_absolute_silhouette.rds")))
val_cindex_all <- safe_read(first_existing(p_valid("val_robust_cindex.rds"), p_valid("val_robust_cindex.csv")))

#2.3.4 Read in cluster stability results
cluster_stability_raw <- read_or_tk(
  first_existing(
    p_stab("stability_z_score_k2_nb1000.csv"),
    p_stab("stability_z_score_k2_nb1000.rds")),
  tk_cols = c("metric","value","percentile","notes"))

#2.3.5 Read in UMAP embeddings
umap_embed <- read_or_tk(
  first_existing(
    p_valid("umap_embedding_k2.csv"),
    p_valid("umap_embedding_k2.rds")),
  tk_cols = c("x","y","cluster"))

#2.4 Read in BD onset results
#2.4.1 Read in all relevant BD survival analysis results with finalized glm-based filenames
bd_hazard_by_wave <- read_or_tk(
  p_bd_surv("bd_dth_hazard_by_wave.csv"),
  tk_cols = c("outcome","model","end_wave","cluster","hazard","lo","hi"))

bd_cumrisk_by_wave <- read_or_tk(
  p_bd_surv("bd_dth_cumrisk_by_wave.csv"),
  tk_cols = c("outcome","model","end_wave","cluster","cumrisk","cumrisk_lo","cumrisk_hi"))

#2.4.2 Read in predicted hazards (kept available for supplement / alternate plotting if needed)
bd_preds_wave_aligned <- read_or_tk(
  p_bd_surv("bd_dth_preds_wave_aligned.csv"),
  tk_cols = c("outcome","model","pred_mode","cluster","end_wave","age_years","pred","lower","upper"))

bd_preds_fixedwave <- read_or_tk(
  p_bd_surv("bd_dth_preds_fixedwave.csv"),
  tk_cols = c("outcome","model","pred_mode","cluster","age_years","end_wave","pred","lower","upper"))

#2.4.3 Read in survival model summaries
bd_dth_wald <- read_or_tk(
  p_bd_surv("bd_dth_wald_reference.csv"),
  tk_cols = c("outcome","model","term","estimate","se","or","ci_lo","ci_hi","p_value"))

bd_dth_cr2 <- read_or_tk(
  p_bd_surv("bd_dth_inference_CR2.csv"),
  tk_cols = c("outcome","model","se_type","term","estimate","se","df","or","ci_lo","ci_hi","p_value"))

bd_dth_hc1 <- read_or_tk(
  p_bd_surv("bd_dth_inference_HC1_participant.csv"),
  tk_cols = c("outcome","model","term","estimate","se","df","or","ci_lo","ci_hi","p_value"))

#2.4.4 Read in emmeans contrasts
bd_dth_emm_age <- read_or_tk(
  p_bd_surv("bd_dth_emmeans_contrasts_by_age.csv"),
  tk_cols = c("outcome","model","age_years","contrast","or","ci_lo","ci_hi","p_value"))

bd_dth_emm_wave <- read_or_tk(
  p_bd_surv("bd_dth_emmeans_contrasts_by_wave.csv"),
  tk_cols = c("outcome","model","end_wave","contrast","or","ci_lo","ci_hi","p_value"))

#2.4.5 Read in wave age medians
bd_wave_age_medians <- read_or_tk(
  p_bd_surv("wave_age_medians.csv"),
  tk_cols = c("end_wave","age_med"))

#2.5 Read in BD prevalence results
#2.5.1 GEE by wave, fixed effects, GLMM (where available), and TMB fixed effects results (where available)
gee_by_wave_bd <- stack_dir(
  p_bd_mix("csv"),
  pattern = "GEE_(AR1|EXC)(_INT)?_cluster_OR_by_wave\\.csv$",
  mutate_f = function(df) {
    df %>%
      janitor::clean_names() %>%
      dplyr::mutate(outcome = norm_bd_outcome(file),
        model = dplyr::case_when(
          stringr::str_detect(file, "GEE_AR1") ~ "GEE_AR1",
          stringr::str_detect(file, "GEE_EXC") ~ "GEE_EXC",
          TRUE ~ "GEE")) %>%
      dplyr::select(-file)
  }
)

#2.5.2 GEE fixed effects
gee_fixed_bd <- stack_dir(
  p_bd_mix("csv"),
  pattern = "GEE_(AR1|EXC)(_INT)?_fixed(.*)\\.csv$",
  mutate_f = function(df) {
    df %>%
      janitor::clean_names() %>%
      dplyr::mutate(outcome = norm_bd_outcome(file),
        model = dplyr::case_when(
          stringr::str_detect(file, "GEE_AR1") ~ "GEE_AR1",
          stringr::str_detect(file, "GEE_EXC") ~ "GEE_EXC",
          TRUE ~ "GEE")) %>%
      dplyr::select(-file)
  }
)

#2.5.3 GLMM fixed effects
glmm_fixed_bd <- stack_dir(
  p_bd_mix("csv"),
  pattern = "GLMM.*fixed_ORs\\.csv$",
  mutate_f = function(df) {
    df %>% janitor::clean_names() %>%
      dplyr::mutate(outcome = norm_bd_outcome(file), model = "GLMM") %>%
      dplyr::select(-file)
  }
)

#2.5.4 glmmTMB fixed effects
tmb_fixed_bd <- stack_dir(
  p_bd_mix("csv"),
  pattern = "glmmTMB.*fixed_ORs\\.csv$",
  mutate_f = function(df) {
    df %>% janitor::clean_names() %>%
      dplyr::mutate(outcome = norm_bd_outcome(file), model = "glmmTMB") %>%
      dplyr::select(-file)
  }
)

#2.5.5 GAMM age-based predictions
bd_gamm_preds <- stack_dir(
  p_bd_mix("pred_csv"),
  pattern = "_GAMM_age_predictions\\.csv$",
  mutate_f = function(df) {
    df %>% janitor::clean_names() %>%
      dplyr::mutate(outcome = norm_bd_outcome(file)) %>%
      dplyr::select(-file)
  }
)

#2.6 CONSORT counts if they exist, else are derived later during construction of CONSORT diagram
consort_counts <- read_or_tk(
  first_existing(
    p_res("consort_counts.csv"),
    p_res("main_analysis","consort_counts.csv")),
  tk_cols = c("stage","label","n","wave","cluster"))

#2.7 Standardize stats data frames
#2.7.1 Function to standardize finalized DTH stats objects
standardize_stats_df <- function(df) {
  if (is.null(df) || is_tk(df)) return(NULL)

  df <- janitor::clean_names(df)

  #2.7.1.1 Harmonize naming
  if ("se_type" %in% names(df) && !"se_method" %in% names(df)) df$se_method <- df$se_type
  if ("p_value" %in% names(df) && !"p.value" %in% names(df)) df$p.value <- df$p_value
  if ("or" %in% names(df) && !"OR" %in% names(df)) df$OR <- df$or
  if ("ci_lo" %in% names(df) && !"conf.low" %in% names(df)) df$conf.low <- df$ci_lo
  if ("ci_hi" %in% names(df) && !"conf.high" %in% names(df)) df$conf.high <- df$ci_hi

  #2.7.1.2 Backfill OR / CI if needed
  if (!"OR" %in% names(df) && "estimate" %in% names(df)) df$OR <- exp(df$estimate)

  #2.7.1.3 Ensure required columns exist
  need <- c("outcome","model","term","estimate","OR","conf.low","conf.high","p.value","se_method","link")
  for (nm in need) if (!nm %in% names(df)) df[[nm]] <- NA

  df
}

#2.7.2 Standardize BD DTH results
bd_dth_cr2_std  <- standardize_stats_df(bd_dth_cr2)
bd_dth_wald_std <- standardize_stats_df(bd_dth_wald)

#2.7.3 Combine and annotate DTH results
bd_dth_sum <- dplyr::bind_rows(
  bd_dth_cr2_std,
  bd_dth_wald_std) %>%
  dplyr::mutate(
    outcome = norm_bd_outcome(dplyr::coalesce(.data$outcome, "")),
    se_method = dplyr::coalesce(se_method, "model-based"),
    link = dplyr::case_when(
      model == "cloglog_offset" ~ "cloglog",
      TRUE ~ "logit"))

#2.7.4 Main-text DTH source = participant-clustered CR2 from primary pooled-logit model
bd_dth_main <- bd_dth_sum %>%
  dplyr::filter(model == "logit_primary", se_method == "CR2")

#2.7.5 Sensitivity DTH source
bd_dth_sens <- bd_dth_sum %>%
  dplyr::filter(!(model == "logit_primary" & se_method == "CR2"))

#2.8 Seed TK data if needed for later use
seed_if_needed <- function() {

  #2.8.1 k_sil: simple TK curve or from val_sil_perk (silhouette by k and scaling)
  if (is.null(val_sil_all) && (is.null(val_sil_perk) || !is.data.frame(val_sil_perk) || nrow(val_sil_perk) == 0)) {
    k_sil_df <- tidyr::expand_grid(k = 2:8, scaling = c("robust_iqr","z_score","none")) %>%
      dplyr::mutate(silhouette = dplyr::case_when(
        scaling == "robust_iqr" ~ dplyr::case_when(k == 2 ~ 0.48, k == 3 ~ 0.43, k == 4 ~ 0.41, TRUE ~ 0.38),
        scaling == "zscore" ~ 0.35 - 0.01*(k-2),
        TRUE ~ 0.30 - 0.015*(k-2)))
  } else if (is.data.frame(val_sil_perk) && nrow(val_sil_perk) > 0) {
    k_sil_df <- val_sil_perk
  } else {

    #2.8.2 Fallback TK if val_sil_all is present but not tidy
    k_sil_df <- tidyr::expand_grid(k = 2:8, scaling = c("robust_iqr","zscore","none")) %>%
      dplyr::mutate(silhouette = 0.4 - 0.01*(k-2))
  }
  assign("k_sil", k_sil_df, envir = .GlobalEnv)

  #2.8.3 k_wss: simple TK curve
  k_wss_df <- tidyr::expand_grid(k = 2:8, scaling = c("robust_iqr","zscore","none")) %>%
    dplyr::group_by(scaling) %>%
    dplyr::mutate(
      wss = 1000/(k-1) + dplyr::case_when(
        scaling == "robust_iqr" ~ 0,
        scaling == "zscore" ~ 50,
        TRUE ~ 100)) %>%
    dplyr::ungroup()
  assign("k_wss", k_wss_df, envir = .GlobalEnv)

  #2.8.4 UMAP: TK cloud if needed
  if (is.null(umap_embed) || is_tk(umap_embed) || !all(c("x","y","cluster") %in% names(umap_embed))) {
    set.seed(123)
    assign("umap_embed",
      tibble::tibble(
      x = c(rnorm(500,-2,1), rnorm(500, 2,1)),
      y = c(rnorm(500, 0,1), rnorm(500, 0,1)),
      cluster = rep(c("C1","C2"), each = 500)),
      envir = .GlobalEnv)
  }

  #2.8.5 CONSORT counts: TK counts if needed
  if (is.null(consort_counts) || is_tk(consort_counts)) {
    assign("consort_counts",
      tibble::tribble(
        ~stage, ~label, ~n, ~wave, ~cluster,
        "start", "ABCD baseline consented", 11878, NA, NA,
        "filter","Eligible (age/QC)", 11500, NA, NA,
        "filter","Complete risk variables", 9800, NA, NA,
        "filter","Complete outcomes & covariates", 9500, NA, NA,
        "filter","Excluded per pre-spec (22.55%)", 2676, NA, NA,
        "end", "Included in risk clustering", 9193, NA, NA),
    envir = .GlobalEnv)
  }
}

#2.9 Seed TK data if needed
seed_if_needed()

```

## 1. CONSORT / Analytic Flow (Figure 1)

In this chunk, we construct a CONSORT-style flow using the consort package. Boxes summarize attrition for the clustered sample and each analysis set (BD DTH; BD prevalence GEE/GLMM/glmmTMB). If required inputs are unavailable, TK text is shown so the document still knits.

### Figure 1 - CONSORT / Analytic Flow

- Purpose: Build CONSORT-style flow summarizing inclusion/exclusion by analysis branch and wave
- Inputs: `bd_pp`, `bd_panel`, `cluster_labels` (plus optional `consort_counts`)
- Outputs: Saves `Figure_01_CONSORT.{pdf,png}` to `results/figures`; returns a grid object (or TK bar chart if `consort` unavailable)
- Notes: Robust to missing inputs; computes counts directly from available data and flags TK placeholders in captions

```{r fig1-consort}

#F1. CONSORT / Analytic Flow
#F1.1 Establish helper functions to be used in CONSORT construction
#F1.1.1 Vector of pretty wave labels for CONSORT bullets/boxes
wave_pretty_map <- c(
  "ses-00A" = "Baseline",
  "ses-02A" = "Year-2",
  "ses-04A" = "Year-4",
  "ses-06A" = "Year-6")

#F1.1.2 Function to prettify wave labels using vector above
wave_lab <- function(w) {
  s <- as.character(w)
  out <- ifelse(s %in% names(wave_pretty_map), wave_pretty_map[s], s)
  unname(out)
}

#F1.1.3 Function to order waves in standard order
wave_order_pretty <- c("Baseline","Year-2","Year-4","Year-6")
order_waves <- function(df, col = "wave") {
  if (!col %in% names(df)) return(df)
  df %>%
    dplyr::mutate(
      !!col := factor(.data[[col]], levels = wave_order_pretty[wave_order_pretty %in% unique(.data[[col]])])
    ) %>%
    dplyr::arrange(.data[[col]]) %>%
    dplyr::mutate(!!col := as.character(.data[[col]]))
}

#F1.1.4 Helper functions to check for presence of columns in data frames
has_any <- function(df, cols) any(cols %in% names(df))
present <- function(df, cols) intersect(cols, names(df))

#F1.1.5 Function to pick wave/timepoint column from data frame
pick_wave_col <- function(df) {
  cands <- c("wave","end_wave","session_id")
  w <- intersect(cands, names(df))
  if (length(w)) w[[1]] else NA_character_
}

#F1.1.6 Row-wise NA check
row_any_na <- function(df_cols) {
  if (is.null(df_cols) || ncol(df_cols) == 0) return(rep(FALSE, ifelse(is.null(df_cols), 0L, nrow(df_cols))))
  rowSums(is.na(df_cols)) > 0
}

#F1.1.7 Keep only participant_id; if only subject_id exists, rename -> participant_id
ensure_ids <- function(df) {
  if (is.null(df)) return(df)
  nms <- names(df)
  if (!"participant_id" %in% nms && "subject_id" %in% nms) {
    names(df)[match("subject_id", nms)] <- "participant_id"
  }

  #F1.1.7.1 Drop any lingering subject_id to avoid accidental use
  df <- dplyr::select(df, -dplyr::any_of("subject_id"))
  df
}

#F1.1.7.2 Ensure IDs in all analysis data frames
bd_pp <- ensure_ids(bd_pp)
bd_panel <- ensure_ids(bd_panel)
cluster_labels <- ensure_ids(cluster_labels)

#F1.1.8 Always coalesce cluster from labels using participant_id only
ensure_cluster <- function(df) {
  if (is.null(df) || is_tk(cluster_labels)) return(df)
  if (!all(c("participant_id","cluster") %in% names(cluster_labels))) return(df)
  df %>%
    dplyr::left_join(
      cluster_labels %>% dplyr::select(participant_id, cluster_lbl = cluster),
      by = "participant_id") %>%
    dplyr::mutate(cluster = dplyr::coalesce(.data$cluster, .data$cluster_lbl)) %>%
    dplyr::select(-dplyr::any_of("cluster_lbl"))
}

#F1.1.9 flexible subset normalizer (expr / quosure / formula / string)
as_quosure_or_null <- function(expr) {
  if (is.null(expr)) return(NULL)
  if (rlang::is_quosure(expr)) return(expr)
  if (inherits(expr, "formula")) return(rlang::as_quosure(rlang::f_rhs(expr), env = rlang::caller_env()))
  if (is.character(expr) && length(expr) == 1L) return(rlang::parse_quo(expr, env = rlang::caller_env()))
  rlang::as_quosure(expr, env = rlang::caller_env())
}

#F1.1.10 Summarize flow + related functions
`%||%` <- function(x, y) if (is.null(x) || (is.data.frame(x) && nrow(x) == 0)) y else x
fmtN <- function(x) format(x, big.mark = ",")
summarize_flow <- function(df,
  outcome_cols,
  covar_candidates = c("age","age_years","age_end","age_wave","sex","sex_at_birth"),
  subset_expr = NULL,
  wave_col = NULL,
  completeness = c("union","intersection"),
  censor_after_first = FALSE) {
  completeness <- match.arg(completeness)

  #F1.1.10.1 Basic checks
  if (is_tk(df) || is.null(df) || !"participant_id" %in% names(df)) {
    return(list(
      text_available = "TK - analysis dataset not available.",
      text_details = "",
      waves = NULL,
      included_by_outcome = tibble::tibble(),
      per_outcome = tibble::tibble()))
  }
  
  dd <- ensure_cluster(ensure_ids(df))
  
  #F1.1.10.2 optional subset
  if (!is.null(subset_expr)) {
    q <- as_quosure_or_null(subset_expr)
    if (!is.null(q)) dd <- dplyr::filter(dd, !!q)
  }
  
  #F1.1.10.3 wave column
  wcol <- wave_col %||% pick_wave_col(dd)
  if (is.na(wcol)) {
    return(list(
      text_available = "TK - no wave/timepoint column detected.",
      text_details = "",
      waves = NULL,
      included_by_outcome = tibble::tibble(),
      per_outcome = tibble::tibble()))
  }
  
  #F1.1.10.4 choose covariates present
  age_var <- intersect(c("age_end","age_wave"), names(dd))
  age_var <- if (length(age_var)) age_var[[1]] else NA_character_
  sex_var <- intersect(c("sex"), names(dd))
  sex_var <- if (length(sex_var)) sex_var[[1]] else NA_character_
  
  #F1.1.10.5 event/status column (for LONG DTH)
  event_col <- intersect(c("event","status"), names(dd))
  event_col <- if (length(event_col)) event_col[[1]] else NA_character_
  
  #F1.1.10.6 detect shapes
  looks_long <- "outcome" %in% names(dd) && !is.na(event_col)
  outs_in_long <- if (looks_long) intersect(outcome_cols, unique(dd$outcome)) else character(0)
  is_long <- looks_long && length(outs_in_long) > 0

  #F1.1.10.7 detect WIDE-DTH event columns if not long
  find_event_col <- function(nm) {
    cands <- c(paste0("event_", nm), paste0("status_", nm))
    hit <- cands[cands %in% names(dd)]
    if (length(hit)) hit[[1]] else NA_character_
  }
  wide_event_map <- setNames(purrr::map_chr(outcome_cols, find_event_col), outcome_cols)
  has_wide_events <- any(!is.na(wide_event_map))

  #F1.1.10.8 DTH censoring (long only)
  if (is_long && censor_after_first) {
    ord_col <- if ("interval_index" %in% names(dd)) "interval_index" else wcol
    dd <- dd %>%
      dplyr::group_by(.data$participant_id, .data$outcome) %>%
      dplyr::arrange(.data[[ord_col]], .by_group = TRUE) %>%
      dplyr::mutate(
        first_event_flag  = dplyr::coalesce(.data[[event_col]] == 1, FALSE),
        at_or_after_first = base::cummax(first_event_flag) > 0,
        after_first = dplyr::lag(at_or_after_first, default = FALSE)) %>%
      dplyr::filter(!after_first) %>%
      dplyr::ungroup()
  }
  
  included_by_outcome <- tibble::tibble()
  
  if (is_long) {

    #F1.1.10.9 keep only requested outcomes
    dd <- dd %>% dplyr::filter(.data$outcome %in% outs_in_long)

    #F1.1.10.10 flags per id x wave
    cl_df <- dd %>%
      dplyr::group_by(.data$participant_id, .data[[wcol]]) %>%
      dplyr::summarise(no_cluster = !any(!is.na(.data$cluster)), .groups = "drop")
    
    age_df <- if (!is.na(age_var)) {
      dd %>% dplyr::group_by(.data$participant_id, .data[[wcol]]) %>%
        dplyr::summarise(miss_age = any(is.na(.data[[age_var]])), .groups = "drop")
    } else dd %>% dplyr::distinct(.data$participant_id, .data[[wcol]]) %>% dplyr::mutate(miss_age = FALSE)
    
    sex_df <- if (!is.na(sex_var)) {
      dd %>% dplyr::group_by(.data$participant_id, .data[[wcol]]) %>%
        dplyr::summarise(miss_sex = any(is.na(.data[[sex_var]])), .groups = "drop")
    } else dd %>% dplyr::distinct(.data$participant_id, .data[[wcol]]) %>% dplyr::mutate(miss_sex = FALSE)

    #F1.1.10.11 presence (non-NA event) without pivot: per id x wave x outcome
    present_long <- dd %>%
      dplyr::group_by(.data$participant_id, .data[[wcol]], .data$outcome) %>%
      dplyr::summarise(has_outcome = any(!is.na(.data[[event_col]])), .groups = "drop")

    #F1.1.10.12 any/all across requested outcomes
    present_anyall <- present_long %>%
      dplyr::group_by(.data$participant_id, .data[[wcol]]) %>%
      dplyr::summarise(
        any_present = any(.data$has_outcome),
        all_present = all(.data$has_outcome),
        .groups = "drop")
    
    #F1.1.10.13 base DF with flags
    base <- cl_df %>%
      dplyr::left_join(age_df,  by = c("participant_id", wcol)) %>%
      dplyr::left_join(sex_df,  by = c("participant_id", wcol)) %>%
      dplyr::left_join(present_anyall, by = c("participant_id", wcol)) %>%
      dplyr::mutate(
        miss_cov = miss_age | miss_sex,
        wave_pretty = wave_lab(.data[[wcol]]),
        miss_outcome = if (completeness == "union")
          !dplyr::coalesce(.data$any_present, FALSE) else
            !dplyr::coalesce(.data$all_present, FALSE))
    
    #F1.1.10.14 elig for per-outcome counts; build a single robust join key and eligibility
    base <- base %>%
      dplyr::mutate(.key = paste0(.data$participant_id, "||", .data[[wcol]]))
    present_long <- present_long %>%
      dplyr::mutate(.key = paste0(.data$participant_id, "||", .data[[wcol]]))
    
    #F1.1.10.15 eligibility DF
    elig_df <- base %>%
      dplyr::transmute(.key, .elig = !.data$no_cluster & !.data$miss_cov)

    #F1.1.10.16 Included-by-outcome counts
    included_by_outcome <- present_long %>%
      dplyr::left_join(elig_df, by = ".key") %>%
      dplyr::filter(.data$.elig & .data$has_outcome) %>%
      dplyr::mutate(wave = wave_lab(.data[[wcol]])) %>%
      dplyr::group_by(wave, outcome = .data$outcome) %>%
      dplyr::summarise(n = dplyr::n_distinct(.data$participant_id), .groups = "drop") %>%
      order_waves("wave")
    
    outcomes_list <- outs_in_long

    #F1.1.10.17 helper: get has_out for a single outcome (long) without using dynamic columns
    get_has_outcome <- function(target_outcome) {
      pw <- present_long %>%
        dplyr::filter(.data$outcome == target_outcome) %>%
        dplyr::select(.key, has_out = has_outcome)
      base %>%
        dplyr::left_join(pw, by = ".key") %>%
        dplyr::mutate(has_out = dplyr::coalesce(.data$has_out, FALSE))
    }
    
    #F1.1.10.18 diagnostic print if no included_by_outcome rows
    if (nrow(included_by_outcome) == 0L) {
      message("[diag] present_long rows: ", nrow(present_long))
      message("[diag] base rows: ", nrow(base))
      print(
        present_long %>%
          dplyr::left_join(elig_df, by = ".key") %>%
          dplyr::mutate(wave = wave_lab(.data[[wcol]])) %>%
          dplyr::count(outcome, wave, has_outcome, .elig) %>%
          dplyr::arrange(outcome, wave))
    }
    
  } else {

    #F1.1.10.19 WIDE (panel) or WIDE-DTH (event_* / status_*)
    base0 <- dd %>%
      dplyr::distinct(participant_id, .data[[wcol]], .keep_all = TRUE) %>%
      dplyr::rename(.wave = !!rlang::sym(wcol)) %>%
      dplyr::mutate(wave = wave_lab(.wave))
    
    #F1.1.10.20 two possibilities:
    #F1.1.10.20.1 classic panel wide: outcome columns exist by their names
    outs_in_wide <- intersect(outcome_cols, names(base0))

    #F1.1.10.20.2 WIDE-DTH: map outcomes to event_* / status_* columns
    wide_map <- if (length(outs_in_wide)) {
      setNames(outs_in_wide, outs_in_wide)  # identity map
    } else if (has_wide_events) {
      purrr::keep(wide_event_map, ~ !is.na(.x))
    } else {
      character(0)
    }
    
    #F1.1.10.21 compute any/all present across requested outcomes
    if (length(wide_map)) {
      sel <- dplyr::select(base0, dplyr::all_of(unname(wide_map)))

      #F1.1.10.21.1 for panel: present = !is.na(value); for event_*: present = !is.na(event col)
      mat <- as.matrix(!is.na(sel))
      any_present <- rowSums(mat) > 0
      all_present <- rowSums(!mat) == 0
    } else {
      any_present <- rep(FALSE, nrow(base0))
      all_present <- rep(FALSE, nrow(base0))
    }
    
    #F1.1.10.21.2 base DF, now with flags
    base <- base0 %>%
      dplyr::mutate(
        no_cluster = is.na(.data$cluster),
        miss_age = if (!is.na(age_var)) is.na(.data[[age_var]]) else FALSE,
        miss_sex = if (!is.na(sex_var)) is.na(.data[[sex_var]]) else FALSE,
        miss_cov = miss_age | miss_sex,
        miss_outcome = if (completeness == "union") !any_present else !all_present)

    #F1.1.10.22 per-outcome included counts
    if (length(wide_map)) {
      included_by_outcome <- purrr::map_dfr(names(wide_map), function(o) {
        col <- wide_map[[o]]
        base %>%
          dplyr::filter(!no_cluster, !miss_cov, !is.na(.data[[col]])) %>%
          dplyr::group_by(wave) %>%
          dplyr::summarise(n = dplyr::n_distinct(participant_id), .groups = "drop") %>%
          dplyr::mutate(outcome = o, .before = 1)
      })
    } else {
      included_by_outcome <- tibble::tibble(wave = character(), outcome = character(), n = integer())
    }
    
    outcomes_list <- names(wide_map)
    
    #F1.1.10.23 helper: get has_out for a single outcome (WIDE), using map to actual column
    get_has_outcome <- function(target_outcome) {
      if (!target_outcome %in% names(wide_map)) {
        return(base %>% dplyr::mutate(has_out = FALSE))
      }
      col <- wide_map[[target_outcome]]
      base %>% dplyr::mutate(has_out = !is.na(.data[[col]]))
    }
  }
  
  #F1.1.10.24 Ensure a pretty display column named 'wave' without breaking joins above
  if ("wave_pretty" %in% names(base)) {
    base <- base %>% dplyr::mutate(wave = .data$wave_pretty)
  }

  #F1.1.10.25 wave summary for overview
  wave_summ <- base %>%
    dplyr::group_by(wave) %>%
    dplyr::summarise(
      start_n = dplyr::n_distinct(participant_id),
      no_cluster = dplyr::n_distinct(participant_id[no_cluster]),
      miss_sex = dplyr::n_distinct(participant_id[miss_sex]),
      miss_age = dplyr::n_distinct(participant_id[miss_age]),
      miss_outcome = dplyr::n_distinct(participant_id[miss_outcome]),
      final_n = dplyr::n_distinct(participant_id[!no_cluster & !miss_cov & !miss_outcome]),
      .groups = "drop") %>%
    dplyr::arrange(wave) %>%
    dplyr::mutate(removed = start_n - final_n)
  wave_summ <- wave_summ %>% order_waves("wave")
  
  #F1.1.10.26 Overview box: "Available for analysis" (cluster + covars present)
  available_by_wave <- base %>%
    dplyr::group_by(wave) %>%
    dplyr::summarise(
      available_n = dplyr::n_distinct(participant_id[!no_cluster & !miss_cov]),
      .groups = "drop") %>% 
      dplyr::arrange(wave)
  
  #F1.1.10.27 Text output: available for analysis
  text_available <- paste(
    "Available for Analysis:",
    paste0("    ",
      paste0(available_by_wave$wave, ": ", format(available_by_wave$available_n, big.mark=",")),
           collapse = "\n"),
    sep = "\n")

  #F1.1.10.28 Per-outcome breakdown (by wave): included + reasons (fixed)
  per_outcome_df <- if (length(outcomes_list)) {
    purrr::map_dfr(outcomes_list, function(o) {
      df_o <- get_has_outcome(o)
      
      #F1.1.10.28.1 compute per-wave counts
      df_o %>%
        dplyr::group_by(wave) %>%
        dplyr::summarise(
          included = dplyr::n_distinct(participant_id[!no_cluster & !miss_cov &  has_out]),
          miss_outcome = dplyr::n_distinct(participant_id[!no_cluster & !miss_cov & !has_out]),
          miss_cluster = dplyr::n_distinct(participant_id[ no_cluster ]),
          miss_cov = dplyr::n_distinct(participant_id[ !no_cluster &  miss_cov ]),
          miss_sex = dplyr::n_distinct(participant_id[ miss_sex ]),
          miss_age = dplyr::n_distinct(participant_id[ miss_age ]),
          .groups = "drop") %>%
        dplyr::arrange(wave) %>%
        dplyr::mutate(outcome = o, .before = 1)
    })
  } else {
    tibble::tibble(outcome = character(), wave = character(),
      included = integer(), miss_outcome = integer(),
      miss_cluster = integer(), miss_cov = integer(),
      miss_sex = integer(), miss_age = integer())
  }
  
  #F1.1.10.29 Text output: per-outcome breakdown
  detail_lines <- purrr::map_chr(unique(base$wave), function(w) {
    rows <- per_outcome_df %>% dplyr::filter(.data$wave == w)
    if (nrow(rows) == 0) return(paste0("  ", w, ": (no requested outcomes present)"))
    inner <- paste0(
      "    ",
      paste0(
        rows$outcome, ": ",
        "included=", format(rows$included, big.mark=","),  "; ",
        "miss outcome=", format(rows$miss_outcome, big.mark=","), "; ",
        "cluster=", format(rows$miss_cluster, big.mark=","), "; ",
        "sex=", format(rows$miss_sex, big.mark=","), "; ",
        "age=", format(rows$miss_age, big.mark=",")),
      collapse = "\n")
    paste0("  ", w, ":\n", inner)
  })
  
  #F1.1.10.30 final text
  text_details <- paste(
    "Per-outcome breakdown (by wave):",
    paste0(detail_lines, collapse = "\n"),
    sep = "\n"
  )
  
  #F1.1.10.31 return everything
  list(
    text_available = text_available,
    text_details = text_details,
    waves = wave_summ,
    included_by_outcome = included_by_outcome,
    per_outcome = per_outcome_df
  )
}

#F1.2.2 Summarize flows
#F1.2.2.1 BD DTH
bd_dth_summary <- summarize_flow(
  df = bd_pp,
  outcome_cols = c("any_bsd","bipolar_I","bipolar_II","bd_nos"),
  covar_candidates = c("age_end","sex"),
  wave_col = "end_wave",
  completeness = "union",
  censor_after_first = TRUE)

#F1.2.2.2 BD prevalence
bd_prev_summary <- summarize_flow(
  df = bd_panel,
  outcome_cols = c("any_bsd","bipolar_I","bipolar_II","bd_nos"),
  covar_candidates = c("age_wave","sex"),
  completeness = "union")

#F1.2.3 Baseline N (prefer any available roster-like table; else max over known frames)
if (!exists("baseline_n", inherits = FALSE) || !is.finite(suppressWarnings(baseline_n))) {
  distinct_n <- function(df) if (!is.null(df) && "participant_id" %in% names(df)) dplyr::n_distinct(df$participant_id) else NA_integer_
  baseline_n <- suppressWarnings(max(c(
    distinct_n(risk_vars),
    distinct_n(outcome_vars),
    distinct_n(cluster_labels),
    distinct_n(bd_panel),
    distinct_n(bd_pp)
  ), na.rm = TRUE))
  if (!is.finite(baseline_n)) baseline_n <- 11878L  # ABCD baseline consented N (ceiling)
}

#F1.2.4 Clustered counts (distinct IDs; accept multiple label conventions)
norm_cluster <- function(x) {
  x <- tolower(as.character(x))
  dplyr::case_when(
    x %in% c("c1","1","high","higher","higher-risk","higher risk","hr") ~ "C1",
    x %in% c("c2","2","low","lower","lower-risk","lower risk","lr") ~ "C2",
    TRUE ~ NA_character_
  )
}

#F1.3 Get clustered counts across all data frames
#F1.3.1 function to get clustered counts
get_cluster_counts <- function() {
  cand <- list(cluster_labels, bd_panel, bd_pp)
  cl_raw <- purrr::compact(purrr::map(cand, \(d) {
    if (is.null(d)) return(NULL)
    if (!all(c("participant_id","cluster") %in% names(d))) return(NULL)
    d %>% dplyr::select(participant_id, cluster) %>% dplyr::filter(!is.na(cluster))
  })) %>% dplyr::bind_rows()
  if (nrow(cl_raw) == 0) return(list(c1 = NA_integer_, c2 = NA_integer_))
  cl_once <- cl_raw %>%
    dplyr::mutate(clu = norm_cluster(cluster)) %>%
    dplyr::filter(!is.na(clu)) %>%
    dplyr::distinct(participant_id, .keep_all = TRUE)
  list(
    c1 = sum(cl_once$clu == "C1"),
    c2 = sum(cl_once$clu == "C2"))
}

#F1.3.2 compute clustered counts
cc <- get_cluster_counts()
c1_n <- cc$c1; c2_n <- cc$c2

#F1.3.3 compute pre-cluster exclusions
inc_total <- sum(c(c1_n, c2_n), na.rm = TRUE)
excl_pre_cluster <- max(baseline_n - inc_total, 0L)

#F1.4 Pretty labels (unchanged; keep this where it is)
pretty_outcome_label <- function(x) dplyr::recode(
  x,
  "any_bsd" = "Any BSD",
  "bipolar_I" = "BD I",
  "bipolar_II" = "BD II",
  "bd_nos" = "BD NOS",
  .default = x)

#F1.5 Helper function to get body-only strings per outcome, one line per wave, with hanging indent
build_ex_inc_blocks <- function(sum_obj, wanted) {
  df <- sum_obj$per_outcome
  ex  <- rep("  (no waves)", length(wanted))
  inc <- rep("  (no waves)", length(wanted))
  names(ex) <- names(inc) <- wanted
  
  #F1.5.1 populate per-outcome blocks
  if (!is.null(df) && nrow(df) > 0) {
    for (o in intersect(wanted, unique(df$outcome))) {
      d <- df %>% dplyr::filter(.data$outcome == o) %>% dplyr::arrange(wave)
      
      #F1.5.2 build lines
      ex_lines <- sprintf("%s: Cl=%s, Cov=%s, Out=%s",
        d$wave, fmtN(d$miss_cluster),
        fmtN(d$miss_cov),
        fmtN(d$miss_outcome))
      inc_lines <- sprintf("%s: %s", d$wave, fmtN(d$included))
      
      #F1.5.3 combine with hanging indent
      ex[o] <- paste0("  ", paste(ex_lines,  collapse = "\n  "))
      inc[o] <- paste0("  ", paste(inc_lines, collapse = "\n  "))
    }
  }
  list(ex = unname(ex[wanted]), inc = unname(inc[wanted]))
}

#F1.6 Build CONSORT diagram
#F1.6.1 Define adjustable knobs for CONSORT layout
CONSORT_KNOBS <- list(
  text_cex = 0.82,    # ↑ for larger text; ↓ for smaller
  lineheight = 0.8,   # ↑ for a bit more breathing room
  width_scale = 1.15, # multiplies all text_width values
  top_width = 36,     # baseline widths (pre-scale) for top row boxes
  side_width = 34,    # side box (excluded before clustering)
  cluster_width = 34, # clustered box
  split_width = 36,   # split titles row
  avail_width = 40,   # "Available for analysis" row
  branch_width = 44)   # EXCLUDED / INCLUDED branch boxes
.w <- function(x) ceiling(x * CONSORT_KNOBS$width_scale)

#F1.6.2 Build CONSORT function
if (requireNamespace("consort", quietly = TRUE)) {

  #F1.6.2.1 Use the knobs above for fast iteration
  tiny_gp <- grid::gpar(
    cex = CONSORT_KNOBS$text_cex,
    lineheight = CONSORT_KNOBS$lineheight)
  
  #F1.6.2.2 Text for top row
  root_txt <- glue::glue(
    "ABCD Baseline Participants\nn = {format(baseline_n, big.mark=',')}")
  
  #F1.6.2.3 Text for exclusions and clustered
  excl_txt <- glue::glue(
    "Excluded before clustering\nn = {format(excl_pre_cluster, big.mark=',')}\n(Missing Risk Variables)")

  #F1.6.2.4 Text for clustered boxes
  clus_txt <- glue::glue(
    "Clustered\nN = {format(inc_total, big.mark=',')}\nHigher-Risk = {format(c1_n, big.mark=',')};  Lower-Risk = {format(c2_n, big.mark=',')}")
  
  #F1.6.2.5 Top row (use adjustable widths)
  g <- consort::add_box(txt = root_txt, text_width = .w(CONSORT_KNOBS$top_width), txt_gp = tiny_gp)
  g <- consort::add_side_box(g, txt = excl_txt, side = "right",
        text_width = .w(CONSORT_KNOBS$side_width), txt_gp = tiny_gp)
  g <- consort::add_box(g, txt = clus_txt, text_width = .w(CONSORT_KNOBS$cluster_width), txt_gp = tiny_gp)
  
  #F1.6.2.6 main analyses
  split_titles <- c(
    "BD First-Onset (DTH)",
    "BD Prevalence (GEE)")
  g <- consort::add_split(g, txt = split_titles,
        text_width = .w(CONSORT_KNOBS$split_width), txt_gp = tiny_gp)
  
  #F1.6.2.7 "Available for analysis" row
  avail_vec <- c(
    bd_dth_summary$text_available,
    bd_prev_summary$text_available)
  g <- consort::add_box(g, txt = avail_vec,
        text_width = .w(CONSORT_KNOBS$avail_width), txt_gp = tiny_gp)
  
  #F1.6.2.8 Outcomes (fixed order)
  bd_outcomes <- c("any_bsd","bipolar_I","bipolar_II","bd_nos")

  #F1.6.2.9 Build body-only blocks per outcome (reuses your helpers)
  bd_dth_blocks <- build_ex_inc_blocks(bd_dth_summary, bd_outcomes)
  bd_prev_blocks <- build_ex_inc_blocks(bd_prev_summary, bd_outcomes)
  
  #F1.6.2.10 Pretty labels helper
  pretty_outcome_label <- function(x) dplyr::recode(
    x, any_bsd="Any BSD", bipolar_I="BD-I", bipolar_II="BD-II", bd_nos="BD-NOS", .default=x)
  
  #F1.6.2.11 Build bullets per branch
  branch_bullets <- function(wanted, ex_vec, inc_vec) {
    ex_bul <- paste(purrr::map2(pretty_outcome_label(wanted), ex_vec,
                ~ paste0("• ", .x, "\n", .y)), collapse = "\n")
    inc_bul<- paste(purrr::map2(pretty_outcome_label(wanted), inc_vec,
                ~ paste0("• ", .x, "\n", .y)), collapse = "\n")
    list(
      ex = paste0("Excluded by wave:\n", ex_bul),
      inc = paste0("Included by wave:\n", inc_bul))
  }
  
  #F1.6.2.12 Get bullets per branch
  bd_dth_p <- branch_bullets(bd_outcomes, bd_dth_blocks$ex, bd_dth_blocks$inc)
  bd_prev_p <- branch_bullets(bd_outcomes, bd_prev_blocks$ex, bd_prev_blocks$inc)
  
  #F1.6.2.13 Row: EXCLUDED per branch
  g <- consort::add_box(
    g,
    txt = c(bd_dth_p$ex, bd_prev_p$ex),
    text_width = .w(CONSORT_KNOBS$branch_width),
    txt_gp = tiny_gp)

  #F1.6.2.14 Row: INCLUDED per branch
  g <- consort::add_box(
    g,
    txt = c(bd_dth_p$inc, bd_prev_p$inc),
    text_width = .w(CONSORT_KNOBS$branch_width),
    txt_gp = tiny_gp)
  
  #F1.6.2.15 Export CONSORT build
  grid_obj <- consort::build_grid(g)
  ggplot2::ggsave(file.path(fig_dir, "Figure_01_CONSORT.pdf"), plot = grid_obj,
            width = 13.3, height = 13.2, bg = "transparent")
  ggplot2::ggsave(file.path(fig_dir, "Figure_01_CONSORT.png"), plot = grid_obj,
            width = 13.1, height = 13.2, dpi = 720, bg = "transparent")
  grid_obj
  
  #F1.6.2.16 Else provide a simple bar plot fallback
} else {
  message("[WARN] Package 'consort' not available; falling back to TK bar.")
  df <- tibble::tibble(stage = c("Excluded pre-cluster","Clustered"),
                       n = c(excl_pre_cluster, inc_total))
  p_consort <- ggplot(df, aes(stage, n, fill = stage)) +
    geom_col(width = 0.6) +
    scale_fill_manual(values = c("Excluded pre-cluster"="#F1.cccccc", "Clustered"="#F1.444444")) +
    labs(title = "CONSORT (TK)", x = NULL, y = "N") +
    theme(legend.position = "none")
  save_fig(p_consort, "Figure_01_CONSORT", width = 11, height = 8.5)
  p_consort
}

```

## Table 1 - Baseline Characteristics of the Clustered Sample

- Purpose: Produce Table 1 aligned to manuscript display.
- Inputs: `data_raw/dataset.csv` for demographics; `outcome_vars` for baseline diagnoses; `cluster_labels` for cluster membership.
- Outputs: Saves `Table_01_Baseline_Characteristics` as HTML/PNG/RTF in `results/tables`; returns a gt object (fallback mini table if `gtsummary` missing).
- Notes: Requires strict raw column names (see `DEM_COLS`) and outcome variables (`DIAG_COLS`); earliest session per participant.

```{r table 1}

## Table 1: Baseline characteristics of the clustered sample ##

#T1.1 Define required raw column mappings
DEM_COLS <- list(
  id = "participant_id",
  session = "session_id",
  age = "ab_g_dyn__visit_age",
  sex = "ab_g_stc__cohort_sex"  # 1=Male, 2=Female
)

#T1.2 Function to read demographics strictly
read_demographics_strict <- function() {
  dem_path <- first_existing(
    p_data("data_raw", "dataset.csv"),
    "/home/exacloud/gscratch/NagelLab/staff/sam/projects/ABCD_MDS_Risk/data/data_raw/dataset.csv")
  dem <- safe_read(dem_path)
  need <- unlist(DEM_COLS, use.names = FALSE)
  miss <- setdiff(need, names(dem))
  if (length(miss)) stop("Missing required columns in raw dataset: ", paste(miss, collapse=", "))

  #T1.2.1 process demographics to get earliest session per participant
  dem %>%
    dplyr::select(
      participant_id = !!DEM_COLS$id,
      session_id = !!DEM_COLS$session,
      age = !!DEM_COLS$age,
      sex_code = !!DEM_COLS$sex) %>%
    dplyr::group_by(participant_id) %>%
    dplyr::arrange(session_id, .by_group = TRUE) %>%
    dplyr::slice(1) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      sex = dplyr::case_when(
        sex_code %in% c(1,"1") ~ "Male",
        sex_code %in% c(2,"2") ~ "Female",
        TRUE ~ NA_character_)) %>%
    dplyr::select(participant_id, age, sex)
}

#T1.3 List diagnosis columns of interest
DIAG_COLS <- c("any_bsd","bipolar_I","bipolar_II","bd_nos")

#T1.4 Function to get baseline outcomes from outcome_vars
baseline_outcomes <- function(ov) {
  need <- c("participant_id","session_id", DIAG_COLS)
  miss <- setdiff(need, names(ov))
  if (length(miss)) stop("Missing required columns in outcome_vars: ", paste(miss, collapse=", "))

  #T1.4.1 get baseline session data of interest per participant
  ov %>%
    dplyr::select(dplyr::all_of(need)) %>%
    dplyr::group_by(participant_id) %>%
    dplyr::arrange(session_id, .by_group = TRUE) %>%
    dplyr::slice(1) %>%
    dplyr::ungroup()
}

#T1.5 Build Table 2 dataset restricted to clustered sample
#T1.5.1 Read demographics
demographics_df <- read_demographics_strict()
ov_base <- baseline_outcomes(outcome_vars)

#T1.5.2 Merge cluster labels, demographics, and baseline outcomes
tbl2_df <- cluster_labels %>%
  dplyr::select(participant_id, cluster) %>%
  dplyr::left_join(demographics_df, by = "participant_id") %>%
  dplyr::left_join(ov_base, by = "participant_id") %>%
  dplyr::mutate(cluster = factor(cluster, levels = c("Higher-Risk","Lower-Risk")))

#T1.5.3 Create readable labels for display
label_list <- list(
  age ~ "Age (years)",
  sex ~ "Sex at birth",
  any_bsd ~ "Any BSD diagnosis",
  bipolar_I ~ "BD I diagnosis",
  bipolar_II ~ "BD II diagnosis",
  bd_nos ~ "BD NOS diagnosis")

#T1.5.4 Create table caption
tbl2_caption <- glue::glue(
  "**Table 1. Baseline characteristics of the clustered sample**  \n",
  "Participants included in clustering (n={scales::comma(dplyr::n_distinct(tbl2_df$participant_id))}).")

#T1.5.5 Function to convert binary variables to yes/no factors
to_yesno <- function(x) {
  if (is.numeric(x) || is.logical(x)) {
    factor(dplyr::case_when(
      is.na(x) ~ NA_character_,
      as.numeric(x) == 1 ~ "Yes",
      TRUE ~ "No"
    ), levels = c("No","Yes"))
  } else x
}

#T1.5.6 Prepare display dataset
tbl2_disp <- tbl2_df %>%
  dplyr::mutate(dplyr::across(dplyr::all_of(DIAG_COLS), to_yesno),
    sex = factor(sex, levels = c("Male","Female")))

#T1.6 Create gtsummary table for table 1
if (requireNamespace("gtsummary", quietly = TRUE)) {

  #T1.6.1 Build the base table with Overall column
  tbl2_gts <-
    gtsummary::tbl_summary(
      data = tbl2_disp %>% dplyr::select(cluster, age, sex, dplyr::all_of(DIAG_COLS)),
      by = cluster,
      type = list(age ~ "continuous"),
      statistic = list(
        age ~ "{mean} ({sd})",
        gtsummary::all_categorical() ~ "{n} ({p}%)"),
      missing = "no",
      label = label_list) %>%
    gtsummary::add_overall() %>%

    #T1.6.2 Clean headers to show only the level names (no N in headers)
    gtsummary::modify_header(all_stat_cols() ~ "**{level}**") %>%

    #T1.6.3 Insert a top "N" row with counts per column (Overall + each cluster)
    gtsummary::modify_table_body(~{
      tb <- .

      #T1.6.3.1 counts in the same column order as the stats columns: stat_0 = Overall, stat_1 = Higher-Risk, stat_2 = Lower-Risk (given factor levels)
      n_overall <- dplyr::n_distinct(tbl2_df$participant_id)
      n_by <- tbl2_df %>%
        dplyr::count(cluster, name = "n") %>%
        dplyr::arrange(factor(cluster, levels = c("Higher-Risk","Lower-Risk"))) %>%
        dplyr::pull(n)
      n_by <- c(n_by, rep(NA_integer_, max(0, 2 - length(n_by))))  # safety pad

      #T1.6.3.2 create N row
      n_row <- tibble::tibble(
        variable = ".N_row",
        var_type = NA_character_,
        var_label = "N",
        row_type = "label",
        label = "N",
        stat_0 = scales::comma(n_overall),
        stat_1 = scales::comma(n_by[1]),
        stat_2 = scales::comma(n_by[2]))
      dplyr::bind_rows(n_row, tb)
    }) %>%
    gtsummary::modify_caption(tbl2_caption) %>%
    gtsummary::as_gt()

  #T1.6.4 Save the gtsummary table in multiple formats
  save_gt(tbl2_gts, "Table_02_Baseline_Characteristics")
  tbl2_gts
} else {

  #T1.6.5 fallback unchanged
  mini <- tbl2_df %>%
    dplyr::count(cluster, name = "n") %>%
    dplyr::mutate(pct = 100 * n / sum(n)) %>%
    dplyr::arrange(cluster)
  mini_gt <- gt::gt(mini) %>%
    gt::fmt_number(columns = "pct", decimals = 1) %>%
    gt::cols_label(cluster = "Cluster", n = "n", pct = "Percent") %>%
    gt::tab_caption(paste0(tbl2_caption, "  (limited: gtsummary not available)."))

  #T1.6.6 Save the fallback mini gt table if necessaruy
  save_gt(mini_gt, "Table_01_Baseline_Characteristics_minimal")
  mini_gt
}

```

## Table 2 - Baseline Risk Variable Descriptives by Cluster (+ effect sizes)

- Purpose: Summarize baseline risk-variable descriptives by cluster with effect sizes
- Inputs: `risk_vars` (wide or long) and `cluster_labels`; optional pretty labels from `data/data_processed/risk_variable_labels.csv`
- Outputs: Saves `Table_02_Risk_Variables_by_Cluster` as HTML/PNG/RTF; returns a gt table
- Notes: Automatically pivots wide `risk_vars` to long; detects variable types (binary vs continuous); uses `effectsize` and `glm`

```{r table 2}

## Table 2: Risk-variable descriptive statistics by cluster (+ effect sizes) ##

#T2.1 Prepare risk variable data (long format with variable + value)
#T2.1.1 If risk_vars is wide, pivot to long; else use as-is
rv0 <- risk_vars
if (!is.null(rv0) && nrow(rv0)) {
  if (!"cluster" %in% names(rv0) && !is_tk(cluster_labels)) {
    rv0 <- rv0 %>% dplyr::left_join(cluster_labels, by = "participant_id")
  }
  rv0 <- rv0 %>%
    dplyr::filter(cluster %in% c("Lower-Risk","Higher-Risk"))

  #T2.1.1.1 Pivot wide -> long (make `variable` + `value`)
  id_cols <- c("participant_id","session_id","site","age","race_ethnicity","sex","family_id","cluster")
  risk_cols <- setdiff(names(rv0), id_cols)
  rv0 <- rv0 %>%
    tidyr::pivot_longer(
      cols = dplyr::all_of(risk_cols),
      names_to  = "variable",
      values_to = "value"
    )
} else {
  rv0 <- read_or_tk("", tk_cols = c("participant_id","variable","value","cluster"))
}

#T2.1.2 Identify variable types (binary vs continuous vs other) and mutate to correct types if not already present as such
var_types <- rv0 %>%
  dplyr::group_by(variable) %>%
  dplyr::summarise(
    n_nonmiss = sum(!is.na(value)),
    is_numeric = all(is.numeric(value) | is.integer(value)),
    uniq_nonNA = dplyr::n_distinct(value[!is.na(value)]),
    all01 = all(value[!is.na(value)] %in% c(0,1)),
    .groups = "drop") %>%
  dplyr::mutate(
    type = dplyr::case_when(
      is_numeric & all01 ~ "Binary",
      is_numeric & uniq_nonNA > 2 ~ "Continuous",
      TRUE ~ "Other"))

#T2.1.3 Filter to only binary and continuous variables with sufficient non-missing data
cont_vars <- var_types %>% dplyr::filter(type == "Continuous", n_nonmiss > 1) %>% dplyr::pull(variable)
bin_vars <- var_types %>% dplyr::filter(type == "Binary", n_nonmiss > 0) %>% dplyr::pull(variable)

#T2.2 Create helper formatting functions
fmt_m_sd <- function(x) sprintf("%.2f (%.2f)", mean(x, na.rm = TRUE), stats::sd(x, na.rm = TRUE))
fmt_n_pct <- function(n1, den) sprintf("%d (%.1f%%)", n1, 100 * n1 / den)
fmt_ci <- function(lo, hi, digits = 2) sprintf("[%.2f, %.2f]", round(lo, digits), round(hi, digits))
fmt_p <- function(p) ifelse(is.na(p), "-", ifelse(p < .001, "<.001", sprintf("%.3f", p)))

#T2.3 Create pretty labels for the table
label_map <- tryCatch({
  lb <- safe_read(p_proc("risk_variable_labels.csv"))
  if (is.data.frame(lb) && all(c("variable","label") %in% names(lb)))
    setNames(lb$label, lb$variable) else NULL
}, error = function(e) NULL)

#T2.4 Analyze continuous risk variables: Cohen's d (HR vs LR) with 95% CI via effectsize; p from Welch t-test
analyze_cont <- function(v) {
  df <- rv0 %>% dplyr::filter(variable == v, !is.na(value)) %>%
    dplyr::mutate(cluster = factor(cluster, levels = c("Higher-Risk","Lower-Risk")))

  #T2.4.1 Require both clusters present
  if (length(unique(df$cluster)) < 2) return(NULL)

  #T2.4.2 Generate grouped descriptive stats
  gstats <- df %>%
    dplyr::group_by(cluster) %>%
    dplyr::summarise(m = mean(value, na.rm = TRUE), s = stats::sd(value, na.rm = TRUE), n = dplyr::n(), .groups="drop")
  m_hr <- gstats$m[gstats$cluster=="Higher-Risk"]; s_hr <- gstats$s[gstats$cluster=="Higher-Risk"]
  m_lr <- gstats$m[gstats$cluster=="Lower-Risk"];  s_lr <- gstats$s[gstats$cluster=="Lower-Risk"]

  #T2.4.3 Effect size: Cohen's d with Hedges' correction + 95% CI
  d_obj <- effectsize::cohens_d(value ~ cluster, data = df, pooled_sd = TRUE,
            hedges.correction = TRUE, ci = 0.95)
  d_val <- as.numeric(d_obj$Cohens_d)
  d_lo <- as.numeric(d_obj$CI_low)
  d_hi <- as.numeric(d_obj$CI_high)

  #T2.4.4 Calculate p-values from Welch t-test
  p_val <- tryCatch({
    broom::tidy(stats::t.test(value ~ cluster, data = df))$p.value[1]
  }, error = function(e) NA_real_)

  #T2.4.5 Build output row
  tibble::tibble(
    Type = "Continuous",
    Variable = if (!is.null(label_map)) dplyr::recode(v, !!!label_map) else v,
    `Lower-risk` = sprintf("%.2f (%.2f)", m_lr, s_lr),
    `Higher-risk` = sprintf("%.2f (%.2f)", m_hr, s_hr),
    `Effect (95% CI)` = sprintf("d = %.2f %s", d_val, fmt_ci(d_lo, d_hi)),
    `p` = fmt_p(p_val))
}

#T2.5 Analyze binary risk variables; create a function to calculate OR (Higher vs Lower) with 95% CI via effectsize on glm, p from glm
analyze_bin <- function(v) {
  df <- rv0 %>% dplyr::filter(variable == v, !is.na(value)) %>%
    dplyr::mutate(
      value = as.integer(value),
      cluster = factor(cluster, levels = c("Lower-Risk","Higher-Risk"))) # LR = ref

  #T2.5.1 Require both clusters present
  if (length(unique(df$cluster)) < 2) return(NULL)

  #T2.5.2 Generate grouped counts
  tab <- df %>%
    dplyr::count(cluster, value) %>%
    tidyr::complete(cluster = c("Lower-Risk","Higher-Risk"), value = c(0,1), fill = list(n = 0)) %>%
    tidyr::pivot_wider(names_from = value, values_from = n, names_prefix = "v") %>%
    dplyr::arrange(cluster)

  #T2.5.3 Extract counts for display
  lr_yes <- tab$v1[tab$cluster=="Lower-Risk"];  lr_den <- rowSums(tab[tab$cluster=="Lower-Risk", c("v0","v1")])
  hr_yes <- tab$v1[tab$cluster=="Higher-Risk"]; hr_den <- rowSums(tab[tab$cluster=="Higher-Risk", c("v0","v1")])

  #T2.5.4 Fit logistic regression model to get OR + CI + p-value
  fit <- stats::glm(value ~ cluster, data = df, family = stats::binomial())

  #T2.5.5 OR + CI from effectsize
  or_es <- tryCatch(effectsize::oddsratio(fit, ci = 0.95), error = function(e) NULL)
  if (!is.null(or_es)) {

    #T2.5.5.1 Extract OR + CI
    row <- which(grepl("clusterHigher-Risk", or_es$Parameter))
    or_val <- as.numeric(or_es$OR[row]); or_lo <- as.numeric(or_es$CI_low[row]); or_hi <- as.numeric(or_es$CI_high[row])
  } else {
    #T2.5.5.2 Extract OR + CI with broom if necessary
    tr <- broom::tidy(fit, conf.int = TRUE, exponentiate = TRUE)
    or_row <- tr[tr$term=="clusterHigher-Risk",]
    or_val <- or_row$estimate; or_lo <- or_row$conf.low; or_hi <- or_row$conf.high
  }
  #T2.5.6 Get p-value from model (Wald)
  p_val <- broom::tidy(fit)$p.value[broom::tidy(fit)$term=="clusterHigher-Risk"]

  #T2.5.7 Build output row
  tibble::tibble(
    Type = "Binary",
    Variable = if (!is.null(label_map)) dplyr::recode(v, !!!label_map) else v,
    `Lower-risk` = fmt_n_pct(lr_yes, lr_den),
    `Higher-risk` = fmt_n_pct(hr_yes, hr_den),
    `Effect (95% CI)` = sprintf("OR = %.2f %s", or_val, fmt_ci(or_lo, or_hi)),
    `p` = fmt_p(p_val))
}

#T2.6 Apply analysis functions to continuous and binary variables; combine results
cont_rows <- purrr::map(cont_vars, analyze_cont) |> purrr::compact()
bin_rows <- purrr::map(bin_vars, analyze_bin) |> purrr::compact()
t2 <- dplyr::bind_rows(cont_rows, bin_rows) |>
  dplyr::arrange(factor(Type, levels = c("Continuous","Binary")), Variable)

#T2.7 Create table caption and build gt table
t2_caption <- paste0(
  "**Table 2. Baseline risk-variable descriptive statistics by cluster**  \n",
  "Continuous: mean (SD); effect size is Cohen's *d* with Hedges' correction and 95% CI (Higher-Lower). ",
  "Binary: n (% “Yes”); effect size is odds ratio (Higher vs Lower) with 95% CI. ",
  "p-values from Welch *t*-test (continuous) and logistic regression Wald test (binary).")

#T2.8 Build gt table with formatting
t2_gt <- gt::gt(t2, groupname_col = "Type") |>
  gt::tab_caption(t2_caption) |>
  gt::cols_align(align = "center", columns = c(`Lower-risk`,`Higher-risk`,`Effect (95% CI)`,`p`))

#T2.9 Save the gt table
save_gt(t2_gt, "Table_02_Risk_Variables_by_Cluster")
t2_gt

```

## Figure 4 - Baseline Risk Variable Distributions by Cluster

- Purpose: Visualize distributions of continuous (violins/boxplots) and binary (stacked bars) risk variables by cluster.
- Inputs: `rv0` and `var_types` built in Table 3; optional `label_map` for display names; significance stars derived from Table 3 p-values.
- Outputs: Saves `Figure_03_Risk_Distributions_by_Cluster.{pdf,png}`; returns a patchwork figure (continuous above, binary below).
- Notes: Facets alphabetized by display label; minimal axes theme; star annotations added where p < .05 if `ggsignif` is available.

```{r figure 3}

## Figure 3: Baseline risk-variable distributions by cluster ##

#F3.1 Define colors for clusters
cluster_cols <- c("Lower-Risk" = "#11A579", "Higher-Risk" = "#7F3C8D")

#F3.2 Create a small theme helper (left Y + bottom X; no grids) for figure 4
axes_only_theme <- function(base_size = 9, axis_text_size = 7) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_blank(),
      axis.line.x.bottom = element_line(color = "black"),
      axis.line.y.left = element_line(color = "black"),
      axis.ticks = element_line(color = "black"),
      axis.ticks.x.top = element_blank(),
      axis.ticks.y.right = element_blank(),
      axis.text = element_text(size = axis_text_size),
      legend.title = element_blank())
}

#F3.3 Identify continuous and binary variables from earlier prep
stopifnot(exists("rv0"), exists("var_types"))
cont_vars <- var_types %>% dplyr::filter(type == "Continuous", n_nonmiss > 1) %>% dplyr::pull(variable)
bin_vars <- var_types %>% dplyr::filter(type == "Binary", n_nonmiss > 0) %>% dplyr::pull(variable)

#F3.4 Create pretty labels for continuous and binary variables (fallback to hardcoded if no CSV)
#F3.4.1 Hardcoded labels for numeric vars (in case CSV not available)
labels_numeric <- c(
  "ACE_index_sum_score" = "ACE Index Sum Score",
  "mh_p_cbcl__synd__aggr_tscore" = "CBCL Aggression Problems T",
  "mh_p_cbcl__synd__attn_tscore" = "CBCL Attention Problems T",
  "mh_p_cbcl__dsm__anx_tscore" = "CBCL DSM-5 Anxiety T",
  "mh_p_cbcl__dsm__dep_tscore" = "CBCL DSM-5 Depression T",
  "le_l_coi__addr1__coi__total__national_zscore" = "COI Z-Score",
  "mh_p_gbi_sum" = "GBI Mania Sum Score",
  "fc_p_nsc__ns_mean" = "Mean Neighborhood Safety",
  "nc_y_nihtb__flnkr__uncor_score" = "NIHTB Flanker Score",
  "nc_y_nihtb__lswmt__uncor_score" = "NIHTB List Sorting Score",
  "nc_y_nihtb__pttcp__uncor_score" = "NIHTB Pattern Comparison Score",
  "sds_total" = "SDSC Sum Score",
  "mh_y_upps__nurg_sum" = "UPPS-P Negative Urgency Score",
  "mh_y_upps__purg_sum" = "UPPS-P Positive Urgency Score")

#F3.4.2 Hardcoded labels for binary vars (in case CSV not available)
labels_binary <- c(
  "bullying" = "Bullying Victimization",
  "family_history_depression" = "Family History of Depression",
  "family_history_mania" = "Family History of Mania")

#F3.4.3 Attempt to read CSV label map (overrides hardcoded if present)
label_map <- tryCatch({
  lb <- safe_read(p_proc("risk_variable_labels.csv"))
  if (is.data.frame(lb) && all(c("variable","label") %in% names(lb)))
    setNames(lb$label, lb$variable) else NULL
}, error = function(e) NULL)

#F3.4.4 Combine hardcoded + CSV label maps (CSV takes precedence)
labeller_vars <- if (!is.null(label_map)) as_labeller(label_map) else label_value

#F3.5 Extract significance stars from Table 2 p-values (if available)
stars_from_t2 <- NULL
if (exists("t2") && is.data.frame(t2) && all(c("Variable","p") %in% names(t2))) {
  
  #F3.5.1 Parse "p" which was formatted (e.g., "<.001", "0.023", "-")
  stars_from_t2 <- t2 %>%
    dplyr::transmute(
      Variable,
      p_num = dplyr::case_when(
        grepl("^<\\s*\\.?0?0*1", p) ~ 0.001,
        p == "-" ~ NA_real_,
        TRUE ~ suppressWarnings(as.numeric(p))),
      stars = dplyr::case_when(
        is.na(p_num) ~ "",
        p_num < 0.001 ~ "***",
        p_num < 0.01 ~ "***",
        p_num < 0.05 ~ "*",
        TRUE ~ "")
)
  
  #F3.5.2 If we have a label_map, make a lookup on label; otherwise join by raw var id
  if (!is.null(label_map)) {

    #F3.5.2.1 Turn raw var -> label, then join on label text in t2$Variable
    stars_from_t2 <- tibble::tibble(
      variable = names(label_map),
      Variable = unname(label_map)) %>%
      dplyr::right_join(stars_from_t2, by = "Variable") %>%
      dplyr::select(variable, stars)
  } else {
    
    #F3.5.2.2 Assume t2$Variable equals the raw var names
    stars_from_t2 <- stars_from_t2 %>%
      dplyr::rename(variable = Variable) %>%
      dplyr::select(variable, stars)
  }
}

#F3.6 Create continuous variable plot (violins + boxplots) with facets alphabetized by label
#F3.6.1 Helper to get display label for a variable
label_for <- function(v) {
  base <- if (!is.null(label_map)) label_map else labels_numeric
  out  <- unname(base[v])
  out[is.na(out)] <- v[is.na(out)]
  out
}

#F3.6.2 Build a key: which vars are present + their display labels
cont_key <- tibble::tibble(
  variable  = cont_vars,
  facet_lab = label_for(cont_vars))

#F3.6.3 Alphabetize by label (case-insensitive), use as the facet level order
cont_levels_alpha <- cont_key$facet_lab[order(tolower(cont_key$facet_lab), cont_key$facet_lab)]

#F3.6.4 Data for plotting (add the facet label column)
cont_df <- rv0 %>%
  dplyr::filter(variable %in% cont_vars,
    !is.na(value),
    cluster %in% c("Lower-Risk","Higher-Risk")) %>%
  dplyr::left_join(cont_key, by = "variable") %>%
  dplyr::mutate(
    cluster = factor(cluster, levels = c("Lower-Risk","Higher-Risk")),
    facet_lab = factor(facet_lab, levels = cont_levels_alpha))

#F3.6.5 Ranges + star positions, computed by facet label
cont_span <- cont_df %>%
  dplyr::group_by(facet_lab) %>%
  dplyr::summarise(
    y_min = min(value, na.rm = TRUE),
    y_max = max(value, na.rm = TRUE),
    .groups = "drop") %>%
  dplyr::mutate(y_lab = y_max + 0.21 * (y_max - y_min))

#F3.6.6 Bring in stars from Table 2 (still keyed by raw variable), then to facet label
cont_star_map <- if (!is.null(stars_from_t2)) {
  cont_key %>% dplyr::left_join(stars_from_t2, by = "variable") %>%
    dplyr::select(facet_lab, stars)
} else {
  tibble::tibble(facet_lab = cont_levels_alpha, stars = "")
}

#F3.6.7 Combine span + star info for annotations
cont_ann <- cont_span %>%
  dplyr::left_join(cont_star_map, by = "facet_lab")

#F3.6.8 Invisible points to grow facet headroom (needs facet_lab present in data)
cont_headroom <- tidyr::crossing(
  cont_ann[, c("facet_lab","y_lab")],
  cluster = factor(c("Lower-Risk","Higher-Risk"), levels = c("Lower-Risk","Higher-Risk")))

#F3.6.9 Plot facet by the label column; row-wise fill via dir = "h"
p_cont <- ggplot(cont_df, aes(x = cluster, y = value, fill = cluster)) +
  geom_violin(trim = FALSE, alpha = 0.5, scale = "width") +
  geom_boxplot(width = 0.18, outlier.size = 0.4) +
  geom_blank(data = cont_headroom, aes(x = cluster, y = y_lab), inherit.aes = FALSE) +
  facet_wrap(~ facet_lab, scales = "free_y", dir = "h") +
  scale_fill_manual(values = cluster_cols, drop = FALSE) +
  labs(title = "A) Numeric Risk Variable Distributions by Risk Cluster",
       x = "Cluster", y = "Value") +
  axes_only_theme(base_size = 12, axis_text_size = 9) +
  coord_cartesian(clip = "off") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.key.height = grid::unit(0.35, "lines"),
    legend.key.width  = grid::unit(0.7, "lines"),
    legend.box.spacing = grid::unit(0.2, "lines"),
    strip.text = element_text(size = 10)) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(alpha = 0.6)))

#F3.6.10 Add significance stars if ggsignif is available and any stars are present
if (requireNamespace("ggsignif", quietly = TRUE) &&
    any(nchar(cont_ann$stars %||% "") > 0)) {
  p_cont <- p_cont +
    ggsignif::geom_signif(
      data = cont_ann,
      aes(xmin = 1, xmax = 2, annotations = stars, y_position = y_lab),
      manual = TRUE, inherit.aes = FALSE,
      tip_length = 0.01, textsize = 2.8, vjust = 0.25
    )
}

#F3.7 Create binary variable plot (stacked bars) with facets alphabetized by label
#F3.7.1 Alphabetize binary variable labels
bin_levels_alpha <- names(sort(labels_binary[names(labels_binary) %in% bin_vars]))

#F3.7.2 Create data for plotting
bin_df <- rv0 %>%
  dplyr::filter(variable %in% bin_vars,
    !is.na(value),
    cluster %in% c("Lower-Risk","Higher-Risk")) %>%
  dplyr::mutate(
    cluster = factor(cluster, levels = c("Lower-Risk","Higher-Risk")),
    value = ifelse(value == 1, "Yes", "No"),
    variable = factor(variable, levels = bin_levels_alpha))

#F3.7.3 Summarize data for plotting (counts + percentages)
bin_plot_df <- bin_df %>%
  dplyr::count(variable, cluster, value) %>%
  dplyr::group_by(variable, cluster) %>%
  dplyr::mutate(pct = 100*n/sum(n)) %>%
  dplyr::ungroup()

#F3.7.4 Prepare star annotations from Table 3 (if available)
bin_ann <- bin_plot_df %>% dplyr::distinct(variable) %>% dplyr::mutate(x = 1.5, y = 103)
if (!is.null(stars_from_t2)) bin_ann <- bin_ann %>% dplyr::left_join(stars_from_t2, by = "variable") else bin_ann$stars <- ""

#F3.7.5 Build binary variable stacked bar plot with facets by variable
p_bin <- ggplot(bin_plot_df, aes(cluster, pct, fill = value)) +
  geom_col() +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5), size = 2.8) +
  facet_wrap(~ variable, labeller = as_labeller(labels_binary, default = label_value)) +
  scale_y_continuous(limits = c(0, 105), expand = expansion(mult = c(0, 0.02))) +
  scale_fill_manual(values = c("No" = "#c7d4ee", "Yes" = "#f2a1a1")) +
  labs(title = "B) Categorical Risk Variable Distributions by Risk Cluster", x = "Cluster", y = "Percent endorsed", fill = "") +
  axes_only_theme(base_size = 12, axis_text_size = 9) +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.key.height = grid::unit(0.35, "lines"),
    legend.key.width = grid::unit(0.7, "lines"),
    legend.box.spacing = grid::unit(0.2, "lines"),
    strip.text = element_text(size = 10)) +
  geom_text(
    data = subset(bin_ann, nchar(stars %||% "") > 0),
    aes(x = x, y = y, label = stars),
    inherit.aes = FALSE, size = 2.8)

#F3.8 Combine continuous + binary plots into figure 3 and save
fig3 <- p_cont / p_bin + patchwork::plot_layout(heights = c(2, 1))
save_fig(fig3, "Figure_03_Risk_Distributions_by_Cluster", width = 14, height = 11.75)
fig3

```

## Table S1: Characteristics of participants in the BD DTH onset analysis

- Purpose: Describe the BD discrete-time hazard (onset) analysis sample by wave and cluster (participants and person-periods)
- Inputs: bd_pp (person-period), including baseline_status/event flags and covariates; uses ever-onset censoring per outcome
- Outputs: Saves Table_S1_Onset_Characteristics_by_Cluster as HTML/PNG/RTF; returns a gt table
- Notes: Applies model-ready filters consistent with survival analyses; reports demographics and interval onset counts per outcome

```{r table s1}

## 6. Table S1: Characteristics of participants in the onset analysis ##

#S1.1 Prepare person-period data for onset analysis results summarization 
#S1.1.1 Ensure BD person-period data is formatted correctly
bd_pp <- bd_pp %>% 
  mutate(
    participant_id = as.character(participant_id),
    family_id = factor(as.character(family_id)),
    site_factor = factor(site_factor),
    cluster = factor(cluster),
    start_wave = factor(start_wave),
    end_wave = factor(end_wave),
    outcome = factor(outcome),
    sex = factor(sex))

#S1.1.2 Refine pre-built data using a model-ready filter that mirrors QC "post" rules from analyses
bd_pp_model <- bd_pp %>%
  filter(event %in% c(0L,1L)) %>%
  filter(!is.na(cluster)) %>%
  filter(!is.na(age_mid), !is.na(sex), !is.na(site_factor), !is.na(family_id))

#S1.1.3 Ensure cluster, wave, sex, and site reference levels are correct
cluster_levels <- levels(bd_pp$cluster)
bd_pp_model <- bd_pp_model %>%
  mutate(
    cluster = forcats::fct_relevel(cluster, "C2"),
    end_wave = forcats::fct_relevel(end_wave, "ses-02A", "ses-04A", "ses-06A"),
    sex = forcats::fct_relevel(sex, "Female"),
    site_factor = forcats::fct_relevel(site_factor, "Site_1"))

#S1.1.4 Establish interval length in years (02A~2, 04A~4, 06A~6); handle missed waves
wave_year <- function(w) {
  as.numeric(stringr::str_extract(as.character(w), "\\d+"))
}

#S1.1.5 Add interval length (years) and log(interval length)
bd_pp_model <- bd_pp_model %>%
  mutate(dt_years = pmax(wave_year(end_wave) - wave_year(start_wave), 1),
    log_dt = log(dt_years))

#S1.2 Establish baseline BD status flags and primary risk set for first onset
#S1.2.1 IMPORTANT: do this BEFORE filtering out baseline-positive so the counts are complete - collapse to one row per participant x outcome x cluster, then count baseline-positive IDs
baseline_prev_ids <- bd_pp_model %>%
  dplyr::group_by(outcome, cluster, participant_id) %>%
  dplyr::summarise(
    baseline_pos = as.integer(any(baseline_status == 1, na.rm = TRUE)),
    .groups = "drop")

#S1.2.2 Build an ID-level baseline flag using the any_bsd outcome (covers BD-I/II/NOS)
baseline_any_bsd_ids <- bd_pp_model %>% 
  dplyr::filter(outcome == "any_bsd") %>%
  dplyr::group_by(participant_id) %>%
  dplyr::summarise(
    all_na = all(is.na(baseline_status)),
    baseline_any_bsd = dplyr::if_else(
      all_na, NA_integer_, as.integer(any(baseline_status == 1L, na.rm = TRUE))),
    .groups = "drop")

#S1.2.3 Join to all outcomes, do not assume 0 for missing data
bd_pp_model <- bd_pp_model %>%
  dplyr::left_join(baseline_any_bsd_ids %>% dplyr::select(participant_id, baseline_any_bsd),
    by = "participant_id")

#S1.2.4 Filter the dataset to contain participants that had not yet had BD
bd_pp_model <- bd_pp_model %>% dplyr::filter(baseline_any_bsd == 0L)

#S1.2.5 Censor after first onset (per outcome x participant); keep all intervals up to and including the first event==1 row; drop any later rows
bd_pp_model <- bd_pp_model %>%
  arrange(outcome, participant_id,
  if ("interval_index" %in% names(.)) interval_index else as.integer(end_wave)) %>%
  group_by(outcome, participant_id) %>%
  mutate(after_onset = dplyr::lag(cummax(event == 1L), default = FALSE)) %>%
  filter(!after_onset) %>%
  ungroup() %>%
  dplyr::select(-after_onset)

#S1.2.6 Establish empirical wave-median ages for plotting/prediction
wave_age_medians <- bd_pp_model %>%
  dplyr::group_by(end_wave) %>%
  dplyr::summarise(age_med = median(age_mid, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(match(end_wave, levels(bd_pp_model$end_wave)))
ages_empirical <- wave_age_medians$age_med %>% as.numeric() %>% round(1)

#S1.3 Create helper functions for Table S1 compilation
#S1.3.1 Map session codes to time labels
time_map <- c("ses-02A" = "2y", "ses-04A" = "4y", "ses-06A" = "6y")
time_levels <- c("2y","4y","6y")  # enforce order

#S1.3.2 Descriptive stat formatting functions
fmt_mean_sd <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) return("-")
  sprintf("%.2f \u00B1 %.2f", mean(x), stats::sd(x))
}
fmt_n <- function(n) ifelse(is.na(n), "-", format(as.integer(n), big.mark = ","))
fmt_n_pct <- function(n, den) {
  out <- ifelse(is.na(den) | den == 0,
    "0 (0.0%)",
    sprintf("%s (%.1f%%)", fmt_n(n), 100 * n / den))
  as.character(out)
}

#S1.4 Compile Table S1: Characteristics of participants in the onset analysis
#S1.4.1 Ensure exists and prepare person-period data for summarization
stopifnot(exists("bd_pp_model"))
pp <- bd_pp_model %>%
  dplyr::mutate(
    cluster_disp = coerce_cluster(cluster),
    time = dplyr::recode(end_wave, !!!time_map),
    time = factor(time, levels = time_levels)) %>%
  dplyr::filter(!is.na(time), cluster_disp %in% c("Lower-Risk","Higher-Risk"))

#S1.4.2 Demographics block (filter single outcome to avoid duplication & capture any BSD)
demo_base <- pp %>% dplyr::filter(outcome == "any_bsd")

#S1.4.3 Compute counts and age stats per time x cluster
demo_counts <- demo_base %>%
  dplyr::group_by(time, cluster_disp) %>%
  dplyr::summarise(
    n_pp = dplyr::n(),
    n_part = dplyr::n_distinct(participant_id),
    age_mean_sd = fmt_mean_sd(age_mid),
    .groups = "drop")

#S1.4.4 Compile sex at birth information
sex_tab <- demo_base %>%
  dplyr::mutate(sex2 = dplyr::case_when(
    !is.na(sex) & sex %in% c("Male","Female") ~ sex,
    TRUE ~ "Other/Unknown")) %>%
  dplyr::distinct(time, cluster_disp, participant_id, sex2) %>%
  dplyr::count(time, cluster_disp, sex2, name = "n_sex") %>%
  dplyr::left_join(demo_counts %>% dplyr::select(time, cluster_disp, n_part), by = c("time","cluster_disp")) %>%
  dplyr::mutate(val = fmt_n_pct(n_sex, n_part)) %>%
  dplyr::filter(sex2 %in% c("Male","Female")) %>%
  dplyr::mutate(Row = paste0("Sex at birth - ", sex2)) %>%
  dplyr::select(time, cluster_disp, Row, val)

#S1.4.5 Compile age information
age_row <- demo_counts %>%
  dplyr::transmute(time, cluster_disp, Row = "Age in years, Mean (SD)", val = age_mean_sd)

#S1.4.6 N rows (participants + person-periods)
n_rows <- demo_counts %>%
  tidyr::pivot_longer(c(n_pp, n_part), names_to = "what", values_to = "val_raw") %>%
  dplyr::mutate(
    Row = dplyr::case_when(
      what == "n_part" ~ "N (participants)",
      what == "n_pp" ~ "N (person-periods)"),
    val = fmt_n(val_raw)) %>%
  dplyr::select(time, cluster_disp, Row, val)

#S1.4.7 Bind demographics rows
demo_rows_cs <- dplyr::bind_rows(n_rows, age_row, sex_tab) %>%
  dplyr::mutate(Section = "Demographics")

#S1.4.8 Compute totals across clusters for demographics
demo_counts_tot <- demo_base %>%
  dplyr::group_by(time) %>%
  dplyr::summarise(
    n_pp = dplyr::n(),
    n_part = dplyr::n_distinct(participant_id),
    age_mean_sd = fmt_mean_sd(age_mid),
    .groups = "drop")

#S1.4.9 Compile information about whole sample
n_rows_total <- demo_counts_tot %>%
  tidyr::pivot_longer(c(n_pp, n_part), names_to = "what", values_to = "val_raw") %>%
  dplyr::mutate(
    Row = dplyr::case_when(
      what == "n_part" ~ "N (participants)",
      what == "n_pp" ~ "N (person-periods)"),
    val = fmt_n(val_raw),
    cluster_disp = "Total") %>%
  dplyr::select(time, cluster_disp, Row, val)

#S1.4.10 Age stats for whole sample
age_row_total <- demo_counts_tot %>%
  dplyr::transmute(
    time, cluster_disp = "Total",
    Row = "Age in years, Mean (SD)",
    val = age_mean_sd)

#S1.4.11 Sex stats for whole sample
sex_tab_total <- demo_base %>%
  dplyr::mutate(sex2 = dplyr::case_when(
    !is.na(sex) & sex %in% c("Male","Female") ~ sex,
    TRUE ~ "Other/Unknown")) %>%
  dplyr::distinct(time, participant_id, sex2) %>%
  dplyr::count(time, sex2, name = "n_sex") %>%
  dplyr::left_join(demo_counts_tot %>% dplyr::select(time, n_part), by = "time") %>%
  dplyr::mutate(val = fmt_n_pct(n_sex, n_part)) %>%
  dplyr::filter(sex2 %in% c("Male","Female")) %>%
  dplyr::mutate(
    cluster_disp = "Total",
    Row = paste0("Sex at birth - ", sex2)) %>%
  dplyr::select(time, cluster_disp, Row, val)

#S1.4.12 Bind total demographics rows for whole sample
demo_rows_total <- dplyr::bind_rows(n_rows_total, age_row_total, sex_tab_total) %>%
  dplyr::mutate(Section = "Demographics")

#S1.5 Compute interval onset events per outcome
#S1.5.1 Define outcome labels
outcome_labels <- c(
  "bipolar_I" = "Interval onset - BD I, n (%)",
  "bipolar_II" = "Interval onset - BD II, n (%)",
  "bipolar_NOS" = "Interval onset - BD NOS, n (%)",
  "bd_nos" = "Interval onset - BD NOS, n (%)",
  "any_bsd" = "Interval onset - Any BSD, n (%)")

#S1.5.2 Compute events per outcome x time x cluster
events_cs <- pp %>%
  dplyr::filter(outcome %in% names(outcome_labels)) %>%
  dplyr::group_by(outcome, time, cluster_disp) %>%
  dplyr::summarise(
    n_at_risk = dplyr::n(),
    n_events = sum(event == 1L, na.rm = TRUE),
    .groups = "drop") %>%
  dplyr::mutate(Row = dplyr::recode(outcome, !!!outcome_labels),
    val = fmt_n_pct(n_events, n_at_risk),
    Section = "Interval onsets")

#S1.5.3 Compute events per outcome x time for total sample
events_total <- pp %>%
  dplyr::filter(outcome %in% names(outcome_labels)) %>%
  dplyr::group_by(outcome, time) %>%
  dplyr::summarise(
    n_at_risk = dplyr::n(),
    n_events = sum(event == 1L, na.rm = TRUE),
    .groups = "drop") %>%
  dplyr::mutate(Row = dplyr::recode(outcome, !!!outcome_labels),
    val = fmt_n_pct(n_events, n_at_risk),
    cluster_disp = "Total",
    Section = "Interval onsets")

#S1.6 Assemble final long-format table and pivot to wide for GT
#S1.6.1 Combine all rows into long format
long_all <- dplyr::bind_rows(
  demo_rows_cs, demo_rows_total,
  events_cs %>% dplyr::select(time, cluster_disp, Row, val, Section),
  events_total %>% dplyr::select(time, cluster_disp, Row, val, Section)) %>%
  dplyr::mutate(
    cluster_disp = factor(cluster_disp, levels = c("Total","Lower-Risk","Higher-Risk")),
    time = factor(time, levels = time_levels))

#S1.6.2 Complete grid to avoid missing cells
grid <- tidyr::crossing(
  Row = unique(long_all$Row),
  Section = unique(long_all$Section),
  cluster_disp = levels(long_all$cluster_disp),
  time = levels(long_all$time))

#S1.6.3 Left-join to complete grid, fill missing with em-dash
long_all <- grid %>% dplyr::left_join(long_all, by = c("Row","Section","cluster_disp","time")) %>%
  dplyr::mutate(val = dplyr::coalesce(val, "\u2014")) # em-dash for empty

#S1.6.4 Pivot to wide format for GT table
wide <- long_all %>%
  dplyr::mutate(col_key = paste0(cluster_disp, "::", time)) %>%
  dplyr::select(Section, Row, col_key, val) %>%
  tidyr::pivot_wider(names_from = col_key, values_from = val)

#S1.6.5 Select and order columns for GT table
col_keys <- c(
  paste0("Total::", time_levels),
  paste0("Lower-Risk::", time_levels),
  paste0("Higher-Risk::", time_levels))
have <- intersect(col_keys, names(wide))
wide <- wide %>% dplyr::select(Section, Row, dplyr::all_of(have))

#S1.7 Create GT table
ts1 <- gt::gt(wide, groupname_col = "Section", rowname_col = "Row") |>

  #S1.7.1 Column labels show only time; spanners show the cluster
  gt::cols_label(.list = setNames(rep(time_levels, 3), have)) |>
  gt::tab_spanner(label = "Total", columns = dplyr::all_of(paste0("Total::", time_levels))) |>
  gt::tab_spanner(label = "Lower-Risk", columns = dplyr::all_of(paste0("Lower-Risk::", time_levels))) |>
  gt::tab_spanner(label = "Higher-Risk", columns = dplyr::all_of(paste0("Higher-Risk::", time_levels))) |>
  gt::cols_align(align = "center", columns = dplyr::all_of(have)) |>
  gt::tab_caption("**Table S1. Characteristics of Participants in the Onset of Bipolar Disorders by Risk Cluster Analysis**  
Counts are shown per assessment interval (2y, 4y, 6y). Demographics are from the *any BSD* risk set (one person-period per participant per interval). Interval onset rows report events within outcome-specific risk sets (post-censoring), shown as *n (percent)* of person-periods at risk.") |>
  gt::tab_options(table.font.size = gt::px(12)) |>
  gt::fmt_missing(columns = dplyr::everything(), missing_text = "\u2014")

#S1.8 Save + print Table S1
save_gt(ts1, "Table_S1_Onset_Characteristics_by_Cluster")
ts1

```

## Figure 4 and Related Stats: Interval and Cumulative Discrete Time Hazard Model Onset by Risk Cluster Incidence Curves for Bipolar Disorders 

- Purpose: Plot discrete-time hazards and cumulative incidence by wave and cluster with CIs; emit inline stats for Results text
- Inputs: Saved `glm` fits (`glm_primary_<outcome>.rds`), `bd_pp_model` (risk set), and `bd_dth_preds` (aligned hazards) when present
- Outputs: Saves `Figure_04_DTH_Hazards_and_Cumulative_by_Cluster.{pdf,png}`; prints Year-6 RD/RR lines and age-term summaries
- Notes: Computes cumulative risk CIs via MVN draws of fixed effects (or boot fallback); overlays observed empirical hazards/cumulative

```{r figure 4 stats}

## Figure 4: Discrete-time hazards and cumulative incidence by cluster ##

#F4.1 Load necessary model fits if not already in memory
#F4.1.1 Define outcomes to plot and fit directory
OUT_FIG <- c("bipolar_I","bipolar_II","bd_nos","any_bsd")

#F4.1.2 Directory where model fits are stored
fit_dir <- res_bd_surv

#F4.1.3 Safe read helper (returns NULL on error)
safe_read_rds <- function(path) tryCatch(readRDS(path), error = function(e) NULL)

#F4.1.4 Load finalized primary glm fits if not already in memory
if (!exists("fits_primary") || !is.list(fits_primary)) fits_primary <- list()
needed <- setdiff(OUT_FIG, names(fits_primary))

#F4.1.5 Load missing fits
if (length(needed)) {
  add <- setNames(
    lapply(needed, function(o) safe_read_rds(file.path(fit_dir, paste0("glm_primary_", o, ".rds")))),
    needed
  )
  fits_primary[names(add)] <- add
}

#F4.1.6 Report loaded fits and warn if any missing
loaded_ok <- names(Filter(Negate(is.null), fits_primary))
message("Loaded primary glm fits: ", paste(loaded_ok, collapse = ", "))
if (!all(OUT_FIG %in% loaded_ok)) {
  warning("Missing primary glm fits for: ", paste(setdiff(OUT_FIG, loaded_ok), collapse = ", "))
}

#F4.1.7 Prepare reference frame and model matrix for glm predictions
make_ref_frame_glm <- function(fit, dat) {
  frm <- model.frame(fit)
  mm0 <- model.matrix(fit)
  contr_all <- attr(mm0, "contrasts")

  get_levs <- function(var) {
    if (var %in% names(frm) && is.factor(frm[[var]])) levels(frm[[var]]) else NULL
  }

  waves_lev <- get_levs("end_wave") %||% levels(dat$end_wave)
  cluster_lev <- get_levs("cluster") %||% levels(dat$cluster)
  sex_lev <- get_levs("sex")
  site_lev <- get_levs("site_factor")
  base_lev <- get_levs("baseline_status")

  ref <- expand.grid(
    end_wave = waves_lev,
    cluster = cluster_lev,
    age_mid_between = 0,
    stringsAsFactors = FALSE)

  if ("age_mid_cwc" %in% names(frm)) ref$age_mid_cwc <- 0
  if ("age_mid_cwc_orth" %in% names(frm)) ref$age_mid_cwc_orth <- 0
  if ("wave_num" %in% names(frm)) ref$wave_num <- match(ref$end_wave, waves_lev) - 1L
  if ("sex" %in% names(frm) && !is.null(sex_lev)) ref$sex <- sex_lev[1]
  if ("site_factor" %in% names(frm) && !is.null(site_lev)) ref$site_factor <- site_lev[1]
  if ("baseline_status" %in% names(frm)) {
    ref$baseline_status <- if (!is.null(base_lev)) base_lev[1] else 0
  }
  if ("log_dt" %in% names(frm) && "log_dt" %in% names(dat)) {
    log_dt_ref <- dat %>%
      dplyr::group_by(end_wave) %>%
      dplyr::summarise(log_dt = median(log_dt, na.rm = TRUE), .groups = "drop")
    ref <- ref %>% dplyr::left_join(log_dt_ref, by = "end_wave")
  }

  #F4.1.7.1 Coerce factor levels to match fitted model
  for (nm in intersect(names(ref), names(frm))) {
    if (is.factor(frm[[nm]])) ref[[nm]] <- factor(ref[[nm]], levels = levels(frm[[nm]]))
  }

  tt <- stats::delete.response(stats::terms(fit))
  X <- stats::model.matrix(tt, ref, contrasts.arg = contr_all)

  beta <- stats::coef(fit)
  V <- as.matrix(stats::vcov(fit))

  miss <- setdiff(names(beta), colnames(X))
  if (length(miss) > 0) {
    X <- cbind(X, matrix(0, nrow(X), length(miss), dimnames = list(NULL, miss)))
  }
  X <- X[, names(beta), drop = FALSE]

  list(ref = ref, X = X, beta = beta, V = V, waves = waves_lev, clus = cluster_lev)
}

#F4.1.8 Function to compute cumulative risk CIs from glm fixed-effect uncertainty
cumrisk_by_wave_ci_mvn_glm <- function(fit, dat, B = 4000, probs = c(0.025, 0.975)) {
  obj <- make_ref_frame_glm(fit, dat)
  ref <- obj$ref
  X <- obj$X
  beta <- obj$beta
  V <- obj$V

  if (requireNamespace("MASS", quietly = TRUE)) {
    beta_draws <- MASS::mvrnorm(n = B, mu = beta, Sigma = V)
  } else {
    L <- chol(V)
    Z <- matrix(stats::rnorm(B * length(beta)), nrow = B)
    beta_draws <- sweep(Z %*% t(L), 2, beta, `+`)
  }

  eta_draws <- X %*% t(beta_draws)
  p_draws <- plogis(eta_draws)

  ref$end_wave <- factor(ref$end_wave, levels = unique(as.character(ref$end_wave)))
  ref$cluster <- factor(ref$cluster, levels = unique(as.character(ref$cluster)))
  idx_by_clu <- split(seq_len(nrow(ref)), ref$cluster)

  out <- lapply(names(idx_by_clu), function(cl) {
    R <- idx_by_clu[[cl]]
    H <- p_draws[R, , drop = FALSE]
    CR <- apply(H, 2, function(col) 1 - cumprod(1 - col))
    if (is.null(dim(CR))) CR <- matrix(CR, nrow = length(R), ncol = 1)

    p_hat <- plogis(as.numeric(X[R, , drop = FALSE] %*% beta))
    cr_hat <- 1 - cumprod(1 - p_hat)

    lo <- apply(CR, 1, stats::quantile, probs = probs[1], na.rm = TRUE)
    hi <- apply(CR, 1, stats::quantile, probs = probs[2], na.rm = TRUE)

    tibble::tibble(
      end_wave = ref$end_wave[R],
      cluster = as.character(ref$cluster[R]),
      cumrisk = cr_hat,
      cumrisk_lo = lo,
      cumrisk_hi = hi)
  }) %>% 
    dplyr::bind_rows()

  out
}

#F4.1.8b Function to compute per-wave hazard CIs from glm fixed-effect uncertainty
hazard_by_wave_ci_mvn_glm <- function(fit, dat, B = 4000, probs = c(0.025, 0.975)) {
    obj <- make_ref_frame_glm(fit, dat)
    ref <- obj$ref
    X <- obj$X
    beta <- obj$beta
    V <- obj$V

    if (requireNamespace("MASS", quietly = TRUE)) {
        beta_draws <- MASS::mvrnorm(n = B, mu = beta, Sigma = V)
    } else {
        L <- chol(V)
        Z <- matrix(stats::rnorm(B * length(beta)), nrow = B)
        beta_draws <- sweep(Z %*% t(L), 2, beta, `+`)
    }

    eta_draws <- X %*% t(beta_draws)
    p_draws <- plogis(eta_draws)
    p_hat <- plogis(as.numeric(X %*% beta))

    tibble::tibble(
        end_wave = factor(ref$end_wave, levels = levels(dat$end_wave)),
        cluster = as.character(ref$cluster),
        hazard = p_hat,
        lo = apply(p_draws, 1, stats::quantile, probs = probs[1], na.rm = TRUE),
        hi = apply(p_draws, 1, stats::quantile, probs = probs[2], na.rm = TRUE)
    )
}

#F4.1.9 Plotting helpers
#F4.1.9.1 Outcome order + pretty labels
OUT_FIG <- c("bipolar_I","bipolar_II","bd_nos","any_bsd")
OUT_LABS <- c(
  bipolar_I = "BD-I",
  bipolar_II = "BD-II",
  bd_nos = "BD-NOS",
  any_bsd = "Any BSD")

#F4.1.9.2 Set the order of levels for waves
wave_levels  <- c("ses-02A","ses-04A","ses-06A")

#F4.1.9.3 Pretty wave labels + median ages for tick labels
wave_pretty <- c("ses-02A" = "Year-2",
                 "ses-04A" = "Year-4",
                 "ses-06A" = "Year-6")

#F4.1.9.4 Axis theme with only axes, no gridlines
axes_only_theme <- function(base_size = 8, axis_text = 6.5) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid = element_blank(),
      axis.line = element_blank(),
      axis.line.x.bottom = element_line(color = "black"),
      axis.line.y.left = element_line(color = "black"),
      axis.ticks = element_line(color = "black"),
      axis.ticks.x.top = element_blank(),
      axis.ticks.y.right = element_blank(),
      axis.text = element_text(size = axis_text),
      legend.position = "bottom",
      legend.box = "horizontal")
}

#F4.1.9.5 Helpers to blank axes on interior panels (use at composition time)
strip_x <- theme(axis.title.x = element_blank(),
            axis.text.x = element_blank(),
            axis.ticks.x = element_blank())
strip_y <- theme(axis.title.y = element_blank(),
            axis.text.y = element_blank(),
            axis.ticks.y = element_blank())

#F4.1.9.6 Robust cluster display helper
to_cluster_disp <- function(x) {
  x_chr <- as.character(x)

  out <- tryCatch(coerce_cluster(x_chr), error = function(e) rep(NA_character_, length(x_chr)))

  out[is.na(out) & x_chr %in% c("C1", "1")] <- "Lower-Risk"
  out[is.na(out) & x_chr %in% c("C2", "2")] <- "Higher-Risk"

  factor(out, levels = c("Lower-Risk", "Higher-Risk"))
}

#F4.2 Build hazard-by-wave directly from saved primary glm fits
haz_by_wave <- purrr::map_dfr(OUT_FIG, function(out) {
  fit <- fits_primary[[out]]
  if (is.null(fit)) return(tibble::tibble())

  dat <- bd_pp_model %>%
    dplyr::filter(outcome == out) %>%
    droplevels() %>%
    dplyr::mutate(
      end_wave = factor(end_wave, levels = wave_levels),
      cluster = factor(cluster)
    )

  if (!nrow(dat)) return(tibble::tibble())

  hazard_by_wave_ci_mvn_glm(fit, dat, B = 2000) %>%
    dplyr::mutate(
      outcome = out,
      model = "logit",
      end_wave = factor(as.character(end_wave), levels = wave_levels),
      cluster = as.character(cluster),
      cluster_disp = to_cluster_disp(cluster),
      hazard = as.numeric(hazard),
      lo = as.numeric(lo),
      hi = as.numeric(hi)
    ) %>%
    dplyr::filter(
      !is.na(cluster_disp),
      is.finite(hazard)
    ) %>%
    dplyr::select(outcome, model, end_wave, cluster, cluster_disp, hazard, lo, hi)
})

message("haz_by_wave rows: ", nrow(haz_by_wave))
print(haz_by_wave %>% dplyr::count(outcome, cluster_disp))

cum_by_wave <- if (!is_tk(bd_cumrisk_by_wave)) {
  bd_cumrisk_by_wave %>%
    janitor::clean_names() %>%
    dplyr::mutate(
      outcome = norm_bd_outcome(outcome),
      model = dplyr::coalesce(model, "logit"),
      end_wave = factor(end_wave, levels = wave_levels),
      cluster = as.character(cluster),
      cluster_disp = factor(coerce_cluster(cluster), levels = c("Lower-Risk","Higher-Risk")))
} else {
  tibble::tibble(
    outcome = character(),
    model = character(),
    end_wave = factor(character(), levels = wave_levels),
    cluster = character(),
    cluster_disp = factor(character(), levels = c("Lower-Risk","Higher-Risk")),
    cumrisk = double(),
    cumrisk_lo = double(),
    cumrisk_hi = double())
}

#F4.2.1 If cumulative-risk CIs are absent in the exported table, backfill from saved glm fits
need_cum_ci <- (
  nrow(cum_by_wave) == 0 ||
  !"cumrisk_lo" %in% names(cum_by_wave) ||
  all(is.na(cum_by_wave$cumrisk_lo)))

if (need_cum_ci) {
  cum_ci_list <- lapply(OUT_FIG, function(out) {
    fit <- fits_primary[[out]]
    if (is.null(fit)) return(NULL)

    dat <- bd_pp_model %>%
      dplyr::filter(outcome == out) %>%
      droplevels() %>%
      dplyr::mutate(
        end_wave = factor(end_wave, levels = wave_levels),
        cluster = factor(cluster))

    if (!nrow(dat)) return(NULL)

    cumrisk_by_wave_ci_mvn_glm(fit, dat, B = 2000) %>%
      dplyr::mutate(
        outcome = out,
        model = "logit",
        end_wave = factor(end_wave, levels = wave_levels),
        cluster_disp = factor(coerce_cluster(cluster), levels = c("Lower-Risk","Higher-Risk"))) %>%
      dplyr::select(outcome, model, end_wave, cluster, cluster_disp, cumrisk, cumrisk_lo, cumrisk_hi)
  })

  cum_by_wave <- dplyr::bind_rows(purrr::compact(cum_ci_list))
}

#F4.3 Prepare age median labels
#F4.3.1 Compute median ages per wave in the model data
age_meds_tbl <- bd_pp_model %>%
  dplyr::group_by(end_wave) %>%
  dplyr::summarise(age_med = median(age_mid, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(end_wave = as.character(end_wave))

#F4.3.2 Create named vector of median ages
age_meds_vec <- setNames(age_meds_tbl$age_med, age_meds_tbl$end_wave)

#F4.3.3 Create x-axis labels with median ages
x_lab_vec <- setNames(
  ifelse(is.finite(age_meds_vec[wave_levels]),
    sprintf("%s\nMedian Mid-Interval Age=%.1f", wave_pretty[wave_levels], age_meds_vec[wave_levels]),
    sprintf("%s\nMedian Mid-Interval Age=-", wave_pretty[wave_levels])),
  wave_levels)

#F4.4 Compute empirical hazards and cumulative incidence by wave and cluster
#F4.4.1 Empirical hazards by wave and cluster
emp_cum_by_wave <- bd_pp_model %>%
  dplyr::filter(outcome %in% OUT_FIG) %>%
  dplyr::mutate(
    cluster_disp = factor(coerce_cluster(cluster), levels = c("Lower-Risk","Higher-Risk")),
    end_wave = factor(end_wave, levels = wave_levels)) %>%
  
  #F4.4.1.1 Per-interval empirical hazard among those at risk
  dplyr::group_by(outcome, cluster_disp, end_wave) %>%
  dplyr::summarise(h = mean(event == 1L, na.rm = TRUE), .groups = "drop") %>%

  #F4.4.1.2 Roll up to cumulative incidence: 1 - Π(1 - h_j)
  dplyr::group_by(outcome, cluster_disp) %>%
  dplyr::arrange(end_wave, .by_group = TRUE) %>%
  dplyr::mutate(cumrisk_obs = 1 - cumprod(1 - pmin(pmax(h, 0), 1))) %>%
  dplyr::ungroup()

#F4.5 Plot predicted and observed hazard by wave and cluster
plot_haz_wave <- function(df, out_key, panel_letter = NULL) {

  d <- df %>%
    dplyr::filter(outcome == out_key, model == "logit") %>%
    dplyr::mutate(
      end_wave = factor(as.character(end_wave), levels = wave_levels),
      cluster_disp = factor(cluster_disp, levels = c("Lower-Risk", "Higher-Risk"))
    ) %>%
    dplyr::arrange(cluster_disp, end_wave) %>%
    dplyr::filter(!is.na(cluster_disp), is.finite(hazard))

  if (is.null(panel_letter)) {
    i <- match(out_key, OUT_FIG)
    panel_letter <- LETTERS[i]
  }

  has_ci <- nrow(d) > 0 &&
    all(c("lo", "hi") %in% names(d)) &&
    any(is.finite(d$lo) & is.finite(d$hi), na.rm = TRUE)

  emp <- emp_cum_by_wave %>%
    dplyr::filter(outcome == out_key) %>%
    dplyr::mutate(
      end_wave = factor(as.character(end_wave), levels = wave_levels),
      cluster_disp = factor(cluster_disp, levels = c("Lower-Risk", "Higher-Risk"))
    )

  ggplot() +
    
    #F4.5.1 Add hazard CI ribbon behind the model-estimated lines
    { if (has_ci) geom_ribbon(
      data = d,
      aes(
        x = end_wave,
        ymin = lo,
        ymax = hi,
        fill = cluster_disp,
        group = cluster_disp
      ),
      inherit.aes = FALSE,
      alpha = 0.15,
      color = NA
    ) } +
    geom_line(
      data = d,
      aes(x = end_wave, y = hazard, color = cluster_disp, group = cluster_disp),
      linewidth = 0.7,
      na.rm = TRUE
    ) +
    geom_point(
      data = d,
      aes(x = end_wave, y = hazard, color = cluster_disp),
      size = 1.4,
      na.rm = TRUE
    ) +
    geom_point(
      data = emp,
      aes(x = end_wave, y = h, color = cluster_disp),
      inherit.aes = FALSE,
      shape = 1,
      size = 1.6
    ) +
    scale_x_discrete(labels = x_lab_vec, drop = FALSE) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    scale_color_manual(
      values = cluster_cols,
      breaks = names(cluster_cols),
      labels = names(cluster_cols),
      drop = FALSE
    ) +
    scale_fill_manual(
      values = cluster_cols,
      breaks = names(cluster_cols),
      labels = names(cluster_cols),
      guide = "none",
      drop = FALSE
    ) +
    labs(
      title = paste0("(", panel_letter, ")  ", OUT_LABS[[out_key]], " Interval Hazard"),
      x = "Wave (Interval End)",
      y = "First-Onset Hazard",
      color = "Cluster"
    ) +
    axes_only_theme()
}

#F4.6 Plot cumulative incidence by wave and cluster
plot_cum_wave <- function(df, out_key) {
  d <- df %>% dplyr::filter(.data$outcome == out_key)
  has_ci <- all(c("cumrisk_lo","cumrisk_hi") %in% names(d)) &&
    any(is.finite(d$cumrisk_lo) & is.finite(d$cumrisk_hi), na.rm = TRUE)
  i <- match(out_key, OUT_FIG)
  panel_letter <- LETTERS[i + 4]
  
  #F4.6.1 Observed cumulative for each outcome
  emp <- emp_cum_by_wave %>% dplyr::filter(outcome == out_key)
  
  #F4.6.2 Plot cumulative incidence by wave and cluster
  ggplot(d, aes(x = end_wave, y = cumrisk, color = cluster_disp, group = cluster_disp)) +
    { if (has_ci) geom_ribbon(aes(ymin = cumrisk_lo, ymax = cumrisk_hi, fill = cluster_disp),
      alpha = 0.15, color = NA) } +
      geom_line(size = 0.9) +
      geom_point(size = 1.8) +
      
      #F4.6.3 Overlay observed empirical cumulative incidence
      geom_point(data = emp,
                aes(x = end_wave, y = cumrisk_obs, color = cluster_disp),
                shape = 1, size = 1.8, inherit.aes = FALSE) +

    #F4.6.4 Scales and labels
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, 0.30)) +
      scale_x_discrete(drop = FALSE, labels = x_lab_vec) +
      scale_color_manual(values = cluster_cols, breaks = names(cluster_cols), labels = names(cluster_cols)) +
      scale_fill_manual(values = cluster_cols, guide = "none") +
      labs(title = paste0("(", panel_letter, ")  ", OUT_LABS[[out_key]], " Cumulative Incidence"),
          x = "Wave (Interval End)", y = "Cumulative Risk", color = "Cluster") +
      axes_only_theme()
}

#F4.7 Build combined by wave and cumulative plot panels
#F4.7.1 Generate hazard and cumulative panels for each outcome
haz_panels <- lapply(seq_along(OUT_FIG), function(i) {
    plot_haz_wave(haz_by_wave, OUT_FIG[i], panel_letter = LETTERS[i])
  })
cum_panels <- lapply(OUT_FIG, function(o) plot_cum_wave(cum_by_wave, o))

#F4.7.2 Keep legend only on the first hazard panel
haz_panels[2:4] <- lapply(haz_panels[2:4], function(p) p + theme(legend.position = "none"))

#F4.7.3 Hide all legends on cumulative panels
cum_panels <- lapply(cum_panels, function(p) p + theme(legend.position = "none"))

#F4.7.4 Strip interior axes (2x2 grids for hazards and cumulative)
haz_grid <- (haz_panels[[1]] + strip_x) + (haz_panels[[2]] + strip_x) +
  (haz_panels[[3]]) + (haz_panels[[4]]) +
  patchwork::plot_layout(ncol = 2)

#F4.7.5 Hide legends on cumulative grid interior panels
cum_grid <- (cum_panels[[1]] + strip_x) + (cum_panels[[2]] + strip_x) +
  (cum_panels[[3]]) + (cum_panels[[4]]) +
  patchwork::plot_layout(ncol = 2)
cum_grid <- cum_grid & theme(legend.position = "none")

#F4.8 Combine hazard and cumulative grids into final figure 5, print, and save
fig4 <- haz_grid / cum_grid +
  patchwork::plot_layout(heights = c(1, 1), guides = "collect") &
  theme(legend.position = "bottom", legend.justification = "center")
print(fig4)
save_fig(fig4, "Figure_04_DTH_Hazards_and_Cumulative_by_Cluster", width = 9.5, height = 11)


## In Line Stats Related to BD DTH Analyses (Figure 4) ##

#F4.9 Helper functions for in-line stats
#F4.9.1 Create reference frame for predictions from fitted primary glm model
make_ref_frame <- function(fit, dat) {
    make_ref_frame_glm(fit = fit, dat = dat)
}

#F4.9.2 Cumulative risk from hazards helper
cumrisk_from_haz <- function(h, k) {
  h <- pmin(pmax(h, 1e-8), 1 - 1e-8)
  1 - exp(sum(log1p(-h[seq_len(k)])))
}

#F4.9.3 Wald CI helper
wald_ci <- function(theta_hat, grad, V, level = 0.95, log_scale = FALSE) {
  
  #F4.9.3.1 Small ridge in case of near-singular V
  if (any(!is.finite(V))) V[!is.finite(V)] <- 0
  V <- V + diag(1e-10, nrow(V))
  se <- sqrt(drop(t(grad) %*% V %*% grad))
  z  <- qnorm((1 + level)/2)
  if (log_scale) {
    lo <- theta_hat * exp(-z * se)
    hi <- theta_hat * exp(+z * se)
  } else {
    lo <- theta_hat - z * se
    hi <- theta_hat + z * se
  }
  list(se = se, lo = lo, hi = hi, z = theta_hat / se, p = 2 * pnorm(-abs(theta_hat / se)))
}

#F4.9.4 Year-6 cumulative risk difference and ratio from primary glm hazards
year6_rd_rr_delta <- function(fit, dat, wave_key = "ses-06A") {
  obj <- make_ref_frame(fit, dat)
  ref <- obj$ref; X <- obj$X; beta <- obj$beta; V <- obj$V
  waves <- obj$waves

  if (!wave_key %in% waves) wave_key <- tail(waves, 1)
  k_idx <- match(wave_key, waves)

  #F4.9.4.1 Map C1/C2 -> Higher/Lower using coerce_cluster()
  cl_levels <- levels(ref$cluster)
  disp <- coerce_cluster(cl_levels)
  role <- setNames(rep(NA_character_, length(cl_levels)), cl_levels)
  role[disp == "Higher-Risk"] <- "HR"
  role[disp == "Lower-Risk"] <- "LR"
  if (anyNA(role)) {
    
    #F4.9.4.2 Fallback: choose HR as the one with larger point CR
    point_cr <- sapply(cl_levels, function(cl) {
      rows <- which(ref$cluster == cl)
      h <- plogis(as.numeric(X[rows, , drop = FALSE] %*% beta))
      cumrisk_from_haz(h, k_idx)
    })
    role[] <- "LR"; role[names(which.max(point_cr))] <- "HR"
  }

  #F4.9.4.3 Functions to compute cumulative risk by cluster
  make_f <- function(cl) {
    rows <- which(ref$cluster == cl)
    function(b) {
      h <- plogis(as.numeric(X[rows, , drop = FALSE] %*% b))
      cumrisk_from_haz(h, k_idx)
    }
  }

  #F4.9.4.4 Cumulative risks, RD, RR, and their CIs
  f_HR <- make_f(names(role)[role == "HR"])
  f_LR <- make_f(names(role)[role == "LR"])
  CR_HR <- f_HR(beta); CR_LR <- f_LR(beta)
  CR_HR <- pmin(pmax(CR_HR, 1e-8), 1 - 1e-8)
  CR_LR <- pmin(pmax(CR_LR, 1e-8), 1 - 1e-8)
  RD <- CR_HR - CR_LR
  RR <- CR_HR / CR_LR
  g_HR <- numDeriv::grad(f_HR, beta)
  g_LR <- numDeriv::grad(f_LR, beta)
  g_RD  <- g_HR - g_LR
  ci_RD <- wald_ci(RD, g_RD, V, level = 0.95, log_scale = FALSE)

  #F4.9.4.5 Log-RR & CI
  g_logRR  <- g_HR / CR_HR - g_LR / CR_LR
  se_logRR <- sqrt(drop(t(g_logRR) %*% V %*% g_logRR))
  z <- qnorm(0.975)
  RR_lo <- RR * exp(-z * se_logRR)
  RR_hi <- RR * exp(+z * se_logRR)
  p_logRR <- 2 * pnorm(-abs(log(RR) / se_logRR))

  #F4.9.4.6 Return results
  list(
    wave = wave_key,
    CR_HR = CR_HR, CR_LR = CR_LR,
    RD = RD, RD_lo = ci_RD$lo, RD_hi = ci_RD$hi, RD_p = ci_RD$p,
    RR = RR, RR_lo = RR_lo, RR_hi = RR_hi, RR_p = p_logRR)
}

#F4.10 Compute Year-6 cumulative risk differences and ratios by outcome and store in a table
cum6 <- lapply(OUT_FIG, function(out) {
  fit <- fits_primary[[out]]; if (is.null(fit)) return(NULL)
  dat <- bd_pp_model %>%
    dplyr::filter(outcome == out) %>%
    droplevels() %>%
    dplyr::mutate(
        
        #F4.10.1 Ensure factors have correct levels
        end_wave = factor(end_wave, levels = wave_levels),
        cluster = factor(cluster),
        sex = factor(sex),
        site_factor = factor(site_factor))

  #F4.10.2 Compute Year-6 cumulative risk RD/RR
  res <- year6_rd_rr_delta(fit, dat, wave_key = "ses-06A")
  tibble::tibble(
    outcome = out,
    out_lab = OUT_LABS[out],
    wave = res$wave,
    HR = res$CR_HR, LR = res$CR_LR,
    RD = res$RD, RD_lo = res$RD_lo, RD_hi = res$RD_hi, RD_p = res$RD_p,
    RR = res$RR, RR_lo = res$RR_lo, RR_hi = res$RR_hi, RR_p = res$RR_p)
    }) %>% 
    dplyr::bind_rows()

#F4.11 Clean output table for in-line reporting
cum6_lines <- cum6 %>%
  mutate(
    line = glue::glue(
      "{out_lab} Year-6 cumulative risk: ",
      "{scales::percent(HR, 0.1)} (HR) vs {scales::percent(LR, 0.1)} (LR); ",
      "RD = {scales::percent(RD, 0.1)} [{scales::percent(RD_lo, 0.1)}, {scales::percent(RD_hi, 0.1)}], p = {sprintf('%.3f', RD_p)}; ",
      "RR = {sprintf('%.2f', RR)} [{sprintf('%.2f', RR_lo)}, {sprintf('%.2f', RR_hi)}], p = {sprintf('%.3f', RR_p)}")) %>% 
      pull(line)

#F4.12 Print cumulative risk difference/ratio lines
cat(paste("•", cum6_lines), sep = "\n")

#F4.13 Extract cluster and age effects for in-line reporting
#F4.13.1 Create a robust outcome-key helper for DTH summary tables
to_bd_outcome_key <- function(x) {
  key <- gsub("[^a-z0-9]", "", tolower(as.character(x)))

  dplyr::case_when(
    key %in% c("bdi", "bipolari") ~ "bipolar_I",
    key %in% c("bdii", "bipolarii") ~ "bipolar_II",
    key %in% c("bdnos", "bipolarnos") ~ "bd_nos",
    key %in% c("bsd", "anybsd", "anybipolarspectrumdisorder") ~ "any_bsd",
    TRUE ~ as.character(x))
}

#F4.13.2 Create a robust outcome-label helper for DTH summary tables
to_bd_out_lab <- function(x) {
  key <- to_bd_outcome_key(x)

  dplyr::case_when(
    key == "bipolar_I" ~ "BD-I",
    key == "bipolar_II" ~ "BD-II",
    key == "bd_nos" ~ "BD-NOS",
    key == "any_bsd" ~ "Any BSD",
    TRUE ~ as.character(x))
}

#F4.13.3 Create a helper to format p-values for inline text
fmt_p_inline <- function(p) {
  dplyr::case_when(
    is.na(p) ~ "-",
    p < .001 ~ "< .001",
    TRUE ~ paste0("= ", sprintf("%.3f", p)))
}

#F4.13.4 Determine whether the robust SE column is named se_method or se_type
se_col <- if ("se_method" %in% names(bd_dth_sum)) {
  "se_method"
} else if ("se_type" %in% names(bd_dth_sum)) {
  "se_type"
} else {
  stop("Could not find either 'se_method' or 'se_type' in bd_dth_sum.")
}

#F4.14 Primary cluster main effect = adjusted pooled interval-onset OR (Higher- vs Lower-Risk)
cluster_effects_primary <- bd_dth_sum %>%
  dplyr::filter(
    model == "logit_primary",
    .data[[se_col]] == "CR2",
    grepl("^cluster", term),
    !grepl(":", term)) %>%
  dplyr::mutate(
    outcome_key = to_bd_outcome_key(outcome),
    out_lab = to_bd_out_lab(outcome),
    OR = dplyr::coalesce(OR, exp(estimate)),
    OR_lo = dplyr::coalesce(ci_lo, exp(estimate - 1.96 * se)),
    OR_hi = dplyr::coalesce(ci_hi, exp(estimate + 1.96 * se)),
    pv = p.value) %>%
  dplyr::arrange(match(outcome_key, OUT_FIG))

cluster_lines_primary <- cluster_effects_primary %>%
  dplyr::transmute(
    line = glue::glue(
      "{out_lab}: primary adjusted Higher- vs Lower-Risk cluster OR = ",
      "{sprintf('%.2f', OR)} [{sprintf('%.2f', OR_lo)}, {sprintf('%.2f', OR_hi)}], ",
      "p {fmt_p_inline(pv)}.")) %>%
  dplyr::pull(line)

if (length(cluster_lines_primary) > 0) {
  cat(paste("•", cluster_lines_primary), sep = "\n")
  cat("\n")
}

#F4.15 Primary age term = between-person age only
age_effects_primary <- bd_dth_sum %>%
  dplyr::filter(
    model == "logit_primary",
    .data[[se_col]] == "CR2",
    term == "age_mid_between") %>%
  dplyr::mutate(
    outcome_key = to_bd_outcome_key(outcome),
    out_lab = to_bd_out_lab(outcome),
    OR = dplyr::coalesce(OR, exp(estimate)),
    pv = p.value,
    lbl = "between-person age") %>%
  dplyr::arrange(match(outcome_key, OUT_FIG))

age_lines_primary <- age_effects_primary %>%
  dplyr::transmute(
    line = glue::glue(
      "{out_lab}: primary {lbl} OR = {sprintf('%.2f', OR)} ",
      "(p {fmt_p_inline(pv)}).")) %>%
  dplyr::pull(line)

if (length(age_lines_primary) > 0) {
  cat(paste("•", age_lines_primary), sep = "\n")
}

#F4.15.1 Sensitivity note for within-person orthogonalized age
age_effects_within <- bd_dth_sum %>%
  dplyr::filter(
    model == "logit_within_orth",
    .data[[se_col]] == "CR2",
    term == "age_mid_cwc_orth") %>%
  dplyr::mutate(
    outcome_key = to_bd_outcome_key(outcome),
    out_lab = to_bd_out_lab(outcome),
    OR = dplyr::coalesce(OR, exp(estimate)),
    pv = p.value,
    lbl = "within-person age (orthogonalized sensitivity)") %>%
  dplyr::arrange(match(outcome_key, OUT_FIG))

if (nrow(age_effects_within) > 0) {
  age_lines_within <- age_effects_within %>%
    dplyr::transmute(
      line = glue::glue(
        "{out_lab}: sensitivity {lbl} OR = {sprintf('%.2f', OR)} ",
        "(p {fmt_p_inline(pv)}).")) %>%
    dplyr::pull(line)

  cat("\n")
  cat(paste("•", age_lines_within), sep = "\n")
}

```

## Table S2: Characteristics of Participants in the Prevalence of Bipolar Disorders Over Time by Risk Cluster Analysis

- Purpose: Describe the GEE prevalence sample by wave and cluster; report demographics and occurrence n (%) per outcome and totals
- Inputs: bd_panel (panel format) with status, age, sex, site, cluster, wave, outcome; produces bd_panel_model via model-ready filters
- Outputs: Saves Table_S2_Prevalence_Characteristics_by_Cluster as HTML/PNG/RTF; prints a gt table
- Notes: Denominator for occurrence rows excludes missing status; demographics taken from Any BSD panel per wave; clusters displayed as Higher-/Lower-Risk

```{r table s2}

## Table S2 - Prevalence (GEE) analysis characteristics ##

#S2.1 Prepare the BD panel dataset
#S2.1.1 Coerce key variables
bd_panel <- bd_panel %>%
  mutate(
    participant_id = as.character(participant_id),
    family_id = factor(as.character(family_id)),
    site_factor = factor(site_factor),
    cluster = factor(cluster),
    wave = factor(wave),
    outcome = factor(outcome),
    sex = factor(sex))

#S2.1.2 Require outcome responses to exist and be 0/1
bd_panel <- bd_panel %>% mutate("status" := as.integer(.data[["status"]]))

#S2.1.3 Model-ready filter (post-QC) data used for BD prevalence analyses
bd_panel_model <- bd_panel %>%
  filter(.data[["status"]] %in% c(0L,1L)) %>%
  filter(!is.na(cluster),
    !is.na(sex),
    !is.na(site_factor),
    !is.na(family_id),
    !is.na(wave)) %>%
  
  #S2.1.3.1 Require age fields present for prediction/interpretation
  filter(!is.na(age_wave), !is.na(age_wave_cwc), !is.na(age_wave_between))

#S2.1.4 Set relevant reference levels of factor variables
bd_panel_model <- bd_panel_model %>%
  mutate(
    cluster = forcats::fct_relevel(cluster, "C2"),
    wave = forcats::fct_relevel(wave, "ses-02A", "ses-04A", "ses-06A"),
    sex = forcats::fct_relevel(sex, "Female"),
    site_factor = forcats::fct_relevel(site_factor, "site01"))

#S2.1.5 Guards
#S2.1.5.1 Guard: ensure bd_panel_model exists and is a data frame
stopifnot(exists("bd_panel_model"), is.data.frame(bd_panel_model))

#S2.1.5.2 Guard: response var (default to "status" if params$response_var not present)
status_var <- if (exists("params") && !is.null(params$response_var)) params$response_var else "status"
stopifnot(status_var %in% names(bd_panel_model))

#S2.1.5.3 Again force outcomes to {0,1,NA}
bd_panel_model[[status_var]] <- as.integer(bd_panel_model[[status_var]])
bd_panel_model[[status_var]][!bd_panel_model[[status_var]] %in% c(0L,1L)] <- NA_integer_

#S2.1.6 Helper functions for Table S2
#S2.1.6.1 Set wave -> time labels (2y/4y/6y)
time_map <- c("ses-02A" = "2y", "ses-04A" = "4y", "ses-06A" = "6y")
time_levels <- c("2y","4y","6y")

#S2.1.6.2 Formatting helpers for descriptive stats
fmt_mean_sd <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) return("\u2014")
  sprintf("%.2f \u00B1 %.2f", mean(x), stats::sd(x))
}
fmt_n <- function(n) ifelse(is.na(n), "\u2014", format(as.integer(n), big.mark = ","))
fmt_n_pct <- function(n, den) {
  out <- ifelse(is.na(den) | den == 0,
    "0 (0.0%)",
    sprintf("%s (%.1f%%)", fmt_n(n), 100 * n / den))
  as.character(out)
}

#S2.1.7 Prepare panel data for Table S2
panel <- bd_panel_model %>%
  dplyr::mutate(
    cluster_disp = coerce_cluster(cluster),
    time = dplyr::recode(as.character(wave), !!!time_map),
    time = factor(time, levels = time_levels),
    sex = as.character(sex)) %>%
  dplyr::filter(!is.na(time), cluster_disp %in% c("Lower-Risk","Higher-Risk"))

#S2.2 Prepare table S2
#S2.2.1 Base dataset for demographics (any_bsd outcome)
demo_base <- panel %>% dplyr::filter(outcome == "any_bsd")

#S2.2.2 Establish N person-periods, N participants, Age
demo_counts <- demo_base %>%
  dplyr::group_by(time, cluster_disp) %>%
  dplyr::summarise(
    n_pp = dplyr::n(),
    n_part = dplyr::n_distinct(participant_id),
    age_mean_sd = fmt_mean_sd(age_wave),
    .groups = "drop")

#S2.2.3 Compute age stats
age_row <- demo_counts %>%
  dplyr::transmute(time, cluster_disp, Row = "Age in years, Mean (SD)", val = age_mean_sd)

#S2.2.4 Compute overall N rows and sex information
n_rows <- demo_counts %>%
  tidyr::pivot_longer(c(n_pp, n_part), names_to = "what", values_to = "val_raw") %>%
  dplyr::mutate(
    Row = dplyr::case_when(
      what == "n_part" ~ "N (participants)",
      what == "n_pp" ~ "N (person-periods)"),
    val = fmt_n(val_raw)) %>%
  dplyr::select(time, cluster_disp, Row, val)
sex_tab <- demo_base %>%
  dplyr::mutate(sex2 = dplyr::case_when(
    !is.na(sex) & sex %in% c("Male","Female") ~ sex,
    TRUE ~ "Other/Unknown")) %>%
  dplyr::distinct(time, cluster_disp, participant_id, sex2) %>%
  dplyr::count(time, cluster_disp, sex2, name = "n_sex") %>%
  dplyr::left_join(demo_counts %>% dplyr::select(time, cluster_disp, n_part), by = c("time","cluster_disp")) %>%
  dplyr::mutate(val = fmt_n_pct(n_sex, n_part)) %>%
  dplyr::filter(sex2 %in% c("Male","Female")) %>%
  dplyr::mutate(Row = paste0("Sex at birth - ", sex2)) %>%
  dplyr::select(time, cluster_disp, Row, val)

#S2.2.5 Combine demographic rows
demo_rows_cs <- dplyr::bind_rows(n_rows, age_row, sex_tab) %>%
  dplyr::mutate(Section = "Demographics")

#S2.2.6 Prepare total (across clusters) demographic rows
demo_counts_tot <- demo_base %>%
  dplyr::group_by(time) %>%
  dplyr::summarise(
    n_pp = dplyr::n(),
    n_part = dplyr::n_distinct(participant_id),
    age_mean_sd = fmt_mean_sd(age_wave),
    .groups = "drop")

#S2.2.7 Compute total N rows across whole sample
n_rows_total <- demo_counts_tot %>%
  tidyr::pivot_longer(c(n_pp, n_part), names_to = "what", values_to = "val_raw") %>%
  dplyr::mutate(
    Row = dplyr::case_when(
      what == "n_part" ~ "N (participants)",
      what == "n_pp" ~ "N (person-periods)"),
    val = fmt_n(val_raw),
    cluster_disp = "Total") %>%
  dplyr::select(time, cluster_disp, Row, val)

#S2.2.8 Compute total age row across whole sample
age_row_total <- demo_counts_tot %>%
  dplyr::transmute(time, cluster_disp = "Total", Row = "Age in years, Mean (SD)", val = age_mean_sd)

#S2.2.9 Compute total sex information across whole sample
sex_tab_total <- demo_base %>%
  dplyr::mutate(sex2 = dplyr::case_when(
    !is.na(sex) & sex %in% c("Male","Female") ~ sex,
    TRUE ~ "Other/Unknown")) %>%
  dplyr::distinct(time, participant_id, sex2) %>%
  dplyr::count(time, sex2, name = "n_sex") %>%
  dplyr::left_join(demo_counts_tot %>% dplyr::select(time, n_part), by = "time") %>%
  dplyr::mutate(val = fmt_n_pct(n_sex, n_part)) %>%
  dplyr::filter(sex2 %in% c("Male","Female")) %>%
  dplyr::mutate(cluster_disp = "Total", Row = paste0("Sex at birth - ", sex2)) %>%
  dplyr::select(time, cluster_disp, Row, val)

#S2.2.10 Combine total demographic rows across whole sample
demo_rows_total <- dplyr::bind_rows(n_rows_total, age_row_total, sex_tab_total) %>%
  dplyr::mutate(Section = "Demographics")

#S2.2.11 Prepare outcome occurrence rows
#S2.2.11.1 Outcome labels
occ_labels <- c(
  "bipolar_I" = "Occurrence - BD I, n (%)",
  "bipolar_II" = "Occurrence - BD II, n (%)",
  "bd_nos" = "Occurrence - BD NOS, n (%)",
  "any_bsd" = "Occurrence - Any BSD, n (%)")

#S2.2.11.2 Occurrence rows by cluster
occ_cs <- panel %>%
  dplyr::filter(outcome %in% names(occ_labels)) %>%
  dplyr::group_by(outcome, time, cluster_disp) %>%
  dplyr::summarise(
    n_obs = sum(!is.na(.data[[status_var]])),
    n_cases = sum(.data[[status_var]] == 1L, na.rm = TRUE),
    .groups = "drop") %>%
  dplyr::mutate(
    Row = dplyr::recode(outcome, !!!occ_labels),
    val = fmt_n_pct(n_cases, n_obs),
    Section = "Occurrence (wave prevalence)") %>%
  dplyr::select(time, cluster_disp, Row, val, Section)

#S2.2.11.3 Occurrence rows total across whole sample
occ_total <- panel %>%
  dplyr::filter(outcome %in% names(occ_labels)) %>%
  dplyr::group_by(outcome, time) %>%
  dplyr::summarise(
    n_obs = sum(!is.na(.data[[status_var]])),
    n_cases = sum(.data[[status_var]] == 1L, na.rm = TRUE),
    .groups = "drop") %>%
  dplyr::mutate(
    Row = dplyr::recode(outcome, !!!occ_labels),
    val = fmt_n_pct(n_cases, n_obs),
    cluster_disp = "Total",
    Section = "Occurrence (wave prevalence)") %>%
  dplyr::select(time, cluster_disp, Row, val, Section)

#S2.2.12 Combine all rows into long format
long_all <- dplyr::bind_rows(
  demo_rows_cs, demo_rows_total,
  occ_cs, occ_total) %>%
  dplyr::mutate(
    cluster_disp = factor(cluster_disp, levels = c("Total","Lower-Risk","Higher-Risk")),
    time = factor(time, levels = time_levels))

#S2.2.13 Create full grid of rows x sections x cluster_disp x time, fill missing with em-dash
grid <- tidyr::crossing(
  Row = unique(long_all$Row),
  Section = unique(long_all$Section),
  cluster_disp = levels(long_all$cluster_disp),
  time = levels(long_all$time))

#S2.3 Build wide table for GT
#S2.3.1 Left-join long_all to grid and fill missing data with em-dash
long_all <- grid %>%
  dplyr::left_join(long_all, by = c("Row","Section","cluster_disp","time")) %>%
  dplyr::mutate(val = dplyr::coalesce(val, "\u2014"))

#S2.3.2 Pivot to wide format with col_keys as cluster_disp::time
wide <- long_all %>%
  dplyr::mutate(col_key = paste0(cluster_disp, "::", time)) %>%
  dplyr::select(Section, Row, col_key, val) %>%
  tidyr::pivot_wider(names_from = col_key, values_from = val)

#S2.3.3 Select columns in desired order (if present)
#S2.3.3.1 Desired column keys
col_keys <- c(
  paste0("Total::", time_levels),
  paste0("Lower-Risk::", time_levels),
  paste0("Higher-Risk::", time_levels))

#S2.3.3.2 Keep only columns that exist in wide
have <- intersect(col_keys, names(wide))

#S2.3.3.3 Select only desired columns
wide <- wide %>% dplyr::select(Section, Row, dplyr::all_of(have))

#S2.4 Create GT table
#S2.4.1 Build GT table with spanners and labels
tS2 <- gt::gt(wide, groupname_col = "Section", rowname_col = "Row") |>
  gt::cols_label(.list = setNames(rep(time_levels, 3), have)) |>
  gt::tab_spanner(label = "Total", columns = dplyr::all_of(paste0("Total::", time_levels))) |>
  gt::tab_spanner(label = "Lower-Risk", columns = dplyr::all_of(paste0("Lower-Risk::", time_levels))) |>
  gt::tab_spanner(label = "Higher-Risk",columns = dplyr::all_of(paste0("Higher-Risk::", time_levels))) |>
  gt::cols_align(align = "center", columns = dplyr::all_of(have)) |>
  gt::tab_caption("**Table S2. Characteristics of Participants in the Prevalence of Bipolar Disorders Over Time by Risk Cluster (GEE Analysis)**
Counts are shown per assessment wave (2y, 4y, 6y). Demographics are derived from the *Any BSD* panel at each wave (one row per participant per wave). Occurrence rows report wave prevalence (status==1) within outcome-specific panels, shown as *n (percent)* of observed person-periods (denominator excludes missing).") |>
  gt::tab_options(table.font.size = gt::px(12)) |>
  gt::fmt_missing(columns = dplyr::everything(), missing_text = "\u2014")

#S2.4.2 Save table S2 (with fallback if save_gt not available)
if (exists("save_gt")) {
  save_gt(tS2, "Table_S2_Prevalence_Characteristics_by_Cluster")
} else {
  out_file <- file.path(getwd(), "Table_S2_Prevalence_Characteristics_by_Cluster.html")
  gt::gtsave(tS2, out_file)
  message("Saved: ", out_file)
}

#S2.4.3 Print GT table for Table S2
tS2

```

## Figure 6: Bipolar Disorder Odds by Risk Cluster Over Time stats

- Purpose: Produce per-wave Higher vs Lower cluster odds ratios (with 95% CI, p) from saved GEE interaction models
- Inputs: *_GEE_EXC_INT.rds models for outcomes (BD I/II/NOS/Any BSD); bd_panel_model used only for wave median age labels
- Outputs: Prints inline lines for manuscript
- Notes: Uses emmeans contrasts on the link scale and robust vcov from geeglm; truncates long CIs with arrows to share a common x-axis

```{r figure 6 stats}

## Figure 6: Bipolar Disorder Odds by Risk Cluster Over Time ##

#F6.1 Setup
#F6.1.1 Directory for saved GEE fits
fit_dir <- "/home/exacloud/gscratch/NagelLab/staff/sam/projects/ABCD_MDS_Risk/results/main_analysis/2_bd_mixed_logit/rds"

#F6.1.2 Directory for each outcome's GEE EXC interaction fit
fit_paths <- c(
  bipolar_I = file.path(fit_dir, "bipolar_I_GEE_EXC_INT.rds"),
  bipolar_II = file.path(fit_dir, "bipolar_II_GEE_EXC_INT.rds"),
  bd_nos = file.path(fit_dir, "bd_nos_GEE_EXC_INT.rds"),
  any_bsd = file.path(fit_dir, "any_bsd_GEE_EXC_INT.rds"))

#F6.1.3 Create a modified safe read function for Figure 6
safe_read <- function(p) tryCatch(readRDS(p), error = function(e) NULL)

#F6.1.4 Establish median age labels per wave
age_meds <- try({
  bd_panel_model %>%
    dplyr::group_by(wave) %>%
    dplyr::summarise(age_med = median(age_wave, na.rm = TRUE), .groups="drop") %>%
    dplyr::transmute(wave = as.character(wave), lab = sprintf("%s (median age %.1f)",
      dplyr::recode(wave, !!!wave_pretty), age_med))
}, silent = TRUE)

#F6.1.5 Fallback if age_meds computation fails
age_meds <- if (inherits(age_meds, "try-error")) tibble::tibble(wave = names(wave_pretty), lab = wave_pretty[names(wave_pretty)]) else age_meds

#F6.2 Helper: compute emmeans odds ratios (Higher- vs Lower-Risk) by wave from each GEE interaction model
em_or_by_wave <- function(fit) {

  #F6.2.1 Use robust vcov from the model (vcov.geeglm is robust by default) to compute emmeans
  V <- try(vcov(fit), silent = TRUE)
  emm <- emmeans::emmeans(fit, ~ cluster | wave, type = "link", vcov. = if (!inherits(V, "try-error")) V else NULL)
  
  #F6.2.2 Force contrast to be C2 - C1 (so OR = odds_C2 / odds_C1) regardless of reference
  lev <- levels(emm@grid$cluster)
  if (!all(c("C1","C2") %in% lev)) stop("Cluster levels not found in fit: ", paste(lev, collapse=", "))
  
  #F6.2.3 Compute contrast and summary
  w <- setNames(numeric(length(lev)), lev); w["C2"] <- 1; w["C1"] <- -1
  cmp <- emmeans::contrast(emm, method = list("C2 - C1" = w))
  sm  <- summary(cmp, infer = TRUE)  # on link (logit) scale
  
  #F6.2.4 Return tidy tibble with OR, CI, p per wave
  tibble::tibble(
    wave = as.character(sm$wave),
    OR = exp(sm$estimate),
    CI_low = exp(sm$lower.CL),
    CI_high= exp(sm$upper.CL),
    p = sm$p.value)
}

#F6.3 Load each GEE fit, compute emmeans OR contrasts, and combine into a single data frame
out_labs <- OUT_LABS
gee_pairs <- purrr::imap_dfr(fit_paths, function(fp, oc) {
  fit <- safe_read(fp)
  if (is.null(fit)) return(NULL)
  em_or_by_wave(fit) %>%
    dplyr::mutate(
      outcome = oc,
      out_lab = out_labs[oc],
      wave_pretty = dplyr::recode(wave, !!!wave_pretty)) %>%
    dplyr::left_join(age_meds, by = c("wave" = "wave")) %>%
    dplyr::mutate(wave_lab = ifelse(!is.na(lab), lab, wave_pretty))
})

#F6.3.1 Check if any results were computed, stop and provide a message why if not
if (!nrow(gee_pairs)) stop("Could not load any *_GEE_EXC_INT.rds fits or compute emmeans contrasts.")

#F6.4 Generate in-line stats for manuscript reporting
fmt_p <- function(p) ifelse(is.na(p), "-", ifelse(p < .001, "<.001", sprintf("%.3f", p)))
fmt_ci <- function(lo, hi) sprintf("%s-%s", sprintf("%.2f", lo), sprintf("%.2f", hi))
inline_lines <- gee_pairs %>%
  dplyr::mutate(line = glue("{out_lab} {wave_pretty}: OR={sprintf('%.2f', OR)}, 95% CI {fmt_ci(CI_low, CI_high)}, p={fmt_p(p)}")) %>%
  dplyr::arrange(factor(out_lab, levels = out_labs), factor(wave, levels = names(wave_pretty))) %>%
  dplyr::pull(line)

#F6.4.1 Print inline lines
cat("Figure 6 in-line stats (Higher- vs Lower-Risk, emmeans, per wave):\n",
    paste0(" • ", inline_lines, collapse = "\n"), "\n\n")

```
