---
title: "Results Compiler - ABCD MDS Risk"
author: "Sam A. Sievertsen"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    code_folding: hide
params:
  seed: 123
  link_primary: "logit"
  wave_ref: "ses-04A"
---

```{r global, include = FALSE}

# Set global env variables
knitr::opts_chunk$set(fig.width = 8, fig.height = 5, dpi = 300, fig.align = "center", echo = TRUE, cache = FALSE, message = TRUE, warning = TRUE, results = "markup", verbose = TRUE, comment = "")

```

```{r setup, include=FALSE}

#1. Script Setup
#1.1 Load in dependent packages from renv
#1.1.1 Create a function to load libraries safely
safe_library <- function(pkg) {
  suppressWarnings({
    suppressPackageStartupMessages({
      tryCatch({
        library(pkg, character.only = TRUE)
      }, error = function(e) {
        message(sprintf("[WARN] Package '%s' not available. Some features may be skipped.", pkg))
      })
    })
  })
}

#1.1.2 Load required packages
pkgs <- c(
  "ggplot2", "dplyr", "tidyr", "stringr", "tibble", "clustMixType", "here",
  "glue", "janitor", "readr", "broom", "broom.mixed", "gt", "gtsummary",
  "kableExtra", "patchwork", "cowplot", "ggtext", "scales", "mgcv", "geepack",
  "lme4", "glmmTMB", "clubSandwich", "DHARMa", "performance", "consort",
  "ggrepel", "GGally", "sf", "viridisLite", "webshot2", "purrr",
  "DiagrammeR", "DiagrammeRsvg", "rsvg", "emmeans", "effectsize", "png",
  "xgboost", "shapviz"
)

#1.1.3 Load all packages safely
invisible(lapply(pkgs, safe_library))

#1.2 Set global options and themes
#1.2.1 Set seed for reproducibility
set.seed(123)

#1.2.2 Set table style to JAMA-like if gtsummary is available
if (requireNamespace("gtsummary", quietly = TRUE)) {
  gtsummary::theme_gtsummary_journal(journal = "jama")
}

#1.2.3 Set ggplot2 text default size
theme_set(theme_minimal(base_size = 12))

#1.3 Establish paths to relevant directories & files as they exist on ARC
#1.3.1 Project root (assumes Rmd is at repo root)
root_dir <- here::here()

#1.3.2 Data roots
data_root <- here::here("data")
data_processed <- here::here("data", "data_processed")
data_raw <- here::here("data", "data_raw")

#1.3.3 Common processed subtrees
analysis_datasets <- here::here("data", "data_processed", "analysis_datasets")
kproto_results <- here::here("data", "data_processed", "kproto_results")
stability_results <- here::here("data", "data_processed", "stability_results")
validation_results<- here::here("data", "data_processed", "validation_results")

#1.3.4 Results (model outputs) roots
results_root <- here::here("results")
results_main <- here::here("results", "main_analysis")
res_bd_surv <- here::here("results", "main_analysis", "2_statistical_analysis", "1_bd_survival")
res_bd_mixed <- here::here("results", "main_analysis", "2_statistical_analysis", "2_bd_mixed_logit")
res_clustering <- here::here("results", "main_analysis", "1_clustering")
res_feat_imp <- here::here("results", "main_analysis", "1_clustering", "4_feature_importance")
res_mvfs <- here::here("results", "main_analysis", "1_clustering", "5_parsimonious_feature_selection")
res_mvfs_k <- here::here("results", "main_analysis", "1_clustering", "6_mvfs_k_calculation")
res_imp_kproto <- here::here("results", "main_analysis", "1_clustering", "7_imputation_kproto")
res_alg_bench <- here::here("results", "main_analysis", "1_clustering", "8_algorithm_benchmark")
res_comorb <- here::here("results", "main_analysis", "2_statistical_analysis", "5_comorbidity_sensitivity_analysis")
res_utility <- here::here("results", "main_analysis", "2_statistical_analysis", "6_risk_vars_vs_cluster_utility")

#1.3.5 Assets (for Figure 0 schematic, etc.)
assets_dir <- here::here("assets")

#1.3.6 Export targets (per §5 filenames)
fig_dir <- here::here("results", "figures")
tab_dir <- here::here("results", "tables")

#1.4 Ensure export directories exist
for (d in c(fig_dir, tab_dir)) if (!dir.exists(d)) dir.create(d, recursive = TRUE)

#1.5 Set default color palette for clusters
cluster_cols <- c("#7F3C8D", "#11A579")

#1.6 Helper functions to be used throughout
#1.6.1 Save figure (PDF + PNG) with transparent background
save_fig <- function(plot, base_filename, width = 9, height = 8, dpi = 720) {
  pdf_path <- file.path(fig_dir, paste0(base_filename, ".pdf"))
  png_path <- file.path(fig_dir, paste0(base_filename, ".png"))
  ggplot2::ggsave(pdf_path, plot = plot, width = width, height = height, bg = "transparent")
  ggplot2::ggsave(png_path, plot = plot, width = width, height = height, dpi = dpi, bg = "transparent")
  invisible(list(pdf = pdf_path, png = png_path))
}

#1.6.2 Format CI labels
ci_label <- function(lo, hi, digits = 2) sprintf("(%.2f, %.2f)", round(lo, digits), round(hi, digits))

#1.6.3 Format OR with CI
fmt_or <- function(est, lo, hi, digits = 2) sprintf("%.2f %s", round(est, digits), ci_label(lo, hi, digits))

#1.6.4 Pretty p-value
strip_p <- function(p) ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

#1.6.5 Patchwork legend bottom helper
legend_bottom <- function(p) p + patchwork::plot_layout(guides = "collect") & theme(legend.position = "bottom")

#1.6.6 Panel tag helper
panel_tag <- function(p, tag = "A") p + ggplot2::labs(tag = tag) + theme(plot.tag = element_text(face = "bold", size = 12))

#1.6.7 GT styled table save helper (HTML/PNG/RTF) with fallbacks
save_gt <- function(gt_tbl, base_filename) {
  html_path <- file.path(tab_dir, paste0(base_filename, ".html"))
  png_path <- file.path(tab_dir, paste0(base_filename, ".png"))
  rtf_path <- file.path(tab_dir, paste0(base_filename, ".rtf"))

  gt::gtsave(gt_tbl, html_path)
  if (requireNamespace("webshot2", quietly = TRUE)) {

    #1.6.7.1 On headless ARC nodes, webshot2 should still work; PNG save is best-effort
    try(gt::gtsave(gt_tbl, png_path), silent = TRUE)
  }
  try(gt::gtsave(gt_tbl, rtf_path), silent = TRUE)
  invisible(list(html = html_path, png = png_path, rtf = rtf_path))
}

#1.6.8 Safe read: CSV or RDS; if missing, return NULL
safe_read <- function(path) {
  if (file.exists(path)) {
    ext <- tolower(tools::file_ext(path))
    if (ext == "rds") return(readRDS(path))
    if (ext %in% c("csv", "tsv")) return(readr::read_csv(path, show_col_types = FALSE))
  }
  return(NULL)
}

#1.6.9 Read or TK placeholder by expected columns
read_or_tk <- function(path, tk_cols, tk_n = 10) {
  df <- safe_read(path)
  if (!is.null(df)) return(df)
  out <- tibble::tibble(.rows = tk_n)
  for (nm in tk_cols) out[[nm]] <- NA
  out <- janitor::remove_empty(out, which = c("rows", "cols"))
  attr(out, "TK") <- TRUE
  out
}

#1.6.10 Small utility function to detect TK data or not 
is_tk <- function(df) isTRUE(isTRUE(attr(df, "TK")))

#1.6.11 CONSORT helper to safely label TK captions
tk_caption <- function(base) paste0(base, ifelse(grepl("TK", base, fixed=TRUE) ||
  grepl("TK", paste(capture.output(str(base)), collapse=" "), fixed=TRUE),
  " (TK)", ""))

#1.6.12 Convenience: quick path joiners for later sections. These keep the Inputs chunk simple when pointing to ARC outputs
p_data <- function(...) file.path(data_root, ...)
p_proc <- function(...) file.path(data_processed, ...)
p_analysis <- function(...) file.path(analysis_datasets, ...)
p_kproto <- function(...) file.path(kproto_results, ...)
p_stab <- function(...) file.path(stability_results, ...)
p_valid <- function(...) file.path(validation_results, ...)
p_res <- function(...) file.path(results_root, ...)
p_main <- function(...) file.path(results_main, ...)
p_bd_surv <- function(...) file.path(res_bd_surv, ...)
p_bd_mix <- function(...) file.path(res_bd_mixed, ...)
p_fi <- function(...) file.path(res_feat_imp, ...)
p_mvfs <- function(...) file.path(res_mvfs, ...)
p_mvfs_k <- function(...) file.path(res_mvfs_k, ...)
p_imp <- function(...) file.path(res_imp_kproto, ...)
p_bench <- function(...) file.path(res_alg_bench, ...)
p_comorb <- function(...) file.path(res_comorb, ...)
p_utility <- function(...) file.path(res_utility, ...)

#1.6.13 Normalize BD outcome names to {BDI, BDII, BSD, BDNOS} (if present)
norm_bd_outcome <- function(x) {
  dplyr::case_when(
    stringr::str_detect(x, stringr::regex("bipolar[_-]?ii", ignore_case = TRUE)) ~ "BDII",
    stringr::str_detect(x, stringr::regex("bipolar[_-]?i(?!i)", ignore_case = TRUE)) ~ "BDI",
    stringr::str_detect(x, stringr::regex("any[_-]?bsd|\\bbd?s?d\\b", ignore_case = TRUE)) ~ "BSD",
    stringr::str_detect(x, stringr::regex("bd[_-]?nos", ignore_case = TRUE)) ~ "BDNOS",
    TRUE ~ x)
}

#1.6.14 Safer directory stacker function: only stack CSVs (or DFs); skip list/vec RDS
stack_dir <- function(dir, pattern, mutate_f = NULL) {
  if (!dir.exists(dir)) return(NULL)
  fs <- list.files(dir, pattern = pattern, full.names = TRUE)
  if (!length(fs)) return(NULL)
  out <- purrr::map(fs, function(f) {
    df <- safe_read(f)

    #1.6.14.1 Skip non-data-frame objects
    if (is.null(df) || !is.data.frame(df)) return(NULL)
    df$file <- basename(f)
    df})
  out <- purrr::compact(out)
  if (!length(out)) return(NULL)
  res <- dplyr::bind_rows(out)
  if (!is.null(mutate_f)) res <- mutate_f(res)
  res
}

#1.6.15 Helper to read val_z_score_silhouette_k*.rds into a tidy DF if possible
read_val_sil_perk <- function(dir) {
  if (!dir.exists(dir)) return(NULL)
  fs <- list.files(dir, pattern = "^val_z_score_silhouette_k[0-9]+\\.rds$", full.names = TRUE)
  if (!length(fs)) return(NULL)
  purrr::map_dfr(fs, function(f) {
    obj <- safe_read(f)
    k <- suppressWarnings(as.integer(stringr::str_match(basename(f), "k(\\d+)")[,2]))
    if (is.data.frame(obj) && "silhouette" %in% names(obj)) {
      tibble::tibble(k = k, scaling = "z_score", silhouette = mean(obj$silhouette, na.rm = TRUE))
    } else if (is.numeric(obj)) {
      tibble::tibble(k = k, scaling = "z_score", silhouette = mean(obj, na.rm = TRUE))
    } else if (is.list(obj)) {
      tibble::tibble(k = k, scaling = "z_score", silhouette = mean(unlist(obj), na.rm = TRUE))
    } else {
      NULL
    }
  })
}

#1.6.16 Coerce ID name
ensure_participant_id <- function(df) {
  if (is.null(df)) return(df)
  nm <- names(df)
  if ("participant_id" %in% nm) return(df)
  if ("participant_id" %in% nm) {
    names(df)[match("participant_id", nm)] <- "participant_id"
  }
  df
}

#1.6.17 Null-coalescing helper
`%||%` <- function(x, y) if (is.null(x) || (is.data.frame(x) && nrow(x) == 0)) y else x

#1.6.18 Helper for file discovery
first_existing <- function(...) {
  xs <- c(...)
  for (p in xs) if (!is.null(p) && nzchar(p) && file.exists(p)) return(p)
  return("")
}

#1.6.19 Canonical pretty labels for the 17 clustering risk variables
risk_label_map_imp <- c(
  "mh_p_cbcl__dsm__dep_tscore" = "CBCL DSM-5 Depression T",
  "mh_p_cbcl__dsm__anx_tscore" = "CBCL DSM-5 Anxiety T",
  "mh_p_cbcl__synd__attn_tscore" = "CBCL Attention Problems T",
  "mh_p_cbcl__synd__aggr_tscore" = "CBCL Aggression Problems T",
  "mh_p_gbi_sum" = "GBI Mania Sum Score",
  "mh_y_upps__nurg_sum" = "UPPS-P Negative Urgency Score",
  "mh_y_upps__purg_sum" = "UPPS-P Positive Urgency Score",
  "sds_total" = "SDSC Total Score",
  "ACE_index_sum_score" = "ACE Index Sum Score",
  "family_history_depression" = "Family History of Depression",
  "family_history_mania" = "Family History of Mania",
  "bullying" = "Bullying Victimization",
  "fc_p_nsc__ns_mean" = "Perceived Neighborhood Safety",
  "le_l_coi__addr1__coi__total__national_zscore" = "Child Opportunity Index National z Score",
  "nc_y_nihtb__lswmt__uncor_score" = "NIHTB List Sorting Working Memory",
  "nc_y_nihtb__flnkr__uncor_score" = "NIHTB Flanker Inhibitory Control",
  "nc_y_nihtb__pttcp__uncor_score" = "NIHTB Pattern Comparison Processing Speed")

```

## Overview and Inputs

This document compiles all manuscript-ready figures and tables for the ABCD MDS Risk project. It is designed to **knit immediately** using TK placeholder data if any source files are missing. Wherever TK placeholders appear, captions are explicitly marked *(TK)* so you can swap in final CSV/RDS files without editing code.

### Inputs configuration

- Purpose: Load all downstream inputs for figures and tables with robust TK fallbacks so the document always knits
- Inputs: CSV/RDS files under `data/data_processed/**/*` and `results/main_analysis/**/*` produced by earlier scripts
- Outputs: In-memory frames/objects (e.g., `risk_vars`, `outcome_vars`, `bd_pp`, `val_sil_*`); TK frames if files are missing
- Notes: `read_or_tk()` seeds placeholders with expected columns; stats frames are standardized for consistent downstream use

```{r input-config}

#2. Inputs Configuration
#2.1 Read in core datasets
#2.1.1 Read in risk variable data
risk_vars <- read_or_tk(
  first_existing(
    p_proc("risk_variable_data.csv"),
    p_proc("risk_variable_data.rds")),
  tk_cols = c("participant_id","variable","value","domain","stage","cluster")) %>% 
  ensure_participant_id()

#2.1.2 Read in outcome variable data
outcome_vars <- read_or_tk(
  first_existing(
    p_proc("outcome_variable_data.csv"),
    p_proc("outcome_variable_data.rds")),
  tk_cols = c(
    "participant_id","session_id","age","sex","site","wave",
    "any_bsd","bipolar_I","bipolar_II","bd_nos")) %>% 
  ensure_participant_id()

#2.1.3 Read in cluster labels
cluster_labels <- read_or_tk(
  first_existing(
    p_analysis("cluster_labels_k2_z_score.csv"),
    p_analysis("cluster_labels_k2_z_score.rds")),
  tk_cols = c("participant_id","cluster")) %>% 
  ensure_participant_id()

#2.1.4 Read in BD panel data
bd_panel <- read_or_tk(
  first_existing(
    p_analysis("bd_panel_k2_z_score.csv"),
    p_analysis("bd_panel_k2_z_score.rds")),
  tk_cols = c(
    "participant_id","family_id","site_factor","wave","cluster","outcome","status",
    "sex","age_wave","age_wave_cwc","age_wave_between",
    "any_bsd","bipolar_I","bipolar_II","bd_nos")) %>%
  ensure_participant_id()

#2.1.5 Read in BD person-period data
bd_pp <- read_or_tk(
  first_existing(
    p_analysis("bd_person_period_k2_z_score.csv"),
    p_analysis("bd_person_period_k2_z_score.rds")),
  tk_cols = c(
    "participant_id","family_id","site_factor","sex","cluster","outcome",
    "start_wave","end_wave","interval_index","event","baseline_status",
    "age_mid","age_mid_between","log_dt")) %>%
  ensure_participant_id()

#2.2 Map C1/C2 cluster labels to manuscript labels (C1 = Lower; C2 = Higher)
#2.2.1 Create a coercion function
coerce_cluster <- function(x) {
  x_std <- x %>%
    as.character() %>%
    stringr::str_to_lower() %>%
    stringr::str_replace_all("[^a-z0-9]+", "")
  dplyr::case_when(
    x_std %in% c("c1","1","low","lower","lowerrisk","lr") ~ "Lower-Risk",
    x_std %in% c("c2","2","high","higher","higherrisk","hr") ~ "Higher-Risk",
    TRUE ~ NA_character_)
}

#2.2.2 Coerce cluster labels where available
if (!is.null(cluster_labels) && nrow(cluster_labels)) {
  cluster_labels <- cluster_labels %>%
    dplyr::mutate(cluster = coerce_cluster(cluster))
}

#2.3 Read in validation results
#2.3.1 All available clustMixType derived index results
val_sil_all <- safe_read(first_existing(
  p_valid("val_z_score_silhouette.rds")
))

#2.3.2 Robust reader for per-k clustMixType validation index files (skips non-DF RDS)
val_sil_perk <- read_val_sil_perk(p_valid("partial_results_in_progress"))

#2.3.3 Optional other scalings (may be non-DF; used only if DF)
val_sil_z <- safe_read(first_existing(p_valid("val_z_score_silhouette.rds")))
val_sil_none <- safe_read(first_existing(p_valid("val_none_silhouette.rds")))
val_sil_minmax <- safe_read(first_existing(p_valid("val_min_max_silhouette.rds")))
val_sil_pct <- safe_read(first_existing(p_valid("val_percentile_silhouette.rds")))
val_sil_abs <- safe_read(first_existing(p_valid("val_max_absolute_silhouette.rds")))
val_cindex_all <- safe_read(first_existing(p_valid("val_robust_cindex.rds"), p_valid("val_robust_cindex.csv")))

#2.3.4 Read in cluster stability results
cluster_stability_raw <- read_or_tk(
  first_existing(
    p_stab("stability_z_score_k2_nb1000.csv"),
    p_stab("stability_z_score_k2_nb1000.rds")),
  tk_cols = c("metric","value","percentile","notes"))

#2.3.5 Read in UMAP embeddings
umap_embed <- read_or_tk(
  first_existing(
    p_valid("umap_embedding_k2.csv"),
    p_valid("umap_embedding_k2.rds")),
  tk_cols = c("x","y","cluster"))

#2.3.6 Read in missing-data imputation sensitivity-analysis results
#2.3.6.1 Read variable-level missingness table
miss_var_imp <- read_or_tk(
  first_existing(
    p_imp("tables", "missingness_by_variable.csv")),
  tk_cols = c("feature", "n_miss", "pct_miss"))

#2.3.6.2 Read imputability summary
imp_inclusion <- read_or_tk(
  first_existing(
    p_imp("tables", "imputation_inclusion_summary.csv")),
  tk_cols = c("N_total", "N_imputable", "pct_imputable", "min_obs_prop"))

#2.3.6.3 Read Little's MCAR summary
little_mcar_imp <- read_or_tk(
  first_existing(
    p_imp("tables", "little_mcar_test_numeric_only.csv")),
  tk_cols = c("statistic", "df", "p.value"))

#2.3.6.4 Read missForest k-grid summary metrics
kgrid_mf <- read_or_tk(
  first_existing(
    p_imp("tables", "kgrid_metrics_missForest.csv")),
  tk_cols = c(
    "imputation", "k", "k_observed", "min_cluster_n", "lambda", "lambda_reason",
    "added_dummy", "objective", "silhouette_huang", "rand_overall", "jacc_overall",
    "rand_mean", "jacc_mean", "rand_sd", "jacc_sd", "ari_vs_anchor_k2",
    "n_overlap_anchor"))

#2.3.7 Read in feature-importance / parsimonious-feature-set sensitivity-analysis results
#2.3.7.1 Read feature-importance master summary
fi_master_sum <- read_or_tk(
  first_existing(
    p_fi("summary", "feature_importance_master_summary.csv")),
  tk_cols = c(
    "feature", "delta_z", "max_cohens_d", "p_anova_fdr",
    "mcr_mean", "mcr_sd", "win_rate", "keep_stability",
    "p_emp_fdr", "mean_abs_shap"))

#2.3.7.2 Read seed-level MCR summaries across 50 seeds
fi_seed_level <- stack_dir(
  p_fi("seed_level"),
  pattern = "^feature_importance_seed[0-9]+\\.csv$",
  mutate_f = function(df) {
    df %>%
      janitor::clean_names() %>%
      dplyr::mutate(
        seed_idx = as.integer(.data$seed_idx),
        seed = as.integer(.data$seed))
  })

#2.3.7.3 Read SHAP global-importance summary
fi_shap_global <- read_or_tk(
  first_existing(
    p_fi("summary", "shap_global_importance.csv")),
  tk_cols = c("feature", "mean_abs_shap"))

#2.3.7.4 Read the exact saved SHAP object from the original feature-importance analysis
fi_shapviz <- safe_read(
  first_existing(
    p_fi("shap", "xgb_surrogate_shapviz.rds")))

#2.3.7.5 Keep the saved SHAP beeswarm PNG path only as a visual fallback
shap_beeswarm_png <- first_existing(
  p_fi("figures", "shap_global_importance_distributions.png"))

#2.3.7.6 Read MVFS feature-priority table
mvfs_priority_tbl <- read_or_tk(
  first_existing(
    p_mvfs("summary", "mvfs_feature_priority_table.csv")),
  tk_cols = c(
    "feature", "mcr_mean", "delta_z", "max_cohens_d", "sep_abs",
    "mean_abs_shap", "rank_MCR", "rank_centroid", "rank_SHAP",
    "rank_final_priority"))

#2.3.7.7 Read MVFS trace summary across subset size
mvfs_trace_all <- read_or_tk(
  first_existing(
    p_mvfs("summary", "mvfs_trace_all.csv")),
  tk_cols = c(
    "search", "lambda_mode", "step_idx", "n_features", "features",
    "lambda", "objective", "ari", "jaccard_mean", "jaccard_min",
    "silhouette_mean", "sig_direction_match", "full_direction_match",
    "full_direction_n_eval", "lambda_reason", "added_dummy",
    "k_observed", "min_cluster_n", "fail_reason"))

#2.3.7.8 Read selected MVFS summary
mvfs_selected_summary <- read_or_tk(
  first_existing(
    p_mvfs("summary", "mvfs_selected_summary.csv")),
  tk_cols = c("search", "lambda_mode", "mvfs_n", "mvfs_row"))

#2.3.7.9 Read selected MVFS solution object (contains the actual retained feature set)
mvfs_selected_solutions <- safe_read(
  first_existing(
    p_mvfs("summary", "mvfs_selected_solutions.rds")))

#2.3.7.10 Read selected MVFS assignment table if needed later
mvfs_selected_assignments <- read_or_tk(
  first_existing(
    p_mvfs("summary", "mvfs_selected_assignments.csv")),
  tk_cols = c("row_id", "anchor_cluster", "mvfs_cluster_mapped", "rowname"))

#2.3.7.11 Read saved 2-D embedding of the selected MVFS7 k=2 solution
mvfs7_umap_k2 <- read_or_tk(
  first_existing(
    p_mvfs_k("tables", "mvfs_mvfs7_umap_k2.csv")),
  tk_cols = c("UMAP1", "UMAP2", "cluster", "set", "k"))

#2.3.7.12 Read cross-algorithm benchmark metrics across candidate k solutions
alg_bench_metrics <- read_or_tk(
  first_existing(
    p_bench("tables", "algorithm_benchmark_metrics_all.csv"),
    p_bench("algorithm_benchmark_metrics_all.csv")),
  tk_cols = c(
    "algorithm", "k", "objective", "bic", "gower_silhouette",
    "ari_vs_anchor_k2", "min_cluster_n", "max_cluster_n",
    "lambda", "lambda_reason"))

#2.3.7.13 Read common-comorbidity sensitivity-analysis ORs by wave
comorb_or_by_wave <- read_or_tk(
  first_existing(
    p_comorb("tables", "gee_or_by_wave.csv"),
    p_res("5_comorbidity_sensitivity_analysis", "tables", "gee_or_by_wave.csv")),
  tk_cols = c(
    "outcome", "wave", "contrast", "log_or", "se", "lcl", "ucl",
    "or", "or_lcl", "or_ucl", "p"))

#2.3.7.14 Read utility-of-alternative-baseline-risk-representations sensitivity-analysis outputs
utility_takeaway <- read_or_tk(
  first_existing(
    p_utility("tables", "cluster_vs_mvfs_vs_raw_takeaway_table.csv"),
    "/mnt/data/cluster_vs_mvfs_vs_raw_takeaway_table.csv"),
  tk_cols = c(
    "domain", "outcome", "primary_metric",
    "cluster_primary", "mvfs_primary", "raw_primary",
    "secondary_metric", "cluster_secondary", "mvfs_secondary", "raw_secondary",
    "delta_mvfs_minus_cluster", "delta_raw_minus_cluster", "delta_raw_minus_mvfs",
    "mvfs_block_p", "raw_block_p"))

dth_utility_coef <- read_or_tk(
  first_existing(
    p_utility("tables", "dth_all_model_robust_coefficients.csv"),
    "/mnt/data/dth_all_model_robust_coefficients.csv"),
  tk_cols = c(
    "outcome", "model", "se_type", "term", "estimate",
    "SE", "df", "OR", "CI_lo", "CI_hi", "p.value"))

prev_utility_coef <- read_or_tk(
  first_existing(
    p_utility("tables", "prevalence_all_model_coefficients.csv"),
    "/mnt/data/prevalence_all_model_coefficients.csv"),
  tk_cols = c(
    "outcome", "model", "term", "estimate",
    "SE", "OR", "CI_lo", "CI_hi", "Wald", "p.value"))

#2.4 Read in BD onset results
#2.4.1 Read in all relevant BD survival analysis results with finalized glm-based filenames
bd_hazard_by_wave <- read_or_tk(
  p_bd_surv("bd_dth_hazard_by_wave.csv"),
  tk_cols = c("outcome","model","end_wave","cluster","hazard","lo","hi"))

bd_cumrisk_by_wave <- read_or_tk(
  p_bd_surv("bd_dth_cumrisk_by_wave.csv"),
  tk_cols = c("outcome","model","end_wave","cluster","cumrisk","cumrisk_lo","cumrisk_hi"))

#2.4.2 Read in predicted hazards (kept available for supplement / alternate plotting if needed)
bd_preds_wave_aligned <- read_or_tk(
  p_bd_surv("bd_dth_preds_wave_aligned.csv"),
  tk_cols = c("outcome","model","pred_mode","cluster","end_wave","age_years","pred","lower","upper"))

bd_preds_fixedwave <- read_or_tk(
  p_bd_surv("bd_dth_preds_fixedwave.csv"),
  tk_cols = c("outcome","model","pred_mode","cluster","age_years","end_wave","pred","lower","upper"))

#2.4.3 Read in survival model summaries
bd_dth_wald <- read_or_tk(
  p_bd_surv("bd_dth_wald_reference.csv"),
  tk_cols = c("outcome","model","term","estimate","se","or","ci_lo","ci_hi","p_value"))

bd_dth_cr2 <- read_or_tk(
  p_bd_surv("bd_dth_inference_CR2.csv"),
  tk_cols = c("outcome","model","se_type","term","estimate","se","df","or","ci_lo","ci_hi","p_value"))

bd_dth_hc1 <- read_or_tk(
  p_bd_surv("bd_dth_inference_HC1_participant.csv"),
  tk_cols = c("outcome","model","term","estimate","se","df","or","ci_lo","ci_hi","p_value"))

#2.4.4 Read in emmeans contrasts
bd_dth_emm_age <- read_or_tk(
  p_bd_surv("bd_dth_emmeans_contrasts_by_age.csv"),
  tk_cols = c("outcome","model","age_years","contrast","or","ci_lo","ci_hi","p_value"))

bd_dth_emm_wave <- read_or_tk(
  p_bd_surv("bd_dth_emmeans_contrasts_by_wave.csv"),
  tk_cols = c("outcome","model","end_wave","contrast","or","ci_lo","ci_hi","p_value"))

#2.4.5 Read in wave age medians
bd_wave_age_medians <- read_or_tk(
  p_bd_surv("wave_age_medians.csv"),
  tk_cols = c("end_wave","age_med"))

#2.5 Read in BD prevalence results
#2.5.1 GEE by wave, fixed effects, GLMM (where available), and TMB fixed effects results (where available)
gee_by_wave_bd <- stack_dir(
  p_bd_mix("csv"),
  pattern = "GEE_(AR1|EXC)(_INT)?_cluster_OR_by_wave\\.csv$",
  mutate_f = function(df) {
    df %>%
      janitor::clean_names() %>%
      dplyr::mutate(outcome = norm_bd_outcome(file),
        model = dplyr::case_when(
          stringr::str_detect(file, "GEE_AR1") ~ "GEE_AR1",
          stringr::str_detect(file, "GEE_EXC") ~ "GEE_EXC",
          TRUE ~ "GEE")) %>%
      dplyr::select(-file)
  }
)

#2.5.2 GEE fixed effects
gee_fixed_bd <- stack_dir(
  p_bd_mix("csv"),
  pattern = "GEE_(AR1|EXC)(_INT)?_fixed(.*)\\.csv$",
  mutate_f = function(df) {
    df %>%
      janitor::clean_names() %>%
      dplyr::mutate(outcome = norm_bd_outcome(file),
        model = dplyr::case_when(
          stringr::str_detect(file, "GEE_AR1") ~ "GEE_AR1",
          stringr::str_detect(file, "GEE_EXC") ~ "GEE_EXC",
          TRUE ~ "GEE")) %>%
      dplyr::select(-file)
  }
)

#2.5.3 GLMM fixed effects
glmm_fixed_bd <- stack_dir(
  p_bd_mix("csv"),
  pattern = "GLMM.*fixed_ORs\\.csv$",
  mutate_f = function(df) {
    df %>% janitor::clean_names() %>%
      dplyr::mutate(outcome = norm_bd_outcome(file), model = "GLMM") %>%
      dplyr::select(-file)
  }
)

#2.5.4 glmmTMB fixed effects
tmb_fixed_bd <- stack_dir(
  p_bd_mix("csv"),
  pattern = "glmmTMB.*fixed_ORs\\.csv$",
  mutate_f = function(df) {
    df %>% janitor::clean_names() %>%
      dplyr::mutate(outcome = norm_bd_outcome(file), model = "glmmTMB") %>%
      dplyr::select(-file)
  }
)

#2.5.5 GAMM age-based predictions
bd_gamm_preds <- stack_dir(
  p_bd_mix("pred_csv"),
  pattern = "_GAMM_age_predictions\\.csv$",
  mutate_f = function(df) {
    df %>% janitor::clean_names() %>%
      dplyr::mutate(outcome = norm_bd_outcome(file)) %>%
      dplyr::select(-file)
  }
)

#2.6 CONSORT counts if they exist, else are derived later during construction of CONSORT diagram
consort_counts <- read_or_tk(
  first_existing(
    p_res("consort_counts.csv"),
    p_res("main_analysis","consort_counts.csv")),
  tk_cols = c("stage","label","n","wave","cluster"))

#2.7 Standardize stats data frames
#2.7.1 Function to standardize finalized DTH stats objects
standardize_stats_df <- function(df) {
  if (is.null(df) || is_tk(df)) return(NULL)

  df <- janitor::clean_names(df)

  #2.7.1.1 Harmonize naming
  if ("se_type" %in% names(df) && !"se_method" %in% names(df)) df$se_method <- df$se_type
  if ("p_value" %in% names(df) && !"p.value" %in% names(df)) df$p.value <- df$p_value
  if ("or" %in% names(df) && !"OR" %in% names(df)) df$OR <- df$or
  if ("ci_lo" %in% names(df) && !"conf.low" %in% names(df)) df$conf.low <- df$ci_lo
  if ("ci_hi" %in% names(df) && !"conf.high" %in% names(df)) df$conf.high <- df$ci_hi

  #2.7.1.2 Backfill OR / CI if needed
  if (!"OR" %in% names(df) && "estimate" %in% names(df)) df$OR <- exp(df$estimate)

  #2.7.1.3 Ensure required columns exist
  need <- c("outcome","model","term","estimate","OR","conf.low","conf.high","p.value","se_method","link")
  for (nm in need) if (!nm %in% names(df)) df[[nm]] <- NA

  df
}

#2.7.2 Standardize BD DTH results
bd_dth_cr2_std  <- standardize_stats_df(bd_dth_cr2)
bd_dth_wald_std <- standardize_stats_df(bd_dth_wald)

#2.7.3 Combine and annotate DTH results
bd_dth_sum <- dplyr::bind_rows(
  bd_dth_cr2_std,
  bd_dth_wald_std) %>%
  dplyr::mutate(
    outcome = norm_bd_outcome(dplyr::coalesce(.data$outcome, "")),
    se_method = dplyr::coalesce(se_method, "model-based"),
    link = dplyr::case_when(
      model == "cloglog_offset" ~ "cloglog",
      TRUE ~ "logit"))

#2.7.4 Main-text DTH source = participant-clustered CR2 from primary pooled-logit model
bd_dth_main <- bd_dth_sum %>%
  dplyr::filter(model == "logit_primary", se_method == "CR2")

#2.7.5 Sensitivity DTH source
bd_dth_sens <- bd_dth_sum %>%
  dplyr::filter(!(model == "logit_primary" & se_method == "CR2"))

#2.8 Seed TK data if needed for later use
seed_if_needed <- function() {

  #2.8.1 k_sil: simple TK curve or from val_sil_perk (silhouette by k and scaling)
  if (is.null(val_sil_all) && (is.null(val_sil_perk) || !is.data.frame(val_sil_perk) || nrow(val_sil_perk) == 0)) {
    k_sil_df <- tidyr::expand_grid(k = 2:8, scaling = c("robust_iqr","z_score","none")) %>%
      dplyr::mutate(silhouette = dplyr::case_when(
        scaling == "robust_iqr" ~ dplyr::case_when(k == 2 ~ 0.48, k == 3 ~ 0.43, k == 4 ~ 0.41, TRUE ~ 0.38),
        scaling == "zscore" ~ 0.35 - 0.01*(k-2),
        TRUE ~ 0.30 - 0.015*(k-2)))
  } else if (is.data.frame(val_sil_perk) && nrow(val_sil_perk) > 0) {
    k_sil_df <- val_sil_perk
  } else {

    #2.8.2 Fallback TK if val_sil_all is present but not tidy
    k_sil_df <- tidyr::expand_grid(k = 2:8, scaling = c("robust_iqr","zscore","none")) %>%
      dplyr::mutate(silhouette = 0.4 - 0.01*(k-2))
  }
  assign("k_sil", k_sil_df, envir = .GlobalEnv)

  #2.8.3 k_wss: simple TK curve
  k_wss_df <- tidyr::expand_grid(k = 2:8, scaling = c("robust_iqr","zscore","none")) %>%
    dplyr::group_by(scaling) %>%
    dplyr::mutate(
      wss = 1000/(k-1) + dplyr::case_when(
        scaling == "robust_iqr" ~ 0,
        scaling == "zscore" ~ 50,
        TRUE ~ 100)) %>%
    dplyr::ungroup()
  assign("k_wss", k_wss_df, envir = .GlobalEnv)

  #2.8.4 UMAP: TK cloud if needed
  if (is.null(umap_embed) || is_tk(umap_embed) || !all(c("x","y","cluster") %in% names(umap_embed))) {
    set.seed(123)
    assign("umap_embed",
      tibble::tibble(
      x = c(rnorm(500,-2,1), rnorm(500, 2,1)),
      y = c(rnorm(500, 0,1), rnorm(500, 0,1)),
      cluster = rep(c("C1","C2"), each = 500)),
      envir = .GlobalEnv)
  }

  #2.8.5 CONSORT counts: TK counts if needed
  if (is.null(consort_counts) || is_tk(consort_counts)) {
    assign("consort_counts",
      tibble::tribble(
        ~stage, ~label, ~n, ~wave, ~cluster,
        "start", "ABCD baseline consented", 11878, NA, NA,
        "filter","Eligible (age/QC)", 11500, NA, NA,
        "filter","Complete risk variables", 9800, NA, NA,
        "filter","Complete outcomes & covariates", 9500, NA, NA,
        "filter","Excluded per pre-spec (22.55%)", 2676, NA, NA,
        "end", "Included in risk clustering", 9193, NA, NA),
    envir = .GlobalEnv)
  }
}

#2.9 Seed TK data if needed
seed_if_needed()

```

## Table 1 - Characteristics of the Two Primary Analytic Samples

- Purpose: Produce the new main-text analytic-sample table replacing the old CONSORT display by summarizing the total columns of the first-onset and occurrence analytic samples in one JAMA-aligned table
- Inputs: `bd_pp` for the first-onset analytic sample and `bd_panel` for the occurrence analytic sample
- Outputs: Saves `Table_01_Analytic_Sample_Characteristics` as HTML/PNG/RTF in `results/tables`; returns a gt table
- Notes: Panel A mirrors the total columns of Table S1 and Panel B mirrors the total columns of Table S2, but both are generated directly from source data so the compiler remains self-contained and updates automatically with the analysis outputs

```{r table 1}

## Table 1: Characteristics of the two primary analytic samples ##

#T1.1 Create shared helpers for main-text analytic-sample summaries
#T1.1.1 Set wave -> time labels
time_map_t1 <- c("ses-02A" = "2y", "ses-04A" = "4y", "ses-06A" = "6y")
time_levels_t1 <- c("2y", "4y", "6y")

#T1.1.2 Formatting helpers
fmt_mean_sd_t1 <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) return("\u2014")
  sprintf("%.2f \u00B1 %.2f", mean(x), stats::sd(x))
}

fmt_n_t1 <- function(n) {
  ifelse(is.na(n), "\u2014", format(as.integer(n), big.mark = ","))
}

fmt_n_pct_t1 <- function(n, den) {
  out <- ifelse(
    is.na(den) | den == 0,
    "0 (0.0%)",
    sprintf("%s (%.1f%%)", fmt_n_t1(n), 100 * n / den))
  as.character(out)
}

#T1.1.3 Desired row ordering within each panel
row_order_onset <- c(
  "N (participants)",
  "N (person-periods)",
  "Age in years, Mean (SD)",
  "Sex at birth - Male",
  "Sex at birth - Female",
  "Interval onset - BD I, n (%)",
  "Interval onset - BD II, n (%)",
  "Interval onset - BD NOS, n (%)",
  "Interval onset - Any BSD, n (%)")

row_order_prev <- c(
  "N (participants)",
  "N (person-periods)",
  "Age in years, Mean (SD)",
  "Sex at birth - Male",
  "Sex at birth - Female",
  "Occurrence - BD I, n (%)",
  "Occurrence - BD II, n (%)",
  "Occurrence - BD NOS, n (%)",
  "Occurrence - Any BSD, n (%)")

#T1.2 Build Panel A totals paralleling the total columns of Table S1
#T1.2.1 Prepare BD person-period data
bd_pp_t1 <- bd_pp %>%
  dplyr::mutate(
    participant_id = as.character(participant_id),
    family_id = factor(as.character(family_id)),
    site_factor = factor(site_factor),
    cluster = factor(cluster),
    start_wave = factor(start_wave),
    end_wave = factor(end_wave),
    outcome = factor(outcome),
    sex = factor(sex))

#T1.2.2 Apply the same model-ready filtering logic used for the onset analysis table
bd_pp_t1_model <- bd_pp_t1 %>%
  dplyr::filter(event %in% c(0L, 1L)) %>%
  dplyr::filter(!is.na(cluster)) %>%
  dplyr::filter(!is.na(age_mid), !is.na(sex), !is.na(site_factor), !is.na(family_id))

#T1.2.3 Derive baseline any-BSD status and restrict to the primary onset risk set
baseline_any_bsd_ids_t1 <- bd_pp_t1_model %>%
  dplyr::filter(outcome == "any_bsd") %>%
  dplyr::group_by(participant_id) %>%
  dplyr::summarise(
    all_na = all(is.na(baseline_status)),
    baseline_any_bsd = dplyr::if_else(
      all_na,
      NA_integer_,
      as.integer(any(baseline_status == 1L, na.rm = TRUE))),
    .groups = "drop")

bd_pp_t1_model <- bd_pp_t1_model %>%
  dplyr::left_join(
    baseline_any_bsd_ids_t1 %>% dplyr::select(participant_id, baseline_any_bsd),
    by = "participant_id") %>%
  dplyr::filter(baseline_any_bsd == 0L)

#T1.2.4 Censor observations after first onset within each outcome
bd_pp_t1_model <- bd_pp_t1_model %>%
  dplyr::arrange(
    outcome,
    participant_id,
    if ("interval_index" %in% names(.)) interval_index else as.integer(end_wave)) %>%
  dplyr::group_by(outcome, participant_id) %>%
  dplyr::mutate(after_onset = dplyr::lag(cummax(event == 1L), default = FALSE)) %>%
  dplyr::filter(!after_onset) %>%
  dplyr::ungroup() %>%
  dplyr::select(-after_onset)

#T1.2.5 Add main-text time labels and retain valid waves
pp_total <- bd_pp_t1_model %>%
  dplyr::mutate(
    time = dplyr::recode(as.character(end_wave), !!!time_map_t1),
    time = factor(time, levels = time_levels_t1)) %>%
  dplyr::filter(!is.na(time))

#T1.2.6 Create demographic rows from the any-BSD onset risk set
demo_onset_total <- pp_total %>%
  dplyr::filter(outcome == "any_bsd")

demo_onset_counts <- demo_onset_total %>%
  dplyr::group_by(time) %>%
  dplyr::summarise(
    n_pp = dplyr::n(),
    n_part = dplyr::n_distinct(participant_id),
    age_mean_sd = fmt_mean_sd_t1(age_mid),
    .groups = "drop")

demo_onset_n_rows <- demo_onset_counts %>%
  tidyr::pivot_longer(c(n_pp, n_part), names_to = "what", values_to = "val_raw") %>%
  dplyr::mutate(
    Row = dplyr::case_when(
      what == "n_part" ~ "N (participants)",
      what == "n_pp" ~ "N (person-periods)"),
    val = fmt_n_t1(val_raw)) %>%
  dplyr::select(time, Row, val)

demo_onset_age_row <- demo_onset_counts %>%
  dplyr::transmute(
    time,
    Row = "Age in years, Mean (SD)",
    val = age_mean_sd)

demo_onset_sex_rows <- demo_onset_total %>%
  dplyr::mutate(
    sex2 = dplyr::case_when(
      !is.na(sex) & sex %in% c("Male", "Female") ~ as.character(sex),
      TRUE ~ "Other/Unknown")) %>%
  dplyr::distinct(time, participant_id, sex2) %>%
  dplyr::count(time, sex2, name = "n_sex") %>%
  dplyr::left_join(
    demo_onset_counts %>% dplyr::select(time, n_part),
    by = "time") %>%
  dplyr::mutate(val = fmt_n_pct_t1(n_sex, n_part)) %>%
  dplyr::filter(sex2 %in% c("Male", "Female")) %>%
  dplyr::mutate(Row = paste0("Sex at birth - ", sex2)) %>%
  dplyr::select(time, Row, val)

#T1.2.7 Create interval-onset rows from the outcome-specific onset risk sets
onset_labels_t1 <- c(
  "bipolar_I" = "Interval onset - BD I, n (%)",
  "bipolar_II" = "Interval onset - BD II, n (%)",
  "bipolar_NOS" = "Interval onset - BD NOS, n (%)",
  "bd_nos" = "Interval onset - BD NOS, n (%)",
  "any_bsd" = "Interval onset - Any BSD, n (%)")

onset_rows_total <- pp_total %>%
  dplyr::filter(outcome %in% names(onset_labels_t1)) %>%
  dplyr::group_by(outcome, time) %>%
  dplyr::summarise(
    n_at_risk = dplyr::n(),
    n_events = sum(event == 1L, na.rm = TRUE),
    .groups = "drop") %>%
  dplyr::mutate(
    Row = dplyr::recode(outcome, !!!onset_labels_t1),
    val = fmt_n_pct_t1(n_events, n_at_risk)) %>%
  dplyr::select(time, Row, val)

panel_a_rows <- dplyr::bind_rows(
  demo_onset_n_rows,
  demo_onset_age_row,
  demo_onset_sex_rows,
  onset_rows_total) %>%
  dplyr::mutate(
    Panel = "Panel A. First-onset analytic sample",
    row_id = match(Row, row_order_onset))

#T1.3 Build Panel B totals paralleling the total columns of Table S2
#T1.3.1 Prepare BD panel data
bd_panel_t1 <- bd_panel %>%
  dplyr::mutate(
    participant_id = as.character(participant_id),
    family_id = factor(as.character(family_id)),
    site_factor = factor(site_factor),
    cluster = factor(cluster),
    wave = factor(wave),
    outcome = factor(outcome),
    sex = factor(sex),
    status = as.integer(.data[["status"]]))

#T1.3.2 Apply the same model-ready filtering logic used for the prevalence analysis table
bd_panel_t1_model <- bd_panel_t1 %>%
  dplyr::filter(.data[["status"]] %in% c(0L, 1L)) %>%
  dplyr::filter(!is.na(cluster), !is.na(sex), !is.na(site_factor), !is.na(family_id), !is.na(wave)) %>%
  dplyr::filter(!is.na(age_wave), !is.na(age_wave_cwc), !is.na(age_wave_between))

#T1.3.3 Add main-text time labels and retain valid waves
panel_total <- bd_panel_t1_model %>%
  dplyr::mutate(
    time = dplyr::recode(as.character(wave), !!!time_map_t1),
    time = factor(time, levels = time_levels_t1),
    sex = as.character(sex)) %>%
  dplyr::filter(!is.na(time))

#T1.3.4 Create demographic rows from the any-BSD prevalence panel
demo_prev_total <- panel_total %>%
  dplyr::filter(outcome == "any_bsd")

demo_prev_counts <- demo_prev_total %>%
  dplyr::group_by(time) %>%
  dplyr::summarise(
    n_pp = dplyr::n(),
    n_part = dplyr::n_distinct(participant_id),
    age_mean_sd = fmt_mean_sd_t1(age_wave),
    .groups = "drop")

demo_prev_n_rows <- demo_prev_counts %>%
  tidyr::pivot_longer(c(n_pp, n_part), names_to = "what", values_to = "val_raw") %>%
  dplyr::mutate(
    Row = dplyr::case_when(
      what == "n_part" ~ "N (participants)",
      what == "n_pp" ~ "N (person-periods)"),
    val = fmt_n_t1(val_raw)) %>%
  dplyr::select(time, Row, val)

demo_prev_age_row <- demo_prev_counts %>%
  dplyr::transmute(
    time,
    Row = "Age in years, Mean (SD)",
    val = age_mean_sd)

demo_prev_sex_rows <- demo_prev_total %>%
  dplyr::mutate(
    sex2 = dplyr::case_when(
      !is.na(sex) & sex %in% c("Male", "Female") ~ sex,
      TRUE ~ "Other/Unknown")) %>%
  dplyr::distinct(time, participant_id, sex2) %>%
  dplyr::count(time, sex2, name = "n_sex") %>%
  dplyr::left_join(
    demo_prev_counts %>% dplyr::select(time, n_part),
    by = "time") %>%
  dplyr::mutate(val = fmt_n_pct_t1(n_sex, n_part)) %>%
  dplyr::filter(sex2 %in% c("Male", "Female")) %>%
  dplyr::mutate(Row = paste0("Sex at birth - ", sex2)) %>%
  dplyr::select(time, Row, val)

#T1.3.5 Create occurrence rows from the outcome-specific prevalence panels
prev_labels_t1 <- c(
  "bipolar_I" = "Occurrence - BD I, n (%)",
  "bipolar_II" = "Occurrence - BD II, n (%)",
  "bd_nos" = "Occurrence - BD NOS, n (%)",
  "any_bsd" = "Occurrence - Any BSD, n (%)")

prev_rows_total <- panel_total %>%
  dplyr::filter(outcome %in% names(prev_labels_t1)) %>%
  dplyr::group_by(outcome, time) %>%
  dplyr::summarise(
    n_obs = sum(!is.na(status)),
    n_cases = sum(status == 1L, na.rm = TRUE),
    .groups = "drop") %>%
  dplyr::mutate(
    Row = dplyr::recode(outcome, !!!prev_labels_t1),
    val = fmt_n_pct_t1(n_cases, n_obs)) %>%
  dplyr::select(time, Row, val)

panel_b_rows <- dplyr::bind_rows(
  demo_prev_n_rows,
  demo_prev_age_row,
  demo_prev_sex_rows,
  prev_rows_total) %>%
  dplyr::mutate(
    Panel = "Panel B. Occurrence analytic sample",
    row_id = match(Row, row_order_prev))

#T1.4 Assemble the final main-text analytic-sample table
#T1.4.1 Bind Panel A and Panel B rows
t1_long <- dplyr::bind_rows(panel_a_rows, panel_b_rows) %>%
  dplyr::mutate(
    time = factor(time, levels = time_levels_t1),
    Panel = factor(
      Panel,
      levels = c(
        "Panel A. First-onset analytic sample",
        "Panel B. Occurrence analytic sample")))

#T1.4.2 Complete panel-specific row x time combinations
panel_row_grid <- t1_long %>%
  dplyr::distinct(Panel, Row, row_id)

t1_long <- tidyr::crossing(
  panel_row_grid,
  time = factor(time_levels_t1, levels = time_levels_t1)) %>%
  dplyr::left_join(
    t1_long %>% dplyr::select(Panel, Row, row_id, time, val),
    by = c("Panel", "Row", "row_id", "time")) %>%
  dplyr::mutate(val = dplyr::coalesce(val, "\u2014"))

#T1.4.3 Pivot to wide format for gt
t1_wide <- t1_long %>%
  dplyr::arrange(Panel, row_id, time) %>%
  tidyr::pivot_wider(
    names_from = time,
    values_from = val) %>%
  dplyr::select(Panel, Row, dplyr::all_of(time_levels_t1))

#T1.4.4 Create caption
t1_caption <- paste0(
  "**Table 1. Characteristics of Participants Included in the Primary Onset and Occurrence Analyses**  \n",
  "Panel A summarizes the first-onset analytic sample and mirrors the total columns of Table S1. ",
  "Panel B summarizes the occurrence analytic sample and mirrors the total columns of Table S2. ",
  "Values are generated directly from source analysis data rather than imported from the supplementary workbook.")

#T1.5 Build, save, and print the gt table
t1_gt <- gt::gt(
  t1_wide,
  groupname_col = "Panel",
  rowname_col = "Row") |>
  gt::cols_label(
    `2y` = "2y",
    `4y` = "4y",
    `6y` = "6y") |>
  gt::cols_align(
    align = "center",
    columns = dplyr::all_of(time_levels_t1)) |>
  gt::tab_caption(t1_caption) |>
  gt::tab_options(table.font.size = gt::px(12)) |>
  gt::fmt_missing(columns = dplyr::everything(), missing_text = "\u2014")

save_gt(t1_gt, "Table_01_Analytic_Sample_Characteristics")
t1_gt

```

## Table 2 - Baseline Characteristics of the Clustered Sample

- Purpose: Produce Table 2 aligned to manuscript display
- Inputs: `data_raw/dataset.csv` for demographics; `outcome_vars` for baseline diagnoses; `cluster_labels` for cluster membership
- Outputs: Saves `Table_02_Baseline_Characteristics` as HTML/PNG/RTF in `results/tables`; returns a gt object (fallback mini table if `gtsummary` missing)
- Notes: Requires strict raw column names (see `DEM_COLS`) and outcome variables (`DIAG_COLS`); earliest session per participant

```{r table 2}

## Table 2: Baseline characteristics of the clustered sample ##

#T2.1 Define required raw column mappings
DEM_COLS <- list(
  id = "participant_id",
  session = "session_id",
  age = "ab_g_dyn__visit_age",
  sex = "ab_g_stc__cohort_sex"  # 1=Male, 2=Female
)

#T2.2 Function to read demographics strictly
read_demographics_strict <- function() {
  dem_path <- first_existing(
    p_data("data_raw", "dataset.csv"),
    "/home/exacloud/gscratch/NagelLab/staff/sam/projects/ABCD_MDS_Risk/data/data_raw/dataset.csv")
  dem <- safe_read(dem_path)
  need <- unlist(DEM_COLS, use.names = FALSE)
  miss <- setdiff(need, names(dem))
  if (length(miss)) stop("Missing required columns in raw dataset: ", paste(miss, collapse=", "))

  #T2.2.1 process demographics to get earliest session per participant
  dem %>%
    dplyr::select(
      participant_id = !!DEM_COLS$id,
      session_id = !!DEM_COLS$session,
      age = !!DEM_COLS$age,
      sex_code = !!DEM_COLS$sex) %>%
    dplyr::group_by(participant_id) %>%
    dplyr::arrange(session_id, .by_group = TRUE) %>%
    dplyr::slice(1) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      sex = dplyr::case_when(
        sex_code %in% c(1,"1") ~ "Male",
        sex_code %in% c(2,"2") ~ "Female",
        TRUE ~ NA_character_)) %>%
    dplyr::select(participant_id, age, sex)
}

#T2.3 List diagnosis columns of interest
DIAG_COLS <- c("any_bsd","bipolar_I","bipolar_II","bd_nos")

#T2.4 Function to get baseline outcomes from outcome_vars
baseline_outcomes <- function(ov) {
  need <- c("participant_id","session_id", DIAG_COLS)
  miss <- setdiff(need, names(ov))
  if (length(miss)) stop("Missing required columns in outcome_vars: ", paste(miss, collapse=", "))

  #T2.4.1 get baseline session data of interest per participant
  ov %>%
    dplyr::select(dplyr::all_of(need)) %>%
    dplyr::group_by(participant_id) %>%
    dplyr::arrange(session_id, .by_group = TRUE) %>%
    dplyr::slice(1) %>%
    dplyr::ungroup()
}

#T2.5 Build Table 2 dataset restricted to clustered sample
#T2.5.1 Read demographics
demographics_df <- read_demographics_strict()
ov_base <- baseline_outcomes(outcome_vars)

#T2.5.2 Merge cluster labels, demographics, and baseline outcomes
tbl2_df <- cluster_labels %>%
  dplyr::select(participant_id, cluster) %>%
  dplyr::left_join(demographics_df, by = "participant_id") %>%
  dplyr::left_join(ov_base, by = "participant_id") %>%
  dplyr::mutate(cluster = factor(cluster, levels = c("Higher-Risk","Lower-Risk")))

#T2.5.3 Create readable labels for display
label_list <- list(
  age ~ "Age (years)",
  sex ~ "Sex at birth",
  any_bsd ~ "Any BSD diagnosis",
  bipolar_I ~ "BD I diagnosis",
  bipolar_II ~ "BD II diagnosis",
  bd_nos ~ "BD NOS diagnosis")

#T2.5.4 Create table caption
tbl2_caption <- glue::glue(
  "**Table 2. Baseline characteristics of the clustered sample**  \n",
  "Participants included in clustering (n={scales::comma(dplyr::n_distinct(tbl2_df$participant_id))}).")

#T2.5.5 Function to convert binary variables to yes/no factors
to_yesno <- function(x) {
  if (is.numeric(x) || is.logical(x)) {
    factor(dplyr::case_when(
      is.na(x) ~ NA_character_,
      as.numeric(x) == 1 ~ "Yes",
      TRUE ~ "No"
    ), levels = c("No","Yes"))
  } else x
}

#T2.5.6 Prepare display dataset
tbl2_disp <- tbl2_df %>%
  dplyr::mutate(dplyr::across(dplyr::all_of(DIAG_COLS), to_yesno),
    sex = factor(sex, levels = c("Male","Female")))

#T2.6 Create gtsummary table for table 1
if (requireNamespace("gtsummary", quietly = TRUE)) {

  #T2.6.1 Build the base table with Overall column
  tbl2_gts <-
    gtsummary::tbl_summary(
      data = tbl2_disp %>% dplyr::select(cluster, age, sex, dplyr::all_of(DIAG_COLS)),
      by = cluster,
      type = list(age ~ "continuous"),
      statistic = list(
        age ~ "{mean} ({sd})",
        gtsummary::all_categorical() ~ "{n} ({p}%)"),
      missing = "no",
      label = label_list) %>%
    gtsummary::add_overall() %>%

    #T2.6.2 Clean headers to show only the level names (no N in headers)
    gtsummary::modify_header(all_stat_cols() ~ "**{level}**") %>%

    #T2.6.3 Insert a top "N" row with counts per column (Overall + each cluster)
    gtsummary::modify_table_body(~{
      tb <- .

      #T2.6.3.1 counts in the same column order as the stats columns: stat_0 = Overall, stat_1 = Higher-Risk, stat_2 = Lower-Risk (given factor levels)
      n_overall <- dplyr::n_distinct(tbl2_df$participant_id)
      n_by <- tbl2_df %>%
        dplyr::count(cluster, name = "n") %>%
        dplyr::arrange(factor(cluster, levels = c("Higher-Risk","Lower-Risk"))) %>%
        dplyr::pull(n)
      n_by <- c(n_by, rep(NA_integer_, max(0, 2 - length(n_by))))  # safety pad

      #T2.6.3.2 create N row
      n_row <- tibble::tibble(
        variable = ".N_row",
        var_type = NA_character_,
        var_label = "N",
        row_type = "label",
        label = "N",
        stat_0 = scales::comma(n_overall),
        stat_1 = scales::comma(n_by[1]),
        stat_2 = scales::comma(n_by[2]))
      dplyr::bind_rows(n_row, tb)
    }) %>%
    gtsummary::modify_caption(tbl2_caption) %>%
    gtsummary::as_gt()

  #T2.6.4 Save the gtsummary table in multiple formats
  save_gt(tbl2_gts, "Table_02_Baseline_Characteristics")
  tbl2_gts
} else {

  #T2.6.5 fallback unchanged
  mini <- tbl2_df %>%
    dplyr::count(cluster, name = "n") %>%
    dplyr::mutate(pct = 100 * n / sum(n)) %>%
    dplyr::arrange(cluster)
  mini_gt <- gt::gt(mini) %>%
    gt::fmt_number(columns = "pct", decimals = 1) %>%
    gt::cols_label(cluster = "Cluster", n = "n", pct = "Percent") %>%
    gt::tab_caption(paste0(tbl2_caption, "  (limited: gtsummary not available)."))

  #T2.6.6 Save the fallback mini gt table if necessaruy
  save_gt(mini_gt, "Table_02_Baseline_Characteristics_minimal")
  mini_gt
}

```

## Table 3 - Baseline Risk Variable Descriptives by Cluster (+ effect sizes)

- Purpose: Summarize baseline risk-variable descriptives by cluster with effect sizes
- Inputs: `risk_vars` (wide or long) and `cluster_labels`; optional pretty labels from `data/data_processed/risk_variable_labels.csv`
- Outputs: Saves `Table_03_Risk_Variables_by_Cluster` as HTML/PNG/RTF; returns a gt table
- Notes: Automatically pivots wide `risk_vars` to long; detects variable types (binary vs continuous); uses `effectsize` and `glm`

```{r table 3}

## Table 3: Risk-variable descriptive statistics by cluster (+ effect sizes) ##

#T3.1 Prepare risk variable data (long format with variable + value)
#T3.1.1 If risk_vars is wide, pivot to long; else use as-is
rv0 <- risk_vars
if (!is.null(rv0) && nrow(rv0)) {
  if (!"cluster" %in% names(rv0) && !is_tk(cluster_labels)) {
    rv0 <- rv0 %>% dplyr::left_join(cluster_labels, by = "participant_id")
  }
  rv0 <- rv0 %>%
    dplyr::filter(cluster %in% c("Lower-Risk","Higher-Risk"))

  #T3.1.1.1 Pivot wide -> long (make `variable` + `value`)
  id_cols <- c("participant_id","session_id","site","age","race_ethnicity","sex","family_id","cluster")
  risk_cols <- setdiff(names(rv0), id_cols)
  rv0 <- rv0 %>%
    tidyr::pivot_longer(
      cols = dplyr::all_of(risk_cols),
      names_to  = "variable",
      values_to = "value"
    )
} else {
  rv0 <- read_or_tk("", tk_cols = c("participant_id","variable","value","cluster"))
}

#T3.1.2 Identify variable types (binary vs continuous vs other) and mutate to correct types if not already present as such
var_types <- rv0 %>%
  dplyr::group_by(variable) %>%
  dplyr::summarise(
    n_nonmiss = sum(!is.na(value)),
    is_numeric = all(is.numeric(value) | is.integer(value)),
    uniq_nonNA = dplyr::n_distinct(value[!is.na(value)]),
    all01 = all(value[!is.na(value)] %in% c(0,1)),
    .groups = "drop") %>%
  dplyr::mutate(
    type = dplyr::case_when(
      is_numeric & all01 ~ "Binary",
      is_numeric & uniq_nonNA > 2 ~ "Continuous",
      TRUE ~ "Other"))

#T3.1.3 Filter to only binary and continuous variables with sufficient non-missing data
cont_vars <- var_types %>% dplyr::filter(type == "Continuous", n_nonmiss > 1) %>% dplyr::pull(variable)
bin_vars <- var_types %>% dplyr::filter(type == "Binary", n_nonmiss > 0) %>% dplyr::pull(variable)

#T3.2 Create helper formatting functions
fmt_m_sd <- function(x) sprintf("%.2f (%.2f)", mean(x, na.rm = TRUE), stats::sd(x, na.rm = TRUE))
fmt_n_pct <- function(n1, den) sprintf("%d (%.1f%%)", n1, 100 * n1 / den)
fmt_ci <- function(lo, hi, digits = 2) sprintf("[%.2f, %.2f]", round(lo, digits), round(hi, digits))
fmt_p <- function(p) ifelse(is.na(p), "-", ifelse(p < .001, "<.001", sprintf("%.3f", p)))

#T3.3 Create pretty labels for the table
label_map <- tryCatch({
  lb <- safe_read(p_proc("risk_variable_labels.csv"))
  if (is.data.frame(lb) && all(c("variable","label") %in% names(lb)))
    setNames(lb$label, lb$variable) else NULL
}, error = function(e) NULL)

#T3.4 Analyze continuous risk variables: Cohen's d (HR vs LR) with 95% CI via effectsize; p from Welch t-test
analyze_cont <- function(v) {
  df <- rv0 %>% dplyr::filter(variable == v, !is.na(value)) %>%
    dplyr::mutate(cluster = factor(cluster, levels = c("Higher-Risk","Lower-Risk")))

  #T3.4.1 Require both clusters present
  if (length(unique(df$cluster)) < 2) return(NULL)

  #T3.4.2 Generate grouped descriptive stats
  gstats <- df %>%
    dplyr::group_by(cluster) %>%
    dplyr::summarise(m = mean(value, na.rm = TRUE), s = stats::sd(value, na.rm = TRUE), n = dplyr::n(), .groups="drop")
  m_hr <- gstats$m[gstats$cluster=="Higher-Risk"]; s_hr <- gstats$s[gstats$cluster=="Higher-Risk"]
  m_lr <- gstats$m[gstats$cluster=="Lower-Risk"];  s_lr <- gstats$s[gstats$cluster=="Lower-Risk"]

  #T3.4.3 Effect size: Cohen's d with Hedges' correction + 95% CI
  d_obj <- effectsize::cohens_d(value ~ cluster, data = df, pooled_sd = TRUE,
            hedges.correction = TRUE, ci = 0.95)
  d_val <- as.numeric(d_obj$Cohens_d)
  d_lo <- as.numeric(d_obj$CI_low)
  d_hi <- as.numeric(d_obj$CI_high)

  #T3.4.4 Calculate p-values from Welch t-test
  p_val <- tryCatch({
    broom::tidy(stats::t.test(value ~ cluster, data = df))$p.value[1]
  }, error = function(e) NA_real_)

  #T3.4.5 Build output row
  tibble::tibble(
    Type = "Continuous",
    Variable = if (!is.null(label_map)) dplyr::recode(v, !!!label_map) else v,
    `Lower-risk` = sprintf("%.2f (%.2f)", m_lr, s_lr),
    `Higher-risk` = sprintf("%.2f (%.2f)", m_hr, s_hr),
    `Effect (95% CI)` = sprintf("d = %.2f %s", d_val, fmt_ci(d_lo, d_hi)),
    `p` = fmt_p(p_val))
}

#T3.5 Analyze binary risk variables; create a function to calculate OR (Higher vs Lower) with 95% CI via effectsize on glm, p from glm
analyze_bin <- function(v) {
  df <- rv0 %>% dplyr::filter(variable == v, !is.na(value)) %>%
    dplyr::mutate(
      value = as.integer(value),
      cluster = factor(cluster, levels = c("Lower-Risk","Higher-Risk"))) # LR = ref

  #T3.5.1 Require both clusters present
  if (length(unique(df$cluster)) < 2) return(NULL)

  #T3.5.2 Generate grouped counts
  tab <- df %>%
    dplyr::count(cluster, value) %>%
    tidyr::complete(cluster = c("Lower-Risk","Higher-Risk"), value = c(0,1), fill = list(n = 0)) %>%
    tidyr::pivot_wider(names_from = value, values_from = n, names_prefix = "v") %>%
    dplyr::arrange(cluster)

  #T3.5.3 Extract counts for display
  lr_yes <- tab$v1[tab$cluster=="Lower-Risk"];  lr_den <- rowSums(tab[tab$cluster=="Lower-Risk", c("v0","v1")])
  hr_yes <- tab$v1[tab$cluster=="Higher-Risk"]; hr_den <- rowSums(tab[tab$cluster=="Higher-Risk", c("v0","v1")])

  #T3.5.4 Fit logistic regression model to get OR + CI + p-value
  fit <- stats::glm(value ~ cluster, data = df, family = stats::binomial())

  #T3.5.5 OR + CI from effectsize
  or_es <- tryCatch(effectsize::oddsratio(fit, ci = 0.95), error = function(e) NULL)
  if (!is.null(or_es)) {

    #T3.5.5.1 Extract OR + CI
    row <- which(grepl("clusterHigher-Risk", or_es$Parameter))
    or_val <- as.numeric(or_es$OR[row]); or_lo <- as.numeric(or_es$CI_low[row]); or_hi <- as.numeric(or_es$CI_high[row])
  } else {
    #T3.5.5.2 Extract OR + CI with broom if necessary
    tr <- broom::tidy(fit, conf.int = TRUE, exponentiate = TRUE)
    or_row <- tr[tr$term=="clusterHigher-Risk",]
    or_val <- or_row$estimate; or_lo <- or_row$conf.low; or_hi <- or_row$conf.high
  }
  #T3.5.6 Get p-value from model (Wald)
  p_val <- broom::tidy(fit)$p.value[broom::tidy(fit)$term=="clusterHigher-Risk"]

  #T3.5.7 Build output row
  tibble::tibble(
    Type = "Binary",
    Variable = if (!is.null(label_map)) dplyr::recode(v, !!!label_map) else v,
    `Lower-risk` = fmt_n_pct(lr_yes, lr_den),
    `Higher-risk` = fmt_n_pct(hr_yes, hr_den),
    `Effect (95% CI)` = sprintf("OR = %.2f %s", or_val, fmt_ci(or_lo, or_hi)),
    `p` = fmt_p(p_val))
}

#T3.6 Apply analysis functions to continuous and binary variables; combine results
cont_rows <- purrr::map(cont_vars, analyze_cont) |> purrr::compact()
bin_rows <- purrr::map(bin_vars, analyze_bin) |> purrr::compact()
t3 <- dplyr::bind_rows(cont_rows, bin_rows) |>
  dplyr::arrange(factor(Type, levels = c("Continuous","Binary")), Variable)

#T3.7 Create table caption and build gt table
t3_caption <- paste0(
  "**Table 3. Baseline risk-variable descriptive statistics by cluster**  \n",
  "Continuous: mean (SD); effect size is Cohen's *d* with Hedges' correction and 95% CI (Higher-Lower). ",
  "Binary: n (% “Yes”); effect size is odds ratio (Higher vs Lower) with 95% CI. ",
  "p-values from Welch *t*-test (continuous) and logistic regression Wald test (binary).")

#T3.8 Build gt table with formatting
t3_gt <- gt::gt(t3, groupname_col = "Type") |>
  gt::tab_caption(t3_caption) |>
  gt::cols_align(align = "center", columns = c(`Lower-risk`,`Higher-risk`,`Effect (95% CI)`,`p`))

#T3.9 Save the gt table
save_gt(t3_gt, "Table_03_Risk_Variables_by_Cluster")
t3_gt

```

## Table S1: Characteristics of participants in the BD DTH onset analysis

- Purpose: Describe the BD discrete-time hazard (onset) analysis sample by wave and cluster (participants and person-periods)
- Inputs: bd_pp (person-period), including baseline_status/event flags and covariates; uses ever-onset censoring per outcome
- Outputs: Saves Table_S1_Onset_Characteristics_by_Cluster as HTML/PNG/RTF; returns a gt table
- Notes: Applies model-ready filters consistent with survival analyses; reports demographics and interval onset counts per outcome

```{r table s1}

## 6. Table S1: Characteristics of participants in the onset analysis ##

#S1.1 Prepare person-period data for onset analysis results summarization 
#S1.1.1 Ensure BD person-period data is formatted correctly
bd_pp <- bd_pp %>% 
  mutate(
    participant_id = as.character(participant_id),
    family_id = factor(as.character(family_id)),
    site_factor = factor(site_factor),
    cluster = factor(cluster),
    start_wave = factor(start_wave),
    end_wave = factor(end_wave),
    outcome = factor(outcome),
    sex = factor(sex))

#S1.1.2 Refine pre-built data using a model-ready filter that mirrors QC "post" rules from analyses
bd_pp_model <- bd_pp %>%
  filter(event %in% c(0L,1L)) %>%
  filter(!is.na(cluster)) %>%
  filter(!is.na(age_mid), !is.na(sex), !is.na(site_factor), !is.na(family_id))

#S1.1.3 Ensure cluster, wave, sex, and site reference levels are correct
cluster_levels <- levels(bd_pp$cluster)
bd_pp_model <- bd_pp_model %>%
  mutate(
    cluster = forcats::fct_relevel(cluster, "C2"),
    end_wave = forcats::fct_relevel(end_wave, "ses-02A", "ses-04A", "ses-06A"),
    sex = forcats::fct_relevel(sex, "Female"),
    site_factor = forcats::fct_relevel(site_factor, "Site_1"))

#S1.1.4 Establish interval length in years (02A~2, 04A~4, 06A~6); handle missed waves
wave_year <- function(w) {
  as.numeric(stringr::str_extract(as.character(w), "\\d+"))
}

#S1.1.5 Add interval length (years) and log(interval length)
bd_pp_model <- bd_pp_model %>%
  mutate(dt_years = pmax(wave_year(end_wave) - wave_year(start_wave), 1),
    log_dt = log(dt_years))

#S1.2 Establish baseline BD status flags and primary risk set for first onset
#S1.2.1 IMPORTANT: do this BEFORE filtering out baseline-positive so the counts are complete - collapse to one row per participant x outcome x cluster, then count baseline-positive IDs
baseline_prev_ids <- bd_pp_model %>%
  dplyr::group_by(outcome, cluster, participant_id) %>%
  dplyr::summarise(
    baseline_pos = as.integer(any(baseline_status == 1, na.rm = TRUE)),
    .groups = "drop")

#S1.2.2 Build an ID-level baseline flag using the any_bsd outcome (covers BD-I/II/NOS)
baseline_any_bsd_ids <- bd_pp_model %>% 
  dplyr::filter(outcome == "any_bsd") %>%
  dplyr::group_by(participant_id) %>%
  dplyr::summarise(
    all_na = all(is.na(baseline_status)),
    baseline_any_bsd = dplyr::if_else(
      all_na, NA_integer_, as.integer(any(baseline_status == 1L, na.rm = TRUE))),
    .groups = "drop")

#S1.2.3 Join to all outcomes, do not assume 0 for missing data
bd_pp_model <- bd_pp_model %>%
  dplyr::left_join(baseline_any_bsd_ids %>% dplyr::select(participant_id, baseline_any_bsd),
    by = "participant_id")

#S1.2.4 Filter the dataset to contain participants that had not yet had BD
bd_pp_model <- bd_pp_model %>% dplyr::filter(baseline_any_bsd == 0L)

#S1.2.5 Censor after first onset (per outcome x participant); keep all intervals up to and including the first event==1 row; drop any later rows
bd_pp_model <- bd_pp_model %>%
  arrange(outcome, participant_id,
  if ("interval_index" %in% names(.)) interval_index else as.integer(end_wave)) %>%
  group_by(outcome, participant_id) %>%
  mutate(after_onset = dplyr::lag(cummax(event == 1L), default = FALSE)) %>%
  filter(!after_onset) %>%
  ungroup() %>%
  dplyr::select(-after_onset)

#S1.2.6 Establish empirical wave-median ages for plotting/prediction
wave_age_medians <- bd_pp_model %>%
  dplyr::group_by(end_wave) %>%
  dplyr::summarise(age_med = median(age_mid, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(match(end_wave, levels(bd_pp_model$end_wave)))
ages_empirical <- wave_age_medians$age_med %>% as.numeric() %>% round(1)

#S1.3 Create helper functions for Table S1 compilation
#S1.3.1 Map session codes to time labels
time_map <- c("ses-02A" = "2y", "ses-04A" = "4y", "ses-06A" = "6y")
time_levels <- c("2y","4y","6y")  # enforce order

#S1.3.2 Descriptive stat formatting functions
fmt_mean_sd <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) return("-")
  sprintf("%.2f \u00B1 %.2f", mean(x), stats::sd(x))
}
fmt_n <- function(n) ifelse(is.na(n), "-", format(as.integer(n), big.mark = ","))
fmt_n_pct <- function(n, den) {
  out <- ifelse(is.na(den) | den == 0,
    "0 (0.0%)",
    sprintf("%s (%.1f%%)", fmt_n(n), 100 * n / den))
  as.character(out)
}

#S1.4 Compile Table S1: Characteristics of participants in the onset analysis
#S1.4.1 Ensure exists and prepare person-period data for summarization
stopifnot(exists("bd_pp_model"))
pp <- bd_pp_model %>%
  dplyr::mutate(
    cluster_disp = coerce_cluster(cluster),
    time = dplyr::recode(end_wave, !!!time_map),
    time = factor(time, levels = time_levels)) %>%
  dplyr::filter(!is.na(time), cluster_disp %in% c("Lower-Risk","Higher-Risk"))

#S1.4.2 Demographics block (filter single outcome to avoid duplication & capture any BSD)
demo_base <- pp %>% dplyr::filter(outcome == "any_bsd")

#S1.4.3 Compute counts and age stats per time x cluster
demo_counts <- demo_base %>%
  dplyr::group_by(time, cluster_disp) %>%
  dplyr::summarise(
    n_pp = dplyr::n(),
    n_part = dplyr::n_distinct(participant_id),
    age_mean_sd = fmt_mean_sd(age_mid),
    .groups = "drop")

#S1.4.4 Compile sex at birth information
sex_tab <- demo_base %>%
  dplyr::mutate(sex2 = dplyr::case_when(
    !is.na(sex) & sex %in% c("Male","Female") ~ sex,
    TRUE ~ "Other/Unknown")) %>%
  dplyr::distinct(time, cluster_disp, participant_id, sex2) %>%
  dplyr::count(time, cluster_disp, sex2, name = "n_sex") %>%
  dplyr::left_join(demo_counts %>% dplyr::select(time, cluster_disp, n_part), by = c("time","cluster_disp")) %>%
  dplyr::mutate(val = fmt_n_pct(n_sex, n_part)) %>%
  dplyr::filter(sex2 %in% c("Male","Female")) %>%
  dplyr::mutate(Row = paste0("Sex at birth - ", sex2)) %>%
  dplyr::select(time, cluster_disp, Row, val)

#S1.4.5 Compile age information
age_row <- demo_counts %>%
  dplyr::transmute(time, cluster_disp, Row = "Age in years, Mean (SD)", val = age_mean_sd)

#S1.4.6 N rows (participants + person-periods)
n_rows <- demo_counts %>%
  tidyr::pivot_longer(c(n_pp, n_part), names_to = "what", values_to = "val_raw") %>%
  dplyr::mutate(
    Row = dplyr::case_when(
      what == "n_part" ~ "N (participants)",
      what == "n_pp" ~ "N (person-periods)"),
    val = fmt_n(val_raw)) %>%
  dplyr::select(time, cluster_disp, Row, val)

#S1.4.7 Bind demographics rows
demo_rows_cs <- dplyr::bind_rows(n_rows, age_row, sex_tab) %>%
  dplyr::mutate(Section = "Demographics")

#S1.4.8 Compute totals across clusters for demographics
demo_counts_tot <- demo_base %>%
  dplyr::group_by(time) %>%
  dplyr::summarise(
    n_pp = dplyr::n(),
    n_part = dplyr::n_distinct(participant_id),
    age_mean_sd = fmt_mean_sd(age_mid),
    .groups = "drop")

#S1.4.9 Compile information about whole sample
n_rows_total <- demo_counts_tot %>%
  tidyr::pivot_longer(c(n_pp, n_part), names_to = "what", values_to = "val_raw") %>%
  dplyr::mutate(
    Row = dplyr::case_when(
      what == "n_part" ~ "N (participants)",
      what == "n_pp" ~ "N (person-periods)"),
    val = fmt_n(val_raw),
    cluster_disp = "Total") %>%
  dplyr::select(time, cluster_disp, Row, val)

#S1.4.10 Age stats for whole sample
age_row_total <- demo_counts_tot %>%
  dplyr::transmute(
    time, cluster_disp = "Total",
    Row = "Age in years, Mean (SD)",
    val = age_mean_sd)

#S1.4.11 Sex stats for whole sample
sex_tab_total <- demo_base %>%
  dplyr::mutate(sex2 = dplyr::case_when(
    !is.na(sex) & sex %in% c("Male","Female") ~ sex,
    TRUE ~ "Other/Unknown")) %>%
  dplyr::distinct(time, participant_id, sex2) %>%
  dplyr::count(time, sex2, name = "n_sex") %>%
  dplyr::left_join(demo_counts_tot %>% dplyr::select(time, n_part), by = "time") %>%
  dplyr::mutate(val = fmt_n_pct(n_sex, n_part)) %>%
  dplyr::filter(sex2 %in% c("Male","Female")) %>%
  dplyr::mutate(
    cluster_disp = "Total",
    Row = paste0("Sex at birth - ", sex2)) %>%
  dplyr::select(time, cluster_disp, Row, val)

#S1.4.12 Bind total demographics rows for whole sample
demo_rows_total <- dplyr::bind_rows(n_rows_total, age_row_total, sex_tab_total) %>%
  dplyr::mutate(Section = "Demographics")

#S1.5 Compute interval onset events per outcome
#S1.5.1 Define outcome labels
outcome_labels <- c(
  "bipolar_I" = "Interval onset - BD I, n (%)",
  "bipolar_II" = "Interval onset - BD II, n (%)",
  "bipolar_NOS" = "Interval onset - BD NOS, n (%)",
  "bd_nos" = "Interval onset - BD NOS, n (%)",
  "any_bsd" = "Interval onset - Any BSD, n (%)")

#S1.5.2 Compute events per outcome x time x cluster
events_cs <- pp %>%
  dplyr::filter(outcome %in% names(outcome_labels)) %>%
  dplyr::group_by(outcome, time, cluster_disp) %>%
  dplyr::summarise(
    n_at_risk = dplyr::n(),
    n_events = sum(event == 1L, na.rm = TRUE),
    .groups = "drop") %>%
  dplyr::mutate(Row = dplyr::recode(outcome, !!!outcome_labels),
    val = fmt_n_pct(n_events, n_at_risk),
    Section = "Interval onsets")

#S1.5.3 Compute events per outcome x time for total sample
events_total <- pp %>%
  dplyr::filter(outcome %in% names(outcome_labels)) %>%
  dplyr::group_by(outcome, time) %>%
  dplyr::summarise(
    n_at_risk = dplyr::n(),
    n_events = sum(event == 1L, na.rm = TRUE),
    .groups = "drop") %>%
  dplyr::mutate(Row = dplyr::recode(outcome, !!!outcome_labels),
    val = fmt_n_pct(n_events, n_at_risk),
    cluster_disp = "Total",
    Section = "Interval onsets")

#S1.6 Assemble final long-format table and pivot to wide for GT
#S1.6.1 Combine all rows into long format
long_all <- dplyr::bind_rows(
  demo_rows_cs, demo_rows_total,
  events_cs %>% dplyr::select(time, cluster_disp, Row, val, Section),
  events_total %>% dplyr::select(time, cluster_disp, Row, val, Section)) %>%
  dplyr::mutate(
    cluster_disp = factor(cluster_disp, levels = c("Total","Lower-Risk","Higher-Risk")),
    time = factor(time, levels = time_levels))

#S1.6.2 Complete grid to avoid missing cells
grid <- tidyr::crossing(
  Row = unique(long_all$Row),
  Section = unique(long_all$Section),
  cluster_disp = levels(long_all$cluster_disp),
  time = levels(long_all$time))

#S1.6.3 Left-join to complete grid, fill missing with em-dash
long_all <- grid %>% dplyr::left_join(long_all, by = c("Row","Section","cluster_disp","time")) %>%
  dplyr::mutate(val = dplyr::coalesce(val, "\u2014")) # em-dash for empty

#S1.6.4 Pivot to wide format for GT table
wide <- long_all %>%
  dplyr::mutate(col_key = paste0(cluster_disp, "::", time)) %>%
  dplyr::select(Section, Row, col_key, val) %>%
  tidyr::pivot_wider(names_from = col_key, values_from = val)

#S1.6.5 Select and order columns for GT table
col_keys <- c(
  paste0("Total::", time_levels),
  paste0("Lower-Risk::", time_levels),
  paste0("Higher-Risk::", time_levels))
have <- intersect(col_keys, names(wide))
wide <- wide %>% dplyr::select(Section, Row, dplyr::all_of(have))

#S1.7 Create GT table
ts1 <- gt::gt(wide, groupname_col = "Section", rowname_col = "Row") |>

  #S1.7.1 Column labels show only time; spanners show the cluster
  gt::cols_label(.list = setNames(rep(time_levels, 3), have)) |>
  gt::tab_spanner(label = "Total", columns = dplyr::all_of(paste0("Total::", time_levels))) |>
  gt::tab_spanner(label = "Lower-Risk", columns = dplyr::all_of(paste0("Lower-Risk::", time_levels))) |>
  gt::tab_spanner(label = "Higher-Risk", columns = dplyr::all_of(paste0("Higher-Risk::", time_levels))) |>
  gt::cols_align(align = "center", columns = dplyr::all_of(have)) |>
  gt::tab_caption("**Table S1. Characteristics of Participants in the Onset of Bipolar Disorders by Risk Cluster Analysis**  
Counts are shown per assessment interval (2y, 4y, 6y). Demographics are from the *any BSD* risk set (one person-period per participant per interval). Interval onset rows report events within outcome-specific risk sets (post-censoring), shown as *n (percent)* of person-periods at risk.") |>
  gt::tab_options(table.font.size = gt::px(12)) |>
  gt::fmt_missing(columns = dplyr::everything(), missing_text = "\u2014")

#S1.8 Save + print Table S1
save_gt(ts1, "Table_S1_Onset_Characteristics_by_Cluster")
ts1

```

## Figure 2 and Related Stats: Interval and Cumulative Discrete Time Hazard Model Onset by Risk Cluster Incidence Curves for Bipolar Disorders 

- Purpose: Plot discrete-time hazards and cumulative incidence by wave and cluster with CIs; emit inline stats for Results text
- Inputs: Saved `glm` fits (`glm_primary_<outcome>.rds`), `bd_pp_model` (risk set), and `bd_dth_preds` (aligned hazards) when present
- Outputs: Saves `Figure_02_DTH_Hazards_and_Cumulative_by_Cluster.{pdf,png}`; prints Year-6 RD/RR lines and age-term summaries
- Notes: Computes cumulative risk CIs via MVN draws of fixed effects (or boot fallback); overlays observed empirical hazards/cumulative

```{r figure 2 stats}

## Figure 2: Discrete-time hazards and cumulative incidence by cluster ##

#F2.1 Load necessary model fits if not already in memory
#F2.1.1 Define outcomes to plot and fit directory
OUT_FIG <- c("bipolar_I","bipolar_II","bd_nos","any_bsd")

#F2.1.2 Directory where model fits are stored
fit_dir <- res_bd_surv

#F2.1.3 Safe read helper (returns NULL on error)
safe_read_rds <- function(path) tryCatch(readRDS(path), error = function(e) NULL)

#F2.1.4 Load finalized primary glm fits if not already in memory
if (!exists("fits_primary") || !is.list(fits_primary)) fits_primary <- list()
needed <- setdiff(OUT_FIG, names(fits_primary))

#F2.1.5 Load missing fits
if (length(needed)) {
  add <- setNames(
    lapply(needed, function(o) safe_read_rds(file.path(fit_dir, paste0("glm_primary_", o, ".rds")))),
    needed)
  fits_primary[names(add)] <- add
}

#F2.1.6 Report loaded fits and warn if any missing
loaded_ok <- names(Filter(Negate(is.null), fits_primary))
message("Loaded primary glm fits: ", paste(loaded_ok, collapse = ", "))
if (!all(OUT_FIG %in% loaded_ok)) {
  warning("Missing primary glm fits for: ", paste(setdiff(OUT_FIG, loaded_ok), collapse = ", "))
}

#F2.1.7 Prepare reference frame and model matrix for glm predictions
make_ref_frame_glm <- function(fit, dat) {
  frm <- model.frame(fit)
  mm0 <- model.matrix(fit)
  contr_all <- attr(mm0, "contrasts")
  
  get_levs <- function(var) {
    if (var %in% names(frm) && is.factor(frm[[var]])) levels(frm[[var]]) else NULL
  }
  
  waves_lev <- get_levs("end_wave") %||% levels(dat$end_wave)
  cluster_lev <- get_levs("cluster") %||% levels(dat$cluster)
  sex_lev <- get_levs("sex")
  site_lev <- get_levs("site_factor")
  base_lev <- get_levs("baseline_status")
  
  ref <- expand.grid(
    end_wave = waves_lev,
    cluster = cluster_lev,
    age_mid_between = 0,
    stringsAsFactors = FALSE)
  
  if ("age_mid_cwc" %in% names(frm)) ref$age_mid_cwc <- 0
  if ("age_mid_cwc_orth" %in% names(frm)) ref$age_mid_cwc_orth <- 0
  if ("wave_num" %in% names(frm)) ref$wave_num <- match(ref$end_wave, waves_lev) - 1L
  if ("sex" %in% names(frm) && !is.null(sex_lev)) ref$sex <- sex_lev[1]
  if ("site_factor" %in% names(frm) && !is.null(site_lev)) ref$site_factor <- site_lev[1]
  if ("baseline_status" %in% names(frm)) {
    ref$baseline_status <- if (!is.null(base_lev)) base_lev[1] else 0
  }
  if ("log_dt" %in% names(frm) && "log_dt" %in% names(dat)) {
    log_dt_ref <- dat %>%
      dplyr::group_by(end_wave) %>%
      dplyr::summarise(log_dt = median(log_dt, na.rm = TRUE), .groups = "drop")
    ref <- ref %>% dplyr::left_join(log_dt_ref, by = "end_wave")
  }
  
  #F2.1.7.1 Coerce factor levels to match fitted model
  for (nm in intersect(names(ref), names(frm))) {
    if (is.factor(frm[[nm]])) ref[[nm]] <- factor(ref[[nm]], levels = levels(frm[[nm]]))
  }
  
  tt <- stats::delete.response(stats::terms(fit))
  X <- stats::model.matrix(tt, ref, contrasts.arg = contr_all)
  
  beta <- stats::coef(fit)
  V <- as.matrix(stats::vcov(fit))
  
  miss <- setdiff(names(beta), colnames(X))
  if (length(miss) > 0) {
    X <- cbind(X, matrix(0, nrow(X), length(miss), dimnames = list(NULL, miss)))
  }
  X <- X[, names(beta), drop = FALSE]
  
  list(ref = ref, X = X, beta = beta, V = V, waves = waves_lev, clus = cluster_lev)
}

#F2.1.8 Estimate CIs with robust helper functions
#F2.1.8.1 Function to compute cumulative risk CIs from glm fixed-effect uncertainty
cumrisk_by_wave_ci_mvn_glm <- function(fit, dat, B = 4000, probs = c(0.025, 0.975)) {
  obj <- make_ref_frame_glm(fit, dat)
  ref <- obj$ref
  X <- obj$X
  beta <- obj$beta
  V <- obj$V
  
  if (requireNamespace("MASS", quietly = TRUE)) {
    beta_draws <- MASS::mvrnorm(n = B, mu = beta, Sigma = V)
  } else {
    L <- chol(V)
    Z <- matrix(stats::rnorm(B * length(beta)), nrow = B)
    beta_draws <- sweep(Z %*% t(L), 2, beta, `+`)
  }
  
  eta_draws <- X %*% t(beta_draws)
  p_draws <- plogis(eta_draws)
  
  ref$end_wave <- factor(ref$end_wave, levels = unique(as.character(ref$end_wave)))
  ref$cluster <- factor(ref$cluster, levels = unique(as.character(ref$cluster)))
  idx_by_clu <- split(seq_len(nrow(ref)), ref$cluster)
  
  out <- lapply(names(idx_by_clu), function(cl) {
    R <- idx_by_clu[[cl]]
    H <- p_draws[R, , drop = FALSE]
    CR <- apply(H, 2, function(col) 1 - cumprod(1 - col))
    if (is.null(dim(CR))) CR <- matrix(CR, nrow = length(R), ncol = 1)
    
    p_hat <- plogis(as.numeric(X[R, , drop = FALSE] %*% beta))
    cr_hat <- 1 - cumprod(1 - p_hat)
    
    lo <- apply(CR, 1, stats::quantile, probs = probs[1], na.rm = TRUE)
    hi <- apply(CR, 1, stats::quantile, probs = probs[2], na.rm = TRUE)
    
    tibble::tibble(
      end_wave = ref$end_wave[R],
      cluster = as.character(ref$cluster[R]),
      cumrisk = cr_hat,
      cumrisk_lo = lo,
      cumrisk_hi = hi)
  }) %>% 
    dplyr::bind_rows()
  
  out
}

#F2.1.8.2 Function to compute per-wave hazard CIs from glm fixed-effect uncertainty
hazard_by_wave_ci_mvn_glm <- function(fit, dat, B = 4000, probs = c(0.025, 0.975)) {
  obj <- make_ref_frame_glm(fit, dat)
  ref <- obj$ref
  X <- obj$X
  beta <- obj$beta
  V <- obj$V
  
  if (requireNamespace("MASS", quietly = TRUE)) {
    beta_draws <- MASS::mvrnorm(n = B, mu = beta, Sigma = V)
  } else {
    L <- chol(V)
    Z <- matrix(stats::rnorm(B * length(beta)), nrow = B)
    beta_draws <- sweep(Z %*% t(L), 2, beta, `+`)
  }
  
  eta_draws <- X %*% t(beta_draws)
  p_draws <- plogis(eta_draws)
  p_hat <- plogis(as.numeric(X %*% beta))
  
  tibble::tibble(
    end_wave = factor(ref$end_wave, levels = levels(dat$end_wave)),
    cluster = as.character(ref$cluster),
    hazard = p_hat,
    lo = apply(p_draws, 1, stats::quantile, probs = probs[1], na.rm = TRUE),
    hi = apply(p_draws, 1, stats::quantile, probs = probs[2], na.rm = TRUE))
}

#F2.1.9 Plotting helpers
#F2.1.9.1 Outcome order + pretty labels
OUT_FIG <- c("bipolar_I","bipolar_II","bd_nos","any_bsd")
OUT_LABS <- c(
  bipolar_I = "BD-I",
  bipolar_II = "BD-II",
  bd_nos = "BD-NOS",
  any_bsd = "Any BSD")

#F2.1.9.2 Set the order of levels for waves
wave_levels  <- c("ses-02A","ses-04A","ses-06A")

#F2.1.9.3 Pretty wave labels + median ages for tick labels
wave_pretty <- c("ses-02A" = "Year-2",
                 "ses-04A" = "Year-4",
                 "ses-06A" = "Year-6")

#F2.1.9.4 Explicit cluster colors for Figure 2
cluster_cols_f2 <- c(
  "Higher-Risk" = "#7F3C8D",
  "Lower-Risk" = "#11A579")

#F2.1.9.5 Axis theme with only axes, no gridlines
axes_only_theme <- function(base_size = 8, axis_text = 6.5) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid = element_blank(),
      axis.line = element_blank(),
      axis.line.x.bottom = element_line(color = "black"),
      axis.line.y.left = element_line(color = "black"),
      axis.ticks = element_line(color = "black"),
      axis.ticks.x.top = element_blank(),
      axis.ticks.y.right = element_blank(),
      axis.text = element_text(size = axis_text),
      legend.position = "bottom",
      legend.box = "horizontal",
      plot.margin = margin(6, 6, 6, 6),
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      legend.background = element_rect(fill = "white", color = NA),
      legend.box.background = element_rect(fill = "white", color = NA))
}

#F2.1.9.6 Helpers to blank axes on interior panels (use at composition time)
strip_x <- theme(axis.title.x = element_blank(),
                 axis.text.x = element_blank(),
                 axis.ticks.x = element_blank())
strip_y <- theme(axis.title.y = element_blank(),
                 axis.text.y = element_blank(),
                 axis.ticks.y = element_blank())

#F2.1.9.7 Robust cluster display helper
to_cluster_disp <- function(x) {
  x_chr <- as.character(x)
  
  out <- tryCatch(coerce_cluster(x_chr), error = function(e) rep(NA_character_, length(x_chr)))
  
  out[is.na(out) & x_chr %in% c("C1", "1")] <- "Lower-Risk"
  out[is.na(out) & x_chr %in% c("C2", "2")] <- "Higher-Risk"
  
  factor(out, levels = c("Higher-Risk", "Lower-Risk"))
}

#F2.2 Build hazard-by-wave directly from saved primary glm fits
haz_by_wave <- purrr::map_dfr(OUT_FIG, function(out) {
  fit <- fits_primary[[out]]
  if (is.null(fit)) return(tibble::tibble())
  
  dat <- bd_pp_model %>%
    dplyr::filter(outcome == out) %>%
    droplevels() %>%
    dplyr::mutate(
      end_wave = factor(end_wave, levels = wave_levels),
      cluster = factor(cluster))
  
  if (!nrow(dat)) return(tibble::tibble())
  
  hazard_by_wave_ci_mvn_glm(fit, dat, B = 2000) %>%
    dplyr::mutate(
      outcome = out,
      model = "logit",
      end_wave = factor(as.character(end_wave), levels = wave_levels),
      cluster = as.character(cluster),
      cluster_disp = to_cluster_disp(cluster),
      hazard = as.numeric(hazard),
      lo = as.numeric(lo),
      hi = as.numeric(hi)) %>%
    dplyr::filter(
      !is.na(cluster_disp),
      is.finite(hazard)) %>%
    dplyr::select(outcome, model, end_wave, cluster, cluster_disp, hazard, lo, hi)
})

message("haz_by_wave rows: ", nrow(haz_by_wave))
print(haz_by_wave %>% dplyr::count(outcome, cluster_disp))

cum_by_wave <- if (!is_tk(bd_cumrisk_by_wave)) {
  bd_cumrisk_by_wave %>%
    janitor::clean_names() %>%
    dplyr::mutate(
      outcome = norm_bd_outcome(outcome),
      model = dplyr::coalesce(model, "logit"),
      end_wave = factor(end_wave, levels = wave_levels),
      cluster = as.character(cluster),
      cluster_disp = factor(coerce_cluster(cluster), levels = names(cluster_cols_f2)))
} else {
  tibble::tibble(
    outcome = character(),
    model = character(),
    end_wave = factor(character(), levels = wave_levels),
    cluster = character(),
    cluster_disp = factor(character(), levels = c("Lower-Risk","Higher-Risk")),
    cumrisk = double(),
    cumrisk_lo = double(),
    cumrisk_hi = double())
}

#F2.2.1 If cumulative-risk CIs are absent in the exported table, backfill from saved glm fits
need_cum_ci <- (
  nrow(cum_by_wave) == 0 ||
    !"cumrisk_lo" %in% names(cum_by_wave) ||
    all(is.na(cum_by_wave$cumrisk_lo)))

if (need_cum_ci) {
  cum_ci_list <- lapply(OUT_FIG, function(out) {
    fit <- fits_primary[[out]]
    if (is.null(fit)) return(NULL)
    
    dat <- bd_pp_model %>%
      dplyr::filter(outcome == out) %>%
      droplevels() %>%
      dplyr::mutate(
        end_wave = factor(end_wave, levels = wave_levels),
        cluster = factor(cluster))
    
    if (!nrow(dat)) return(NULL)
    
    cumrisk_by_wave_ci_mvn_glm(fit, dat, B = 2000) %>%
      dplyr::mutate(
        outcome = out,
        model = "logit",
        end_wave = factor(end_wave, levels = wave_levels),
        cluster_disp = factor(coerce_cluster(cluster), levels = names(cluster_cols_f2))) %>%
      dplyr::select(outcome, model, end_wave, cluster, cluster_disp, cumrisk, cumrisk_lo, cumrisk_hi)
  })
  
  cum_by_wave <- dplyr::bind_rows(purrr::compact(cum_ci_list))
}

#F2.3 Prepare age median labels
#F2.3.1 Compute median ages per wave in the model data
age_meds_tbl <- bd_pp_model %>%
  dplyr::group_by(end_wave) %>%
  dplyr::summarise(age_med = median(age_mid, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(end_wave = as.character(end_wave))

#F2.3.2 Create named vector of median ages
age_meds_vec <- setNames(age_meds_tbl$age_med, age_meds_tbl$end_wave)

#F2.3.3 Create x-axis labels with median ages
x_lab_vec <- setNames(
  ifelse(is.finite(age_meds_vec[wave_levels]),
         sprintf("%s\nMedian Mid-Interval Age=%.1f", wave_pretty[wave_levels], age_meds_vec[wave_levels]),
         sprintf("%s\nMedian Mid-Interval Age=-", wave_pretty[wave_levels])),
  wave_levels)

#F2.4 Compute empirical hazards and cumulative incidence by wave and cluster
#F2.4.1 Empirical hazards by wave and cluster
emp_cum_by_wave <- bd_pp_model %>%
  dplyr::filter(outcome %in% OUT_FIG) %>%
  dplyr::mutate(
    cluster_disp = factor(coerce_cluster(cluster), levels = names(cluster_cols_f2)),
    end_wave = factor(end_wave, levels = wave_levels)) %>%
  
  #F2.4.1.1 Per-interval empirical hazard among those at risk
  dplyr::group_by(outcome, cluster_disp, end_wave) %>%
  dplyr::summarise(h = mean(event == 1L, na.rm = TRUE), .groups = "drop") %>%
  
  #F2.4.1.2 Roll up to cumulative incidence: 1 - Π(1 - h_j)
  dplyr::group_by(outcome, cluster_disp) %>%
  dplyr::arrange(end_wave, .by_group = TRUE) %>%
  dplyr::mutate(cumrisk_obs = 1 - cumprod(1 - pmin(pmax(h, 0), 1))) %>%
  dplyr::ungroup()

#F2.5 Plot predicted and observed hazard by wave and cluster
plot_haz_wave <- function(df, out_key, panel_letter = NULL) {
  
  d <- df %>%
    dplyr::filter(outcome == out_key, model == "logit") %>%
    dplyr::mutate(
      end_wave = factor(as.character(end_wave), levels = wave_levels),
      cluster_disp = factor(cluster_disp, levels = names(cluster_cols_f2))) %>%
    dplyr::arrange(cluster_disp, end_wave) %>%
    dplyr::filter(!is.na(cluster_disp), is.finite(hazard))
  
  if (is.null(panel_letter)) {
    i <- match(out_key, OUT_FIG)
    panel_letter <- LETTERS[i]
  }
  
  has_ci <- nrow(d) > 0 &&
    all(c("lo", "hi") %in% names(d)) &&
    any(is.finite(d$lo) & is.finite(d$hi), na.rm = TRUE)
  
  emp <- emp_cum_by_wave %>%
    dplyr::filter(outcome == out_key) %>%
    dplyr::mutate(
      end_wave = factor(as.character(end_wave), levels = wave_levels),
      cluster_disp = factor(cluster_disp, levels = names(cluster_cols_f2)))
  
  ggplot() +
    
    #F2.5.1 Add hazard CI ribbon behind the model-estimated lines
    { if (has_ci) geom_ribbon(
      data = d,
      aes(
        x = end_wave,
        ymin = lo,
        ymax = hi,
        fill = cluster_disp,
        group = cluster_disp),
      inherit.aes = FALSE,
      alpha = 0.15,
      color = NA) } +
    geom_line(
      data = d,
      aes(x = end_wave, y = hazard, color = cluster_disp, group = cluster_disp),
      linewidth = 0.7,
      na.rm = TRUE) +
    geom_point(
      data = d,
      aes(x = end_wave, y = hazard, color = cluster_disp),
      size = 1.4,
      na.rm = TRUE) +
    geom_point(
      data = emp,
      aes(x = end_wave, y = h, color = cluster_disp),
      inherit.aes = FALSE,
      shape = 1,
      size = 1.6) +
    scale_x_discrete(labels = x_lab_vec, drop = FALSE) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    scale_color_manual(
      values = cluster_cols_f2,
      breaks = names(cluster_cols_f2),
      labels = names(cluster_cols_f2),
      drop = FALSE) +
    scale_fill_manual(
      values = cluster_cols_f2,
      breaks = names(cluster_cols_f2),
      labels = names(cluster_cols_f2),
      guide = "none",
      drop = FALSE) +
    labs(
      title = paste0("(", panel_letter, ")  ", OUT_LABS[[out_key]], " Interval Hazard"),
      x = "Wave (Interval End)",
      y = "First-Onset Hazard",
      color = "Cluster") +
    axes_only_theme()
}

#F2.6 Plot cumulative incidence by wave and cluster
plot_cum_wave <- function(df, out_key) {
  d <- df %>%
    dplyr::filter(.data$outcome == out_key) %>%
    dplyr::mutate(
      end_wave = factor(as.character(end_wave), levels = wave_levels),
      cluster_disp = factor(cluster_disp, levels = names(cluster_cols_f2)))
  has_ci <- all(c("cumrisk_lo","cumrisk_hi") %in% names(d)) &&
    any(is.finite(d$cumrisk_lo) & is.finite(d$cumrisk_hi), na.rm = TRUE)
  i <- match(out_key, OUT_FIG)
  panel_letter <- LETTERS[i + 4]
  
  #F2.6.1 Observed cumulative for each outcome
  emp <- emp_cum_by_wave %>%
    dplyr::filter(outcome == out_key) %>%
    dplyr::mutate(
      end_wave = factor(as.character(end_wave), levels = wave_levels),
      cluster_disp = factor(cluster_disp, levels = names(cluster_cols_f2)))
  
  #F2.6.2 Plot cumulative incidence by wave and cluster
  ggplot(d, aes(x = end_wave, y = cumrisk, color = cluster_disp, group = cluster_disp)) +
    { if (has_ci) geom_ribbon(aes(ymin = cumrisk_lo, ymax = cumrisk_hi, fill = cluster_disp),
                              alpha = 0.15, color = NA) } +
    geom_line(size = 0.9) +
    geom_point(size = 1.8) +
    
    #F2.6.3 Overlay observed empirical cumulative incidence
    geom_point(data = emp,
               aes(x = end_wave, y = cumrisk_obs, color = cluster_disp),
               shape = 1, size = 1.8, inherit.aes = FALSE) +
    
    #F2.6.4 Scales and labels
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, 0.30)) +
    scale_x_discrete(drop = FALSE, labels = x_lab_vec) +
    scale_color_manual(
      values = cluster_cols_f2,
      breaks = names(cluster_cols_f2),
      labels = names(cluster_cols_f2),
      drop = FALSE) +
    scale_fill_manual(
      values = cluster_cols_f2,
      breaks = names(cluster_cols_f2),
      labels = names(cluster_cols_f2),
      guide = "none",
      drop = FALSE) +
    labs(title = paste0("(", panel_letter, ")  ", OUT_LABS[[out_key]], " Cumulative Incidence"),
         x = "Wave (Interval End)", y = "Cumulative Risk", color = "Cluster") +
    axes_only_theme()
}

#F2.7 Build combined by wave and cumulative plot panels
#F2.7.1 Generate hazard and cumulative panels for each outcome
haz_panels <- lapply(seq_along(OUT_FIG), function(i) {
  plot_haz_wave(haz_by_wave, OUT_FIG[i], panel_letter = LETTERS[i])
})
cum_panels <- lapply(OUT_FIG, function(o) plot_cum_wave(cum_by_wave, o))

#F2.7.2 Prepare the panel versions used in the final layout
haz_layout_panels <- list(
  haz_panels[[1]] + strip_x,
  haz_panels[[2]] + strip_x,
  haz_panels[[3]],
  haz_panels[[4]]
)

cum_layout_panels <- list(
  cum_panels[[1]] + strip_x,
  cum_panels[[2]] + strip_x,
  cum_panels[[3]],
  cum_panels[[4]]
)

#F2.7.3 Extract a shared legend with explicit white background
legend_f2 <- cowplot::get_legend(
  haz_panels[[1]] +
    theme(
      legend.position = "bottom",
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      legend.background = element_rect(fill = "white", color = NA),
      legend.box.background = element_rect(fill = "white", color = NA))
)

#F2.7.4 Remove legends from the actual panel plots
haz_layout_panels <- lapply(haz_layout_panels, function(p) p + theme(legend.position = "none"))
cum_layout_panels <- lapply(cum_layout_panels, function(p) p + theme(legend.position = "none"))

#F2.7.5 Align all 8 panels simultaneously so the top and bottom blocks share identical geometry
aligned_f2 <- cowplot::align_plots(
  plotlist = c(haz_layout_panels, cum_layout_panels),
  align = "hv",
  axis = "tblr",
  greedy = FALSE
)

#F2.7.6 Rebuild the top and bottom 2x2 grids from the already-aligned grobs
haz_grid <- cowplot::plot_grid(
  plotlist = aligned_f2[1:4],
  ncol = 2,
  rel_widths = c(1, 1),
  rel_heights = c(1, 1))

cum_grid <- cowplot::plot_grid(
  plotlist = aligned_f2[5:8],
  ncol = 2,
  rel_widths = c(1, 1),
  rel_heights = c(1, 1))

#F2.7.7 Stack the two aligned grids
fig2_main <- cowplot::plot_grid(
  haz_grid,
  cum_grid,
  ncol = 1,
  rel_heights = c(1, 1),
  align = "v",
  axis = "lr")

#F2.7.8 Wrap the legend in a white drawing surface
legend_plot_f2 <- cowplot::ggdraw() +
  cowplot::draw_grob(legend_f2, x = 0, y = 0, width = 1, height = 1) +
  theme(
    plot.background = element_rect(fill = "white", color = NA))

#F2.8 Combine figure + legend
fig2 <- cowplot::plot_grid(
  fig2_main,
  legend_plot_f2,
  ncol = 1,
  rel_heights = c(1, 0.08))

print(fig2)

#F2.9 Save Figure 2 with white background (avoid transparent legend band artifacts)
ggplot2::ggsave(filename = file.path(fig_dir, "Figure_02_DTH_Hazards_and_Cumulative_by_Cluster.pdf"), plot = fig2, width = 9.5, height = 11, bg = "white")

ggplot2::ggsave(filename = file.path(fig_dir, "Figure_02_DTH_Hazards_and_Cumulative_by_Cluster.png"), plot = fig2, width = 9.5, height = 11, dpi = 720, bg = "white")


## In Line Stats Related to BD DTH Analyses (Figure 2) ##

#F2.10 Helper functions for in-line stats
#F2.10.1 Create reference frame for predictions from fitted primary glm model
make_ref_frame <- function(fit, dat) {
    make_ref_frame_glm(fit = fit, dat = dat)
}

#F2.10.2 Cumulative risk from hazards helper
cumrisk_from_haz <- function(h, k) {
  h <- pmin(pmax(h, 1e-8), 1 - 1e-8)
  1 - exp(sum(log1p(-h[seq_len(k)])))
}

#F2.10.3 Wald CI helper
wald_ci <- function(theta_hat, grad, V, level = 0.95, log_scale = FALSE) {
  
  #F2.10.3.1 Small ridge in case of near-singular V
  if (any(!is.finite(V))) V[!is.finite(V)] <- 0
  V <- V + diag(1e-10, nrow(V))
  se <- sqrt(drop(t(grad) %*% V %*% grad))
  z  <- qnorm((1 + level)/2)
  if (log_scale) {
    lo <- theta_hat * exp(-z * se)
    hi <- theta_hat * exp(+z * se)
  } else {
    lo <- theta_hat - z * se
    hi <- theta_hat + z * se
  }
  list(se = se, lo = lo, hi = hi, z = theta_hat / se, p = 2 * pnorm(-abs(theta_hat / se)))
}

#F2.10.4 Year-6 cumulative risk difference and ratio from primary glm hazards
year6_rd_rr_delta <- function(fit, dat, wave_key = "ses-06A") {
  obj <- make_ref_frame(fit, dat)
  ref <- obj$ref; X <- obj$X; beta <- obj$beta; V <- obj$V
  waves <- obj$waves

  if (!wave_key %in% waves) wave_key <- tail(waves, 1)
  k_idx <- match(wave_key, waves)

  #F2.10.4.1 Map C1/C2 -> Higher/Lower using coerce_cluster()
  cl_levels <- levels(ref$cluster)
  disp <- coerce_cluster(cl_levels)
  role <- setNames(rep(NA_character_, length(cl_levels)), cl_levels)
  role[disp == "Higher-Risk"] <- "HR"
  role[disp == "Lower-Risk"] <- "LR"
  if (anyNA(role)) {
    
    #F2.10.4.2 Fallback: choose HR as the one with larger point CR
    point_cr <- sapply(cl_levels, function(cl) {
      rows <- which(ref$cluster == cl)
      h <- plogis(as.numeric(X[rows, , drop = FALSE] %*% beta))
      cumrisk_from_haz(h, k_idx)
    })
    role[] <- "LR"; role[names(which.max(point_cr))] <- "HR"
  }

  #F2.10.4.3 Functions to compute cumulative risk by cluster
  make_f <- function(cl) {
    rows <- which(ref$cluster == cl)
    function(b) {
      h <- plogis(as.numeric(X[rows, , drop = FALSE] %*% b))
      cumrisk_from_haz(h, k_idx)
    }
  }

  #F2.10.4.4 Cumulative risks, RD, RR, and their CIs
  f_HR <- make_f(names(role)[role == "HR"])
  f_LR <- make_f(names(role)[role == "LR"])
  CR_HR <- f_HR(beta); CR_LR <- f_LR(beta)
  CR_HR <- pmin(pmax(CR_HR, 1e-8), 1 - 1e-8)
  CR_LR <- pmin(pmax(CR_LR, 1e-8), 1 - 1e-8)
  RD <- CR_HR - CR_LR
  RR <- CR_HR / CR_LR
  g_HR <- numDeriv::grad(f_HR, beta)
  g_LR <- numDeriv::grad(f_LR, beta)
  g_RD  <- g_HR - g_LR
  ci_RD <- wald_ci(RD, g_RD, V, level = 0.95, log_scale = FALSE)

  #F2.10.4.5 Log-RR & CI
  g_logRR  <- g_HR / CR_HR - g_LR / CR_LR
  se_logRR <- sqrt(drop(t(g_logRR) %*% V %*% g_logRR))
  z <- qnorm(0.975)
  RR_lo <- RR * exp(-z * se_logRR)
  RR_hi <- RR * exp(+z * se_logRR)
  p_logRR <- 2 * pnorm(-abs(log(RR) / se_logRR))

  #F2.10.4.6 Return results
  list(
    wave = wave_key,
    CR_HR = CR_HR, CR_LR = CR_LR,
    RD = RD, RD_lo = ci_RD$lo, RD_hi = ci_RD$hi, RD_p = ci_RD$p,
    RR = RR, RR_lo = RR_lo, RR_hi = RR_hi, RR_p = p_logRR)
}

#F2.11 Compute Year-6 cumulative risk differences and ratios by outcome and store in a table
cum6 <- lapply(OUT_FIG, function(out) {
  fit <- fits_primary[[out]]; if (is.null(fit)) return(NULL)
  dat <- bd_pp_model %>%
    dplyr::filter(outcome == out) %>%
    droplevels() %>%
    dplyr::mutate(
        
        #F2.11.1 Ensure factors have correct levels
        end_wave = factor(end_wave, levels = wave_levels),
        cluster = factor(cluster),
        sex = factor(sex),
        site_factor = factor(site_factor))

  #F2.11.2 Compute Year-6 cumulative risk RD/RR
  res <- year6_rd_rr_delta(fit, dat, wave_key = "ses-06A")
  tibble::tibble(
    outcome = out,
    out_lab = OUT_LABS[out],
    wave = res$wave,
    HR = res$CR_HR, LR = res$CR_LR,
    RD = res$RD, RD_lo = res$RD_lo, RD_hi = res$RD_hi, RD_p = res$RD_p,
    RR = res$RR, RR_lo = res$RR_lo, RR_hi = res$RR_hi, RR_p = res$RR_p)
    }) %>% 
    dplyr::bind_rows()

#F2.12 Clean output table for in-line reporting
cum6_lines <- cum6 %>%
  mutate(
    line = glue::glue(
      "{out_lab} Year-6 cumulative risk: ",
      "{scales::percent(HR, 0.1)} (HR) vs {scales::percent(LR, 0.1)} (LR); ",
      "RD = {scales::percent(RD, 0.1)} [{scales::percent(RD_lo, 0.1)}, {scales::percent(RD_hi, 0.1)}], p = {sprintf('%.3f', RD_p)}; ",
      "RR = {sprintf('%.2f', RR)} [{sprintf('%.2f', RR_lo)}, {sprintf('%.2f', RR_hi)}], p = {sprintf('%.3f', RR_p)}")) %>% 
      pull(line)

#F2.13 Print cumulative risk difference/ratio lines
cat(paste("•", cum6_lines), sep = "\n")

#F2.14 Extract cluster and age effects for in-line reporting
#F2.14.1 Create a robust outcome-key helper for DTH summary tables
to_bd_outcome_key <- function(x) {
  key <- gsub("[^a-z0-9]", "", tolower(as.character(x)))

  dplyr::case_when(
    key %in% c("bdi", "bipolari") ~ "bipolar_I",
    key %in% c("bdii", "bipolarii") ~ "bipolar_II",
    key %in% c("bdnos", "bipolarnos") ~ "bd_nos",
    key %in% c("bsd", "anybsd", "anybipolarspectrumdisorder") ~ "any_bsd",
    TRUE ~ as.character(x))
}

#F2.14.2 Create a robust outcome-label helper for DTH summary tables
to_bd_out_lab <- function(x) {
  key <- to_bd_outcome_key(x)

  dplyr::case_when(
    key == "bipolar_I" ~ "BD-I",
    key == "bipolar_II" ~ "BD-II",
    key == "bd_nos" ~ "BD-NOS",
    key == "any_bsd" ~ "Any BSD",
    TRUE ~ as.character(x))
}

#F2.14.3 Create a helper to format p-values for inline text
fmt_p_inline <- function(p) {
  dplyr::case_when(
    is.na(p) ~ "-",
    p < .001 ~ "< .001",
    TRUE ~ paste0("= ", sprintf("%.3f", p)))
}

#F2.14.4 Determine whether the robust SE column is named se_method or se_type
se_col <- if ("se_method" %in% names(bd_dth_sum)) {
  "se_method"
} else if ("se_type" %in% names(bd_dth_sum)) {
  "se_type"
} else {
  stop("Could not find either 'se_method' or 'se_type' in bd_dth_sum.")
}

#F2.15 Primary cluster main effect = adjusted pooled interval-onset OR (Higher- vs Lower-Risk)
cluster_effects_primary <- bd_dth_sum %>%
  dplyr::filter(
    model == "logit_primary",
    .data[[se_col]] == "CR2",
    grepl("^cluster", term),
    !grepl(":", term)) %>%
  dplyr::mutate(
    outcome_key = to_bd_outcome_key(outcome),
    out_lab = to_bd_out_lab(outcome),
    OR = dplyr::coalesce(OR, exp(estimate)),
    OR_lo = dplyr::coalesce(ci_lo, exp(estimate - 1.96 * se)),
    OR_hi = dplyr::coalesce(ci_hi, exp(estimate + 1.96 * se)),
    pv = p.value) %>%
  dplyr::arrange(match(outcome_key, OUT_FIG))

cluster_lines_primary <- cluster_effects_primary %>%
  dplyr::transmute(
    line = glue::glue(
      "{out_lab}: primary adjusted Higher- vs Lower-Risk cluster OR = ",
      "{sprintf('%.2f', OR)} [{sprintf('%.2f', OR_lo)}, {sprintf('%.2f', OR_hi)}], ",
      "p {fmt_p_inline(pv)}.")) %>%
  dplyr::pull(line)

if (length(cluster_lines_primary) > 0) {
  cat(paste("•", cluster_lines_primary), sep = "\n")
  cat("\n")
}

#F2.16 Primary age term = between-person age only
age_effects_primary <- bd_dth_sum %>%
  dplyr::filter(
    model == "logit_primary",
    .data[[se_col]] == "CR2",
    term == "age_mid_between") %>%
  dplyr::mutate(
    outcome_key = to_bd_outcome_key(outcome),
    out_lab = to_bd_out_lab(outcome),
    OR = dplyr::coalesce(OR, exp(estimate)),
    pv = p.value,
    lbl = "between-person age") %>%
  dplyr::arrange(match(outcome_key, OUT_FIG))

age_lines_primary <- age_effects_primary %>%
  dplyr::transmute(
    line = glue::glue(
      "{out_lab}: primary {lbl} OR = {sprintf('%.2f', OR)} ",
      "(p {fmt_p_inline(pv)}).")) %>%
  dplyr::pull(line)

if (length(age_lines_primary) > 0) {
  cat(paste("•", age_lines_primary), sep = "\n")
}

#F2.16.1 Sensitivity note for within-person orthogonalized age
age_effects_within <- bd_dth_sum %>%
  dplyr::filter(
    model == "logit_within_orth",
    .data[[se_col]] == "CR2",
    term == "age_mid_cwc_orth") %>%
  dplyr::mutate(
    outcome_key = to_bd_outcome_key(outcome),
    out_lab = to_bd_out_lab(outcome),
    OR = dplyr::coalesce(OR, exp(estimate)),
    pv = p.value,
    lbl = "within-person age (orthogonalized sensitivity)") %>%
  dplyr::arrange(match(outcome_key, OUT_FIG))

if (nrow(age_effects_within) > 0) {
  age_lines_within <- age_effects_within %>%
    dplyr::transmute(
      line = glue::glue(
        "{out_lab}: sensitivity {lbl} OR = {sprintf('%.2f', OR)} ",
        "(p {fmt_p_inline(pv)}).")) %>%
    dplyr::pull(line)

  cat("\n")
  cat(paste("•", age_lines_within), sep = "\n")
}

```

## Table S2: Characteristics of Participants in the Prevalence of Bipolar Disorders Over Time by Risk Cluster Analysis

- Purpose: Describe the GEE prevalence sample by wave and cluster; report demographics and occurrence n (%) per outcome and totals
- Inputs: bd_panel (panel format) with status, age, sex, site, cluster, wave, outcome; produces bd_panel_model via model-ready filters
- Outputs: Saves Table_S2_Prevalence_Characteristics_by_Cluster as HTML/PNG/RTF; prints a gt table
- Notes: Denominator for occurrence rows excludes missing status; demographics taken from Any BSD panel per wave; clusters displayed as Higher-/Lower-Risk

```{r table s2}

## Table S2 - Prevalence (GEE) analysis characteristics ##

#S2.1 Prepare the BD panel dataset
#S2.1.1 Coerce key variables
bd_panel <- bd_panel %>%
  mutate(
    participant_id = as.character(participant_id),
    family_id = factor(as.character(family_id)),
    site_factor = factor(site_factor),
    cluster = factor(cluster),
    wave = factor(wave),
    outcome = factor(outcome),
    sex = factor(sex))

#S2.1.2 Require outcome responses to exist and be 0/1
bd_panel <- bd_panel %>% mutate("status" := as.integer(.data[["status"]]))

#S2.1.3 Model-ready filter (post-QC) data used for BD prevalence analyses
bd_panel_model <- bd_panel %>%
  filter(.data[["status"]] %in% c(0L,1L)) %>%
  filter(!is.na(cluster),
    !is.na(sex),
    !is.na(site_factor),
    !is.na(family_id),
    !is.na(wave)) %>%
  
  #S2.1.3.1 Require age fields present for prediction/interpretation
  filter(!is.na(age_wave), !is.na(age_wave_cwc), !is.na(age_wave_between))

#S2.1.4 Set relevant reference levels of factor variables
bd_panel_model <- bd_panel_model %>%
  mutate(
    cluster = forcats::fct_relevel(cluster, "C2"),
    wave = forcats::fct_relevel(wave, "ses-02A", "ses-04A", "ses-06A"),
    sex = forcats::fct_relevel(sex, "Female"),
    site_factor = forcats::fct_relevel(site_factor, "site01"))

#S2.1.5 Guards
#S2.1.5.1 Guard: ensure bd_panel_model exists and is a data frame
stopifnot(exists("bd_panel_model"), is.data.frame(bd_panel_model))

#S2.1.5.2 Guard: response var (default to "status" if params$response_var not present)
status_var <- if (exists("params") && !is.null(params$response_var)) params$response_var else "status"
stopifnot(status_var %in% names(bd_panel_model))

#S2.1.5.3 Again force outcomes to {0,1,NA}
bd_panel_model[[status_var]] <- as.integer(bd_panel_model[[status_var]])
bd_panel_model[[status_var]][!bd_panel_model[[status_var]] %in% c(0L,1L)] <- NA_integer_

#S2.1.6 Helper functions for Table S2
#S2.1.6.1 Set wave -> time labels (2y/4y/6y)
time_map <- c("ses-02A" = "2y", "ses-04A" = "4y", "ses-06A" = "6y")
time_levels <- c("2y","4y","6y")

#S2.1.6.2 Formatting helpers for descriptive stats
fmt_mean_sd <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) return("\u2014")
  sprintf("%.2f \u00B1 %.2f", mean(x), stats::sd(x))
}
fmt_n <- function(n) ifelse(is.na(n), "\u2014", format(as.integer(n), big.mark = ","))
fmt_n_pct <- function(n, den) {
  out <- ifelse(is.na(den) | den == 0,
    "0 (0.0%)",
    sprintf("%s (%.1f%%)", fmt_n(n), 100 * n / den))
  as.character(out)
}

#S2.1.7 Prepare panel data for Table S2
panel <- bd_panel_model %>%
  dplyr::mutate(
    cluster_disp = coerce_cluster(cluster),
    time = dplyr::recode(as.character(wave), !!!time_map),
    time = factor(time, levels = time_levels),
    sex = as.character(sex)) %>%
  dplyr::filter(!is.na(time), cluster_disp %in% c("Lower-Risk","Higher-Risk"))

#S2.2 Prepare table S2
#S2.2.1 Base dataset for demographics (any_bsd outcome)
demo_base <- panel %>% dplyr::filter(outcome == "any_bsd")

#S2.2.2 Establish N person-periods, N participants, Age
demo_counts <- demo_base %>%
  dplyr::group_by(time, cluster_disp) %>%
  dplyr::summarise(
    n_pp = dplyr::n(),
    n_part = dplyr::n_distinct(participant_id),
    age_mean_sd = fmt_mean_sd(age_wave),
    .groups = "drop")

#S2.2.3 Compute age stats
age_row <- demo_counts %>%
  dplyr::transmute(time, cluster_disp, Row = "Age in years, Mean (SD)", val = age_mean_sd)

#S2.2.4 Compute overall N rows and sex information
n_rows <- demo_counts %>%
  tidyr::pivot_longer(c(n_pp, n_part), names_to = "what", values_to = "val_raw") %>%
  dplyr::mutate(
    Row = dplyr::case_when(
      what == "n_part" ~ "N (participants)",
      what == "n_pp" ~ "N (person-periods)"),
    val = fmt_n(val_raw)) %>%
  dplyr::select(time, cluster_disp, Row, val)
sex_tab <- demo_base %>%
  dplyr::mutate(sex2 = dplyr::case_when(
    !is.na(sex) & sex %in% c("Male","Female") ~ sex,
    TRUE ~ "Other/Unknown")) %>%
  dplyr::distinct(time, cluster_disp, participant_id, sex2) %>%
  dplyr::count(time, cluster_disp, sex2, name = "n_sex") %>%
  dplyr::left_join(demo_counts %>% dplyr::select(time, cluster_disp, n_part), by = c("time","cluster_disp")) %>%
  dplyr::mutate(val = fmt_n_pct(n_sex, n_part)) %>%
  dplyr::filter(sex2 %in% c("Male","Female")) %>%
  dplyr::mutate(Row = paste0("Sex at birth - ", sex2)) %>%
  dplyr::select(time, cluster_disp, Row, val)

#S2.2.5 Combine demographic rows
demo_rows_cs <- dplyr::bind_rows(n_rows, age_row, sex_tab) %>%
  dplyr::mutate(Section = "Demographics")

#S2.2.6 Prepare total (across clusters) demographic rows
demo_counts_tot <- demo_base %>%
  dplyr::group_by(time) %>%
  dplyr::summarise(
    n_pp = dplyr::n(),
    n_part = dplyr::n_distinct(participant_id),
    age_mean_sd = fmt_mean_sd(age_wave),
    .groups = "drop")

#S2.2.7 Compute total N rows across whole sample
n_rows_total <- demo_counts_tot %>%
  tidyr::pivot_longer(c(n_pp, n_part), names_to = "what", values_to = "val_raw") %>%
  dplyr::mutate(
    Row = dplyr::case_when(
      what == "n_part" ~ "N (participants)",
      what == "n_pp" ~ "N (person-periods)"),
    val = fmt_n(val_raw),
    cluster_disp = "Total") %>%
  dplyr::select(time, cluster_disp, Row, val)

#S2.2.8 Compute total age row across whole sample
age_row_total <- demo_counts_tot %>%
  dplyr::transmute(time, cluster_disp = "Total", Row = "Age in years, Mean (SD)", val = age_mean_sd)

#S2.2.9 Compute total sex information across whole sample
sex_tab_total <- demo_base %>%
  dplyr::mutate(sex2 = dplyr::case_when(
    !is.na(sex) & sex %in% c("Male","Female") ~ sex,
    TRUE ~ "Other/Unknown")) %>%
  dplyr::distinct(time, participant_id, sex2) %>%
  dplyr::count(time, sex2, name = "n_sex") %>%
  dplyr::left_join(demo_counts_tot %>% dplyr::select(time, n_part), by = "time") %>%
  dplyr::mutate(val = fmt_n_pct(n_sex, n_part)) %>%
  dplyr::filter(sex2 %in% c("Male","Female")) %>%
  dplyr::mutate(cluster_disp = "Total", Row = paste0("Sex at birth - ", sex2)) %>%
  dplyr::select(time, cluster_disp, Row, val)

#S2.2.10 Combine total demographic rows across whole sample
demo_rows_total <- dplyr::bind_rows(n_rows_total, age_row_total, sex_tab_total) %>%
  dplyr::mutate(Section = "Demographics")

#S2.2.11 Prepare outcome occurrence rows
#S2.2.11.1 Outcome labels
occ_labels <- c(
  "bipolar_I" = "Occurrence - BD I, n (%)",
  "bipolar_II" = "Occurrence - BD II, n (%)",
  "bd_nos" = "Occurrence - BD NOS, n (%)",
  "any_bsd" = "Occurrence - Any BSD, n (%)")

#S2.2.11.2 Occurrence rows by cluster
occ_cs <- panel %>%
  dplyr::filter(outcome %in% names(occ_labels)) %>%
  dplyr::group_by(outcome, time, cluster_disp) %>%
  dplyr::summarise(
    n_obs = sum(!is.na(.data[[status_var]])),
    n_cases = sum(.data[[status_var]] == 1L, na.rm = TRUE),
    .groups = "drop") %>%
  dplyr::mutate(
    Row = dplyr::recode(outcome, !!!occ_labels),
    val = fmt_n_pct(n_cases, n_obs),
    Section = "Occurrence (wave prevalence)") %>%
  dplyr::select(time, cluster_disp, Row, val, Section)

#S2.2.11.3 Occurrence rows total across whole sample
occ_total <- panel %>%
  dplyr::filter(outcome %in% names(occ_labels)) %>%
  dplyr::group_by(outcome, time) %>%
  dplyr::summarise(
    n_obs = sum(!is.na(.data[[status_var]])),
    n_cases = sum(.data[[status_var]] == 1L, na.rm = TRUE),
    .groups = "drop") %>%
  dplyr::mutate(
    Row = dplyr::recode(outcome, !!!occ_labels),
    val = fmt_n_pct(n_cases, n_obs),
    cluster_disp = "Total",
    Section = "Occurrence (wave prevalence)") %>%
  dplyr::select(time, cluster_disp, Row, val, Section)

#S2.2.12 Combine all rows into long format
long_all <- dplyr::bind_rows(
  demo_rows_cs, demo_rows_total,
  occ_cs, occ_total) %>%
  dplyr::mutate(
    cluster_disp = factor(cluster_disp, levels = c("Total","Lower-Risk","Higher-Risk")),
    time = factor(time, levels = time_levels))

#S2.2.13 Create full grid of rows x sections x cluster_disp x time, fill missing with em-dash
grid <- tidyr::crossing(
  Row = unique(long_all$Row),
  Section = unique(long_all$Section),
  cluster_disp = levels(long_all$cluster_disp),
  time = levels(long_all$time))

#S2.3 Build wide table for GT
#S2.3.1 Left-join long_all to grid and fill missing data with em-dash
long_all <- grid %>%
  dplyr::left_join(long_all, by = c("Row","Section","cluster_disp","time")) %>%
  dplyr::mutate(val = dplyr::coalesce(val, "\u2014"))

#S2.3.2 Pivot to wide format with col_keys as cluster_disp::time
wide <- long_all %>%
  dplyr::mutate(col_key = paste0(cluster_disp, "::", time)) %>%
  dplyr::select(Section, Row, col_key, val) %>%
  tidyr::pivot_wider(names_from = col_key, values_from = val)

#S2.3.3 Select columns in desired order (if present)
#S2.3.3.1 Desired column keys
col_keys <- c(
  paste0("Total::", time_levels),
  paste0("Lower-Risk::", time_levels),
  paste0("Higher-Risk::", time_levels))

#S2.3.3.2 Keep only columns that exist in wide
have <- intersect(col_keys, names(wide))

#S2.3.3.3 Select only desired columns
wide <- wide %>% dplyr::select(Section, Row, dplyr::all_of(have))

#S2.4 Create GT table
#S2.4.1 Build GT table with spanners and labels
tS2 <- gt::gt(wide, groupname_col = "Section", rowname_col = "Row") |>
  gt::cols_label(.list = setNames(rep(time_levels, 3), have)) |>
  gt::tab_spanner(label = "Total", columns = dplyr::all_of(paste0("Total::", time_levels))) |>
  gt::tab_spanner(label = "Lower-Risk", columns = dplyr::all_of(paste0("Lower-Risk::", time_levels))) |>
  gt::tab_spanner(label = "Higher-Risk",columns = dplyr::all_of(paste0("Higher-Risk::", time_levels))) |>
  gt::cols_align(align = "center", columns = dplyr::all_of(have)) |>
  gt::tab_caption("**Table S2. Characteristics of Participants in the Prevalence of Bipolar Disorders Over Time by Risk Cluster (GEE Analysis)**
Counts are shown per assessment wave (2y, 4y, 6y). Demographics are derived from the *Any BSD* panel at each wave (one row per participant per wave). Occurrence rows report wave prevalence (status==1) within outcome-specific panels, shown as *n (percent)* of observed person-periods (denominator excludes missing).") |>
  gt::tab_options(table.font.size = gt::px(12)) |>
  gt::fmt_missing(columns = dplyr::everything(), missing_text = "\u2014")

#S2.4.2 Save table S2 (with fallback if save_gt not available)
if (exists("save_gt")) {
  save_gt(tS2, "Table_S2_Prevalence_Characteristics_by_Cluster")
} else {
  out_file <- file.path(getwd(), "Table_S2_Prevalence_Characteristics_by_Cluster.html")
  gt::gtsave(tS2, out_file)
  message("Saved: ", out_file)
}

#S2.4.3 Print GT table for Table S2
tS2

```

## Bipolar Disorder Odds by Risk Cluster Over Time stats

- Purpose: Produce per-wave Higher vs Lower cluster odds ratios (with 95% CI, p) from saved GEE interaction models
- Inputs: *_GEE_EXC_INT.rds models for outcomes (BD I/II/NOS/Any BSD); bd_panel_model used only for wave median age labels
- Outputs: Prints inline lines for manuscript
- Notes: Uses emmeans contrasts on the link scale and robust vcov from geeglm; truncates long CIs with arrows to share a common x-axis

```{r bd odds over time stats}

## Bipolar Disorder Odds by Risk Cluster Over Time ##

#OR.1 Setup
#OR.1.1 Directory for saved GEE fits
fit_dir <- "/home/exacloud/gscratch/NagelLab/staff/sam/projects/ABCD_MDS_Risk/results/main_analysis/2_bd_mixed_logit/rds"

#OR.1.2 Directory for each outcome's GEE EXC interaction fit
fit_paths <- c(
  bipolar_I = file.path(fit_dir, "bipolar_I_GEE_EXC_INT.rds"),
  bipolar_II = file.path(fit_dir, "bipolar_II_GEE_EXC_INT.rds"),
  bd_nos = file.path(fit_dir, "bd_nos_GEE_EXC_INT.rds"),
  any_bsd = file.path(fit_dir, "any_bsd_GEE_EXC_INT.rds"))

#OR.1.3 Create a modified safe read function for this section
safe_read_rds_gee <- function(p) tryCatch(readRDS(p), error = function(e) NULL)

#OR.1.4 Establish median age labels per wave
age_meds <- try({
  bd_panel_model %>%
    dplyr::group_by(wave) %>%
    dplyr::summarise(age_med = median(age_wave, na.rm = TRUE), .groups="drop") %>%
    dplyr::transmute(wave = as.character(wave), lab = sprintf("%s (median age %.1f)",
      dplyr::recode(wave, !!!wave_pretty), age_med))
}, silent = TRUE)

#OR.1.5 Fallback if age_meds computation fails
age_meds <- if (inherits(age_meds, "try-error")) tibble::tibble(wave = names(wave_pretty), lab = wave_pretty[names(wave_pretty)]) else age_meds

#OR.2 Helper: compute emmeans odds ratios (Higher- vs Lower-Risk) by wave from each GEE interaction model
em_or_by_wave <- function(fit) {

  #OR.2.1 Use robust vcov from the model (vcov.geeglm is robust by default) to compute emmeans
  V <- try(vcov(fit), silent = TRUE)
  emm <- emmeans::emmeans(fit, ~ cluster | wave, type = "link", vcov. = if (!inherits(V, "try-error")) V else NULL)
  
  #OR.2.2 Force contrast to be C2 - C1 (so OR = odds_C2 / odds_C1) regardless of reference
  lev <- levels(emm@grid$cluster)
  if (!all(c("C1","C2") %in% lev)) stop("Cluster levels not found in fit: ", paste(lev, collapse=", "))
  
  #OR.2.3 Compute contrast and summary
  w <- setNames(numeric(length(lev)), lev); w["C2"] <- 1; w["C1"] <- -1
  cmp <- emmeans::contrast(emm, method = list("C2 - C1" = w))
  sm  <- summary(cmp, infer = TRUE)  # on link (logit) scale
  
  #OR.2.4 Return tidy tibble with OR, CI, p per wave
  tibble::tibble(
    wave = as.character(sm$wave),
    OR = exp(sm$estimate),
    CI_low = exp(sm$lower.CL),
    CI_high= exp(sm$upper.CL),
    p = sm$p.value)
}

#OR.3 Load each GEE fit, compute emmeans OR contrasts, and combine into a single data frame
out_labs <- OUT_LABS
gee_pairs <- purrr::imap_dfr(fit_paths, function(fp, oc) {
  fit <- safe_read_rds_gee(fp)
  if (is.null(fit)) return(NULL)
  em_or_by_wave(fit) %>%
    dplyr::mutate(
      outcome = oc,
      out_lab = out_labs[oc],
      wave_pretty = dplyr::recode(wave, !!!wave_pretty)) %>%
    dplyr::left_join(age_meds, by = c("wave" = "wave")) %>%
    dplyr::mutate(wave_lab = ifelse(!is.na(lab), lab, wave_pretty))
})

#OR.3.1 Check if any results were computed, stop and provide a message why if not
if (!nrow(gee_pairs)) stop("Could not load any *_GEE_EXC_INT.rds fits or compute emmeans contrasts.")

#OR.4 Generate in-line stats for manuscript reporting
fmt_p <- function(p) ifelse(is.na(p), "-", ifelse(p < .001, "<.001", sprintf("%.3f", p)))
fmt_ci <- function(lo, hi) sprintf("%s-%s", sprintf("%.2f", lo), sprintf("%.2f", hi))
inline_lines <- gee_pairs %>%
  dplyr::mutate(line = glue("{out_lab} {wave_pretty}: OR={sprintf('%.2f', OR)}, 95% CI {fmt_ci(CI_low, CI_high)}, p={fmt_p(p)}")) %>%
  dplyr::arrange(factor(out_lab, levels = out_labs), factor(wave, levels = names(wave_pretty))) %>%
  dplyr::pull(line)

#OR.4.1 Print inline lines
cat("GEE in-line stats (Higher- vs Lower-Risk, emmeans, per wave):\n",
    paste0(" • ", inline_lines, collapse = "\n"), "\n\n")

```

## Missing-Data Imputation Sensitivity Inputs

- Purpose: Use the missForest missing-data sensitivity-analysis outputs that were read in during `#2. Inputs Configuration`
- Inputs: `miss_var_imp`, `imp_inclusion`, `little_mcar_imp`, and `kgrid_mf`
- Outputs: In-memory objects passed directly to Table S3, Table S4, and Figure S2
- Notes: All file discovery and reading occur upstream in the centralized inputs section so downstream supplement code remains display-focused

## Table S3 - Baseline Missingness Across the 17 Risk Variables Used for Clustering

- Purpose: Reproduce the current supplementary missingness table using the exact missForest sensitivity-analysis export
- Inputs: `miss_var_imp`
- Outputs: Saves `Table_S3_Baseline_Missingness_Clustering_Variables` as HTML/PNG/RTF in results/tables; returns a gt table
- Notes: The saved pct_miss field is a proportion, so it is displayed here as a percentage for publication-quality formatting

```{r table s3}

## Table S3: Baseline missingness across the 17 risk variables used for clustering ##

#S3.1 Prepare the missingness table using exact saved outputs
tS3_df <- miss_var_imp %>%
  dplyr::transmute(
    `Baseline Risk Variable` = dplyr::recode(feature, !!!risk_label_map_imp),
    `N Missing` = as.integer(n_miss),
    `Percent Missing` = as.numeric(pct_miss)) %>%
 dplyr::arrange(dplyr::desc(`Percent Missing`), `Baseline Risk Variable`)

#S3.2 Create table caption
tS3_caption <- paste0(
  "**Table S3. Baseline missingness across the 17 risk variables used for clustering**  \n",
  "Missingness is summarized across all baseline participants screened for the imputation sensitivity analysis. ",
  "Percent missing is displayed as a percentage of the baseline screening denominator.")

#S3.3 Build gt table
tS3_gt <- gt::gt(tS3_df) |>
  gt::fmt_number(columns = `N Missing`, decimals = 0, use_seps = TRUE) |>
  gt::fmt_percent(columns = `Percent Missing`, decimals = 2) |>
  gt::cols_align(
    align = "center",
    columns = c(`N Missing`, `Percent Missing`)) |>
  gt::tab_caption(tS3_caption) |>
  gt::tab_options(table.font.size = gt::px(12))

#S3.4 Save and print Table S3
save_gt(tS3_gt, "Table_S3_Baseline_Missingness_Clustering_Variables")
tS3_gt

```

## Table S4 - k-Prototypes Fit, Stability, and Anchor-Concordance Metrics for MissForest-Imputed Baseline Data Across Candidate k Solutions

- Purpose: Reproduce the current supplementary missForest k-grid summary table using the exact saved output from the imputation script
- Inputs: `kgrid_mf`
- Outputs: Saves `Table_S4_MissForest_KGrid_Metrics` as HTML/PNG/RTF in results/tables; returns a gt table
- Notes: This table uses only the missForest solution, consistent with the current supplement and revised analysis plan

```{r table s4}

## Table S4: k-prototypes fit, stability, and anchor-concordance metrics for missForest-imputed baseline data across candidate k solutions ##

#S4.1 Prepare the exact missForest k-grid summary table
tS4_df <- kgrid_mf %>%
  dplyr::transmute(
    `k Value` = as.integer(k),
    `Smallest Cluster n` = as.integer(min_cluster_n),
    `lambda` = as.numeric(lambda),
    `Mean Silhouette` = as.numeric(silhouette_huang),
    `Mean Jaccard` = as.numeric(jacc_mean),
    `ARI vs Anchor` = as.numeric(ari_vs_anchor_k2)) %>%
  dplyr::arrange(`k Value`)

#S4.2 Create caption
tS4_caption <- paste0(
  "**Table S4. k-prototypes fit, stability, and anchor-concordance metrics for missForest-imputed baseline data across candidate k solutions**  \n",
  "Candidate k-prototypes solutions were refit after missForest imputation of eligible baseline participants. ",
  "Mean Jaccard reflects bootstrap stability, and ARI reflects agreement with the original complete-case k=2 anchor partition.")

#S4.3 Build gt table
tS4_gt <- gt::gt(tS4_df) |>
  gt::fmt_number(columns = `k Value`, decimals = 0, use_seps = FALSE) |>
  gt::fmt_number(columns = `Smallest Cluster n`, decimals = 0, use_seps = TRUE) |>
  gt::fmt_number(
    columns = c(`lambda`, `Mean Silhouette`, `Mean Jaccard`, `ARI vs Anchor`),
    decimals = 3) |>
  gt::cols_align(
    align = "center",
    columns = c(`k Value`, `Smallest Cluster n`, `lambda`, `Mean Silhouette`, `Mean Jaccard`, `ARI vs Anchor`)) |>
  gt::tab_caption(tS4_caption) |>
  gt::tab_options(table.font.size = gt::px(12))

#S4.4 Save and print Table S4
save_gt(tS4_gt, "Table_S4_MissForest_KGrid_Metrics")
tS4_gt

```

## Figure S1 - MissForest-Imputed Clustering Sensitivity Metrics Across Candidate k Solutions

- Purpose: Visualize the saved missForest clustering-summary metrics across candidate k values
- Inputs: `kgrid_mf`
- Outputs: Saves `Figure_S1_MissForest_KGrid_Metrics.{pdf,png}` in results/figures; returns a ggplot object
- Notes: This is the figure supported by the saved outputs. The current imputation script does not save replicate-level silhouette or ARI distributions, so this figure summarizes the exported metric values across k rather than plotting distributions

```{r figure s1}

## Figure S1: MissForest-imputed clustering sensitivity metrics across candidate k solutions ##

#FS1.1 Reshape saved k-grid metrics into long plotting format
figS1_df <- dplyr::bind_rows(
  kgrid_mf %>%
    dplyr::transmute(
      k = as.integer(k),
      metric = "Mean silhouette",
      value = as.numeric(silhouette_huang),
      ymin = NA_real_,
      ymax = NA_real_),
  kgrid_mf %>%
    dplyr::transmute(
      k = as.integer(k),
      metric = "Mean Jaccard",
      value = as.numeric(jacc_mean),
      ymin = pmax(as.numeric(jacc_mean) - as.numeric(jacc_sd), 0),
      ymax = pmin(as.numeric(jacc_mean) + as.numeric(jacc_sd), 1)),
  kgrid_mf %>%
    dplyr::transmute(
      k = as.integer(k),
      metric = "ARI vs anchor",
      value = as.numeric(ari_vs_anchor_k2),
      ymin = NA_real_,
      ymax = NA_real_)) %>%
  dplyr::mutate(
    metric = factor(metric, levels = c("Mean silhouette", "Mean Jaccard", "ARI vs anchor")))

#FS1.2 Build a clean multi-panel summary figure
figS1 <- ggplot(figS1_df, aes(x = k, y = value)) +
  geom_vline(xintercept = 2, linetype = "dashed", linewidth = 0.6, alpha = 0.7) +
  geom_line(linewidth = 0.6) +
  geom_point(
    shape = 21,
    size = 3.0,
    stroke = 0.7,
    fill = "white") +
  scale_x_continuous(breaks = sort(unique(figS1_df$k))) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    expand = expansion(mult = c(0.015, 0.03))) +
  facet_wrap(
    ~ metric,
    ncol = 3,
    scales = "fixed",
    labeller = as_labeller(c(
      "Mean silhouette" = "A. Mean silhouette",
      "Mean Jaccard" = "B. Mean Jaccard",
      "ARI vs anchor" = "C. ARI vs anchor"))) +
  labs(
    title = "Figure S1. MissForest-imputed clustering sensitivity metrics across candidate k solutions",
    x = "Candidate number of clusters (k)",
    y = NULL) +
  theme_classic(base_size = 18) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 16, margin = margin(t = 2, b = 2)),
    plot.title = element_text(face = "bold", size = 14),
    axis.title.x = element_text(size = 14),
    axis.text = element_text(size = 12),
    panel.spacing = grid::unit(0.55, "lines"),
    plot.margin = margin(4, 8, 4, 8))

#FS1.3 Save and print Figure S1
save_fig(figS1, "Figure_S1_MissForest_KGrid_Metrics", width = 11.5, height = 7)
figS1

```

## Table S5 - Feature importance summary across MCR, centroid separability, and XGBoost + SHAP analyses and resulting minimally viable feature subset (MVFS) composition

- Purpose: Summarize the integrated feature-priority results used for parsimonious feature selection and identify which variables were retained in the final MVFS
- Inputs: `mvfs_priority_tbl`, `mvfs_selected_solutions`, and `risk_label_map_imp`
- Outputs: Saves `Table_S5_Feature_Importance_and_MVFS_Composition` as HTML/PNG/RTF in `results/tables`; returns a gt table
- Notes: The final MVFS composition is taken directly from the saved selected-solution RDS when available; if that object is unavailable, the code falls back to the first `mvfs_n` features by final priority rank

```{r table s5}

## Table S5: Feature importance summary and resulting MVFS composition ##

#S5.1 Determine the final retained MVFS feature set from the saved selected solution
#S5.1.1 Prefer the actual saved selected-solution object
mvfs_features_final <- character(0)

if (is.list(mvfs_selected_solutions) && length(mvfs_selected_solutions) >= 1) {
  sol1 <- mvfs_selected_solutions[[1]]

  if (is.list(sol1) && "features" %in% names(sol1)) {
    mvfs_features_final <- sol1$features %>% as.character()
  }
}

#S5.1.2 Fallback: if the selected-solution object is unavailable, use the selected MVFS size and the ranked priority table
if (!length(mvfs_features_final)) {
  mvfs_n_fallback <- mvfs_selected_summary %>%
    dplyr::filter(!is.na(mvfs_n)) %>%
    dplyr::slice(1) %>%
    dplyr::pull(mvfs_n)

  if (length(mvfs_n_fallback) == 1 && is.finite(mvfs_n_fallback)) {
    mvfs_features_final <- mvfs_priority_tbl %>%
      dplyr::arrange(rank_final_priority) %>%
      dplyr::slice_head(n = as.integer(mvfs_n_fallback)) %>%
      dplyr::pull(feature) %>%
      as.character()
  }
}

#S5.2 Prepare publication-ready Table S5 data
tS5_df <- mvfs_priority_tbl %>%
  dplyr::transmute(
    `Variable` = dplyr::recode(feature, !!!risk_label_map_imp),
    `MCR Mean` = as.numeric(mcr_mean),
    `|Δz|` = as.numeric(delta_z),
    `Max Cohen's d` = as.numeric(max_cohens_d),
    `Mean |SHAP|` = as.numeric(mean_abs_shap),
    `Ranked Importance` = as.integer(rank_final_priority),
    `Included in Final MVFS?` = dplyr::if_else(
      feature %in% mvfs_features_final,
      "Yes",
      "No",
      missing = "No")) %>%
  dplyr::arrange(`Ranked Importance`)

#S5.3 Create caption
tS5_caption <- paste0(
  "**Table S5. Feature importance summary across misclassification rate (MCR), centroid separability, and XGBoost + SHAP analyses and resulting minimally viable feature subset (MVFS) composition**  \n",
  "Final MVFS membership is taken from the saved selected parsimonious solution when available. ",
  "For binary features, centroid-separation statistics based on standardized mean separation are not applicable and are displayed as em dashes.")

#S5.4 Build gt table
tS5_gt <- gt::gt(tS5_df) |>
  gt::fmt_number(
    columns = c(`MCR Mean`, `|Δz|`, `Max Cohen's d`, `Mean |SHAP|`),
    decimals = 3) |>
  gt::fmt_number(
    columns = `Ranked Importance`,
    decimals = 0,
    use_seps = FALSE) |>
  gt::sub_missing(
    columns = c(`|Δz|`, `Max Cohen's d`),
    missing_text = "\u2014") |>
  gt::cols_align(
    align = "center",
    columns = c(`MCR Mean`, `|Δz|`, `Max Cohen's d`, `Mean |SHAP|`, `Ranked Importance`, `Included in Final MVFS?`)) |>
  gt::tab_caption(tS5_caption) |>
  gt::tab_options(table.font.size = gt::px(12))

#S5.5 Save and print Table S5
save_gt(tS5_gt, "Table_S5_Feature_Importance_and_MVFS_Composition")
tS5_gt

```

## Figure S2 - Feature-importance and parsimonious-feature-set sensitivity displays

- Purpose: Display the seed-level MCR distributions, the saved SHAP beeswarm plot, and the 2-D embedding of the selected MVFS solution with cluster centroids
- Inputs: `fi_seed_level`, `mvfs_priority_tbl`, `shap_beeswarm_png`, and `mvfs7_umap_k2`
- Outputs: Saves `Figure_S2_Feature_Importance_and_MVFS_Sensitivity.{pdf,png}` in `results/figures`; returns a combined figure object
- Notes: The SHAP beeswarm panel is embedded from the saved analysis figure because the feature-importance pipeline saved the visualization but not an exported observation-level SHAP matrix suitable for exact downstream reconstruction

```{r figure s2}

## Figure S2: Feature-importance and parsimonious-feature-set sensitivity displays ##

#FS2.1 Create a stable feature ordering shared across all three panels
#FS2.1.1 Order features by final integrated priority rank
feature_order_tbl <- mvfs_priority_tbl %>%
  dplyr::transmute(
    feature,
    feature_label = dplyr::recode(feature, !!!risk_label_map_imp),
    rank_final_priority = as.integer(rank_final_priority),
    mcr_mean = as.numeric(mcr_mean),
    mean_abs_shap = as.numeric(mean_abs_shap),
    sep_abs = as.numeric(sep_abs)) %>%
  dplyr::arrange(rank_final_priority)

feature_levels_plot <- rev(feature_order_tbl$feature_label)
feature_names_plot <- feature_order_tbl$feature
pretty_feature_order <- dplyr::recode(
  feature_names_plot,
  !!!risk_label_map_imp,
  .default = feature_names_plot)

#FS2.2 Create a shared clean theme for all panels
panel_theme_s2 <- function(base_size = 16) {
  theme_classic(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size - 1),
      plot.subtitle = element_text(size = base_size - 1),
      axis.title.x = element_text(size = base_size - 1),
      axis.title.y = element_text(size = base_size - 1.5),
      axis.text.x = element_text(size = base_size - 2),
      axis.text.y = element_text(size = base_size - 2),
      legend.title = element_text(size = base_size - 1),
      legend.text = element_text(size = base_size - 2),
      plot.margin = margin(6, 8, 6, 8))
}

#FS2.3 Panel A: seed-level MCR distributions across features
#FS2.3.1 Prepare plotting data
mcr_seed_plot_df <- fi_seed_level %>%
  dplyr::filter(!is.na(feature), !is.na(mcr_mean)) %>%
  dplyr::transmute(
    feature,
    feature_label = dplyr::recode(feature, !!!risk_label_map_imp),
    mcr_mean = as.numeric(mcr_mean),
    seed_idx = as.integer(seed_idx)) %>%
  dplyr::mutate(
    feature_label = factor(feature_label, levels = feature_levels_plot))

#FS2.3.2 Compute per-feature mean MCR for overlay
mcr_seed_means <- mcr_seed_plot_df %>%
  dplyr::group_by(feature_label) %>%
  dplyr::summarise(
    mcr_mean_overall = mean(mcr_mean, na.rm = TRUE),
    .groups = "drop")

#FS2.3.3 Build Panel A
p_mcr <- ggplot(mcr_seed_plot_df, aes(x = mcr_mean, y = feature_label)) +
  geom_violin(
    fill = "#D9D9D9",
    alpha = 0.75,
    color = NA,
    trim = TRUE,
    scale = "width") +
  geom_boxplot(
    width = 0.2,
    outlier.shape = NA,
    fill = "white",
    color = "black",
    linewidth = 0.4) +
  geom_point(
    data = mcr_seed_means,
    aes(x = mcr_mean_overall, y = feature_label),
    inherit.aes = FALSE,
    shape = 21,
    size = 2.7,
    stroke = 0.6,
    fill = "black",
    color = "black") +
  labs(
    title = "A. Permutation MCR Distributions",
    x = "Misclassification rate (MCR)",
    y = NULL) +
  panel_theme_s2()

#FS2.4 Panel B: ranked centroid separability display
#FS2.4.1 Prepare separability plotting data
sep_plot_df <- feature_order_tbl %>%
  dplyr::transmute(
    feature_label = factor(feature_label, levels = feature_levels_plot),
    sep_abs = as.numeric(sep_abs),
    sep_missing = is.na(sep_abs))

#FS2.4.2 Build Panel B
p_sep <- ggplot(sep_plot_df, aes(x = sep_abs, y = feature_label)) +
  geom_segment(
    data = sep_plot_df %>% dplyr::filter(!sep_missing),
    aes(x = 0, xend = sep_abs, y = feature_label, yend = feature_label),
    linewidth = 0.5,
    color = "grey55") +
  geom_point(
    data = sep_plot_df %>% dplyr::filter(!sep_missing),
    shape = 21,
    size = 2.7,
    stroke = 0.6,
    fill = "white",
    color = "black") +
  labs(
    title = "B. Centroid Separability",
    x = "|Δz|",
    y = NULL) +
  panel_theme_s2() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank())

#FS2.5 Panel C: exact SHAP beeswarm from the saved feature-importance analysis
if (inherits(fi_shapviz, "shapviz")) {

  sv_exact <- fi_shapviz

  #FS2.5.1 Apply pretty labels to the saved SHAP object
  old_names <- colnames(sv_exact$S)
  new_names <- dplyr::recode(
    old_names,
    !!!risk_label_map_imp,
    .default = old_names)

  colnames(sv_exact$S) <- new_names

  if (!is.null(sv_exact$X)) {
    if (is.data.frame(sv_exact$X)) {
      names(sv_exact$X) <- dplyr::recode(
        names(sv_exact$X),
        !!!risk_label_map_imp,
        .default = names(sv_exact$X))
    } else if (is.matrix(sv_exact$X)) {
      colnames(sv_exact$X) <- dplyr::recode(
        colnames(sv_exact$X),
        !!!risk_label_map_imp,
        .default = colnames(sv_exact$X))
    }
  }

  #FS2.5.2 Preserve the final hierarchical feature order used in Panels A and B
  keep_cols <- pretty_feature_order[pretty_feature_order %in% colnames(sv_exact$S)]
  sv_exact$S <- sv_exact$S[, keep_cols, drop = FALSE]

  if (!is.null(sv_exact$X)) {
    if (is.data.frame(sv_exact$X)) {
      keep_x <- keep_cols[keep_cols %in% names(sv_exact$X)]
      sv_exact$X <- sv_exact$X[, keep_x, drop = FALSE]
    } else if (is.matrix(sv_exact$X)) {
      keep_x <- keep_cols[keep_cols %in% colnames(sv_exact$X)]
      sv_exact$X <- sv_exact$X[, keep_x, drop = FALSE]
    }
  }

  #FS2.5.3 Build the exact SHAP beeswarm
  p_shap <- shapviz::sv_importance(
    sv_exact,
    kind = "beeswarm",
    max_display = ncol(sv_exact$S),
    show_numbers = FALSE,
    sort_features = FALSE) +
    ggplot2::scale_y_discrete(limits = feature_levels_plot) +
    ggplot2::scale_color_viridis_c(option = "inferno") +
    labs(
      title = "C. SHAP Importance Distributions",
      x = "SHAP value",
      y = NULL,
      color = "Relative feature\nvalue") +
    panel_theme_s2() +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      panel.background = element_rect(fill = "transparent", color = NA),
      plot.background = element_rect(fill = "transparent", color = NA),
      legend.background = element_rect(fill = "transparent", color = NA),
      legend.box.background = element_rect(fill = "transparent", color = NA))

} else {
  stop(
    "Exact Figure S2 SHAP reproduction requires ",
    "results/main_analysis/1_clustering/4_feature_importance/shap/xgb_surrogate_shapviz.rds. ",
    "Finish re-running 4_feature_importance.rmd, then knit again.")
}

#FS2.6 Combine panels into the final figure
figS2 <- p_mcr + p_sep + p_shap +
  patchwork::plot_layout(ncol = 3, widths = c(1.15, 0.85, 1.55)) +
  patchwork::plot_annotation(
    title = "Figure S2. Feature-importance and parsimonious-feature-set sensitivity displays",
    theme = theme(
      plot.title = element_text(face = "bold", size = 15)))

#FS2.7 Save and print Figure S2
save_fig(figS2, "Figure_S2_Feature_Importance_and_MVFS_Sensitivity", width = 18, height = 10, dpi = 720)
figS2

```

## Figure S3 - MVFS agreement and stability trace across candidate subset size

- Purpose: Show how agreement with the anchor partition and minimum clusterwise stability vary across candidate MVFS subset size for the selected primary search trace
- Inputs: `mvfs_trace_all` and `mvfs_selected_summary`
- Outputs: Saves `Figure_S3_MVFS_Trace_ARI_and_Jaccard.{pdf,png}` in `results/figures`; returns a combined figure object
- Notes: This figure intentionally focuses on the selected primary trace rather than retaining facet labels from the intermediate analysis plots, which is cleaner and more aligned with publication display standards

```{r figure s3}

## Figure S3: MVFS agreement and stability trace across candidate subset size ##

#FS3.1 Identify the selected primary MVFS trace
primary_mvfs_trace <- mvfs_selected_summary %>%
  dplyr::filter(!is.na(mvfs_n))

#FS3.1.1 Guard against multiple selected traces to avoid silently plotting the wrong one
if (nrow(primary_mvfs_trace) != 1) {
  stop(
    "Figure S3 expects exactly one selected MVFS trace in mvfs_selected_summary; found ",
    nrow(primary_mvfs_trace), ".")
}

#FS3.1.2 Extract the selected trace metadata
primary_search <- primary_mvfs_trace$search[[1]]
primary_lambda <- primary_mvfs_trace$lambda_mode[[1]]
primary_mvfs_n <- as.integer(primary_mvfs_trace$mvfs_n[[1]])

#FS3.2 Restrict the full trace table to the selected primary trace
mvfs_trace_plot_df <- mvfs_trace_all %>%
  dplyr::filter(
    search == primary_search,
    lambda_mode == primary_lambda) %>%
  dplyr::arrange(n_features)

mvfs_trace_point <- mvfs_trace_plot_df %>%
  dplyr::filter(n_features == primary_mvfs_n)

#FS3.3 Build Panel A: ARI trace
p_mvfs_ari <- ggplot(mvfs_trace_plot_df, aes(x = n_features, y = ari)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.9) +
  geom_hline(yintercept = 0.80, linetype = "dashed", linewidth = 0.4) +
  geom_vline(xintercept = primary_mvfs_n, linetype = "dashed", linewidth = 0.4) +
  geom_point(
    data = mvfs_trace_point,
    size = 3,
    shape = 21,
    stroke = 0.8,
    fill = "white") +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    expand = expansion(mult = c(0.02, 0.04))) +
  labs(
    title = "A. Agreement with anchor partition",
    x = NULL,
    y = "Adjusted Rand index") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"))

#FS3.4 Build Panel B: minimum Jaccard trace
p_mvfs_jacc <- ggplot(mvfs_trace_plot_df, aes(x = n_features, y = jaccard_min)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.9) +
  geom_hline(yintercept = 0.80, linetype = "dashed", linewidth = 0.4) +
  geom_vline(xintercept = primary_mvfs_n, linetype = "dashed", linewidth = 0.4) +
  geom_point(
    data = mvfs_trace_point,
    size = 3,
    shape = 21,
    stroke = 0.8,
    fill = "white") +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    expand = expansion(mult = c(0.02, 0.04))) +
  labs(
    title = "B. Minimum clusterwise stability",
    x = "Candidate subset size, |S| (number of features)",
    y = "Minimum Jaccard similarity") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"))

#FS3.5 Combine into final Figure S3
figS3 <- p_mvfs_ari / p_mvfs_jacc +
  patchwork::plot_layout(heights = c(1, 1)) +
  patchwork::plot_annotation(
    title = "Figure S3. Minimally viable feature subset (MVFS) agreement and stability trace across subset size",
    subtitle = glue::glue(
      "Primary trace: search = {primary_search}; lambda mode = {primary_lambda}; selected MVFS = {primary_mvfs_n}; thresholds shown at 0.80"),
    theme = theme(
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 10)))

#FS3.6 Save and print Figure S3
save_fig(figS3, "Figure_S3_MVFS_Trace_ARI_and_Jaccard", width = 9.5, height = 8.5)
figS3

```

## Table S6: Cross-algorithm clustering benchmark metrics across candidate k solutions

- Purpose: Summarize cross-algorithm benchmark results across candidate k solutions and identify the preferred solution within each clustering algorithm under the shared comparison framework
- Inputs: `alg_bench_metrics`, containing algorithm-specific benchmark outputs across candidate k values
- Outputs: Saves `Table_S6_Algorithm_Benchmark_Metrics` as HTML/PNG/RTF in results/tables; returns a gt table
- Notes: Standardizes algorithm labels across k-prototypes, KAMILA, and clustMD; uses mean Gower silhouette as the common primary comparison metric; bolds the preferred row within each algorithm; retains BIC descriptively for clustMD only because it is not available for the other algorithms

```{r table s6}

## Table S6: Cross-algorithm clustering benchmark metrics across candidate k solutions ##

#S6.1 Prepare the benchmark table
#S6.1.1 Clean and standardize algorithm labels
tS6_df0 <- alg_bench_metrics %>%
  janitor::clean_names() %>%
  dplyr::mutate(
    algorithm_key = stringr::str_to_lower(as.character(algorithm)),
    algorithm_disp = dplyr::case_when(
      algorithm_key %in% c("kproto", "k-prototypes", "kprototypes") ~ "k-prototypes",
      algorithm_key %in% c("kamila") ~ "KAMILA",
      algorithm_key %in% c("clustmd") ~ "clustMD",
      TRUE ~ as.character(algorithm)),
    algorithm_disp = factor(
      algorithm_disp,
      levels = c("k-prototypes", "KAMILA", "clustMD")),
    k = as.integer(k),
    bic = as.numeric(bic),
    gower_silhouette = as.numeric(gower_silhouette),
    ari_vs_anchor_k2 = as.numeric(ari_vs_anchor_k2),
    min_cluster_n = as.integer(min_cluster_n),
    max_cluster_n = as.integer(max_cluster_n)) %>%
  dplyr::group_by(algorithm_disp) %>%
  dplyr::arrange(dplyr::desc(gower_silhouette), k, .by_group = TRUE) %>%
  dplyr::mutate(preferred_k = dplyr::row_number() == 1L) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(algorithm_disp, k)

#S6.1.2 Build publication-ready display data
tS6_df <- tS6_df0 %>%
  dplyr::transmute(
    Algorithm = algorithm_disp,
    `k Value` = k,
    `BIC` = bic,
    `Mean Gower Silhouette` = gower_silhouette,
    `ARI vs Anchor` = ari_vs_anchor_k2,
    `Minimum Cluster n` = min_cluster_n,
    `Maximum Cluster n` = max_cluster_n,
    preferred_k = preferred_k)

#S6.1.3 Save row indices for preferred solutions so they can be emphasized in the table
preferred_rows_s6 <- which(tS6_df$preferred_k)

#S6.2 Create caption
tS6_caption <- paste0(
  "**Table S6. Cross-algorithm clustering benchmark metrics across candidate k solutions**  \n",
  "Candidate solutions from k-prototypes, KAMILA, and clustMD were compared on the same complete-case, z-scored mixed-data frame. ",
  "The common primary comparison metric was mean silhouette computed on Gower distance. ",
  "Within each algorithm, the preferred solution was defined as the candidate maximizing mean Gower silhouette, with smaller *k* used to break ties.")

#S6.3 Build gt table
tS6_gt <- gt::gt(
  tS6_df %>% dplyr::select(-preferred_k)) |>
  gt::fmt_number(
    columns = `k Value`,
    decimals = 0,
    use_seps = FALSE) |>
  gt::fmt_number(
    columns = `BIC`,
    decimals = 3,
    use_seps = TRUE) |>
  gt::fmt_number(
    columns = c(`Mean Gower Silhouette`, `ARI vs Anchor`),
    decimals = 3) |>
  gt::fmt_number(
    columns = c(`Minimum Cluster n`, `Maximum Cluster n`),
    decimals = 0,
    use_seps = TRUE) |>
  gt::sub_missing(
    columns = `BIC`,
    missing_text = "\u2014") |>
  gt::cols_align(
    align = "center",
    columns = c(
      `Algorithm`,
      `k Value`,
      `BIC`,
      `Mean Gower Silhouette`,
      `ARI vs Anchor`,
      `Minimum Cluster n`,
      `Maximum Cluster n`)) |>
  gt::tab_caption(tS6_caption) |>
  gt::tab_source_note(
  gt::md(paste(
    "Rows shown in **bold** indicate the preferred solution within each algorithm under the shared Gower-silhouette framework.",
    "ARI is a descriptive agreement metric relative to the original k-prototypes *k*=2 anchor partition and was not the optimization target.",
    "BIC was available only for clustMD and is shown descriptively.",
    sep = "  \n"))) |>
  gt::tab_options(table.font.size = gt::px(12))

#S6.4 Emphasize preferred rows
tS6_gt <- tS6_gt |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_body(
      rows = preferred_rows_s6,
      columns = dplyr::everything()))

#S6.5 Save and print Table S6
save_gt(tS6_gt, "Table_S6_Algorithm_Benchmark_Metrics")
tS6_gt

```

## Figure S4: 

- Purpose: Visualize how the shared benchmark metrics vary across candidate k values for each clustering algorithm
- Inputs: `alg_bench_metrics`
- Outputs: Saves `Figure_S4_Algorithm_Benchmark_Metrics.{pdf,png}` in results/figures; returns a combined figure object
- Notes: Panel A displays mean Gower silhouette, the shared primary cross-algorithm comparison metric; Panel B displays ARI versus the original k-prototypes k=2 anchor partition as a descriptive concordance metric; open circles mark the preferred solution within each algorithm under the shared Gower-silhouette framework

```{r figure s4}

## Figure S4: Cross-algorithm benchmark sensitivity metrics across candidate k solutions ##

#FS4.1 Prepare plotting data
alg_bench_plot_df <- alg_bench_metrics %>%
  janitor::clean_names() %>%
  dplyr::mutate(
    algorithm_key = stringr::str_to_lower(as.character(algorithm)),
    algorithm_disp = dplyr::case_when(
      algorithm_key %in% c("kproto", "k-prototypes", "kprototypes") ~ "k-prototypes",
      algorithm_key %in% c("kamila") ~ "KAMILA",
      algorithm_key %in% c("clustmd") ~ "clustMD",
      TRUE ~ as.character(algorithm)),
    algorithm_disp = factor(
      algorithm_disp,
      levels = c("k-prototypes", "KAMILA", "clustMD")),
    k = as.integer(k),
    gower_silhouette = as.numeric(gower_silhouette),
    ari_vs_anchor_k2 = as.numeric(ari_vs_anchor_k2)) %>%
  dplyr::arrange(algorithm_disp, k)

#FS4.1.1 Identify preferred solutions under the shared Gower-silhouette rule
alg_bench_pref_df <- alg_bench_plot_df %>%
  dplyr::group_by(algorithm_disp) %>%
  dplyr::arrange(dplyr::desc(gower_silhouette), k, .by_group = TRUE) %>%
  dplyr::slice(1) %>%
  dplyr::ungroup()

#FS4.2 Set colors and theme
alg_cols_s4 <- setNames(
  viridisLite::viridis(3, option = "D"),
  c("k-prototypes", "KAMILA", "clustMD"))

panel_theme_s4 <- function(base_size = 12) {
  theme_classic(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "bottom",
      legend.title = element_text(size = base_size - 1),
      legend.text = element_text(size = base_size - 1),
      axis.title = element_text(size = base_size),
      axis.text = element_text(size = base_size - 1),
      plot.margin = margin(6, 8, 6, 8))
}

#FS4.3 Panel A: shared primary benchmark metric
p_alg_sil <- ggplot(
  alg_bench_plot_df,
  aes(x = k, y = gower_silhouette, color = algorithm_disp)) +
  geom_vline(
    xintercept = 2,
    linetype = "dashed",
    linewidth = 0.4,
    alpha = 0.7) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.9) +
  geom_point(
    data = alg_bench_pref_df,
    shape = 21,
    size = 2.8,
    stroke = 0.8,
    fill = "white") +
  scale_x_continuous(
    breaks = sort(unique(alg_bench_plot_df$k))) +
  scale_color_manual(
    values = alg_cols_s4,
    breaks = names(alg_cols_s4),
    labels = names(alg_cols_s4)) +
  labs(
    title = "A. Mean Gower silhouette by candidate k",
    x = "Candidate number of clusters (k)",
    y = "Mean Gower silhouette",
    color = "Algorithm") +
  panel_theme_s4()

#FS4.4 Panel B: descriptive agreement with the anchor partition
p_alg_ari <- ggplot(
  alg_bench_plot_df,
  aes(x = k, y = ari_vs_anchor_k2, color = algorithm_disp)) +
  geom_vline(
    xintercept = 2,
    linetype = "dashed",
    linewidth = 0.4,
    alpha = 0.7) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.9) +
  geom_point(
    data = alg_bench_pref_df,
    aes(x = k, y = ari_vs_anchor_k2, color = algorithm_disp),
    shape = 21,
    size = 2.8,
    stroke = 0.8,
    fill = "white",
    inherit.aes = FALSE) +
  scale_x_continuous(
    breaks = sort(unique(alg_bench_plot_df$k))) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    expand = expansion(mult = c(0.02, 0.04))) +
  scale_color_manual(
    values = alg_cols_s4,
    breaks = names(alg_cols_s4),
    labels = names(alg_cols_s4)) +
  labs(
    title = "B. ARI vs original k-prototypes k=2 anchor",
    x = "Candidate number of clusters (k)",
    y = "Adjusted Rand index",
    color = "Algorithm") +
  panel_theme_s4()

#FS4.5 Combine into final Figure S4
figS4 <- p_alg_sil + p_alg_ari +
  patchwork::plot_layout(guides = "collect", ncol = 2) +
  patchwork::plot_annotation(
    title = "Figure S4. Cross-algorithm benchmark sensitivity metrics across candidate k solutions",
    subtitle = paste0(
      "Panel A shows the shared primary cross-algorithm comparison metric. ",
      "Panel B shows descriptive agreement with the original k-prototypes k=2 anchor partition. ",
      "Open circles indicate the preferred solution within each algorithm under the shared Gower-silhouette framework."),
    theme = theme(
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 10))) &
  theme(legend.position = "bottom")

#FS4.6 Save and print Figure S4
save_fig(figS4, "Figure_S4_Algorithm_Benchmark_Metrics", width = 12.5, height = 5.4)
figS4

```

## Figure S5: Common-comorbidity sensitivity-analysis odds ratios by wave

- Purpose: Display the wave-specific higher-risk vs lower-risk cluster odds ratios for the common-comorbidity sensitivity-analysis contrasts emphasized in the supplement text
- Inputs: `comorb_or_by_wave`, containing emmeans-derived cluster contrasts by wave from the common-comorbidity GEE models
- Outputs: Saves `Figure_S5_Common_Comorbidity_Sensitivity_ORs.{pdf,png}` in results/figures; returns a combined figure object
- Notes: Restricts display to the 3 contrasts reported in the supplement results; uses a log-scale odds-ratio axis with a reference line at OR=1; Panel A shows contrasts versus no diagnosis, and Panel B shows the direct pairwise comparison between BSD plus comorbidity and other diagnosis only

```{r figure s5}

## Figure S5: Common-comorbidity sensitivity-analysis odds ratios by wave ##

#FS5.1 Guard against placeholder input
if (is_tk(comorb_or_by_wave)) {
  stop(
    "Figure S5 requires results/main_analysis/2_statistical_analysis/5_comorbidity_sensitivity_analysis/tables/gee_or_by_wave.csv. ",
    "Run 5_comorbidity_sensitivity_analysis.Rmd first, then knit again.")
}

#FS5.2 Prepare plotting data
#FS5.2.1 Keep only the contrasts reported in the supplementary results
figS5_df <- comorb_or_by_wave %>%
  janitor::clean_names() %>%
  dplyr::filter(outcome %in% c(
    "other_only_vs_nodx",
    "bsd_plus_other_vs_nodx",
    "bsd_plus_vs_other_only")) %>%
  dplyr::mutate(
    outcome_label = dplyr::recode(
      outcome,
      "other_only_vs_nodx" = "Other diagnosis only vs no diagnosis",
      "bsd_plus_other_vs_nodx" = "BSD plus any other diagnosis vs no diagnosis",
      "bsd_plus_vs_other_only" = "BSD plus any other diagnosis vs other diagnosis only"),
    panel_label = dplyr::case_when(
      outcome %in% c("other_only_vs_nodx", "bsd_plus_other_vs_nodx") ~ "A. Contrasts versus no diagnosis",
      outcome == "bsd_plus_vs_other_only" ~ "B. Direct pairwise comparison"),
    wave_label = dplyr::recode(
      as.character(wave),
      "ses-02A" = "Year-2",
      "ses-04A" = "Year-4",
      "ses-06A" = "Year-6"),
    wave_label = factor(wave_label, levels = c("Year-6", "Year-4", "Year-2")),
    outcome_label = factor(
      outcome_label,
      levels = c(
        "Other diagnosis only vs no diagnosis",
        "BSD plus any other diagnosis vs no diagnosis",
        "BSD plus any other diagnosis vs other diagnosis only")),
    panel_label = factor(
      panel_label,
      levels = c(
        "A. Contrasts versus no diagnosis",
        "B. Direct pairwise comparison")),
    or = as.numeric(or),
    or_lcl = as.numeric(or_lcl),
    or_ucl = as.numeric(or_ucl),
    p = as.numeric(p))

#FS5.2.2 Save a clean plotting/stats file for auditing if desired
readr::write_csv(
  figS5_df,
  file.path(tab_dir, "Figure_S5_Common_Comorbidity_Sensitivity_ORs_SourceData.csv"))

#FS5.2.3 Define a shared linear x-axis range across both panels
fs5_xmax <- max(figS5_df$or_ucl, na.rm = TRUE)
fs5_xmax <- ceiling(fs5_xmax / 5) * 5
fs5_x_breaks <- seq(0, fs5_xmax, by = 2.5)

#FS5.3 Shared theme helper
forest_theme_s5 <- function(base_size = 11, strip_size = 9.5) {
  ggplot2::theme_classic(base_size = base_size) +
    ggplot2::theme(
      panel.grid = ggplot2::element_blank(),
      axis.title.y = ggplot2::element_blank(),
      axis.text.y = ggplot2::element_text(size = base_size - 1),
      axis.title.x = ggplot2::element_text(size = base_size),
      axis.text.x = ggplot2::element_text(size = base_size - 1),
      strip.background = ggplot2::element_blank(),
      strip.text.y.left = ggplot2::element_text(
        face = "bold",
        angle = 0,
        size = strip_size,
        margin = ggplot2::margin(r = 8)),
      strip.text = ggplot2::element_text(
        face = "bold",
        size = strip_size),
      strip.placement = "outside",
      plot.title = ggplot2::element_text(face = "bold", size = base_size + 0.5),
      plot.subtitle = ggplot2::element_text(size = base_size - 0.5),
      plot.margin = ggplot2::margin(6, 8, 6, 8))
}

#FS5.4 Build Panel A: contrasts versus no diagnosis
figS5_A_df <- figS5_df %>%
  dplyr::filter(panel_label == "A. Contrasts versus no diagnosis")

p_figS5_A <- ggplot2::ggplot(
  figS5_A_df,
  ggplot2::aes(x = or, y = wave_label)) +
  ggplot2::geom_vline(
    xintercept = 1,
    linetype = "dashed",
    linewidth = 0.4,
    color = "grey40") +
  ggplot2::geom_segment(
    ggplot2::aes(x = or_lcl, xend = or_ucl, yend = wave_label),
    linewidth = 0.4,
    color = "black") +
  ggplot2::geom_point(
    size = 2.2,
    color = "#7F3C8D") +
  ggplot2::scale_x_continuous(
    limits = c(0, fs5_xmax),
    breaks = fs5_x_breaks,
    expand = expansion(mult = c(0, 0.01))) +
  ggplot2::facet_grid(
    outcome_label ~ .,
    switch = "y",
    labeller = ggplot2::label_wrap_gen(width = 30)) +
  ggplot2::labs(
    title = "A. Contrasts versus no diagnosis",
    x = NULL) +
  forest_theme_s5() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank())

#FS5.5 Build Panel B: direct pairwise comparison
figS5_B_df <- figS5_df %>%
  dplyr::filter(panel_label == "B. Direct pairwise comparison")

p_figS5_B <- ggplot2::ggplot(
  figS5_B_df,
  ggplot2::aes(x = or, y = wave_label)) +
  ggplot2::geom_vline(
    xintercept = 1,
    linetype = "dashed",
    linewidth = 0.4,
    color = "grey40") +
  ggplot2::geom_segment(
    ggplot2::aes(x = or_lcl, xend = or_ucl, yend = wave_label),
    linewidth = 0.4,
    color = "black") +
  ggplot2::geom_point(
    size = 2.2,
    color = "#7F3C8D") +
  ggplot2::scale_x_continuous(
    limits = c(0, fs5_xmax),
    breaks = fs5_x_breaks,
    expand = expansion(mult = c(0, 0.01))) +
  ggplot2::facet_grid(
    outcome_label ~ .,
    switch = "y",
    labeller = ggplot2::label_wrap_gen(width = 30)) +
  ggplot2::labs(
    title = "B. Direct pairwise comparison",
    x = "Odds ratio (Higher-Risk vs Lower-Risk)") +
  forest_theme_s5()

#FS5.6 Combine into final Figure S5
figS5 <- p_figS5_A / p_figS5_B +
  patchwork::plot_layout(heights = c(2.2, 1.15)) +
  patchwork::plot_annotation(
    title = "Figure S5. Common-comorbidity sensitivity-analysis odds ratios by wave",
    subtitle = paste0(
      "Odds ratios compare Higher-Risk versus Lower-Risk cluster membership across the diagnostic-group contrasts ",
      "reported in the supplementary results. Horizontal lines indicate 95% confidence intervals; the dashed ",
      "vertical line marks the null value (OR = 1)."),
    theme = theme(
      plot.title = ggplot2::element_text(face = "bold", size = 13),
      plot.subtitle = ggplot2::element_text(size = 10)))

#FS5.7 Save and print Figure S5
save_fig(figS5, "Figure_S5_Common_Comorbidity_Sensitivity_ORs", width = 8.8, height = 7.4)
figS5

```

## Utility of Alternative Baseline Risk Representations Sensitivity-Analysis Inputs

- Purpose: Use the saved comparison-summary and coefficient tables from the alternative baseline-risk representation sensitivity-analysis script
- Inputs: `utility_takeaway`, `dth_utility_coef`, and `prev_utility_coef`
- Outputs: In-memory objects passed directly to Table S7, Table S8, Table S9, and optional Figure S6
- Notes: These objects compare 3 competing baseline-risk representations on identical analytic rows complete for the full 17-variable baseline predictor block, consistent with the supplementary methods

## Table S7 - Relative downstream BSD-related information retained by alternative baseline risk representations in first-onset and occurrence models

- Purpose: Summarize relative downstream information retained by the cluster-label, MVFS, and full 17-variable baseline representations across first-onset and occurrence BSD models
- Inputs: `utility_takeaway`
- Outputs: Saves `Table_S7_Alternative_Baseline_Risk_Representation_Utility` as HTML/PNG/RTF in `results/tables`; returns a gt table
- Notes: Primary metrics are McFadden pseudo-R2 for first-onset models and row-level in-sample AUC for occurrence models; joint block-test p-values correspond to the MVFS and full 17-variable predictor blocks beyond shared covariates

```{r table s7}

## Table S7: Relative downstream BSD-related information retained by alternative baseline risk representations ##

#S7.1 Guard against placeholder input
if (is_tk(utility_takeaway)) {
  stop(
    "Table S7 requires results/main_analysis/2_statistical_analysis/6_risk_vars_vs_cluster_utility/tables/cluster_vs_mvfs_vs_raw_takeaway_table.csv. ",
    "Run 6_risk_vars_vs_cluster_utility.Rmd first, then knit again.")
}

#S7.2 Helper functions
fmt_p_s7 <- function(p) {
  dplyr::case_when(
    is.na(p) ~ "\u2014",
    p < 0.001 ~ "<.001",
    TRUE ~ sprintf("%.3f", round(p, 3)))
}

#S7.3 Prepare publication-ready Table S7 data
tS7_df <- utility_takeaway %>%
  janitor::clean_names() %>%
  dplyr::mutate(
    domain = dplyr::recode(
      domain,
      "DTH" = "Onset",
      "Prevalence" = "Prevalence"),
    outcome = factor(
      outcome,
      levels = c("BD-I", "BD-II", "BD-NOS", "Any BSD")),
    domain = factor(
      domain,
      levels = c("Onset", "Prevalence"))) %>%
  dplyr::arrange(domain, outcome) %>%
  dplyr::transmute(
    `Model Type` = domain,
    `Outcome` = outcome,
    `Primary Metric Type` = primary_metric,
    `Cluster-Label Model Primary Metric Value` = as.numeric(cluster_primary),
    `MVFS 7-Variable Model Primary Metric Value` = as.numeric(mvfs_primary),
    `Full 17-Variable Model Primary Metric Value` = as.numeric(raw_primary),
    `Secondary Metric Type` = secondary_metric,
    `Cluster-Label Model Secondary Metric Value` = as.numeric(cluster_secondary),
    `MVFS 7-Variable Model Secondary Metric Value` = as.numeric(mvfs_secondary),
    `Full 17-Variable Model Secondary Metric Value` = as.numeric(raw_secondary),
    `MVFS 7-Variable Model Block Test P-Value` = fmt_p_s7(mvfs_block_p),
    `Full 17-Variable Model Block Test P-Value` = fmt_p_s7(raw_block_p))

#S7.4 Create caption
tS7_caption <- paste0(
  "**Table S7. Relative downstream BSD-related information retained by alternative baseline risk representations in first-onset and occurrence models**  \n",
  "All 3 baseline-risk representations were compared on identical analytic rows complete for the full 17-variable baseline predictor block. ",
  "Primary descriptive metrics were McFadden pseudo-R2 for first-onset models and row-level in-sample AUC for occurrence models.")

#S7.5 Build gt table
tS7_gt <- gt::gt(tS7_df) |>
  gt::fmt_number(
    columns = c(
      `Cluster-Label Model Primary Metric Value`,
      `MVFS 7-Variable Model Primary Metric Value`,
      `Full 17-Variable Model Primary Metric Value`,
      `Cluster-Label Model Secondary Metric Value`,
      `MVFS 7-Variable Model Secondary Metric Value`,
      `Full 17-Variable Model Secondary Metric Value`),
    decimals = 3) |>
  gt::cols_align(
    align = "center",
    columns = dplyr::everything()) |>
  gt::tab_caption(tS7_caption) |>
  gt::tab_source_note(
    gt::md(paste(
      "For onset models, the secondary metric is the Brier score; for prevalence models, the secondary metric is log loss.",
      "Block-test p-values correspond to joint Wald tests for the MVFS or full 17-variable baseline predictor block beyond shared covariates.",
      "Occurrence-model AUCs are descriptive row-level in-sample summaries and should not be interpreted as internally validated subject-level predictive performance.",
      sep = "  \n"))) |>
  gt::tab_options(table.font.size = gt::px(12))

#S7.6 Save and print Table S7
save_gt(tS7_gt, "Table_S7_Alternative_Baseline_Risk_Representation_Utility")
tS7_gt

```

## Table S8 - Participant-clustered robust-SE adjusted odds ratios for baseline risk representations in discrete-time hazard models of first-onset bipolar spectrum outcomes

- Purpose: Reproduce the substantive predictor rows from the alternative baseline-risk representation discrete-time hazard models
- Inputs: `dth_utility_coef`
- Outputs: Saves `Table_S8_DTH_Alternative_Baseline_Risk_Representation_ORs` as HTML/PNG/RTF in `results/tables`; returns a gt table
- Notes: Displays only the baseline-risk representation terms of interest; nuisance covariates such as sex, wave, age, and intercept terms are omitted for readability

```{r table s8}

## Table S8: Participant-clustered robust-SE adjusted odds ratios for first-onset models ##

#S8.1 Guard against placeholder input
if (is_tk(dth_utility_coef)) {
  stop(
    "Table S8 requires results/main_analysis/2_statistical_analysis/6_risk_vars_vs_cluster_utility/tables/dth_all_model_robust_coefficients.csv. ",
    "Run 6_risk_vars_vs_cluster_utility.Rmd first, then knit again.")
}

#S8.2 Helper functions and ordering objects
utility_outcome_map <- c(
  "bipolar_I" = "BD-I",
  "bipolar_II" = "BD-II",
  "bd_nos" = "BD-NOS",
  "any_bsd" = "Any BSD")

utility_model_map <- c(
  "cluster_model" = "Cluster Label Model",
  "mvfs_model" = "MVFS 7-Variable Model",
  "raw_model" = "Full 17-Variable Model")

utility_term_map <- c(
  "clusterC2" = "Risk Cluster",
  "mh_p_cbcl__dsm__dep_tscore" = "CBCL DSM-5 Depression T",
  "mh_p_cbcl__dsm__anx_tscore" = "CBCL DSM-5 Anxiety T",
  "mh_p_cbcl__synd__attn_tscore" = "CBCL Attention Problems T",
  "mh_p_cbcl__synd__aggr_tscore" = "CBCL Aggression Problems T",
  "mh_p_gbi_sum" = "GBI Mania Sum Score",
  "sds_total" = "SDSC Sum Score",
  "family_history_depression1" = "Family History of Depression",
  "mh_y_upps__nurg_sum" = "UPPS-P Negative Urgency Score",
  "mh_y_upps__purg_sum" = "UPPS-P Positive Urgency Score",
  "le_l_coi__addr1__coi__total__national_zscore" = "COI Z-Score",
  "fc_p_nsc__ns_mean" = "Mean Neighborhood Safety",
  "family_history_mania1" = "Family History of Mania",
  "bullying1" = "Bullying Victimization",
  "nc_y_nihtb__lswmt__uncor_score" = "NIHTB List Sorting Score",
  "nc_y_nihtb__flnkr__uncor_score" = "NIHTB Flanker Score",
  "nc_y_nihtb__pttcp__uncor_score" = "NIHTB Pattern Comparison Score",
  "ACE_index_sum_score" = "ACE Index Sum Score")

mvfs_term_order_s8 <- c(
  "mh_p_cbcl__dsm__dep_tscore",
  "mh_p_cbcl__synd__aggr_tscore",
  "mh_p_cbcl__synd__attn_tscore",
  "mh_p_cbcl__dsm__anx_tscore",
  "mh_p_gbi_sum",
  "sds_total",
  "family_history_depression1")

raw_term_order_s8 <- c(
  "mh_p_cbcl__dsm__dep_tscore",
  "mh_p_cbcl__dsm__anx_tscore",
  "mh_p_cbcl__synd__attn_tscore",
  "mh_p_cbcl__synd__aggr_tscore",
  "mh_p_gbi_sum",
  "mh_y_upps__nurg_sum",
  "mh_y_upps__purg_sum",
  "le_l_coi__addr1__coi__total__national_zscore",
  "fc_p_nsc__ns_mean",
  "sds_total",
  "family_history_depression1",
  "family_history_mania1",
  "bullying1",
  "nc_y_nihtb__lswmt__uncor_score",
  "nc_y_nihtb__flnkr__uncor_score",
  "nc_y_nihtb__pttcp__uncor_score",
  "ACE_index_sum_score")

fmt_or_ci_s8 <- function(or, lo, hi) {
  sprintf("%.3f [%.3f, %.3f]", or, lo, hi)
}

fmt_p_s8 <- function(p) {
  dplyr::case_when(
    is.na(p) ~ "\u2014",
    p < 0.001 ~ "<.001",
    TRUE ~ sprintf("%.3f", round(p, 3)))
}

#S8.3 Prepare publication-ready Table S8 data
tS8_df <- dth_utility_coef %>%
  janitor::clean_names() %>%
  dplyr::filter(
    (model == "cluster_model" & term == "clusterc2") |
      (model == "mvfs_model" & term %in% tolower(mvfs_term_order_s8)) |
      (model == "raw_model" & term %in% tolower(raw_term_order_s8))) %>%
  dplyr::mutate(
    term = dplyr::case_when(
      term == "clusterc2" ~ "clusterC2",
      term == "family_history_depression1" ~ "family_history_depression1",
      term == "family_history_mania1" ~ "family_history_mania1",
      term == "bullying1" ~ "bullying1",
      TRUE ~ term),
    outcome_label = dplyr::recode(outcome, !!!utility_outcome_map),
    model_label = dplyr::recode(model, !!!utility_model_map),
    term_label = dplyr::recode(term, !!!utility_term_map),
    outcome_order = match(outcome, c("bipolar_I", "bipolar_II", "bd_nos", "any_bsd")),
    model_order = match(model, c("cluster_model", "mvfs_model", "raw_model")),
    term_order = dplyr::case_when(
      model == "cluster_model" ~ 1L,
      model == "mvfs_model" ~ match(term, mvfs_term_order_s8),
      model == "raw_model" ~ match(term, raw_term_order_s8),
      TRUE ~ 999L)) %>%
  dplyr::arrange(outcome_order, model_order, term_order) %>%
  dplyr::transmute(
    `Outcome` = outcome_label,
    `Model` = model_label,
    `Variable` = term_label,
    `SE Type` = se_type,
    `SE` = as.numeric(se),
    `OR [95% CI]` = fmt_or_ci_s8(as.numeric(or), as.numeric(ci_lo), as.numeric(ci_hi)),
    `P-Value` = fmt_p_s8(p_value))

#S8.4 Create caption
tS8_caption <- paste0(
  "**Table S8. Participant-clustered robust-SE adjusted odds ratios for baseline risk representations in discrete-time hazard models of first-onset bipolar spectrum outcomes**  \n",
  "Only baseline-risk representation terms are shown. Shared adjustment covariates were sex assigned at birth, interval indicator, and between-person age.")

#S8.5 Build gt table
tS8_gt <- gt::gt(tS8_df) |>
  gt::fmt_number(
    columns = `SE`,
    decimals = 3) |>
  gt::cols_align(
    align = "center",
    columns = c(`Outcome`, `Model`, `Variable`, `SE Type`, `SE`, `OR [95% CI]`, `P-Value`)) |>
  gt::tab_caption(tS8_caption) |>
  gt::tab_source_note(
    gt::md(paste(
      "Cluster-label rows use participant-clustered CR2 robust standard errors.",
      "MVFS and full 17-variable rows use participant-clustered HC1 robust standard errors when CR2 was not estimable in the higher-parameter specifications.",
      sep = "  \n"))) |>
  gt::tab_options(table.font.size = gt::px(12))

#S8.6 Save and print Table S8
save_gt(tS8_gt, "Table_S8_DTH_Alternative_Baseline_Risk_Representation_ORs")
tS8_gt

```

## Table S9 - Robust-SE adjusted odds ratios for baseline risk representations in generalized estimating equation models of bipolar spectrum outcome occurrence

- Purpose: Reproduce the substantive predictor rows from the alternative baseline-risk representation GEE occurrence models
- Inputs: `prev_utility_coef`
- Outputs: Saves `Table_S9_GEE_Alternative_Baseline_Risk_Representation_ORs` as HTML/PNG/RTF in `results/tables`; returns a gt table
- Notes: Displays only the baseline-risk representation terms of interest; nuisance covariates such as wave, sex, baseline outcome status, age, and intercept terms are omitted for readability

```{r table s9}

## Table S9: Robust-SE adjusted odds ratios for occurrence models ##

#S9.1 Guard against placeholder input
if (is_tk(prev_utility_coef)) {
  stop(
    "Table S9 requires results/main_analysis/2_statistical_analysis/6_risk_vars_vs_cluster_utility/tables/prevalence_all_model_coefficients.csv. ",
    "Run 6_risk_vars_vs_cluster_utility.Rmd first, then knit again.")
}

#S9.2 Helper functions
fmt_or_ci_s9 <- function(or, lo, hi) {
  sprintf("%.3f [%.3f, %.3f]", or, lo, hi)
}

fmt_p_s9 <- function(p) {
  dplyr::case_when(
    is.na(p) ~ "\u2014",
    p < 0.001 ~ "<.001",
    TRUE ~ sprintf("%.3f", round(p, 3)))
}

#S9.3 Prepare publication-ready Table S9 data
tS9_df <- prev_utility_coef %>%
  janitor::clean_names() %>%
  dplyr::filter(
    (model == "cluster_model" & term == "clusterc2") |
      (model == "mvfs_model" & term %in% tolower(mvfs_term_order_s8)) |
      (model == "raw_model" & term %in% tolower(raw_term_order_s8))) %>%
  dplyr::mutate(
    term = dplyr::case_when(
      term == "clusterc2" ~ "clusterC2",
      term == "family_history_depression1" ~ "family_history_depression1",
      term == "family_history_mania1" ~ "family_history_mania1",
      term == "bullying1" ~ "bullying1",
      TRUE ~ term),
    outcome_label = dplyr::recode(outcome, !!!utility_outcome_map),
    model_label = dplyr::recode(model, !!!utility_model_map),
    term_label = dplyr::recode(term, !!!utility_term_map),
    outcome_order = match(outcome, c("bipolar_I", "bipolar_II", "bd_nos", "any_bsd")),
    model_order = match(model, c("cluster_model", "mvfs_model", "raw_model")),
    term_order = dplyr::case_when(
      model == "cluster_model" ~ 1L,
      model == "mvfs_model" ~ match(term, mvfs_term_order_s8),
      model == "raw_model" ~ match(term, raw_term_order_s8),
      TRUE ~ 999L)) %>%
  dplyr::arrange(outcome_order, model_order, term_order) %>%
  dplyr::transmute(
    `Outcome` = outcome_label,
    `Model` = model_label,
    `Variable` = term_label,
    `SE` = as.numeric(se),
    `OR [95% CI]` = fmt_or_ci_s9(as.numeric(or), as.numeric(ci_lo), as.numeric(ci_hi)),
    `P-Value` = fmt_p_s9(p_value))

#S9.4 Create caption
tS9_caption <- paste0(
  "**Table S9. Robust-SE adjusted odds ratios for baseline risk representations in generalized estimating equation models of bipolar spectrum outcome occurrence**  \n",
  "Only baseline-risk representation terms are shown. Shared adjustment covariates were wave, sex assigned at birth, baseline outcome status, and grand-mean centered age.")

#S9.5 Build gt table
tS9_gt <- gt::gt(tS9_df) |>
  gt::fmt_number(
    columns = `SE`,
    decimals = 3) |>
  gt::cols_align(
    align = "center",
    columns = c(`Outcome`, `Model`, `Variable`, `SE`, `OR [95% CI]`, `P-Value`)) |>
  gt::tab_caption(tS9_caption) |>
  gt::tab_source_note(
    gt::md(
      "Standard errors are the robust sandwich standard errors returned by the generalized estimating equation models.")) |>
  gt::tab_options(table.font.size = gt::px(12))

#S9.6 Save and print Table S9
save_gt(tS9_gt, "Table_S9_GEE_Alternative_Baseline_Risk_Representation_ORs")
tS9_gt

```

## Figure S6 - Cluster-label effect magnitudes versus strongest single-predictor effect magnitudes across alternative baseline risk representations

- Purpose: Distill the main interpretive insight from the alternative baseline-risk representation sensitivity analysis by comparing the risk-cluster effect size with the strongest single-predictor effect size from the MVFS and full 17-variable models
- Inputs: `dth_utility_coef` and `prev_utility_coef`
- Outputs: Saves `Figure_S6_Cluster_vs_Strongest_Single_Predictor_Effect_Magnitudes.{pdf,png}` in results/figures; returns a ggplot object
- Notes: For the MVFS and full 17-variable models, the plotted point is the single baseline predictor with the largest odds-ratio magnitude relative to 1.0 within that outcome/model, defined as exp(abs(log(OR))); inferential details remain in Tables S8 and S9

```{r figure s6}

## Figure S6: Cluster-label versus strongest single-predictor ORs across alternative baseline risk representations ##

#FS6.1 Guard against placeholder inputs
if (is_tk(dth_utility_coef) || is_tk(prev_utility_coef)) {
  stop(
    "Figure S6 requires results/main_analysis/2_statistical_analysis/6_risk_vars_vs_cluster_utility/tables/",
    "dth_all_model_robust_coefficients.csv and prevalence_all_model_coefficients.csv. ",
    "Run 6_risk_vars_vs_cluster_utility.Rmd first, then knit again.")
}

#FS6.2 Define pretty-label maps
utility_outcome_map_s6 <- c(
  "bipolar_I" = "BD-I",
  "bipolar_II" = "BD-II",
  "bd_nos" = "BD-NOS",
  "any_bsd" = "Any BSD")

utility_term_map_s6 <- c(
  "clusterc2" = "Risk Cluster",
  "mh_p_cbcl__dsm__dep_tscore" = "CBCL DSM-5 Depression T",
  "mh_p_cbcl__dsm__anx_tscore" = "CBCL DSM-5 Anxiety T",
  "mh_p_cbcl__synd__attn_tscore" = "CBCL Attention Problems T",
  "mh_p_cbcl__synd__aggr_tscore" = "CBCL Aggression Problems T",
  "mh_p_gbi_sum" = "GBI Mania Sum Score",
  "sds_total" = "SDSC Sum Score",
  "family_history_depression1" = "Family History of Depression",
  "mh_y_upps__nurg_sum" = "UPPS-P Negative Urgency Score",
  "mh_y_upps__purg_sum" = "UPPS-P Positive Urgency Score",
  "le_l_coi__addr1__coi__total__national_zscore" = "COI Z-Score",
  "fc_p_nsc__ns_mean" = "Mean Neighborhood Safety",
  "family_history_mania1" = "Family History of Mania",
  "bullying1" = "Bullying Victimization",
  "nc_y_nihtb__lswmt__uncor_score" = "NIHTB List Sorting Score",
  "nc_y_nihtb__flnkr__uncor_score" = "NIHTB Flanker Score",
  "nc_y_nihtb__pttcp__uncor_score" = "NIHTB Pattern Comparison Score",
  "ace_index_sum_score" = "ACE Index Sum Score")

#FS6.3 Define the baseline predictor sets
mvfs_terms_s6 <- c(
  "mh_p_cbcl__dsm__dep_tscore",
  "mh_p_cbcl__synd__aggr_tscore",
  "mh_p_cbcl__synd__attn_tscore",
  "mh_p_cbcl__dsm__anx_tscore",
  "mh_p_gbi_sum",
  "sds_total",
  "family_history_depression1")

raw_terms_s6 <- c(
  "mh_p_cbcl__dsm__dep_tscore",
  "mh_p_cbcl__dsm__anx_tscore",
  "mh_p_cbcl__synd__attn_tscore",
  "mh_p_cbcl__synd__aggr_tscore",
  "mh_p_gbi_sum",
  "mh_y_upps__nurg_sum",
  "mh_y_upps__purg_sum",
  "le_l_coi__addr1__coi__total__national_zscore",
  "fc_p_nsc__ns_mean",
  "sds_total",
  "family_history_depression1",
  "family_history_mania1",
  "bullying1",
  "nc_y_nihtb__lswmt__uncor_score",
  "nc_y_nihtb__flnkr__uncor_score",
  "nc_y_nihtb__pttcp__uncor_score",
  "ace_index_sum_score")

#FS6.4 Create helpers to standardize coefficient tables and normalize keys
norm_key_s6 <- function(x) {
  stringr::str_replace_all(
    stringr::str_to_lower(as.character(x)),
    "[^a-z0-9]+",
    "")
}

mvfs_terms_key_s6 <- norm_key_s6(mvfs_terms_s6)
raw_terms_key_s6  <- norm_key_s6(raw_terms_s6)

prep_coef_s6 <- function(df, domain_label) {
  df <- df %>% janitor::clean_names()

  if (!"or" %in% names(df) && "estimate" %in% names(df)) {
    df <- df %>% dplyr::mutate(or = exp(estimate))
  }

  if (!"ci_lo" %in% names(df)) {
    if ("conf_low" %in% names(df)) {
      df <- df %>% dplyr::mutate(ci_lo = conf_low)
    } else if ("ci_low" %in% names(df)) {
      df <- df %>% dplyr::mutate(ci_lo = ci_low)
    } else {
      df <- df %>% dplyr::mutate(ci_lo = NA_real_)
    }
  }

  if (!"ci_hi" %in% names(df)) {
    if ("conf_high" %in% names(df)) {
      df <- df %>% dplyr::mutate(ci_hi = conf_high)
    } else if ("ci_high" %in% names(df)) {
      df <- df %>% dplyr::mutate(ci_hi = ci_high)
    } else {
      df <- df %>% dplyr::mutate(ci_hi = NA_real_)
    }
  }

  df %>%
    dplyr::mutate(
      domain_label = domain_label,
      model_key = norm_key_s6(model),
      term_key = norm_key_s6(term),
      outcome_label = dplyr::recode(outcome, !!!utility_outcome_map_s6, .default = outcome),
      term_label = dplyr::recode(term, !!!utility_term_map_s6, .default = term),
      or = as.numeric(or),
      ci_lo = as.numeric(ci_lo),
      ci_hi = as.numeric(ci_hi),
      effect_abs_log = abs(log(or)))
}

#FS6.5 Extract the cluster effect and strongest single predictor per outcome/model
extract_top_effects_s6 <- function(df, domain_label) {

  df0 <- prep_coef_s6(df, domain_label)

  #FS6.5.1 Keep the cluster-label model row
  clus_df <- df0 %>%
    dplyr::filter(
      stringr::str_detect(model_key, "cluster"),
      term_key %in% c("clusterc2", "riskcluster")) %>%
    dplyr::group_by(outcome_label) %>%
    dplyr::slice_head(n = 1) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      display_group = "Risk Cluster",
      y_pos = 3,
      row_label = "Risk Cluster")

  #FS6.5.2 Keep the strongest MVFS predictor by absolute log-OR magnitude
  mvfs_df <- df0 %>%
    dplyr::filter(
      stringr::str_detect(model_key, "mvfs"),
      term_key %in% mvfs_terms_key_s6) %>%
    dplyr::group_by(outcome_label) %>%
    dplyr::slice_max(order_by = effect_abs_log, n = 1, with_ties = FALSE) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      display_group = "Strongest MVFS Predictor",
      y_pos = 2,
      row_label = paste0(term_label))

  #FS6.5.3 Keep the strongest full-model predictor by absolute log-OR magnitude
  raw_df <- df0 %>%
    dplyr::filter(
      (stringr::str_detect(model_key, "17") |
         stringr::str_detect(model_key, "raw") |
         stringr::str_detect(model_key, "full")) &
        !stringr::str_detect(model_key, "mvfs") &
        !stringr::str_detect(model_key, "cluster"),
      term_key %in% raw_terms_key_s6) %>%
    dplyr::group_by(outcome_label) %>%
    dplyr::slice_max(order_by = effect_abs_log, n = 1, with_ties = FALSE) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      display_group = "Strongest Full-Model Predictor",
      y_pos = 1,
      row_label = paste0(term_label))

  dplyr::bind_rows(clus_df, mvfs_df, raw_df) %>%
    dplyr::mutate(
      row_label_plot = stringr::str_wrap(row_label, width = 16),
      display_group = factor(
        display_group,
        levels = c(
          "Risk Cluster",
          "Strongest MVFS Predictor",
          "Strongest Full-Model Predictor")),
      outcome_label = factor(
        outcome_label,
        levels = c("BD-I", "BD-II", "BD-NOS", "Any BSD")))
}

#FS6.6 Build plotting data
figS6_df <- dplyr::bind_rows(
  extract_top_effects_s6(
    dth_utility_coef,
    "A. First-onset models"),
  extract_top_effects_s6(
    prev_utility_coef,
    "B. Occurrence models"))

#FS6.6.1 Save source data for auditing
readr::write_csv(
  figS6_df %>%
    dplyr::select(
      domain_label,
      outcome_label,
      display_group,
      row_label,
      term,
      term_label,
      or,
      ci_lo,
      ci_hi),
  file.path(tab_dir, "Figure_S6_Cluster_vs_Strongest_Predictor_ORs_SourceData.csv"))

#FS6.7 Set figure bounds and display helpers
fs6_xmax <- max(figS6_df$ci_hi, na.rm = TRUE)
fs6_xmax <- ceiling(fs6_xmax / 0.5) * 0.5
fs6_xmax <- max(fs6_xmax, 5)

#FS6.7.1 Create point-adjacent labels positioned slightly above/right of each OR
fs6_label_dx <- 0.15
fs6_label_xmax_buffer <- 0.50

figS6_df <- figS6_df %>%
  dplyr::mutate(
    row_label_plot = stringr::str_wrap(row_label, width = 16),
    label_x = pmin(or + fs6_label_dx, fs6_xmax - fs6_label_xmax_buffer),
    label_y = y_pos + dplyr::case_when(
      y_pos == 3 ~ 0.18,
      y_pos == 2 ~ 0.15,
      y_pos == 1 ~ 0.18,
      TRUE ~ 0.16))

fs6_cols <- c(
  "Risk Cluster" = "#7F3C8D",
  "Strongest MVFS Predictor" = "#11A579",
  "Strongest Full-Model Predictor" = "#3969AC")

fig_theme_s6 <- function(base_size = 16) {
  ggplot2::theme_classic(base_size = base_size) +
    ggplot2::theme(
      strip.background = ggplot2::element_blank(),
      strip.text = ggplot2::element_text(face = "bold", size = base_size + 0.5),
      axis.title.y = ggplot2::element_blank(),
      axis.text.y = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank(),
      axis.title.x = ggplot2::element_text(size = base_size),
      axis.text.x = ggplot2::element_text(size = base_size - 0.5),
      legend.position = "bottom",
      legend.title = ggplot2::element_blank(),
      legend.text = ggplot2::element_text(size = base_size - 0.5),
      plot.title = ggplot2::element_text(face = "bold", size = base_size + 1.6),
      plot.subtitle = ggplot2::element_text(size = base_size - 0.1),
      panel.spacing.x = grid::unit(1.2, "lines"),
      panel.spacing.y = grid::unit(1.0, "lines"),
      plot.margin = ggplot2::margin(10, 14, 10, 14))
}

#FS6.8 Build the figure
figS6 <- ggplot2::ggplot(
  figS6_df,
  ggplot2::aes(
    x = or,
    y = y_pos,
    color = display_group)) +
  ggplot2::geom_vline(
    xintercept = 1,
    linetype = "dashed",
    linewidth = 0.5,
    color = "grey45") +
  ggplot2::geom_segment(
    ggplot2::aes(x = ci_lo, xend = ci_hi, yend = y_pos),
    linewidth = 0.7) +
  ggplot2::geom_point(
    size = 2.8) +
  ggplot2::geom_label(
    data = figS6_df,
    ggplot2::aes(
      x = label_x,
      y = label_y,
      label = row_label_plot),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 0,
    lineheight = 0.90,
    size = 4.0,
    color = "black",
    fill = "white",
    label.size = NA,
    label.padding = grid::unit(0.08, "lines")) +
  ggplot2::facet_grid(
    domain_label ~ outcome_label) +
  ggplot2::scale_x_continuous(
    breaks = seq(0, fs6_xmax, by = 1),
    expand = ggplot2::expansion(mult = c(0.02, 0.03))) +
  ggplot2::scale_y_continuous(
    limits = c(0.4, 3.75),
    breaks = c(1, 2, 3),
    labels = NULL) +
  ggplot2::scale_color_manual(
    values = fs6_cols,
    breaks = names(fs6_cols),
    labels = c(
      "Risk Cluster",
      "Strongest MVFS Predictor",
      "Strongest Full-Model Predictor")) +
  ggplot2::coord_cartesian(
    xlim = c(0, fs6_xmax),
    clip = "on") +
  ggplot2::labs(
    title = "Figure S6. Cluster-label odds ratios versus strongest single-predictor odds ratios",
    subtitle = paste0(
      "Within each outcome, the adjusted higher- versus lower-risk cluster odds ratio is shown alongside the single ",
      "baseline predictor with the largest absolute log-OR magnitude from the MVFS and full 17-variable models. ",
      "Horizontal bars denote 95% confidence intervals."),
    x = "Odds Ratio") +
  fig_theme_s6()

#FS6.9 Save and print Figure S6
save_fig(figS6, "Figure_S6_Cluster_vs_Strongest_Predictor_ORs", width = 14.75, height = 9.75, dpi = 1440)

figS6

```
